### 2023-09-28 pldi24/paper.tex
Initial commit for paper

%%
%% This is file `sample-acmsmall-submission.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall-submission')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall-submission.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen,review]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%%
%% These commands are for a JOURNAL article.
\acmJournal{JACM}
\acmVolume{37}
\acmNumber{4}
\acmArticle{111}
\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Denotational Semantics for SSA with Weak Memory Operations}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Neel Krishnaswami}
\email{nk480@cl.cam.ac.uk}
\orcid{0000-0003-2838-5865}

\author{Jad Ghalayini}
\email{ik404@cl.cam.ac.uk}
\orcid{0000-0002-6905-1303}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>00000000.0000000.0000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Type Inference, System F, Call-by-Push-Value, Polarized Typing, Focalisation, Subtypingr}

\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
ACM's consolidated article template, introduced in 2017, provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. Numerous ACM and
SIG-specific \LaTeX\ templates have been examined, and their unique
features incorporated into this single new template.

If you are new to publishing with ACM, this document is a valuable
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-blind initial submission of a full-length technical paper, a
two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
journal article, a SIGCHI Extended Abstract, and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

\subsection{Template Styles}

The primary parameter given to the ``\verb|acmart|'' document class is
the {\itshape template style} which corresponds to the kind of publication
or SIG publishing the work. This parameter is enclosed in square
brackets and is a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[STYLE]{acmart}
\end{verbatim}

Journals use one of three template styles. All but three ACM journals
use the {\verb|acmsmall|} template style:
\begin{itemize}
\item {\texttt{acmsmall}}: The default journal template style.
\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
\item {\texttt{acmtog}}: Used by TOG.
\end{itemize}

The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
\begin{itemize}
\item {\texttt{acmconf}}: The default proceedings template style.
\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
\end{itemize}

\subsection{Template Parameters}

In addition to specifying the {\itshape template style} to be used in
formatting your work, there are a number of {\itshape template parameters}
which modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \texttt{\acmSubmissionID} command to print the
  submission's unique ID on each page of the work.
\item{\texttt{authorversion}}: Produces a version of the work suitable
  for posting by the author.
\item{\texttt{screen}}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[acmsmall,screen,review]{acmart}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification.  As an exception, multiple authors may share one
affiliation. Authors' names should not be abbreviated; use full first
names wherever possible. Include authors' e-mail addresses whenever
possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail
alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note
to apply to multiple authors --- for example, if the first two authors
of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of
the list of authors to be used in the page headers, to prevent
overlapping text. The following command should be placed just after
the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all
of the authors' names, which may result in overlapping text in the
page headers.

The article template's documentation, available at
\url{https://www.acm.org/publications/proceedings-template}, has a
complete explanation of these commands and tips for their effective
use.

Note that authors' addresses are mandatory for journal articles.

\section{Rights Information}

Authors of any work published by ACM will need to complete a rights
form. Depending on the kind of work, and the rights management choice
made by the author, this may be copyright transfer, permission,
license, or an OA (open access) agreement.

Regardless of the rights management choice, the author will receive a
copy of the completed rights form once it has been submitted. This
form contains \LaTeX\ commands that must be copied into the source
document. When the document source is compiled, these commands and
their parameters add formatted text to several areas of the final
document:
\begin{itemize}
\item the ``ACM Reference Format'' text on the first page.
\item the ``rights management'' text on the first page.
\item the conference information in the page header(s).
\end{itemize}

Rights information is unique to the work; if you are preparing several
works for an event, make sure to use the correct set of commands with
each of the works.

The ACM Reference Format text is required for all articles over one
page in length, and is optional for one-page articles (abstracts).

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful
taxonomic tools for you to help readers find your work in an online
search.

The ACM Computing Classification System ---
\url{https://www.acm.org/publications/class-2012} --- is a set of
classifiers and concepts that describe the computing
discipline. Authors can select entries from this classification
system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases
of the authors' choosing, providing a more flexible way of describing
the research being presented.

CCS concepts and user-defined keywords are required for for all
articles over two pages in length, and are optional for one- and
two-page articles (or abstracts).

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \textbf{math} environment,
which can be invoked with the usual
\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
    Ewing, Inc. [Public domain], via Wikimedia
    Commons. (\url{https://goo.gl/VLCRBB}).}
  \Description{A woman and a girl in white dresses sit in an open car.}
\end{figure}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that
are placed after all author and affiliation information, and before
the body of the article, spanning the page. If you wish to have such a
figure in your article, place the command immediately before the
\verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

Citations and references are numbered by default. A small number of
ACM publications have citations and references formatted in the
``author year'' style; for these exceptions, please include this
command in the {\bfseries preamble} (before the command
``\verb|\begin{document}|'') of your \LaTeX\ source:
\begin{verbatim}
  \citestyle{acmauthoryear}
\end{verbatim}

  Some examples.  A paginated journal article \cite{Abril07}, an
  enumerated journal article \cite{Cohen07}, a reference to an entire
  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
  monograph/whole book in a series (see 2a in spec. document)
  \cite{Harel79}, a divisible-book such as an anthology or compilation
  \cite{Editor00} followed by the same example, however we only output
  the series if the volume number is given \cite{Editor00a} (so
  Editor00a's series should NOT be present since it has no vol. no.),
  a chapter in a divisible book \cite{Spector90}, a chapter in a
  divisible book in a series \cite{Douglass98}, a multi-volume work as
  book \cite{Knuth97}, a couple of articles in a proceedings (of a
  conference, symposium, workshop for example) (paginated proceedings
  article) \cite{Andler79, Hagerup1993}, a proceedings article with
  all possible elements \cite{Smith10}, an example of an enumerated
  proceedings article \cite{VanGundy07}, an informally published work
  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
  master's thesis: \cite{anisi03}, an online document / world wide web
  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
  publication \cite{rous08}, 'YYYYb'-test for prolific author
  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
  contain 'duplicate' DOI and URLs (some SIAM articles)
  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
  couple of citations with DOIs:
  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
  Artifacts: \cite{R} and \cite{UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\verb|\end{document}|'' command at the conclusion of your source
document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
  \appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not
numbered. This document has two appendices, demonstrating the section
and subsection identification method.

\section{Multi-language papers}

Papers may be written in languages other than English or include
titles, subtitles, keywords and abstracts in different languages (as a
rule, a paper in a language other than English should include an
English title and an English abstract).  Use \verb|language=...| for
every language used in the paper.  The last language indicated is the
main language of the paper.  For example, a French paper with
additional titles and abstracts in English and German may start with
the following command
\begin{verbatim}
\documentclass[sigconf, language=english, language=german,
               language=french]{acmart}
\end{verbatim}

The title, subtitle, keywords and abstract will be typeset in the main
language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
begin title, subtitle and keywords, can be used to set these elements
in the other languages.  The environment \verb|translatedabstract| is
used to set the translation of the abstract.  These commands and
environment have a mandatory first argument: the language of the
second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
of their usage.

\section{SIGCHI Extended Abstracts}

The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
not in Word) produces a landscape-orientation formatted article, with
a wide left margin. Three environments are available for use with the
``\verb|sigchi-a|'' template style, and produce formatted output in
the margin:
\begin{description}
\item[\texttt{sidebar}:]  Place formatted text in the margin.
\item[\texttt{marginfigure}:] Place a figure in the margin.
\item[\texttt{margintable}:] Place a table in the margin.
\end{description}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\section{Online Resources}

Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
enim maximus. Vestibulum gravida massa ut felis suscipit
congue. Quisque mattis elit a risus ultrices commodo venenatis eget
dui. Etiam sagittis eleifend elementum.

Nam interdum magna at lectus dignissim, ac dignissim lorem
rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
massa et mattis lacinia.

\end{document}
\endinput
%%
%% End of file `sample-acmsmall-submission.tex'.

### 2023-10-02 pldi24/paper.tex
Tap

@@
-%%
-%% This is file `sample-acmsmall-submission.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `acmsmall-submission')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-acmsmall-submission.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
 %% Commands for TeXCount
@@
 
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
 %% Rights management information.  This information is sent to you
@@
 %% These commands are for a JOURNAL article.
-\acmJournal{JACM}
-\acmVolume{37}
-\acmNumber{4}
-\acmArticle{111}
-\acmMonth{8}
+% \acmJournal{JACM}
+% \acmVolume{37}
+% \acmNumber{4}
+% \acmArticle{111}
+% \acmMonth{8}
 
@@
 \author{Jad Ghalayini}
-\email{ik404@cl.cam.ac.uk}
+\email{jeg74@cl.cam.ac.uk}
 \orcid{0000-0002-6905-1303}
@@
 %% of authors' names for this purpose.
-\renewcommand{\shortauthors}{Trovato et al.}
+% \renewcommand{\shortauthors}{Trovato et al.}
 
@@
 \begin{abstract}
-  A clear and well-documented \LaTeX\ document is presented as an
-  article formatted for publication by ACM in a conference proceedings
-  or journal publication. Based on the ``acmart'' document class, this
-  article presents and explains many of the common variations, as well
-  as many of the formatting elements an author may use in the
-  preparation of the documentation of their work.
+  TODO THIS
 \end{abstract}
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{Type Inference, System F, Call-by-Push-Value, Polarized Typing, Focalisation, Subtypingr}
+\keywords{TODO PUT KEYWORDS HERE}
 
-\received{20 February 2007}
-\received[revised]{12 March 2009}
-\received[accepted]{5 June 2009}
+% \received{20 February 2007}
+% \received[revised]{12 March 2009}
+% \received[accepted]{5 June 2009}
 
@@
 \section{Introduction}
-ACM's consolidated article template, introduced in 2017, provides a
-consistent \LaTeX\ style for use across ACM publications, and
-incorporates accessibility and metadata-extraction functionality
-necessary for future Digital Library endeavors. Numerous ACM and
-SIG-specific \LaTeX\ templates have been examined, and their unique
-features incorporated into this single new template.
 
-If you are new to publishing with ACM, this document is a valuable
-guide to the process of preparing your work for publication. If you
-have published with ACM before, this document provides insight and
-instruction into more recent changes to the article template.
-
-The ``\verb|acmart|'' document class can be used to prepare articles
-for any ACM publication --- conference or journal, and for any stage
-of publication, from review to final ``camera-ready'' copy, to the
-author's own version, with {\itshape very} few changes to the source.
-
-\section{Template Overview}
-As noted in the introduction, the ``\verb|acmart|'' document class can
-be used to prepare many different kinds of documentation --- a
-double-blind initial submission of a full-length technical paper, a
-two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
-journal article, a SIGCHI Extended Abstract, and more --- all by
-selecting the appropriate {\itshape template style} and {\itshape
-  template parameters}.
-
-This document will explain the major features of the document
-class. For further information, the {\itshape \LaTeX\ User's Guide} is
-available from
-\url{https://www.acm.org/publications/proceedings-template}.
-
-\subsection{Template Styles}
-
-The primary parameter given to the ``\verb|acmart|'' document class is
-the {\itshape template style} which corresponds to the kind of publication
-or SIG publishing the work. This parameter is enclosed in square
-brackets and is a part of the {\verb|documentclass|} command:
-\begin{verbatim}
-  \documentclass[STYLE]{acmart}
-\end{verbatim}
-
-Journals use one of three template styles. All but three ACM journals
-use the {\verb|acmsmall|} template style:
-\begin{itemize}
-\item {\texttt{acmsmall}}: The default journal template style.
-\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
-\item {\texttt{acmtog}}: Used by TOG.
-\end{itemize}
-
-The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
-\begin{itemize}
-\item {\texttt{acmconf}}: The default proceedings template style.
-\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
-\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
-\end{itemize}
-
-\subsection{Template Parameters}
-
-In addition to specifying the {\itshape template style} to be used in
-formatting your work, there are a number of {\itshape template parameters}
-which modify some part of the applied template style. A complete list
-of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}
-
-Frequently-used parameters, or combinations of parameters, include:
-\begin{itemize}
-\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
-  conference submission. Anonymizes the work and includes line
-  numbers. Use with the \texttt{\acmSubmissionID} command to print the
-  submission's unique ID on each page of the work.
-\item{\texttt{authorversion}}: Produces a version of the work suitable
-  for posting by the author.
-\item{\texttt{screen}}: Produces colored hyperlinks.
-\end{itemize}
-
-This document uses the following string as the first command in the
-source file:
-\begin{verbatim}
-\documentclass[acmsmall,screen,review]{acmart}
-\end{verbatim}
-
-\section{Modifications}
-
-Modifying the template --- including but not limited to: adjusting
-margins, typeface sizes, line spacing, paragraph and list definitions,
-and the use of the \verb|\vspace| command to manually adjust the
-vertical spacing between elements of your work --- is not allowed.
-
-{\bfseries Your document will be returned to you for revision if
-  modifications are discovered.}
-
-\section{Typefaces}
-
-The ``\verb|acmart|'' document class requires the use of the
-``Libertine'' typeface family. Your \TeX\ installation should include
-this set of packages. Please do not substitute other typefaces. The
-``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
-as they will override the built-in typeface families.
-
-\section{Title Information}
-
-The title of your work should use capital letters appropriately -
-\url{https://capitalizemytitle.com/} has useful rules for
-capitalization. Use the {\verb|title|} command to define the title of
-your work. If your work has a subtitle, define it with the
-{\verb|subtitle|} command.  Do not insert line breaks in your title.
-
-If your title is lengthy, you must define a short version to be used
-in the page headers, to prevent overlapping text. The \verb|title|
-command has a ``short title'' parameter:
-\begin{verbatim}
-  \title[short title]{full title}
-\end{verbatim}
-
-\section{Authors and Affiliations}
-
-Each author must be defined separately for accurate metadata
-identification.  As an exception, multiple authors may share one
-affiliation. Authors' names should not be abbreviated; use full first
-names wherever possible. Include authors' e-mail addresses whenever
-possible.
-
-Grouping authors' names or e-mail addresses, or providing an ``e-mail
-alias,'' as shown below, is not acceptable:
-\begin{verbatim}
-  \author{Brooke Aster, David Mehldau}
-  \email{dave,judy,steve@university.edu}
-  \email{firstname.lastname@phillips.org}
-\end{verbatim}
-
-The \verb|authornote| and \verb|authornotemark| commands allow a note
-to apply to multiple authors --- for example, if the first two authors
-of an article contributed equally to the work.
-
-If your author list is lengthy, you must define a shortened version of
-the list of authors to be used in the page headers, to prevent
-overlapping text. The following command should be placed just after
-the last \verb|\author{}| definition:
-\begin{verbatim}
-  \renewcommand{\shortauthors}{McCartney, et al.}
-\end{verbatim}
-Omitting this command will force the use of a concatenated list of all
-of the authors' names, which may result in overlapping text in the
-page headers.
-
-The article template's documentation, available at
-\url{https://www.acm.org/publications/proceedings-template}, has a
-complete explanation of these commands and tips for their effective
-use.
-
-Note that authors' addresses are mandatory for journal articles.
-
-\section{Rights Information}
-
-Authors of any work published by ACM will need to complete a rights
-form. Depending on the kind of work, and the rights management choice
-made by the author, this may be copyright transfer, permission,
-license, or an OA (open access) agreement.
-
-Regardless of the rights management choice, the author will receive a
-copy of the completed rights form once it has been submitted. This
-form contains \LaTeX\ commands that must be copied into the source
-document. When the document source is compiled, these commands and
-their parameters add formatted text to several areas of the final
-document:
-\begin{itemize}
-\item the ``ACM Reference Format'' text on the first page.
-\item the ``rights management'' text on the first page.
-\item the conference information in the page header(s).
-\end{itemize}
-
-Rights information is unique to the work; if you are preparing several
-works for an event, make sure to use the correct set of commands with
-each of the works.
-
-The ACM Reference Format text is required for all articles over one
-page in length, and is optional for one-page articles (abstracts).
-
-\section{CCS Concepts and User-Defined Keywords}
-
-Two elements of the ``acmart'' document class provide powerful
-taxonomic tools for you to help readers find your work in an online
-search.
-
-The ACM Computing Classification System ---
-\url{https://www.acm.org/publications/class-2012} --- is a set of
-classifiers and concepts that describe the computing
-discipline. Authors can select entries from this classification
-system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
-commands to be included in the \LaTeX\ source.
-
-User-defined keywords are a comma-separated list of words and phrases
-of the authors' choosing, providing a more flexible way of describing
-the research being presented.
-
-CCS concepts and user-defined keywords are required for for all
-articles over two pages in length, and are optional for one- and
-two-page articles (or abstracts).
-
-\section{Sectioning Commands}
-
-Your work should use standard \LaTeX\ sectioning commands:
-\verb|section|, \verb|subsection|, \verb|subsubsection|, and
-\verb|paragraph|. They should be numbered; do not remove the numbering
-from the commands.
-
-Simulating a sectioning command by setting the first word or words of
-a paragraph in boldface or italicized text is {\bfseries not allowed.}
-
-\section{Tables}
-
-The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
-package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
-high-quality tables.
-
-Table captions are placed {\itshape above} the table.
-
-Because tables cannot be split across pages, the best placement for
-them is typically the top of the page nearest their initial cite.  To
-ensure this proper ``floating'' placement of tables, use the
-environment \textbf{table} to enclose the table's contents and the
-table caption.  The contents of the table itself must go in the
-\textbf{tabular} environment, to be aligned properly in rows and
-columns, with the desired horizontal and vertical rules.  Again,
-detailed instructions on \textbf{tabular} material are found in the
-\textit{\LaTeX\ User's Guide}.
-
-Immediately following this sentence is the point at which
-Table~\ref{tab:freq} is included in the input file; compare the
-placement of the table here with the table in the printed output of
-this document.
-
-\begin{table}
-  \caption{Frequency of Special Characters}
-  \label{tab:freq}
-  \begin{tabular}{ccl}
-    \toprule
-    Non-English or Math&Frequency&Comments\\
-    \midrule
-    \O & 1 in 1,000& For Swedish names\\
-    $\pi$ & 1 in 5& Common in math\\
-    \$ & 4 in 5 & Used in business\\
-    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-  \bottomrule
-\end{tabular}
-\end{table}
-
-To set a wider table, which takes up the whole width of the page's
-live area, use the environment \textbf{table*} to enclose the table's
-contents and the table caption.  As with a single-column table, this
-wide table will ``float'' to a location deemed more
-desirable. Immediately following this sentence is the point at which
-Table~\ref{tab:commands} is included in the input file; again, it is
-instructive to compare the placement of the table here with the table
-in the printed output of this document.
-
-\begin{table*}
-  \caption{Some Typical Commands}
-  \label{tab:commands}
-  \begin{tabular}{ccl}
-    \toprule
-    Command &A Number & Comments\\
-    \midrule
-    \texttt{{\char'134}author} & 100& Author \\
-    \texttt{{\char'134}table}& 300 & For tables\\
-    \texttt{{\char'134}table*}& 400& For wider tables\\
-    \bottomrule
-  \end{tabular}
-\end{table*}
-
-Always use midrule to separate table header rows from data rows, and
-use it only for this purpose. This enables assistive technologies to
-recognise table headers and support their users in navigating tables
-more easily.
-
-\section{Math Equations}
-You may want to display math equations in three distinct styles:
-inline, numbered or non-numbered display.  Each of the three are
-discussed in the next sections.
-
-\subsection{Inline (In-text) Equations}
-A formula that appears in the running text is called an inline or
-in-text formula.  It is produced by the \textbf{math} environment,
-which can be invoked with the usual
-\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
-the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
-and structures, from $\alpha$ to $\omega$, available in
-\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
-examples of in-text equations in context. Notice how this equation:
-\begin{math}
-  \lim_{n\rightarrow \infty}x=0
-\end{math},
-set here in in-line math style, looks slightly different when
-set in display style.  (See next section).
-
-\subsection{Display Equations}
-A numbered display equation---one set off by vertical space from the
-text and centered horizontally---is produced by the \textbf{equation}
-environment. An unnumbered display equation is produced by the
-\textbf{displaymath} environment.
-
-Again, in either environment, you can use any of the symbols and
-structures available in \LaTeX\@; this section will just give a couple
-of examples of display equations in context.  First, consider the
-equation, shown as an inline equation above:
-\begin{equation}
-  \lim_{n\rightarrow \infty}x=0
-\end{equation}
-Notice how it is formatted somewhat differently in
-the \textbf{displaymath}
-environment.  Now, we'll enter an unnumbered equation:
-\begin{displaymath}
-  \sum_{i=0}^{\infty} x + 1
-\end{displaymath}
-and follow it with another numbered equation:
-\begin{equation}
-  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
-\end{equation}
-just to demonstrate \LaTeX's able handling of numbering.
-
-\section{Figures}
-
-The ``\verb|figure|'' environment should be used for figures. One or
-more images can be placed within a figure. If your figure contains
-third-party material, you must clearly identify it as such, as shown
-in the example below.
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{sample-franklin}
-  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
-    Ewing, Inc. [Public domain], via Wikimedia
-    Commons. (\url{https://goo.gl/VLCRBB}).}
-  \Description{A woman and a girl in white dresses sit in an open car.}
-\end{figure}
-
-Your figures should contain a caption which describes the figure to
-the reader.
-
-Figure captions are placed {\itshape below} the figure.
-
-Every figure should also have a figure description unless it is purely
-decorative. These descriptions convey what’s in the image to someone
-who cannot see it. They are also used by search engine crawlers for
-indexing images, and when images cannot be loaded.
-
-A figure description must be unformatted plain text less than 2000
-characters long (including spaces).  {\bfseries Figure descriptions
-  should not repeat the figure caption – their purpose is to capture
-  important information that is not already provided in the caption or
-  the main text of the paper.} For figures that convey important and
-complex new information, a short text description may not be
-adequate. More complex alternative descriptions can be placed in an
-appendix and referenced in a short figure description. For example,
-provide a data table capturing the information in a bar chart, or a
-structured list representing a graph.  For additional information
-regarding how best to write figure descriptions and why doing this is
-so important, please see
-\url{https://www.acm.org/publications/taps/describing-figures/}.
-
-\subsection{The ``Teaser Figure''}
-
-A ``teaser figure'' is an image, or set of images in one figure, that
-are placed after all author and affiliation information, and before
-the body of the article, spanning the page. If you wish to have such a
-figure in your article, place the command immediately before the
-\verb|\maketitle| command:
-\begin{verbatim}
-  \begin{teaserfigure}
-    \includegraphics[width=\textwidth]{sampleteaser}
-    \caption{figure caption}
-    \Description{figure description}
-  \end{teaserfigure}
-\end{verbatim}
-
-\section{Citations and Bibliographies}
-
-The use of \BibTeX\ for the preparation and formatting of one's
-references is strongly recommended. Authors' names should be complete
---- use full first names (``Donald E. Knuth'') not initials
-(``D. E. Knuth'') --- and the salient identifying features of a
-reference should be included: title, year, volume, number, pages,
-article DOI, etc.
-
-The bibliography is included in your source document with these two
-commands, placed just before the \verb|\end{document}| command:
-\begin{verbatim}
-  \bibliographystyle{ACM-Reference-Format}
-  \bibliography{bibfile}
-\end{verbatim}
-where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
-suffix, of the \BibTeX\ file.
-
-Citations and references are numbered by default. A small number of
-ACM publications have citations and references formatted in the
-``author year'' style; for these exceptions, please include this
-command in the {\bfseries preamble} (before the command
-``\verb|\begin{document}|'') of your \LaTeX\ source:
-\begin{verbatim}
-  \citestyle{acmauthoryear}
-\end{verbatim}
-
-  Some examples.  A paginated journal article \cite{Abril07}, an
-  enumerated journal article \cite{Cohen07}, a reference to an entire
-  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
-  monograph/whole book in a series (see 2a in spec. document)
-  \cite{Harel79}, a divisible-book such as an anthology or compilation
-  \cite{Editor00} followed by the same example, however we only output
-  the series if the volume number is given \cite{Editor00a} (so
-  Editor00a's series should NOT be present since it has no vol. no.),
-  a chapter in a divisible book \cite{Spector90}, a chapter in a
-  divisible book in a series \cite{Douglass98}, a multi-volume work as
-  book \cite{Knuth97}, a couple of articles in a proceedings (of a
-  conference, symposium, workshop for example) (paginated proceedings
-  article) \cite{Andler79, Hagerup1993}, a proceedings article with
-  all possible elements \cite{Smith10}, an example of an enumerated
-  proceedings article \cite{VanGundy07}, an informally published work
-  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
-    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
-  master's thesis: \cite{anisi03}, an online document / world wide web
-  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
-  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
-  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
-  publication \cite{rous08}, 'YYYYb'-test for prolific author
-  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
-  contain 'duplicate' DOI and URLs (some SIAM articles)
-  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
-  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
-  couple of citations with DOIs:
-  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
-  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
-  Artifacts: \cite{R} and \cite{UMassCitations}.
-
-\section{Acknowledgments}
-
-Identification of funding sources and other support, and thanks to
-individuals and groups that assisted in the research and the
-preparation of the work should be included in an acknowledgment
-section, which is placed just before the reference section in your
-document.
-
-This section has a special environment:
-\begin{verbatim}
-  \begin{acks}
-  ...
-  \end{acks}
-\end{verbatim}
-so that the information contained therein can be more easily collected
-during the article metadata extraction phase, and to ensure
-consistency in the spelling of the section heading.
-
-Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.
-
-\section{Appendices}
-
-If your work needs an appendix, add it before the
-``\verb|\end{document}|'' command at the conclusion of your source
-document.
-
-Start the appendix with the ``\verb|appendix|'' command:
-\begin{verbatim}
-  \appendix
-\end{verbatim}
-and note that in the appendix, sections are lettered, not
-numbered. This document has two appendices, demonstrating the section
-and subsection identification method.
-
-\section{Multi-language papers}
-
-Papers may be written in languages other than English or include
-titles, subtitles, keywords and abstracts in different languages (as a
-rule, a paper in a language other than English should include an
-English title and an English abstract).  Use \verb|language=...| for
-every language used in the paper.  The last language indicated is the
-main language of the paper.  For example, a French paper with
-additional titles and abstracts in English and German may start with
-the following command
-\begin{verbatim}
-\documentclass[sigconf, language=english, language=german,
-               language=french]{acmart}
-\end{verbatim}
-
-The title, subtitle, keywords and abstract will be typeset in the main
-language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
-begin title, subtitle and keywords, can be used to set these elements
-in the other languages.  The environment \verb|translatedabstract| is
-used to set the translation of the abstract.  These commands and
-environment have a mandatory first argument: the language of the
-second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
-of their usage.
-
-\section{SIGCHI Extended Abstracts}
-
-The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
-not in Word) produces a landscape-orientation formatted article, with
-a wide left margin. Three environments are available for use with the
-``\verb|sigchi-a|'' template style, and produce formatted output in
-the margin:
-\begin{description}
-\item[\texttt{sidebar}:]  Place formatted text in the margin.
-\item[\texttt{marginfigure}:] Place a figure in the margin.
-\item[\texttt{margintable}:] Place a table in the margin.
-\end{description}
-
-%%
-%% The acknowledgments section is defined using the "acks" environment
-%% (and NOT an unnumbered section). This ensures the proper
-%% identification of the section in the article metadata, and the
-%% consistent spelling of the heading.
-\begin{acks}
-To Robert, for the bagels and explaining CMYK and color spaces.
-\end{acks}
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{sample-base}
-
-%%
-%% If your work has an appendix, this is the place to put it.
-\appendix
-
-\section{Research Methods}
-
-\subsection{Part One}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
-malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
-sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
-vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
-lacinia dolor. Integer ultricies commodo sem nec semper.
-
-\subsection{Part Two}
-
-Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
-ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
-ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
-eros. Vivamus non purus placerat, scelerisque diam eu, cursus
-ante. Etiam aliquam tortor auctor efficitur mattis.
-
-\section{Online Resources}
-
-Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
-pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
-enim maximus. Vestibulum gravida massa ut felis suscipit
-congue. Quisque mattis elit a risus ultrices commodo venenatis eget
-dui. Etiam sagittis eleifend elementum.
-
-Nam interdum magna at lectus dignissim, ac dignissim lorem
-rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
-massa et mattis lacinia.
+TODO THIS
 
@@
 \endinput
-%%
-%% End of file `sample-acmsmall-submission.tex'.

### 2023-10-03 pldi24/paper.tex
Tap
Initial paper blocking

@@
 %TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
+
 \documentclass[acmsmall,screen,review]{acmart}
+
+\usepackage{syntax}
+\renewcommand{\syntleft}{\normalfont\itshape}
+\renewcommand{\syntright}{\normalfont\itshape}
+
+\usepackage{prftree}
+
+\newcounter{todos}
+\newcommand{\TODO}[1]{{
+  \stepcounter{todos}
+  \begin{center}\large{\textcolor{red}{\textbf{TODO \arabic{todos}:} #1}}\end{center}
+}}
+
+% Math fonts
+\newcommand{\mc}[1]{\ensuremath{\mathcal{#1}}}
+\newcommand{\mb}[1]{\ensuremath{\mathbf{#1}}}
+\newcommand{\ms}[1]{\ensuremath{\mathsf{#1}}}
+
+% Syntax atoms
+\newcommand{\lbl}[1]{{`#1}}
+\newcommand{\lto}{\Rightarrow}
+\newcommand{\ctt}{\ms{tt}}
+\newcommand{\cff}{\ms{ff}}
+
+% Syntax
+\newcommand{\letexpr}[3]{\ensuremath{\ms{let}\;#1 = #2\;\ms{in}\;#3}}
+\newcommand{\letstmt}[3]{\ensuremath{\ms{let}\;#1 = #2; #3}}
+\newcommand{\jmpb}[2]{\ms{jmp}\;#1\;#2}
+\newcommand{\ljmpb}[2]{\jmpb{\lbl{#1}}{#2}}
+\newcommand{\ite}[3]{\ms{if}\;#1\;\{#2\}\;\ms{else}\;\{#3\}}
+\newcommand{\ewhere}[2]{\ms{then}\;#1\;\ms{where}\;#2}
+\newcommand{\where}[2]{#1\;\ms{where}\;#2}
+\newcommand{\wbranch}[3]{#1(#2) \lto #3}
+\newcommand{\lwbranch}[3]{\wbranch{\lbl{#1}}{#2}{#3}}
+\newcommand{\bsplice}[3]{#1(#2)\;\{#3\}}
+\newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
+
+% Branding
+\newcommand{\isotopessa}{\ms{isotope_{SSA}}}
 
@@
 %%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
 %% The majority of ACM publications use numbered citations and
@@
 
-%%
-%% end of the preamble, start of the body of the document source.
 \begin{document}
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
 \title{Denotational Semantics for SSA with Weak Memory Operations}
 
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
 \author{Neel Krishnaswami}
@@
 
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-% \renewcommand{\shortauthors}{Trovato et al.}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
 \begin{abstract}
@@
 
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
 \maketitle
@@
 
+\section{SSA Syntax}
+
+\TODO{text}
+
+\begin{figure}
+  \begin{center}
+    \begin{grammar}
+      <\(a, b, c, e\)> ::= \(x\) 
+      \;|\; \(f\;a\)
+      \;|\; \((a, b)\) 
+      \;|\; \(()\) 
+      \;|\; \(\ctt\) 
+      \;|\; \(\cff\)
+      \;|\; \(\letexpr{x}{a}{e}\)
+      \;|\; \(\letexpr{(x, y)}{a}{e}\)
+      \;|\; \(\lbsplice{\ell}{x: A}{t}\)
+
+      <\(s, t\)> ::= \(\ljmpb{\ell}{a}\) 
+      \;|\; \(\ite{e}{s}{t}\)
+      \;|\; \(\letstmt{x}{a}{t}\)
+      \;|\; \(\letstmt{(x, y)}{a}{t}\)
+      \;|\; \(\ewhere{t}{L}\)
+
+      <\(L\)> ::= \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{t}, L\)
+
+      <\(\Gamma\)> ::= \(\cdot\) \;|\; \(\Gamma, x: A\)
+
+      <\(\ms{L}\)> ::= \(\cdot\) \;|\; \(\ms{L}, \lbl{\ell}[\Gamma](x: A)\)
+    \end{grammar}
+  \end{center}
+  \caption{Grammar for \isotopessa}
+  \Description{Grammar for isotope-SSA}
+  \label{fig:ssa-grammar}
+\end{figure}
+
+\TODO{late \ms{where}-binding, and other sugar (?)}
+
+\TODO{top-level functions}
+
+\TODO{table of typing judgements}
+
+\TODO{contexts, label-contexts}
+
+\begin{figure}
+  \begin{center}        
+    \begingroup
+    \renewcommand{\arraystretch}{1.5}
+    \setlength{\tabcolsep}{2em}
+    \begin{tabular}{rl}
+        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
+        \(\Gamma \vdash_p a: A\) &
+        \(a\) is a term of type \(A\) in context \(\Gamma\) with purity \(p \in \{0, 1\}\) \\
+        \(\Gamma \vdash t \rhd \ms{L}\) &
+        \(t\) is a block targeting label-set \(\ms{L}\) in context \(\Gamma\) \\
+        \(\ms{L} \vdash L \rhd \ms{K}\) &
+        The labels \(L\) send label-set \(\ms{L}\) to label-set \(\ms{K}\)
+    \end{tabular}
+    \endgroup
+  \end{center}
+  \caption{Typing judgements for \isotopessa}
+  \Description{Typing judgements for isotope-SSA}
+  \label{fig:ssa-judgements}
+\end{figure}
+
+\begin{figure}
+  \begin{center}    
+    \TODO{this}
+  \end{center}
+  \caption{Typing rules for \isotopessa}
+  \Description{Typing rules for isotope-SSA}
+  \label{fig:ssa-typing}
+\end{figure}
+
+\section{SSA Semantics}
+
+\TODO{premonoidal categories}
+
+\TODO{Elgot structure}
+
+\begin{figure}
+  \begin{center}
+    \TODO{this}
+  \end{center}
+  \caption{Semantics for \isotopessa}
+  \Description{Semantics for isotope-SSA}
+  \label{fig:ssa-semantics}
+\end{figure}
+
+\TODO{metatheory: substitution, rewriting, etc}
+\TODO{\(\implies\) E-graph optimization}
+
+\section{Basic Models}
+
+\TODO{Ye Olde Trace Monad}
+
+\TODO{Ye Olde State Transformer}
+
+\TODO{Ye Olde Nondeterministic Trace Monad}
+
+\section{Weak Memory}
+
+\TODO{Pomsets}
+
+\TODO{SC Monad}
+
+\TODO{Building The Weak Memory Monad}
+
+\TODO{Karoubi envelope as premonoidal demonstration?}
+
+\section{Implementation}
+
+\TODO{this}
+
+\section{Related Work}
+
+\TODO{\cite{promonad}}
+
+\TODO{\cite{linear-state-usage}}
+
+\TODO{\cite{ssa-is-fun}}
+
+\TODO{\cite{sparky}}
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{references}
+
 \end{document}

### 2023-10-04 pldi24/paper.tex
Began copying over typing rules
Copied over typing rules

@@
 \newcommand{\letstmt}[3]{\ensuremath{\ms{let}\;#1 = #2; #3}}
-\newcommand{\jmpb}[2]{\ms{jmp}\;#1\;#2}
-\newcommand{\ljmpb}[2]{\jmpb{\lbl{#1}}{#2}}
+\newcommand{\brb}[2]{\ms{br}\;#1\;#2}
+\newcommand{\lbrb}[2]{\brb{\lbl{#1}}{#2}}
 \newcommand{\ite}[3]{\ms{if}\;#1\;\{#2\}\;\ms{else}\;\{#3\}}
@@
 \newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
+\newcommand{\csplits}[3]{#1 \mapsto #2;#3}
+\newcommand{\cwk}[2]{#1 \mapsto #2}
+\newcommand{\lwk}[2]{#1 \rightsquigarrow #2}
+\newcommand{\tlin}[2]{\ms{lin}_{#2}(#1)}
+\newcommand{\ltlin}[3]{\ms{lin}_{#3}(#1^{#2})}
+\newcommand{\thyp}[3]{#1: {#2}^{#3}}
+\newcommand{\lhyp}[3]{#1[#2](#3)}
+\newcommand{\llhyp}[3]{\lhyp{\lbl{#1}}{#2}{#3}}
+\newcommand{\rle}[1]{{\scriptsize\textsf{#1}}}
+\newcommand{\taff}{\ms{a}}
+\newcommand{\trel}{\ms{r}}
+\newcommand{\tint}{\infty}
+\newcommand{\hasty}[4]{#1 \vdash_{#2} #3: #4}
+\newcommand{\haslb}[3]{#1 \vdash #2 \rhd #3}
+\newcommand{\lhaslb}[3]{#1 \vdash #2 \rhd #3}
 
@@
     \begin{grammar}
+      <\(A, B, C\)> ::= \(X\)
+      \;|\; \(A \otimes B\)
+
       <\(a, b, c, e\)> ::= \(x\) 
@@
 
-      <\(s, t\)> ::= \(\ljmpb{\ell}{a}\) 
+      <\(s, t\)> ::= \(\lbrb{\ell}{a}\) 
       \;|\; \(\ite{e}{s}{t}\)
@@
 
-      <\(\Gamma\)> ::= \(\cdot\) \;|\; \(\Gamma, x: A\)
+      <\(\Gamma\)> ::= \(\cdot\) \;|\; \(\Gamma, \thyp{x}{A}{q}\)
 
@@
         \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
-        \(\Gamma \vdash_p a: A\) &
+        \(\hasty{\Gamma}{p}{x}{A}\) &
         \(a\) is a term of type \(A\) in context \(\Gamma\) with purity \(p \in \{0, 1\}\) \\
-        \(\Gamma \vdash t \rhd \ms{L}\) &
+        \(\haslb{\Gamma}{t}{\ms{L}}\) &
         \(t\) is a block targeting label-set \(\ms{L}\) in context \(\Gamma\) \\
-        \(\ms{L} \vdash L \rhd \ms{K}\) &
-        The labels \(L\) send label-set \(\ms{L}\) to label-set \(\ms{K}\)
+        \(\lhaslb{\ms{L}}{L}{\ms{K}}\) &
+        The labels \(L\) send label-set \(\ms{L}\) to label-set \(\ms{K}\) \\
+        \(\csplits{\Gamma}{\Delta}{\Xi}\) &
+        The context \(\Gamma\) splits into \(\Delta\) and \(\Xi\) \\
+        \(\tlin{A}{q}\) &
+        The type \(A\) has linearity \(q\) \\
+        \(\ltlin{A}{r}{q}\) &
+        \(\tlin{A}{q} \land q \subseteq r\) \\
+        \(\tlin{\Gamma}{q}\) &
+        The context \(\Gamma\) has linearity \(q\) \\
+        \(\cwk{\Gamma}{\Delta}\) &
+        \(\Gamma\) is a weakening of \(\Delta\) (i.e. \(\csplits{\Gamma}{\Delta}{\cdot}\)) \\
+        \(\lwk{\ms{L}}{\ms{K}}\) &
+        The label-set \(\ms{L}\) weakens to the label-set \(\ms{K}\)
     \end{tabular}
@@
 \begin{figure}
-  \begin{center}    
-    \TODO{this}
-  \end{center}
-  \caption{Typing rules for \isotopessa}
-  \Description{Typing rules for isotope-SSA}
-  \label{fig:ssa-typing}
+  \begin{gather*}    
+    \prftree[r]{\rle{base-lin}}{q \subseteq \ms{lin}(X)}{\tlin{X}{q}} \qquad
+    \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} \qquad
+    \prftree[r]{\rle{nil-lin}}{\tlin{\cdot}{q}} \qquad
+    \prftree[r]{\rle{cons-lin}}{\ltlin{A}{r}{q}}{\tlin{\Gamma}{q}}
+      {\tlin{\Gamma, \thyp{x}{A}{r}}{q}} \\
+    \prftree[r]{\rle{split-nil}}{\csplits{\cdot}{\cdot}{\cdot}} \qquad
+    \prftree[r]{\rle{split-left}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {r \subseteq q}
+      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta, \thyp{x}{A}{r}}{\Xi}} \qquad
+    \prftree[r]{\rle{split-right}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {r \subseteq q}
+      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta}{\Xi, \thyp{x}{A}{r}}} \\
+    \prftree[r]{\rle{split-dup}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\ltlin{A}{q}{\trel}}
+      {r, s \subseteq q}
+      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta, \thyp{x}{A}{r}}{\Xi, \thyp{x}{A}{s}}}
+      \qquad
+    \prftree[r]{\rle{split-drop}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\ltlin{A}{q}{\taff}}
+      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta}{\Xi}}
+      \\
+    \prftree[r]{\rle{join-nil}}{\lwk{\cdot}{\cdot}} \qquad
+    \prftree[r]{\rle{join-cons}}
+      {\lwk{\ms{L}}{\ms{K}}}
+      {\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}} 
+      \qquad
+    \prftree[r]{\rle{join-zero}}
+      {\lwk{\ms{L}}{\ms{K}}}
+      {\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}} 
+  \end{gather*}
+  \caption{Structural rules for \isotopessa}
+  \Description{Structural rules for isotope-SSA}
+  \label{fig:ssa-structural}
+\end{figure}
+
+\begin{figure}
+  \begin{gather*}    
+    \prftree[r]{\rle{var}}
+      {\cwk{\Gamma}{\thyp{x}{A}{q}}}
+      {\hasty{\Gamma}{p}{x}{A}} \qquad
+    \prftree[r]{\rle{app}}
+      {f \in \mc{I}_p(A, B)}
+      {\hasty{\Gamma}{1}{a}{A}}
+      {\hasty{\Gamma}{p}{f\;a}{B}} \qquad
+    \prftree[r]{\rle{pair}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\hasty{\Delta}{1}{a}{A}}
+      {\hasty{\Xi}{1}{b}{B}}
+      {\hasty{\Gamma}{1}{(a, b)}{A \otimes B}} \\
+    \prftree[r]{\rle{unit}}
+      {\cwk{\Gamma}{\cdot}}
+      {\hasty{\Gamma}{1}{()}{\mb{1}}} \qquad
+    \prftree[r]{\rle{true}}
+      {\cwk{\Gamma}{\cdot}}
+      {\hasty{\Gamma}{1}{\ctt}{\mb{2}}} \qquad
+    \prftree[r]{\rle{false}}
+      {\cwk{\Gamma}{\cdot}}
+      {\hasty{\Gamma}{1}{\cff}{\mb{2}}} \\
+    \prftree[r]{\rle{let}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}}
+      {\hasty{\Xi}{1}{a}{A}}
+      {\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}} \qquad
+    \prftree[r]{\rle{blk}}
+      {\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}}
+      {\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}} \\
+    \prftree[r]{\rle{let2}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}}
+      {\hasty{\Xi}{1}{a}{A \otimes B}}
+      {\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}}
+  \end{gather*}
+  \caption{Typing rules for \isotopessa terms}
+  \Description{Typing rules for isotope-SSA terms}
+  \label{fig:ssa-term-typing}
+\end{figure}
+
+\begin{figure}
+  \begin{gather*}    
+    \prftree[r]{\rle{br}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
+      %TODO: consider allowing binding here, or should it be purely for lets?
+      {\hasty{\Xi}{1}{a}{A}}
+      {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} \qquad
+    \prftree[r]{\rle{ite}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\hasty{\Delta}{1}{e}{\mb{2}}}
+      {\haslb{\Xi}{s}{\ms{L}}}
+      {\haslb{\Xi}{t}{\ms{L}}}
+      {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} \\
+    \prftree[r]{\rle{let-blk}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
+      {\hasty{\Xi}{p}{a}{A}}
+      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} \\
+    \prftree[r]{\rle{let2-blk}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
+      {\hasty{\Xi}{p}{a}{A \otimes B}}
+      {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \\
+    \prftree[r]{\rle{where}}
+      {\haslb{\Gamma}{t}{\ms{L}}}
+      {\lhaslb{\ms{L}}{L}{\ms{K}}}
+      {\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}} \qquad
+    \prftree[r]{\rle{nil-br}}
+      {\lwk{\ms{L}}{\ms{K}}}
+      {\lhaslb{\ms{L}}{\cdot}{\ms{K}}} \qquad
+    \prftree[r]{\rle{cons-br}}
+      {\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
+      {\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}
+      {\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
+  \end{gather*}
+  \caption{Typing rules for \isotopessa blocks}
+  \Description{Typing rules for isotope-SSA blocks}
+  \label{fig:ssa-block-typing}
 \end{figure}

### 2023-10-05 pldi24/paper.tex
Tap
foo

@@
 
-TODO THIS
+\TODO{THIS}
+
+\paragraph{Contributions}
+
+\begin{itemize}
+\item We have a syntax and type system for SSA-style programs.
+We include support for substructural types, and our type-theoretic syntax makes formulating inlining and rewriting easier than in the traditional CFG-based formulations.
+\item We show that our language has a compositional intepretation in any effectful category with an Elgot structure.
+We use this structure to give generic proofs of the expected substitution properties as well as prove the soundness of an equational theory for the language.
+\item Our generic equational theory justifies a variety of powerful optimisation techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and strength reduction. 
+\item We show how Elgot monads give rise to models of our language.
+We use this to derive semantics from a general trace monad, and show that various monad transformers such as state transformers preserve the needed Elgot structure.
+\item We use these tools to demonstrate the existence of a variety of concrete models satisfying these categorical axioms.
+Starting first from simple languges such as state plus printing, we also show how more challenging semantics such as the TSO semantics for weak memory also fit into our framework.
+This thus gives rise to an SSA-based IR with support for weak memory operations, which is fully semantically-justified.
+\item \TODO{We also have a small implementation, which \ldots}
+\item Our weak memory semantics and our results about Elgot monads are formalised in the Lean theorem prover. \TODO{Other stuff}
+\end{itemize}
 
@@
         The context \(\Gamma\) splits into \(\Delta\) and \(\Xi\) \\
+        \(\lwk{\ms{L}}{\ms{K}}\) &
+        The label-set \(\ms{L}\) weakens to the label-set \(\ms{K}\) \\
         \(\tlin{A}{q}\) &
@@
         \(\ltlin{A}{r}{q}\) &
-        \(\tlin{A}{q} \land q \subseteq r\) \\
+        The type \(A\) has linearity \(q \subseteq r\)
+        (i.e. \(\tlin{A}{q} \land q \subseteq r\)) \\
         \(\tlin{\Gamma}{q}\) &
@@
         \(\cwk{\Gamma}{\Delta}\) &
-        \(\Gamma\) is a weakening of \(\Delta\) (i.e. \(\csplits{\Gamma}{\Delta}{\cdot}\)) \\
-        \(\lwk{\ms{L}}{\ms{K}}\) &
-        The label-set \(\ms{L}\) weakens to the label-set \(\ms{K}\)
+        \(\Gamma\) is a weakening of \(\Delta\) 
+        (i.e. \(\csplits{\Gamma}{\Delta}{\cdot}\))
     \end{tabular}

### 2023-10-10 pldi24/paper.tex
Added more stubs

@@
 
+\begin{figure}
+  \TODO{this}
+  \caption{Typing rules for \isotopessa substitutions}
+  \Description{Typing rules for isotope-SSA substitutions}
+  \label{fig:ssa-block-typing}
+\end{figure}
+
+\begin{theorem}[Substitution] \
+  \TODO{this}
+\end{theorem}
+
 \section{SSA Semantics}
@@
 
-\TODO{metatheory: substitution, rewriting, etc}
+\begin{theorem}[Semantic Substitution] \
+  \TODO{this}
+\end{theorem}
+
 \TODO{\(\implies\) E-graph optimization}
+
+\begin{figure}
+  \begin{center}
+    \TODO{this}
+  \end{center}
+  \caption{Grammar for \isotopessa expressions with holes}
+  \Description{Grammar for isotope-SSA expressions with holes}
+  \label{fig:ssa-semantics}
+\end{figure}
+
+\begin{theorem}[Rewriting] \
+  \TODO{this}
+\end{theorem}
 
@@
 
-\TODO{Ye Olde State Transformer}
+\TODO{Is Elgot}
 
@@
 
+\TODO{Is Elgot}
+
+\TODO{Ye Olde Reader/Writer/State Transformer}
+
 \section{Weak Memory}
 
-\TODO{Pomsets}
+\TODO{Pomsets...}
 
-\TODO{SC Monad}
+\TODO{Form a monoid, so we can define...}
 
-\TODO{Building The Weak Memory Monad}
+\TODO{The SC Monad, which, via composition w/ the state transformer, lets us}
+
+\TODO{Build The Weak Memory Monad, which is too powerful, so we do...}
 

### 2023-10-11 pldi24/paper.tex
Work on structural judgements, stubs

@@
 \newcommand{\lhaslb}[3]{#1 \vdash #2 \rhd #3}
+
+% Denotational semantics
+\newcommand{\dnt}[1]{\llbracket{#1}\rrbracket}
+\newcommand{\ednt}[1]{\left\llbracket{#1}\right\rrbracket}
+\newcommand{\upg}[1]{{#1}^\uparrow}
 
@@
   \Description{Typing rules for isotope-SSA substitutions}
-  \label{fig:ssa-block-typing}
+  \label{fig:ssa-subst-typing}
 \end{figure}
@@
 \begin{figure}
-  \begin{center}
+  \begin{gather*}
+    \boxed{\dnt{A}}: |\mc{C}_1| \\
+    \dnt{X} = \ms{base}(X) 
+      \text{ where } 
+      \dnt{\mb{1}} = \mb{1}, 
+      \dnt{\mb{2}} = \mb{2}
+    \qquad \dnt{A \otimes B} = \dnt{A} \otimes \dnt{B} \\
+    \boxed{\dnt{\Gamma}}: |\mc{C}_1| \\
+    \dnt{\cdot} = \mb{1} \qquad 
+    \dnt{\Gamma, \thyp{x}{A}{q}} = \dnt{\Gamma} \otimes \dnt{A} \\
+    \boxed{\dnt{\ms{L}}}: |\mc{C}_0| \\
+    \dnt{\cdot} = \mb{0} \qquad
+    \dnt{\ms{L}, \llhyp{\ell}{\Gamma}{A}} = 
+      \dnt{\ms{L}} + \dnt{\Gamma} \otimes \dnt{A}
+  \end{gather*}
+  \caption{Semantics for \isotopessa types and contexts}
+  \Description{Semantics for isotope-SSA types and contexts}
+  \label{fig:ssa-type-semantics}
+\end{figure}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{\dnt{\tlin{A}{\taff}}: \mc{C}_1(\dnt{A}, \mb{1})} \\
+    \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
+    \dnt{\tlin{A \otimes B}{\taff}} 
+      = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
+      ; \lambda \\
+    \boxed{\dnt{\tlin{A}{\trel}}}
+      : \mc{C}_1(\dnt{A}, \dnt{A} \otimes \dnt{A}) \\
+    \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
+    \dnt{\tlin{A \otimes B}{\trel}} = 
+      \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
+      ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha \\
+    \boxed{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
+      : \mc{C}_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi}) \\
+    \dnt{\csplits{\cdot}{\cdot}{\cdot}} = \lambda^{-1} \qquad
+    \dnt{\csplits
+      {\Gamma, \thyp{x}{A}{q}}
+      {\Delta, \thyp{x}{A}{q}}
+      {\Xi}} 
+      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha;\dnt{\Delta} \otimes \sigma;\alpha \\
+    \dnt{\csplits
+      {\Gamma, \thyp{x}{A}{q}}
+      {\Delta}
+      {\Xi, \thyp{x}{A}{q}}} 
+      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha \\
+    \dnt{\csplits
+      {\Gamma, \thyp{x}{A}{q}}
+      {\Delta}
+      {\Xi, \thyp{x}{A}{q}}} \
+      = \dnt{\Gamma} \otimes \dnt{\tlin{A}{\trel}};
+        \alpha;
+        \dnt{\Delta} \otimes \sigma \otimes \dnt{A};
+        \alpha \\
+    \dnt{\csplits
+      {\Gamma, \thyp{x}{A}{q}}
+      {\Delta}
+      {\Xi}}
+      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+        \otimes \dnt{\tlin{A}{\taff}}
+      ; \rho \\
+    \boxed{\dnt{\cwk{\Gamma}{\Delta}}
+      : \mc{C}_1(\dnt{\Gamma}, \dnt{\Delta})} \\
+      = \dnt{\csplits{\Gamma}{\Delta}{\cdot}};\rho \\
+    \boxed{\dnt{\lwk{\ms{L}}{\ms{K}}}
+      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} \\
+      \dnt{\lwk{\cdot}{\cdot}} = \ms{id} \qquad
+      \dnt{\lwk
+        {\ms{L}, \llhyp{\ell}{\Gamma}{A}}
+        {\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
+        = \dnt{\lwk{\ms{L}}{\ms{K}}} + \dnt{\Gamma} \otimes \dnt{A} \\
+      \dnt{\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
+        = \dnt{\lwk{\ms{L}}{\ms{K}}};\rho_+^{-1}
+        ; \dnt{\ms{K}} \oplus 0_{\dnt{\Gamma} \otimes \dnt{A}}
+  \end{gather*}
+  \caption{Semantics for \isotopessa structural judgements}
+  \Description{Semantics for isotope-SSA structural judgements}
+  \label{fig:ssa-structural-semantics}
+\end{figure}
+
+\begin{figure}
   \TODO{this}
-  \end{center}
-  \caption{Semantics for \isotopessa}
-  \Description{Semantics for isotope-SSA}
-  \label{fig:ssa-semantics}
+  \caption{Semantics for \isotopessa terms}
+  \Description{Semantics for isotope-SSA terms}
+  \label{fig:ssa-structural-terms}
+\end{figure}
+
+\begin{figure}
+  \TODO{this}
+  \caption{Semantics for \isotopessa blocks}
+  \Description{Semantics for isotope-SSA blocks}
+  \label{fig:ssa-structural-blocks}
 \end{figure}
@@
   \end{center}
-  \caption{Grammar for \isotopessa expressions with holes}
-  \Description{Grammar for isotope-SSA expressions with holes}
+  \caption{Grammar for \isotopessa terms and blocks with holes}
+  \Description{Grammar for isotope-SSA terms and blocks with holes}
   \label{fig:ssa-semantics}

### 2023-10-12 pldi24/paper.tex
Term semantics
Block semantics
Statements for coherence and semantic upgrade, rules and statement for substitution
Began work on semantic substitution; forgot about label substitutions...
Typo
Tap

@@
 }}
+\newcommand{\sorry}{\textcolor{red}{\textbf{sorry}}}
 
@@
 \newcommand{\lhaslb}[3]{#1 \vdash #2 \rhd #3}
+\newcommand{\issubst}[3]{#1: #2 \mapsto #3}
+\newcommand{\substcons}[3]{[{#2 \mapsto #3}]{#1}}
 
@@
 \newcommand{\ednt}[1]{\left\llbracket{#1}\right\rrbracket}
-\newcommand{\upg}[1]{{#1}^\uparrow}
+\newcommand{\upg}[2]{{#1}^{\uparrow #2}}
 
@@
       {\lwk{\ms{L}}{\ms{K}}}
-      {\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}} 
+      {\cwk{\Gamma}{\Delta}}
+      {\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}} 
       \qquad
@@
       {\hasty{\Xi}{1}{b}{B}}
-      {\hasty{\Gamma}{1}{(a, b)}{A \otimes B}} \\
+      {\hasty{\Gamma}{p}{(a, b)}{A \otimes B}} \\
     \prftree[r]{\rle{unit}}
       {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{1}{()}{\mb{1}}} \qquad
+      {\hasty{\Gamma}{p}{()}{\mb{1}}} \qquad
     \prftree[r]{\rle{true}}
       {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{1}{\ctt}{\mb{2}}} \qquad
+      {\hasty{\Gamma}{p}{\ctt}{\mb{2}}} \qquad
     \prftree[r]{\rle{false}}
       {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{1}{\cff}{\mb{2}}} \\
+      {\hasty{\Gamma}{p}{\cff}{\mb{2}}} \\
     \prftree[r]{\rle{let}}
@@
       %TODO: consider allowing binding here, or should it be purely for lets?
-      {\hasty{\Xi}{1}{a}{A}}
+      {\hasty{\Xi}{0}{a}{A}}
       {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} \qquad
@@
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta}{1}{e}{\mb{2}}}
+      {\hasty{\Delta}{0}{e}{\mb{2}}}
       {\haslb{\Xi}{s}{\ms{L}}}
@@
       {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
-      {\hasty{\Xi}{p}{a}{A}}
+      {\hasty{\Xi}{1}{a}{A}}
       {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} \\
@@
       {\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
-      {\hasty{\Xi}{p}{a}{A \otimes B}}
+      {\hasty{\Xi}{1}{a}{A \otimes B}}
       {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \\
@@
     \prftree[r]{\rle{nil-br}}
-      {\lwk{\ms{L}}{\ms{K}}}
-      {\lhaslb{\ms{L}}{\cdot}{\ms{K}}} \qquad
+      {\lhaslb{\ms{L}}{\cdot}{\ms{L}}} \qquad
     \prftree[r]{\rle{cons-br}}
@@
 \begin{figure}
-  \TODO{this}
+  \begin{gather*}
+    \prftree[r]{\rle{subset-nil}}
+      {\cdot \subseteq \cdot}
+      \qquad
+    \prftree[r]{\rle{subset-cons}}
+      {\Gamma \subseteq \Delta}
+      {\Gamma, \thyp{x}{A}{q} \subseteq \Delta, \thyp{x}{A}{q}}
+      \qquad
+    \prftree[r]{\rle{subset-drop}}
+      {\Gamma \subseteq \Delta}
+      {\Gamma \subseteq \Delta, \thyp{x}{A}{q}} 
+    \\
+    \prftree[r]{\rle{in-head}}
+      {\thyp{x}{A}{q} \in \Gamma, \thyp{x}{A}{q}}
+      \qquad
+    \prftree[r]{\rle{in-cons}}
+      {\thyp{x}{A}{q} \in \Gamma}
+      {\thyp{x}{A}{q} \in \Gamma, \thyp{y}{B}{r}}
+    \\
+    \prftree[r]{\rle{var-head}}
+      {x \in \Gamma, \thyp{x}{A}{q}}
+      \qquad
+    \prftree[r]{\rle{var-cons}}
+      {x \in \Gamma}
+      {x \in \Gamma, \thyp{y}{B}{r}} 
+    \\
+    \Gamma \cup \cdot = \Gamma \qquad
+    \Gamma \cup (\Delta, \thyp{x}{A}{q}) = (\Gamma \cup \Delta), \thyp{x}{A}{q} \text{ if } x \notin \Gamma \qquad
+    \Gamma \cup (\Delta, \thyp{x}{A}{q}) = \Gamma \cup \Delta \text{ otherwise} 
+    % \\
+    %   \cdot^\Gamma_\Delta = \cdot 
+    %   \qquad
+    %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta
+    %     = \Theta^\Gamma_\Delta, 
+    %       \thyp{x}{A}{q} \text{ if } x \in \Delta \lor x \notin \Gamma
+    %   \qquad
+    %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta = \Theta^\Gamma_\Delta \text{ otherwise}
+    % \\
+    %   \cdot^\Gamma_\Delta = \cdot 
+    %   \qquad
+    %   (\ms{L}, \llhyp{\ell}{\Theta}{A})^\Gamma_\Delta
+    %   = \ms{L}^\Gamma_\Delta, \llhyp{\ell}{\Theta^\Gamma_\Delta}{A}
+  \end{gather*}
+  \caption{Manipulation of \isotopessa contexts}
+  \Description{Manipulation of isotope-SSA contexts}
+  \label{fig:context-manipulation}
+\end{figure}
+
+\begin{theorem}[Weakening] \
+  \begin{itemize}
+    \item If \(\cwk{\Gamma}{\Delta}\), \(\hasty{\Delta}{p}{a}{A}\), then \(\hasty{\Gamma}{p}{a}{A}\)
+    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
+    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then \(\haslb{\Gamma}{t}{\ms{K}}\)
+  \end{itemize}
+\end{theorem}
+
+\begin{theorem}[Minimal Context] \
+  For all \(a, A\), there exists \(\Delta\) such that \(\forall \Gamma, \hasty{\Gamma}{p}{a}{A} \implies \cwk{\Gamma}{\Delta} \land \hasty{\Delta}{p}{a}{A}\).
+If such a \(\Gamma\) exists, then \(\Delta\) is unique; in this case we call \(\Delta\) \textbf{minimal}.
+\end{theorem}
+
+\begin{figure}
+  \begin{gather*}
+    \prftree[r]{\rle{subst-nil}}
+      {\issubst{\cdot}{\cdot}{\cdot}}
+      \qquad
+    \prftree[r]{\rle{subst-cons}}
+      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
+      {\hasty{\Theta_x}{0}{a}{A} \text{ minimal}}
+      {\tlin{\Theta_x}{q}}
+      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
+      {\issubst{\substcons{\gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
+  \end{gather*}
   \caption{Typing rules for \isotopessa substitutions}
@@
 
-\begin{theorem}[Substitution] \
+\TODO{label substitution}
+
+\begin{theorem}[Substitution Weakening] \
   \TODO{this}
+\end{theorem}
+
+\begin{theorem}[Substitution] 
+  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  \begin{itemize}
+    \item If \(\hasty{\Gamma}{p}{a}{A}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}\)
+    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Theta}{[\gamma]t}{[\gamma]\ms{L}}\)
+  \end{itemize}
 \end{theorem}
@@
         {\ms{L}, \llhyp{\ell}{\Gamma}{A}}
-        {\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-        = \dnt{\lwk{\ms{L}}{\ms{K}}} + \dnt{\Gamma} \otimes \dnt{A} \\
+        {\ms{K}, \llhyp{\ell}{\Delta}{A}}}
+        = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\cwk{\Gamma}{\Delta} \otimes \dnt{A})}{} \\
       \dnt{\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
@@
 \begin{figure}
-  \TODO{this}
+  \begin{gather*}
+    \boxed{\dnt{\hasty{\Gamma}{p}{a}{A}}
+      : \mc{C}_p(\dnt{\Gamma}, \dnt{A})} \\
+    \dnt{\hasty{\Gamma}{p}{x}{A}} 
+      = \dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}
+      \qquad
+    \dnt{\hasty{\Gamma}{p}{f\;a}{B}}
+      = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}}}{p}
+      ; \ms{inst}_p(f) \\
+    \dnt{\hasty{\Gamma}{p}{(a, b)}{A \otimes B}}
+      = \upg{(
+        \dnt{\csplits{\Gamma}{\Delta}{\Xi}};
+        \dnt{\hasty{\Delta}{1}{a}{A}} \otimes
+        \dnt{\hasty{\Xi}{1}{b}{B}}
+      )}{\uparrow p} \\
+    \dnt{\hasty{\Gamma}{p}{()}{\mb{1}}}
+      = \upg{\dnt{\cwk{\Gamma}{\cdot}}}{p} 
+      \qquad
+    \dnt{\hasty{\Gamma}{p}{\ctt}{\mb{2}}}
+      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\ctt)}{p}
+      \qquad
+    \dnt{\hasty{\Gamma}{p}{\cff}{\mb{2}}}
+      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\cff)}{p} \\
+    \dnt{\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}}
+      = \upg{(
+        \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}}
+      )}{p};\dnt{\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}} \\
+    \dnt{\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}}
+      = \upg{(
+        \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A \otimes B}}
+      )}{p}
+      ; 
+      \\ \qquad \qquad \qquad \qquad \qquad \alpha
+      ; \dnt{\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}}
+      \\
+    \dnt{\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}}
+      = \dnt{\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}};\alpha_+
+  \end{gather*}
   \caption{Semantics for \isotopessa terms}
@@
 \begin{figure}
-  \TODO{this}
+  \begin{gather*}
+    \boxed{\dnt{\haslb{\Gamma}{t}{\ms{L}}}
+      : \mc{C}_1(\dnt{\Gamma}, \dnt{\ms{L}})} \\
+    \dnt{\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}}
+      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{0}{a}{A}})}
+      ; \dnt{\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}} \\
+    \dnt{\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}}
+      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+      ; \dnt{\hasty{\Delta}{0}{e}{\mb{2}}} \otimes \dnt{\Xi})}
+      ; \ms{ite}_{\dnt{\Xi}} ;
+      \\ \qquad \qquad \qquad \qquad
+      [
+        \dnt{\haslb{\Xi}{s}{\ms{L}}}, 
+        \dnt{\haslb{\Xi}{t}{\ms{L}}} 
+      ]
+      \\
+    \dnt{\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}}
+      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A}}
+      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
+      \\
+    \dnt{\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}}
+      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
+        ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A \otimes B}}
+        ; \alpha
+        ; \dnt{\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
+      \\
+    \dnt{\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}}
+      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+      ; \dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger
+      \\
+    \boxed{\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}
+      : \mc{C}_1(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
+      \\
+    \dnt{\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
+      = \dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{A}{t}}}
+      ; [[\iota_0, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{L}}], \iota_1]
+      \\
+    \dnt{\lhaslb{\ms{L}}{\cdot}{\ms{L}}}
+      = \iota_0
+  \end{gather*}
   \caption{Semantics for \isotopessa blocks}
@@
 
-\begin{theorem}[Semantic Substitution] \
-  \TODO{this}
+\begin{theorem}[Coherence]
+  \begin{itemize}
+    \item Given any two derivations \(D_1, D_2: \hasty{\Gamma}{p}{a}{A}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{theorem}
+
+\begin{theorem}[Semantic Upgrade]
+  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{0}{a}{A}\), then
+  \[\dnt{\hasty{\Gamma}{1}{a}{A}} = \upg{\hasty{\Gamma}{0}{a}{A}}{}\]
+\end{theorem}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_0(\dnt{\Theta}, \dnt{\Gamma})} \\
+    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id} \qquad
+    \dnt{\issubst{\substcons{\Gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
+      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{0}{a}{A}}
+  \end{gather*}
+  \caption{Semantics for \isotopessa substitutions}
+  \Description{Semantics for isotope-SSA substitutions}
+  \label{fig:ssa-subst-semantics}
+\end{figure}
+
+\begin{theorem}[Semantic Substitution] 
+  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  \begin{itemize}
+    \item \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
+      ;\dnt{\hasty{\Gamma}{p}{a}{A}} 
+      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}}\)
+    \item \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
+      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+      = \dnt{\haslb{\Theta}{[\gamma]t}{[\gamma]\ms{L}}}
+      ; \sorry \)
+  \end{itemize}
 \end{theorem}
@@
   \end{center}
-  \caption{Grammar for \isotopessa terms and blocks with holes}
-  \Description{Grammar for isotope-SSA terms and blocks with holes}
-  \label{fig:ssa-semantics}
+  \caption{Grammar for \isotopessa blocks with holes}
+  \Description{Grammar for isotope-SSA blocks with holes}
+  \label{fig:blocks-with-holes}
 \end{figure}

### 2023-10-13 pldi24/paper.tex
Added linearity tags to typing rules

@@
 \newcommand{\tint}{\infty}
-\newcommand{\hasty}[4]{#1 \vdash_{#2} #3: #4}
+\newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
 \newcommand{\haslb}[3]{#1 \vdash #2 \rhd #3}
@@
       <\(\ms{L}\)> ::= \(\cdot\) \;|\; \(\ms{L}, \lbl{\ell}[\Gamma](x: A)\)
+
+      <\(p\)> ::= 0 \;|\; 1
+
+      <\(q\)> ::= \(\varnothing\) 
+      \;|\; \(\taff\) 
+      \;|\; \(\trel\) 
+      \;|\; \(\tint\)
     \end{grammar}
@@
         \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
-        \(\hasty{\Gamma}{p}{x}{A}\) &
-        \(a\) is a term of type \(A\) in context \(\Gamma\) with purity \(p \in \{0, 1\}\) \\
+        \(\hasty{\Gamma}{p}{x}{A}{q}\) &
+        \(a\) is a term of type \(A\) in context \(\Gamma\) with purity \(p\) and linearity \(q\) \\
         \(\haslb{\Gamma}{t}{\ms{L}}\) &
@@
       {\cwk{\Gamma}{\thyp{x}{A}{q}}}
-      {\hasty{\Gamma}{p}{x}{A}} \qquad
+      {\hasty{\Gamma}{p}{x}{A}{q}} \qquad
     \prftree[r]{\rle{app}}
-      {f \in \mc{I}_p(A, B)}
-      {\hasty{\Gamma}{1}{a}{A}}
-      {\hasty{\Gamma}{p}{f\;a}{B}} \qquad
+      {f \in \mc{I}_p^q(A, B)}
+      {\hasty{\Gamma}{1}{a}{A}{q}}
+      {\hasty{\Gamma}{p}{f\;a}{B}{q}} \qquad
     \prftree[r]{\rle{pair}}
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta}{1}{a}{A}}
-      {\hasty{\Xi}{1}{b}{B}}
-      {\hasty{\Gamma}{p}{(a, b)}{A \otimes B}} \\
+      {\hasty{\Delta}{1}{a}{A}{q}}
+      {\hasty{\Xi}{1}{b}{B}{q}}
+      {\hasty{\Gamma}{p}{(a, b)}{A \otimes B}{q}} \\
     \prftree[r]{\rle{unit}}
       {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{p}{()}{\mb{1}}} \qquad
+      {\hasty{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
     \prftree[r]{\rle{true}}
       {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{p}{\ctt}{\mb{2}}} \qquad
+      {\hasty{\Gamma}{p}{\ctt}{\mb{2}}{q}} \qquad
     \prftree[r]{\rle{false}}
       {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{p}{\cff}{\mb{2}}} \\
+      {\hasty{\Gamma}{p}{\cff}{\mb{2}}{q}} \\
     \prftree[r]{\rle{let}}
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}}
-      {\hasty{\Xi}{1}{a}{A}}
-      {\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}} \qquad
+      {\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}{q}}
+      {\hasty{\Xi}{1}{a}{A}{q}}
+      {\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}{q}} \qquad
     \prftree[r]{\rle{blk}}
       {\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}}
-      {\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}} \\
+      {\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}} \\
     \prftree[r]{\rle{let2}}
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}}
-      {\hasty{\Xi}{1}{a}{A \otimes B}}
-      {\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}}
+      {\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}{q}}
+      {\hasty{\Xi}{1}{a}{A \otimes B}{q}}
+      {\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}{q}}
   \end{gather*}
@@
       %TODO: consider allowing binding here, or should it be purely for lets?
-      {\hasty{\Xi}{0}{a}{A}}
+      {\hasty{\Xi}{0}{a}{A}{\tint}}
       {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} \qquad
@@
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta}{0}{e}{\mb{2}}}
+      {\hasty{\Delta}{0}{e}{\mb{2}}{\tint}}
       {\haslb{\Xi}{s}{\ms{L}}}
@@
       {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
-      {\hasty{\Xi}{1}{a}{A}}
+      {\hasty{\Xi}{1}{a}{A}{q}}
       {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} \\
@@
       {\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
-      {\hasty{\Xi}{1}{a}{A \otimes B}}
+      {\hasty{\Xi}{1}{a}{A \otimes B}{q}}
       {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \\
@@
   \label{fig:ssa-block-typing}
+\end{figure}
+
+\begin{theorem}[Minimal Context] \
+  For all \(a, A\), there exists \(\Delta\) such that \(\forall \Gamma, \hasty{\Gamma}{p}{a}{A}{q} \implies \cwk{\Gamma}{\Delta} \land \hasty{\Delta}{p}{a}{A}{q}\).
+If such a \(\Gamma\) exists, then \(\Delta\) is unique; in this case we call \(\Delta\) \textbf{minimal}.
+\end{theorem}
+
+\begin{figure}
+  \begin{gather*}
+    \prftree[r]{\rle{subst-nil}}
+      {\issubst{\cdot}{\cdot}{\cdot}}
+      \qquad
+    \prftree[r]{\rle{subst-cons}}
+      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
+      {\hasty{\Theta_x}{0}{a}{A}{q} \text{ minimal}}
+      {\tlin{\Theta_x}{q}}
+      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
+      {\issubst{\substcons{\gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
+  \end{gather*}
+  \caption{Typing rules for \isotopessa substitutions}
+  \Description{Typing rules for isotope-SSA substitutions}
+  \label{fig:ssa-subst-typing}
 \end{figure}
@@
 
+\TODO{rewrite union in rule-style?}
+
 \begin{theorem}[Weakening] \
   \begin{itemize}
-    \item If \(\cwk{\Gamma}{\Delta}\), \(\hasty{\Delta}{p}{a}{A}\), then \(\hasty{\Gamma}{p}{a}{A}\)
+    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\), \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
     \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
@@
 
-\begin{theorem}[Minimal Context] \
-  For all \(a, A\), there exists \(\Delta\) such that \(\forall \Gamma, \hasty{\Gamma}{p}{a}{A} \implies \cwk{\Gamma}{\Delta} \land \hasty{\Delta}{p}{a}{A}\).
-If such a \(\Gamma\) exists, then \(\Delta\) is unique; in this case we call \(\Delta\) \textbf{minimal}.
-\end{theorem}
+\TODO{restriction of substitutions to contexts, further context manipulation}
 
-\begin{figure}
-  \begin{gather*}
-    \prftree[r]{\rle{subst-nil}}
-      {\issubst{\cdot}{\cdot}{\cdot}}
-      \qquad
-    \prftree[r]{\rle{subst-cons}}
-      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
-      {\hasty{\Theta_x}{0}{a}{A} \text{ minimal}}
-      {\tlin{\Theta_x}{q}}
-      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
-      {\issubst{\substcons{\gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa substitutions}
-  \Description{Typing rules for isotope-SSA substitutions}
-  \label{fig:ssa-subst-typing}
-\end{figure}
+\TODO{substitution is a RELATION}
 
 \TODO{label substitution}
+
+\TODO{lifting of substitutions to label substitutions}
 
@@
   \begin{itemize}
-    \item If \(\hasty{\Gamma}{p}{a}{A}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}\)
+    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
     \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Theta}{[\gamma]t}{[\gamma]\ms{L}}\)
@@
 \end{theorem}
+
+\begin{figure}
+  \begin{center}
+    \TODO{this}
+  \end{center}
+  \caption{Grammar for \isotopessa blocks with holes}
+  \Description{Grammar for isotope-SSA blocks with holes}
+  \label{fig:blocks-with-holes-grammar}
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \TODO{this}
+  \end{center}
+  \caption{Typing rules for \isotopessa blocks with holes}
+  \Description{Typing rules for isotope-SSA blocks with holes}
+  \label{fig:blocks-with-holes-typing}
+\end{figure}
+
+\TODO{rewriting}
 
@@
   \begin{gather*}
-    \boxed{\dnt{\tlin{A}{\taff}}: \mc{C}_1(\dnt{A}, \mb{1})} \\
+    \boxed{\dnt{\tlin{A}{\taff}: \mc{C}_1^\tint(\dnt{A}, \mb{1})}}
+      \\
     \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
@@
       ; \lambda \\
-    \boxed{\dnt{\tlin{A}{\trel}}}
-      : \mc{C}_1(\dnt{A}, \dnt{A} \otimes \dnt{A}) \\
+    \boxed{\dnt{\tlin{A}{\trel}}
+      : \mc{C}_1^\tint(\dnt{A}, \dnt{A} \otimes \dnt{A})}
+      \\
     \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
@@
       \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
-      ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha \\
-    \boxed{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
-      : \mc{C}_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi}) \\
+      ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
+      \\
+    \boxed{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})}
+      \\
     \dnt{\csplits{\cdot}{\cdot}{\cdot}} = \lambda^{-1} \qquad
@@
       {\Xi}} 
-      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha;\dnt{\Delta} \otimes \sigma;\alpha \\
+      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha;\dnt{\Delta} \otimes \sigma;\alpha 
+      \\
     \dnt{\csplits
@@
       {\Xi, \thyp{x}{A}{q}}} 
-      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha \\
+      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha 
+      \\
     \dnt{\csplits
@@
         \dnt{\Delta} \otimes \sigma \otimes \dnt{A};
-        \alpha \\
+        \alpha 
+        \\
     \dnt{\csplits
@@
         \otimes \dnt{\tlin{A}{\taff}}
-      ; \rho \\
+      ; \rho 
+      \\
     \boxed{\dnt{\cwk{\Gamma}{\Delta}}
-      : \mc{C}_1(\dnt{\Gamma}, \dnt{\Delta})} \\
-      = \dnt{\csplits{\Gamma}{\Delta}{\cdot}};\rho \\
+      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta})} 
+      \\
+      \dnt{\cwk{\Gamma}{\Delta}} 
+      = \dnt{\csplits{\Gamma}{\Delta}{\cdot}};\rho 
+      \\
     \boxed{\dnt{\lwk{\ms{L}}{\ms{K}}}
-      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} \\
+      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} 
+      \\
     \dnt{\lwk{\cdot}{\cdot}} = \ms{id} \qquad
@@
       {\ms{K}, \llhyp{\ell}{\Delta}{A}}}
-        = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\cwk{\Gamma}{\Delta} \otimes \dnt{A})}{} \\
+      = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\cwk{\Gamma}{\Delta} \otimes \dnt{A})}{} 
+      \\
     \dnt{\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
@@
   \begin{gather*}
-    \boxed{\dnt{\hasty{\Gamma}{p}{a}{A}}
-      : \mc{C}_p(\dnt{\Gamma}, \dnt{A})} \\
-    \dnt{\hasty{\Gamma}{p}{x}{A}} 
+    \boxed{\dnt{\hasty{\Gamma}{p}{a}{A}{q}}
+      : \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})} 
+      \\
+    \dnt{\hasty{\Gamma}{p}{x}{A}{q}} 
       = \dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}
       \qquad
-    \dnt{\hasty{\Gamma}{p}{f\;a}{B}}
-      = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}}}{p}
-      ; \ms{inst}_p(f) \\
-    \dnt{\hasty{\Gamma}{p}{(a, b)}{A \otimes B}}
+    \dnt{\hasty{\Gamma}{p}{f\;a}{B}{q}}
+      = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{p}
+      ; \ms{inst}_p(f) 
+      \\
+    \dnt{\hasty{\Gamma}{p}{(a, b)}{A \otimes B}{q}}
       = \upg{(
         \dnt{\csplits{\Gamma}{\Delta}{\Xi}};
-        \dnt{\hasty{\Delta}{1}{a}{A}} \otimes
-        \dnt{\hasty{\Xi}{1}{b}{B}}
-      )}{\uparrow p} \\
-    \dnt{\hasty{\Gamma}{p}{()}{\mb{1}}}
+        \dnt{\hasty{\Delta}{1}{a}{A}{q}} \otimes
+        \dnt{\hasty{\Xi}{1}{b}{B}{q}}
+      )}{\uparrow p} 
+      \\
+    \dnt{\hasty{\Gamma}{p}{()}{\mb{1}}{q}}
       = \upg{\dnt{\cwk{\Gamma}{\cdot}}}{p} 
-      \qquad
-    \dnt{\hasty{\Gamma}{p}{\ctt}{\mb{2}}}
+      \\
+    \dnt{\hasty{\Gamma}{p}{\ctt}{\mb{2}}{q}}
       = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\ctt)}{p}
       \qquad
-    \dnt{\hasty{\Gamma}{p}{\cff}{\mb{2}}}
-      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\cff)}{p} \\
-    \dnt{\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}}
+    \dnt{\hasty{\Gamma}{p}{\cff}{\mb{2}}{q}}
+      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\cff)}{p} 
+      \\
+    \dnt{\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}{q}}
       = \upg{(
         \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}}
-      )}{p};\dnt{\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}} \\
-    \dnt{\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}}
+        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{q}}
+      )}{p};\dnt{\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}{q}} 
+      \\
+    \dnt{\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}{q}}
       = \upg{(
         \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A \otimes B}}
+        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A \otimes B}{q}}
       )}{p}
@@
       \\ \qquad \qquad \qquad \qquad \qquad \alpha
-      ; \dnt{\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}}
+      ; \dnt{\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}{q}}
       \\
-    \dnt{\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}}
+    \dnt{\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}}
       = \dnt{\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}};\alpha_+
@@
     \boxed{\dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      : \mc{C}_1(\dnt{\Gamma}, \dnt{\ms{L}})} \\
+      : \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}})} \\
     \dnt{\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}}
       = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{0}{a}{A}})}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{0}{a}{A}{\tint}})}
       ; \dnt{\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}} \\
@@
       = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\hasty{\Delta}{0}{e}{\mb{2}}} \otimes \dnt{\Xi})}
+      ; \dnt{\hasty{\Delta}{0}{e}{\mb{2}}{\tint}} \otimes \dnt{\Xi})}
       ; \ms{ite}_{\dnt{\Xi}} ;
@@
       = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A}}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A}{q}}
       ; \dnt{\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
@@
       = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
-        ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A \otimes B}}
+        ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A \otimes B}{q}}
         ; \alpha
@@
     \boxed{\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}
-      : \mc{C}_1(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
+      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
       \\
@@
   \begin{itemize}
-    \item Given any two derivations \(D_1, D_2: \hasty{\Gamma}{p}{a}{A}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
     \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
@@
 \begin{theorem}[Semantic Upgrade]
-  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{0}{a}{A}\), then
-  \[\dnt{\hasty{\Gamma}{1}{a}{A}} = \upg{\hasty{\Gamma}{0}{a}{A}}{}\]
+  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
+  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\hasty{\Gamma}{1}{a}{A}{q}}{}\]
 \end{theorem}
@@
     \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_0(\dnt{\Theta}, \dnt{\Gamma})} \\
-    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id} \qquad
     \dnt{\issubst{\substcons{\Gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
-      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{0}{a}{A}}
+      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{0}{a}{A}{q}} \\
+    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id}
   \end{gather*}
@@
     \item \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
-      ;\dnt{\hasty{\Gamma}{p}{a}{A}} 
-      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}}\)
+      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}\)
     \item \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
@@
   \end{center}
-  \caption{Grammar for \isotopessa blocks with holes}
-  \Description{Grammar for isotope-SSA blocks with holes}
-  \label{fig:blocks-with-holes}
+  \caption{Semantics for \isotopessa blocks with holes}
+  \Description{Semantics for isotope-SSA blocks with holes}
+  \label{fig:blocks-with-holes-semantics}
 \end{figure}

### 2023-10-17 pldi24/paper.tex
Grammar and typing rules for blocks and terms with holes
Tap

@@
 \newcommand{\issubst}[3]{#1: #2 \mapsto #3}
+\newcommand{\lbsubst}[3]{#1: #2 \rightsquigarrow #3}
 \newcommand{\substcons}[3]{[{#2 \mapsto #3}]{#1}}
+\newcommand{\lbsubstcons}[4]{[{#2(#3) \mapsto #4}]{#1}}
+\newcommand{\llbsubstcons}[4]{\lbsubstcons{#1}{\lbl{#2}}{#3}{#4}}
+\newcommand{\exprletsubst}[2]{{#1};{#2}}
+\newcommand{\stmtletsubst}[2]{{#1};{#2}}
+\newcommand{\mhole}[1]{{#1}^?}
+\newcommand{\lhole}[1]{?#1}
+\newcommand{\mhasty}[6]{#1;#2 \vdash_{#3} #4: {#5}^{#6}}
+\newcommand{\mhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
+\newcommand{\mlhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
+\newcommand{\tyhole}[5]{#1: #2 \mapsto_{#3} {#4}^{#5}}
+\newcommand{\blkhole}[3]{#1: #2 \mapsto #3}
+\newcommand{\cfghole}[3]{#1: #2 \mapsto #3}
 
@@
       {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
-      %TODO: consider allowing binding here, or should it be purely for lets?
       {\hasty{\Xi}{0}{a}{A}{\tint}}
@@
 
-\begin{theorem}[Minimal Context] \
-  For all \(a, A\), there exists \(\Delta\) such that \(\forall \Gamma, \hasty{\Gamma}{p}{a}{A}{q} \implies \cwk{\Gamma}{\Delta} \land \hasty{\Delta}{p}{a}{A}{q}\).
-If such a \(\Gamma\) exists, then \(\Delta\) is unique; in this case we call \(\Delta\) \textbf{minimal}.
-\end{theorem}
-
 \begin{figure}
-  \begin{gather*}
-    \prftree[r]{\rle{subst-nil}}
-      {\issubst{\cdot}{\cdot}{\cdot}}
-      \qquad
-    \prftree[r]{\rle{subst-cons}}
-      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
-      {\hasty{\Theta_x}{0}{a}{A}{q} \text{ minimal}}
-      {\tlin{\Theta_x}{q}}
-      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
-      {\issubst{\substcons{\gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa substitutions}
-  \Description{Typing rules for isotope-SSA substitutions}
-  \label{fig:ssa-subst-typing}
+  \begin{center}        
+    \begingroup
+    \renewcommand{\arraystretch}{1.5}
+    \setlength{\tabcolsep}{2em}
+    \begin{tabular}{rl}
+        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
+        \(\issubst{\gamma}{\Theta}{\Gamma}\) &
+        ... \\
+        \(\lbsubst{L}{\ms{L}}{\ms{K}}\) &
+        ... \\
+        \(\mhasty{\Gamma}{H}{p}{\mhole{a}}{A}{q}\) &
+        ... \\
+        \(\mhaslb{\Gamma}{H}{\mhole{t}}{\ms{L}}\) &
+        ... \\
+        \(\mlhaslb{\ms{L}}{H}{\mhole{L}}{\ms{K}}\) &
+        ...
+    \end{tabular}
+    \endgroup
+  \end{center}
+  \caption{Metatheoretic typing judgements for \isotopessa}
+  \Description{Metatheoretic typing judgements for isotope-SSA}
+  \label{fig:ssa-meta-judgements}
 \end{figure}
 
-\begin{figure}
-  \begin{gather*}
-    \prftree[r]{\rle{subset-nil}}
-      {\cdot \subseteq \cdot}
-      \qquad
-    \prftree[r]{\rle{subset-cons}}
-      {\Gamma \subseteq \Delta}
-      {\Gamma, \thyp{x}{A}{q} \subseteq \Delta, \thyp{x}{A}{q}}
-      \qquad
-    \prftree[r]{\rle{subset-drop}}
-      {\Gamma \subseteq \Delta}
-      {\Gamma \subseteq \Delta, \thyp{x}{A}{q}} 
-    \\
-    \prftree[r]{\rle{in-head}}
-      {\thyp{x}{A}{q} \in \Gamma, \thyp{x}{A}{q}}
-      \qquad
-    \prftree[r]{\rle{in-cons}}
-      {\thyp{x}{A}{q} \in \Gamma}
-      {\thyp{x}{A}{q} \in \Gamma, \thyp{y}{B}{r}}
-    \\
-    \prftree[r]{\rle{var-head}}
-      {x \in \Gamma, \thyp{x}{A}{q}}
-      \qquad
-    \prftree[r]{\rle{var-cons}}
-      {x \in \Gamma}
-      {x \in \Gamma, \thyp{y}{B}{r}} 
-    \\
-    \Gamma \cup \cdot = \Gamma \qquad
-    \Gamma \cup (\Delta, \thyp{x}{A}{q}) = (\Gamma \cup \Delta), \thyp{x}{A}{q} \text{ if } x \notin \Gamma \qquad
-    \Gamma \cup (\Delta, \thyp{x}{A}{q}) = \Gamma \cup \Delta \text{ otherwise} 
+% \begin{theorem}[Minimal Context] \
+%   For all \(a, A\), there exists \(\Delta\) such that \(\forall \Gamma, \hasty{\Gamma}{p}{a}{A}{q} \implies \cwk{\Gamma}{\Delta} \land \hasty{\Delta}{p}{a}{A}{q}\).
+If such a \(\Gamma\) exists, then \(\Delta\) is unique; in this case we call \(\Delta\) \textbf{minimal}.
+% \end{theorem}
+
+% \begin{figure}
+%   \begin{gather*}
+%     \prftree[r]{\rle{subset-nil}}
+%       {\cdot \subseteq \cdot}
+%       \qquad
+%     \prftree[r]{\rle{subset-cons}}
+%       {\Gamma \subseteq \Delta}
+%       {\Gamma, \thyp{x}{A}{q} \subseteq \Delta, \thyp{x}{A}{q}}
+%       \qquad
+%     \prftree[r]{\rle{subset-drop}}
+%       {\Gamma \subseteq \Delta}
+%       {\Gamma \subseteq \Delta, \thyp{x}{A}{q}} 
 %     \\
-    %   \cdot^\Gamma_\Delta = \cdot 
+%     \prftree[r]{\rle{in-head}}
+%       {\thyp{x}{A}{q} \in \Gamma, \thyp{x}{A}{q}}
 %       \qquad
-    %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta
-    %     = \Theta^\Gamma_\Delta, 
-    %       \thyp{x}{A}{q} \text{ if } x \in \Delta \lor x \notin \Gamma
+%     \prftree[r]{\rle{in-cons}}
+%       {\thyp{x}{A}{q} \in \Gamma}
+%       {\thyp{x}{A}{q} \in \Gamma, \thyp{y}{B}{r}}
+%     \\
+%     \prftree[r]{\rle{var-head}}
+%       {x \in \Gamma, \thyp{x}{A}{q}}
 %       \qquad
-    %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta = \Theta^\Gamma_\Delta \text{ otherwise}
+%     \prftree[r]{\rle{var-cons}}
+%       {x \in \Gamma}
+%       {x \in \Gamma, \thyp{y}{B}{r}} 
 %     \\
-    %   \cdot^\Gamma_\Delta = \cdot 
-    %   \qquad
-    %   (\ms{L}, \llhyp{\ell}{\Theta}{A})^\Gamma_\Delta
-    %   = \ms{L}^\Gamma_\Delta, \llhyp{\ell}{\Theta^\Gamma_\Delta}{A}
-  \end{gather*}
-  \caption{Manipulation of \isotopessa contexts}
-  \Description{Manipulation of isotope-SSA contexts}
-  \label{fig:context-manipulation}
-\end{figure}
-
-\TODO{rewrite union in rule-style?}
+%     \Gamma \cup \cdot = \Gamma \qquad
+%     \Gamma \cup (\Delta, \thyp{x}{A}{q}) = (\Gamma \cup \Delta), \thyp{x}{A}{q} \text{ if } x \notin \Gamma \qquad
+%     \Gamma \cup (\Delta, \thyp{x}{A}{q}) = \Gamma \cup \Delta \text{ otherwise} 
+%     % \\
+%     %   \cdot^\Gamma_\Delta = \cdot 
+%     %   \qquad
+%     %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta
+%     %     = \Theta^\Gamma_\Delta, 
+%     %       \thyp{x}{A}{q} \text{ if } x \in \Delta \lor x \notin \Gamma
+%     %   \qquad
+%     %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta = \Theta^\Gamma_\Delta \text{ otherwise}
+%     % \\
+%     %   \cdot^\Gamma_\Delta = \cdot 
+%     %   \qquad
+%     %   (\ms{L}, \llhyp{\ell}{\Theta}{A})^\Gamma_\Delta
+%     %   = \ms{L}^\Gamma_\Delta, \llhyp{\ell}{\Theta^\Gamma_\Delta}{A}
+%   \end{gather*}
+%   \caption{Manipulation of \isotopessa contexts}
+%   \Description{Manipulation of isotope-SSA contexts}
+%   \label{fig:context-manipulation}
+% \end{figure}
 
@@
 
-\TODO{restriction of substitutions to contexts, further context manipulation}
+\begin{figure}
+  \begin{grammar}
+    <\(\gamma, \delta, \theta\)> ::= \(\cdot\)
+      \;|\; \(\substcons{\gamma}{x}{a}\)
+  \end{grammar}
+  \caption{Grammar for \isotopessa substitutions}
+  \Description{Grammar for isotope-SSA substitutions}
+  \label{fig:ssa-subst-grammar}
+\end{figure}
 
-\TODO{substitution is a RELATION}
+\begin{figure}
+  \begin{gather*}
+    \prftree[r]{\rle{subst-nil}}
+      {\issubst{\cdot}{\cdot}{\cdot}}
+      \qquad
+    \prftree[r]{\rle{subst-cons}}
+      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
+      {\hasty{\Theta_x}{0}{a}{A}{q}}
+      {\tlin{\Theta_x}{q}}
+      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
+      {\issubst{\substcons{\gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
+      \\
+    \prftree[r]{\rle{lb-subst-nil}}
+    {\lbsubst{\cdot}{\cdot}{\ms{K}}}
+      \qquad
+    \prftree[r]{\rle{lb-subst-cons}}
+      {\lbsubst{L}{\ms{L}}{\ms{K}}}
+      {\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{K}}}
+      {\lbsubst{\llbsubstcons{L}{\ell}{x}{t}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
+  \end{gather*}
+  \caption{Typing rules for \isotopessa substitutions}
+  \Description{Typing rules for isotope-SSA substitutions}
+  \label{fig:ssa-subst-typing}
+\end{figure}
 
-\TODO{label substitution}
+\begin{figure}
+  \begin{gather*}
+    \gamma_\cdot = \cdot
+    \qquad
+    \gamma_{\Delta, \thyp{x}{A}{q}} = \substcons{\gamma_\Delta}{x}{a}
+    \text{ if } x \mapsto a \in \gamma
+    \qquad
+    \gamma_{\Delta, \thyp{x}{A}{q}} = \substcons{\gamma_\Delta}{x}{x}
+    \text{ otherwise} 
+    \\
+    \stmtletsubst{\cdot}{t} = t
+    \qquad
+    \stmtletsubst{\substcons{\gamma}{x}{a}}{t} 
+      = \letstmt{x}{a}{\stmtletsubst{\gamma}{t}}
+    \qquad
+    \exprletsubst{\cdot}{e} = e
+    \qquad
+    \exprletsubst{\substcons{\gamma}{x}{a}}{e} 
+      = \letstmt{x}{a}{\stmtletsubst{\gamma}{e}}
+    \\
+    \gamma^{\cdot} = \cdot
+    \qquad
+    \gamma^{\ms{L}, \llhyp{\ell}{\Delta}{A}} 
+      = \llbsubstcons{\gamma^{\ms{L}}}{\ell}{x}
+        {\stmtletsubst{\gamma^\Delta}{\lbrb{\ell}{x}}}
+  \end{gather*}
+  \caption{Operations on \isotopessa substitutions}
+  \Description{Operations on isotope-SSA substitutions}
+  \label{fig:ssa-subst-lift}
+\end{figure}
 
-\TODO{lifting of substitutions to label substitutions}
-
-\begin{theorem}[Substitution Weakening] \
-  \TODO{this}
+\begin{theorem}[Substitution Weakening] 
+  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
+  \begin{itemize}
+    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
+    \item \(\issubst{\gamma_\Delta}{\Theta_\Delta}{\Delta}\)
+    \item \(\issubst{\gamma_\Xi}{\Theta_\Xi}{\Xi}\)
+  \end{itemize}
 \end{theorem}
@@
 
+\TODO{substitution/label substitution grammar?}
+
+\TODO{do we need to define capture avoiding substitution vs. rewriting?}
+
+\TODO{substitution composition/label substitution composition?}
+
+\TODO{substitution weakening}
+
 \begin{figure}
-  \begin{center}
-    \TODO{this}
-  \end{center}
-  \caption{Grammar for \isotopessa blocks with holes}
+  \begin{grammar}
+    <\(\mhole{a}, \mhole{b}, \mhole{c}, \mhole{e}\)> ::= \(\lhole{X}\)
+    \;|\; \(x\) 
+    \;|\; \(f\;\mhole{a}\)
+    \;|\; \((\mhole{a}, \mhole{b})\) 
+    \;|\; \(()\) 
+    \;|\; \(\ctt\) 
+    \;|\; \(\cff\)
+    \;|\; \(\letexpr{x}{\mhole{a}}{\mhole{e}}\)
+    \;|\; \(\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}\)
+    \alt \(\lbsplice{\ell}{x: A}{\mhole{t}}\)
+
+    <\(\mhole{s}, \mhole{t}\)> ::= \(\lhole{T}\) 
+    \;|\; \(\lbrb{\ell}{\mhole{a}}\) 
+    \;|\; \(\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}\)
+    \;|\; \(\letstmt{x}{\mhole{a}}{\mhole{t}}\)
+    \;|\; \(\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}\)
+    \;|\; \(\ewhere{\mhole{t}}{\mhole{L}}\)
+
+    <\(\mhole{L}\)> ::= \(\lhole{W}\) \;|\; \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{\mhole{t}}, \mhole{L}\)
+
+    <\(H\)> ::= \(\cdot\) 
+    \;|\; \(H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}\)
+    \;|\; \(H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}\)
+    \;|\; \(H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\)
+  \end{grammar}
+  \caption{Grammar for \isotopessa terms and blocks with holes}
   \Description{Grammar for isotope-SSA blocks with holes}
@@
 \begin{figure}
-  \begin{center}
-    \TODO{this}
-  \end{center}
+  \begin{gather*}    
+    \prftree[r]{\rle{hole-tm}}
+      {\tyhole{\lhole{X}}{\Delta}{p}{A}{q}}
+      {\cwk{\Gamma}{\Delta}}
+      {\mhasty{\Gamma}{H}{p}{\lhole{X}}{A}{q}} \qquad
+    \prftree[r]{\rle{var}}
+      {\cwk{\Gamma}{\thyp{x}{A}{q}}}
+      {\mhasty{\Gamma}{H}{p}{x}{A}{q}} \qquad
+    \prftree[r]{\rle{app}}
+      {f \in \mc{I}_p^q(A, B)}
+      {\mhasty{\Gamma}{H}{1}{\mhole{a}}{A}{q}}
+      {\mhasty{\Gamma}{H}{p}{f\;\mhole{a}}{B}{q}} \\
+    \prftree[r]{\rle{pair}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhasty{\Delta}{H}{1}{\mhole{a}}{A}{q}}
+      {\mhasty{\Xi}{H}{1}{\mhole{b}}{B}{q}}
+      {\mhasty{\Gamma}{H}{p}{(\mhole{a}, \mhole{b})}{A \otimes B}{q}} \\
+    \prftree[r]{\rle{unit}}
+      {\cwk{\Gamma}{\cdot}}
+      {\mhasty{\Gamma}{H}{p}{()}{\mb{1}}{q}} \qquad
+    \prftree[r]{\rle{true}}
+      {\cwk{\Gamma}{\cdot}}
+      {\mhasty{\Gamma}{H}{p}{\ctt}{\mb{2}}{q}} \qquad
+    \prftree[r]{\rle{false}}
+      {\cwk{\Gamma}{\cdot}}
+      {\mhasty{\Gamma}{H}{p}{\cff}{\mb{2}}{q}} \\
+    \prftree[r]{\rle{let}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhasty{\Delta, \thyp{x}{A}{}}{H}{p}{\mhole{e}}{B}{q}}
+      {\mhasty{\Xi}{H}{1}{\mhole{a}}{A}{q}}
+      {\mhasty{\Gamma}{H}{p}{\letexpr{x}{\mhole{a}}{\mhole{e}}}{B}{q}} \qquad
+    \prftree[r]{\rle{blk}}
+      {\mhaslb{\Gamma}{H}{\mhole{t}}{\llhyp{\ell}{\cdot}{A}}}
+      {\mhasty{\Gamma}{H}{0}{\lbsplice{\ell}{A}{\mhole{t}}}{A}{\varnothing}} \\
+    \prftree[r]{\rle{let2}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{H}{p}{\mhole{e}}{C}{q}}
+      {\mhasty{\Xi}{1}{H}{\mhole{a}}{A \otimes B}{q}}
+      {\mhasty{\Gamma}{H}{p}{\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}}{C}{q}}
+  \end{gather*}
+  \caption{Typing rules for \isotopessa terms with holes}
+  \Description{Typing rules for isotope-SSA terms with holes}
+  \label{fig:terms-with-holes-typing}
+\end{figure}
+
+\begin{figure}
+  \begin{gather*}    
+    \prftree[r]{\rle{hole-blk}}
+      {\blkhole{\lhole{T}}{\Delta}{\ms{L}} \in H}
+      {\cwk{\Gamma}{\Delta}}
+      {\mhaslb{\Gamma}{H}{\lhole{T}}{\ms{L}}} 
+      \qquad
+    \prftree[r]{\rle{br}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
+      {\mhasty{\Xi}{H}{0}{\mhole{a}}{A}{\tint}}
+      {\mhaslb{\Gamma}{H}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
+      \\
+    \prftree[r]{\rle{ite}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhasty{\Delta}{H}{0}{\mhole{e}}{\mb{2}}{\tint}}
+      {\mhaslb{\Xi}{H}{\mhole{s}}{\ms{L}}}
+      {\mhaslb{\Xi}{H}{\mhole{t}}{\ms{L}}}
+      {\mhaslb{\Gamma}{H}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\
+    \prftree[r]{\rle{let-blk}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhaslb{\Delta, \thyp{x}{A}{}}{H}{\mhole{t}}{\ms{L}}}
+      {\mhasty{\Xi}{H}{1}{\mhole{a}}{A}{q}}
+      {\mhaslb{\Gamma}{H}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
+    \prftree[r]{\rle{let2-blk}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhaslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{H}{\mhole{t}}{\ms{L}}}
+      {\mhasty{\Xi}{H}{1}{\mhole{a}}{A \otimes B}{q}}
+      {\mhaslb{\Gamma}{H}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
+    \prftree[r]{\rle{where}}
+      {\mhaslb{\Gamma}{H}{\mhole{t}}{\ms{L}}}
+      {\mlhaslb{\ms{L}}{H}{\mhole{L}}{\ms{K}}}
+      {\mhaslb{\Gamma}{H}{\ewhere{\mhole{t}}{\mhole{L}}}{\ms{K}}} 
+      \qquad
+    \prftree[r]{\rle{hole-cfg}}
+      {\cfghole{\lhole{W}}{\ms{L}}{\ms{K}} \in H}
+      {\lwk{\ms{M}}{\ms{K}}}
+      {\mlhaslb{\ms{L}}{H}{\lhole{W}}{\ms{K}}}
+      \\
+    \prftree[r]{\rle{nil-br}}
+      {\mlhaslb{\ms{L}}{H}{\cdot}{\ms{L}}} \qquad
+    \prftree[r]{\rle{cons-br}}
+      {\mlhaslb{\ms{L}}{H}{\mhole{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
+      {\mhaslb{\Gamma, \thyp{x}{A}{}}{H}{\mhole{t}}{\ms{L}}}
+      {\mlhaslb{\ms{L}}{H}{\mhole{L}, \lwbranch{\ell}{x: A}{\mhole{t}}}{\ms{K}}}
+  \end{gather*}
   \caption{Typing rules for \isotopessa blocks with holes}
@@
     \dnt{\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
-      = \dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{A}{t}}}
-      ; [[\iota_0, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{L}}], \iota_1]
+      = \dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
+      ; [[\iota_0, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}], \iota_1]
       \\
@@
   \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\hasty{\Gamma}{1}{a}{A}{q}}{}\]
+\end{theorem}
+
+\begin{theorem}[Semantic Weakening] \
+  \TODO{this}
 \end{theorem}
@@
 
-\TODO{\(\implies\) E-graph optimization}
+\begin{figure}
+  \begin{align*}
+    \dnt{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}
+      &= \dnt{H} \times \mc{C}_p^q(\dnt{\Gamma}, \dnt{A}) 
+      \\
+    \dnt{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}
+      &= \dnt{H} \times \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}}) 
+      \\
+    \dnt{H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
+      &= \dnt{H} \times \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
+      \\
+    \dnt{\cdot} 
+      &= {()}
+  \end{align*}
+  \caption{Semantics for \isotopessa holes}
+  \Description{Semantics for isotope-SSA holes}
+  \label{fig:holes-semantics}
+\end{figure}
 
 \begin{figure}
-  \begin{center}
-    \TODO{this}
-  \end{center}
+  \begin{gather*}
+    \boxed{
+      \dnt{\mhasty{\Gamma}{H}{p}{\mhole{a}}{A}{q}}
+      : \dnt{H} \to \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})}
+      \\
+    \sorry
+  \end{gather*}
+  \caption{Semantics for \isotopessa terms with holes}
+  \Description{Semantics for isotope-SSA blocks with holes}
+  \label{fig:terms-with-holes-semantics}
+\end{figure}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{
+      \dnt{\mhaslb{\Gamma}{H}{\mhole{t}}{\ms{L}}}
+      : \dnt{H} \to \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}})}
+      \\
+    \sorry
+      \\  
+    \boxed{
+      \dnt{\mlhaslb{\ms{L}}{H}{\mhole{L}}{\ms{K}}}
+      : \dnt{H} \to \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
+      \\
+    \sorry
+  \end{gather*}
   \caption{Semantics for \isotopessa blocks with holes}
@@
 \end{theorem}
+
+\TODO{\(\implies\) E-graph optimization}
 

### 2023-10-18 pldi24/paper.tex
Functional tap
Rewriting skeleton

--- 2023-10-13 pldi24/paper.tex
@@
 \newcommand{\issubst}[3]{#1: #2 \mapsto #3}
-\newcommand{\substcons}[3]{[{#2 \mapsto #3}]{#1}}
+\newcommand{\lbsubst}[3]{#1: #2 \rightsquigarrow #3}
+\newcommand{\exprletsubst}[2]{{#1};{#2}}
+\newcommand{\stmtletsubst}[2]{{#1};{#2}}
+\newcommand{\mhole}[1]{{#1}^?}
+\newcommand{\lhole}[1]{?#1}
+\newcommand{\mhasty}[6]{#1;#2 \vdash_{#3} #4: {#5}^{#6}}
+\newcommand{\mhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
+\newcommand{\mlhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
+\newcommand{\tyhole}[5]{#1: #2 \mapsto_{#3} {#4}^{#5}}
+\newcommand{\blkhole}[3]{#1: #2 \mapsto #3}
+\newcommand{\cfghole}[3]{#1: #2 \mapsto #3}
+\newcommand{\substctx}[2]{{#1}^{#2}}
+\newcommand{\substlbs}[2]{{#1}^{#2}}
+\newcommand{\restrictsubst}[2]{{#1}_{#2}}
+\newcommand{\subsubst}[2]{#1 \subseteq #2}
+\newcommand{\isrw}[3]{#1: #2 \mapsto #3}
+\newcommand{\strictlbsubst}[3]{#1: #2 \rightsquigarrow_= #3}
 
@@
 
-\begin{theorem}[Minimal Context] \
-  For all \(a, A\), there exists \(\Delta\) such that \(\forall \Gamma, \hasty{\Gamma}{p}{a}{A}{q} \implies \cwk{\Gamma}{\Delta} \land \hasty{\Delta}{p}{a}{A}{q}\).
-If such a \(\Gamma\) exists, then \(\Delta\) is unique; in this case we call \(\Delta\) \textbf{minimal}.
-\end{theorem}
-
 \begin{figure}
-  \begin{gather*}
-    \prftree[r]{\rle{subst-nil}}
-      {\issubst{\cdot}{\cdot}{\cdot}}
-      \qquad
-    \prftree[r]{\rle{subst-cons}}
-      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
-      {\hasty{\Theta_x}{0}{a}{A}{q} \text{ minimal}}
-      {\tlin{\Theta_x}{q}}
-      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
-      {\issubst{\substcons{\gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa substitutions}
-  \Description{Typing rules for isotope-SSA substitutions}
-  \label{fig:ssa-subst-typing}
+  \begin{center}        
+    \begingroup
+    \renewcommand{\arraystretch}{1.5}
+    \setlength{\tabcolsep}{2em}
+    \begin{tabular}{rl}
+        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
+        \(\issubst{\gamma}{\Theta}{\Gamma}\) &
+        ... \\
+        \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) &
+        ... \\
+        \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) &
+        ... \\
+        \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) &
+        ... \\
+        \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) &
+        ...
+    \end{tabular}
+    \endgroup
+  \end{center}
+  \caption{Metatheoretic typing judgements for \isotopessa}
+  \Description{Metatheoretic typing judgements for isotope-SSA}
+  \label{fig:ssa-meta-judgements}
 \end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \prftree[r]{\rle{subset-nil}}
-      {\cdot \subseteq \cdot}
-      \qquad
-    \prftree[r]{\rle{subset-cons}}
-      {\Gamma \subseteq \Delta}
-      {\Gamma, \thyp{x}{A}{q} \subseteq \Delta, \thyp{x}{A}{q}}
-      \qquad
-    \prftree[r]{\rle{subset-drop}}
-      {\Gamma \subseteq \Delta}
-      {\Gamma \subseteq \Delta, \thyp{x}{A}{q}} 
-    \\
-    \prftree[r]{\rle{in-head}}
-      {\thyp{x}{A}{q} \in \Gamma, \thyp{x}{A}{q}}
-      \qquad
-    \prftree[r]{\rle{in-cons}}
-      {\thyp{x}{A}{q} \in \Gamma}
-      {\thyp{x}{A}{q} \in \Gamma, \thyp{y}{B}{r}}
-    \\
-    \prftree[r]{\rle{var-head}}
-      {x \in \Gamma, \thyp{x}{A}{q}}
-      \qquad
-    \prftree[r]{\rle{var-cons}}
-      {x \in \Gamma}
-      {x \in \Gamma, \thyp{y}{B}{r}} 
-    \\
-    \Gamma \cup \cdot = \Gamma \qquad
-    \Gamma \cup (\Delta, \thyp{x}{A}{q}) = (\Gamma \cup \Delta), \thyp{x}{A}{q} \text{ if } x \notin \Gamma \qquad
-    \Gamma \cup (\Delta, \thyp{x}{A}{q}) = \Gamma \cup \Delta \text{ otherwise} 
-    % \\
-    %   \cdot^\Gamma_\Delta = \cdot 
-    %   \qquad
-    %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta
-    %     = \Theta^\Gamma_\Delta, 
-    %       \thyp{x}{A}{q} \text{ if } x \in \Delta \lor x \notin \Gamma
-    %   \qquad
-    %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta = \Theta^\Gamma_\Delta \text{ otherwise}
-    % \\
-    %   \cdot^\Gamma_\Delta = \cdot 
-    %   \qquad
-    %   (\ms{L}, \llhyp{\ell}{\Theta}{A})^\Gamma_\Delta
-    %   = \ms{L}^\Gamma_\Delta, \llhyp{\ell}{\Theta^\Gamma_\Delta}{A}
-  \end{gather*}
-  \caption{Manipulation of \isotopessa contexts}
-  \Description{Manipulation of isotope-SSA contexts}
-  \label{fig:context-manipulation}
-\end{figure}
-
-\TODO{rewrite union in rule-style?}
 
@@
 
-\TODO{restriction of substitutions to contexts, further context manipulation}
+\begin{figure}
+  \begin{gather*}
+    \boxed{\gamma: \ms{Var} \to \ms{Expr}}
+    \\
+    \prftree[r]{\rle{subst-nil}}
+      {\issubst{\gamma}{\cdot}{\cdot}}
+      \qquad
+    \prftree[r]{\rle{subst-cons}}
+      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
+      {\hasty{\Theta_x}{0}{\gamma(x)}{A}{q}}
+      {\tlin{\Theta_x}{q}}
+      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
+      {\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
+    \\
+    \boxed{\mc{L}: \ms{Label} \to \ms{Expr} \to \ms{Block} \text{ s.t. for fresh } x, [a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)}
+    \\
+    \prftree[r]{\rle{lb-subst-nil}}
+    {\lbsubst{\mc{L}}{\cdot}{\ms{K}}}
+      \qquad
+    \prftree[r]{\rle{lb-subst-cons}}
+      {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      {\haslb{\Gamma, \thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}}
+      {\lbsubst{\mc{L}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
+    \\
+    \prftree[r]{\rle{lb-strict}}
+      {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      {\ms{labels}(\ms{L}) = \ms{labels}(\ms{K})}
+      {\strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+  \end{gather*}
+  \caption{Typing rules for \isotopessa substitutions. We assume that for all but finitely many \(x\), \(\lbl{\ell}\), \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) = \lbrb{\ell}{x}\).}
+  \Description{Typing rules for isotope-SSA substitutions}
+  \label{fig:ssa-subst-typing}
+\end{figure}
 
-\TODO{substitution is a RELATION}
+\begin{figure}
+  \begin{gather*}
+    \subsubst{\gamma}{\gamma'} \iff 
+      \forall x, \gamma(x) \in \{x, \gamma'(x)\}
+    \\
+    \restrictsubst{\gamma}{\Gamma}(x) 
+      =
+      (\ms{if}\; x \in \Gamma
+      \;\ms{then}\;\gamma(x)
+      \;\ms{else}\;x)
+    \\
+    \stmtletsubst{\substctx{\gamma}{\cdot}}{t} = t
+    \qquad
+    \stmtletsubst{\substctx{\gamma}{\Gamma, x: A^q}}{t} 
+      = \letstmt{x}{\gamma(x)}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}
+    \qquad
+    \exprletsubst{\cdot}{e} = e
+    \qquad
+    \exprletsubst{\substctx{\gamma}{\Gamma, x: A^q}}{e} 
+      = \letstmt{x}{\gamma(x)}{\substctx{\gamma}{\Gamma}}
+    \\
+    \substlbs{\gamma}{\cdot}(\lbl{\ell}, x) = \lbrb{\ell}{x}
+    \qquad
+    \substlbs{\gamma}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}(\lbl{k}, x)
+      = 
+      (\ms{if}\; \lbl{k} = \lbl{\ell} 
+      \;\ms{then}\;\stmtletsubst{\substctx{\gamma}{\Gamma}}{\lbrb{\ell}{x}}
+      \;\ms{else}\;\substlbs{\gamma}{\ms{L}}(\lbl{\ell}, x))
+    \\
+    [\mc{L}](\lbrb{\ell}{a}) = \mc{L}(\lbl{\ell}, a)
+    \qquad
+    [\mc{L}]\ite{e}{s}{t} = \ite{e}{[\mc{L}]s}{[\mc{L}]t}
+    \\
+    [\mc{L}](\letstmt{x}{a}{t}) = (\letstmt{x}{a}{[\mc{L}]t})
+    \qquad
+    [\mc{L}](\letstmt{(x, y)}{a}{t}) = (\letstmt{(x, y)}{a}{[\mc{L}]t})
+    \\
+    [\mc{L}]\ewhere{t}{L} = \ewhere{[\mc{L}]t}{[\mc{L}]L}
+    \\
+    [\mc{L}]\cdot = \cdot
+    \qquad
+    [\mc{L}](\lwbranch{\ell}{x: A}{t})
+    = (\lwbranch{\ell}{x: A}{[\mc{L}]t})
+  \end{gather*}
+  \caption{Operations on \isotopessa substitutions. In particular, note that for all \(\Gamma\), \(\subsubst{\restrictsubst{\gamma}{\Gamma}}{\gamma}\)}
+  \Description{Operations on isotope-SSA substitutions}
+  \label{fig:ssa-subst-ops}
+\end{figure}
 
-\TODO{label substitution}
-
-\TODO{lifting of substitutions to label substitutions}
-
-\begin{theorem}[Substitution Weakening] \
-  \TODO{this}
-\end{theorem}
+\begin{lemma}[Substitution Weakening] 
+  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
+  \begin{itemize}
+    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
+    \item \(\issubst{\gamma_\Delta}{\Theta_\Delta}{\Delta}\)
+    \item \(\issubst{\gamma_\Xi}{\Theta_\Xi}{\Xi}\)
+  \end{itemize}
+  Furthermore, if \(\subsubst{\gamma}{\gamma'}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then \(\issubst{\gamma'}{\Theta}{\Gamma}\).
+\end{lemma}
 
@@
     \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
-    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Theta}{[\gamma]t}{[\gamma]\ms{L}}\)
+    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), \(\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}\), and then \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
   \end{itemize}
@@
 
+\begin{theorem}[Label Substitution] 
+  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), if \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
+\end{theorem}
+
 \begin{figure}
-  \begin{center}
-    \TODO{this}
-  \end{center}
-  \caption{Grammar for \isotopessa blocks with holes}
+  \begin{grammar}
+    <\(\mhole{a}, \mhole{b}, \mhole{c}, \mhole{e}\)> ::= \(\lhole{X}\)
+    \;|\; \(x\) 
+    \;|\; \(f\;\mhole{a}\)
+    \;|\; \((\mhole{a}, \mhole{b})\) 
+    \;|\; \(()\) 
+    \;|\; \(\ctt\) 
+    \;|\; \(\cff\)
+    \;|\; \(\letexpr{x}{\mhole{a}}{\mhole{e}}\)
+    \;|\; \(\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}\)
+    \alt \(\lbsplice{\ell}{x: A}{\mhole{t}}\)
+
+    <\(\mhole{s}, \mhole{t}\)> ::= \(\lhole{T}\) 
+    \;|\; \(\lbrb{\ell}{\mhole{a}}\) 
+    \;|\; \(\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}\)
+    \;|\; \(\letstmt{x}{\mhole{a}}{\mhole{t}}\)
+    \;|\; \(\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}\)
+    \;|\; \(\ewhere{\mhole{t}}{\mhole{L}}\)
+
+    <\(\mhole{L}\)> ::= \(\lhole{W}\) \;|\; \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{\mhole{t}}, \mhole{L}\)
+
+    <\(H\)> ::= \(\cdot\) 
+    \;|\; \(H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}\)
+    \;|\; \(H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}\)
+    \;|\; \(H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\)
+  \end{grammar}
+  \caption{Grammar for \isotopessa terms and blocks with holes}
   \Description{Grammar for isotope-SSA blocks with holes}
@@
 \begin{figure}
-  \begin{center}
-    \TODO{this}
-  \end{center}
+  \begin{gather*}    
+    \prftree[r]{\rle{hole-tm}}
+      {\tyhole{\lhole{X}}{\Delta}{p}{A}{q}}
+      {\cwk{\Gamma}{\Delta}}
+      {\mhasty{H}{\Gamma}{p}{\lhole{X}}{A}{q}} \qquad
+    \prftree[r]{\rle{var}}
+      {\cwk{\Gamma}{\thyp{x}{A}{q}}}
+      {\mhasty{H}{\Gamma}{p}{x}{A}{q}} \qquad
+    \prftree[r]{\rle{app}}
+      {f \in \mc{I}_p^q(A, B)}
+      {\mhasty{H}{\Gamma}{1}{\mhole{a}}{A}{q}}
+      {\mhasty{H}{\Gamma}{p}{f\;\mhole{a}}{B}{q}} \\
+    \prftree[r]{\rle{pair}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhasty{H}{\Delta}{1}{\mhole{a}}{A}{q}}
+      {\mhasty{H}{\Xi}{1}{\mhole{b}}{B}{q}}
+      {\mhasty{H}{\Gamma}{p}{(\mhole{a}, \mhole{b})}{A \otimes B}{q}} \\
+    \prftree[r]{\rle{unit}}
+      {\cwk{\Gamma}{\cdot}}
+      {\mhasty{H}{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
+    \prftree[r]{\rle{true}}
+      {\cwk{\Gamma}{\cdot}}
+      {\mhasty{H}{\Gamma}{p}{\ctt}{\mb{2}}{q}} \qquad
+    \prftree[r]{\rle{false}}
+      {\cwk{\Gamma}{\cdot}}
+      {\mhasty{H}{\Gamma}{p}{\cff}{\mb{2}}{q}} \\
+    \prftree[r]{\rle{let}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhasty{H}{\Delta, \thyp{x}{A}{}}{p}{\mhole{e}}{B}{q}}
+      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
+      {\mhasty{H}{\Gamma}{p}{\letexpr{x}{\mhole{a}}{\mhole{e}}}{B}{q}} \qquad
+    \prftree[r]{\rle{blk}}
+      {\mhaslb{H}{\Gamma}{\mhole{t}}{\llhyp{\ell}{\cdot}{A}}}
+      {\mhasty{H}{\Gamma}{0}{\lbsplice{\ell}{A}{\mhole{t}}}{A}{\varnothing}} \\
+    \prftree[r]{\rle{let2}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhasty{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{\mhole{e}}{C}{q}}
+      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
+      {\mhasty{H}{\Gamma}{p}{\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}}{C}{q}}
+  \end{gather*}
+  \caption{Typing rules for \isotopessa terms with holes}
+  \Description{Typing rules for isotope-SSA terms with holes}
+  \label{fig:terms-with-holes-typing}
+\end{figure}
+
+\begin{figure}
+  \begin{gather*}    
+    \prftree[r]{\rle{hole-blk}}
+      {\blkhole{\lhole{T}}{\Delta}{\ms{L}} \in H}
+      {\cwk{\Gamma}{\Delta}}
+      {\mhaslb{H}{\Gamma}{\lhole{T}}{\ms{L}}} 
+      \qquad
+    \prftree[r]{\rle{br}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
+      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{\tint}}
+      {\mhaslb{H}{\Gamma}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
+      \\
+    \prftree[r]{\rle{ite}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhasty{H}{\Delta}{0}{\mhole{e}}{\mb{2}}{\tint}}
+      {\mhaslb{H}{\Xi}{\mhole{s}}{\ms{L}}}
+      {\mhaslb{H}{\Xi}{\mhole{t}}{\ms{L}}}
+      {\mhaslb{H}{\Gamma}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\
+    \prftree[r]{\rle{let-blk}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhaslb{H}{\Delta, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
+      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
+      {\mhaslb{H}{\Gamma}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
+    \prftree[r]{\rle{let2-blk}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\mhaslb{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{\mhole{t}}{\ms{L}}}
+      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
+      {\mhaslb{H}{\Gamma}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
+    \prftree[r]{\rle{where}}
+      {\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
+      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
+      {\mhaslb{H}{\Gamma}{\ewhere{\mhole{t}}{\mhole{L}}}{\ms{K}}} 
+      \qquad
+    \prftree[r]{\rle{hole-cfg}}
+      {\cfghole{\lhole{W}}{\ms{L}}{\ms{K}} \in H}
+      {\lwk{\ms{M}}{\ms{K}}}
+      {\mlhaslb{\ms{L}}{H}{\lhole{W}}{\ms{K}}}
+      \\
+    \prftree[r]{\rle{nil-br}}
+      {\mlhaslb{H}{\ms{L}}{\cdot}{\ms{L}}} \qquad
+    \prftree[r]{\rle{cons-br}}
+      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
+      {\mhaslb{H}{\Gamma, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
+      {\mlhaslb{H}{\ms{L}}{\mhole{L}, \lwbranch{\ell}{x: A}{\mhole{t}}}{\ms{K}}}
+  \end{gather*}
   \caption{Typing rules for \isotopessa blocks with holes}
@@
 
-\TODO{rewriting}
+\begin{figure}
+  \begin{gather*}
+    \boxed{H: \ms{Hole} \to \ms{MExpr} \sqcup \ms{MBlock}}
+    \\
+    \sorry
+  \end{gather*}
+  \caption{Typing rules for \isotopessa rewrites}
+  \Description{Typing rules for isotope-SSA rewrites}
+  \label{fig:rewrite-typing}
+\end{figure}
+
+\begin{theorem}[Rewriting]
+  Given \(\isrw{\mc{H}}{H}{I}\), we have
+  \begin{itemize}
+    \item If \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), then \(\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}\)
+    \item If \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), then \(\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}\)
+  \end{itemize}
+  Furthermore, we have that 
+  \begin{itemize}
+    \item \(\mhasty{\cdot}{\Gamma}{p}{a}{A}{q} \iff \hasty{\Gamma}{p}{a}{A}{q}\)
+    \item \(\mhaslb{\cdot}{\Gamma}{t}{\ms{L}} \iff \haslb{\Gamma}{t}{\ms{L}}\)
+  \end{itemize}
+\end{theorem}
 
@@
 
-\begin{theorem}[Coherence]
+\begin{lemma}[Coherence] \
   \begin{itemize}
@@
     \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
   \end{itemize}
-\end{theorem}
+\end{lemma}
 
-\begin{theorem}[Semantic Upgrade]
+\begin{lemma}[Semantic Upgrade]
   For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
   \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\hasty{\Gamma}{1}{a}{A}{q}}{}\]
-\end{theorem}
+\end{lemma}
+
+\begin{lemma}[Semantic Weakening] 
+  If \(\cwk{\Gamma}{\Delta}\) and \(\hasty{\Gamma}{p}{a}{A}{q}\), then
+  \[
+    \dnt{\hasty{\Gamma}{p}{a}{A}{q}} = \dnt{\cwk{\Gamma}{\Delta}};\dnt{\hasty{\Delta}{p}{a}{A}{q}}
+  \]
+\end{lemma}
 
@@
   \begin{gather*}
-    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_0(\dnt{\Theta}, \dnt{\Gamma})} \\
-    \dnt{\issubst{\substcons{\Gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
-      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{0}{a}{A}{q}} \\
+    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_1(\dnt{\Theta}, \dnt{\Gamma})} 
+    \\
+    \dnt{\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
+      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{1}{a}{A}{q}} \\
     \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id}
+    \\
+    \boxed{\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}: \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} 
   \end{gather*}
@@
 \end{figure}
+
+\begin{lemma}[Substitution Coherence] \
+  \begin{itemize}
+    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1: \strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \strictlbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{lemma}
+
+\begin{lemma}[Semantic Substitution Splitting] 
+  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
+  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
+  \[
+    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
+      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
+      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
+  \]
+\end{lemma}
 
@@
       = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}\)
-    \item \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
+    \item Given \(\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}\), \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
       ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      = \dnt{\haslb{\Theta}{[\gamma]t}{[\gamma]\ms{L}}}
-      ; \sorry \)
+      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
+      ; \dnt{\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}\)
   \end{itemize}
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+  \[
+    \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+    = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
+  \]
 \end{theorem}
 
-\TODO{\(\implies\) E-graph optimization}
+\begin{figure}
+  \begin{align*}
+    \dnt{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}
+      &= \dnt{H} \times \mc{C}_p^q(\dnt{\Gamma}, \dnt{A}) 
+      \\
+    \dnt{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}
+      &= \dnt{H} \times \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}}) 
+      \\
+    \dnt{H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
+      &= \dnt{H} \times \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
+      \\
+    \dnt{\cdot} 
+      &= \mb{1}
+  \end{align*}
+  \caption{Semantics for \isotopessa holes}
+  \Description{Semantics for isotope-SSA holes}
+  \label{fig:holes-semantics}
+\end{figure}
 
 \begin{figure}
-  \begin{center}
-    \TODO{this}
-  \end{center}
+  \begin{gather*}
+    \boxed{
+      \dnt{\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}}
+      : \dnt{H} \to \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})}
+      \\
+    \sorry
+  \end{gather*}
+  \caption{Semantics for \isotopessa terms with holes}
+  \Description{Semantics for isotope-SSA blocks with holes}
+  \label{fig:terms-with-holes-semantics}
+\end{figure}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{
+      \dnt{\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
+      : \dnt{H} \to \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}})}
+      \\
+    \sorry
+      \\  
+    \boxed{
+      \dnt{\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
+      : \dnt{H} \to \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
+      \\
+    \sorry
+  \end{gather*}
   \caption{Semantics for \isotopessa blocks with holes}
@@
 
-\begin{theorem}[Rewriting] \
-  \TODO{this}
+\begin{figure}
+  \begin{gather*}
+    \boxed{
+      \dnt{\isrw{\mc{H}}{I}{H}}
+      : \dnt{I} \to \dnt{H} 
+    }
+    \\
+    \sorry
+  \end{gather*}
+  \caption{Semantics for \isotopessa rewrites}
+  \Description{Semantics for isotope-SSA rewrites}
+  \label{fig:rewrite-semantics}
+\end{figure}
+
+\begin{lemma}[Rewriting Coherence] \
+  \begin{itemize}
+    \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{lemma}
+
+\begin{theorem}[Semantic Rewriting]
+  Given \(\isrw{\mc{H}}{I}{H}\), we have that
+  \begin{itemize}
+    \item For all \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(m \in \dnt{I}\), we have
+    \[
+      \dnt{\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}}(m)
+      = \dnt{\mhasty{H}{\Gamma}{p}{a}{A}{q}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
+    \]
+    \item For all \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(m \in \dnt{I}\), we have
+    \[
+      \dnt{\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}}(m)
+      = \dnt{\mhaslb{H}{\Gamma}{t}{\ms{L}}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
+    \]
+  \end{itemize}
+
+  Furthermore, we have that
+  \[
+    \dnt{\mhasty{\cdot}{\Gamma}{p}{a}{A}{q}}() = \dnt{\hasty{\Gamma}{p}{a}{A}{q}}
+    \qquad
+    \dnt{\mhaslb{\cdot}{\Gamma}{t}{\ms{L}}}() = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+  \]
 \end{theorem}
+
+\TODO{\(\implies\) E-graph optimization}
+
+\TODO{rewriting substitution + weakening?}
 

### 2023-10-18 pldi24/paper.tex
Tweaked substitution/semantic substitution, text stubs

@@
 \newcommand{\isrw}[3]{#1: #2 \mapsto #3}
-\newcommand{\strictlbsubst}[3]{#1: #2 \rightsquigarrow_= #3}
+% \newcommand{\strictlbsubst}[3]{#1: #2 \rightsquigarrow_= #3}
 
@@
 
+\TODO{some text about each typing judgement}
+
 \begin{figure}
@@
 \end{figure}
+
+\TODO{some text about each structural rule}
 
@@
 
+\TODO{some text about each typing rule}
+
 \begin{figure}
@@
 \end{figure}
+
+\TODO{some text about each block typing rule}
+
+\subsection{Syntactic Metatheory}
+
+\begin{theorem}[Weakening] \
+  \begin{itemize}
+    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\), \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
+    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
+    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then \(\haslb{\Gamma}{t}{\ms{K}}\)
+  \end{itemize}
+\end{theorem}
 
@@
         \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) &
-        ...
+        ... \\
+        \(\isrw{\mc{H}}{I}{H}\) &
+        ... \\
     \end{tabular}
@@
 
-\begin{theorem}[Weakening] \
-  \begin{itemize}
-    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\), \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
-    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
-    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then \(\haslb{\Gamma}{t}{\ms{K}}\)
-  \end{itemize}
-\end{theorem}
+\TODO{some text about each metatheoretic judgement}
 
@@
       {\lbsubst{\mc{L}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
-    \\
-    \prftree[r]{\rle{lb-strict}}
-      {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      {\ms{labels}(\ms{L}) = \ms{labels}(\ms{K})}
-      {\strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+    % \\
+    % \prftree[r]{\rle{lb-strict}}
+    %   {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+    %   {\ms{labels}(\ms{L}) = \ms{labels}(\ms{K})}
+    %   {\strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
   \end{gather*}
@@
 \end{figure}
+
+\TODO{some text about substitution}
 
@@
 \end{figure}
+
+\TODO{some text about substitution operations}
 
@@
 
+\TODO{some text about substitution weakening}
+
 \begin{theorem}[Substitution] 
@@
     \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
-    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), \(\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}\), and then \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
+    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{L}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
   \end{itemize}
 \end{theorem}
+
+\TODO{some text about substitution}
 
@@
 \end{theorem}
+
+\TODO{some text about label substitution}
 
@@
 \end{figure}
+
+\TODO{some text about terms with holes}
 
@@
 
+\TODO{some text about term with hole typing rules}
+
 \begin{figure}
@@
 
+\TODO{some text about block with hole typing rules}
+
 \begin{figure}
@@
 \end{figure}
+
+\TODO{some text about rewrite typing rules}
 
@@
   \end{itemize}
+
   Furthermore, we have that 
@@
 \end{theorem}
+'
+\TODO{some text about rewriting}
 
@@
 
-\TODO{premonoidal categories}
+\TODO{premonoidal/effectful categories}
 
 \TODO{Elgot structure}
+
+\TODO{relevant/affine/intuitionistic subcategories}
 
@@
 
+\TODO{some text about type/context semantics}
+
 \begin{figure}
@@
 
+\TODO{some text about structural semantics}
+
 \begin{figure}
@@
 \end{figure}
+
+\TODO{some text about term semantics}
 
@@
 
+\TODO{some text about block semantics}
+
+\subsection{Semantic Metatheory}
+
+\TODO{some text about semantic metatheory}
+
 \begin{lemma}[Coherence] \
@@
 
+\TODO{some text about coherence, implicit use above}
+
 \begin{lemma}[Semantic Upgrade]
@@
 
+\TODO{some text about upgrade}
+
 \begin{lemma}[Semantic Weakening] 
-  If \(\cwk{\Gamma}{\Delta}\) and \(\hasty{\Gamma}{p}{a}{A}{q}\), then
+  If \(\cwk{\Gamma}{\Delta}\), then 
+  \begin{itemize}
+    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
     \[
-    \dnt{\hasty{\Gamma}{p}{a}{A}{q}} = \dnt{\cwk{\Gamma}{\Delta}};\dnt{\hasty{\Delta}{p}{a}{A}{q}}
+      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
+      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+    \]
+    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
+    \[
+      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
+      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
+    \]
+  \end{itemize}
+
+  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
+  \[
+    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
+    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
   \]
 \end{lemma}
+
+\TODO{some text about semantic weakening}
 
@@
 \end{figure}
+
+\TODO{some text about substitution semantics}
 
@@
     \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1: \strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \strictlbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+    % \item Given any two derivations \(D_1: \strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \strictlbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
   \end{itemize}
@@
   \begin{itemize}
-    \item \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
+    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
+    \[
+      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
       ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}\)
-    \item Given \(\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}\), \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
+      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
+    \]
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}\), \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
+    \[
+      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
       ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
       = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
-      ; \dnt{\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}\)
+      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
+    \]
   \end{itemize}
@@
 
+\TODO{some text about semantic substitution}
+
 \begin{figure}
@@
 
+\TODO{some text about semantics of holes}
+
 \begin{figure}
@@
 
+\TODO{fill}
+
+\TODO{some text about semantics of terms with holes}
+
 \begin{figure}
@@
 \end{figure}
+
+\TODO{fill}
+
+\TODO{some text about semantics of blocks with holes}
 
@@
 \end{figure}
+
+\TODO{fill}
+
+\TODO{some text about semantics of rewrites}
 

### 2023-10-25 pldi24/paper.tex
Tap

@@
 
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix}
+
 \begin{figure}
@@
       {\lbsubst{\mc{L}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
-    % \\
-    % \prftree[r]{\rle{lb-strict}}
-    %   {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-    %   {\ms{labels}(\ms{L}) = \ms{labels}(\ms{K})}
-    %   {\strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
   \end{gather*}
@@
 
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix?}
+
 \TODO{some text about substitution weakening}
@@
     \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
-    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{L}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
+    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
+    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
+    \begin{itemize}
+      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
+      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
+      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
+    \end{itemize}
   \end{itemize}
@@
 
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix}
+
 \TODO{some text about substitution}
@@
 \begin{theorem}[Label Substitution] 
-  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), if \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
+  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
+  \begin{itemize}
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
+    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
+  \end{itemize}
 \end{theorem}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix}
 
@@
 \end{theorem}
-'
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix}
+
 \TODO{some text about rewriting}
@@
 
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix? This is just a trivial induction.}
+
 \TODO{some text about coherence, implicit use above}
@@
 
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix}
+
 \TODO{some text about semantic weakening}
@@
     \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-    % \item Given any two derivations \(D_1: \strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \strictlbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
   \end{itemize}
 \end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix? just trivial induction.}
 
@@
     \]
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}\), \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
+    \item For all 
+      \(\haslb{\Gamma}{t}{\ms{L}}\), 
+      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
+      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
     \[
@@
     \]
-  \end{itemize}
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+    \item For all 
+      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
+      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
+      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
+      \[
+        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
+        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
+        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
+        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+      \]
+  \end{itemize}
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
+  \begin{itemize} 
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
     \[
@@
     \]
+    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), we have
+    \[
+      \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      = \dnt{\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{L}}}
+    \]
+  \end{itemize}
 \end{theorem}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+\TODO{float proof to appendix}
 
@@
 \end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix? just trivial induction}
 
@@
 
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\TODO{pull down to appendix}
+
 \TODO{\(\implies\) E-graph optimization}
@@
 \TODO{Ye Olde Reader/Writer/State Transformer}
+
+\TODO{Is Elgot}
 

### 2023-10-31 abstract-freyd/paper.tex
Tap

--- 2023-10-18 pldi24/paper.tex
@@
 
-\title{Denotational Semantics for SSA with Weak Memory Operations}
+\title{Freyd Categories are SSA}
 
@@
 \orcid{0000-0002-6905-1303}
-
-\begin{abstract}
-  TODO THIS
-\end{abstract}
 
@@
 
-\section{Introduction}
-
-\TODO{THIS}
-
-\paragraph{Contributions}
-
-\begin{itemize}
-\item We have a syntax and type system for SSA-style programs.
-We include support for substructural types, and our type-theoretic syntax makes formulating inlining and rewriting easier than in the traditional CFG-based formulations.
-\item We show that our language has a compositional intepretation in any effectful category with an Elgot structure.
-We use this structure to give generic proofs of the expected substitution properties as well as prove the soundness of an equational theory for the language.
-\item Our generic equational theory justifies a variety of powerful optimisation techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and strength reduction. 
-\item We show how Elgot monads give rise to models of our language.
-We use this to derive semantics from a general trace monad, and show that various monad transformers such as state transformers preserve the needed Elgot structure.
-\item We use these tools to demonstrate the existence of a variety of concrete models satisfying these categorical axioms.
-Starting first from simple languges such as state plus printing, we also show how more challenging semantics such as the TSO semantics for weak memory also fit into our framework.
-This thus gives rise to an SSA-based IR with support for weak memory operations, which is fully semantically-justified.
-\item \TODO{We also have a small implementation, which \ldots}
-\item Our weak memory semantics and our results about Elgot monads are formalised in the Lean theorem prover. \TODO{Other stuff}
-\end{itemize}
-
-\section{SSA Syntax}
-
-\TODO{text}
-
-\begin{figure}
-  \begin{center}
-    \begin{grammar}
-      <\(A, B, C\)> ::= \(X\)
-      \;|\; \(A \otimes B\)
-
-      <\(a, b, c, e\)> ::= \(x\) 
-      \;|\; \(f\;a\)
-      \;|\; \((a, b)\) 
-      \;|\; \(()\) 
-      \;|\; \(\ctt\) 
-      \;|\; \(\cff\)
-      \;|\; \(\letexpr{x}{a}{e}\)
-      \;|\; \(\letexpr{(x, y)}{a}{e}\)
-      \;|\; \(\lbsplice{\ell}{x: A}{t}\)
-
-      <\(s, t\)> ::= \(\lbrb{\ell}{a}\) 
-      \;|\; \(\ite{e}{s}{t}\)
-      \;|\; \(\letstmt{x}{a}{t}\)
-      \;|\; \(\letstmt{(x, y)}{a}{t}\)
-      \;|\; \(\ewhere{t}{L}\)
-
-      <\(L\)> ::= \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{t}, L\)
-
-      <\(\Gamma\)> ::= \(\cdot\) \;|\; \(\Gamma, \thyp{x}{A}{q}\)
-
-      <\(\ms{L}\)> ::= \(\cdot\) \;|\; \(\ms{L}, \lbl{\ell}[\Gamma](x: A)\)
-
-      <\(p\)> ::= 0 \;|\; 1
-
-      <\(q\)> ::= \(\varnothing\) 
-      \;|\; \(\taff\) 
-      \;|\; \(\trel\) 
-      \;|\; \(\tint\)
-    \end{grammar}
-  \end{center}
-  \caption{Grammar for \isotopessa}
-  \Description{Grammar for isotope-SSA}
-  \label{fig:ssa-grammar}
-\end{figure}
-
-\TODO{late \ms{where}-binding, and other sugar (?)}
-
-\TODO{top-level functions}
-
-\TODO{table of typing judgements}
-
-\TODO{contexts, label-contexts}
-
-\begin{figure}
-  \begin{center}        
-    \begingroup
-    \renewcommand{\arraystretch}{1.5}
-    \setlength{\tabcolsep}{2em}
-    \begin{tabular}{rl}
-        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
-        \(\hasty{\Gamma}{p}{x}{A}{q}\) &
-        \(a\) is a term of type \(A\) in context \(\Gamma\) with purity \(p\) and linearity \(q\) \\
-        \(\haslb{\Gamma}{t}{\ms{L}}\) &
-        \(t\) is a block targeting label-set \(\ms{L}\) in context \(\Gamma\) \\
-        \(\lhaslb{\ms{L}}{L}{\ms{K}}\) &
-        The labels \(L\) send label-set \(\ms{L}\) to label-set \(\ms{K}\) \\
-        \(\csplits{\Gamma}{\Delta}{\Xi}\) &
-        The context \(\Gamma\) splits into \(\Delta\) and \(\Xi\) \\
-        \(\lwk{\ms{L}}{\ms{K}}\) &
-        The label-set \(\ms{L}\) weakens to the label-set \(\ms{K}\) \\
-        \(\tlin{A}{q}\) &
-        The type \(A\) has linearity \(q\) \\
-        \(\ltlin{A}{r}{q}\) &
-        The type \(A\) has linearity \(q \subseteq r\)
-        (i.e. \(\tlin{A}{q} \land q \subseteq r\)) \\
-        \(\tlin{\Gamma}{q}\) &
-        The context \(\Gamma\) has linearity \(q\) \\
-        \(\cwk{\Gamma}{\Delta}\) &
-        \(\Gamma\) is a weakening of \(\Delta\) 
-        (i.e. \(\csplits{\Gamma}{\Delta}{\cdot}\))
-    \end{tabular}
-    \endgroup
-  \end{center}
-  \caption{Typing judgements for \isotopessa}
-  \Description{Typing judgements for isotope-SSA}
-  \label{fig:ssa-judgements}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}    
-    \prftree[r]{\rle{base-lin}}{q \subseteq \ms{lin}(X)}{\tlin{X}{q}} \qquad
-    \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} \qquad
-    \prftree[r]{\rle{nil-lin}}{\tlin{\cdot}{q}} \qquad
-    \prftree[r]{\rle{cons-lin}}{\ltlin{A}{r}{q}}{\tlin{\Gamma}{q}}
-      {\tlin{\Gamma, \thyp{x}{A}{r}}{q}} \\
-    \prftree[r]{\rle{split-nil}}{\csplits{\cdot}{\cdot}{\cdot}} \qquad
-    \prftree[r]{\rle{split-left}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {r \subseteq q}
-      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta, \thyp{x}{A}{r}}{\Xi}} \qquad
-    \prftree[r]{\rle{split-right}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {r \subseteq q}
-      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta}{\Xi, \thyp{x}{A}{r}}} \\
-    \prftree[r]{\rle{split-dup}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\ltlin{A}{q}{\trel}}
-      {r, s \subseteq q}
-      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta, \thyp{x}{A}{r}}{\Xi, \thyp{x}{A}{s}}}
-      \qquad
-    \prftree[r]{\rle{split-drop}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\ltlin{A}{q}{\taff}}
-      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta}{\Xi}}
-      \\
-    \prftree[r]{\rle{join-nil}}{\lwk{\cdot}{\cdot}} \qquad
-    \prftree[r]{\rle{join-cons}}
-      {\lwk{\ms{L}}{\ms{K}}}
-      {\cwk{\Gamma}{\Delta}}
-      {\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}} 
-      \qquad
-    \prftree[r]{\rle{join-zero}}
-      {\lwk{\ms{L}}{\ms{K}}}
-      {\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}} 
-  \end{gather*}
-  \caption{Structural rules for \isotopessa}
-  \Description{Structural rules for isotope-SSA}
-  \label{fig:ssa-structural}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}    
-    \prftree[r]{\rle{var}}
-      {\cwk{\Gamma}{\thyp{x}{A}{q}}}
-      {\hasty{\Gamma}{p}{x}{A}{q}} \qquad
-    \prftree[r]{\rle{app}}
-      {f \in \mc{I}_p^q(A, B)}
-      {\hasty{\Gamma}{1}{a}{A}{q}}
-      {\hasty{\Gamma}{p}{f\;a}{B}{q}} \qquad
-    \prftree[r]{\rle{pair}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta}{1}{a}{A}{q}}
-      {\hasty{\Xi}{1}{b}{B}{q}}
-      {\hasty{\Gamma}{p}{(a, b)}{A \otimes B}{q}} \\
-    \prftree[r]{\rle{unit}}
-      {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
-    \prftree[r]{\rle{true}}
-      {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{p}{\ctt}{\mb{2}}{q}} \qquad
-    \prftree[r]{\rle{false}}
-      {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{p}{\cff}{\mb{2}}{q}} \\
-    \prftree[r]{\rle{let}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}{q}}
-      {\hasty{\Xi}{1}{a}{A}{q}}
-      {\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}{q}} \qquad
-    \prftree[r]{\rle{blk}}
-      {\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}}
-      {\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}} \\
-    \prftree[r]{\rle{let2}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}{q}}
-      {\hasty{\Xi}{1}{a}{A \otimes B}{q}}
-      {\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}{q}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa terms}
-  \Description{Typing rules for isotope-SSA terms}
-  \label{fig:ssa-term-typing}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}    
-    \prftree[r]{\rle{br}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
-      {\hasty{\Xi}{0}{a}{A}{\tint}}
-      {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} \qquad
-    \prftree[r]{\rle{ite}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta}{0}{e}{\mb{2}}{\tint}}
-      {\haslb{\Xi}{s}{\ms{L}}}
-      {\haslb{\Xi}{t}{\ms{L}}}
-      {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} \\
-    \prftree[r]{\rle{let-blk}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
-      {\hasty{\Xi}{1}{a}{A}{q}}
-      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} \\
-    \prftree[r]{\rle{let2-blk}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
-      {\hasty{\Xi}{1}{a}{A \otimes B}{q}}
-      {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \\
-    \prftree[r]{\rle{where}}
-      {\haslb{\Gamma}{t}{\ms{L}}}
-      {\lhaslb{\ms{L}}{L}{\ms{K}}}
-      {\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}} \qquad
-    \prftree[r]{\rle{nil-br}}
-      {\lhaslb{\ms{L}}{\cdot}{\ms{L}}} \qquad
-    \prftree[r]{\rle{cons-br}}
-      {\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      {\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}
-      {\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa blocks}
-  \Description{Typing rules for isotope-SSA blocks}
-  \label{fig:ssa-block-typing}
-\end{figure}
-
-\begin{figure}
-  \begin{center}        
-    \begingroup
-    \renewcommand{\arraystretch}{1.5}
-    \setlength{\tabcolsep}{2em}
-    \begin{tabular}{rl}
-        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
-        \(\issubst{\gamma}{\Theta}{\Gamma}\) &
-        ... \\
-        \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) &
-        ... \\
-        \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) &
-        ... \\
-        \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) &
-        ... \\
-        \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) &
-        ...
-    \end{tabular}
-    \endgroup
-  \end{center}
-  \caption{Metatheoretic typing judgements for \isotopessa}
-  \Description{Metatheoretic typing judgements for isotope-SSA}
-  \label{fig:ssa-meta-judgements}
-\end{figure}
-
-\begin{theorem}[Weakening] \
-  \begin{itemize}
-    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\), \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
-    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
-    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then \(\haslb{\Gamma}{t}{\ms{K}}\)
-  \end{itemize}
-\end{theorem}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\gamma: \ms{Var} \to \ms{Expr}}
-    \\
-    \prftree[r]{\rle{subst-nil}}
-      {\issubst{\gamma}{\cdot}{\cdot}}
-      \qquad
-    \prftree[r]{\rle{subst-cons}}
-      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
-      {\hasty{\Theta_x}{0}{\gamma(x)}{A}{q}}
-      {\tlin{\Theta_x}{q}}
-      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
-      {\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
-    \\
-    \boxed{\mc{L}: \ms{Label} \to \ms{Expr} \to \ms{Block} \text{ s.t. for fresh } x, [a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)}
-    \\
-    \prftree[r]{\rle{lb-subst-nil}}
-    {\lbsubst{\mc{L}}{\cdot}{\ms{K}}}
-      \qquad
-    \prftree[r]{\rle{lb-subst-cons}}
-      {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      {\haslb{\Gamma, \thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}}
-      {\lbsubst{\mc{L}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
-    \\
-    \prftree[r]{\rle{lb-strict}}
-      {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      {\ms{labels}(\ms{L}) = \ms{labels}(\ms{K})}
-      {\strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa substitutions. We assume that for all but finitely many \(x\), \(\lbl{\ell}\), \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) = \lbrb{\ell}{x}\).}
-  \Description{Typing rules for isotope-SSA substitutions}
-  \label{fig:ssa-subst-typing}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \subsubst{\gamma}{\gamma'} \iff 
-      \forall x, \gamma(x) \in \{x, \gamma'(x)\}
-    \\
-    \restrictsubst{\gamma}{\Gamma}(x) 
-      =
-      (\ms{if}\; x \in \Gamma
-      \;\ms{then}\;\gamma(x)
-      \;\ms{else}\;x)
-    \\
-    \stmtletsubst{\substctx{\gamma}{\cdot}}{t} = t
-    \qquad
-    \stmtletsubst{\substctx{\gamma}{\Gamma, x: A^q}}{t} 
-      = \letstmt{x}{\gamma(x)}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}
-    \qquad
-    \exprletsubst{\cdot}{e} = e
-    \qquad
-    \exprletsubst{\substctx{\gamma}{\Gamma, x: A^q}}{e} 
-      = \letstmt{x}{\gamma(x)}{\substctx{\gamma}{\Gamma}}
-    \\
-    \substlbs{\gamma}{\cdot}(\lbl{\ell}, x) = \lbrb{\ell}{x}
-    \qquad
-    \substlbs{\gamma}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}(\lbl{k}, x)
-      = 
-      (\ms{if}\; \lbl{k} = \lbl{\ell} 
-      \;\ms{then}\;\stmtletsubst{\substctx{\gamma}{\Gamma}}{\lbrb{\ell}{x}}
-      \;\ms{else}\;\substlbs{\gamma}{\ms{L}}(\lbl{\ell}, x))
-    \\
-    [\mc{L}](\lbrb{\ell}{a}) = \mc{L}(\lbl{\ell}, a)
-    \qquad
-    [\mc{L}]\ite{e}{s}{t} = \ite{e}{[\mc{L}]s}{[\mc{L}]t}
-    \\
-    [\mc{L}](\letstmt{x}{a}{t}) = (\letstmt{x}{a}{[\mc{L}]t})
-    \qquad
-    [\mc{L}](\letstmt{(x, y)}{a}{t}) = (\letstmt{(x, y)}{a}{[\mc{L}]t})
-    \\
-    [\mc{L}]\ewhere{t}{L} = \ewhere{[\mc{L}]t}{[\mc{L}]L}
-    \\
-    [\mc{L}]\cdot = \cdot
-    \qquad
-    [\mc{L}](\lwbranch{\ell}{x: A}{t})
-    = (\lwbranch{\ell}{x: A}{[\mc{L}]t})
-  \end{gather*}
-  \caption{Operations on \isotopessa substitutions. In particular, note that for all \(\Gamma\), \(\subsubst{\restrictsubst{\gamma}{\Gamma}}{\gamma}\)}
-  \Description{Operations on isotope-SSA substitutions}
-  \label{fig:ssa-subst-ops}
-\end{figure}
-
-\begin{lemma}[Substitution Weakening] 
-  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
-  \begin{itemize}
-    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
-    \item \(\issubst{\gamma_\Delta}{\Theta_\Delta}{\Delta}\)
-    \item \(\issubst{\gamma_\Xi}{\Theta_\Xi}{\Xi}\)
-  \end{itemize}
-  Furthermore, if \(\subsubst{\gamma}{\gamma'}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then \(\issubst{\gamma'}{\Theta}{\Gamma}\).
-\end{lemma}
-
-\begin{theorem}[Substitution] 
-  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  \begin{itemize}
-    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
-    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), \(\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}\), and then \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
-  \end{itemize}
-\end{theorem}
-
-\begin{theorem}[Label Substitution] 
-  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), if \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
-\end{theorem}
-
-\begin{figure}
-  \begin{grammar}
-    <\(\mhole{a}, \mhole{b}, \mhole{c}, \mhole{e}\)> ::= \(\lhole{X}\)
-    \;|\; \(x\) 
-    \;|\; \(f\;\mhole{a}\)
-    \;|\; \((\mhole{a}, \mhole{b})\) 
-    \;|\; \(()\) 
-    \;|\; \(\ctt\) 
-    \;|\; \(\cff\)
-    \;|\; \(\letexpr{x}{\mhole{a}}{\mhole{e}}\)
-    \;|\; \(\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}\)
-    \alt \(\lbsplice{\ell}{x: A}{\mhole{t}}\)
-
-    <\(\mhole{s}, \mhole{t}\)> ::= \(\lhole{T}\) 
-    \;|\; \(\lbrb{\ell}{\mhole{a}}\) 
-    \;|\; \(\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}\)
-    \;|\; \(\letstmt{x}{\mhole{a}}{\mhole{t}}\)
-    \;|\; \(\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}\)
-    \;|\; \(\ewhere{\mhole{t}}{\mhole{L}}\)
-
-    <\(\mhole{L}\)> ::= \(\lhole{W}\) \;|\; \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{\mhole{t}}, \mhole{L}\)
-
-    <\(H\)> ::= \(\cdot\) 
-    \;|\; \(H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}\)
-    \;|\; \(H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}\)
-    \;|\; \(H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\)
-  \end{grammar}
-  \caption{Grammar for \isotopessa terms and blocks with holes}
-  \Description{Grammar for isotope-SSA blocks with holes}
-  \label{fig:blocks-with-holes-grammar}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}    
-    \prftree[r]{\rle{hole-tm}}
-      {\tyhole{\lhole{X}}{\Delta}{p}{A}{q}}
-      {\cwk{\Gamma}{\Delta}}
-      {\mhasty{H}{\Gamma}{p}{\lhole{X}}{A}{q}} \qquad
-    \prftree[r]{\rle{var}}
-      {\cwk{\Gamma}{\thyp{x}{A}{q}}}
-      {\mhasty{H}{\Gamma}{p}{x}{A}{q}} \qquad
-    \prftree[r]{\rle{app}}
-      {f \in \mc{I}_p^q(A, B)}
-      {\mhasty{H}{\Gamma}{1}{\mhole{a}}{A}{q}}
-      {\mhasty{H}{\Gamma}{p}{f\;\mhole{a}}{B}{q}} \\
-    \prftree[r]{\rle{pair}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta}{1}{\mhole{a}}{A}{q}}
-      {\mhasty{H}{\Xi}{1}{\mhole{b}}{B}{q}}
-      {\mhasty{H}{\Gamma}{p}{(\mhole{a}, \mhole{b})}{A \otimes B}{q}} \\
-    \prftree[r]{\rle{unit}}
-      {\cwk{\Gamma}{\cdot}}
-      {\mhasty{H}{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
-    \prftree[r]{\rle{true}}
-      {\cwk{\Gamma}{\cdot}}
-      {\mhasty{H}{\Gamma}{p}{\ctt}{\mb{2}}{q}} \qquad
-    \prftree[r]{\rle{false}}
-      {\cwk{\Gamma}{\cdot}}
-      {\mhasty{H}{\Gamma}{p}{\cff}{\mb{2}}{q}} \\
-    \prftree[r]{\rle{let}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta, \thyp{x}{A}{}}{p}{\mhole{e}}{B}{q}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
-      {\mhasty{H}{\Gamma}{p}{\letexpr{x}{\mhole{a}}{\mhole{e}}}{B}{q}} \qquad
-    \prftree[r]{\rle{blk}}
-      {\mhaslb{H}{\Gamma}{\mhole{t}}{\llhyp{\ell}{\cdot}{A}}}
-      {\mhasty{H}{\Gamma}{0}{\lbsplice{\ell}{A}{\mhole{t}}}{A}{\varnothing}} \\
-    \prftree[r]{\rle{let2}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{\mhole{e}}{C}{q}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
-      {\mhasty{H}{\Gamma}{p}{\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}}{C}{q}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa terms with holes}
-  \Description{Typing rules for isotope-SSA terms with holes}
-  \label{fig:terms-with-holes-typing}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}    
-    \prftree[r]{\rle{hole-blk}}
-      {\blkhole{\lhole{T}}{\Delta}{\ms{L}} \in H}
-      {\cwk{\Gamma}{\Delta}}
-      {\mhaslb{H}{\Gamma}{\lhole{T}}{\ms{L}}} 
-      \qquad
-    \prftree[r]{\rle{br}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{\tint}}
-      {\mhaslb{H}{\Gamma}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
-      \\
-    \prftree[r]{\rle{ite}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta}{0}{\mhole{e}}{\mb{2}}{\tint}}
-      {\mhaslb{H}{\Xi}{\mhole{s}}{\ms{L}}}
-      {\mhaslb{H}{\Xi}{\mhole{t}}{\ms{L}}}
-      {\mhaslb{H}{\Gamma}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\
-    \prftree[r]{\rle{let-blk}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhaslb{H}{\Delta, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
-      {\mhaslb{H}{\Gamma}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
-    \prftree[r]{\rle{let2-blk}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhaslb{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{\mhole{t}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
-      {\mhaslb{H}{\Gamma}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
-    \prftree[r]{\rle{where}}
-      {\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
-      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
-      {\mhaslb{H}{\Gamma}{\ewhere{\mhole{t}}{\mhole{L}}}{\ms{K}}} 
-      \qquad
-    \prftree[r]{\rle{hole-cfg}}
-      {\cfghole{\lhole{W}}{\ms{L}}{\ms{K}} \in H}
-      {\lwk{\ms{M}}{\ms{K}}}
-      {\mlhaslb{\ms{L}}{H}{\lhole{W}}{\ms{K}}}
-      \\
-    \prftree[r]{\rle{nil-br}}
-      {\mlhaslb{H}{\ms{L}}{\cdot}{\ms{L}}} \qquad
-    \prftree[r]{\rle{cons-br}}
-      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      {\mhaslb{H}{\Gamma, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
-      {\mlhaslb{H}{\ms{L}}{\mhole{L}, \lwbranch{\ell}{x: A}{\mhole{t}}}{\ms{K}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa blocks with holes}
-  \Description{Typing rules for isotope-SSA blocks with holes}
-  \label{fig:blocks-with-holes-typing}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{H: \ms{Hole} \to \ms{MExpr} \sqcup \ms{MBlock}}
-    \\
-    \sorry
-  \end{gather*}
-  \caption{Typing rules for \isotopessa rewrites}
-  \Description{Typing rules for isotope-SSA rewrites}
-  \label{fig:rewrite-typing}
-\end{figure}
-
-\begin{theorem}[Rewriting]
-  Given \(\isrw{\mc{H}}{H}{I}\), we have
-  \begin{itemize}
-    \item If \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), then \(\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}\)
-    \item If \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), then \(\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}\)
-  \end{itemize}
-  Furthermore, we have that 
-  \begin{itemize}
-    \item \(\mhasty{\cdot}{\Gamma}{p}{a}{A}{q} \iff \hasty{\Gamma}{p}{a}{A}{q}\)
-    \item \(\mhaslb{\cdot}{\Gamma}{t}{\ms{L}} \iff \haslb{\Gamma}{t}{\ms{L}}\)
-  \end{itemize}
-\end{theorem}
-
-\section{SSA Semantics}
-
-\TODO{premonoidal categories}
-
-\TODO{Elgot structure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{A}}: |\mc{C}_1| \\
-    \dnt{X} = \ms{base}(X) 
-      \text{ where } 
-      \dnt{\mb{1}} = \mb{1}, 
-      \dnt{\mb{2}} = \mb{2}
-    \qquad \dnt{A \otimes B} = \dnt{A} \otimes \dnt{B} \\
-    \boxed{\dnt{\Gamma}}: |\mc{C}_1| \\
-    \dnt{\cdot} = \mb{1} \qquad 
-    \dnt{\Gamma, \thyp{x}{A}{q}} = \dnt{\Gamma} \otimes \dnt{A} \\
-    \boxed{\dnt{\ms{L}}}: |\mc{C}_0| \\
-    \dnt{\cdot} = \mb{0} \qquad
-    \dnt{\ms{L}, \llhyp{\ell}{\Gamma}{A}} = 
-      \dnt{\ms{L}} + \dnt{\Gamma} \otimes \dnt{A}
-  \end{gather*}
-  \caption{Semantics for \isotopessa types and contexts}
-  \Description{Semantics for isotope-SSA types and contexts}
-  \label{fig:ssa-type-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{\tlin{A}{\taff}: \mc{C}_1^\tint(\dnt{A}, \mb{1})}}
-      \\
-    \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
-    \dnt{\tlin{A \otimes B}{\taff}} 
-      = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
-      ; \lambda \\
-    \boxed{\dnt{\tlin{A}{\trel}}
-      : \mc{C}_1^\tint(\dnt{A}, \dnt{A} \otimes \dnt{A})}
-      \\
-    \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
-    \dnt{\tlin{A \otimes B}{\trel}} = 
-      \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
-      ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
-      \\
-    \boxed{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})}
-      \\
-    \dnt{\csplits{\cdot}{\cdot}{\cdot}} = \lambda^{-1} \qquad
-    \dnt{\csplits
-      {\Gamma, \thyp{x}{A}{q}}
-      {\Delta, \thyp{x}{A}{q}}
-      {\Xi}} 
-      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha;\dnt{\Delta} \otimes \sigma;\alpha 
-      \\
-    \dnt{\csplits
-      {\Gamma, \thyp{x}{A}{q}}
-      {\Delta}
-      {\Xi, \thyp{x}{A}{q}}} 
-      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha 
-      \\
-    \dnt{\csplits
-      {\Gamma, \thyp{x}{A}{q}}
-      {\Delta}
-      {\Xi, \thyp{x}{A}{q}}} \
-      = \dnt{\Gamma} \otimes \dnt{\tlin{A}{\trel}};
-        \alpha;
-        \dnt{\Delta} \otimes \sigma \otimes \dnt{A};
-        \alpha 
-        \\
-    \dnt{\csplits
-      {\Gamma, \thyp{x}{A}{q}}
-      {\Delta}
-      {\Xi}}
-      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-        \otimes \dnt{\tlin{A}{\taff}}
-      ; \rho 
-      \\
-    \boxed{\dnt{\cwk{\Gamma}{\Delta}}
-      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta})} 
-      \\
-      \dnt{\cwk{\Gamma}{\Delta}} 
-      = \dnt{\csplits{\Gamma}{\Delta}{\cdot}};\rho 
-      \\
-    \boxed{\dnt{\lwk{\ms{L}}{\ms{K}}}
-      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} 
-      \\
-    \dnt{\lwk{\cdot}{\cdot}} = \ms{id} \qquad
-    \dnt{\lwk
-      {\ms{L}, \llhyp{\ell}{\Gamma}{A}}
-      {\ms{K}, \llhyp{\ell}{\Delta}{A}}}
-      = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\cwk{\Gamma}{\Delta} \otimes \dnt{A})}{} 
-      \\
-    \dnt{\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      = \dnt{\lwk{\ms{L}}{\ms{K}}};\rho_+^{-1}
-      ; \dnt{\ms{K}} \oplus 0_{\dnt{\Gamma} \otimes \dnt{A}}
-  \end{gather*}
-  \caption{Semantics for \isotopessa structural judgements}
-  \Description{Semantics for isotope-SSA structural judgements}
-  \label{fig:ssa-structural-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{\hasty{\Gamma}{p}{a}{A}{q}}
-      : \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{x}{A}{q}} 
-      = \dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}
-      \qquad
-    \dnt{\hasty{\Gamma}{p}{f\;a}{B}{q}}
-      = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{p}
-      ; \ms{inst}_p(f) 
-      \\
-    \dnt{\hasty{\Gamma}{p}{(a, b)}{A \otimes B}{q}}
-      = \upg{(
-        \dnt{\csplits{\Gamma}{\Delta}{\Xi}};
-        \dnt{\hasty{\Delta}{1}{a}{A}{q}} \otimes
-        \dnt{\hasty{\Xi}{1}{b}{B}{q}}
-      )}{\uparrow p} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{()}{\mb{1}}{q}}
-      = \upg{\dnt{\cwk{\Gamma}{\cdot}}}{p} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{\ctt}{\mb{2}}{q}}
-      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\ctt)}{p}
-      \qquad
-    \dnt{\hasty{\Gamma}{p}{\cff}{\mb{2}}{q}}
-      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\cff)}{p} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}{q}}
-      = \upg{(
-        \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{q}}
-      )}{p};\dnt{\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}{q}} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}{q}}
-      = \upg{(
-        \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A \otimes B}{q}}
-      )}{p}
-      ; 
-      \\ \qquad \qquad \qquad \qquad \qquad \alpha
-      ; \dnt{\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}{q}}
-      \\
-    \dnt{\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}}
-      = \dnt{\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}};\alpha_+
-  \end{gather*}
-  \caption{Semantics for \isotopessa terms}
-  \Description{Semantics for isotope-SSA terms}
-  \label{fig:ssa-structural-terms}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      : \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}})} \\
-    \dnt{\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}}
-      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{0}{a}{A}{\tint}})}
-      ; \dnt{\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}} \\
-    \dnt{\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}}
-      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\hasty{\Delta}{0}{e}{\mb{2}}{\tint}} \otimes \dnt{\Xi})}
-      ; \ms{ite}_{\dnt{\Xi}} ;
-      \\ \qquad \qquad \qquad \qquad
-      [
-        \dnt{\haslb{\Xi}{s}{\ms{L}}}, 
-        \dnt{\haslb{\Xi}{t}{\ms{L}}} 
-      ]
-      \\
-    \dnt{\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}}
-      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A}{q}}
-      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
-      \\
-    \dnt{\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}}
-      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A \otimes B}{q}}
-      ; \alpha
-      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
-      \\
-    \dnt{\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}}
-      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      ; \dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger
-      \\
-    \boxed{\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}
-      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
-      \\
-    \dnt{\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
-      = \dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      ; [[\iota_0, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}], \iota_1]
-      \\
-    \dnt{\lhaslb{\ms{L}}{\cdot}{\ms{L}}}
-      = \iota_0
-  \end{gather*}
-  \caption{Semantics for \isotopessa blocks}
-  \Description{Semantics for isotope-SSA blocks}
-  \label{fig:ssa-structural-blocks}
-\end{figure}
-
-\begin{lemma}[Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\begin{lemma}[Semantic Upgrade]
-  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
-  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\hasty{\Gamma}{1}{a}{A}{q}}{}\]
-\end{lemma}
-
-\begin{lemma}[Semantic Weakening] 
-  If \(\cwk{\Gamma}{\Delta}\) and \(\hasty{\Gamma}{p}{a}{A}{q}\), then
-  \[
-    \dnt{\hasty{\Gamma}{p}{a}{A}{q}} = \dnt{\cwk{\Gamma}{\Delta}};\dnt{\hasty{\Delta}{p}{a}{A}{q}}
-  \]
-\end{lemma}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_1(\dnt{\Theta}, \dnt{\Gamma})} 
-    \\
-    \dnt{\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
-      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{1}{a}{A}{q}} \\
-    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id}
-    \\
-    \boxed{\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}: \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} 
-  \end{gather*}
-  \caption{Semantics for \isotopessa substitutions}
-  \Description{Semantics for isotope-SSA substitutions}
-  \label{fig:ssa-subst-semantics}
-\end{figure}
-
-\begin{lemma}[Substitution Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1: \strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \strictlbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\begin{lemma}[Semantic Substitution Splitting] 
-  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
-  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
-  \[
-    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
-      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
-      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
-  \]
-\end{lemma}
-
-\begin{theorem}[Semantic Substitution] 
-  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  \begin{itemize}
-    \item \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
-      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}\)
-    \item Given \(\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}\), \(\upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
-      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
-      ; \dnt{\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}\)
-  \end{itemize}
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), we have
-  \[
-    \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-    = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
-  \]
-\end{theorem}
-
-\begin{figure}
-  \begin{align*}
-    \dnt{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}
-      &= \dnt{H} \times \mc{C}_p^q(\dnt{\Gamma}, \dnt{A}) 
-      \\
-    \dnt{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}
-      &= \dnt{H} \times \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}}) 
-      \\
-    \dnt{H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
-      &= \dnt{H} \times \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
-      \\
-    \dnt{\cdot} 
-      &= \mb{1}
-  \end{align*}
-  \caption{Semantics for \isotopessa holes}
-  \Description{Semantics for isotope-SSA holes}
-  \label{fig:holes-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}}
-      : \dnt{H} \to \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})}
-      \\
-    \sorry
-  \end{gather*}
-  \caption{Semantics for \isotopessa terms with holes}
-  \Description{Semantics for isotope-SSA blocks with holes}
-  \label{fig:terms-with-holes-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
-      : \dnt{H} \to \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}})}
-      \\
-    \sorry
-      \\  
-    \boxed{
-      \dnt{\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
-      : \dnt{H} \to \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
-      \\
-    \sorry
-  \end{gather*}
-  \caption{Semantics for \isotopessa blocks with holes}
-  \Description{Semantics for isotope-SSA blocks with holes}
-  \label{fig:blocks-with-holes-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\isrw{\mc{H}}{I}{H}}
-      : \dnt{I} \to \dnt{H} 
-    }
-    \\
-    \sorry
-  \end{gather*}
-  \caption{Semantics for \isotopessa rewrites}
-  \Description{Semantics for isotope-SSA rewrites}
-  \label{fig:rewrite-semantics}
-\end{figure}
-
-\begin{lemma}[Rewriting Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\begin{theorem}[Semantic Rewriting]
-  Given \(\isrw{\mc{H}}{I}{H}\), we have that
-  \begin{itemize}
-    \item For all \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(m \in \dnt{I}\), we have
-    \[
-      \dnt{\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}}(m)
-      = \dnt{\mhasty{H}{\Gamma}{p}{a}{A}{q}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
-    \]
-    \item For all \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(m \in \dnt{I}\), we have
-    \[
-      \dnt{\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}}(m)
-      = \dnt{\mhaslb{H}{\Gamma}{t}{\ms{L}}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
-    \]
-  \end{itemize}
-
-  Furthermore, we have that
-  \[
-    \dnt{\mhasty{\cdot}{\Gamma}{p}{a}{A}{q}}() = \dnt{\hasty{\Gamma}{p}{a}{A}{q}}
-    \qquad
-    \dnt{\mhaslb{\cdot}{\Gamma}{t}{\ms{L}}}() = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-  \]
-\end{theorem}
-
-\TODO{\(\implies\) E-graph optimization}
-
-\TODO{rewriting substitution + weakening?}
-
-\section{Basic Models}
-
-\TODO{Ye Olde Trace Monad}
-
-\TODO{Is Elgot}
-
-\TODO{Ye Olde Nondeterministic Trace Monad}
-
-\TODO{Is Elgot}
-
-\TODO{Ye Olde Reader/Writer/State Transformer}
-
-\section{Weak Memory}
-
-\TODO{Pomsets...}
-
-\TODO{Form a monoid, so we can define...}
-
-\TODO{The SC Monad, which, via composition w/ the state transformer, lets us}
-
-\TODO{Build The Weak Memory Monad, which is too powerful, so we do...}
-
-\TODO{Karoubi envelope as premonoidal demonstration?}
-
-\section{Implementation}
-
 \TODO{this}

### 2023-10-31 pldi24/paper.tex
Tap

@@
 \newcommand{\lwk}[2]{#1 \rightsquigarrow #2}
-\newcommand{\tlin}[2]{\ms{lin}_{#2}(#1)}
-\newcommand{\ltlin}[3]{\ms{lin}_{#3}(#1^{#2})}
+\newcommand{\tlin}[2]{#2 \subseteq \ms{lin}(#1)}
+\newcommand{\ltlin}[3]{#3 \subseteq \ms{lin}(#1) \cap #2}
 \newcommand{\thyp}[3]{#1: {#2}^{#3}}
@@
 
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-%%\citestyle{acmauthoryear}
-
 \begin{document}
 
-\title{Denotational Semantics for SSA with Weak Memory Operations}
+\title{Effectful Semantics for Substructural SSA}
 
@@
 
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
 \begin{CCSXML}
@@
   <concept>
-  <concept_id>00000000.0000000.0000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+  <concept_id>10003752.10010124.10010131.10010133</concept_id>
+  <concept_desc>Theory of computation~Denotational semantics</concept_desc>
   <concept_significance>500</concept_significance>
@@
   <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>300</concept_significance>
+  <concept_id>10003752.10010124.10010131.10010137</concept_id>
+  <concept_desc>Theory of computation~Categorical semantics</concept_desc>
+  <concept_significance>500</concept_significance>
   </concept>
   <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
+  <concept_id>10003752.10003790.10011740</concept_id>
+  <concept_desc>Theory of computation~Type theory</concept_desc>
+  <concept_significance>500</concept_significance>
   </concept>
   <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
+  <concept_id>10003752.10003790.10011742</concept_id>
+  <concept_desc>Theory of computation~Separation logic</concept_desc>
+  <concept_significance>300</concept_significance>
   </concept>
@@
 
-\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
+\ccsdesc[500]{Theory of computation~Denotational semantics}
+\ccsdesc[500]{Theory of computation~Categorical semantics}
+\ccsdesc[500]{Theory of computation~Type theory}
+\ccsdesc[300]{Theory of computation~Separation logic}
 
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{TODO PUT KEYWORDS HERE}
+\keywords{SSA, Categorical Semantics, Elgot Structure, Effectful Category}
 
@@
 
-\TODO{THIS}
+Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, and MLIR are based.
+One of the core advantages of putting programs in SSA form is the ability to reason about programs in an equational manner by
+...
 
@@
     \begin{grammar}
-      <\(A, B, C\)> ::= \(X\)
+      <\(A, B, C\)> ::= 
+      \(X\)
+      \;|\; \(\mathbf{1}\)
+      \;|\; \(\mathbf{2}\)
       \;|\; \(A \otimes B\)
@@
       <\(q\)> ::= \(\varnothing\) 
-      \;|\; \(\taff\) 
-      \;|\; \(\trel\) 
-      \;|\; \(\tint\)
+      \;|\; \(\{\taff\}\) 
+      \;|\; \(\{\trel\}\) 
+      \;|\; \(\{\taff, \trel\}\)
     \end{grammar}
   \end{center}
-  \caption{Grammar for \isotopessa}
+  \caption{Grammar for \isotopessa, parametrized over a set of instructions \(f \in \mc{I}\)}
   \Description{Grammar for isotope-SSA}
@@
         \(\hasty{\Gamma}{p}{x}{A}{q}\) &
-        \(a\) is a term of type \(A\) in context \(\Gamma\) with purity \(p\) and linearity \(q\) \\
+        \(a\) is a term of type \(A\) in context \(\Gamma\) with purity \(p\) and quantity \(q\) \\
         \(\haslb{\Gamma}{t}{\ms{L}}\) &
@@
         The label-set \(\ms{L}\) weakens to the label-set \(\ms{K}\) \\
-        \(\tlin{A}{q}\) &
-        The type \(A\) has linearity \(q\) \\
-        \(\ltlin{A}{r}{q}\) &
-        The type \(A\) has linearity \(q \subseteq r\)
-        (i.e. \(\tlin{A}{q} \land q \subseteq r\)) \\
-        \(\tlin{\Gamma}{q}\) &
-        The context \(\Gamma\) has linearity \(q\) \\
+        % \(\tlin{A}{q}\) &
+        % The type \(A\) can be used with linearity \(q\) \\
+        % \(\ltlin{A}{r}{q}\) &
+        % The type \(A\) can be used with linearity \(q \subseteq r\)
+        % (i.e. \(\tlin{A}{q} \land q \subseteq r\)) \\
+        % \(\tlin{\Gamma}{q}\) &
+        % The context \(\Gamma\) has linearity \(q\) \\
         \(\cwk{\Gamma}{\Delta}\) &
@@
   \begin{gather*}    
-    \prftree[r]{\rle{base-lin}}{q \subseteq \ms{lin}(X)}{\tlin{X}{q}} \qquad
-    \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} \qquad
+    % \prftree[r]{\rle{base-lin}}{q \subseteq \ms{lin}(X)}{\tlin{X}{q}} \qquad
+    \prftree[r]{\rle{unit-lin}}{}{\tlin{\mathbf{1}}{q}} 
+    \qquad
+    \prftree[r]{\rle{bool-lin}}{}{\tlin{\mathbf{2}}{q}} 
+    \qquad
+    \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} \\
     \prftree[r]{\rle{nil-lin}}{\tlin{\cdot}{q}} \qquad
@@
       {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
-      {\hasty{\Xi}{0}{a}{A}{\tint}}
+      {\hasty{\Xi}{1}{a}{A}{\tint}}
       {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} \qquad
@@
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta}{0}{e}{\mb{2}}{\tint}}
+      {\hasty{\Delta}{1}{e}{\mb{2}}{\tint}}
       {\haslb{\Xi}{s}{\ms{L}}}
@@
       {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
-      {\hasty{\Xi}{1}{a}{A}{q}}
+      {\hasty{\Xi}{0}{a}{A}{q}}
       {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} \\
@@
       {\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
-      {\hasty{\Xi}{1}{a}{A \otimes B}{q}}
+      {\hasty{\Xi}{0}{a}{A \otimes B}{q}}
       {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \\
@@
       {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
-      {\hasty{\Theta_x}{0}{\gamma(x)}{A}{q}}
+      {\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}}
       {\tlin{\Theta_x}{q}}
@@
   \end{gather*}
-  \caption{Typing rules for \isotopessa substitutions. We assume that for all but finitely many \(x\), \(\lbl{\ell}\), \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) = \lbrb{\ell}{x}\).}
+  \caption{
+    Typing rules for \isotopessa substitutions. 
+    We generally assume that for all but finitely many variables \(x\) and labels \(\lbl{\ell}\), \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) = \lbrb{\ell}{x}\).
+  }
   \Description{Typing rules for isotope-SSA substitutions}
@@
   \begin{gather*}
-    \subsubst{\gamma}{\gamma'} \iff 
-      \forall x, \gamma(x) \in \{x, \gamma'(x)\}
-    \\
-    \restrictsubst{\gamma}{\Gamma}(x) 
-      =
-      (\ms{if}\; x \in \Gamma
-      \;\ms{then}\;\gamma(x)
-      \;\ms{else}\;x)
-    \\
+    % \restrictsubst{\gamma}{\Gamma}(x) 
+    %   =
+    %   (\ms{if}\; x \in \Gamma
+    %   \;\ms{then}\;\gamma(x)
+    %   \;\ms{else}\;x)
+    % \\
     \stmtletsubst{\substctx{\gamma}{\cdot}}{t} = t
@@
   \end{gather*}
-  \caption{Operations on \isotopessa substitutions. In particular, note that for all \(\Gamma\), \(\subsubst{\restrictsubst{\gamma}{\Gamma}}{\gamma}\)}
+  \caption{Operations on \isotopessa substitutions.}
   \Description{Operations on isotope-SSA substitutions}
@@
 
-\begin{lemma}[Substitution Weakening] 
+\begin{lemma}[Substitution Splitting] 
   If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
@@
     \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
-    \item \(\issubst{\gamma_\Delta}{\Theta_\Delta}{\Delta}\)
-    \item \(\issubst{\gamma_\Xi}{\Theta_\Xi}{\Xi}\)
-  \end{itemize}
-  Furthermore, if \(\subsubst{\gamma}{\gamma'}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then \(\issubst{\gamma'}{\Theta}{\Gamma}\).
+    \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
+    \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
+  \end{itemize}
 \end{lemma}
@@
       {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{\tint}}
+      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{\tint}}
       {\mhaslb{H}{\Gamma}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
@@
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta}{0}{\mhole{e}}{\mb{2}}{\tint}}
+      {\mhasty{H}{\Delta}{1}{\mhole{e}}{\mb{2}}{\tint}}
       {\mhaslb{H}{\Xi}{\mhole{s}}{\ms{L}}}
@@
       {\mhaslb{H}{\Delta, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
+      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{q}}
       {\mhaslb{H}{\Gamma}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
@@
       {\mhaslb{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{\mhole{t}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
+      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A \otimes B}{q}}
       {\mhaslb{H}{\Gamma}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
@@
       = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{0}{a}{A}{\tint}})}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{\tint}})}
       ; \dnt{\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}} \\
@@
       = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\hasty{\Delta}{0}{e}{\mb{2}}{\tint}} \otimes \dnt{\Xi})}
+      ; \dnt{\hasty{\Delta}{1}{e}{\mb{2}}{\tint}} \otimes \dnt{\Xi})}
       ; \ms{ite}_{\dnt{\Xi}} ;
@@
       = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A}{q}}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A}{q}}
       ; \dnt{\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
@@
       = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{1}{a}{A \otimes B}{q}}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A \otimes B}{q}}
       ; \alpha

### 2023-11-01 abstract-freyd/paper.tex
Began talk abstract

@@
 %% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
+% \setcopyright{acmcopyright}
+% \copyrightyear{2018}
+% \acmYear{2018}
+% \acmDOI{XXXXXXX.XXXXXXX}
 
@@
 
-\title{Freyd Categories are SSA}
-
-\author{Neel Krishnaswami}
-\email{nk480@cl.cam.ac.uk}
-\orcid{0000-0003-2838-5865}
+\title{SSA is Freyd Categories}
 
@@
 \orcid{0000-0002-6905-1303}
-
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-\begin{CCSXML}
-<ccs2012>
- <concept>
-  <concept_id>00000000.0000000.0000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>500</concept_significance>
- </concept>
- <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
 
@@
 
-\TODO{this}
+Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure \cite{llvm}, GCC's GIMPLE IR, Cranelift \cite{cranelift}, and MLIR \cite{mlir} are based.
+One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.
 
-\section{Related Work}
+One of the interesting observations of the MLIR project is that SSA is very widely applicable, being useful for reasoning about everything from high-level machine learning operations to low-level GPU code, and even hardware design and simulation, as in the CIRCT project \cite{circt}.
+This hints that SSA is a natural syntax for some fundamental, hopefully compositional, structure, much like how the simply typed lambda calculus is a natural syntax for arbitrary Cartesian closed categories.
+In this talk, we would like to argue that SSA without control-flow is a natural syntax for \textit{Freyd categories}, which we then provide a graphical syntax for in terms of string diagrams as in \cite{promonad}.
+We then argue that SSA with \textit{terminating} control-flow (i.e., in which the basic blocks form a DAG) corresponds exactly to \textit{distributive} Freyd categories, while allowing general control-flow corresponds to equipping a distributive Freyd category with an \textit{Elgot structure}; we then give a string-diagrammatic graphical representation of such programs similar to traditional control-flow graphs.
 
-\TODO{\cite{promonad}}
+We do all this by giving an inductive, type-theoretic presentation of SSA which we equip with a categorical semantics, proving substitution and the soundness of various basic optimizations and rewrites.
+Finally, we wish to give some examples of various models and their applications, highlighting how they appear in diverse areas of computer science including classical compiler development, weak memory, quantum circuits, and hardware design.
 
-\TODO{\cite{linear-state-usage}}
-
-\TODO{\cite{ssa-is-fun}}
-
-\TODO{\cite{sparky}}
+Time permitting, we also hope to explain how to generalize our IR with support for both substructural types and impure substitutions, which require us to move to the more general setting of \textit{distributive effectful categories}, and give a few applications of this more general system in separation logic and reasoning about nondeterminism.
 

### 2023-11-01 pldi24/paper.tex
Began talk abstract

@@
 \setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
+\copyrightyear{2023}
+\acmYear{2023}
 \acmDOI{XXXXXXX.XXXXXXX}
@@
 
-Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, and MLIR are based.
-One of the core advantages of putting programs in SSA form is the ability to reason about programs in an equational manner by
-...
+Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, Cranelift, and MLIR are based.
+One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.
+The semantics of SSA, however, can be quite complex to reason about, especially when considering the somewhat unintuitive scoping rules of variables and \(\phi\)-function arguments.
+
+In this paper, we give a type-theoretic account of SSA-style programs, which we extend with support for substructural types.
+This allows us to perform proofs and inductive definitions over SSA programs much more easily, which we take advantage of to give SSA a categorical semantics in terms of distributive effectful categories equipped with an Elgot structure.
+We then give some models, and prove various optimizations sound on them.
 

### 2023-11-01 abstract-freyd/paper.tex
Tap

@@
 
-Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure \cite{llvm}, GCC's GIMPLE IR, Cranelift \cite{cranelift}, and MLIR \cite{mlir} are based.
+Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure \cite{llvm}, GCC's GIMPLE IR \cite{gcc}, Cranelift \cite{cranelift}, and MLIR \cite{mlir} are based.
 One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.
 
-One of the interesting observations of the MLIR project is that SSA is very widely applicable, being useful for reasoning about everything from high-level machine learning operations to low-level GPU code, and even hardware design and simulation, as in the CIRCT project \cite{circt}.
-This hints that SSA is a natural syntax for some fundamental, hopefully compositional, structure, much like how the simply typed lambda calculus is a natural syntax for arbitrary Cartesian closed categories.
-In this talk, we would like to argue that SSA without control-flow is a natural syntax for \textit{Freyd categories}, which we then provide a graphical syntax for in terms of string diagrams as in \cite{promonad}.
-We then argue that SSA with \textit{terminating} control-flow (i.e., in which the basic blocks form a DAG) corresponds exactly to \textit{distributive} Freyd categories, while allowing general control-flow corresponds to equipping a distributive Freyd category with an \textit{Elgot structure}; we then give a string-diagrammatic graphical representation of such programs similar to traditional control-flow graphs.
+One of the core observations of the MLIR project is that SSA is very widely applicable, being useful for representing everything from high-level machine learning operations to low-level GPU code, and even hardware designs, as in the CIRCT project \cite{circt}.
+This hints that SSA is a natural syntax for some fundamental compositional structure, much like how the simply typed lambda calculus is a natural syntax for Cartesian closed categories.
 
-We do all this by giving an inductive, type-theoretic presentation of SSA which we equip with a categorical semantics, proving substitution and the soundness of various basic optimizations and rewrites.
-Finally, we wish to give some examples of various models and their applications, highlighting how they appear in diverse areas of computer science including classical compiler development, weak memory, quantum circuits, and hardware design.
+In this talk, we would like to argue that SSA without control-flow is a syntax for \textit{Freyd categories}, which we then provide a graphical syntax for in terms of string diagrams as in \cite{promonad}.
+We then argue that SSA with \textit{terminating} control-flow (i.e., in which the basic blocks form a DAG) corresponds exactly to \textit{distributive} Freyd categories, while SSA with general control-flow corresponds to equipping a distributive Freyd category with an \textit{Elgot structure}; we then give a string-diagrammatic graphical representation of such programs based on traditional control-flow graphs.
 
-Time permitting, we also hope to explain how to generalize our IR with support for both substructural types and impure substitutions, which require us to move to the more general setting of \textit{distributive effectful categories}, and give a few applications of this more general system in separation logic and reasoning about nondeterminism.
+We do this by providing a type-theoretic presentation of SSA, which we call \isotopessa, which is designed to be amenable to inductive proofs and definitions.
+By using this formalization, we can equip SSA with a categorical semantics, which we show satisfies important properties such as semantic substitution and congruence.
+We then show how various optimizations, such as strength reduction, constant propagation, global value numbering, loop hoisting, and E-graph rewriting are compatible with our semantics.
+Finally, we give various models of our categorical semantics exhibiting interesting features, building up to a simple model of TSO-style weak memory based on \cite{sparky} which treats the interactions between threads in a compositional manner.
+
+Time permitting, we also hope to explain how to generalize our IR with support for substructural types and impure substitutions, which require us to move to the more general setting of \textit{distributive effectful categories}, and give a few applications of this more general system relating to separation logic, quantum computing, and nondeterminism.
 

### 2023-11-01 pldi24/paper.tex
Tap

@@
 
+\TODO{abstract?}
+
 \section{Introduction}
@@
 
-In this paper, we give a type-theoretic account of SSA-style programs, which we extend with support for substructural types.
-This allows us to perform proofs and inductive definitions over SSA programs much more easily, which we take advantage of to give SSA a categorical semantics in terms of distributive effectful categories equipped with an Elgot structure.
-We then give some models, and prove various optimizations sound on them.
+\TODO{write about ``what is SSA", article style}
+
+\TODO{basic blocks, strange scoping rules...}
 
@@
 
+\section{Separation Logic}
+
+\TODO{State transformer on traces}
+
+\TODO{Monoidal functor, yielding sliding}
+
 \section{Weak Memory}

### 2023-11-01 pldi24/paper.tex
Tap

@@
 One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.
+
+In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks of the following form:
+\TODO{this}
+A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
+\TODO{this}
+is not in SSA form, but can easily be converted into SSA form by rewriting it as
+\TODO{this}
+Unfortunately, things get a little more tricky when we try to define variables in loops or branches. For example, consider the following code:
+\TODO{this}
+There are two competing definitions of x in the left and right blocks, and it’s unclear how to disambiguate between them.
+The usual solution is to introduce a \(\phi\)-node.
+In particular, we can write:
+\TODO{this}
+\TODO{segue}
 The semantics of SSA, however, can be quite complex to reason about, especially when considering the somewhat unintuitive scoping rules of variables and \(\phi\)-function arguments.
 
-\TODO{write about ``what is SSA", article style}
-
-\TODO{basic blocks, strange scoping rules...}
+\TODO{basic-blocks with arguments}
+
+\TODO{dominance frontiers}
+
+\TODO{inductive presentation}
+
+\TODO{what we're doing}
 

### 2023-11-01 pldi24/paper.tex
Tap

@@
 \usepackage{prftree}
+
+\usepackage{listings}
+\usepackage{xcolor}
+
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
+
+\lstdefinestyle{mystyle}{
+%    backgroundcolor=\color{backcolour},   
+    commentstyle=\color{codegreen},
+    keywordstyle=\color{magenta},
+    numberstyle=\tiny\color{codegray},
+    stringstyle=\color{codepurple},
+    basicstyle=\ttfamily\footnotesize,
+    breakatwhitespace=false,         
+    breaklines=true,                 
+    captionpos=b,                    
+    keepspaces=true,                 
+    numbers=left,                    
+    numbersep=5pt,                  
+    showspaces=false,                
+    showstringspaces=false,
+    showtabs=false,                  
+    tabsize=2
+}
+
+\lstset{style=mystyle}
 
@@
 
-In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks of the following form:
-\TODO{this}
-A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
-\TODO{this}
-is not in SSA form, but can easily be converted into SSA form by rewriting it as
-\TODO{this}
-Unfortunately, things get a little more tricky when we try to define variables in loops or branches. For example, consider the following code:
-\TODO{this}
-There are two competing definitions of x in the left and right blocks, and it’s unclear how to disambiguate between them.
-The usual solution is to introduce a \(\phi\)-node.
-In particular, we can write:
-\TODO{this}
-\TODO{segue}
-The semantics of SSA, however, can be quite complex to reason about, especially when considering the somewhat unintuitive scoping rules of variables and \(\phi\)-function arguments.
-
-\TODO{basic-blocks with arguments}
-
-\TODO{dominance frontiers}
-
-\TODO{inductive presentation}
-
-\TODO{what we're doing}
+\TODO{text...}
 
@@
 \end{itemize}
+
+\section{Background}
+
+In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks of the following form:
+\begin{lstlisting}
+'label:
+  var = instruction
+  ...
+  var = instruction
+  terminator
+\end{lstlisting}
+A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
+\begin{lstlisting}
+'entry:
+  x = 7
+  x = add x 2
+  ret x
+\end{lstlisting}
+is not in SSA form, but can easily be converted into SSA form by rewriting it as
+\begin{lstlisting}
+'entry:
+  x0 = 7
+  x1 = add x0 2
+  ret x1
+\end{lstlisting}
+Unfortunately, things get a little more tricky when we try to define variables in loops or branches. For example, consider the following code:
+\begin{lstlisting}
+'entry:
+  br arg0 'left `right
+'left:
+  x = 7
+  jmp 'exit
+'right:
+  x = 3
+  jmp 'exit
+'exit:
+  ret x
+\end{lstlisting}
+There are two competing definitions of \(x\) in the left and right blocks, and it’s unclear how to disambiguate between them.
+The usual solution is to introduce a \(\phi\)-node.
+In particular, we can write:
+\begin{lstlisting}
+'entry:
+  br arg0 'left `right
+'left:
+  x0 = 7
+  jmp 'exit
+'right:
+  x1 = 3
+  jmp 'exit
+'exit:
+  x2 = phi 'left x0 'right x1
+  ret x2
+\end{lstlisting}
+\TODO{make above into graphs?}
+\TODO{segue}
+The semantics of SSA, however, can be quite complex to reason about, especially when considering the somewhat unintuitive scoping rules of variables and \(\phi\)-function arguments.
+
+\TODO{basic-blocks with arguments}
+
+\TODO{dominance frontiers}
+
+\TODO{inductive presentation}
+
+\TODO{what we're doing}
 

### 2023-11-01 pldi24/paper.tex
Rewrite typing rules, rewrite semantics

@@
   \begin{gather*}
-    \boxed{H: \ms{Hole} \to \ms{MExpr} \sqcup \ms{MBlock}}
-    \\
-    \sorry
+    \boxed{H: \ms{Hole} \to \ms{MExpr} \sqcup \ms{MBlock} \sqcup \ms{MCfg}}
+    \\
+    \prftree[r]{\rle{rw-nil}}
+      {\isrw{\mc{H}}{H}{\cdot}}
+    \qquad
+    \prftree[r]{\rle{rw-tm}}
+      {\isrw{\mc{H}}{I}{H}}
+      {\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}
+      {\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}
+    \\
+    \prftree[r]{\rle{rw-blk}}
+      {\isrw{\mc{H}}{I}{H}}
+      {\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}
+      {\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
+    \qquad
+    \prftree[r]{\rle{rw-cfg}}
+      {\isrw{\mc{H}}{I}{H}}
+      {\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}
+      {\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
   \end{gather*}
@@
 
+\TODO{define a substructural effectful category, rewrite as ``substructural effectful category \(\mc{C}\)" equipped with proper structrure}
+
+The semantics for an \isotopessa dialect are given by:
+\begin{itemize}
+  \item A family of categories \(\mc{C}^q_1\) parametrized by quantity \(q \subseteq \{\trel, \taff\}\)
+  \item \(q \subseteq q'\) implies \(\mc{C}^{q'}_1\) is a wide subcategory of \(\mc{C}^q_1\)
+  \item All \(\mc{C}^q_1\) share symmetric monoidal structure (i.e., the inclusion functors between them are strict symmetric monoidal functors).
+  \item A symmetric premonoidal category \(\mc{C}_0\)
+  \item An identity on objects functor functor \(\upg{\cdot}{}: \mc{C}^\varnothing_1 \to \mc{C}_0\) which preserves all symmetric premonoidal structure
+  \item A mapping from base types \(A\) to objects in \(|\mc{C}|\)
+  \item A mapping from functions \(f \in \mc{I}_p^q(A, B)\) to morphisms \(\dnt{f}: \mc{C}_p^q(A, B)\), where \(\mc{C}_0^q = \mc{C}_0\)
+  \item 
+\end{itemize}
+
 \TODO{premonoidal/effectful categories}
@@
   \begin{gather*}
-    \boxed{\dnt{A}}: |\mc{C}_1| \\
+    \boxed{\dnt{A}}: |\mc{C}| \\
     \dnt{X} = \ms{base}(X) 
@@
     \qquad \dnt{A \otimes B} = \dnt{A} \otimes \dnt{B} \\
-    \boxed{\dnt{\Gamma}}: |\mc{C}_1| \\
+    \boxed{\dnt{\Gamma}}: |\mc{C}| \\
     \dnt{\cdot} = \mb{1} \qquad 
     \dnt{\Gamma, \thyp{x}{A}{q}} = \dnt{\Gamma} \otimes \dnt{A} \\
-    \boxed{\dnt{\ms{L}}}: |\mc{C}_0| \\
+    \boxed{\dnt{\ms{L}}}: |\mc{C}| \\
     \dnt{\cdot} = \mb{0} \qquad
@@
   \begin{gather*}
-    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_1(\dnt{\Theta}, \dnt{\Gamma})} 
+    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_1^\varnothing(\dnt{\Theta}, \dnt{\Gamma})} 
     \\
@@
     \\
-    \sorry
+    \dnt{\isrw{\mc{H}}{I}{\cdot}}\;i = ()
+    \qquad
+    \dnt{\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}\;i
+    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}\;i)
+    \\
+    \dnt{\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
+    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}\;i)
+    \\
+    \dnt{\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
+    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}\;i)
   \end{gather*}

### 2023-11-01 pldi24/paper.tex
Tap

@@
   \item All \(\mc{C}^q_1\) share symmetric monoidal structure (i.e., the inclusion functors between them are strict symmetric monoidal functors).
-  \item A symmetric premonoidal category \(\mc{C}_0\)
+  \item A symmetric premonoidal category \(\mc{C}_0\) equipped with an Elgot structure
+  \item An object \(\mb{2}\), morphisms \(\ms{tt}, \ms{ff}: \mc{C}_1^\tint(\mb{1}, \mb{2})\), and a family of isomorphisms \(\ms{ite}_A: 2 \otimes A \simeq A + A\) such that 
+  \begin{itemize}
+    \item \(\ms{ite}_A^{-1} = [\lambda_A^{-1};\ms{tt} \otimes A, \lambda_A^{-1};\ms{ff} \otimes A]\)
+    \item \(\mb{2} \otimes f;\ms{ite}_A = \ms{ite}_A;f + f\)
+  \end{itemize}
   \item An identity on objects functor functor \(\upg{\cdot}{}: \mc{C}^\varnothing_1 \to \mc{C}_0\) which preserves all symmetric premonoidal structure
-  \item A mapping from base types \(A\) to objects in \(|\mc{C}|\)
-  \item A mapping from functions \(f \in \mc{I}_p^q(A, B)\) to morphisms \(\dnt{f}: \mc{C}_p^q(A, B)\), where \(\mc{C}_0^q = \mc{C}_0\)
-  \item 
+  \item A mapping from base types \(X\) to objects in \(\dnt{X}: |\mc{C}|\)
+  \item A mapping from functions \(f \in \mc{I}_p^q(A, B)\) to morphisms \(\dnt{f}: \mc{C}_p^q(\dnt{A}, \dnt{B})\), where \(\mc{C}_0^q = \mc{C}_0\)
+  \item A set of \textit{affine} objects \(A, B \in \ms{Aff}(\mc{C}) \subseteq |\mc{C}|\) equipped with maps \(\ms{drop}(A): \mc{C}_1^\tint(A, \mb{1})\) such that:
+  \begin{itemize}
+    \item \(\ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B); \lambda_\mb{1}\)
+    \item For all \(f: \mc{C}^\taff_1(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
+  \end{itemize}
+  \item A set of \textit{relevant} objects \(A, B \in \ms{Rel}(\mc{C}) \subseteq |\mc{C}|\) equipped with maps \(\ms{split}(A): \mc{C}_1^\tint(A, A \times A)\) such that:
+  \begin{itemize}
+    \item \(\ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
+    \item For all \(f: \mc{C}^\trel_1(A, B), f;\ms{split}(B) = \ms{split}(A);f \otimes f\)
+    \item \textit{Commutativity}: \(\ms{split}(A);\sigma = \ms{split}(A)\)
+    \item \textit{Associativity}: \(\ms{split}(A);A \otimes \ms{split}(A) = \ms{split}(A);\ms{split}(A) \otimes A;\alpha\)
+    \item \textit{Comonoid}: if \(A \in \ms{Aff}(\mc{C})\), \(\ms{split}(A);\ms{drop}(A) \otimes A = \lambda^{-1}\) and \(\ms{split}(A);A \otimes \ms{drop}(A) = \rho^{-1}\)
+  \end{itemize}
 \end{itemize}

### 2023-11-02 pldi24/paper.tex
Defined substructural monoidal category

@@
 
-\TODO{define a substructural effectful category, rewrite as ``substructural effectful category \(\mc{C}\)" equipped with proper structrure}
+\begin{definition}[Substructural Monoidal Category]
+  A \textbf{substructural monoidal category} is a tuple \((\mc{C}, \mc{A}, \mc{R})\) where:
+  \begin{itemize}
+    \item \(\mc{C}\) is a family of categories \(\mc{C}^q\) parametrized by subsets \(q \subseteq \{\ms{a}, \ms{r}\}\) such that
+    \begin{itemize}
+      \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the same. We will write this set as \(|\mc{C}|\)
+      \item All \(\mc{C}^q\) share symmetric monoidal structure, and are individually closed under the tensor products of morphisms
+      \item \(\mc{C}^{q \cap q'} = \mc{C}^q \cap \mc{C}^{q'}\)
+    \end{itemize}
+    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of objects closed under tensor products, such that
+    \begin{itemize}
+      \item \(\mb{1} \in \mc{A}\)
+      \item Every \(A\) in \(\mc{A}\) is equipped with a morphism \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
+      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
+      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\{\ms{a}\}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
+    \end{itemize}
+    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of objects closed under tensor products, such that
+    \begin{itemize}
+      \item \(\mb{1} \in \mc{R}\)
+      \item Every \(A\) in \(\mc{R}\) is equipped with a morphism \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
+      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
+      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
+      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
+      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
+    \end{itemize}
+  \end{itemize}
+  We call a substructural monoidal category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
+Many of our examples will satisfy this structure.
+\end{definition}
+A simple example of a substructural Cartesian category is the category of relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\).
+In particular, we can define:
+\begin{equation}
+  \begin{aligned}
+    \ms{Rel}^\varnothing(A, B) 
+    &= \ms{Rel}(A, B) 
+    &&= A \to \mc{P}(B) \\
+    \ms{Rel}^{\{\ms{a}\}}(A, B) 
+    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
+    &&= A \to \mc{P}^+(B) \\
+    \ms{Rel}^{\{\ms{r}\}}(A, B) 
+    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
+    &&= A \to \ms{Option}(B) \\
+    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
+    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
+    &&= A \to B
+  \end{aligned}
+\end{equation}
+Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and \(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and that they satisfy the axioms of a commutative comonoid, as desired.
+
+Note that we cannot simply have \(\ms{Rel}^{\{\ms{a}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) = \varnothing\), we have
+\begin{equation}
+  (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a) = \{()\}
+\end{equation}
+We will see that this corresponds to the fact that the following programs are not semantically equal
+\begin{equation}
+  \ms{let}\;x = \ms{loop}(); e \not\simeq e \qquad \text{where} \qquad x \notin \ms{fv}(e)
+\end{equation}
+Similarly, we cannot simply have \(\ms{Rel}^{\{\ms{r}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)| \supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
+\begin{equation}
+  (R;\ms{split}(B))(a) = \{(b, b) \mid b \in R(a)\} \neq (\ms{split}(A);R)(a) = R(a) \times R(a) 
+\end{equation}
+since only the left-hand side contains the pair \((b_0, b_1)\), this corresponding to the fact that the following programs are not semantically equal
+\begin{equation}
+  \ms{let}\;x = \ms{rand}(); (x, x) \not\simeq (\ms{rand}(), \ms{rand}())
+\end{equation}
+Rather than directly give a concrete example of a general substructural monoidal category, it makes more sense to first jump to the next definition, which will provide us with one more naturally:
+\begin{definition}[Susbtructural Effectful Category]
+  \ \TODO{this}
+\end{definition}
+
+\TODO{state monad}
+
+\TODO{separation logic}
 

### 2023-11-02 pldi24/paper.tex
Defined substructural effectful category

@@
 \begin{definition}[Susbtructural Effectful Category]
-  \ \TODO{this}
+  A \textbf{substructural effectful category} \(\mc{C}_1 \to \mc{C}_0\) consists of
+  \begin{itemize}
+    \item A substructural monoidal category \(\mc{C}_1\)
+    \item A symmetric premonoidal category \(\mc{C}_0\)
+    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0\) preserving all symmetric premonoidal structure (which naturally induces functors \(\upg{\cdot}{}: \mc{C}_1^q \to \mc{C}_0\) which do the same)
+  \end{itemize}
+  We'll define \(\mc{C}_0^q\) to simply be \(\mc{C}_0\) to allow us to write \(\mc{C}_p^q\) for all \(p \in \{0, 1\}\), \(q \subseteq \{\ms{a}, \ms{r}\}\).
+Following this convention, we define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\) and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply be the identity functor on \(\mc{C}_1^q\).
 \end{definition}

### 2023-11-03 pldi24/paper.tex
more intro

--- 2023-11-01 pldi24/paper.tex
@@
 \newcommand{\sorry}{\textcolor{red}{\textbf{sorry}}}
+
+\newcommand{\todo}[1]{\stepcounter{todos} \textcolor{red}{TODO \arabic{todos}:} #1}
 
@@
 \begin{abstract}
-  TODO THIS
+  Static single-assignment form (SSA) is one of the most widely used
+  compiler intermediate representations. In this paper, we build on
+  the observation that SSA can be seen as a language of basic blocks
+  with arguments (or first-order procedures tail-calling one another)
+  to give a type theory for SSA programs. Notably, the type system and
+  equational theory of our language is designed to support
+  substituting impure terms for variables, which makes it easy to
+  validate compiler optimisations such as hoisting assignments out of
+  loops.
+
+  We give a categorical axiomatisation of our type theory and show
+  that our equational theory is sound with respect to it. Our
+  categorical semantics generalises the notions of premonoidal
+  category and Freyd category to support substructural features,
+  including both support for substructural types, as well as
+  F\"{u}hrmann's notion of central, copyable and discardable
+  morphisms.
+
+  Finally, we exhibit a number of concrete models of our categorical
+  axiomatisation, including in particular a model of TSO-style weak
+  memory based on the semantics of \citet{sparky}. This demonstrates
+  that our approach is strong enough to connect machine models of
+  concurrency to compiler IRs supporting a rich collection of sound
+  rewrites.
 \end{abstract}
@@
 
-\TODO{abstract?}
-
 \section{Introduction}
 
-Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, Cranelift, and MLIR are based.
+Static Single Assignment, or SSA, has been a cornerstone of compiler
+design ever since its introduction by \cite{ssa-original} in the late
+1980s. Older compiler intermediate representations permitted the same
+local variable to be assigned multiple times, whereas SSA requires
+each variable bindings to be immutable -- they cannot be modified
+after initialisation. This makes many dataflow analyses easier to
+compute, and (just as in functional languages) makes it much easier to
+justify compiler optimisations in terms of equational rewrites.
+
+In fact, the relationship between SSA form and functional programming
+has been known for decades: \citet{kelsey-ssa-cps} showed that SSA can
+be seen as a subset of continuation-passing style, and
+\citet{appel-ssa} built on this to observe that every SSA program can
+be viewed as a collection of first-order functions which tail-call one
+another. This makes the equivalence of the functional view to the
+traditional $\phi$-node presentation very easy to see. In the
+functional style, each tail call lists its arguments. In contrast, a
+$\phi$-node collects together the arguments from every callsite, which
+lets the tail-calls/jumps to omit their arguments. Thus, the main
+difference between SSA and a functional program is where in the
+program text the arguments to a tail call are listed.
+
+\TODO{Example here?}
+
+This ``basic blocks with arguments'' presentation of SSA has become
+very popular in industry, with compiler backends such as
+Cranelift~\cite{cranelift}, MLIR~\cite{mlir}, and SIL~\cite{SIL} all
+using this representation. This tight connection might suggest that
+the semantics of SSA representations is trivial: we can just interpret
+an SSA program as a corresponding functional program, and then rewrite
+the SSA programs using the equational theory of functional programs.
+
+While this is a perfectly sound thing to do, it is not strong enough
+to write a good compiler. Compiler IRs like SSA are fundamentally
+about effectful programs, and the equational theory of effectful
+functional programs is very weak. Instead of a full substitution rule,
+the equational theory of impure functional languages only justifies
+substituting values for variables (the $\beta$-value rule). This weak
+substitution rule suffices to validate some basic compiler
+optimisations such as constant propagation and copy propagation, but
+is not strong enough to justify many other essential compiler
+optimisations. For example, hoisting a store out of a loop or
+conditional involves moving an effectful term (an assignment) out of
+its original position in the program. As another example, dead store
+elimination removes an effectful term from the program altogether. \todo{Another example?}
+
+These examples illustrate that many optimisations require taking
+\emph{effectful} terms and moving them around and substituting them
+for variables.
+
+% 
+% , on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, Cranelift, and MLIR are based.
 One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.
-
+% 
 \TODO{text...}

### 2023-11-03 pldi24/paper.tex
Tao

--- 2023-11-02 pldi24/paper.tex
@@
     \end{itemize}
-    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of objects closed under tensor products, such that
+    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects closed under tensor products, such that
     \begin{itemize}
@@
     \end{itemize}
-    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of objects closed under tensor products, such that
+    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects closed under tensor products, such that
     \begin{itemize}
@@
   \end{itemize}
+  We call the morphisms in \(\mc{C}^{\{\ms{a}\}}\) \textbf{affine}, the morphisms in \(\mc{C}^{\{\ms{r}\}}\) \textbf{relevant}, and the morphisms in \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}.
+We call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the other \(\mc{C}^q\) \textbf{linear}.
 We call a substructural monoidal category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
-Many of our examples will satisfy this structure.
 \end{definition}
@@
   \begin{itemize}
-    \item A substructural monoidal category \(\mc{C}_1\)
+    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms in this category \textbf{central}
     \item A symmetric premonoidal category \(\mc{C}_0\)
@@
 
-\TODO{state monad}
+We can obtain an example of a substructural effectful category quite easily by, considering the power-set monad above, applying the state transformer to yield the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to \mc{P}(A \times S)\).
+Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
+\(
+  \upg{R}{} = \lambda a, s. R(a) \times \{s\}
+\), we can verify that this satisfies the axioms of a substructural effectful category.
+Note that here the \textit{central} morphisms are precisely those which do not read from or write to the state, and therefore which commute with all other morphisms.
+
+We can further use this example to get a substructural \textit{monoidal} category by considering a primitive form of separation logic.
 
 \TODO{separation logic}
+
+\TODO{distributive category}
+
+\TODO{Elgot category}
 

### 2023-11-04 pldi24/paper.tex
Fiddled with citations
Separation logic example

@@
 
-We can further use this example to get a substructural \textit{monoidal} category by considering a primitive form of separation logic.
-
-\TODO{separation logic}
+We can further use this example to get a substructural \textit{monoidal} category by considering a primitive form of separation logic: let's introduce a grammar of predicates \(\varphi, \psi ::= \top, \ms{S}, \bot\) for programs which cannot access state, can access state, and are invalid, respectively.
+The separating conjunction operator on these predicates is then given by
+\begin{equation}
+  \top * \varphi = \varphi * \top = \varphi
+  \qquad
+  \bot * \varphi = \varphi * \bot = \bot
+  \qquad
+  \ms{S} * \ms{S} = \bot
+\end{equation}
+We'll then define a category \(\ms{SRel}\) with objects of the form \((A, \varphi)\) and morphisms
+\begin{equation}
+  \begin{gathered}
+    \boxed{
+      \ms{SRel}((A, \varphi), (B, \psi)) 
+      \subseteq A \to \ms{S} \to \mc{P}(A \times S) 
+    }
+    \\
+    \begin{aligned}
+      \ms{SRel}((A, \top), (B, \top)) &= \{f \in A \to S \to \mc{P}(A \times S) \mid \exists g. f = \lambda a, s. g(a) \times \{s\}\} \\
+      \ms{SRel}((A, \ms{S}), (B, \ms{S})) &= A \to S \to \mc{P}(A \times S) \\
+      \ms{SRel}((A, \varphi), (B, \psi)) &= \{\lambda a, s. \varnothing\} \quad \text{otherwise}
+    \end{aligned}
+  \end{gathered}
+\end{equation}
+Note that it is precisely the \textit{central} morphisms \(\ms{C}_0(A, B)\) which lie in \(\ms{SRel}((A, \top), (B, \top))\).
+We can then define the tensor product on objects as \((A, \varphi) \otimes (B, \psi) = (A \otimes B, \varphi * \psi)\), and, for \(f \in \mathsf{SRel}((A, \varphi), (B, \psi))\), and tensor product functors
+\begin{equation}
+  \begin{aligned}
+  f \otimes (C, \phi) &= f \otimes C,
+  & (C, \phi) \otimes f &= C \otimes f
+  && \text{if}\; \varphi * \phi \neq \bot, \psi * \phi \neq \bot \\
+  f \otimes (C, \phi) &= \lambda a, s. \varnothing,
+  & (C, \phi) \otimes f &=  \lambda a, s. \varnothing
+  && \text{otherwise}
+  \end{aligned} 
+\end{equation}
+Lifting associators, unitors, and symmetries from \(\mc{C}_0\) in the obvious manner (setting them to \(\lambda a, s.
+\varnothing\) where necessary), we can verify that \(\ms{SRel}\) is a \textit{premonoidal} category.
+we can see that it is in fact \textit{monoidal} by nothing that, for any \(f, g\), \(f \ltimes g = f \rtimes g\) since either \(f\) is central, \(g\) is central, or their product is the empty morphism \(\lambda a, s.
+\varnothing\).
+Hence, to get a \textit{substructural} monoidal category, we simply need to define:
+\begin{equation}
+  \begin{gathered}
+  \ms{Aff}(\ms{SRel}) = \ms{Rel}(\ms{SRel}) = \{(A, \top) | A \in \ms{Set}\} \neq |\ms{SRel}|
+  \\
+  \begin{aligned}
+    \ms{SRel}^\varnothing((A, \phi), (B, \psi))
+    &= \ms{SRel}((A, \phi), (B, \psi)) \\
+    \ms{SRel}^{\{\ms{a}\}}((A, \phi), (B, \psi)) 
+    &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \geq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
+    \ms{SRel}^{\{\ms{r}\}}((A, \phi), (B, \psi)) 
+    &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \leq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
+    \implies \ms{SRel}^{\{\ms{a}, \ms{r}\}}((A, \phi), (B, \psi)) 
+    &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| = 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
+  \end{aligned}
+  \end{gathered}
+\end{equation}
+We can verify that the associators, unitors, and symmetry lie in \(\ms{SRel}^{\{\ms{a}, \ms{r}\}}\) and that all other desired axioms are satisfied.
 
@@
 
-\TODO{\cite{ssa-is-fun}}
-
 \TODO{\cite{sparky}}

### 2023-11-04 pldi24/paper.tex
Tap
Taps

@@
 
+\subsection{Isotope Models}
+
+\TODO{intro text}
+
+\TODO{define premonoidal category}
+
+\TODO{pull Elgot structure up here?}
+
 \begin{definition}[Substructural Monoidal Category]
@@
 
-\TODO{distributive category}
-
-\TODO{Elgot category}
+\TODO{relate separation logic trick to \cite{promonad}, \cite{linear-state-usage}}
+
+\begin{definition}[Distributive category] 
+  A \textbf{distributive category} is a premonoidal category with coproducts equipped with a natural family of central isomorphisms \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\).
+\end{definition}
+
+\TODO{text}
+
+\begin{definition}[Elgot category] 
+  An \textbf{Elgot category} is a category \(\mc{C}\) with coproducts equipped with an operator \(\cdot^\dagger: \mc{C}(A, B + A) \to \mc{C}(A, B)\) satisfying the following properties:
+  \begin{itemize}
+    \item \textbf{Fixpoint:} ...
+    \item \textbf{Naturality:} ...
+    \item \textbf{Codiagonal:} ...
+    \item \textbf{Uniformity:} ...
+  \end{itemize}
+  A monad \(\mb{T}\) on \(\mc{C}\) such that its Kleisli cateogry \(\mc{C}_{\mb{T}}\) is Elgot is called an \textbf{Elgot monad}.
+\end{definition}
+
+\TODO{text}
+
+\TODO{distributivity of monads}
+
+\TODO{result: if a monad is Elgot, the state/reader/writer transformers yield Elgot monads}
+
+\TODO{example of Elgot monads: traces}
+
+\subsection{Denotational Semantics}
+
+\TODO{text}
+
+\TODO{refactor below}
 
@@
 \end{itemize}
-
-\TODO{premonoidal/effectful categories}
-
-\TODO{Elgot structure}
-
-\TODO{relevant/affine/intuitionistic subcategories}
 
@@
 
-\section{Basic Models}
+\subsection{Optimizations}
+
+\TODO{this...}
+
+\section{Concrete Models}
+
+\subsection{Basic Models}
 
@@
 
-\section{Separation Logic}
+\TODO{already done above?}
+
+\subsection{Separation Logic}
 
@@
 
-\section{Weak Memory}
+\subsection{Weak Memory}
+
+\TODO{\cite{sparky}}
 
@@
 
+\TODO{remove? time + space...}
+
 \section{Related Work}
 
-\TODO{\cite{promonad}}
-
-\TODO{\cite{linear-state-usage}}
-
-\TODO{\cite{sparky}}
+\TODO{this}
 

### 2023-11-04 pldi24/paper.tex
Elgot category properties
Began work on refactoring

@@
 \newcommand{\trel}{\ms{r}}
-\newcommand{\tint}{\infty}
+\newcommand{\tint}{\{\ms{a}, \ms{r}\}}
 \newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
@@
 
-\TODO{define premonoidal category}
+\TODO{define premonoidal category?}
 
 \TODO{pull Elgot structure up here?}
+
+\TODO{substructural premonoidal category, relationship?
+This makes \(\mc{C}_0^q\) have structure, which is useful at least for things that are affine but not central (see: dead code elimination).
+Duplication is Less Useful (TM) but nice for symmetry.}
 
@@
 \end{definition}
+\TODO{Define \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) syntax? Make lattice generated by syntax?}
 A simple example of a substructural Cartesian category is the category of relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\).
@@
   \begin{itemize}
-    \item \textbf{Fixpoint:} ...
-    \item \textbf{Naturality:} ...
-    \item \textbf{Codiagonal:} ...
-    \item \textbf{Uniformity:} ...
+    \item \textbf{Fixpoint:} \(f;[\ms{id}, f^\dagger] = f^\dagger\)
+    \item \textbf{Naturality:} \((f;g + \ms{id})^\dagger = f^\dagger;g\)
+    \item \textbf{Codiagonal:} \(\forall f \in \mc{C}(A, (B + A) + A), (f^\dagger)^\dagger = (f;[\ms{id}, \iota_1])^\dagger\)
+    \item \textbf{Uniformity:}
+    \(
+      h;f = g;\ms{id} + h \implies h;f^\dagger = g^\dagger
+    \)
   \end{itemize}
@@
 \end{definition}
+\TODO{why do these matter? they show up in the proofs, but...}
+Some important properties of all Elgot categories include:
+\begin{itemize}
+  \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id}, (h;[\iota_0, g])^\dagger]\).
+\TODO{Proposition 16 of \cite{goncharov-guarded-unguarded}; also Lemma 31 of \cite{goncharov-squaring}}
+  \item \textbf{Squaring:} \((f;[\iota_0, f])^\dagger = f^\dagger\). \TODO{Lemma 31 of \cite{goncharov-squaring}}
+\end{itemize}
+\TODO{Hyland and Hasegawa traces?}
+\TODO{intuition? Can draw diagrams, but those take up space}
+
+\TODO{results on distributivity of monads}
+
+\TODO{result: if a monad is Elgot, the state/reader/writer transformers yield Elgot monads}
+
+\TODO{example of Elgot monads: traces}
+
+\subsection{Denotational Semantics}
 
@@
 
-\TODO{distributivity of monads}
-
-\TODO{result: if a monad is Elgot, the state/reader/writer transformers yield Elgot monads}
-
-\TODO{example of Elgot monads: traces}
-
-\subsection{Denotational Semantics}
-
-\TODO{text}
-
-\TODO{refactor below}
-
-The semantics for an \isotopessa dialect are given by:
-\begin{itemize}
-  \item A family of categories \(\mc{C}^q_1\) parametrized by quantity \(q \subseteq \{\trel, \taff\}\)
-  \item \(q \subseteq q'\) implies \(\mc{C}^{q'}_1\) is a wide subcategory of \(\mc{C}^q_1\)
-  \item All \(\mc{C}^q_1\) share symmetric monoidal structure (i.e., the inclusion functors between them are strict symmetric monoidal functors).
-  \item A symmetric premonoidal category \(\mc{C}_0\) equipped with an Elgot structure
-  \item An object \(\mb{2}\), morphisms \(\ms{tt}, \ms{ff}: \mc{C}_1^\tint(\mb{1}, \mb{2})\), and a family of isomorphisms \(\ms{ite}_A: 2 \otimes A \simeq A + A\) such that 
-  \begin{itemize}
-    \item \(\ms{ite}_A^{-1} = [\lambda_A^{-1};\ms{tt} \otimes A, \lambda_A^{-1};\ms{ff} \otimes A]\)
-    \item \(\mb{2} \otimes f;\ms{ite}_A = \ms{ite}_A;f + f\)
-  \end{itemize}
-  \item An identity on objects functor functor \(\upg{\cdot}{}: \mc{C}^\varnothing_1 \to \mc{C}_0\) which preserves all symmetric premonoidal structure
-  \item A mapping from base types \(X\) to objects in \(\dnt{X}: |\mc{C}|\)
-  \item A mapping from functions \(f \in \mc{I}_p^q(A, B)\) to morphisms \(\dnt{f}: \mc{C}_p^q(\dnt{A}, \dnt{B})\), where \(\mc{C}_0^q = \mc{C}_0\)
-  \item A set of \textit{affine} objects \(A, B \in \ms{Aff}(\mc{C}) \subseteq |\mc{C}|\) equipped with maps \(\ms{drop}(A): \mc{C}_1^\tint(A, \mb{1})\) such that:
-  \begin{itemize}
-    \item \(\ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B); \lambda_\mb{1}\)
-    \item For all \(f: \mc{C}^\taff_1(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
-  \end{itemize}
-  \item A set of \textit{relevant} objects \(A, B \in \ms{Rel}(\mc{C}) \subseteq |\mc{C}|\) equipped with maps \(\ms{split}(A): \mc{C}_1^\tint(A, A \times A)\) such that:
-  \begin{itemize}
-    \item \(\ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
-    \item For all \(f: \mc{C}^\trel_1(A, B), f;\ms{split}(B) = \ms{split}(A);f \otimes f\)
-    \item \textit{Commutativity}: \(\ms{split}(A);\sigma = \ms{split}(A)\)
-    \item \textit{Associativity}: \(\ms{split}(A);A \otimes \ms{split}(A) = \ms{split}(A);\ms{split}(A) \otimes A;\alpha\)
-    \item \textit{Comonoid}: if \(A \in \ms{Aff}(\mc{C})\), \(\ms{split}(A);\ms{drop}(A) \otimes A = \lambda^{-1}\) and \(\ms{split}(A);A \otimes \ms{drop}(A) = \rho^{-1}\)
-  \end{itemize}
-\end{itemize}
+The semantics for an \isotopessa dialect with a set of types \(\mc{T}\) and instructions \(\mc{I}\) are given by:
+\begin{itemize}
+  \item A \textit{substructural effectful category} \(\mc{C}\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot}, equipped with
+  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies X \in \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies X \in \ms{Rel}(\mc{C})\)
+  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\) to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
+\end{itemize}
+\TODO{make aff and rel syntax nicer/more consistent here? Subcategory syntax would be nice for lemmas and just coherence.}
 

### 2023-11-04 pldi24/paper.tex
Commented out splices, work on substructural category definitions

@@
 \newcommand{\bsplice}[3]{#1(#2)\;\{#3\}}
-\newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
+%\newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
 \newcommand{\csplits}[3]{#1 \mapsto #2;#3}
@@
 \newcommand{\trel}{\ms{r}}
-\newcommand{\tint}{\{\ms{a}, \ms{r}\}}
+\newcommand{\tint}{{\{\ms{a}, \ms{r}\}}}
 \newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
@@
       \;|\; \(\letexpr{(x, y)}{a}{e}\)
-      \;|\; \(\lbsplice{\ell}{x: A}{t}\)
+      % \;|\; \(\lbsplice{\ell}{x: A}{t}\)
 
@@
       {\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}{q}} \qquad
-    \prftree[r]{\rle{blk}}
-      {\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}}
-      {\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}} \\
+    % \prftree[r]{\rle{blk}}
+    %   {\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}}
+    %   {\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}} 
+      \\
     \prftree[r]{\rle{let2}}
@@
     \;|\; \(\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}\)
-    \alt \(\lbsplice{\ell}{x: A}{\mhole{t}}\)
+    % \alt \(\lbsplice{\ell}{x: A}{\mhole{t}}\)
 
@@
       {\mhasty{H}{\Gamma}{p}{\letexpr{x}{\mhole{a}}{\mhole{e}}}{B}{q}} \qquad
-    \prftree[r]{\rle{blk}}
-      {\mhaslb{H}{\Gamma}{\mhole{t}}{\llhyp{\ell}{\cdot}{A}}}
-      {\mhasty{H}{\Gamma}{0}{\lbsplice{\ell}{A}{\mhole{t}}}{A}{\varnothing}} \\
+    % \prftree[r]{\rle{blk}}
+    %   {\mhaslb{H}{\Gamma}{\mhole{t}}{\llhyp{\ell}{\cdot}{A}}}
+    %   {\mhasty{H}{\Gamma}{0}{\lbsplice{\ell}{A}{\mhole{t}}}{A}{\varnothing}} 
+      \\
     \prftree[r]{\rle{let2}}
@@
 
-\subsection{Isotope Models}
-
 \TODO{intro text}
 
+\subsection{Substructural Categories}
+
+\TODO{intro text}
+
 \TODO{define premonoidal category?}
 
-\TODO{pull Elgot structure up here?}
-
-\TODO{substructural premonoidal category, relationship?
-This makes \(\mc{C}_0^q\) have structure, which is useful at least for things that are affine but not central (see: dead code elimination).
-Duplication is Less Useful (TM) but nice for symmetry.}
-
-\begin{definition}[Substructural Monoidal Category]
-  A \textbf{substructural monoidal category} is a tuple \((\mc{C}, \mc{A}, \mc{R})\) where:
+\begin{definition}[Substructural (Pre)monoidal Category]
+  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A}, \mc{R})\) where:
   \begin{itemize}
@@
       \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the same. We will write this set as \(|\mc{C}|\)
-      \item All \(\mc{C}^q\) share symmetric monoidal structure, and are individually closed under the tensor products of morphisms
+      \item All \(\mc{C}^q\) share symmetric (pre)monoidal structure, and are individually closed under the tensor products of morphisms
       \item \(\mc{C}^{q \cap q'} = \mc{C}^q \cap \mc{C}^{q'}\)
@@
       \item \(\mb{1} \in \mc{A}\)
-      \item Every \(A\) in \(\mc{A}\) is equipped with a morphism \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
+      \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
       \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
@@
       \item \(\mb{1} \in \mc{R}\)
-      \item Every \(A\) in \(\mc{R}\) is equipped with a morphism \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
+      \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
       \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
@@
       \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
+      \item \(\forall A, B \in \mc{R}, \forall f: \mc{C}^{\{\ms{r}\}}(A, B), f;\ms{split}(B) = \ms{split}(A);f \ltimes f = \ms{split}(A);f\rtimes f\)
     \end{itemize}
@@
 We call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the other \(\mc{C}^q\) \textbf{linear}.
-We call a substructural monoidal category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
+We call a substructural \textit{monoidal} category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
 \end{definition}
@@
 \end{equation}
-Rather than directly give a concrete example of a general substructural monoidal category, it makes more sense to first jump to the next definition, which will provide us with one more naturally:
+To get a substructural premonoidal category, we can apply the state monad transformer to the power set monad to obtain the Kleisli category with morphisms of the form \(\ms{Set}_{\ms{St}\;\mc{P}\;S}(A, B) = A \to S \to \mc{P}(B \times S)\).
+Once again, it is clear that all objects have split and drop morphisms inherited from \(\ms{Set}\), and therefore are affine and relevant
+
 \begin{definition}[Susbtructural Effectful Category]
@@
   \begin{itemize}
-    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms in this category \textbf{central}
-    \item A symmetric premonoidal category \(\mc{C}_0\)
-    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0\) preserving all symmetric premonoidal structure (which naturally induces functors \(\upg{\cdot}{}: \mc{C}_1^q \to \mc{C}_0\) which do the same)
-  \end{itemize}
-  We'll define \(\mc{C}_0^q\) to simply be \(\mc{C}_0\) to allow us to write \(\mc{C}_p^q\) for all \(p \in \{0, 1\}\), \(q \subseteq \{\ms{a}, \ms{r}\}\).
-Following this convention, we define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\) and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply be the identity functor on \(\mc{C}_1^q\).
+    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms in this category \textbf{central} or \textbf{linear}
+    \item A substructural premonoidal category \(\mc{C}_0\)
+    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0^\varnothing\) which:
+    \begin{itemize}
+      \item Preserves all symmetric premonoidal structure (which naturally induces functors \(\upg{\cdot}{}: \mc{C}_1^q \to \mc{C}_0\) which do the same)
+      \item Sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\)
+    \end{itemize}
+  \end{itemize}
+  We define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\) and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply be the identity functor on \(\mc{C}_1^q\).
+
+  Morphisms in \(\mc{C}_p^{\{\ms{a}\}}\) are called \textbf{copyable}, morphisms in \(\mc{C}_p^{\{\ms{r}\}}\) are called \textbf{discardable},
+  while morphisms in \(\mc{C}_1^{\{\ms{a}\}}, \mc{C}_1^{\{\ms{r}\}}, \mc{C}_1^{\{\ms{a}, \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and \textbf{intuitionistic} or \textbf{pure} respectively.
 \end{definition}
@@
 
+In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a substructural premonoidal category, with \(\mc{C}^\varnothing = \mc{C}\) and \(\mc{C}^{\{\ms{a}, \ms{r}\}}, \mc{C}^{\{\ms{a}\}}, \mc{C}^{\{\ms{r}\}}\) the subcategory consisting only of the associators, unitors, and symmetry, and only the identity object considered affine and relevant.
+On the other extreme, every \textit{Cartesian} category \(\mc{C}\) can be viewed as a substructural Cartesian category with all morphisms both relevant and affine; in fact, this is the case \textit{if and only if} a category is Cartesian, hence the term "substructural Cartesian category."
+
+We can use these observations to deduce that every strong monad \(\mb{T}\) on a substructural monoidal category \(\mc{C}\) induces a substructural premonoidal category \(\mc{C}_{\mb{T}}\) (the Kleisli category) with affine/relevant objects the same as in \(\mc{C}\), and affine/relevant morphisms containing the image of all affine/relevant morphisms in \(\mc{C}\) under the inclusion functor \(\mc{C} \to \mc{C}_{\mb{T}}\).
+
 \begin{definition}[Distributive category] 
@@
 \end{definition}
+
+\TODO{text}
+
+\subsection{Elgot Categories}
 
@@
       ; \dnt{\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}{q}}
-      \\
-    \dnt{\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}}
-      = \dnt{\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}};\alpha_+
+    %   \\
+    % \dnt{\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}}
+    %   = \dnt{\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}};\alpha_+
   \end{gather*}
@@
     \boxed{\dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      : \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}})} \\
+      : \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})} \\
     \dnt{\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}}
@@
     \boxed{\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}
-      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
+      : \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
       \\
@@
     \\
-    \boxed{\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}: \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} 
+    \boxed{\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}: \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}})} 
   \end{gather*}
@@
       \dnt{\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
-      : \dnt{H} \to \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}})}
+      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})}
       \\
@@
       \dnt{\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
-      : \dnt{H} \to \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
+      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
       \\

### 2023-11-05 pldi24/paper.tex
Tao

@@
 
-\TODO{relate separation logic trick to \cite{promonad}, \cite{linear-state-usage}}
+\TODO{relate separation logic trick to \cite{promonad}, \cite{linear-state-usage}, \cite{mellies-ftrs}}
 
@@
 \end{definition}
+
+\TODO{results on distributivity of monads}
 
@@
 \TODO{why do these matter? they show up in the proofs, but...}
-Some important properties of all Elgot categories include:
-\begin{itemize}
-  \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id}, (h;[\iota_0, g])^\dagger]\).
-\TODO{Proposition 16 of \cite{goncharov-guarded-unguarded}; also Lemma 31 of \cite{goncharov-squaring}}
-  \item \textbf{Squaring:} \((f;[\iota_0, f])^\dagger = f^\dagger\). \TODO{Lemma 31 of \cite{goncharov-squaring}}
-\end{itemize}
+Some important properties which hold for all Elgot categories include:
+\begin{itemize}
+  \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id}, (h;[\iota_0, g])^\dagger]\)
+  \item \textbf{Squaring:} \((f;[\iota_0, f])^\dagger = f^\dagger\)
+\end{itemize}
+Proofs can be found in Lemma 31 of \cite{goncharov-squaring}.
+
 \TODO{Hyland and Hasegawa traces?}
+
 \TODO{intuition? Can draw diagrams, but those take up space}
 
-\TODO{results on distributivity of monads}
-
 \TODO{result: if a monad is Elgot, the state/reader/writer transformers yield Elgot monads}
-
-\TODO{example of Elgot monads: traces}
 

### 2023-11-05 pldi24/paper.tex
Tap at distributivity structure

@@
 
-\begin{definition}[Distributive category] 
-  A \textbf{distributive category} is a premonoidal category with coproducts equipped with a natural family of central isomorphisms \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\).
+\subsection{Distributivity}
+
+\TODO{text}
+
+\begin{definition}[If-then-else category] \
+  \TODO{this}
 \end{definition}
 
-\TODO{results on distributivity of monads}
+\TODO{results on if-then-else of monads}
+
+\begin{definition}[Distributive (pre)monoidal category] 
+  A \textbf{distributive (pre)monoidal category} is a (pre)monoidal category with all finite coproducts equipped with a natural family of central isomorphisms \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\).
+  \TODO{complex coherence conditions in \cite{laplaza-distributivity}, what we need for if-then-else}
+\end{definition}
 

### 2023-11-05 pldi24/paper.tex
More work on distributivity

--- 2023-11-05 pldi24/paper.tex
@@
 \newcommand{\trel}{\ms{r}}
-\newcommand{\tint}{{\{\ms{a}, \ms{r}\}}}
+\newcommand{\tint}{\ms{i}}
 \newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
@@
 \begin{definition}[Susbtructural Effectful Category]
-  A \textbf{substructural effectful category} \(\mc{C}_1 \to \mc{C}_0\) consists of
+  A \textbf{substructural effectful category} \(\mc{C}\) consists of
   \begin{itemize}
     \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms in this category \textbf{central} or \textbf{linear}
-    \item A substructural premonoidal category \(\mc{C}_0\)
-    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0^\varnothing\) which:
-    \begin{itemize}
-      \item Preserves all symmetric premonoidal structure (which naturally induces functors \(\upg{\cdot}{}: \mc{C}_1^q \to \mc{C}_0\) which do the same)
-      \item Sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\)
-    \end{itemize}
+    \item A substructural premonoidal category \(\mc{C}_0\) sharing the same affine and relevant objects as \(\mc{C}_1\)
+    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0^\varnothing\) preserving all substructural premonoidal structure, i.e., preserves all symmetries, associators, unitors, drops, and splits and sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\).
+    In particular, restricting \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to \mc{C}_0^q\) into effectful categories.
   \end{itemize}
@@
 
-\begin{definition}[Distributive category] 
-  A \textbf{distributive category} is a premonoidal category with coproducts equipped with a natural family of central isomorphisms \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\).
+\subsection{Distributivity}
+
+% \TODO{text}
+
+% \begin{definition}[If-then-else category] \
+%   \TODO{this}
+% \end{definition}
+
+% \TODO{results on if-then-else of monads}
+
+\begin{definition}[Distributive (pre)monoidal category] 
+  A \textbf{distributive (pre)monoidal category} is a (pre)monoidal category with all finite coproducts equipped with a natural family of central isomorphisms (the \textbf{distributor}) \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\).
+This isomorphism is subject to rather complex coherence conditions, worked out in full for the more general case of \textit{rig categories} in \cite{laplaza-distributivity}, but for our purposes we will only make use of the following:
+  \begin{equation*}
+    \delta_{A, B, C}^{-1} = [\iota_0 \otimes C, \iota_1 \otimes C] 
+    \qquad
+    \sigma^+_{A, B} \otimes C;\delta_{B, A, C} = \delta_{A, B, C};\sigma^+_{A \otimes C, B \otimes C}
+  \end{equation*}
+  Note in particular that this implies that \(\iota_i \otimes C;\delta_{A, B, C} = \iota_i\).
+A substructural (pre)monodial category is \textbf{distributive} if it is distributive as a (pre)monoidal category, and its distributor is \textit{pure}.
+A substructural effectful category \(\mc{C}\) is \textbf{distributive} if \(\mc{C}_0\) is distributive (as a substructural (pre)monoidal category).
 \end{definition}
 
+\TODO{can swap be derived from the previous two? Does this imply some of the other equations from \cite{laplaza-distributivity}? Alas for modularity!}
+
 \TODO{results on distributivity of monads}
+
+\TODO{results on distributivity of effectful things}
 
@@
 \subsection{Denotational Semantics}
+
+\TODO{coherent syntax for affine/relevant morphisms... see below...}
 

### 2023-11-05 pldi24/paper.tex
Tap

@@
 
-We can use these observations to deduce that every strong monad \(\mb{T}\) on a substructural monoidal category \(\mc{C}\) induces a substructural premonoidal category \(\mc{C}_{\mb{T}}\) (the Kleisli category) with affine/relevant objects the same as in \(\mc{C}\), and affine/relevant morphisms containing the image of all affine/relevant morphisms in \(\mc{C}\) under the inclusion functor \(\mc{C} \to \mc{C}_{\mb{T}}\).
+We can also deduce the following:
+\begin{proposition}
+  If \(\mc{C}\) is a substructural monoidal category then the Kleisli category \(\mc{C}_{\mb{T}}\) is a substructural premonoidal category with:
+  \begin{itemize}
+    \item \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\), \(\ms{Rel}(\mc{C}_{\mb{T}}) = \ms{Rel}(\mc{C})\)
+    \item \(\mc{C}_{\mb{T}}^q(A, B) = \upg{\mc{C}^q(A, B)}{}\), where \(\upg{f}{} = f;\eta\)
+    \item Drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\)
+  \end{itemize} 
+  In particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a substructural effectful category.
+\end{proposition}
+
+\TODO{work into text?}
 
 \subsection{Distributivity}
-
-% \TODO{text}
-
-% \begin{definition}[If-then-else category] \
-%   \TODO{this}
-% \end{definition}
-
-% \TODO{results on if-then-else of monads}
 
@@
 \end{definition}
-
-\TODO{can swap be derived from the previous two? Does this imply some of the other equations from \cite{laplaza-distributivity}? Alas for modularity!}
-
-\TODO{results on distributivity of monads}
-
-\TODO{results on distributivity of effectful things}
+Many (pre)monoidal categories are distributive, including of course \(\ms{Set}\), but also categories which can be derived from monads on more primitive categories such as \(\ms{PFun}\) and \(\ms{Rel}\), as can be deduced from the following proposition:
+\begin{proposition}
+  If \(\mc{C}\) is a distributive substructural monoidal category and \(\mb{T}\) is a strong monad on \(\mc{C}\), then the Kleisli category \(\mc{C}_{\mb{T}}\) is a distributive substructural effectful category (with affine and relevant objects and morphisms inherited from \(\mc{C}\))
+\end{proposition}
+\begin{proof}
+  Let \(\delta_{A, B, C}: \mc{C}((A + B) \otimes C, A \otimes C + B \otimes C)\) denote \(\mc{C}\)'s distributor.
+We will write composition in \(\mc{C}\) as "\(;\)" and composition in the Kleisli category \(\mb{T}\) as "\(\gg\)".
+  We will show that
+  \(\upg{\delta_{A, B, C}}{} = \delta_{A, B, C};\ms{pure}: \mc{C}_{\mb{T}}((A + B) \otimes C, A \otimes C + B \otimes C)\) satisfies the desired properties of a distributor. In particular, it is:
+  \begin{itemize}
+    \item Central: this follows immediately from being the image of a pure morphism
+    \item Affine: \(\upg{\delta}{} \gg \upg{\ms{drop}}{} = \upg{(\delta;\ms{drop})}{} = \upg{\ms{drop}}{}\)
+    \item Relevant: \(\upg{\delta}{} \gg \upg{\ms{split}}{} = \upg{(\delta;\ms{split})}{} = \upg{(\ms{split};\delta \otimes \delta)}{} = \upg{\ms{split}}{} \gg \upg{\delta}{} \otimes \upg{\delta}{}\)
+    \item Left inverse: \(\upg{\delta}{} \gg \upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \upg{\delta;[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \ms{id}\)
+    \item Right inverse: \(\upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} \gg \upg{\delta}{} = \upg{[\iota_0 \otimes C, \iota_1 \otimes C];\delta}{} = \ms{id}\)
+    \item Symmetric: \(\upg{\sigma^+}{} \otimes C \gg \upg{\delta}{} = \upg{\sigma^+ \otimes C;\delta}{} = \upg{\delta;\sigma^+}{} = \upg{\delta}{} \gg \upg{\sigma^+}{}\)
+  \end{itemize}
+\end{proof}
 

### 2023-11-05 pldi24/paper.tex
Tap

@@
 \newcommand{\rle}[1]{{\scriptsize\textsf{#1}}}
-\newcommand{\taff}{\ms{a}}
-\newcommand{\trel}{\ms{r}}
-\newcommand{\tint}{\ms{i}}
+\newcommand{\taff}{{\{\ms{a}\}}}
+\newcommand{\trel}{{\{\ms{r}\}}}
+\newcommand{\tint}{{\{\ms{a}, \ms{r}\}}}
 \newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
@@
       {\hasty{\Xi}{1}{a}{A}{\tint}}
-      {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} \qquad
+      {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} 
+    \\
     \prftree[r]{\rle{ite}}
@@
       {\haslb{\Xi}{t}{\ms{L}}}
-      {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} \\
+      {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} 
+    \\
     \prftree[r]{\rle{let-blk}}
@@
       {\hasty{\Xi}{0}{a}{A}{q}}
-      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} \\
+      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} 
+    \\
     \prftree[r]{\rle{let2-blk}}
@@
       {\hasty{\Xi}{0}{a}{A \otimes B}{q}}
-      {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \\
+      {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \qquad
     \prftree[r]{\rle{where}}
@@
       {\lhaslb{\ms{L}}{L}{\ms{K}}}
-      {\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}} \qquad
+      {\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}}
+    \\
     \prftree[r]{\rle{nil-br}}
@@
 
-\TODO{result: if a monad is Elgot, the state/reader/writer transformers yield Elgot monads}
+\begin{proposition}
+  If \(\mb{T}\) is an Elgot monad on \(\ms{Set}\), then \(\ms{StateT}\;\mb{T}\;S = \lambda A.
+S \to \ms{T}(A \times S)\), \(\ms{ReaderT}\;\mb{T}\;S = \lambda A.
+S \to \ms{T}(A)\), and \(\ms{WriterT}\;\mb{T}\;S = \lambda A.
+\ms{T}(A \times S)\) are all Elgot
+\end{proposition}
 
@@
 
-\TODO{coherent syntax for affine/relevant morphisms... see below...}
-
-\TODO{text}
-
-The semantics for an \isotopessa dialect with a set of types \(\mc{T}\) and instructions \(\mc{I}\) are given by:
-\begin{itemize}
-  \item A \textit{substructural effectful category} \(\mc{C}\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot}, equipped with
-  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies X \in \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies X \in \ms{Rel}(\mc{C})\)
+\TODO{make aff and rel syntax nicer/more consistent here? Subcategory syntax would be nice for lemmas and just coherence.}
+
+We're now ready to give a semantics for an \isotopessa dialect with a set of types \(\mc{T}\) and instructions \(\mc{I}\). We'll need:
+\begin{itemize}
+  \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot}, equipped with
+  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies \ms{base}(X) \in \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies \ms{base}(X) \in \ms{Rel}(\mc{C})\)
   \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\) to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
 \end{itemize}
-\TODO{make aff and rel syntax nicer/more consistent here? Subcategory syntax would be nice for lemmas and just coherence.}
+We begin by giving a semantics for types and contexts in terms of objects in \(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}; this allows us to give a semantics for structural judgements in Figure \ref{fig:ssa-structural-semantics}.
+
+\TODO{some text about what the semantics actually are...}
+
+We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics} followed by a semantics for blocks and label-sets in Figure \ref{fig:ssa-block-semantics}.
+
+\TODO{ditto}
+
+\TODO{rename blocks to statements? label-sets to CFGs?}
 
@@
     \dnt{X} = \ms{base}(X) 
-      \text{ where } 
-      \dnt{\mb{1}} = \mb{1}, 
-      \dnt{\mb{2}} = \mb{2}
+    \qquad
+    \dnt{\mb{1}} = \mb{1}
+    \qquad
+    \dnt{\mb{2}} = \mb{2} = \mb{1} + \mb{1}
     \qquad \dnt{A \otimes B} = \dnt{A} \otimes \dnt{B} \\
@@
 
-\TODO{some text about type/context semantics}
-
 \begin{figure}
@@
       \\
-    \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
-    \dnt{\tlin{A \otimes B}{\taff}} 
-      = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
-      ; \lambda \\
+      \dnt{\tlin{A}{\taff}} = \ms{drop}(\dnt{A})
+      % \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
+      % \dnt{\tlin{A \otimes B}{\taff}} 
+      %   = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
+      %   ; \lambda 
+      \\
     \boxed{\dnt{\tlin{A}{\trel}}
@@
       \\
-    \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
-    \dnt{\tlin{A \otimes B}{\trel}} = 
-      \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
-      ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
+      \dnt{\tlin{A}{\trel}} = \ms{split}(\dnt{A})
+      % \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
+      % \dnt{\tlin{A \otimes B}{\trel}} = 
+      %   \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
+      %   ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
       \\
@@
 \end{figure}
-
-\TODO{some text about structural semantics}
 
@@
   \Description{Semantics for isotope-SSA terms}
-  \label{fig:ssa-structural-terms}
+  \label{fig:ssa-term-semantics}
 \end{figure}
@@
   \Description{Semantics for isotope-SSA blocks}
-  \label{fig:ssa-structural-blocks}
+  \label{fig:ssa-block-semantics}
 \end{figure}

### 2023-11-05 pldi24/paper.tex
Tap
Literally 100 TODOs rip

@@
   For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
-  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\hasty{\Gamma}{1}{a}{A}{q}}{}\]
+  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
 \end{lemma}
@@
 
+\subsection{Weak Memory}
+
+\TODO{\cite{sparky}}
+
+\TODO{Pomsets...}
+
+\TODO{Form a monoid, so we can define...}
+
+\TODO{The SC Monad, which, via composition w/ the state transformer, lets us}
+
+\TODO{Build The Weak Memory Monad, which is too powerful, so we do...}
+
+\TODO{Karoubi envelope as premonoidal demonstration?}
+
 \subsection{Separation Logic}
 
-\TODO{State transformer on traces}
-
-\TODO{Monoidal functor, yielding sliding}
-
-\subsection{Weak Memory}
-
-\TODO{\cite{sparky}}
-
-\TODO{Pomsets...}
-
-\TODO{Form a monoid, so we can define...}
-
-\TODO{The SC Monad, which, via composition w/ the state transformer, lets us}
-
-\TODO{Build The Weak Memory Monad, which is too powerful, so we do...}
-
-\TODO{Karoubi envelope as premonoidal demonstration?}
+Building on the ideas in \cite{mellies-ftrs}, we will show how a simple separation logic based on \cite{reynolds-separation-2002} slightly adapted to the categorical setting gives us a very rich \isotopessa model supporting a basic notion of refinement types, while at the same time giving us a nontrivial notion of nonlinear types.
+
+We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{String}^* + \ms{String}^\omega)\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the infinite monoid of lists of strings described in the previous section.
+
+\TODO{by the above, this is an \isotopessa model w/ nice properties; moving on!}
+
+We now define a separation logic with predicates having the following grammar:
+\begin{grammar}
+  <\(\varphi, \psi\)> ::= 
+    \(\varnothing\) 
+    \;|\; \(e \mapsto a\) 
+    \;|\; \(\varphi * \psi\)
+    \;|\; \(\forall x: A, \varphi\)
+    \;|\; \(\exists x: A, \varphi\)
+    \;|\; \(\varphi \land \psi\)
+    \;|\; \(\varphi \lor \psi\)
+    \;|\; \(\varphi \Rightarrow \psi\)
+    \;|\; \(\top\)
+    \;|\; \(\bot\)
+\end{grammar}
+where \(e, a\) denote regular \isotopessa expressions and \(A, B\) regular \isotopessa types.
+
+We can now do something somewhat tricky with \isotopessa dialects:
+\TODO{truly pure expressions \(\implies\) semantics for expressions}
+\TODO{giving us a semantics for separation logic stuff}
+\TODO{now we define the category \(\ms{Sep}\) w/ objects contexts + predicates}
+\TODO{tensor product, niceties \(\implies\) category, yey}
+\TODO{sliding}
+\TODO{therefore, substructural \textit{effectful} w/ nontrivial linear objects}
 

### 2023-11-06 pldi24/paper.tex
Tap...

--- 2023-11-05 pldi24/paper.tex
@@
 
+\label{sec:semantics}
+
 \TODO{intro text}
@@
 
+In this section, we will build on the bare-bones models given in Section \ref{sec:semantics} to develop more sophisticated models capable of underpinning \isotopessa dialects supporting features such as weak memory and separation logic.
+
 \subsection{Basic Models}
@@
 
+\subsection{Weak Memory}
+
+\TODO{\cite{sparky}}
+
+\TODO{Pomsets...}
+
+\TODO{Form a monoid, so we can define...}
+
+\TODO{The SC Monad, which, via composition w/ the state transformer, lets us}
+
+\TODO{Build The Weak Memory Monad, which is too powerful, so we do...}
+
+\TODO{Karoubi envelope as premonoidal demonstration?}
+
 \subsection{Separation Logic}
 
-\TODO{State transformer on traces}
-
-\TODO{Monoidal functor, yielding sliding}
-
-\subsection{Weak Memory}
-
-\TODO{\cite{sparky}}
-
-\TODO{Pomsets...}
-
-\TODO{Form a monoid, so we can define...}
-
-\TODO{The SC Monad, which, via composition w/ the state transformer, lets us}
-
-\TODO{Build The Weak Memory Monad, which is too powerful, so we do...}
-
-\TODO{Karoubi envelope as premonoidal demonstration?}
+Building on the ideas in \cite{mellies-ftrs}, we will show how a simple separation logic based on \cite{reynolds-separation-2002} slightly adapted to the categorical setting gives us a very rich \isotopessa model supporting a basic notion of refinement types, while at the same time giving us a nontrivial notion of nonlinear types.
+
+We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{String}^* + \ms{String}^\omega)\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the infinite monoid of lists of strings described in the previous section.
+
+% \TODO{by the above, this is an \isotopessa model w/ nice properties; moving on!}
+
+% We now define a separation logic with predicates having the following grammar:
+% \begin{grammar}
+%   <\(\varphi, \psi\)> ::= 
+%     \(\varnothing\) 
+%     \;|\; \(e \mapsto a\) 
+%     \;|\; \(\varphi * \psi\)
+%     \;|\; \(\forall x: A, \varphi\)
+%     \;|\; \(\exists x: A, \varphi\)
+%     \;|\; \(\varphi \land \psi\)
+%     \;|\; \(\varphi \lor \psi\)
+%     \;|\; \(\varphi \Rightarrow \psi\)
+%     \;|\; \(\top\)
+%     \;|\; \(\bot\)
+% \end{grammar}
+% where \(e, a\) denote regular \isotopessa expressions and \(A, B\) regular \isotopessa types.
+
+% We can now do something somewhat tricky with \isotopessa dialects:
+% \TODO{truly pure expressions \(\implies\) semantics for expressions}
+% \TODO{giving us a semantics for separation logic stuff}
+% \TODO{now we define the category \(\ms{Sep}\) w/ objects contexts + predicates}
+% \TODO{tensor product, niceties \(\implies\) category, yey}
+% \TODO{sliding}
+% \TODO{therefore, substructural \textit{effectful} w/ nontrivial linear objects}
 

### 2023-11-06 pldi24/paper.tex
Tap

@@
 \newcommand{\ms}[1]{\ensuremath{\mathsf{#1}}}
+
+% Math
+\newcommand{\nats}{\mathbb{N}}
 
@@
 
-\TODO{\(\implies\) E-graph optimization}
-
-\TODO{rewriting substitution + weakening?}
-
 \subsection{Optimizations}
 
-\TODO{this...}
+\TODO{this:}
+
+\TODO{substitution \(\implies\): loop hoisting, global value numbering, some dead code elimination}
+
+\TODO{rewriting \(\implies\) strength reduction, E-graph optimization, see \cite{cranelift}}
+
+\TODO{inlining}
+
+\TODO{loop fusion}
+
+\TODO{unreachable code elimination}
+
+\TODO{more dead code elimination}
 
@@
 
-\TODO{Ye Olde Trace Monad}
-
-\TODO{Is Elgot}
-
-\TODO{Ye Olde Nondeterministic Trace Monad}
-
-\TODO{Is Elgot}
-
-\TODO{Ye Olde Reader/Writer/State Transformer}
-
-\TODO{Is Elgot}
-
-\TODO{already done above?}
+We begin by defining the monad of \textit{nondeterministic traces} generated by a monoid \(M\) and a set \(I\) on \(M\) equipped with an "infinitary action" as follows:
+\begin{definition}[Infinitary Action]
+  An \textbf{infinitary action} of a monoid \(M\) on a set \(I\) is a pair \((\Sigma, \cdot)\) where \(\Sigma: (\nats \to M) \to I\) is a map from streams of \(M\) to \(I\) and \(\cdot: M \to I \to I\) is a monoid action such that
+  \(
+    m \cdot \Sigma \ell = \Sigma (m: \ell) 
+  \)
+\end{definition}
+Some examples include:
+\begin{itemize}
+  \item Any monoid \(M\) equipped with the unique maps \(\Sigma: (\nats \to M) \to \mb{1}\), \(\cdot: M \to \mb{1} \to \mb{1}\)
+  \item The monoid of lists of \(A\) equipped with the concatenation map \(\Sigma: (\nats \to A^*) \to A^\omega\) and the concatenation action \(\cdot: A^* \to A^\omega \to A^\omega\)
+\end{itemize}
+\TODO{better syntax, think how to write \(A^\omega\)...}
+
+\begin{definition}[Trace Monad]
+  Given an infinitary action \((\Sigma, \cdot)\) of \(M\) on \(I\), we can define the \textbf{trace monad} \(\ms{Trace}\;\Sigma\;A = A \times M + I\), with
+  \TODO{bind}
+  \TODO{pure}
+\end{definition}
+
+\TODO{is Elgot...}
+
+\begin{definition}[Nondeterministic Trace Monad]
+  Given an infinitary action \((\Sigma, \cdot)\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Trace}\;\Sigma\;A = \mc{P}^+(A \times M + I)\), with
+  \TODO{bind}
+  \TODO{pure}
+\end{definition}
+
+\TODO{is Elgot...}
+
+\TODO{induces a substructural effectful category...}
 
@@
 
-\TODO{Pomsets...}
+\begin{definition}[Pomset] \
+  \ \TODO{this}
+\end{definition}
 
@@
 
-\TODO{The SC Monad, which, via composition w/ the state transformer, lets us}
+\TODO{The SC Monad, which, via composition w/ the state transformer w/ a buffer as state, lets us}
 
@@
 
-\TODO{Karoubi envelope as premonoidal demonstration?}
+\begin{definition}[Monadic Predicate] \
+  \TODO{this and naming}
+\end{definition}
+
+\TODO{Proper Weak Memory Monad, preserves the Elgot structure (yey)}
 
@@
 We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{String}^* + \ms{String}^\omega)\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the infinite monoid of lists of strings described in the previous section.
+
+\TODO{Name for ND trace monad}
 

### 2023-11-07 pldi24/paper.tex
Work on trace monad section, citations

--- 2023-11-06 pldi24/paper.tex
@@
 \newcommand{\isrw}[3]{#1: #2 \mapsto #3}
+\newcommand{\mbind}{\mathbin{{>}\hspace{-0.1em}{>}\hspace{-0.1em}{=}}}
 % \newcommand{\strictlbsubst}[3]{#1: #2 \rightsquigarrow_= #3}
@@
 \TODO{text...}
+
+\TODO{cite \cite{rosen-gvn-1988} somewhere?}
 
@@
 
+\TODO{clean up, and generalize to arbitrary premonoidal categories (w/ initial object? how to deal w/ coproducts?)}
+
 In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a substructural premonoidal category, with \(\mc{C}^\varnothing = \mc{C}\) and \(\mc{C}^{\{\ms{a}, \ms{r}\}}, \mc{C}^{\{\ms{a}\}}, \mc{C}^{\{\ms{r}\}}\) the subcategory consisting only of the associators, unitors, and symmetry, and only the identity object considered affine and relevant.
@@
 
-\TODO{Ye Olde Trace Monad}
-
-\TODO{Is Elgot}
-
-\TODO{Ye Olde Nondeterministic Trace Monad}
-
-\TODO{Is Elgot}
-
-\TODO{Ye Olde Reader/Writer/State Transformer}
-
-\TODO{Is Elgot}
-
-\TODO{already done above?}
+\TODO{text}
+
+\begin{definition}[Stream Action]
+  A \textbf{stream action} of a monoid \(M\) on a set \(I\) is a pair \((\Pi, \cdot)\) where \(\Pi: M^\omega \to I\) is a map from streams of \(M\) to \(I\) and \(\cdot: M \to I \to I\) is a monoid action such that
+  \(
+    m \cdot \Pi \ell = \Pi (m: \ell) 
+  \)
+  where \(m:\ell\) denotes the stream formed by prefixing \(m\) to \(\ell\).
+\end{definition}
+Two important examples include:
+\begin{itemize}
+  \item For any monoid \(M\), the unique maps \(\Pi: (\nats \to M) \to \mb{1}\), \(\cdot: M \to \mb{1} \to \mb{1}\) form a stream action on \(\mb{1}\)
+  \item The concatenation map \(\Pi: (A^*)^\omega \to A^\omega\) and the concatenation action \(\cdot: A^* \to A^{\leq \omega} \to A^{\leq \omega}\) form a stream action of the monoid of lists of \(A\), \(A^*\), on the set of potentially infinite sequences of \(A\), \(A^{\leq \omega}\).
+\end{itemize}
+
+\TODO{text}
+
+\begin{definition}[Trace Monad]
+  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{trace monad} \(\ms{Trace}\;\Pi\;A = A \times M + I\), with
+  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
+  and \(\mu_A = [\lambda (a, m). m \cdot a, \ms{id}]\)
+  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with
+  \[f^\dagger(a) = ...\]
+\end{definition}
+
+\TODO{text}
+
+\begin{definition}[Nondeterministic Trace Monad]
+  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Trace}\;\Sigma\;A = \mc{P}^+(A \times M + I)\), with
+  \(\eta_A = \lambda a.\{\iota_0 (a, 1)\}\)
+  and \(\mu_A = ...\)
+  This is an Elgot monad, with
+  \[f^\dagger(a) = ...\]
+\end{definition}
+
+\TODO{text}
+
+\TODO{induces a substructural effectful category...}
+
+\TODO{central = commutative, "commutes with infinity" (i.e. is absorbed by infinity)}
+
+\TODO{relevant = deterministic}
+
+\TODO{affine = no effect on trace}
 
@@
 
-\TODO{\cite{sparky}}
-
-\TODO{Pomsets...}
+\TODO{text, \cite{sparky}}
+
+\begin{definition}[Pomset] \
+  \ \TODO{this}
+\end{definition}
 
@@
 
-We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{String}^* + \ms{String}^\omega)\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the infinite monoid of lists of strings described in the previous section.
+We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{Bytes}^* + \ms{Bytes}^{\leq \omega})\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the infinite monoid of lists of strings described in the previous section.
+
+\TODO{Name for ND trace monad}
 
@@
 
+\TODO{\cite{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
+
+\TODO{\cite{schneider-phd}}
+
+\TODO{\cite{chakravarty-functional-ssa-2003}}
+
+\TODO{\cite{barthe-compcert-ssa-2014}}
+
+\TODO{\cite{buchwald-verified-ssa-2016}}
+
+\TODO{\cite{herklotz-gsa-2023}}
+
+\TODO{\cite{pop-ssa-inout-2009}}
+
+\TODO{\cite{garbuzov-structural-cfg-2018}}
+
+\TODO{\cite{hua-explicit-ssa-2010}}
+
 \TODO{this}
@@
 
+\TODO{start appendix}
+
 \end{document}

### 2023-11-07 pldi24/paper.tex
Tap

@@
 This thus gives rise to an SSA-based IR with support for weak memory operations, which is fully semantically-justified.
-\item \TODO{We also have a small implementation, which \ldots}
+\item \todo{We also have a small implementation, which \ldots}
 \item Our weak memory semantics and our results about Elgot monads are formalised in the Lean theorem prover. \TODO{Other stuff}
@@
 
-\TODO{define premonoidal category?}
+\begin{definition}[Binoidal Category] \
+  \TODO{this}
+\end{definition}
+
+\begin{definition}[Premonoidal Category] \
+  \TODO{this}
+\end{definition}
+
+\TODO{text}
 
@@
 
+\TODO{find better notation than \(\Pi\)?}
+
 \begin{definition}[Stream Action]
@@
 
-\begin{definition}[Pomset] \
-  \ \TODO{this}
+\begin{definition}[Pomset] 
+  A \textbf{pomset} over a set of actions \(\mc{A}\) with a distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\)
+  \TODO{finish, clean}
 \end{definition}
@@
 \TODO{Form a monoid, so we can define...}
+
+Let \(\ms{Pom}\) denote the monoid of pomsets...
+
+\TODO{stream action... trace monad \(\ms{Traces}\;\ms{Pom}\;\ms{Pom}\)...}
 
@@
 
-We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{Bytes}^* + \ms{Bytes}^{\leq \omega})\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the infinite monoid of lists of strings described in the previous section.
-
-\TODO{Name for ND trace monad}
-
-% \TODO{by the above, this is an \isotopessa model w/ nice properties; moving on!}
-
-% We now define a separation logic with predicates having the following grammar:
-% \begin{grammar}
-%   <\(\varphi, \psi\)> ::= 
-%     \(\varnothing\) 
-%     \;|\; \(e \mapsto a\) 
-%     \;|\; \(\varphi * \psi\)
-%     \;|\; \(\forall x: A, \varphi\)
-%     \;|\; \(\exists x: A, \varphi\)
-%     \;|\; \(\varphi \land \psi\)
-%     \;|\; \(\varphi \lor \psi\)
-%     \;|\; \(\varphi \Rightarrow \psi\)
-%     \;|\; \(\top\)
-%     \;|\; \(\bot\)
-% \end{grammar}
-% where \(e, a\) denote regular \isotopessa expressions and \(A, B\) regular \isotopessa types.
-
-% We can now do something somewhat tricky with \isotopessa dialects:
-% \TODO{truly pure expressions \(\implies\) semantics for expressions}
-% \TODO{giving us a semantics for separation logic stuff}
-% \TODO{now we define the category \(\ms{Sep}\) w/ objects contexts + predicates}
-% \TODO{tensor product, niceties \(\implies\) category, yey}
-% \TODO{sliding}
-% \TODO{therefore, substructural \textit{effectful} w/ nontrivial linear objects}
+We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{Bytes}^* + \ms{Bytes}^{\leq \omega})\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the monoid of lists of bytestrings described in the previous section.
+
+\TODO{say something better than "bytestrings"}
+
+\TODO{what is a \(\ms{Heap}\)}
+
+\TODO{Use name \ms{Traces}}
+
+\TODO{by the above, this is an \isotopessa model w/ nice properties; moving on!}
+
+Given a set \(A\), we can define a predicate \(\varphi\) on \(A\), \(\mb{Heap}\) pairs to have the following type:
+\TODO{this}
+\TODO{Objects \((A, \varphi)\), \((B, \psi)\)}
+\TODO{arrows}
+\TODO{note on magic wand}
+\TODO{Tensor product and separating conjunction}
+\TODO{Coproduct, Elgot structure inherited from base category}
+\TODO{sliding w/ properly restricted equality}
+
+\TODO{so, it's a substructural effectful category mmkay}
+\TODO{central = no printing}
+\TODO{linear objects iff non duplicable}
+\TODO{relevant morphisms iff deterministic}
+\TODO{affine morphisms always (if no heap effects?)}
 
@@
 \TODO{\cite{hua-explicit-ssa-2010}}
+
+\TODO{\cite{rvsdg}}
 
@@
 
-\TODO{start appendix}
+\clearpage
+
+\appendix
+
+\section{Appendix}
 

### 2023-11-07 pldi24/paper.tex
Tap

@@
 
-\subsection{Basic Models}
+\subsection{Trace Models}
 
@@
 
-\TODO{find better notation than \(\Pi\)?}
-
 \begin{definition}[Stream Action]
-  A \textbf{stream action} of a monoid \(M\) on a set \(I\) is a pair \((\Pi, \cdot)\) where \(\Pi: M^\omega \to I\) is a map from streams of \(M\) to \(I\) and \(\cdot: M \to I \to I\) is a monoid action such that
+  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to I\) is a map from streams of \(M\) to \(I\) satisfying
   \(
-    m \cdot \Pi \ell = \Pi (m: \ell) 
-  \)
-  where \(m:\ell\) denotes the stream formed by prefixing \(m\) to \(\ell\).
+    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
+  \),
+  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the stream \(\ell\).
 \end{definition}
-Two important examples include:
-\begin{itemize}
-  \item For any monoid \(M\), the unique maps \(\Pi: (\nats \to M) \to \mb{1}\), \(\cdot: M \to \mb{1} \to \mb{1}\) form a stream action on \(\mb{1}\)
-  \item The concatenation map \(\Pi: (A^*)^\omega \to A^\omega\) and the concatenation action \(\cdot: A^* \to A^{\leq \omega} \to A^{\leq \omega}\) form a stream action of the monoid of lists of \(A\), \(A^*\), on the set of potentially infinite sequences of \(A\), \(A^{\leq \omega}\).
-\end{itemize}
+Some basic examples include:
+\begin{enumerate}
+  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\) forms a stream action on the one element set \(\mathbf{1}\).
+  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite) sequence \(s\).
+  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts on \(\nats \cup \{\infty\}\) by addition.
+\end{enumerate}
+We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted \(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\).
+For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq \omega}) = Z(A^*) = \{[\,]\}\), while \(\mc{SZ}(\nats, \nats \cup \{\infty\}) = \{0\} \subseteq Z(\nats) = \nats\).
+Generally, \(\mc{SZ}(M, I)\) always contains the unit of \(M\).
 
@@
   and \(\mu_A = [\lambda (a, m). m \cdot a, \ms{id}]\)
-  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with
-  \[f^\dagger(a) = ...\]
+  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\).
 \end{definition}
@@
 \TODO{text}
+
+\TODO{induces a substructural effectful category...}
+
+\TODO{central = commutative, "commutes with infinity" (i.e. is absorbed by infinity)}
+
+\TODO{relevant = idempotent}
+
+\TODO{affine = no effect on trace}
 
@@
   This is an Elgot monad, with
-  \[f^\dagger(a) = ...\]
+  \[f^\dagger(a) = \{\} \cup \{\} \cup \{\}\]
 \end{definition}
@@
 
-\TODO{relevant = deterministic}
+\TODO{relevant = deterministic + idempotent; inclusion of trace monad}
 
@@
 
-\subsection{Weak Memory}
-
-\TODO{text, \cite{sparky}}
-
-\begin{definition}[Pomset] 
-  A \textbf{pomset} over a set of actions \(\mc{A}\) with a distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\)
-  \TODO{finish, clean}
-\end{definition}
-
-\TODO{Form a monoid, so we can define...}
-
-Let \(\ms{Pom}\) denote the monoid of pomsets...
-
-\TODO{stream action... trace monad \(\ms{Traces}\;\ms{Pom}\;\ms{Pom}\)...}
-
-\TODO{The SC Monad, which, via composition w/ the state transformer w/ a buffer as state, lets us}
-
-\TODO{Build The Weak Memory Monad, which is too powerful, so we do...}
-
-\begin{definition}[Monadic Predicate] \
-  \TODO{this and naming}
-\end{definition}
-
-\TODO{Proper Weak Memory Monad, preserves the Elgot structure (yey)}
-
-\subsection{Separation Logic}
+\subsection{Heaps and Separation Logic}
 
@@
 
+\TODO{equality vs \cite{mellies-ftrs}}
+\TODO{potential solutions}
+
+\subsection{Weak Memory}
+
+\TODO{text, \cite{sparky}}
+
+\begin{definition}[Pomset] 
+  A \textbf{pomset} over a set of actions \(\mc{A}\) with a distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\)
+  \TODO{finish, clean}
+\end{definition}
+
+\TODO{Form a monoid, so we can define...}
+
+Let \(\ms{Pom}\) denote the monoid of pomsets...
+
+\TODO{stream action... trace monad \(\ms{Traces}\;\ms{Pom}\;\ms{Pom}\)...}
+
+\TODO{The SC Monad, which, via composition w/ the state transformer w/ a buffer as state, lets us}
+
+\TODO{Build The Weak Memory Monad, which is too powerful, so we do...}
+
+\begin{definition}[Monadic Predicate] \
+  \TODO{this and naming}
+\end{definition}
+
+\TODO{Proper Weak Memory Monad, preserves the Elgot structure (yey)}
+
 \section{Implementation}

### 2023-11-07 pldi24/paper.tex
Twiddle, bump Typst version
Tap

@@
 
-\TODO{text}
-
+We will begin by giving a useful foundation for a wide class of \isotopessa models: the "trace models", which accumulate a (potentially infinite) nondeterministic trace of monoidal "events".
+This will later come in handy for reasoning about output behaviour and weak memory.
+
+We begin with some definitions:
 \begin{definition}[Stream Action]
@@
   \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\) forms a stream action on the one element set \(\mathbf{1}\).
+  \item \todo{product, giving monoidal structure?}
   \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite) sequence \(s\).
@@
 
+We can now define the \textit{trace monad} of a stream action as follows:
+\begin{definition}[Trace Monad]
+  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A \times M + I)\), with
+  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
+  and 
+  \[\mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}\]
+  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with iteration operator
+  \(f^\dagger(a) = \bigcup_{n \in \nats \cup \{\infty\}}f_n(a)\), where
+  \TODO{this}
+
+  \TODO{deterministic trace monad; inherited Elgot structure: use as example of "Elgot submonad"? Go cite \cite{carette-central-2022}}
+\end{definition}
+
 \TODO{text}
 
-\begin{definition}[Trace Monad]
-  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{trace monad} \(\ms{Trace}\;\Pi\;A = A \times M + I\), with
-  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
-  and \(\mu_A = [\lambda (a, m). m \cdot a, \ms{id}]\)
-  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\).
-\end{definition}
-
-\TODO{text}
+\TODO{merge det/nondet}
 
@@
 \TODO{affine = no effect on trace}
-
-\begin{definition}[Nondeterministic Trace Monad]
-  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Trace}\;\Sigma\;A = \mc{P}^+(A \times M + I)\), with
-  \(\eta_A = \lambda a.\{\iota_0 (a, 1)\}\)
-  and \(\mu_A = ...\)
-  This is an Elgot monad, with
-  \[f^\dagger(a) = \{\} \cup \{\} \cup \{\}\]
-\end{definition}
 
@@
 \TODO{sliding w/ properly restricted equality}
+\TODO{equality vs \cite{mellies-ftrs}}
+\TODO{potential solutions}
 
@@
 
-\TODO{equality vs \cite{mellies-ftrs}}
-\TODO{potential solutions}
-
 \subsection{Weak Memory}

### 2023-11-07 pldi24/paper.tex
Tap

@@
 
-We will begin by giving a useful foundation for a wide class of \isotopessa models: the "trace models", which accumulate a (potentially infinite) nondeterministic trace of monoidal "events".
-This will later come in handy for reasoning about output behaviour and weak memory.
+\TODO{fix this text}
+
+We will begin by giving a useful foundation for a wide class of \isotopessa models: the "trace models" over \(\mb{Set}\), which accumulate a (potentially infinite) nondeterministic trace of monoidal "events".
+These will later come in handy for reasoning about output behaviour and weak memory.
 
@@
   \TODO{this}
-
-  \TODO{deterministic trace monad; inherited Elgot structure: use as example of "Elgot submonad"? Go cite \cite{carette-central-2022}}
 \end{definition}
@@
 
-\TODO{merge det/nondet}
-
-\TODO{induces a substructural effectful category...}
-
-\TODO{central = commutative, "commutes with infinity" (i.e. is absorbed by infinity)}
-
-\TODO{relevant = idempotent}
-
-\TODO{affine = no effect on trace}
-
-\TODO{text}
-
-\TODO{induces a substructural effectful category...}
-
-\TODO{central = commutative, "commutes with infinity" (i.e. is absorbed by infinity)}
-
-\TODO{relevant = deterministic + idempotent; inclusion of trace monad}
-
-\TODO{affine = no effect on trace}
+\begin{definition}[Submonad] 
+  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\).
+We say that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and, for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where \(f^\dagger\) is computed as in \(\mb{T}\).
+\end{definition}
+Some important examples of submonads include:
+\begin{itemize}
+  \item \todo{unit submonad of every monad}
+  \item \todo{union, intersection of (Elgot) submonads; lattice of submonads}
+  \item \todo{commuting absorbtive submonad of trace monad: example of...}
+  \item \todo{central submonad. Go cite \cite{carette-central-2022}}
+  \item \todo{deterministic, idempotent trace monad: relevant submonad}
+  \item \todo{noneffectful trace monad: affine submonad}
+  \item \todo{cap and centralize, cap and purify}
+\end{itemize}
+\TODO{in particular, these induce a substructural effectful category...}
 

### 2023-11-07 pldi24/paper.tex
Tap

@@
   \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\) forms a stream action on the one element set \(\mathbf{1}\).
-  \item \todo{product, giving monoidal structure?}
+  \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta = \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
+  \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta = \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i \cdot_i l_i)_i\).
   \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite) sequence \(s\).

### 2023-11-07 pldi24/paper.tex
Tap...
Tap

@@
   where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with iteration operator
-  \(f^\dagger(a) = \bigcup_{n \in \nats \cup \{\infty\}}f_n(a)\), where
-  \TODO{this}
+  \(f^\dagger(a) = f^\infty(a) \cup \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t \in f_n(a)\}\), where
+  \[
+    f_0 = f, \qquad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n]
+  \]
+  \[
+    f^\infty(a) = \{\Theta (\lambda n. m_n) \mid a_0 = a \land \forall n, \exists a_{n + 1}. \iota_0 (\iota_1\;a_{n + 1}, m_n) \in f(a_n)\}
+  \]
+  \TODO{clean this up}
 \end{definition}
@@
 \begin{itemize}
-  \item \todo{unit submonad of every monad}
-  \item \todo{union, intersection of (Elgot) submonads; lattice of submonads}
+  \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\); similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
+  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise unions \(\bigvee_i\mb{S}_i = \lambda A.
+\bigcup_i \mb{S}_i\;A\) and intersections \(\bigwedge_i\mb{S}_i = \lambda A.
+\bigcap_i \mb{S}_i\;A\) are submonads, forming a lattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\)
   \item \todo{commuting absorbtive submonad of trace monad: example of...}

### 2023-11-07 pldi24/paper.tex
Tap

@@
   \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\); similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
-  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise unions \(\bigvee_i\mb{S}_i = \lambda A.
-\bigcup_i \mb{S}_i\;A\) and intersections \(\bigwedge_i\mb{S}_i = \lambda A.
-\bigcap_i \mb{S}_i\;A\) are submonads, forming a lattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\)
+  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise intersections \(\bigwedge_i\mb{S}_i = \lambda A.
+\bigcap_i \mb{S}_i\;A\) are submonads, forming a meet-semilattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\)
   \item \todo{commuting absorbtive submonad of trace monad: example of...}

### 2023-11-08 pldi24/paper.tex
Tap

@@
 
+\begin{definition}[Effectful Category] \
+  \TODO{this}
+\end{definition}
+
+\begin{definition}[(Symmetric) Premonoidal Subcategory]
+  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is closed under whiskering and contains all associators, unitors, and symmetries.
+We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\), \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is \textbf{central} if this generalizes to the case where only one of \(f, g\) is in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
+\end{definition} 
+Note that the set of symmetric premonoidal subcategories, monoidal subcategories, and central subcategories of a category \(\mc{C}\) are all closed under intersections and therefore, in particular, form a complete meet-semilattice with shared bottom element \(\ms{Skeleton}(\mc{C})\), where \(\ms{Skeleton}(C)\) is the smallest possible symmetric premonodial subcategory.
+The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the largest central subcategory is the center of \(\mc{C}\) \(Z(\mb{C})\); the existence of a top-element means that in these cases we may induce the structure of a complete lattice in the usual manner.
+
+\begin{definition}[Strictly Effectful Category] \
+  \TODO{this}
+\end{definition}
+
 \TODO{text}
@@
   \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise intersections \(\bigwedge_i\mb{S}_i = \lambda A.
-\bigcap_i \mb{S}_i\;A\) are submonads, forming a meet-semilattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\)
+\bigcap_i \mb{S}_i\;A\) are submonads, forming a complete meet-semilattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete lattice in the standard manner
   \item \todo{commuting absorbtive submonad of trace monad: example of...}
@@
 \end{itemize}
+\TODO{a submonad induces a wide subcategory}
+
 \TODO{in particular, these induce a substructural effectful category...}

### 2023-11-09 pldi24/paper.tex
Tap

@@
 
-\TODO{text}
+In Figure \ref{fig:ssa-grammar}, we present a formal grammar for our SSA syntax.
+We begin by giving a grammar for \textit{types} \(A, B, C\), parametrized by a set of \textit{base types} \(X\).
+We require the existence of a \textit{unit type} \(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all tensor products \(A \otimes B\); we write \(\otimes\) to underscore the fact that types are allowed to be substructural.
+
+We separate program fragments into 3 main syntactic categories: \textit{expressions} \(a, b, c, e\), \textit{blocks} \(s, t\), and \textit{label-sets} \(L\).
+While we make a distinction between \textit{pure} and \textit{impure} expressions, this is done at the level of typing rules rather than via syntax.
+A \textit{block} is conceptualized as a series of \textit{let-bindings} followed by a \textit{terminator}.
+Here, a terminator is either an unconditional branch to another block with an argument or an if-statement.
+A block may be encapsulated within a \textit{where-block}, which defines other blocks the inner block can branch to; these blocks can also branch to each other in a mutually recursive fashion.
+Similarly to blocks, expressions may also contain let-bindings; this is somewhat of a departure from classical SSA, in which there is only one kind of variable binding, but facilitates various optimizations and corresponds more closely to the in-memory representation, where pure expressions are represented as E-graph nodes.
+Note that we do \textit{not} give a grammar or semantics for top-level functions, which we can view as simply a list of arguments and a block.
+Except for inlining, we will only consider the optimization and analysis of a single function body at a time.
+
+We proceed to give a grammar for \textit{contexts} \(\Gamma\), which consist of a list of variables \(x\) given a type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\).
+This tells us whether the variable is \textit{affine} (i.e., can be ignored) and/or \textit{relevant} (i.e., whether it can be used multiple times).
+A variable which is both affine and relevant is called \textit{intuitionistic}.
+Whether we can use a variable also depends on the \textit{linearity} of its type.
+
+We can then give a grammar for \textit{label-contexts} \(\ms{L}\), which consist of a list of \textit{labels} each annotated with a context \(\Gamma\), denoting the variables which must be live on entry to the block (corresponding to the set a dominance frontier would compute), and an \textit{argument} \(A\).
 
@@
 \end{figure}
-
-\TODO{late \ms{where}-binding, and other sugar (?)}
-
-\TODO{top-level functions}
-
-\TODO{table of typing judgements}
-
-\TODO{contexts, label-contexts}
 
@@
 \end{figure}
+
+We can now list out our primary typing judgements, which are given in Figure \ref{fig:ssa-judgements}.
 

### 2023-11-09 pldi24/paper.tex
Tap

@@
         \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
-        \(\hasty{\Gamma}{p}{x}{A}{q}\) &
+        \(\hasty{\Gamma}{p}{a}{A}{q}\) &
         \(a\) is a term of type \(A\) in context \(\Gamma\) with purity \(p\) and quantity \(q\) \\
         \(\haslb{\Gamma}{t}{\ms{L}}\) &
-        \(t\) is a block targeting label-set \(\ms{L}\) in context \(\Gamma\) \\
+        \(t\) is a block targeting labels \(\ms{L}\) in context \(\Gamma\) \\
         \(\lhaslb{\ms{L}}{L}{\ms{K}}\) &
-        The labels \(L\) send label-set \(\ms{L}\) to label-set \(\ms{K}\) \\
+        The label-set \(L\) sends labels \(\ms{L}\) to labels \(\ms{K}\) \\
         \(\csplits{\Gamma}{\Delta}{\Xi}\) &
@@
 We can now list out our primary typing judgements, which are given in Figure \ref{fig:ssa-judgements}.
-
-\TODO{some text about each typing judgement}
+We have three main typing judgements: \(\hasty{\Gamma}{p}{x}{A}{q}\) is the typing judgement for expressions, which says that the expression \(a\) has type \(A\) in context \(\Gamma\).
+The annotation \(q\) means that the expression itself has quantity \(q \subseteq \tint\), while the annotation \(p\) means that the expression has purity \(p\) (1 for pure, 0 for impure).
+\(\haslb{\Gamma}{t}{\ms{L}}\) is the typing block for blocks, which says that, using variables in \(\Gamma\), the block \(t\) exits by jumping to labels in \(\ms{L}\).
+Finally, \(\lhaslb{\ms{L}}{L}{\ms{K}}\) says that the label-set \(L\) defines the labels \(\ms{L}\) in terms of the labels \(\ms{K}\) (blocks in \(L\) are allowed to recursively call into \(\ms{L}\) or exit to \(\ms{K}\)).
+
+To define these, we will need a few basic structural judgements.
+\(\csplits{\Gamma}{\Delta}{\Xi}\) means that the context \(\Gamma\) can be split into contexts \(\Delta\) and \(\Xi\).
+We use the notion of splitting, rather than weakening, a context as primitive to be able to deal with linearity, since not all variables can be dropped or duplicated.
+This can be used to define weakening \(\cwk{\Gamma}{\Delta}\) as simply "splitting \(\Gamma\) into \(\Delta\) and the empty context", since a split is allowed to drop affine variables.
+Finally, \(\lwk{\ms{L}}{\ms{K}}\) means that the label-set \(\ms{L}\) weakens to the label-set \(\ms{K}\), i.e.
+that \(\ms{K}\) has more labels, but for shared labels, needs less variables to be live (but the same non-affine variables!) The typing rules for these judgements are given in Figure \ref{fig:ssa-structural}.
 

### 2023-11-09 pldi24/paper.tex
Figure

@@
 \usepackage{xcolor}
+\usepackage{subfig}
 
@@
 
-In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks of the following form:
-\begin{lstlisting}
-'label:
-  var = instruction
-  ...
-  var = instruction
-  terminator
+In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks, which can be viewed as \textit{label}, followed by a sequence of \textit{instructions}, each defining a \textit{variable}, followed by a \textit{terminator}.
+A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
+\TODO{text}
+
+\newsavebox\notssafib
+
+\begin{lrbox}{\notssafib}
+\begin{minipage}{0.3\textwidth}
+\begin{lstlisting}[frame=none,numbers=none]
+'entry:
+  m = 0
+  n = 1
+  brz i 'exit 'loop
+'loop:
+  t = add m n
+  m = n
+  n = t
+  N = sub i 1
+  brz i 'exit 'loop
+'exit:
+  ret m
 \end{lstlisting}
-A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
-\begin{lstlisting}
+\end{minipage}
+\end{lrbox}
+
+\newsavebox\ssafib
+
+\begin{lrbox}{\ssafib}
+\begin{minipage}{0.3\textwidth}
+\begin{lstlisting}[frame=none,numbers=none]
 'entry:
-  x = 7
-  x = add x 2
-  ret x
+  m0 = 0
+  n0 = 1
+  brz N 
+    'exit (m0) 
+    'loop (i, m0, n0)
+'loop(i0, m1, n1):
+  m2 = n1
+  n2 = add m1 n1
+  i1 = sub i0 1
+  brz z1 
+    'exit (m2) 
+    'loop (i1, m2, n2)
+'exit(m3):
+  ret m3
 \end{lstlisting}
-is not in SSA form, but can easily be converted into SSA form by rewriting it as
-\begin{lstlisting}
-'entry:
-  x0 = 7
-  x1 = add x0 2
-  ret x1
+\end{minipage}
+\end{lrbox}
+
+\newsavebox\ourfib
+
+\begin{lrbox}{\ourfib}
+\begin{minipage}{0.3\textwidth}
+\begin{lstlisting}[frame=none,numbers=none]
+let m0 = 0;
+let n1 = 1;
+if eq i 0 {
+  br 'exit(m0)
+} else {
+  br 'loop(i, m0, n0)
+} where
+'loop(i0, m1, n1) =>
+  let m2 = n1;
+  let n2 = add m1 n1;
+  let i1 = sub i0 1;
+  if eq i1 0 {
+    br 'exit(m2)
+  } else {
+    br 'loop(i1, m2, n2)
+  }
+'exit(m3) => ret m3
 \end{lstlisting}
-Unfortunately, things get a little more tricky when we try to define variables in loops or branches. For example, consider the following code:
-\begin{lstlisting}
-'entry:
-  br arg0 'left `right
-'left:
-  x = 7
-  jmp 'exit
-'right:
-  x = 3
-  jmp 'exit
-'exit:
-  ret x
-\end{lstlisting}
-There are two competing definitions of \(x\) in the left and right blocks, and it’s unclear how to disambiguate between them.
-The usual solution is to introduce a \(\phi\)-node.
-In particular, we can write:
-\begin{lstlisting}
-'entry:
-  br arg0 'left `right
-'left:
-  x0 = 7
-  jmp 'exit
-'right:
-  x1 = 3
-  jmp 'exit
-'exit:
-  x2 = phi 'left x0 'right x1
-  ret x2
-\end{lstlisting}
-\TODO{make above into graphs?}
-\TODO{segue}
-The semantics of SSA, however, can be quite complex to reason about, especially when considering the somewhat unintuitive scoping rules of variables and \(\phi\)-function arguments.
+\end{minipage}
+\end{lrbox}
+
+\begin{figure}
+  \centering
+  \subfloat[3-address code]{\usebox\notssafib}
+  \quad
+  \subfloat[Basic blocks with arguments]{\usebox\ssafib}
+  \quad
+  \subfloat[Our syntax]{\usebox\ourfib}
+  \caption{A simple program for computing the \texttt{i}th Fibonacci number, represented as 3-address code, basic-blocks with arguments, and in our SSA syntax}
+  \Description{}
+  \label{fig:ssa-examples}
+\end{figure}
 

### 2023-11-09 pldi24/paper.tex
intro

@@
 Static Single Assignment, or SSA, has been a cornerstone of compiler
-design ever since its introduction by \cite{ssa-original} in the late
-1980s. Older compiler intermediate representations permitted the same
-local variable to be assigned multiple times, whereas SSA requires
-each variable bindings to be immutable -- they cannot be modified
-after initialisation. This makes many dataflow analyses easier to
-compute, and (just as in functional languages) makes it much easier to
-justify compiler optimisations in terms of equational rewrites.
+design ever since its introduction by
+\cite{ssa-original,rosen-gvn-1988} in the late 1980s. Older compiler
+intermediate representations permitted the same local variable to be
+assigned multiple times, whereas SSA requires each variable bindings
+to be immutable -- they cannot be modified after initialisation. This
+makes many dataflow analyses easier to compute, and (just as in
+functional languages) makes it much easier to justify compiler
+optimisations in terms of equational rewrites.
 
@@
 
-\TODO{Example here?}
-
 This ``basic blocks with arguments'' presentation of SSA has become
@@
 substituting values for variables (the $\beta$-value rule). This weak
-substitution rule suffices to validate some basic compiler
-optimisations such as constant propagation and copy propagation, but
-is not strong enough to justify many other essential compiler
-optimisations. For example, hoisting a store out of a loop or
-conditional involves moving an effectful term (an assignment) out of
-its original position in the program. As another example, dead store
-elimination removes an effectful term from the program altogether. \todo{Another example?}
+equational rule suffices to validate some basic compiler optimisations
+such as constant propagation and copy propagation, but is not strong
+enough to justify many other essential compiler optimisations. For
+example, hoisting a store out of a loop or conditional involves moving
+an effectful term (an assignment) out of its original position in the
+program. As another example, dead store elimination removes an
+effectful term from the program altogether. \todo{Add link to figure with code for these examples}
 
@@
 \emph{effectful} terms and moving them around and substituting them
-for variables.
-
-% 
-% , on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, Cranelift, and MLIR are based.
-One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.
-% 
-\TODO{text...}
-
-\TODO{cite \cite{rosen-gvn-1988} somewhere?}
+for variables. Since this is not a generally valid program
+transformation, we need a finer classification of programs than
+merely pure and effectful to judge when a substitution is allowed.
+
+A useful taxonomy was introduced by \citet{fuhrmann}, who classified
+effects based on (1) whether or not the order of effects mattered,
+(2) whether it was sound to discard an effectful operation, and (3)
+whether it was sound to duplicate an effectful operation.
+
+Suppose we substitute \texttt{print("hello")} for $x$ in the program
+below:
+\begin{verbatim}
+let x = print("hello");            let y = print("world");
+let y = print("world");   =/=>     let z = print("hello");
+let z = x;                         ...
+...
+\end{verbatim}
+
+In this case, substituting a term for a variable changes the order the
+I/O actions are performed, which changes the program's observable
+behaviour. On the other hand, suppose we have an effect which increments
+a counter. Then the same transformation is sound. 
+\begin{verbatim}
+let x = increment(a);            let y = increment(b);
+let y = increment(b);   ===>     let z = increment(a);
+let z = x;                        ...
+...
+\end{verbatim}
+
+F\"{u}hrmann called these commutative effects \emph{central}, which we
+also call \emph{linear}.
+
+It can also matter whether an effect is performed at all or not. For example:
+\begin{verbatim}
+let x = may_loop();    =/=>      return 5;
+return 5 
+\end{verbatim}
+In this example, even though $x$ is not used, the call to \texttt{may_loop()} cannot
+be erased, as the observable behaviour of the program can change if it goes into an
+infinite loop.
+
+On the other hand, if the effectful operation is merely nondeterministic, then it is
+safe to discard results. 
+\begin{verbatim}
+let x = rand();    ===>          return 5;
+return 5 
+\end{verbatim}
+F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use his
+terminology, and also call such effects \emph{affine}.
+
+Now consider the following example: 
+\begin{verbatim}
+let x = rand();    =/=>     let y = rand();
+let y = x;                  let z = rand();
+let z = x;                  return y == z;
+return y == z
+\end{verbatim}
+
+In this case, substituting for $x$ duplicates the call to \texttt{rand()}, and can change
+this from a program which always returns true to one which can return either true or false.
+
+On the other hand, if the effect is the possibility of nontermination, then the
+substitution is sound. 
+\begin{verbatim}
+let x = may_loop();    ===>     let y = may_loop();
+let y = x;                      let z = may_loop();
+let z = x;                      return y == z;
+return y == z
+\end{verbatim}
+F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
+terminology, and also call such effects \emph{relevant}.
+
+With these distinction in place, we can \emph{define} a pure
+expression as one which is all three of central, copyable, and
+discardable.
+
+In addition to working with effects, compiler intermediate
+representations also have to track accesses to memory. This
+information is typically gathered via ad-hoc alias analyses, but a
+compositional alternative to these analyses is to track this
+information via substructural type systems such as linear and
+affine types, which associate aliasing and ownership information
+with individual substructural values. Having aliasing information
+justifies important optimisations such as loop hoisting: 
+\begin{verbatim}
+while(e) {              x = read(a);
+  x = read(a);          while(e) { 
+  body             ==>    body
+}                       }
+\end{verbatim}
+This transformation is only sound if the body of the loop \texttt{body}
+does not access \texttt{a}, which is precisely the kind of frame condition
+that linear types can track.
+
+Moreover, since we already want to track whether effectful
+\emph{expressions} can be duplicated or discarded, the very same
+machinery can be reused to track whether \emph{values} are
+linear/affine/relevant.
+
+Historically, compiler writers have treated the semantics of their IR
+quite informally, since the "intended model" of the language is a
+simple sequential machine operating over a heap which is a flat array
+of bytes. Unfortunately, over time both the intended model and the
+actual hardware have drifted away from this simple mental model.
+Compiler frameworks have to model concurrent accesses to memory in
+order to compile multithreaded code, and furthermore the IR's memory
+model has to be sound with respect to the actual weak memory models
+supported by hardware.
+
+As a result, it is now much harder to judge whether a particular
+optimisation is sound or not. Moreover, different kinds of hardware
+such as x86, GPUs, and ARM all offer different memory models. This
+means that we need an abstract semantics for compiler IRs, which can
+permit a separation of concerns. This will let compiler writers
+validate their optimisations against the abstract semantics, and will
+let us separately show that a particular machine model faithfully
+implements the semantics.
+
+In this paper, we show that this is possible. We give a type system
+for SSA which uses substructural types to both track effects at a
+fine-grained level, and to track ownership of values. We then
+rigorously justify a variety of optimisations by reference to an
+abstract categorical axiomatisation of this language. We then show
+that our categorical axioms have a variety of concrete models,
+particularly including one of TSO-style weak memory originally
+introduced by \citet{sparky}. This lets us argue that a host of
+SSA-based optimisations are sound with respect to the x86 memory
+model.
 
@@
 \begin{itemize}
-\item We have a syntax and type system for SSA-style programs.
-We include support for substructural types, and our type-theoretic syntax makes formulating inlining and rewriting easier than in the traditional CFG-based formulations.
-\item We show that our language has a compositional intepretation in any effectful category with an Elgot structure.
-We use this structure to give generic proofs of the expected substitution properties as well as prove the soundness of an equational theory for the language.
-\item Our generic equational theory justifies a variety of powerful optimisation techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and strength reduction. 
-\item We show how Elgot monads give rise to models of our language.
+\item We give a type system for SSA-style programs, which has a full substitution theorem and equational theory.
+This language includes substructural types, both for controlling effects and tracking ownership in values, and our strong substitution principle makes formulating inlining and rewriting optimisations easier than in the traditional CFG-based formulations.
+\item We give an axiomatic semantics for our language, by showing that it has a compositional intepretation in any effectful category with an Elgot structure.
+We use this structure to give generic proofs of the soundness of substitution and rewriting properties.
+\item Our abstract semantics justifies a variety of powerful optimisation techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and strength reduction. 
+\item We show how Elgot monads give rise to instances of our categorical semantics.
 We use this to derive semantics from a general trace monad, and show that various monad transformers such as state transformers preserve the needed Elgot structure.
@@
 This thus gives rise to an SSA-based IR with support for weak memory operations, which is fully semantically-justified.
-\item \todo{We also have a small implementation, which \ldots}
-\item Our weak memory semantics and our results about Elgot monads are formalised in the Lean theorem prover. \TODO{Other stuff}
+\item Our weak memory semantics and our results about Elgot monads are formalised in the Lean theorem prover.
 \end{itemize}

### 2023-11-10 pldi24/paper.tex
Cleaning, improved example programs

@@
 \usepackage{subfig}
+\usepackage{fancyvrb}
 
@@
 
-Static Single Assignment, or SSA, has been a cornerstone of compiler
-design ever since its introduction by
-\cite{ssa-original,rosen-gvn-1988} in the late 1980s. Older compiler
-intermediate representations permitted the same local variable to be
-assigned multiple times, whereas SSA requires each variable bindings
-to be immutable -- they cannot be modified after initialisation. This
-makes many dataflow analyses easier to compute, and (just as in
-functional languages) makes it much easier to justify compiler
-optimisations in terms of equational rewrites.
-
-In fact, the relationship between SSA form and functional programming
-has been known for decades: \citet{kelsey-ssa-cps} showed that SSA can
-be seen as a subset of continuation-passing style, and
-\citet{appel-ssa} built on this to observe that every SSA program can
-be viewed as a collection of first-order functions which tail-call one
-another. This makes the equivalence of the functional view to the
-traditional $\phi$-node presentation very easy to see. In the
-functional style, each tail call lists its arguments. In contrast, a
-$\phi$-node collects together the arguments from every callsite, which
-lets the tail-calls/jumps to omit their arguments. Thus, the main
-difference between SSA and a functional program is where in the
-program text the arguments to a tail call are listed.
-
-This ``basic blocks with arguments'' presentation of SSA has become
-very popular in industry, with compiler backends such as
-Cranelift~\cite{cranelift}, MLIR~\cite{mlir}, and SIL~\cite{SIL} all
-using this representation. This tight connection might suggest that
-the semantics of SSA representations is trivial: we can just interpret
-an SSA program as a corresponding functional program, and then rewrite
-the SSA programs using the equational theory of functional programs.
-
-While this is a perfectly sound thing to do, it is not strong enough
-to write a good compiler. Compiler IRs like SSA are fundamentally
-about effectful programs, and the equational theory of effectful
-functional programs is very weak. Instead of a full substitution rule,
-the equational theory of impure functional languages only justifies
-substituting values for variables (the $\beta$-value rule). This weak
-equational rule suffices to validate some basic compiler optimisations
-such as constant propagation and copy propagation, but is not strong
-enough to justify many other essential compiler optimisations. For
-example, hoisting a store out of a loop or conditional involves moving
-an effectful term (an assignment) out of its original position in the
-program. As another example, dead store elimination removes an
-effectful term from the program altogether. \todo{Add link to figure with code for these examples}
+Static Single Assignment, or SSA, has been a cornerstone of compiler design ever
+since its introduction by \cite{ssa-original,rosen-gvn-1988} in the late 1980s.
+Older compiler intermediate representations permitted the same local variable to
+be assigned multiple times, whereas SSA requires each variable bindings to be
+immutable -- they cannot be modified after initialisation. This makes many
+dataflow analyses easier to compute, and (just as in functional languages) makes
+it much easier to justify compiler optimisations in terms of equational
+rewrites.
+
+In fact, the relationship between SSA form and functional programming has been
+known for decades: \citet{kelsey-ssa-cps} showed that SSA can be seen as a
+subset of continuation-passing style, and \citet{appel-ssa} built on this to
+observe that every SSA program can be viewed as a collection of first-order
+functions which tail-call one another. This makes the equivalence of the
+functional view to the traditional $\phi$-node presentation very easy to see. In
+the functional style, each tail call lists its arguments. In contrast, a
+$\phi$-node collects together the arguments from every callsite, which lets the
+tail-calls/jumps to omit their arguments. Thus, the main difference between SSA
+and a functional program is where in the program text the arguments to a tail
+call are listed.
+
+This "basic blocks with arguments" presentation of SSA has become very popular
+in industry, with compiler backends such as Cranelift~\cite{cranelift},
+MLIR~\cite{mlir}, and SIL~\cite{SIL} all using this representation. This tight
+connection might suggest that the semantics of SSA representations is trivial:
+we can just interpret an SSA program as a corresponding functional program, and
+then rewrite the SSA programs using the equational theory of functional
+programs.
+
+While this is a perfectly sound thing to do, it is not strong enough to write a
+good compiler. Compiler IRs like SSA are fundamentally about effectful programs,
+and the equational theory of effectful functional programs is very weak. Instead
+of a full substitution rule, the equational theory of impure functional
+languages only justifies substituting values for variables (the $\beta$-value
+rule). This weak equational rule suffices to validate some basic compiler
+optimisations such as constant propagation and copy propagation, but is not
+strong enough to justify many other essential compiler optimisations. For
+example, hoisting a store out of a loop or conditional involves moving an
+effectful term (an assignment) out of its original position in the program. As
+another example, dead store elimination removes an effectful term from the
+program altogether. \todo{Add link to figure with code for these examples}
 
 These examples illustrate that many optimisations require taking
-\emph{effectful} terms and moving them around and substituting them
-for variables. Since this is not a generally valid program
-transformation, we need a finer classification of programs than
-merely pure and effectful to judge when a substitution is allowed.
-
-A useful taxonomy was introduced by \citet{fuhrmann}, who classified
-effects based on (1) whether or not the order of effects mattered,
-(2) whether it was sound to discard an effectful operation, and (3)
-whether it was sound to duplicate an effectful operation.
-
-Suppose we substitute \texttt{print("hello")} for $x$ in the program
-below:
-\begin{verbatim}
-let x = print("hello");            let y = print("world");
-let y = print("world");   =/=>     let z = print("hello");
-let z = x;                         ...
+\emph{effectful} terms and moving them around and substituting them for
+variables. Since this is not a generally valid program transformation, we need a
+finer classification of programs than merely pure and effectful to judge when a
+substitution is allowed.
+
+A useful taxonomy was introduced by \citet{fuhrmann-direct-1999}, classifying
+effects based on (1) whether or not the order of effects mattered, (2) whether
+it was sound to discard an effectful operation, and (3) whether it was sound to
+duplicate an effectful operation. Suppose we substitute \texttt{print("hello")}
+for $x$ in the program below:
+\begin{equation}
+\begin{BVerbatim}[baseline=c]
+let x = print("hello");
+let y = print("world");
+let z = x;   
 ...
-\end{verbatim}
-
-In this case, substituting a term for a variable changes the order the
-I/O actions are performed, which changes the program's observable
-behaviour. On the other hand, suppose we have an effect which increments
-a counter. Then the same transformation is sound. 
-\begin{verbatim}
-let x = increment(a);            let y = increment(b);
-let y = increment(b);   ===>     let z = increment(a);
-let z = x;                        ...
+\end{BVerbatim}
+\qquad \qquad \centernot\implies \qquad \qquad
+\begin{BVerbatim}[baseline=c]
+let y = print("world");
+let x = print("hello");
 ...
-\end{verbatim}
-
-F\"{u}hrmann called these commutative effects \emph{central}, which we
-also call \emph{linear}.
-
-It can also matter whether an effect is performed at all or not. For example:
-\begin{verbatim}
-let x = may_loop();    =/=>      return 5;
+\end{BVerbatim}
+\end{equation}
+In this case, substituting a term for a variable changes the order the I/O
+actions are performed, which changes the program's observable behaviour. On the
+other hand, suppose we have an effect which increments a counter. Then the same
+transformation is sound, since \(x\) is used exactly once:
+\begin{equation}
+\begin{BVerbatim}[baseline=c]
+let x = increment(a);
+let y = increment(b);
+let z = x;
+...
+\end{BVerbatim}
+\qquad \qquad \implies \qquad \qquad
+\begin{BVerbatim}[baseline=c]
+let y = increment(b);
+let z = increment(a);
+...
+\end{BVerbatim}
+\end{equation}
+F\"{u}hrmann called such commutative effects \emph{central}, which we also call
+\emph{linear}. It can also matter whether an effect is performed at all or not.
+For example, consider:
+\begin{equation}
+\begin{BVerbatim}[baseline=c]
+let x = may_loop();
 return 5 
-\end{verbatim}
-In this example, even though $x$ is not used, the call to \texttt{may_loop()} cannot
-be erased, as the observable behaviour of the program can change if it goes into an
-infinite loop.
-
-On the other hand, if the effectful operation is merely nondeterministic, then it is
-safe to discard results. 
-\begin{verbatim}
-let x = rand();    ===>          return 5;
+\end{BVerbatim}
+\qquad \qquad \centernot\implies \qquad \qquad
+\begin{BVerbatim}[baseline=c]
 return 5
-\end{verbatim}
-F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use his
-terminology, and also call such effects \emph{affine}.
-
-Now consider the following example: 
-\begin{verbatim}
-let x = rand();    =/=>     let y = rand();
-let y = x;                  let z = rand();
-let z = x;                  return y == z;
+\end{BVerbatim}
+\end{equation}
+In this example, even though $x$ is not used, the call to \texttt{may_loop()}
+cannot be erased, as the observable behaviour of the program can change if it
+goes into an infinite loop. On the other hand, if the effectful operation is
+merely nondeterministic, then it is safe to discard results:
+\begin{equation}
+\begin{BVerbatim}[baseline=c]
+let x = rand();
+return 5 
+\end{BVerbatim}
+\qquad \qquad \implies \qquad \qquad
+\begin{BVerbatim}[baseline=c]
+return 5
+\end{BVerbatim}
+\end{equation}
+F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use
+his terminology, and also call such effects \emph{affine}. Now consider the
+following example: 
+\begin{equation}
+\begin{BVerbatim}[baseline=c]
+let x = rand();
+let y = x;
+let z = x;
 return y == z 
-\end{verbatim}
-
-In this case, substituting for $x$ duplicates the call to \texttt{rand()}, and can change
-this from a program which always returns true to one which can return either true or false.
-
-On the other hand, if the effect is the possibility of nontermination, then the
-substitution is sound. 
-\begin{verbatim}
-let x = may_loop();    ===>     let y = may_loop();
-let y = x;                      let z = may_loop();
-let z = x;                      return y == z;
+\end{BVerbatim}
+\qquad \qquad \centernot\implies \qquad \qquad
+\begin{BVerbatim}[baseline=c]
+let y = rand();
+let z = rand();
 return y == z
-\end{verbatim}
+\end{BVerbatim}
+\end{equation}
+In this case, substituting for $x$ duplicates the call to \texttt{rand()}, and
+can change this from a program which always returns true to one which can return
+either true or false. On the other hand, if the effect is the possibility of
+nontermination, then the substitution is sound. 
+\begin{equation}
+\begin{BVerbatim}[baseline=c]
+let x = may_loop();
+let y = x;
+let z = x;
+return y == z 
+\end{BVerbatim}
+\qquad \qquad \implies \qquad \qquad
+\begin{BVerbatim}[baseline=c]
+let y = may_loop();
+let z = may_loop();
+return y == z
+\end{BVerbatim}
+\end{equation}
 F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
-terminology, and also call such effects \emph{relevant}.
-
-With these distinction in place, we can \emph{define} a pure
-expression as one which is all three of central, copyable, and
-discardable.
-
-In addition to working with effects, compiler intermediate
-representations also have to track accesses to memory. This
-information is typically gathered via ad-hoc alias analyses, but a
-compositional alternative to these analyses is to track this
-information via substructural type systems such as linear and
-affine types, which associate aliasing and ownership information
-with individual substructural values. Having aliasing information
-justifies important optimisations such as loop hoisting: 
-\begin{verbatim}
-while(e) {              x = read(a);
-  x = read(a);          while(e) { 
-  body             ==>    body
-}                       }
-\end{verbatim}
-This transformation is only sound if the body of the loop \texttt{body}
-does not access \texttt{a}, which is precisely the kind of frame condition
-that linear types can track.
-
-Moreover, since we already want to track whether effectful
-\emph{expressions} can be duplicated or discarded, the very same
-machinery can be reused to track whether \emph{values} are
-linear/affine/relevant.
-
-Historically, compiler writers have treated the semantics of their IR
-quite informally, since the "intended model" of the language is a
-simple sequential machine operating over a heap which is a flat array
-of bytes. Unfortunately, over time both the intended model and the
-actual hardware have drifted away from this simple mental model.
-Compiler frameworks have to model concurrent accesses to memory in
-order to compile multithreaded code, and furthermore the IR's memory
-model has to be sound with respect to the actual weak memory models
+terminology, and also call such effects \emph{relevant}. With these distinction
+in place, we can \emph{define} a pure expression as one which is all three of
+central, copyable, and discardable.
+
+In addition to working with effects, compiler intermediate representations also
+have to track accesses to memory. This information is typically gathered via
+ad-hoc alias analyses, but a compositional alternative to these analyses is to
+track this information via substructural type systems such as linear and affine
+types, which associate aliasing and ownership information with individual
+substructural values. Having aliasing information justifies important
+optimisations such as loop hoisting: 
+\begin{equation}
+\begin{BVerbatim}[baseline=c]
+while (e) {
+  x = read(a);
+  body
+}
+\end{BVerbatim}
+\qquad \qquad \implies \qquad \qquad
+\begin{BVerbatim}[baseline=c]
+x = read(a);
+while(e) {
+  body
+}
+\end{BVerbatim}
+\end{equation}
+This transformation is only sound if the body of the loop \texttt{body} does not
+access \texttt{a}, which is precisely the kind of frame condition that linear
+types can track.
+
+Moreover, since we already want to track whether effectful \emph{expressions}
+can be duplicated or discarded, the very same machinery can be reused to track
+whether \emph{values} are linear/affine/relevant.
+
+Historically, compiler writers have treated the semantics of their IR quite
+informally, since the "intended model" of the language is a simple sequential
+machine operating over a heap which is a flat array of bytes. Unfortunately,
+over time both the intended model and the actual hardware have drifted away from
+this simple mental model. Compiler frameworks have to model concurrent accesses
+to memory in order to compile multithreaded code, and furthermore the IR's
+memory model has to be sound with respect to the actual weak memory models
 supported by hardware.
 
-As a result, it is now much harder to judge whether a particular
-optimisation is sound or not. Moreover, different kinds of hardware
-such as x86, GPUs, and ARM all offer different memory models. This
-means that we need an abstract semantics for compiler IRs, which can
-permit a separation of concerns. This will let compiler writers
-validate their optimisations against the abstract semantics, and will
-let us separately show that a particular machine model faithfully
+As a result, it is now much harder to judge whether a particular optimisation is
+sound or not. Moreover, different kinds of hardware such as x86, GPUs, and ARM
+all offer different memory models. This means that we need an abstract semantics
+for compiler IRs, which can permit a separation of concerns. This will let
+compiler writers validate their optimisations against the abstract semantics,
+and will let us separately show that a particular machine model faithfully
 implements the semantics.
 
-In this paper, we show that this is possible. We give a type system
-for SSA which uses substructural types to both track effects at a
-fine-grained level, and to track ownership of values. We then
-rigorously justify a variety of optimisations by reference to an
-abstract categorical axiomatisation of this language. We then show
-that our categorical axioms have a variety of concrete models,
-particularly including one of TSO-style weak memory originally
-introduced by \citet{sparky}. This lets us argue that a host of
-SSA-based optimisations are sound with respect to the x86 memory
-model.
+In this paper, we show that this is possible. We give a type system for SSA
+which uses substructural types to both track effects at a fine-grained level,
+and to track ownership of values. We then rigorously justify a variety of
+optimisations by reference to an abstract categorical axiomatisation of this
+language. We then show that our categorical axioms have a variety of concrete
+models, particularly including one of TSO-style weak memory originally
+introduced by \citet{sparky}. This lets us argue that a host of SSA-based
+optimisations are sound with respect to the x86 memory model.
 
@@
 \begin{itemize}
-\item We give a type system for SSA-style programs, which has a full substitution theorem and equational theory.
-This language includes substructural types, both for controlling effects and tracking ownership in values, and our strong substitution principle makes formulating inlining and rewriting optimisations easier than in the traditional CFG-based formulations.
-\item We give an axiomatic semantics for our language, by showing that it has a compositional intepretation in any effectful category with an Elgot structure.
-We use this structure to give generic proofs of the soundness of substitution and rewriting properties.
-\item Our abstract semantics justifies a variety of powerful optimisation techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and strength reduction. 
-\item We show how Elgot monads give rise to instances of our categorical semantics.
-We use this to derive semantics from a general trace monad, and show that various monad transformers such as state transformers preserve the needed Elgot structure.
-\item We use these tools to demonstrate the existence of a variety of concrete models satisfying these categorical axioms.
-Starting first from simple languges such as state plus printing, we also show how more challenging semantics such as the TSO semantics for weak memory also fit into our framework.
-This thus gives rise to an SSA-based IR with support for weak memory operations, which is fully semantically-justified.
-\item Our weak memory semantics and our results about Elgot monads are formalised in the Lean theorem prover.
+\item We give a type system for SSA-style programs, which has a full
+substitution theorem and equational theory. This language includes substructural
+types, both for controlling effects and tracking ownership in values, and our
+strong substitution principle makes formulating inlining and rewriting
+optimisations easier than in the traditional CFG-based formulations. 
+\item We give an axiomatic semantics for our language, by showing that it has a
+compositional intepretation in any effectful category with an Elgot structure.
+We use this structure to give generic proofs of the soundness of substitution
+and rewriting properties.
+\item Our abstract semantics justifies a variety of powerful optimisation
+techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and
+strength reduction. 
+\item We show how Elgot monads give rise to instances of our categorical
+semantics. We use this to derive semantics from a general trace monad, and show
+that various monad transformers such as state transformers preserve the needed
+Elgot structure. 
+\item We use these tools to demonstrate the existence of a variety of concrete
+models satisfying these categorical axioms. Starting first from simple languges
+such as state plus printing, we also show how more challenging semantics such as
+the TSO semantics for weak memory also fit into our framework. This thus gives
+rise to an SSA-based IR with support for weak memory operations, which is fully
+semantically-justified.
+\item Our weak memory semantics and our results about Elgot monads are
+formalised in the Lean theorem prover.
 \end{itemize}
@@
 
-In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks, which can be viewed as \textit{label}, followed by a sequence of \textit{instructions}, each defining a \textit{variable}, followed by a \textit{terminator}.
-A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
-\TODO{text}
+In a standard SSA-based compiler, to represent a function, we begin by
+decomposing it into basic blocks, which can be viewed as \textit{label},
+followed by a sequence of \textit{instructions}, each defining a
+\textit{variable}, followed by a \textit{terminator}. A program written in this
+manner being in ``SSA form'' simply means that every variable name is defined
+exactly once. For example, \TODO{text}
 
@@
 In Figure \ref{fig:ssa-grammar}, we present a formal grammar for our SSA syntax.
-We begin by giving a grammar for \textit{types} \(A, B, C\), parametrized by a set of \textit{base types} \(X\).
-We require the existence of a \textit{unit type} \(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all tensor products \(A \otimes B\); we write \(\otimes\) to underscore the fact that types are allowed to be substructural.
-
-We separate program fragments into 3 main syntactic categories: \textit{expressions} \(a, b, c, e\), \textit{blocks} \(s, t\), and \textit{label-sets} \(L\).
-While we make a distinction between \textit{pure} and \textit{impure} expressions, this is done at the level of typing rules rather than via syntax.
-A \textit{block} is conceptualized as a series of \textit{let-bindings} followed by a \textit{terminator}.
-Here, a terminator is either an unconditional branch to another block with an argument or an if-statement.
-A block may be encapsulated within a \textit{where-block}, which defines other blocks the inner block can branch to; these blocks can also branch to each other in a mutually recursive fashion.
-Similarly to blocks, expressions may also contain let-bindings; this is somewhat of a departure from classical SSA, in which there is only one kind of variable binding, but facilitates various optimizations and corresponds more closely to the in-memory representation, where pure expressions are represented as E-graph nodes.
-Note that we do \textit{not} give a grammar or semantics for top-level functions, which we can view as simply a list of arguments and a block.
-Except for inlining, we will only consider the optimization and analysis of a single function body at a time.
-
-We proceed to give a grammar for \textit{contexts} \(\Gamma\), which consist of a list of variables \(x\) given a type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\).
-This tells us whether the variable is \textit{affine} (i.e., can be ignored) and/or \textit{relevant} (i.e., whether it can be used multiple times).
-A variable which is both affine and relevant is called \textit{intuitionistic}.
-Whether we can use a variable also depends on the \textit{linearity} of its type.
-
-We can then give a grammar for \textit{label-contexts} \(\ms{L}\), which consist of a list of \textit{labels} each annotated with a context \(\Gamma\), denoting the variables which must be live on entry to the block (corresponding to the set a dominance frontier would compute), and an \textit{argument} \(A\).
+We begin by giving a grammar for \textit{types} \(A, B, C\), parametrized by a
+set of \textit{base types} \(X\). We require the existence of a \textit{unit
+type} \(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all tensor products
+\(A \otimes B\); we write \(\otimes\) to underscore the fact that types are
+allowed to be substructural.
+
+We separate program fragments into 3 main syntactic categories:
+\textit{expressions} \(a, b, c, e\), \textit{blocks} \(s, t\), and
+\textit{label-sets} \(L\). While we make a distinction between \textit{pure} and
+\textit{impure} expressions, this is done at the level of typing rules rather
+than via syntax. A \textit{block} is conceptualized as a series of
+\textit{let-bindings} followed by a \textit{terminator}. Here, a terminator is
+either an unconditional branch to another block with an argument or an
+if-statement. A block may be encapsulated within a \textit{where-block}, which
+defines other blocks the inner block can branch to; these blocks can also branch
+to each other in a mutually recursive fashion. Similarly to blocks, expressions
+may also contain let-bindings; this is somewhat of a departure from classical
+SSA, in which there is only one kind of variable binding, but facilitates
+various optimizations and corresponds more closely to the in-memory
+representation, where pure expressions are represented as E-graph nodes. Note
+that we do \textit{not} give a grammar or semantics for top-level functions,
+which we can view as simply a list of arguments and a block. Except for
+inlining, we will only consider the optimization and analysis of a single
+function body at a time.
+
+We proceed to give a grammar for \textit{contexts} \(\Gamma\), which consist of
+a list of variables \(x\) given a type \(A\) annotated with a \textit{quantity}
+\(q \subseteq \tint\). This tells us whether the variable is \textit{affine}
+(i.e., can be ignored) and/or \textit{relevant} (i.e., whether it can be used
+multiple times). A variable which is both affine and relevant is called
+\textit{intuitionistic}. Whether we can use a variable also depends on the
+\textit{linearity} of its type.
+
+We can then give a grammar for \textit{label-contexts} \(\ms{L}\), which consist
+of a list of \textit{labels} each annotated with a context \(\Gamma\), denoting
+the variables which must be live on entry to the block (corresponding to the set
+a dominance frontier would compute), and an \textit{argument} \(A\).
 
@@
 
-We can now list out our primary typing judgements, which are given in Figure \ref{fig:ssa-judgements}.
-We have three main typing judgements: \(\hasty{\Gamma}{p}{x}{A}{q}\) is the typing judgement for expressions, which says that the expression \(a\) has type \(A\) in context \(\Gamma\).
-The annotation \(q\) means that the expression itself has quantity \(q \subseteq \tint\), while the annotation \(p\) means that the expression has purity \(p\) (1 for pure, 0 for impure).
-\(\haslb{\Gamma}{t}{\ms{L}}\) is the typing block for blocks, which says that, using variables in \(\Gamma\), the block \(t\) exits by jumping to labels in \(\ms{L}\).
-Finally, \(\lhaslb{\ms{L}}{L}{\ms{K}}\) says that the label-set \(L\) defines the labels \(\ms{L}\) in terms of the labels \(\ms{K}\) (blocks in \(L\) are allowed to recursively call into \(\ms{L}\) or exit to \(\ms{K}\)).
+We can now list out our primary typing judgements, which are given in Figure
+\ref{fig:ssa-judgements}. We have three main typing judgements:
+\(\hasty{\Gamma}{p}{x}{A}{q}\) is the typing judgement for expressions, which
+says that the expression \(a\) has type \(A\) in context \(\Gamma\). The
+annotation \(q\) means that the expression itself has quantity \(q \subseteq
+\tint\), while the annotation \(p\) means that the expression has purity \(p\)
+(1 for pure, 0 for impure). \(\haslb{\Gamma}{t}{\ms{L}}\) is the typing block
+for blocks, which says that, using variables in \(\Gamma\), the block \(t\)
+exits by jumping to labels in \(\ms{L}\). Finally,
+\(\lhaslb{\ms{L}}{L}{\ms{K}}\) says that the label-set \(L\) defines the labels
+\(\ms{L}\) in terms of the labels \(\ms{K}\) (blocks in \(L\) are allowed to
+recursively call into \(\ms{L}\) or exit to \(\ms{K}\)).
 
 To define these, we will need a few basic structural judgements.
-\(\csplits{\Gamma}{\Delta}{\Xi}\) means that the context \(\Gamma\) can be split into contexts \(\Delta\) and \(\Xi\).
-We use the notion of splitting, rather than weakening, a context as primitive to be able to deal with linearity, since not all variables can be dropped or duplicated.
-This can be used to define weakening \(\cwk{\Gamma}{\Delta}\) as simply "splitting \(\Gamma\) into \(\Delta\) and the empty context", since a split is allowed to drop affine variables.
-Finally, \(\lwk{\ms{L}}{\ms{K}}\) means that the label-set \(\ms{L}\) weakens to the label-set \(\ms{K}\), i.e.
-that \(\ms{K}\) has more labels, but for shared labels, needs less variables to be live (but the same non-affine variables!) The typing rules for these judgements are given in Figure \ref{fig:ssa-structural}.
+\(\csplits{\Gamma}{\Delta}{\Xi}\) means that the context \(\Gamma\) can be split
+into contexts \(\Delta\) and \(\Xi\). We use the notion of splitting, rather
+than weakening, a context as primitive to be able to deal with linearity, since
+not all variables can be dropped or duplicated. This can be used to define
+weakening \(\cwk{\Gamma}{\Delta}\) as simply "splitting \(\Gamma\) into
+\(\Delta\) and the empty context", since a split is allowed to drop affine
+variables. Finally, \(\lwk{\ms{L}}{\ms{K}}\) means that the label-set \(\ms{L}\)
+weakens to the label-set \(\ms{K}\), i.e. that \(\ms{K}\) has more labels, but
+for shared labels, needs less variables to be live (but the same non-affine
+variables!) The typing rules for these judgements are given in Figure
+\ref{fig:ssa-structural}.
 
@@
 
-\section{Implementation}
-
-\TODO{this}
-
-\TODO{remove? time + space...}
-
 \section{Related Work}

### 2023-11-10 pldi24/paper.tex
Cleaned up figure slightly

@@
 \usepackage{xcolor}
-\usepackage{subfig}
+\usepackage{subcaption}
 \usepackage{fancyvrb}
@@
 
-\newsavebox\notssafib
-
-\begin{lrbox}{\notssafib}
-\begin{minipage}{0.3\textwidth}
-\begin{lstlisting}[frame=none,numbers=none]
+\begin{figure}
+  \centering
+  \begin{subfigure}[t]{0.33\textwidth}
+    \begin{BVerbatim}[baseline=t, gobble=3]
     'entry:
@@
       ret m
-\end{lstlisting}
-\end{minipage}
-\end{lrbox}
-
-\newsavebox\ssafib
-
-\begin{lrbox}{\ssafib}
-\begin{minipage}{0.3\textwidth}
-\begin{lstlisting}[frame=none,numbers=none]
+    \end{BVerbatim}
+    \caption{3-address code}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.32\textwidth}
+    \begin{BVerbatim}[baseline=t, gobble=3]
       'entry:
@@
       ret m3
-\end{lstlisting}
-\end{minipage}
-\end{lrbox}
-
-\newsavebox\ourfib
-
-\begin{lrbox}{\ourfib}
-\begin{minipage}{0.3\textwidth}
-\begin{lstlisting}[frame=none,numbers=none]
+    \end{BVerbatim}
+    \caption{Basic blocks with arguments}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.33\textwidth}
+    \begin{BVerbatim}[baseline=t, gobble=3]
     let m0 = 0;
@@
     if eq i 0 {
-  br 'exit(m0)
+      br 'exit (m0)
     } else {
-  br 'loop(i, m0, n0)
+      br 'loop (i, m0, n0)
     } where
@@
       if eq i1 0 {
-    br 'exit(m2)
+        br 'exit (m2)
       } else {
-    br 'loop(i1, m2, n2)
+        br 'loop (i1, m2, n2)
       }
     'exit(m3) => ret m3
-\end{lstlisting}
-\end{minipage}
-\end{lrbox}
-
-\begin{figure}
-  \centering
-  \subfloat[3-address code]{\usebox\notssafib}
-  \quad
-  \subfloat[Basic blocks with arguments]{\usebox\ssafib}
-  \quad
-  \subfloat[Our syntax]{\usebox\ourfib}
+    \end{BVerbatim}
+    \caption{Our syntax}
+  \end{subfigure}
   \caption{A simple program for computing the \texttt{i}th Fibonacci number, represented as 3-address code, basic-blocks with arguments, and in our SSA syntax}

### 2023-11-11 pldi24/paper.tex
Tap

@@
 access \texttt{a}, which is precisely the kind of frame condition that linear
-types can track.
-
-Moreover, since we already want to track whether effectful \emph{expressions}
-can be duplicated or discarded, the very same machinery can be reused to track
-whether \emph{values} are linear/affine/relevant.
+types can track. Moreover, since we already want to track whether effectful
+\emph{expressions} can be duplicated or discarded, the very same machinery can
+be reused to track whether \emph{values} are linear/affine/relevant.
 
@@
 
-\section{Background}
+\section{SSA Syntax}
 
@@
 \textit{variable}, followed by a \textit{terminator}. A program written in this
-manner being in ``SSA form'' simply means that every variable name is defined
-exactly once. For example, \TODO{text}
+manner being in "SSA form" simply means that every variable name is defined
+exactly once.
+For example, the program in Figure \ref{fig:3-addr} has 3 basic blocks, one of which is a distinguished \textit{entry block}, but is \textit{not} in SSA form, since the variables \(m, n\) and \(i\) are each modified through subsequent assignments (\(t\) is recomputed on every iteration of the loop, but since it only has one definition, it \textit{is} in SSA form).
+To convert this program to SSA form, we would traditionally use \(\phi\)-functions, or, equivalently, annotate each basic block with arguments, as in Figure \ref{fig:bb-arg}.
 
@@
       n = t
-      N = sub i 1
+      i = sub i 1
       brz i 'exit 'loop
@@
     \caption{3-address code}
+    \label{fig:3-addr}
   \end{subfigure}
@@
     \caption{Basic blocks with arguments}
+    \label{fig:bb-arg}
   \end{subfigure}
@@
     \end{BVerbatim}
+    \label{fig:our-syn}
     \caption{Our syntax}
@@
 
-\TODO{basic-blocks with arguments}
-
-\TODO{dominance frontiers}
+\TODO{dominance frontiers?}
 
@@
 
-\TODO{what we're doing}
-
-\section{SSA Syntax}
+\TODO{seque into what we're doing}
 
@@
       <\(q\)> ::= \(\varnothing\) 
-      \;|\; \(\{\taff\}\) 
-      \;|\; \(\{\trel\}\) 
-      \;|\; \(\{\taff, \trel\}\)
+      \;|\; \(\taff\) 
+      \;|\; \(\trel\) 
+      \;|\; \(\tint\)
     \end{grammar}

### 2023-11-11 pldi24/paper.tex
Tap

@@
     \end{BVerbatim}
+    \caption{Our syntax}
     \label{fig:our-syn}
-    \caption{Our syntax}
   \end{subfigure}
@@
 
-\TODO{dominance frontiers?}
-
-\TODO{inductive presentation}
-
-\TODO{seque into what we're doing}
+For our type theoretic presentation of SSA, we will treat a very slight generalization of the standard basic blocks with arguments syntax, in that we will support \textit{nested scopes} and, in particular, \textit{if-statements}.
+The program in Figure \ref{fig:bb-arg} translated to our syntax is given in Figure \ref{fig:our-syn}.
+These do not introduce any new expressivity: an if-statement can be viewed as syntactic sugar for a conditional branch to one of two anonymous basic blocks, while nested scopes simply \textit{restrict} where we can branch to.
+In particular, in Section \ref{ssec:ssa-norm}, we will formalize interconversion with traditional SSA and show how it is semantically sound to simply merge all nested scopes.
+The point of the slightly modified syntax is simply to make some of our metatheoretic results more practical to state.
 
@@
 
+\subsection{Normalization}
+
+\label{ssec:ssa-norm}
+
+\TODO{this}
+
 \subsection{Optimizations}

### 2023-11-11 pldi24/paper.tex
Tap

@@
 
+\TODO{note that label-set weakening is "backwards"}
+
+\TODO{be consistent with label-context vs label-set}
+
+\TODO{rewrite \textit{all} this \textit{way} better}
+
+The rules \rle{unit-lin} and \rle{bool-lin} state that both the unit and boolean types must be intuitionistic; the rule \rle{pair-lin} says that the linearity of a pair is simply the intersection of the linearity of its components.
+The rules \rle{nil-lin} and \rle{cons-lin} together say that the linearity of a context is the intersection of the linearity of its variables, where the linearity of a variable is the intersection of the linearity of its type and its quantity annotation.
+
+The rules for context splitting \(\csplits{\Gamma}{\Delta}{\Xi}\) work by partitioning variables in \(\Gamma\) into either \(\Delta\) or \(\Xi\), only permitting a variable to be in both if it is \textit{relevant} and only permitting a variable to be ignored (dropped) if it is \textit{affine}.
+In particular, the rule \rle{split-nil} provides a base case by saying that the empty context can be split into two empty contexts, while the rules \rle{split-left} and \rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively; both these rules also allow us to reduce the quantity \(q\) to a quantity \(r \subseteq q\), "forgetting" some of our capabilities.
+The rule \rle{split-dup} allows the duplication of a variable which has a relevant type \(A\) \textit{and} a relevant quantity \(q\), with the option to weaken the quantity to \textit{different} quantities \(r\) and \(s\) on the left-hand side and right-hand sides respectively.
+Finally, the rule \rle{split-aff} simply allows us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.
+The rule \rle{join-nil} is the analogue to \rle{split-nil} for \textit{label-contexts}, providing a base case in that the empty label weakens into itself.
+The rule \rle{join-cons} says that if \(\ms{L}\) weakens into \(\ms{K}\) and \(\Delta\) weakens into \(\Gamma\), then \(\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}\), allowing us to "forget" any extra variables dropped by \(\cwk{\Gamma}{\Delta}\).
+Finally, the rule \rle{join-zero} simply says that if \(\lwk{\ms{L}}{\ms{K}}\), then \(\ms{L}\) also weakens to any extension of \(\ms{K}\).
+
 \begin{figure}
@@
 \end{figure}
-
-\TODO{some text about each structural rule}
 

### 2023-11-12 pldi24/paper.tex
Tap at SSA text

--- 2023-11-11 pldi24/paper.tex
@@
 
-For our type theoretic presentation of SSA, we will treat a very slight generalization of the standard basic blocks with arguments syntax, in that we will support \textit{nested scopes} and, in particular, \textit{if-statements}.
-The program in Figure \ref{fig:bb-arg} translated to our syntax is given in Figure \ref{fig:our-syn}.
-These do not introduce any new expressivity: an if-statement can be viewed as syntactic sugar for a conditional branch to one of two anonymous basic blocks, while nested scopes simply \textit{restrict} where we can branch to.
-In particular, in Section \ref{ssec:ssa-norm}, we will formalize interconversion with traditional SSA and show how it is semantically sound to simply merge all nested scopes.
-The point of the slightly modified syntax is simply to make some of our metatheoretic results more practical to state.
-
-In Figure \ref{fig:ssa-grammar}, we present a formal grammar for our SSA syntax.
-We begin by giving a grammar for \textit{types} \(A, B, C\), parametrized by a
-set of \textit{base types} \(X\). We require the existence of a \textit{unit
-type} \(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all tensor products
-\(A \otimes B\); we write \(\otimes\) to underscore the fact that types are
-allowed to be substructural.
-
-We separate program fragments into 3 main syntactic categories:
-\textit{expressions} \(a, b, c, e\), \textit{blocks} \(s, t\), and
-\textit{label-sets} \(L\). While we make a distinction between \textit{pure} and
-\textit{impure} expressions, this is done at the level of typing rules rather
-than via syntax. A \textit{block} is conceptualized as a series of
-\textit{let-bindings} followed by a \textit{terminator}. Here, a terminator is
-either an unconditional branch to another block with an argument or an
-if-statement. A block may be encapsulated within a \textit{where-block}, which
-defines other blocks the inner block can branch to; these blocks can also branch
-to each other in a mutually recursive fashion. Similarly to blocks, expressions
-may also contain let-bindings; this is somewhat of a departure from classical
-SSA, in which there is only one kind of variable binding, but facilitates
-various optimizations and corresponds more closely to the in-memory
-representation, where pure expressions are represented as E-graph nodes. Note
-that we do \textit{not} give a grammar or semantics for top-level functions,
-which we can view as simply a list of arguments and a block. Except for
-inlining, we will only consider the optimization and analysis of a single
-function body at a time.
-
-We proceed to give a grammar for \textit{contexts} \(\Gamma\), which consist of
-a list of variables \(x\) given a type \(A\) annotated with a \textit{quantity}
-\(q \subseteq \tint\). This tells us whether the variable is \textit{affine}
-(i.e., can be ignored) and/or \textit{relevant} (i.e., whether it can be used
-multiple times). A variable which is both affine and relevant is called
-\textit{intuitionistic}. Whether we can use a variable also depends on the
-\textit{linearity} of its type.
-
-We can then give a grammar for \textit{label-contexts} \(\ms{L}\), which consist
-of a list of \textit{labels} each annotated with a context \(\Gamma\), denoting
-the variables which must be live on entry to the block (corresponding to the set
-a dominance frontier would compute), and an \textit{argument} \(A\).
+For our type theoretic presentation of SSA, we will treat a very slight
+generalization of the standard basic blocks with arguments syntax, in that we
+will support \textit{nested scopes} and, in particular, \textit{if-statements}.
+The program in Figure \ref{fig:bb-arg} translated to our syntax is given in
+Figure \ref{fig:our-syn}. These do not introduce any new expressivity: an
+if-statement can be viewed as syntactic sugar for a conditional branch to one of
+two anonymous basic blocks, while nested scopes simply \textit{restrict} where
+we can branch to. In particular, in Section \ref{ssec:ssa-norm}, we will
+formalize interconversion with traditional SSA and show how it is semantically
+sound to simply merge all nested scopes. The point of the slightly modified
+syntax is simply to make some of our metatheoretic results more practical to
+state.
+
+The grammar for our syntax is given in Figure \ref{fig:ssa-grammar}, while the
+typing judgements themselves are given in \ref{fig:ssa-judgements}. We begin by
+giving a grammar for \textit{types} \(A, B, C\), parametrized by a set of
+\textit{base types} \(X\); all syntactically well-formed types are also
+semantically valid. We postulate the existence of a \textit{unit type}
+\(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all tensor products \(A
+\otimes B\); we write \(\otimes\) to underscore the fact that types are allowed
+to be substructural, assigning every type a linearity \(\ms{lin}(A)\) satisfying
+the constraints in Figure \ref{fig:ssa-structural}. In particular, the rules
+\rle{unit-lin} and \rle{bool-lin} state that both the unit and boolean types
+must be intuitionistic; the rule \rle{pair-lin} says that the linearity of a
+pair is simply the intersection of the linearity of its components. The rules
+\rle{nil-lin} and \rle{cons-lin} together say that the linearity of a context is
+the intersection of the linearity of its variables, where the linearity of a
+variable is the intersection of the linearity of its type and its quantity
+annotation.
+
+Our grammar proper is split into three main syntactic categories:
+\begin{itemize}
+  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
+  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
+  is of type \(A\) with \textit{purity} \(p = 1\) if \textit{pure}, or \(p = 0\)
+  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). The rules
+  for this judgement are given in Figure \ref{fig:ssa-term-typing}; in
+  particular, we note that a impure terms cannot be \textit{nested}; that is,
+  all proper subterms of any term are pure.
+  \item \textit{Blocks} \(s, t\), which are typed via the judgement
+  \(\haslb{\ms{L}}{t}{\Gamma}\), which says that the block \(t\) is well-formed
+  with respect to the context \(\Gamma\) and branches only to labels in
+  \(\ms{L}\). The rules for this judgement are given in Figure
+  \ref{fig:ssa-block-typing}.
+  \item \textit{Label-sets} or \textit{control-flow graphs} \(L\), which are
+  typed via the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\)
+  defines a control-flow graph with inputs \(\ms{L}\) and outputs \(\ms{K}\)
+  (some of which may also be in \(\ms{L}\), indicating recursive control-flow).
+  The rules for this judgement are given in Figure \ref{fig:ssa-block-typing}.
+\end{itemize}
+Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
+type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). A variable
+\(x: A^q\) is called \textit{affine} if its type \(A\) is affine (i.e.
+\(\tlin{A}{\taff}\)) \textit{and} \(\taff \subseteq q\), in which case it can be
+freely ignored. Similarly, \(x\) is \textit{relevant} if its type \(A\) relevant
+(i.e. \(\tlin{A}{\trel}\)) \textit{and} \(\trel \subseteq q\), in which case it
+can be freely ignored. This is formalized in the structural judgement
+\(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
+subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special case
+\(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
+typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
+\rle{split-nil} provides a base case by saying that the empty context can be
+split into two empty contexts, while the rules \rle{split-left} and
+\rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively;
+both these rules also allow us to reduce the quantity \(q\) to a quantity \(r
+\subseteq q\), "forgetting" some of our capabilities. The rule \rle{split-dup}
+allows the duplication of a variable which has a relevant type \(A\)
+\textit{and} a relevant quantity \(q\), with the option to weaken the quantity
+to \textit{different} quantities \(r\) and \(s\) on the left-hand side and
+right-hand sides respectively. Finally, the rule \rle{split-aff} simply allows
+us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.
+
+\TODO{segue to label-contexts \(\ms{L}\)}
+
+The rule \rle{join-nil} is the analogue to \rle{split-nil} for \textit{label-contexts}, providing a base case in that the empty label weakens into itself.
+The rule \rle{join-cons} says that if \(\ms{L}\) weakens into \(\ms{K}\) and \(\Delta\) weakens into \(\Gamma\), then \(\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}\), allowing us to "forget" any extra variables dropped by \(\cwk{\Gamma}{\Delta}\).
+Finally, the rule \rle{join-zero} simply says that if \(\lwk{\ms{L}}{\ms{K}}\), then \(\ms{L}\) also weakens to any extension of \(\ms{K}\).
+
+\TODO{label-weakening is backwards}
 
@@
 
-We can now list out our primary typing judgements, which are given in Figure
-\ref{fig:ssa-judgements}. We have three main typing judgements:
-\(\hasty{\Gamma}{p}{x}{A}{q}\) is the typing judgement for expressions, which
-says that the expression \(a\) has type \(A\) in context \(\Gamma\). The
-annotation \(q\) means that the expression itself has quantity \(q \subseteq
-\tint\), while the annotation \(p\) means that the expression has purity \(p\)
-(1 for pure, 0 for impure). \(\haslb{\Gamma}{t}{\ms{L}}\) is the typing block
-for blocks, which says that, using variables in \(\Gamma\), the block \(t\)
-exits by jumping to labels in \(\ms{L}\). Finally,
-\(\lhaslb{\ms{L}}{L}{\ms{K}}\) says that the label-set \(L\) defines the labels
-\(\ms{L}\) in terms of the labels \(\ms{K}\) (blocks in \(L\) are allowed to
-recursively call into \(\ms{L}\) or exit to \(\ms{K}\)).
-
-To define these, we will need a few basic structural judgements.
-\(\csplits{\Gamma}{\Delta}{\Xi}\) means that the context \(\Gamma\) can be split
-into contexts \(\Delta\) and \(\Xi\). We use the notion of splitting, rather
-than weakening, a context as primitive to be able to deal with linearity, since
-not all variables can be dropped or duplicated. This can be used to define
-weakening \(\cwk{\Gamma}{\Delta}\) as simply "splitting \(\Gamma\) into
-\(\Delta\) and the empty context", since a split is allowed to drop affine
-variables. Finally, \(\lwk{\ms{L}}{\ms{K}}\) means that the label-set \(\ms{L}\)
-weakens to the label-set \(\ms{K}\), i.e. that \(\ms{K}\) has more labels, but
-for shared labels, needs less variables to be live (but the same non-affine
-variables!) The typing rules for these judgements are given in Figure
-\ref{fig:ssa-structural}.
-
 \begin{figure}

### 2023-11-12 pldi24/paper.tex
Tap

@@
 
-\TODO{segue to label-contexts \(\ms{L}\)}
-
-The rule \rle{join-nil} is the analogue to \rle{split-nil} for \textit{label-contexts}, providing a base case in that the empty label weakens into itself.
-The rule \rle{join-cons} says that if \(\ms{L}\) weakens into \(\ms{K}\) and \(\Delta\) weakens into \(\Gamma\), then \(\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}\), allowing us to "forget" any extra variables dropped by \(\cwk{\Gamma}{\Delta}\).
-Finally, the rule \rle{join-zero} simply says that if \(\lwk{\ms{L}}{\ms{K}}\), then \(\ms{L}\) also weakens to any extension of \(\ms{K}\).
-
-\TODO{label-weakening is backwards}
+\textit{Label-contexts} \(\ms{L}\) are defined to consist of a list of
+\textit{labels} \(\lbl{\ell}\) annotated with a context \(\Gamma\) and an
+argument type \(A\). The judgement \(\lwk{\ms{L}}{\ms{K}}\) states that
+\(\ms{L}\) is a subcontext of \(\ms{K}\), i.e., that every program which
+"branches to \(\ms{L}\)" (i.e., finishes by either nontermination or branching
+to one of the labels \(\lbl{\ell}\) with the appropriate preserved variables
+\(\Gamma\) and argument \(A\)) can also be viewed as branching to \(\ms{K}\).
+The conditions for this are formalized in Figure \ref{fig:ssa-structural} as
+follows: the rule \rle{join-nil} provides a base case stating that the empty
+label weakens into itself. The rule \rle{join-cons} says that if \(\ms{L}\)
+weakens into \(\ms{K}\) and \(\Delta\) weakens into \(\Gamma\), then
+\(\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}\),
+allowing us to "forget" any extra variables dropped by \(\cwk{\Gamma}{\Delta}\).
+Finally, the rule \rle{join-zero} simply says that if \(\lwk{\ms{L}}{\ms{K}}\),
+then \(\ms{L}\) also weakens to any extension of \(\ms{K}\). Note that this
+judgement is "backwards," in that the "larger" context is on the \textit{right};
+this corresponds to the placement of the label context on the right-hand side of
+the typing judgement for blocks, and is by analogy to our semantics for blocks
+and label-sets in Section \ref{sec:semantics}.
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
   otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). The rules
-  for this judgement are given in Figure \ref{fig:ssa-term-typing}; in
-  particular, we note that a impure terms cannot be \textit{nested}; that is,
-  all proper subterms of any term are pure.
+  for this judgement are given in Figure \ref{fig:ssa-term-typing}.
   \item \textit{Blocks} \(s, t\), which are typed via the judgement
@@
 
+We can now get to the typing judgement for expressions in Figure
+\ref{fig:ssa-term-typing}. The rule \rle{var} is essentially the standard rule
+for variables, with the requirement that, other than \(x\), all variables in
+\(\Gamma\) are affine. Note that the quantity \(q\) of the \textit{expression}
+\(x\) is must satisfy \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this
+reflects that we're allowed to \textit{forget} capabilities, but not
+\textit{add} them. The \rle{app} rule is also mostly standard: given an
+instruction \(f\) which takes in an argument \(A\) and has a return type of
+\(B\) and a \textit{pure} expression \(\hasty{\Gamma}{1}{a}{A}{q}\), we conclude
+that \(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that \(f\) itself
+have purity \(p\) and quantity \(q\), i.e. that \(f \in \mc{I}_p^q(A, B)\). The
+\rle{pair} rule says that we can type a pair \((a, b)\) in \(\Gamma\) if
+\(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\) and \(\Xi\) types
+\(b\); the quantity of a pair is simply the intersection of the quantity of its
+components (this is expressed in the rule by allowing any quantity which types
+\textit{both} components), and both components must be pure. Finally, we have
+the rules \rle{let} and \rle{let2}, corresponding to let-bindings and
+destructuring let-bindings respectively. In both cases, we split the context
+\(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is used to type a pure
+subterm \(a\), which is bound to variables \(x\) or destructured to variables
+\(x, y\) respectively, in a potentially impure term \(e\) which is otherwise
+typed in \(\Delta\).
+
 \begin{figure}
@@
 \end{figure}
-
-\TODO{some text about each typing rule}
 

### 2023-11-13 pldi24/paper.tex
Blurbs about typing rules

@@
 
+The rules for block typing are given in Figure \ref{fig:ssa-block-typing}.
+\rle{br} says that, given a pure term \(a: A\) which is \textit{both} affine and
+relevant (i.e. intuitionistic), we can type a block \(\lbrb{\ell}{a}\) which
+branches to \(\ell\) with argument \(a\), where \(\ms{L}\) contains the context
+\(\ell\) with an argument type of \(A\) and an appropriate set of unused
+variables. The rule \rle{ite} typechecks if-then-else statements by stating
+that, if \(\Gamma\) splits into \(\Delta\) and \(\Xi\), where \(\Xi\) typechecks
+a pure, intuitionistic discriminant \(e: \mb{2}\) and \(\Delta\) typechecks both
+\(s, t\) to branch into \(\ms{L}\), then the whole if-statement
+\(\ite{e}{s}{t}\) also typechecks as branching into \(\ms{L}\) in constext
+\(\Gamma\). \rle{let-blk} and \rle{let2-blk} are quite similar to their
+expression analogues, except that the bound expression \(e\) is allowed to be
+impure (otherwise, programs could only make use of pure functions!), and the
+remainder of the context plus the bound variable is used to type a block rather
+than a term. Finally, probably the most interesting rule is \rle{where}, which
+says that, given a block \(t\) which branches to \(\ms{L}\) in context
+\(\Gamma\), and labels \(L\) mapping \(\ms{L}\) to \(\ms{K}\), \(\ewhere{t}{L}\)
+branches to \(\ms{K}\). Here, \(L\) mapping \(\ms{L}\) to \(\ms{K}\) is defined
+using the typing rules \rle{nil-br}, which says that an empty list of blocks
+leaves the target label-set unchanged, and \rle{cons-br}, which says that if
+\(L\) maps \(\ms{L}\) to either \(\ms{K}\) or the label
+\(\lbl{\ell}[\Gamma](A)\), and \(t\) branches to \(\ms{L}\) in the context
+\(\Gamma, \thyp{x}{A}{}\), then \(L\) maps \(\ms{L}\) to \(\ms{K}\), having
+(potentially recursively) defined \(\lbl{\ell}\).
+
 \begin{figure}
@@
 \end{figure}
-
-\TODO{some text about each block typing rule}
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
 
-\TODO{fix this text}
-
 We will begin by giving a useful foundation for a wide class of \isotopessa models: the "trace models" over \(\mb{Set}\), which accumulate a (potentially infinite) nondeterministic trace of monoidal "events".
@@
 
+\begin{definition}[Relevant, Affine Monad]
+  A monad \(\mb{T}\) on \(\ms{Set}\) is \textbf{relevant} if its Kleisli category satisfies \(f;\ms{split} = \ms{split};f \ltimes f = \ms{split};f \rtimes f\), where \(\ms{split} = \upg{\lambda a.
+(a, a)}{}\).
+\(\mb{T}\) is \textbf{affine} if its Kleisli category satisfies \(f;\ms{drop} = \ms{drop}\), where \(\ms{drop} = \upg{\lambda a.()}{}\).
+A monad is \textbf{intuitionistic} if it is both relevant and affine.
+\end{definition}
+
 \begin{definition}[Submonad] 
@@
 We say that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and, for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where \(f^\dagger\) is computed as in \(\mb{T}\).
+Note that \(\mb{S}\) is also an (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\) from \(\mb{T}\).
 \end{definition}
 Some important examples of submonads include:
-\begin{itemize}
+\begin{enumerate}
   \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\); similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
   \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise intersections \(\bigwedge_i\mb{S}_i = \lambda A.
-\bigcap_i \mb{S}_i\;A\) are submonads, forming a complete meet-semilattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete lattice in the standard manner
-  \item \todo{commuting absorbtive submonad of trace monad: example of...}
-  \item \todo{central submonad. Go cite \cite{carette-central-2022}}
-  \item \todo{deterministic, idempotent trace monad: relevant submonad}
-  \item \todo{noneffectful trace monad: affine submonad}
-  \item \todo{cap and centralize, cap and purify}
-\end{itemize}
-\TODO{a submonad induces a wide subcategory}
+\bigcap_i \mb{S}_i\;A\) are submonads, forming a complete meet-semilattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete lattice in the standard manner.
+  \item In particular, the intersection of a relevant/affine/commutative submonad and another submonad is relevant/affine/commutative respectively.
+Since every commutative submonad induces a central subcategory of the Kleisli category of the parent monad, it follows that a triple of a relevant, affine, and commutative submonad make the Kleisli category of a monad \(\mb{T}\) into a substructural effectful category with all objects both affine and relevant.
+  \item If \(C \subseteq Z(M)\) is submonoid of \(M\), then \(A \mapsto \mc{P}^+(A \times C), \mc{P}^+(A \times C \times I) \subseteq \ms{Traces}\;\Theta\;A\) are submonads
+  \item In particular, if \(M_C \subseteq Z(M)\) is a central submonoid of \(M\) such that \(\forall m \in M_C, \forall i \in I, m \cdot i = i\), then \(A \mapsto \mc{P}^+(A \times M_C) \subseteq \ms{Traces}\;\Theta\;A\) is a submonad whose Kleisli category is a central subcategory of the Kleisli category of \(\ms{Traces}\;\Theta\;A\).
+Note \(M_C = \{\ms{id}\}\) is always such a submonoid.
+  \label{item:central-submonad}
+  \item Similarly, if \(M_R \subseteq Z(M)\) is a submonoid of \(M\) such that \(\forall m \in M_R, m \cdot m = m\), then \(A \mapsto \{S \in \mc{P}^+(A \times M_R + I) \mid |S| = 1\} \subseteq \ms{Traces}\;\Theta\;A\) is a \textit{relevant} submonad.
+Note \(M_R = \{\ms{id}\}\) is always such a submonoid.
+  \label{item:relevant-submonad}
+  \item Finally, \(A \mapsto \mc{P}^+(A) \subseteq \ms{Traces}\;\Theta\;A\) is a \textit{relevant} submonad.
+  \label{item:affine-submonad}
+\end{enumerate}
+In particular, \ref{item:central-submonad}, \ref{item:relevant-submonad}, \ref{item:affine-submonad} equip the Kleisli category of \(\ms{Traces}\;\Theta\;A\) with the structure of a substructural effectful category given choices of \(M_C\), \(M_R\).
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
 
-\TODO{in particular, these induce a substructural effectful category...}
-
-\subsection{Heaps and Separation Logic}
+\subsection{Heaps, Printing, and Separation Logic}
+
+As a concrete example, we can now introduce the \textbf{printing monad}, which is simply defined as \(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where \(\Sigma: (\ms{Bytes}^*)^\omega \to \ms{Bytes}^{\leq \omega}\) denotes concatenation of lists of bytestrings.
+
+\TODO{substructural effectful category}
+
+\TODO{heap via \(\ms{StateT}\), lifting of central/relevant/affine submonads}
+
+\TODO{what is a \(\ms{Heap}\)}
+
+\TODO{segue to separation logic}
 
 Building on the ideas in \cite{mellies-ftrs}, we will show how a simple separation logic based on \cite{reynolds-separation-2002} slightly adapted to the categorical setting gives us a very rich \isotopessa model supporting a basic notion of refinement types, while at the same time giving us a nontrivial notion of nonlinear types.
-
-We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{Bytes}^* + \ms{Bytes}^{\leq \omega})\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the monoid of lists of bytestrings described in the previous section.
-
-\TODO{say something better than "bytestrings"}
-
-\TODO{what is a \(\ms{Heap}\)}
-
-\TODO{Use name \ms{Traces}}
 
@@
 \TODO{Coproduct, Elgot structure inherited from base category}
-\TODO{sliding w/ properly restricted equality}
-\TODO{equality vs \cite{mellies-ftrs}}
-\TODO{potential solutions}
 
 \TODO{so, it's a substructural effectful category mmkay}
-\TODO{central = no printing}
+\TODO{central = no printing, no heap}
 \TODO{linear objects iff non duplicable}
@@
 \TODO{affine morphisms always (if no heap effects?)}
+
+\TODO{equality vs \cite{mellies-ftrs}}
+\TODO{sliding w/ properly restricted equality}
+\TODO{\(\implies\) central = no printing, heap allowed}
 

### 2023-11-13 pldi24/paper.tex
Printing monad

--- 2023-11-13 pldi24/paper.tex
@@
 
+\begin{definition}[Relevant, Affine Monad]
+  A monad \(\mb{T}\) on \(\ms{Set}\) is \textbf{relevant} if its Kleisli
+  category satisfies \(f;\ms{split} = \ms{split};f \ltimes f = \ms{split};f
+  \rtimes f\), where \(\ms{split} = \upg{\lambda a. (a, a)}{}\). \(\mb{T}\) is
+  \textbf{affine} if its Kleisli category satisfies \(f;\ms{drop} = \ms{drop}\),
+  where \(\ms{drop} = \upg{\lambda a.()}{}\). A monad is \textbf{intuitionistic}
+  if it is both relevant and affine.
+\end{definition}
+
 \begin{definition}[Submonad] 
-  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\).
-We say that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and, for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where \(f^\dagger\) is computed as in \(\mb{T}\).
+  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a
+  collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq
+  \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in
+  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say
+  that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and,
+  for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where
+  \(f^\dagger\) is computed as in \(\mb{T}\). Note that \(\mb{S}\) is also an
+  (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
+  from \(\mb{T}\).
 \end{definition}
 Some important examples of submonads include:
-\begin{itemize}
-  \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\); similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
-  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise intersections \(\bigwedge_i\mb{S}_i = \lambda A.
-\bigcap_i \mb{S}_i\;A\) are submonads, forming a complete meet-semilattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete lattice in the standard manner
-  \item \todo{commuting absorbtive submonad of trace monad: example of...}
-  \item \todo{central submonad. Go cite \cite{carette-central-2022}}
-  \item \todo{deterministic, idempotent trace monad: relevant submonad}
-  \item \todo{noneffectful trace monad: affine submonad}
-  \item \todo{cap and centralize, cap and purify}
-\end{itemize}
-\TODO{a submonad induces a wide subcategory}
-
-\TODO{in particular, these induce a substructural effectful category...}
-
-\subsection{Heaps and Separation Logic}
+\begin{enumerate}
+  \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad
+  \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\);
+  similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
+  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
+  intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
+  submonads, forming a complete meet-semilattice with bottom element
+  \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete
+  lattice in the standard manner.
+  \item In particular, the intersection of a relevant/affine/commutative
+  submonad and another submonad is relevant/affine/commutative respectively.
+  Since every commutative submonad induces a central subcategory of the Kleisli
+  category of the parent monad, it follows that a triple of a relevant, affine,
+  and commutative submonad make the Kleisli category of a monad \(\mb{T}\) into
+  a substructural effectful category with all objects both affine and relevant.
+  \item If \(C \subseteq Z(M)\) is submonoid of \(M\), then \(A \mapsto
+  \mc{P}^+(A \times C + \mb{0}), \mc{P}^+(A \times C \times I) \subseteq
+  \ms{Traces}\;\Theta\;A\) are submonads
+  \item In particular, if \(M_C \subseteq Z(M)\) is a central submonoid of \(M\)
+  such that \(\forall m \in M_C, \forall i \in I, m \cdot i = i\), then \(A
+  \mapsto \mc{P}^+(A \times M_C + \mb{0}) \subseteq \ms{Traces}\;\Theta\;A\) is
+  a submonad whose Kleisli category is a central subcategory of the Kleisli
+  category of \(\ms{Traces}\;\Theta\;A\). Note \(M_C = \{1\}\) is always such a
+  submonoid.
+  \label{item:central-submonad}
+  \item Similarly, if \(M_R \subseteq M_C\) is a submonoid such that \(\forall m
+  \in M_R, m \cdot m = m\), then \(A \mapsto \{S \in \mc{P}^+(A \times M_R + I)
+  \mid |S| = 1\} \subseteq \ms{Traces}\;\Theta\;A\) is a \textit{relevant}
+  submonad. Note \(M_R = \{1\}\) is always such a submonoid.
+  \label{item:relevant-submonad}
+  \item Finally, \(A \mapsto \mc{P}^+(A \times \{1\} + \mb{0}) \subseteq
+  \ms{Traces}\;\Theta\;A\) is an \textit{affine} submonad.
+  \label{item:affine-submonad}
+\end{enumerate}
+In particular, \ref{item:central-submonad}, \ref{item:relevant-submonad},
+\ref{item:affine-submonad} equip the Kleisli category of
+\(\ms{Traces}\;\Theta\;A\) with the structure of a substructural effectful
+category given choices of \(M_C\), \(M_R\).
+
+\subsection{Heaps, Printing, and Separation Logic}
+
+As a concrete example, we can now introduce the \textbf{printing monad}, which
+is simply defined as \(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where
+\(\Sigma: (\ms{Bytes}^*)^\omega \to \ms{Bytes}^{\leq \omega}\) denotes
+concatenation of lists of bytestrings. This can be equipped with the structure
+of a substructural effectful category by choosing \(M_C = M_R = \{[]\}\), giving
+us an \isotopessa model supporting the instructions \(\ms{print} \in
+\mc{I}^\varnothing_0(\ms{Bytes}, \mb{1})\) and \(\ms{rand} \in
+\mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard semantics
+\(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), [b])\}\), \(\dnt{\ms{rand}} =
+\lambda (). \{\iota_0 (a, p[]) \mid a \in A\}\).
+
+We can make things a little more interesting by adding a heap to the mix, defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where \(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function.
+The Kleisli category of this monad can be equipped with the structure of a substructural effectful category by simply taking the affine/relevant morphisms to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\); as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we therefore have an \isotopessa model supporting the instructions \(\ms{get}: \mc{I}^\varnothing_0(\nats, \nats)\) and \(\ms{set}: \mc{I}^\varnothing_0(\nats \times \nats, ())\).
+We can assign semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} = \lambda (p, v)\;h.
+\{\iota_0((), [p \mapsto v]h, [])\}\).
+\(\ms{get}\) is a bit more tricky, since it is unclear what to do when we try to access uninitialized memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}} = \lambda p\;h.
+\{\iota_0(v, []) \mid v = h\;p \lor p \notin h\}\); for now, this will do.
+
+\TODO{segue to separation logic}
 
 Building on the ideas in \cite{mellies-ftrs}, we will show how a simple separation logic based on \cite{reynolds-separation-2002} slightly adapted to the categorical setting gives us a very rich \isotopessa model supporting a basic notion of refinement types, while at the same time giving us a nontrivial notion of nonlinear types.
-
-We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{Bytes}^* + \ms{Bytes}^{\leq \omega})\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the monoid of lists of bytestrings described in the previous section.
-
-\TODO{say something better than "bytestrings"}
-
-\TODO{what is a \(\ms{Heap}\)}
-
-\TODO{Use name \ms{Traces}}
 

### 2023-11-13 pldi24/paper.tex
Cut semisep blurb

@@
 
-We can further use this example to get a substructural \textit{monoidal} category by considering a primitive form of separation logic: let's introduce a grammar of predicates \(\varphi, \psi ::= \top, \ms{S}, \bot\) for programs which cannot access state, can access state, and are invalid, respectively.
+% We can further use this example to get a substructural \textit{monoidal} category by considering a primitive form of separation logic: let's introduce a grammar of predicates \(\varphi, \psi ::= \top, \ms{S}, \bot\) for programs which cannot access state, can access state, and are invalid, respectively.
 The separating conjunction operator on these predicates is then given by
-\begin{equation}
-  \top * \varphi = \varphi * \top = \varphi
-  \qquad
-  \bot * \varphi = \varphi * \bot = \bot
-  \qquad
-  \ms{S} * \ms{S} = \bot
-\end{equation}
-We'll then define a category \(\ms{SRel}\) with objects of the form \((A, \varphi)\) and morphisms
-\begin{equation}
-  \begin{gathered}
-    \boxed{
-      \ms{SRel}((A, \varphi), (B, \psi)) 
-      \subseteq A \to \ms{S} \to \mc{P}(A \times S) 
-    }
-    \\
-    \begin{aligned}
-      \ms{SRel}((A, \top), (B, \top)) &= \{f \in A \to S \to \mc{P}(A \times S) \mid \exists g. f = \lambda a, s. g(a) \times \{s\}\} \\
-      \ms{SRel}((A, \ms{S}), (B, \ms{S})) &= A \to S \to \mc{P}(A \times S) \\
-      \ms{SRel}((A, \varphi), (B, \psi)) &= \{\lambda a, s. \varnothing\} \quad \text{otherwise}
-    \end{aligned}
-  \end{gathered}
-\end{equation}
-Note that it is precisely the \textit{central} morphisms \(\ms{C}_0(A, B)\) which lie in \(\ms{SRel}((A, \top), (B, \top))\).
+% \begin{equation}
+%   \top * \varphi = \varphi * \top = \varphi
+%   \qquad
+%   \bot * \varphi = \varphi * \bot = \bot
+%   \qquad
+%   \ms{S} * \ms{S} = \bot
+% \end{equation}
+% We'll then define a category \(\ms{SRel}\) with objects of the form \((A, \varphi)\) and morphisms
+% \begin{equation}
+%   \begin{gathered}
+%     \boxed{
+%       \ms{SRel}((A, \varphi), (B, \psi)) 
+%       \subseteq A \to \ms{S} \to \mc{P}(A \times S) 
+%     }
+%     \\
+%     \begin{aligned}
+%       \ms{SRel}((A, \top), (B, \top)) &= \{f \in A \to S \to \mc{P}(A \times S) \mid \exists g. f = \lambda a, s. g(a) \times \{s\}\} \\
+%       \ms{SRel}((A, \ms{S}), (B, \ms{S})) &= A \to S \to \mc{P}(A \times S) \\
+%       \ms{SRel}((A, \varphi), (B, \psi)) &= \{\lambda a, s. \varnothing\} \quad \text{otherwise}
+%     \end{aligned}
+%   \end{gathered}
+% \end{equation}
+% Note that it is precisely the \textit{central} morphisms \(\ms{C}_0(A, B)\) which lie in \(\ms{SRel}((A, \top), (B, \top))\).
 We can then define the tensor product on objects as \((A, \varphi) \otimes (B, \psi) = (A \otimes B, \varphi * \psi)\), and, for \(f \in \mathsf{SRel}((A, \varphi), (B, \psi))\), and tensor product functors
-\begin{equation}
-  \begin{aligned}
-  f \otimes (C, \phi) &= f \otimes C,
-  & (C, \phi) \otimes f &= C \otimes f
-  && \text{if}\; \varphi * \phi \neq \bot, \psi * \phi \neq \bot \\
-  f \otimes (C, \phi) &= \lambda a, s. \varnothing,
-  & (C, \phi) \otimes f &=  \lambda a, s. \varnothing
-  && \text{otherwise}
-  \end{aligned} 
-\end{equation}
-Lifting associators, unitors, and symmetries from \(\mc{C}_0\) in the obvious manner (setting them to \(\lambda a, s.
+% \begin{equation}
+%   \begin{aligned}
+%   f \otimes (C, \phi) &= f \otimes C,
+%   & (C, \phi) \otimes f &= C \otimes f
+%   && \text{if}\; \varphi * \phi \neq \bot, \psi * \phi \neq \bot \\
+%   f \otimes (C, \phi) &= \lambda a, s. \varnothing,
+%   & (C, \phi) \otimes f &=  \lambda a, s. \varnothing
+%   && \text{otherwise}
+%   \end{aligned} 
+% \end{equation}
+% Lifting associators, unitors, and symmetries from \(\mc{C}_0\) in the obvious manner (setting them to \(\lambda a, s.
 \varnothing\) where necessary), we can verify that \(\ms{SRel}\) is a \textit{premonoidal} category.
@@
 Hence, to get a \textit{substructural} monoidal category, we simply need to define:
-\begin{equation}
-  \begin{gathered}
-  \ms{Aff}(\ms{SRel}) = \ms{Rel}(\ms{SRel}) = \{(A, \top) | A \in \ms{Set}\} \neq |\ms{SRel}|
-  \\
-  \begin{aligned}
-    \ms{SRel}^\varnothing((A, \phi), (B, \psi))
-    &= \ms{SRel}((A, \phi), (B, \psi)) \\
-    \ms{SRel}^{\{\ms{a}\}}((A, \phi), (B, \psi)) 
-    &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \geq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
-    \ms{SRel}^{\{\ms{r}\}}((A, \phi), (B, \psi)) 
-    &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \leq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
-    \implies \ms{SRel}^{\{\ms{a}, \ms{r}\}}((A, \phi), (B, \psi)) 
-    &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| = 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
-  \end{aligned}
-  \end{gathered}
-\end{equation}
-We can verify that the associators, unitors, and symmetry lie in \(\ms{SRel}^{\{\ms{a}, \ms{r}\}}\) and that all other desired axioms are satisfied.
-
-\TODO{relate separation logic trick to \cite{promonad}, \cite{linear-state-usage}, \cite{mellies-ftrs}}
-
-\TODO{clean up, and generalize to arbitrary premonoidal categories (w/ initial object? how to deal w/ coproducts?)}
+% \begin{equation}
+%   \begin{gathered}
+%   \ms{Aff}(\ms{SRel}) = \ms{Rel}(\ms{SRel}) = \{(A, \top) | A \in \ms{Set}\} \neq |\ms{SRel}|
+%   \\
+%   \begin{aligned}
+%     \ms{SRel}^\varnothing((A, \phi), (B, \psi))
+%     &= \ms{SRel}((A, \phi), (B, \psi)) \\
+%     \ms{SRel}^{\{\ms{a}\}}((A, \phi), (B, \psi)) 
+%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \geq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
+%     \ms{SRel}^{\{\ms{r}\}}((A, \phi), (B, \psi)) 
+%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \leq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
+%     \implies \ms{SRel}^{\{\ms{a}, \ms{r}\}}((A, \phi), (B, \psi)) 
+%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| = 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
+%   \end{aligned}
+%   \end{gathered}
+% \end{equation}
+% We can verify that the associators, unitors, and symmetry lie in \(\ms{SRel}^{\{\ms{a}, \ms{r}\}}\) and that all other desired axioms are satisfied.
+
+% \TODO{relate separation logic trick to \cite{promonad}, \cite{linear-state-usage}, \cite{mellies-ftrs}}
+
+% \TODO{clean up, and generalize to arbitrary premonoidal categories (w/ initial object? how to deal w/ coproducts?)}
 
@@
 
+\TODO{ssec:separation}
+
 As a concrete example, we can now introduce the \textbf{printing monad}, which

### 2023-11-13 pldi24/paper.tex
Basic separation model

@@
 We can make things a little more interesting by adding a heap to the mix, defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where \(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function.
-The Kleisli category of this monad can be equipped with the structure of a substructural effectful category by simply taking the affine/relevant morphisms to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\); as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we therefore have an \isotopessa model supporting the instructions \(\ms{get}: \mc{I}^\varnothing_0(\nats, \nats)\) and \(\ms{set}: \mc{I}^\varnothing_0(\nats \times \nats, ())\).
+The Kleisli category of this monad can be equipped with the structure of a substructural effectful category by simply taking the affine/relevant morphisms to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\); as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we therefore have an \isotopessa model supporting the instructions \(\ms{set}: \mc{I}^\varnothing_0(\nats \times \nats, \mb{1})\), \(\ms{get}: \mc{I}^\varnothing_0(\nats, \nats)\), \(\ms{alloc}: \mc{I}^\varnothing_0(\nats, \nats)\), and \(\ms{free}: \mc{I}^\varnothing_0(\nats, \mb{1})\).
 We can assign semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} = \lambda (p, v)\;h.
@@
 \(\ms{get}\) is a bit more tricky, since it is unclear what to do when we try to access uninitialized memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}} = \lambda p\;h.
-\{\iota_0(v, []) \mid v = h\;p \lor p \notin h\}\); for now, this will do.
+\{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\); for now, this will do.
+Finally, \(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', []) \mid h' = h \sqcup p \mapsto v\}\) simply fills a random empty heap cell with the provided value, returning a pointer to the cell, while \(\dnt{\ms{free}} = \lambda p\;h.
+\{\iota_0\;((), h \setminus p, [])\}\)
 
@@
 
-\TODO{by the above, this is an \isotopessa model w/ nice properties; moving on!}
-
-Given a set \(A\), we can define a predicate \(\varphi\) on \(A\), \(\mb{Heap}\) pairs to have the following type:
-\TODO{this}
-\TODO{Objects \((A, \varphi)\), \((B, \psi)\)}
-\TODO{arrows}
-\TODO{note on magic wand}
-\TODO{Tensor product and separating conjunction}
-\TODO{Coproduct, Elgot structure inherited from base category}
-
-\TODO{so, it's a substructural effectful category mmkay}
-\TODO{central = no printing, no heap}
-\TODO{linear objects iff non duplicable}
-\TODO{relevant morphisms iff deterministic}
-\TODO{affine morphisms always (if no heap effects?)}
+Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of type \(A \to \ms{Heap} \to \ms{Prop}\). 
+Taking as objects pairs \((A, \Phi)\), we can define the category \(\ms{Sep}\) to have morphisms
+\begin{equation}
+  \ms{Sep}((A, \Phi), (B, \Psi)) = \{f: A \to \ms{Comp}\;B 
+    \mid \forall \phi: \ms{Heap} \to \ms{Prop}, \ms{hoare}\;f\;(\Phi \ast \phi)\;(\Psi \ast \phi)\}
+\end{equation}
+where
+\begin{equation}
+  \ms{hoare}\;f\;\Phi\;\Psi = \forall a\;h. \Phi\;a\;h \implies \forall \iota_0\;(b, h') \in f\;a\;h. \Psi\;b\;h'
+\end{equation}
+is the usual partial correctness triple definition and the separating conjunction of heap-predicates is defined as usual to be 
+\begin{equation}
+  \begin{gathered}
+  \varphi \ast \psi = \lambda h. \exists h_1, h_2. h = h_1 \sqcup h_2 \land \varphi h_1 \land \psi h_2 
+  \qquad
+  \varphi \ast \Phi = \lambda a. \varphi \ast \Phi\;a
+  \qquad
+  \Phi \ast \varphi = \lambda a. \Phi\;a \ast \varphi \\
+  \Phi \ast \Psi = \lambda (a, b). \Phi\;a \ast \Psi\;b: A \times B \to \ms{Heap} \to \ms{Prop}
+  \end{gathered}
+\end{equation}
+We can naturally define a tensor product on objects \((A, \Phi) \otimes (B, \Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking
+\begin{itemize}
+  \item Monoidal unit \((\mb{1}, \ms{emp})\), where \(\ms{emp}\) is the predicate which holds only for the empty heap
+  \item Associators and inverse associators given by \(\alpha_{(A, \Phi), (B, \Psi), (C, \Theta)} = \alpha_{A, B, C}\), which are valid since \(\ast\) is associative
+  \item Unitors and inverse unitors given by \(\lambda_{(A, \Phi)} = \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\), which are valid since \(\ms{emp}\) is the unit of \(\ast\)
+  \item \((C, \Theta) \otimes f = C \times f\) and \(f \otimes (C, \Theta) = f \times C\), which are valid definitions by the frame rule
+\end{itemize}
+equips \(\ms{Sep}\) with the structure of a premonoidal category. We can give \(\ms{Sep}\) the structure of a substructural effectful category by taking:
+\begin{itemize}
+  \item Affine objects to be precisely those for which \(\ms{drop} = \upg{\lambda a.
+()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), which for this logic corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend on the heap.
+  \item Relevant objects to be precisely those for which \(\ms{split} = \upg{\lambda a.
+(a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), again in this case those for which \(\Phi\) does not depend on the heap.
+  \item Central morphisms to be those which do not print \textit{or} modify the heap. We will consider how to relax this condition later.
+  \item Relevant and affine morphisms to be the lifts of relevant/affine morphisms in \(\ms{Print}\)
+\end{itemize}
+\(\ms{Sep}\) also has coproducts given by \((A, \Phi) + (B, \Psi) = (A + B, [\Phi, \Psi])\), which allow us to equip it with an Elgot structure by taking advantage of the fact that
+\begin{equation}
+  f \in \ms{Sep}((A, \Phi), (B + A, [\Psi, \Phi])) \implies f^\dagger \in \ms{Sep}((A, \Phi), (B, \Psi))
+\end{equation}
+This implies that \(\ms{Sep}\) is an \isotopessa model supporting the following interesting operations
+\begin{itemize}
+  \item \(\ms{coe}: \mc{I}^\tint_1((A, \Phi), (A, \Psi))\) where \(\forall a, \Phi\;a \implies \Psi\;a\), implemented by the identity morphism
+  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Bytes}, \ms{emp}), (\mb{1}, \ms{emp}))\)
+  \item \(\ms{rand}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for \(A\) nonempty
+  \item \(\ms{set}: \mc{I}^\varnothing_0((\nats \times \nats, \lambda (\ell, \cdot).
+\exists v.
+\ell \mapsto v), (\nats, \lambda \ell.
+\exists v.
+\ell \mapsto v))\), where \(\ms{set}\) is modified slightly to return the address at which the value passed in was stored to overcome a lack of dependency between the source and target objects.
+  \item \(\ms{get}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)
+  \item \(\ms{alloc}: \mc{I}^\varnothing_0((\nats, \ms{emp}), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)
+  \item \(\ms{free}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\mb{1}, \ms{emp}))\)
+\end{itemize}
 
@@
 \TODO{sliding w/ properly restricted equality}
-\TODO{\(\implies\) central = no printing, heap allowed}
+\TODO{\(\implies\) central = no printing, heap allowed: can make heap operations relevant/affine too!}
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
 
-\TODO{ssec:separation}
+\label{ssec:separation}
 
@@
 
-\TODO{segue to separation logic}
-
-Building on the ideas in \cite{mellies-ftrs}, we will show how a simple separation logic based on \cite{reynolds-separation-2002} slightly adapted to the categorical setting gives us a very rich \isotopessa model supporting a basic notion of refinement types, while at the same time giving us a nontrivial notion of nonlinear types.
+We can use this construction to, building on the ideas in \cite{mellies-ftrs}, show how a simple separation logic based on \cite{reynolds-separation-2002} slightly adapted to the categorical setting gives us a very rich \isotopessa model supporting a basic notion of refinement types, while at the same time giving us a nontrivial notion of nonlinear types.
 
@@
 \end{itemize}
-
-\TODO{equality vs \cite{mellies-ftrs}}
+We can clearly see that this is a refinement type system in the sence of \cite{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi) \mapsto A\) which is simply the identity on morphisms.
+Unfortunately, this means that our notion of equality is a little too stringent, since we need to consider equality on inputs which do not match our preconditions.
+
+\TODO{example, + example set}
+
+\TODO{restricted equality}
+
 \TODO{sliding w/ properly restricted equality}
+
 \TODO{\(\implies\) central = no printing, heap allowed: can make heap operations relevant/affine too!}

### 2023-11-13 pldi24/paper.tex
Tap

@@
 
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\TODO{pull down to appendix}
-
 \begin{figure}
@@
 
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\TODO{pull down to appendix?}
-
 \TODO{some text about substitution weakening}
@@
 
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\TODO{pull down to appendix}
-
 \TODO{some text about substitution}
@@
 \end{theorem}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\TODO{pull down to appendix}
 
@@
 
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\TODO{pull down to appendix? This is just a trivial induction.}
-
 \TODO{some text about coherence, implicit use above}
@@
 
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\TODO{pull down to appendix}
-
 \TODO{some text about semantic weakening}
@@
 \end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
 
@@
 
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-\TODO{float proof to appendix}
-
 \TODO{some text about semantic substitution}
@@
 \end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\TODO{pull down to appendix? just trivial induction}
 
@@
 
-We can make things a little more interesting by adding a heap to the mix, defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where \(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function.
-The Kleisli category of this monad can be equipped with the structure of a substructural effectful category by simply taking the affine/relevant morphisms to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\); as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we therefore have an \isotopessa model supporting the instructions \(\ms{set}: \mc{I}^\varnothing_0(\nats \times \nats, \mb{1})\), \(\ms{get}: \mc{I}^\varnothing_0(\nats, \nats)\), \(\ms{alloc}: \mc{I}^\varnothing_0(\nats, \nats)\), and \(\ms{free}: \mc{I}^\varnothing_0(\nats, \mb{1})\).
-We can assign semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} = \lambda (p, v)\;h.
-\{\iota_0((), [p \mapsto v]h, [])\}\).
-\(\ms{get}\) is a bit more tricky, since it is unclear what to do when we try to access uninitialized memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}} = \lambda p\;h.
-\{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\); for now, this will do.
-Finally, \(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', []) \mid h' = h \sqcup p \mapsto v\}\) simply fills a random empty heap cell with the provided value, returning a pointer to the cell, while \(\dnt{\ms{free}} = \lambda p\;h.
-\{\iota_0\;((), h \setminus p, [])\}\)
-
-We can use this construction to, building on the ideas in \cite{mellies-ftrs}, show how a simple separation logic based on \cite{reynolds-separation-2002} slightly adapted to the categorical setting gives us a very rich \isotopessa model supporting a basic notion of refinement types, while at the same time giving us a nontrivial notion of nonlinear types.
-
-Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of type \(A \to \ms{Heap} \to \ms{Prop}\). 
-Taking as objects pairs \((A, \Phi)\), we can define the category \(\ms{Sep}\) to have morphisms
+We can make things a little more interesting by adding a heap to the mix,
+defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where
+\(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function. The
+Kleisli category of this monad can be equipped with the structure of a
+substructural effectful category by simply taking the affine/relevant morphisms
+to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\);
+as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we
+therefore have an \isotopessa model supporting the instructions \(\ms{set}:
+\mc{I}^\varnothing_0(\nats \times \nats, \mb{1})\), \(\ms{get}:
+\mc{I}^\varnothing_0(\nats, \nats)\), \(\ms{alloc}: \mc{I}^\varnothing_0(\nats,
+\nats)\), and \(\ms{free}: \mc{I}^\varnothing_0(\nats, \mb{1})\). We can assign
+semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} =
+\lambda (p, v)\;h. \{\iota_0((), [p \mapsto v]h, [])\}\). \(\ms{get}\) is a bit
+more tricky, since it is unclear what to do when we try to access uninitialized
+memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}}
+= \lambda p\;h. \{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\); for now,
+this will do. Finally, \(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', [])
+\mid h' = h \sqcup p \mapsto v\}\) simply fills a random empty heap cell with
+the provided value, returning a pointer to the cell, while \(\dnt{\ms{free}} =
+\lambda p\;h. \{\iota_0\;((), h \setminus p, [])\}\)
+
+We can use this construction to, building on the ideas in \cite{mellies-ftrs},
+show how a simple separation logic based on \cite{reynolds-separation-2002}
+slightly adapted to the categorical setting gives us a very rich \isotopessa
+model supporting a basic notion of refinement types, while at the same time
+giving us a nontrivial notion of nonlinear types.
+
+Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of
+type \(A \to \ms{Heap} \to \ms{Prop}\). Taking as objects pairs \((A, \Phi)\),
+we can define the category \(\ms{Sep}\) to have morphisms
 \begin{equation}
@@
 \end{equation}
-is the usual partial correctness triple definition and the separating conjunction of heap-predicates is defined as usual to be 
+is the usual partial correctness triple definition and the separating
+conjunction of heap-predicates is defined as usual to be 
 \begin{equation}
@@
 \end{equation}
-We can naturally define a tensor product on objects \((A, \Phi) \otimes (B, \Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking
-\begin{itemize}
-  \item Monoidal unit \((\mb{1}, \ms{emp})\), where \(\ms{emp}\) is the predicate which holds only for the empty heap
-  \item Associators and inverse associators given by \(\alpha_{(A, \Phi), (B, \Psi), (C, \Theta)} = \alpha_{A, B, C}\), which are valid since \(\ast\) is associative
-  \item Unitors and inverse unitors given by \(\lambda_{(A, \Phi)} = \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\), which are valid since \(\ms{emp}\) is the unit of \(\ast\)
-  \item \((C, \Theta) \otimes f = C \times f\) and \(f \otimes (C, \Theta) = f \times C\), which are valid definitions by the frame rule
+We can naturally define a tensor product on objects \((A, \Phi) \otimes (B,
+\Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking
+\begin{itemize}
+  \item Monoidal unit \((\mb{1}, \ms{emp})\), where \(\ms{emp}\) is the
+  predicate which holds only for the empty heap
+  \item Associators and inverse associators given by \(\alpha_{(A, \Phi), (B,
+  \Psi), (C, \Theta)} = \alpha_{A, B, C}\), which are valid since \(\ast\) is
+  associative
+  \item Unitors and inverse unitors given by \(\lambda_{(A, \Phi)} =
+  \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\), which are valid since \(\ms{emp}\)
+  is the unit of \(\ast\)
+  \item \((C, \Theta) \otimes f = C \times f\) and \(f \otimes (C, \Theta) = f
+  \times C\), which are valid definitions by the frame rule
 \end{itemize}
@@
 \begin{itemize}
-  \item Affine objects to be precisely those for which \(\ms{drop} = \upg{\lambda a.
-()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), which for this logic corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend on the heap.
-  \item Relevant objects to be precisely those for which \(\ms{split} = \upg{\lambda a.
-(a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), again in this case those for which \(\Phi\) does not depend on the heap.
-  \item Central morphisms to be those which do not print \textit{or} modify the heap. We will consider how to relax this condition later.
+  \item Affine objects to be precisely those for which \(\ms{drop} =
+  \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), which for
+  this logic corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend
+  on the heap.
+  \item Relevant objects to be precisely those for which \(\ms{split} =
+  \upg{\lambda a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), again in
+  this case those for which \(\Phi\) does not depend on the heap.
+  \item Central morphisms to be those which do not print, diverge, \textit{or} modify the
+  heap. We will consider how to relax this condition later.
   \item Relevant and affine morphisms to be the lifts of relevant/affine morphisms in \(\ms{Print}\)
@@
 \begin{itemize}
-  \item \(\ms{coe}: \mc{I}^\tint_1((A, \Phi), (A, \Psi))\) where \(\forall a, \Phi\;a \implies \Psi\;a\), implemented by the identity morphism
-  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Bytes}, \ms{emp}), (\mb{1}, \ms{emp}))\)
+  \item \(\ms{coe}: \mc{I}^\tint_1((A, \Phi), (A, \Psi))\) where \(\forall a,
+  \Phi\;a \implies \Psi\;a\), implemented by the identity morphism
+  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Bytes}, \ms{emp}), (\mb{1},
+  \ms{emp}))\)
   \item \(\ms{rand}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for \(A\) nonempty
-  \item \(\ms{set}: \mc{I}^\varnothing_0((\nats \times \nats, \lambda (\ell, \cdot).
-\exists v.
-\ell \mapsto v), (\nats, \lambda \ell.
-\exists v.
-\ell \mapsto v))\), where \(\ms{set}\) is modified slightly to return the address at which the value passed in was stored to overcome a lack of dependency between the source and target objects.
+  \item \(\ms{set}: \mc{I}^\varnothing_0((\nats \times \nats, \lambda (\ell,
+  \cdot). \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell
+  \mapsto v))\), where \(\ms{set}\) is modified slightly to return the address
+  at which the value passed in was stored to overcome a lack of dependency
+  between the source and target objects.
   \item \(\ms{get}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)
-  \item \(\ms{alloc}: \mc{I}^\varnothing_0((\nats, \ms{emp}), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)
+  \item \(\ms{alloc}: \mc{I}^\varnothing_0((\nats, \ms{emp}), (\nats, \lambda
+  \ell. \exists v. \ell \mapsto v))\)
   \item \(\ms{free}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\mb{1}, \ms{emp}))\)
 \end{itemize}
-We can clearly see that this is a refinement type system in the sence of \cite{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi) \mapsto A\) which is simply the identity on morphisms.
-Unfortunately, this means that our notion of equality is a little too stringent, since we need to consider equality on inputs which do not match our preconditions.
-
-\TODO{example, + example set}
-
-\TODO{restricted equality}
-
-\TODO{sliding w/ properly restricted equality}
-
-\TODO{\(\implies\) central = no printing, heap allowed: can make heap operations relevant/affine too!}
+We can clearly see that this is a refinement type system in the sence of
+\cite{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi)
+\mapsto A\) which is simply the identity on morphisms. Unfortunately, this means
+that our notion of equality is a little too stringent, since we need to consider
+equality on inputs which do not match our preconditions. Thankfully, we can fix
+this problem quite easily by quotienting morphisms over the equivalence relation
+\begin{equation}
+  \forall f, g: \ms{Sep}((A, \Phi), (B, \Psi)).
+  f \simeq g \iff \forall a, h. \Phi\;a\;h \implies f\;a\;h = g\;a\;h
+\end{equation}
+yielding the new category \(\ms{SComp}\). In particular, this allows us to widen
+our definition of central morphisms to include all morphisms which do not print
+or diverge, including in particular \(\ms{set}, \ms{get}, \ms{alloc}\) and
+\(\ms{free}\), which we can in fact mark relevant and affine as well, allowing
+us, for example, to make use of intermediate allocations in an otherwise pure
+computation.
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
 
-\TODO{text, \cite{sparky}}
-
+Another interesting use of the trace monad to build a sophisticated \isotopessa model is to build a model of TSO-style weak memory based on \citet{sparky}. We begin with the following definitions:
 \begin{definition}[Pomset] 
-  A \textbf{pomset} over a set of actions \(\mc{A}\) with a distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\)
-  \TODO{finish, clean}
+  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
+  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
+  partially-ordered \textbf{carrier set} \(P\) such that every \(p \in P\) has
+  finitely many predecessors equipped with a mapping \(\alpha: P \to \mc{A}\)
+  quotiented under the equivalence relation \(\alpha: P \to \mc{A} \simeq
+  \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\) from \(P\)
+  and \(Q\), there exists a partial order isomorphism between them which
+  respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite} if its
+  carrier set is.
 \end{definition}
-
-\TODO{Form a monoid, so we can define...}
-
-Let \(\ms{Pom}\) denote the monoid of pomsets...
-
-\TODO{stream action... trace monad \(\ms{Traces}\;\ms{Pom}\;\ms{Pom}\)...}
-
-\TODO{The SC Monad, which, via composition w/ the state transformer w/ a buffer as state, lets us}
-
-\TODO{Build The Weak Memory Monad, which is too powerful, so we do...}
-
-\begin{definition}[Monadic Predicate] \
-  \TODO{this and naming}
-\end{definition}
-
-\TODO{Proper Weak Memory Monad, preserves the Elgot structure (yey)}
+(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\), which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to \mc{A}\), where \(P + Q\) is given the lexicographic ordering and \(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many predecessors.
+They also form a monoid under parallel composition \(\alpha || \beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\) where \(P + Q\) is given the standard partial ordering (with elements of \(P\) and \(Q\) incomparable).
+
+Given a partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n \in N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by the function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\), where the dependent product \(\Sigma_nP_n\) is given the lexicographic order.
+In particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega \to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on finite pomsets, which allows us to define the \textbf{sequential consistency monad} \(\ms{SC}\;A = \ms{Traces}\;\Sigma\;A\).
+
+\TODO{how to do reads}
+
+\TODO{how to do writes}
+
+\TODO{weak memory monad}
+
+\TODO{new read/write}
+
+\TODO{fence}
+
+\TODO{submonad rune}
+
+\TODO{Buffer continuity}
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
 
-Another interesting use of the trace monad to build a sophisticated \isotopessa model is to build a model of TSO-style weak memory based on \citet{sparky}. We begin with the following definitions:
+Another interesting use of the trace monad to build a sophisticated \isotopessa model is to build a model of TSO-style weak memory based on \citet{sparky}.
+For simplicity, we will consider an infinite set of named locations \(x, y, z \in \ms{Loc}\) which are subject to concurrent modification by all processors via TSO reads, writes, and fences.
+
+We begin with the following definitions:
 \begin{definition}[Pomset] 
@@
 Given a partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n \in N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by the function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\), where the dependent product \(\Sigma_nP_n\) is given the lexicographic order.
-In particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega \to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on finite pomsets, which allows us to define the \textbf{sequential consistency monad} \(\ms{SC}\;A = \ms{Traces}\;\Sigma\;A\).
+In particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega \to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on finite pomsets.
+
+We define a \textbf{program order pomset} to be a pomset with \(\mc{A} =
+\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
+\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
+\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
+\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
+\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
+program order pomsets given by sequential composition for \(\mc{A}\).
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
 Another interesting use of the trace monad to build a sophisticated \isotopessa model is to build a model of TSO-style weak memory based on \citet{sparky}.
-For simplicity, we will consider an infinite set of named locations \(x, y, z \in \ms{Loc}\) which are subject to concurrent modification by all processors via TSO reads, writes, and fences.
+For simplicity, we will consider an infinite set of named, fixed locations \(x, y, z \in \ms{Loc}\) which are subject to concurrent modification by all processors via TSO reads, writes, and fences.
 
@@
   respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite} if its
-  carrier set is.
+  carrier set is. We will write unordered pomsets using multiset notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list notation, e.g. \([a, a, b]\).
 \end{definition}
@@
 \ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
-program order pomsets given by sequential composition for \(\mc{A}\).
-
-\TODO{how to do reads}
-
-\TODO{how to do writes}
+program order pomsets given by sequential composition for \(\mc{A}\). This
+yields an \isotopessa model with support for concurrent read and write
+operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
+\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
+semantics
+\begin{equation}
+  \dnt{\ms{read}_x} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
+  \qquad
+  \dnt{\ms{write}_x} = \lambda v. \{((), \{x := v\})\}
+\end{equation}
+The semantics of \(\ms{read}\) in particular give a hint as to how this model
+works: rather than tracking the state of the heap, we simply \textit{emit} a
+pomset of the events that would have to happen for a given execution (in the
+case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
+the case of a write, the single write event \(x := v\) for a write of \(v\)),
+and later post-filter on the pomset of an entire program to get a sequentially
+consistent execution. Hence, our semantics remains compositional, allowing us to
+reason about each program fragment individually, while still allowing us to add
+arbitrary external state (e.g. another program concurrently modifying the heap)
+before finally considering the execution of an entire program.
+
+On that note, we support the parallel execution of multiple (potentially
+nonterminating) morphisms as follows:
+\begin{equation}
+  \begin{aligned}
+  f || g = \lambda (a, b). 
+  & \{\iota_0 ((a', b'), \alpha || \beta) 
+    \mid \iota_0 (a', \alpha) \in f\;a 
+    \land \iota_0 (b', \beta) \in f\;b\} 
+  \\ & \cup \{\iota_1 (\alpha || \beta) \mid (\iota_0 (a', \alpha) \in f\;a \lor \iota_1\;\alpha \in f\;a) \land \iota_1 \beta \in f\;b\} 
+  \\ & \cup \{\iota_1 (\alpha || \beta) \mid \iota_1\;\alpha \in f\;a \land \iota_0 (b', \beta) \in f\;b\}: \ms{Set}_{\ms{PO}}(A \otimes B, A' \otimes B')
+  \end{aligned}
+\end{equation}
+It's tempting to try to use this as a tensor product of morphisms and so
+graduate to a bona fide monoidal category rather than our current premonoidal
+setting, since spatial concurrency is one of the headline use-cases of the
+latter, but this definition does \textit{not} support sliding:
+\begin{equation}
+  (\ms{write}_x || \ms{id}) ; (\ms{id} || \ms{write}_y)
+  \neq (\ms{id} || \ms{write}_y) ; (\ms{write}_x || \ms{id})
+\end{equation}
+since the left-hand side will emit the pomset \([x := a, y := b]\) while the right-hand side will emit the pomset \([y := b, x := a]\).
+If we did want this behaviour, we'd need to have a significantly more sophisticated model of composition which effectively takes nontermination into account, which we will leave to future work.
+
+To graduate from SC concurrency to a genuine, if maximally simple, weak memory model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}^*\) of lists of actions, which we will interpret as linear pomsets ordered by index.
 
@@
 
-\TODO{submonad rune}
-
-\TODO{Buffer continuity}
+\TODO{too powerful, submonad rune of buffer continuity}
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
   distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
-  partially-ordered \textbf{carrier set} \(P\) such that every \(p \in P\) has
-  finitely many predecessors equipped with a mapping \(\alpha: P \to \mc{A}\)
-  quotiented under the equivalence relation \(\alpha: P \to \mc{A} \simeq
-  \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\) from \(P\)
-  and \(Q\), there exists a partial order isomorphism between them which
-  respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite} if its
-  carrier set is. We will write unordered pomsets using multiset notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list notation, e.g. \([a, a, b]\).
+  \textit{nonempty} partially-ordered \textbf{carrier set} \(P\) such that every
+  \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
+  \to \mc{A}\) quotiented under the equivalence relation \(\alpha: P \to \mc{A}
+  \simeq \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\)
+  from \(P\) and \(Q\), there exists a partial order isomorphism between them
+  which respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite}
+  if its carrier set is. We will write unordered pomsets using multiset
+  notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list
+  notation, e.g. \([a, a, b]\).
 \end{definition}
-(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\), which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to \mc{A}\), where \(P + Q\) is given the lexicographic ordering and \(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many predecessors.
-They also form a monoid under parallel composition \(\alpha || \beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\) where \(P + Q\) is given the standard partial ordering (with elements of \(P\) and \(Q\) incomparable).
+(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\),
+which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to
+\mc{A}\), where \(P + Q\) is given the lexicographic ordering and
+\(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many
+predecessors. They also form a monoid under parallel composition \(\alpha ||
+\beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\)
+where \(P + Q\) is given the standard partial ordering (with elements of \(P\)
+and \(Q\) incomparable). Note in both cases the monoidal unit is \(\{\delta\}\),
+since we only consider nonempty pomsets but allow the removal of finitely many
+ticks.
 
@@
 setting, since spatial concurrency is one of the headline use-cases of the
-latter, but this definition does \textit{not} support sliding:
+former, but this definition does \textit{not} support sliding:
 \begin{equation}
@@
 
-To graduate from SC concurrency to a genuine, if maximally simple, weak memory model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}^*\) of lists of actions, which we will interpret as linear pomsets ordered by index.
-
-\TODO{weak memory monad}
+To graduate from SC concurrency to a genuine, if maximally simple, weak memory
+model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}^*\) of
+lists of actions, which we will interpret as linear pomsets ordered by index.
+In particular, we define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;\ms{PO}\), which gives us another \isotopessa model in the usual manner.
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
 \begin{equation}
-  \dnt{\ms{read}_x} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
-  \qquad
-  \dnt{\ms{write}_x} = \lambda v. \{((), \{x := v\})\}
+  \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
+  \qquad
+  \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
 \end{equation}
@@
 setting, since spatial concurrency is one of the headline use-cases of the
-former, but this definition does \textit{not} support sliding:
+former, but unfortunately, sliding still fails:
+\(
+  (W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})
+  \neq (\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}} || \ms{id})
+\),
+since the left-hand side will emit the pomset \([x := a, y := b]\) while the
+right-hand side will emit the pomset \([y := b, x := a]\). If we did want this
+behaviour, we'd need to have a significantly more sophisticated model of
+composition which effectively takes nontermination into account, which we will
+leave to future work.
+
+To graduate from SC concurrency to a genuine, if maximally simple, weak memory
+model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}_w^*\)
+of lists of write actions, which we will interpret as linear pomsets ordered by
+index. In particular, we define the monad \(\ms{TSO} =
+\ms{StateT}\;\ms{Buf}\;\ms{PO}\). Given a buffer \(\ms{Buf}\), we can define the
+result of reads \([\cdot]_x: \ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the
+buffer as follows:
 \begin{equation}
-  (\ms{write}_x || \ms{id}) ; (\ms{id} || \ms{write}_y)
-  \neq (\ms{id} || \ms{write}_y) ; (\ms{write}_x || \ms{id})
+  (L;\{x := v\})[x] = v
+  \qquad
+  (L;\{y := v\})[x] = L[x]
+  \qquad
+  [][x] = \bot
 \end{equation}
-since the left-hand side will emit the pomset \([x := a, y := b]\) while the right-hand side will emit the pomset \([y := b, x := a]\).
-If we did want this behaviour, we'd need to have a significantly more sophisticated model of composition which effectively takes nontermination into account, which we will leave to future work.
-
-To graduate from SC concurrency to a genuine, if maximally simple, weak memory
-model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}^*\) of
-lists of actions, which we will interpret as linear pomsets ordered by index.
-In particular, we define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;\ms{PO}\), which gives us another \isotopessa model in the usual manner.
+Note that writes at the \textit{end} of the buffer are prioritized, since later
+writes overwrite earlier ones! We will implement TSO ordering by buffering all
+our writes, in which case they are only visible to the local thread. At any
+point, we may choose to \textit{flush} some of the writes in buffer to the
+global pomset of events. This is implemented via the morphism (called
+\(\ms{split}\) in \cite{sparky})
+\begin{equation}
+  \ms{pflush} = \lambda ()\;L. \{((), L', L'') | L = L';L''\}
+  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
+\end{equation}
+On the other hand, reads events will first attempt to read from the buffer, and,
+if there is no corresponding write in the buffer, will read an arbitrary value,
+in both cases pushing an event to the global pomset; that is, the semantics of
+reads and writes are given by
+\begin{equation}
+  \begin{aligned}
+  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
+    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
+  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
+    &= \ms{pflush};(\lambda v\;L. \{(v, L)\});\ms{pflush}
+  \end{aligned}
+\end{equation}
 
@@
 
+\TODO{in the presence of parallel composition}
+
 \TODO{fence}
 
-\TODO{too powerful, submonad rune of buffer continuity}
+\TODO{big power, such transactional memory, much wow}
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
 \newcommand{\upg}[2]{{#1}^{\uparrow #2}}
+
+% Weak memory
+\newcommand{\bufloc}[1]{\overline{#1}}
 
@@
 
-We define a \textbf{program order pomset} to be a pomset with \(\mc{A} =
+We define a \textbf{program order pomset} to be a pomset with \(\mc{A}_{\ms{PO}} =
 \mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
@@
 To graduate from SC concurrency to a genuine, if maximally simple, weak memory
-model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}_w^*\)
+model, we will to introduce a \textit{buffer} of write actions. We will
+implement TSO ordering by buffering all our writes, in which case they are only
+visible to the local thread. At any point, we may choose to \textit{flush} some
+of the writes in buffer to the global pomset of events.
+\TODO{\(\mc{A}_{\ms{TSO}}\), \(\mc{A}_b\)}
+of type \(\ms{Buf} = \mc{A}_b^*\)
 of lists of write actions, which we will interpret as linear pomsets ordered by
 index. In particular, we define the monad \(\ms{TSO} =
-\ms{StateT}\;\ms{Buf}\;\ms{PO}\). Given a buffer \(\ms{Buf}\), we can define the
+\ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\). 
+\TODO{using \(\mc{A}_{\ms{TSO}}\)}
+Given a buffer \(\ms{Buf}\), we can define the
 result of reads \([\cdot]_x: \ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the
@@
 \begin{equation}
-  (L;\{x := v\})[x] = v
-  \qquad
-  (L;\{y := v\})[x] = L[x]
+  (L;\{\bufloc{x} := v\})[x] = v
+  \qquad
+  (L;\{\bufloc{y} := v\})[x] = L[x]
   \qquad
@@
 Note that writes at the \textit{end} of the buffer are prioritized, since later
-writes overwrite earlier ones! We will implement TSO ordering by buffering all
-our writes, in which case they are only visible to the local thread. At any
-point, we may choose to \textit{flush} some of the writes in buffer to the
-global pomset of events. This is implemented via the morphism (called
+writes overwrite earlier ones! 
+\TODO{segue}
+This is implemented via the morphism (called
 \(\ms{split}\) in \cite{sparky})
 \begin{equation}
-  \ms{pflush} = \lambda ()\;L. \{((), L', L'') | L = L';L''\}
+  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
   : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
 \end{equation}
+\TODO{reflow}
 On the other hand, reads events will first attempt to read from the buffer, and,
@@
   \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
-    &= \ms{pflush};(\lambda v\;L. \{(v, L)\});\ms{pflush}
+    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
   \end{aligned}
 \end{equation}
-
-\TODO{new read/write}
-
+\TODO{text explaining this}
 \TODO{in the presence of parallel composition}

### 2023-11-13 pldi24/paper.tex
Tap

@@
 visible to the local thread. At any point, we may choose to \textit{flush} some
-of the writes in buffer to the global pomset of events.
-\TODO{\(\mc{A}_{\ms{TSO}}\), \(\mc{A}_b\)}
-of type \(\ms{Buf} = \mc{A}_b^*\)
-of lists of write actions, which we will interpret as linear pomsets ordered by
-index. In particular, we define the monad \(\ms{TSO} =
-\ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\). 
-\TODO{using \(\mc{A}_{\ms{TSO}}\)}
-Given a buffer \(\ms{Buf}\), we can define the
-result of reads \([\cdot]_x: \ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the
-buffer as follows:
+of the writes in buffer to the global pomset of events. In particular, we will
+introduce the set \(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write
+actions}, where an action \(\bufloc{x} := v\) by a thread denotes adding a write
+\(x := v\) to the thread's write buffer. We may then define the set of TSO
+actions \(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be
+defined to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will
+interpret as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index. In
+particular, we define the monad \(\ms{TSO} =
+\ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\), where \(\Sigma\) is the stream
+action of finite pomsets over \(\mc{A}_{\ms{TSO}}\) on pomsets over
+\(\mc{A}_{\ms{TSO}}\); we can view \(\ms{PO}\) as a submonad of this monad in
+the standard manner.
+
+Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
+\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
 \begin{equation}

### 2023-11-13 pldi24/paper.tex
Tap

@@
 implement TSO ordering by buffering all our writes, in which case they are only
-visible to the local thread. At any point, we may choose to \textit{flush} some
-of the writes in buffer to the global pomset of events. In particular, we will
-introduce the set \(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write
-actions}, where an action \(\bufloc{x} := v\) by a thread denotes adding a write
-\(x := v\) to the thread's write buffer. We may then define the set of TSO
-actions \(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be
-defined to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will
-interpret as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index. In
-particular, we define the monad \(\ms{TSO} =
-\ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\), where \(\Sigma\) is the stream
-action of finite pomsets over \(\mc{A}_{\ms{TSO}}\) on pomsets over
-\(\mc{A}_{\ms{TSO}}\); we can view \(\ms{PO}\) as a submonad of this monad in
-the standard manner.
+visible to the local thread. On the other hand, reads events will first attempt
+to read from the buffer, and, if there is no corresponding write in the buffer,
+will read an arbitrary value, in both cases pushing an event to the global
+pomset. At any point, we may choose to \textit{flush} some of the writes in
+buffer to the global pomset of events. In particular, we will introduce the set
+\(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write actions}, where an
+action \(\bufloc{x} := v\) by a thread denotes adding a write \(x := v\) to the
+thread's write buffer. We may then define the set of TSO actions
+\(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be defined
+to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret
+as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index. In particular, we
+define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
+where \(\Sigma\) is the stream action of finite pomsets over
+\(\mc{A}_{\ms{TSO}}\) on pomsets over \(\mc{A}_{\ms{TSO}}\); we can view
+\(\ms{PO}\) as a submonad of this monad in the standard manner.
 
@@
 Note that writes at the \textit{end} of the buffer are prioritized, since later
-writes overwrite earlier ones! 
-\TODO{segue}
-This is implemented via the morphism (called
-\(\ms{split}\) in \cite{sparky})
-\begin{equation}
-  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
-  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
-\end{equation}
-\TODO{reflow}
-On the other hand, reads events will first attempt to read from the buffer, and,
-if there is no corresponding write in the buffer, will read an arbitrary value,
-in both cases pushing an event to the global pomset; that is, the semantics of
-reads and writes are given by
+writes overwrite earlier ones! The semantics of reads and writes are then given
+by
 \begin{equation}
@@
 \end{equation}
+where buffer flushing is implemented via the morphism (called \(\ms{split}\) in
+\cite{sparky})
+\begin{equation}
+  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
+  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
+\end{equation}
 \TODO{text explaining this}

### 2023-11-13 pldi24/paper.tex
Tap

@@
   \begin{aligned}
-  f || g = \lambda (a, b). 
-  & \{\iota_0 ((a', b'), \alpha || \beta) 
-    \mid \iota_0 (a', \alpha) \in f\;a 
-    \land \iota_0 (b', \beta) \in f\;b\} 
-  \\ & \cup \{\iota_1 (\alpha || \beta) \mid (\iota_0 (a', \alpha) \in f\;a \lor \iota_1\;\alpha \in f\;a) \land \iota_1 \beta \in f\;b\} 
-  \\ & \cup \{\iota_1 (\alpha || \beta) \mid \iota_1\;\alpha \in f\;a \land \iota_0 (b', \beta) \in f\;b\}: \ms{Set}_{\ms{PO}}(A \otimes B, A' \otimes B')
+  f_0 || f_1 = \lambda (a_0, a_1). 
+  & \{\iota_0 ((b_0, b_1), \alpha_0 || \alpha_1) 
+    \mid \iota_0 (b_i, \alpha_i) \in f_i\;a_i\} 
+  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0) \land \iota_1 \alpha_1 \in f_1\;a_1\} 
+  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0 \land \iota_0 (a_1, \alpha_1) \in f_1\;a_1\}: \ms{Set}_{\ms{PO}}(A \otimes A', B \otimes B')
   \end{aligned}
@@
 \end{equation}
-\TODO{text explaining this}
-\TODO{in the presence of parallel composition}
 
@@
 \TODO{big power, such transactional memory, much wow}
+
+We can now define the parallel composition of morphisms in a "fork-join" style as follows: we first flush the buffer completely (i.e., taking it and sticking it at the beginning of our pomset), then execute \(f\) and \(g\) in parallel with separate buffers, \textit{filtering out executions which completely flush the buffer}.
+Since both threads end up with an empty buffer, the resulting joined buffer is also empty, giving us the following:
+\begin{equation}
+  \begin{aligned}
+    f_0 || f_1 = \lambda (a_0, a_1)\;L. 
+    & \{\iota_0 ((b_0, b_1), [], L;(\alpha_0 || \alpha_1)) 
+      \mid \iota_0 (b_i, [], \alpha_i) \in f_i\;a_i\;[]\} 
+    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, [], \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0\;[]) \land \iota_1 \alpha_1 \in f_1\;a_1\;[]\} 
+    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0\;[] \land \iota_0 (a_1, [], \alpha_1) \in f_1\;a_1\;[]\}
+    \end{aligned}
+\end{equation}
 

### 2023-11-13 pldi24/paper.tex
Tap

@@
 to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret
-as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index. In particular, we
+as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index (with the empty list corresponding to \(\{\delta\}\)). In particular, we
 define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
@@
 \end{equation}
-
-\TODO{fence}
-
-\TODO{big power, such transactional memory, much wow}
-
+To be able to perform synchronization, we will also need to introduce a \(\ms{fence}: \mc{I}^\varnothing_0(\mb{1}, \mb{1})\) instruction, which simply causes all actions before the fence to be observed before any actions after the fence.
+For \(\ms{TSO}\), implementing this is as simple as flushing the buffer, giving
+\begin{equation}
+  \dnt{\ms{fence}} = \lambda ()\;L. \{((), [], L;\{\delta\})\}
+\end{equation}
 We can now define the parallel composition of morphisms in a "fork-join" style as follows: we first flush the buffer completely (i.e., taking it and sticking it at the beginning of our pomset), then execute \(f\) and \(g\) in parallel with separate buffers, \textit{filtering out executions which completely flush the buffer}.

### 2023-11-14 pldi24/paper.tex
Moving things down to appendix
Tap

@@
 
+\TODO{some text about rewriting}
+
+\section{SSA Semantics}
+
+\label{sec:semantics}
+
+\TODO{intro text}
+
+\subsection{Substructural Categories}
+
+\TODO{intro text}
+
+\begin{definition}[Binoidal Category] \
+  \TODO{this}
+\end{definition}
+
+\begin{definition}[Premonoidal Category] \
+  \TODO{this}
+\end{definition}
+
+\begin{definition}[Effectful Category] \
+  \TODO{this}
+\end{definition}
+
+\begin{definition}[(Symmetric) Premonoidal Subcategory]
+  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is closed under whiskering and contains all associators, unitors, and symmetries.
+We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\), \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is \textbf{central} if this generalizes to the case where only one of \(f, g\) is in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
+\end{definition} 
+Note that the set of symmetric premonoidal subcategories, monoidal subcategories, and central subcategories of a category \(\mc{C}\) are all closed under intersections and therefore, in particular, form a complete meet-semilattice with shared bottom element \(\ms{Skeleton}(\mc{C})\), where \(\ms{Skeleton}(C)\) is the smallest possible symmetric premonodial subcategory.
+The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the largest central subcategory is the center of \(\mc{C}\) \(Z(\mb{C})\); the existence of a top-element means that in these cases we may induce the structure of a complete lattice in the usual manner.
+
+\begin{definition}[Strictly Effectful Category] \
+  \TODO{this}
+\end{definition}
+
+\TODO{text}
+
+\begin{definition}[Substructural (Pre)monoidal Category]
+  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A}, \mc{R})\) where:
+  \begin{itemize}
+    \item \(\mc{C}\) is a family of categories \(\mc{C}^q\) parametrized by subsets \(q \subseteq \{\ms{a}, \ms{r}\}\) such that
+    \begin{itemize}
+      \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the same. We will write this set as \(|\mc{C}|\)
+      \item All \(\mc{C}^q\) share symmetric (pre)monoidal structure, and are individually closed under the tensor products of morphisms
+      \item \(\mc{C}^{q \cap q'} = \mc{C}^q \cap \mc{C}^{q'}\)
+    \end{itemize}
+    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects closed under tensor products, such that
+    \begin{itemize}
+      \item \(\mb{1} \in \mc{A}\)
+      \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
+      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
+      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\{\ms{a}\}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
+    \end{itemize}
+    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects closed under tensor products, such that
+    \begin{itemize}
+      \item \(\mb{1} \in \mc{R}\)
+      \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
+      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
+      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
+      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
+      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
+      \item \(\forall A, B \in \mc{R}, \forall f: \mc{C}^{\{\ms{r}\}}(A, B), f;\ms{split}(B) = \ms{split}(A);f \ltimes f = \ms{split}(A);f\rtimes f\)
+    \end{itemize}
+  \end{itemize}
+  We call the morphisms in \(\mc{C}^{\{\ms{a}\}}\) \textbf{affine}, the morphisms in \(\mc{C}^{\{\ms{r}\}}\) \textbf{relevant}, and the morphisms in \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}.
+We call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the other \(\mc{C}^q\) \textbf{linear}.
+We call a substructural \textit{monoidal} category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
+\end{definition}
+\TODO{Define \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) syntax? Make lattice generated by syntax?}
+A simple example of a substructural Cartesian category is the category of relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\).
+In particular, we can define:
+\begin{equation}
+  \begin{aligned}
+    \ms{Rel}^\varnothing(A, B) 
+    &= \ms{Rel}(A, B) 
+    &&= A \to \mc{P}(B) \\
+    \ms{Rel}^{\{\ms{a}\}}(A, B) 
+    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
+    &&= A \to \mc{P}^+(B) \\
+    \ms{Rel}^{\{\ms{r}\}}(A, B) 
+    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
+    &&= A \to \ms{Option}(B) \\
+    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
+    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
+    &&= A \to B
+  \end{aligned}
+\end{equation}
+Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and \(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and that they satisfy the axioms of a commutative comonoid, as desired.
+
+Note that we cannot simply have \(\ms{Rel}^{\{\ms{a}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) = \varnothing\), we have
+\begin{equation}
+  (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a) = \{()\}
+\end{equation}
+We will see that this corresponds to the fact that the following programs are not semantically equal
+\begin{equation}
+  \ms{let}\;x = \ms{loop}(); e \not\simeq e \qquad \text{where} \qquad x \notin \ms{fv}(e)
+\end{equation}
+Similarly, we cannot simply have \(\ms{Rel}^{\{\ms{r}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)| \supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
+\begin{equation}
+  (R;\ms{split}(B))(a) = \{(b, b) \mid b \in R(a)\} \neq (\ms{split}(A);R)(a) = R(a) \times R(a) 
+\end{equation}
+since only the left-hand side contains the pair \((b_0, b_1)\), this corresponding to the fact that the following programs are not semantically equal
+\begin{equation}
+  \ms{let}\;x = \ms{rand}(); (x, x) \not\simeq (\ms{rand}(), \ms{rand}())
+\end{equation}
+To get a substructural premonoidal category, we can apply the state monad transformer to the power set monad to obtain the Kleisli category with morphisms of the form \(\ms{Set}_{\ms{St}\;\mc{P}\;S}(A, B) = A \to S \to \mc{P}(B \times S)\).
+Once again, it is clear that all objects have split and drop morphisms inherited from \(\ms{Set}\), and therefore are affine and relevant
+
+\begin{definition}[Susbtructural Effectful Category]
+  A \textbf{substructural effectful category} \(\mc{C}\) consists of
+  \begin{itemize}
+    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms in this category \textbf{central} or \textbf{linear}
+    \item A substructural premonoidal category \(\mc{C}_0\) sharing the same affine and relevant objects as \(\mc{C}_1\)
+    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0^\varnothing\) preserving all substructural premonoidal structure, i.e., preserves all symmetries, associators, unitors, drops, and splits and sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\).
+    In particular, restricting \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to \mc{C}_0^q\) into effectful categories.
+  \end{itemize}
+  We define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\) and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply be the identity functor on \(\mc{C}_1^q\).
+
+  Morphisms in \(\mc{C}_p^{\{\ms{a}\}}\) are called \textbf{copyable}, morphisms in \(\mc{C}_p^{\{\ms{r}\}}\) are called \textbf{discardable},
+  while morphisms in \(\mc{C}_1^{\{\ms{a}\}}, \mc{C}_1^{\{\ms{r}\}}, \mc{C}_1^{\{\ms{a}, \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and \textbf{intuitionistic} or \textbf{pure} respectively.
+\end{definition}
+
+We can obtain an example of a substructural effectful category quite easily by, considering the power-set monad above, applying the state transformer to yield the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to \mc{P}(A \times S)\).
+Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
+\(
+  \upg{R}{} = \lambda a, s. R(a) \times \{s\}
+\), we can verify that this satisfies the axioms of a substructural effectful category.
+Note that here the \textit{central} morphisms are precisely those which do not read from or write to the state, and therefore which commute with all other morphisms.
+
+In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a substructural premonoidal category, with \(\mc{C}^\varnothing = \mc{C}\) and \(\mc{C}^{\{\ms{a}, \ms{r}\}}, \mc{C}^{\{\ms{a}\}}, \mc{C}^{\{\ms{r}\}}\) the subcategory consisting only of the associators, unitors, and symmetry, and only the identity object considered affine and relevant.
+On the other extreme, every \textit{Cartesian} category \(\mc{C}\) can be viewed as a substructural Cartesian category with all morphisms both relevant and affine; in fact, this is the case \textit{if and only if} a category is Cartesian, hence the term "substructural Cartesian category."
+
+We can also deduce the following:
+\begin{proposition}
+  If \(\mc{C}\) is a substructural monoidal category then the Kleisli category \(\mc{C}_{\mb{T}}\) is a substructural premonoidal category with:
+  \begin{itemize}
+    \item \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\), \(\ms{Rel}(\mc{C}_{\mb{T}}) = \ms{Rel}(\mc{C})\)
+    \item \(\mc{C}_{\mb{T}}^q(A, B) = \upg{\mc{C}^q(A, B)}{}\), where \(\upg{f}{} = f;\eta\)
+    \item Drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\)
+  \end{itemize} 
+  In particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a substructural effectful category.
+\end{proposition}
+
+\TODO{work into text?}
+
+\subsection{Distributivity}
+
+\begin{definition}[Distributive (pre)monoidal category] 
+  A \textbf{distributive (pre)monoidal category} is a (pre)monoidal category with all finite coproducts equipped with a natural family of central isomorphisms (the \textbf{distributor}) \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\).
+This isomorphism is subject to rather complex coherence conditions, worked out in full for the more general case of \textit{rig categories} in \cite{laplaza-distributivity}, but for our purposes we will only make use of the following:
+  \begin{equation*}
+    \delta_{A, B, C}^{-1} = [\iota_0 \otimes C, \iota_1 \otimes C] 
+    \qquad
+    \sigma^+_{A, B} \otimes C;\delta_{B, A, C} = \delta_{A, B, C};\sigma^+_{A \otimes C, B \otimes C}
+  \end{equation*}
+  Note in particular that this implies that \(\iota_i \otimes C;\delta_{A, B, C} = \iota_i\).
+A substructural (pre)monodial category is \textbf{distributive} if it is distributive as a (pre)monoidal category, and its distributor is \textit{pure}.
+A substructural effectful category \(\mc{C}\) is \textbf{distributive} if \(\mc{C}_0\) is distributive (as a substructural (pre)monoidal category).
+\end{definition}
+Many (pre)monoidal categories are distributive, including of course \(\ms{Set}\), but also categories which can be derived from monads on more primitive categories such as \(\ms{PFun}\) and \(\ms{Rel}\), as can be deduced from the following proposition:
+\begin{proposition}
+  If \(\mc{C}\) is a distributive substructural monoidal category and \(\mb{T}\) is a strong monad on \(\mc{C}\), then the Kleisli category \(\mc{C}_{\mb{T}}\) is a distributive substructural effectful category (with affine and relevant objects and morphisms inherited from \(\mc{C}\))
+\end{proposition}
+\begin{proof}
+  Let \(\delta_{A, B, C}: \mc{C}((A + B) \otimes C, A \otimes C + B \otimes C)\) denote \(\mc{C}\)'s distributor.
+We will write composition in \(\mc{C}\) as "\(;\)" and composition in the Kleisli category \(\mb{T}\) as "\(\gg\)".
+  We will show that
+  \(\upg{\delta_{A, B, C}}{} = \delta_{A, B, C};\ms{pure}: \mc{C}_{\mb{T}}((A + B) \otimes C, A \otimes C + B \otimes C)\) satisfies the desired properties of a distributor. In particular, it is:
+  \begin{itemize}
+    \item Central: this follows immediately from being the image of a pure morphism
+    \item Affine: \(\upg{\delta}{} \gg \upg{\ms{drop}}{} = \upg{(\delta;\ms{drop})}{} = \upg{\ms{drop}}{}\)
+    \item Relevant: \(\upg{\delta}{} \gg \upg{\ms{split}}{} = \upg{(\delta;\ms{split})}{} = \upg{(\ms{split};\delta \otimes \delta)}{} = \upg{\ms{split}}{} \gg \upg{\delta}{} \otimes \upg{\delta}{}\)
+    \item Left inverse: \(\upg{\delta}{} \gg \upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \upg{\delta;[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \ms{id}\)
+    \item Right inverse: \(\upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} \gg \upg{\delta}{} = \upg{[\iota_0 \otimes C, \iota_1 \otimes C];\delta}{} = \ms{id}\)
+    \item Symmetric: \(\upg{\sigma^+}{} \otimes C \gg \upg{\delta}{} = \upg{\sigma^+ \otimes C;\delta}{} = \upg{\delta;\sigma^+}{} = \upg{\delta}{} \gg \upg{\sigma^+}{}\)
+  \end{itemize}
+\end{proof}
+
+\TODO{text}
+
+\subsection{Elgot Categories}
+
+\TODO{text}
+
+\begin{definition}[Elgot category] 
+  An \textbf{Elgot category} is a category \(\mc{C}\) with coproducts equipped with an operator \(\cdot^\dagger: \mc{C}(A, B + A) \to \mc{C}(A, B)\) satisfying the following properties:
+  \begin{itemize}
+    \item \textbf{Fixpoint:} \(f;[\ms{id}, f^\dagger] = f^\dagger\)
+    \item \textbf{Naturality:} \((f;g + \ms{id})^\dagger = f^\dagger;g\)
+    \item \textbf{Codiagonal:} \(\forall f \in \mc{C}(A, (B + A) + A), (f^\dagger)^\dagger = (f;[\ms{id}, \iota_1])^\dagger\)
+    \item \textbf{Uniformity:}
+    \(
+      h;f = g;\ms{id} + h \implies h;f^\dagger = g^\dagger
+    \)
+  \end{itemize}
+  A monad \(\mb{T}\) on \(\mc{C}\) such that its Kleisli cateogry \(\mc{C}_{\mb{T}}\) is Elgot is called an \textbf{Elgot monad}.
+\end{definition}
+\TODO{why do these matter? they show up in the proofs, but...}
+Some important properties which hold for all Elgot categories include:
+\begin{itemize}
+  \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id}, (h;[\iota_0, g])^\dagger]\)
+  \item \textbf{Squaring:} \((f;[\iota_0, f])^\dagger = f^\dagger\)
+\end{itemize}
+Proofs can be found in Lemma 31 of \cite{goncharov-squaring}.
+
+\TODO{Hyland and Hasegawa traces?}
+
+\TODO{intuition? Can draw diagrams, but those take up space}
+
+\begin{proposition}
+  If \(\mb{T}\) is an Elgot monad on \(\ms{Set}\), then \(\ms{StateT}\;\mb{T}\;S = \lambda A.
+S \to \ms{T}(A \times S)\), \(\ms{ReaderT}\;\mb{T}\;S = \lambda A.
+S \to \ms{T}(A)\), and \(\ms{WriterT}\;\mb{T}\;S = \lambda A.
+\ms{T}(A \times S)\) are all Elgot
+\end{proposition}
+
+\subsection{Denotational Semantics}
+
+\TODO{make aff and rel syntax nicer/more consistent here? Subcategory syntax would be nice for lemmas and just coherence.}
+
+We're now ready to give a semantics for an \isotopessa dialect with a set of types \(\mc{T}\) and instructions \(\mc{I}\). We'll need:
+\begin{itemize}
+  \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot}, equipped with
+  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies \ms{base}(X) \in \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies \ms{base}(X) \in \ms{Rel}(\mc{C})\)
+  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\) to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
+\end{itemize}
+We begin by giving a semantics for types and contexts in terms of objects in \(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}; this allows us to give a semantics for structural judgements in Figure \ref{fig:ssa-structural-semantics}.
+
+\TODO{some text about what the semantics actually are...}
+
+We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics} followed by a semantics for blocks and label-sets in Figure \ref{fig:ssa-block-semantics}.
+
+\TODO{ditto}
+
+\TODO{rename blocks to statements? label-sets to CFGs?}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{\dnt{A}}: |\mc{C}| \\
+    \dnt{X} = \ms{base}(X) 
+    \qquad
+    \dnt{\mb{1}} = \mb{1}
+    \qquad
+    \dnt{\mb{2}} = \mb{2} = \mb{1} + \mb{1}
+    \qquad \dnt{A \otimes B} = \dnt{A} \otimes \dnt{B} \\
+    \boxed{\dnt{\Gamma}}: |\mc{C}| \\
+    \dnt{\cdot} = \mb{1} \qquad 
+    \dnt{\Gamma, \thyp{x}{A}{q}} = \dnt{\Gamma} \otimes \dnt{A} \\
+    \boxed{\dnt{\ms{L}}}: |\mc{C}| \\
+    \dnt{\cdot} = \mb{0} \qquad
+    \dnt{\ms{L}, \llhyp{\ell}{\Gamma}{A}} = 
+      \dnt{\ms{L}} + \dnt{\Gamma} \otimes \dnt{A}
+  \end{gather*}
+  \caption{Semantics for \isotopessa types and contexts}
+  \Description{Semantics for isotope-SSA types and contexts}
+  \label{fig:ssa-type-semantics}
+\end{figure}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{\dnt{\tlin{A}{\taff}: \mc{C}_1^\tint(\dnt{A}, \mb{1})}}
+      \\
+      \dnt{\tlin{A}{\taff}} = \ms{drop}(\dnt{A})
+      % \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
+      % \dnt{\tlin{A \otimes B}{\taff}} 
+      %   = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
+      %   ; \lambda 
+      \\
+    \boxed{\dnt{\tlin{A}{\trel}}
+      : \mc{C}_1^\tint(\dnt{A}, \dnt{A} \otimes \dnt{A})}
+      \\
+      \dnt{\tlin{A}{\trel}} = \ms{split}(\dnt{A})
+      % \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
+      % \dnt{\tlin{A \otimes B}{\trel}} = 
+      %   \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
+      %   ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
+      \\
+    \boxed{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})}
+      \\
+    \dnt{\csplits{\cdot}{\cdot}{\cdot}} = \lambda^{-1} \qquad
+    \dnt{\csplits
+      {\Gamma, \thyp{x}{A}{q}}
+      {\Delta, \thyp{x}{A}{q}}
+      {\Xi}} 
+      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha;\dnt{\Delta} \otimes \sigma;\alpha 
+      \\
+    \dnt{\csplits
+      {\Gamma, \thyp{x}{A}{q}}
+      {\Delta}
+      {\Xi, \thyp{x}{A}{q}}} 
+      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha 
+      \\
+    \dnt{\csplits
+      {\Gamma, \thyp{x}{A}{q}}
+      {\Delta}
+      {\Xi, \thyp{x}{A}{q}}} \
+      = \dnt{\Gamma} \otimes \dnt{\tlin{A}{\trel}};
+        \alpha;
+        \dnt{\Delta} \otimes \sigma \otimes \dnt{A};
+        \alpha 
+        \\
+    \dnt{\csplits
+      {\Gamma, \thyp{x}{A}{q}}
+      {\Delta}
+      {\Xi}}
+      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+        \otimes \dnt{\tlin{A}{\taff}}
+      ; \rho 
+      \\
+    \boxed{\dnt{\cwk{\Gamma}{\Delta}}
+      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta})} 
+      \\
+      \dnt{\cwk{\Gamma}{\Delta}} 
+      = \dnt{\csplits{\Gamma}{\Delta}{\cdot}};\rho 
+      \\
+    \boxed{\dnt{\lwk{\ms{L}}{\ms{K}}}
+      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} 
+      \\
+    \dnt{\lwk{\cdot}{\cdot}} = \ms{id} \qquad
+    \dnt{\lwk
+      {\ms{L}, \llhyp{\ell}{\Gamma}{A}}
+      {\ms{K}, \llhyp{\ell}{\Delta}{A}}}
+      = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\cwk{\Gamma}{\Delta} \otimes \dnt{A})}{} 
+      \\
+    \dnt{\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
+      = \dnt{\lwk{\ms{L}}{\ms{K}}};\rho_+^{-1}
+      ; \dnt{\ms{K}} \oplus 0_{\dnt{\Gamma} \otimes \dnt{A}}
+  \end{gather*}
+  \caption{Semantics for \isotopessa structural judgements}
+  \Description{Semantics for isotope-SSA structural judgements}
+  \label{fig:ssa-structural-semantics}
+\end{figure}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{\dnt{\hasty{\Gamma}{p}{a}{A}{q}}
+      : \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})} 
+      \\
+    \dnt{\hasty{\Gamma}{p}{x}{A}{q}} 
+      = \dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}
+      \qquad
+    \dnt{\hasty{\Gamma}{p}{f\;a}{B}{q}}
+      = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{p}
+      ; \ms{inst}_p(f) 
+      \\
+    \dnt{\hasty{\Gamma}{p}{(a, b)}{A \otimes B}{q}}
+      = \upg{(
+        \dnt{\csplits{\Gamma}{\Delta}{\Xi}};
+        \dnt{\hasty{\Delta}{1}{a}{A}{q}} \otimes
+        \dnt{\hasty{\Xi}{1}{b}{B}{q}}
+      )}{\uparrow p} 
+      \\
+    \dnt{\hasty{\Gamma}{p}{()}{\mb{1}}{q}}
+      = \upg{\dnt{\cwk{\Gamma}{\cdot}}}{p} 
+      \\
+    \dnt{\hasty{\Gamma}{p}{\ctt}{\mb{2}}{q}}
+      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\ctt)}{p}
+      \qquad
+    \dnt{\hasty{\Gamma}{p}{\cff}{\mb{2}}{q}}
+      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\cff)}{p} 
+      \\
+    \dnt{\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}{q}}
+      = \upg{(
+        \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{q}}
+      )}{p};\dnt{\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}{q}} 
+      \\
+    \dnt{\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}{q}}
+      = \upg{(
+        \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A \otimes B}{q}}
+      )}{p}
+      ; 
+      \\ \qquad \qquad \qquad \qquad \qquad \alpha
+      ; \dnt{\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}{q}}
+    %   \\
+    % \dnt{\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}}
+    %   = \dnt{\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}};\alpha_+
+  \end{gather*}
+  \caption{Semantics for \isotopessa terms}
+  \Description{Semantics for isotope-SSA terms}
+  \label{fig:ssa-term-semantics}
+\end{figure}
+
+\TODO{some text about term semantics}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{\dnt{\haslb{\Gamma}{t}{\ms{L}}}
+      : \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})} \\
+    \dnt{\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}}
+      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{\tint}})}
+      ; \dnt{\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}} \\
+    \dnt{\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}}
+      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+      ; \dnt{\hasty{\Delta}{1}{e}{\mb{2}}{\tint}} \otimes \dnt{\Xi})}
+      ; \ms{ite}_{\dnt{\Xi}} ;
+      \\ \qquad \qquad \qquad \qquad
+      [
+        \dnt{\haslb{\Xi}{s}{\ms{L}}}, 
+        \dnt{\haslb{\Xi}{t}{\ms{L}}} 
+      ]
+      \\
+    \dnt{\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}}
+      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A}{q}}
+      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
+      \\
+    \dnt{\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}}
+      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A \otimes B}{q}}
+      ; \alpha
+      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
+      \\
+    \dnt{\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}}
+      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+      ; \dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger
+      \\
+    \boxed{\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}
+      : \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
+      \\
+    \dnt{\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
+      = \dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
+      ; [[\iota_0, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}], \iota_1]
+      \\
+    \dnt{\lhaslb{\ms{L}}{\cdot}{\ms{L}}}
+      = \iota_0
+  \end{gather*}
+  \caption{Semantics for \isotopessa blocks}
+  \Description{Semantics for isotope-SSA blocks}
+  \label{fig:ssa-block-semantics}
+\end{figure}
+
+\TODO{some text about block semantics}
+
+\subsection{Semantic Metatheory}
+
+\TODO{some text about semantic metatheory}
+
+\begin{lemma}[Coherence] \
+  \begin{itemize}
+    \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{lemma}
+
+\TODO{some text about coherence, implicit use above}
+
+\begin{lemma}[Semantic Upgrade]
+  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
+  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
+\end{lemma}
+
+\TODO{some text about upgrade}
+
+\begin{lemma}[Semantic Weakening] 
+  If \(\cwk{\Gamma}{\Delta}\), then 
+  \begin{itemize}
+    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
+    \[
+      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
+      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+    \]
+    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
+    \[
+      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
+      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
+    \]
+  \end{itemize}
+
+  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
+  \[
+    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
+    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
+  \]
+\end{lemma}
+
+\TODO{some text about semantic weakening}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_1^\varnothing(\dnt{\Theta}, \dnt{\Gamma})} 
+    \\
+    \dnt{\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
+      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{1}{a}{A}{q}} \\
+    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id}
+    \\
+    \boxed{\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}: \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}})} 
+  \end{gather*}
+  \caption{Semantics for \isotopessa substitutions}
+  \Description{Semantics for isotope-SSA substitutions}
+  \label{fig:ssa-subst-semantics}
+\end{figure}
+
+\TODO{some text about substitution semantics}
+
+\begin{theorem}[Semantic Substitution] 
+  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  \begin{itemize}
+    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
+    \[
+      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
+      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
+    \]
+    \item For all 
+      \(\haslb{\Gamma}{t}{\ms{L}}\), 
+      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
+      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
+    \[
+      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
+      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
+      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
+    \]
+    \item For all 
+      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
+      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
+      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
+      \[
+        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
+        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
+        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
+        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+      \]
+  \end{itemize}
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
+  \begin{itemize} 
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+    \[
+      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
+    \]
+    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), we have
+    \[
+      \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      = \dnt{\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{L}}}
+    \]
+  \end{itemize}
+\end{theorem}
+
+\TODO{some text about semantic substitution}
+
+\subsection{Normalization}
+
+\label{ssec:ssa-norm}
+
+\TODO{this}
+
+\subsection{Optimizations}
+
+\TODO{this:}
+
+\TODO{substitution \(\implies\): loop hoisting, global value numbering, some dead code elimination}
+
+\TODO{rewriting \(\implies\) strength reduction, E-graph optimization, see \cite{cranelift}}
+
+\TODO{inlining}
+
+\TODO{loop fusion}
+
+\TODO{unreachable code elimination}
+
+\TODO{more dead code elimination}
+
+\section{Concrete Models}
+
+In this section, we will build on the bare-bones models given in Section
+\ref{sec:semantics} to develop more sophisticated models capable of underpinning
+\isotopessa dialects supporting features such as weak memory and separation
+logic.
+
+\subsection{Trace Models}
+
+We will begin by giving a useful foundation for a wide class of \isotopessa
+models: the "trace models" over \(\mb{Set}\), which accumulate a (potentially
+infinite) nondeterministic trace of monoidal "events". These will later come in
+handy for reasoning about output behaviour and weak memory.
+
+We begin with some definitions:
+\begin{definition}[Stream Action]
+  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a
+  monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to
+  I\) mapping (infinite) streams of \(M\) to elements of \(I\) satisfying
+  \(
+    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
+  \),
+  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the
+  stream \(\ell\).
+\end{definition}
+Some basic examples include:
+\begin{enumerate}
+  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\) forms a stream action on the one element set \(\mathbf{1}\).
+  \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta = \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
+  \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta = \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i \cdot_i l_i)_i\).
+  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite) sequence \(s\).
+  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts on \(\nats \cup \{\infty\}\) by addition.
+\end{enumerate}
+We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted
+\(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of
+those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\).
+For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq
+\omega}) = Z(A^*) = \{[\,]\}\), while \(\mc{SZ}(\nats, \nats \cup \{\infty\}) =
+\{0\} \subseteq Z(\nats) = \nats\). Generally, \(\mc{SZ}(M, I)\) always contains
+the unit of \(M\).
+
+We can now define the \textit{trace monad} of a stream action as follows:
+\begin{definition}[Trace Monad]
+  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the
+  \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A
+  \times M + I)\), with \(\eta_A = \lambda a.\iota_0 (a, 1)\) and 
+  \[
+    \mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}
+  \]
+  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') =
+  \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an
+  Elgot monad, with iteration operator \(f^\dagger(a) = f^\infty(a) \cup
+  \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t
+  \in f_n(a)\}\), where
+  \[
+    f_0 = f, \qquad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n]
+  \]
+  \[
+    f^\infty(a) = \{\Theta (\lambda n. m_n) \mid a_0 = a \land \forall n, \exists a_{n + 1}. \iota_0 (\iota_1\;a_{n + 1}, m_n) \in f(a_n)\}
+  \]
+  \TODO{clean this up}
+\end{definition}
+
+\TODO{text}
+
+\begin{definition}[Relevant, Affine Monad]
+  A monad \(\mb{T}\) on \(\ms{Set}\) is \textbf{relevant} if its Kleisli
+  category satisfies \(f;\ms{split} = \ms{split};f \ltimes f = \ms{split};f
+  \rtimes f\), where \(\ms{split} = \upg{\lambda a. (a, a)}{}\). \(\mb{T}\) is
+  \textbf{affine} if its Kleisli category satisfies \(f;\ms{drop} = \ms{drop}\),
+  where \(\ms{drop} = \upg{\lambda a.()}{}\). A monad is \textbf{intuitionistic}
+  if it is both relevant and affine.
+\end{definition}
+
+\begin{definition}[Submonad] 
+  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a
+  collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq
+  \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in
+  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say
+  that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and,
+  for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where
+  \(f^\dagger\) is computed as in \(\mb{T}\). Note that \(\mb{S}\) is also an
+  (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
+  from \(\mb{T}\).
+\end{definition}
+Some important examples of submonads include:
+\begin{enumerate}
+  \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad
+  \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\);
+  similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
+  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
+  intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
+  submonads, forming a complete meet-semilattice with bottom element
+  \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete
+  lattice in the standard manner.
+  \item In particular, the intersection of a relevant/affine/commutative
+  submonad and another submonad is relevant/affine/commutative respectively.
+  Since every commutative submonad induces a central subcategory of the Kleisli
+  category of the parent monad, it follows that a triple of a relevant, affine,
+  and commutative submonad make the Kleisli category of a monad \(\mb{T}\) into
+  a substructural effectful category with all objects both affine and relevant.
+  \item If \(C \subseteq Z(M)\) is submonoid of \(M\), then \(A \mapsto
+  \mc{P}^+(A \times C + \mb{0}), \mc{P}^+(A \times C \times I) \subseteq
+  \ms{Traces}\;\Theta\;A\) are submonads
+  \item In particular, if \(M_C \subseteq Z(M)\) is a central submonoid of \(M\)
+  such that \(\forall m \in M_C, \forall i \in I, m \cdot i = i\), then \(A
+  \mapsto \mc{P}^+(A \times M_C + \mb{0}) \subseteq \ms{Traces}\;\Theta\;A\) is
+  a submonad whose Kleisli category is a central subcategory of the Kleisli
+  category of \(\ms{Traces}\;\Theta\;A\). Note \(M_C = \{1\}\) is always such a
+  submonoid.
+  \label{item:central-submonad}
+  \item Similarly, if \(M_R \subseteq M_C\) is a submonoid such that \(\forall m
+  \in M_R, m \cdot m = m\), then \(A \mapsto \{S \in \mc{P}^+(A \times M_R + I)
+  \mid |S| = 1\} \subseteq \ms{Traces}\;\Theta\;A\) is a \textit{relevant}
+  submonad. Note \(M_R = \{1\}\) is always such a submonoid.
+  \label{item:relevant-submonad}
+  \item Finally, \(A \mapsto \mc{P}^+(A \times \{1\} + \mb{0}) \subseteq
+  \ms{Traces}\;\Theta\;A\) is an \textit{affine} submonad.
+  \label{item:affine-submonad}
+\end{enumerate}
+In particular, \ref{item:central-submonad}, \ref{item:relevant-submonad},
+\ref{item:affine-submonad} equip the Kleisli category of
+\(\ms{Traces}\;\Theta\;A\) with the structure of a substructural effectful
+category given choices of \(M_C\), \(M_R\).
+
+\subsection{Heaps, Printing, and Separation Logic}
+
+\label{ssec:separation}
+
+As a concrete example, we can now introduce the \textbf{printing monad}, which
+is simply defined as \(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where
+\(\Sigma: (\ms{Bytes}^*)^\omega \to \ms{Bytes}^{\leq \omega}\) denotes
+concatenation of lists of bytestrings. This can be equipped with the structure
+of a substructural effectful category by choosing \(M_C = M_R = \{[]\}\), giving
+us an \isotopessa model supporting the instructions \(\ms{print} \in
+\mc{I}^\varnothing_0(\ms{Bytes}, \mb{1})\) and \(\ms{rand} \in
+\mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard semantics
+\(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), [b])\}\), \(\dnt{\ms{rand}} =
+\lambda (). \{\iota_0 (a, p[]) \mid a \in A\}\).
+
+We can make things a little more interesting by adding a heap to the mix,
+defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where
+\(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function. The
+Kleisli category of this monad can be equipped with the structure of a
+substructural effectful category by simply taking the affine/relevant morphisms
+to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\);
+as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we
+therefore have an \isotopessa model supporting the instructions \(\ms{set}:
+\mc{I}^\varnothing_0(\nats \times \nats, \mb{1})\), \(\ms{get}:
+\mc{I}^\varnothing_0(\nats, \nats)\), \(\ms{alloc}: \mc{I}^\varnothing_0(\nats,
+\nats)\), and \(\ms{free}: \mc{I}^\varnothing_0(\nats, \mb{1})\). We can assign
+semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} =
+\lambda (p, v)\;h. \{\iota_0((), [p \mapsto v]h, [])\}\). \(\ms{get}\) is a bit
+more tricky, since it is unclear what to do when we try to access uninitialized
+memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}}
+= \lambda p\;h. \{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\); for now,
+this will do. Finally, \(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', [])
+\mid h' = h \sqcup p \mapsto v\}\) simply fills a random empty heap cell with
+the provided value, returning a pointer to the cell, while \(\dnt{\ms{free}} =
+\lambda p\;h. \{\iota_0\;((), h \setminus p, [])\}\)
+
+We can use this construction to, building on the ideas in \cite{mellies-ftrs},
+show how a simple separation logic based on \cite{reynolds-separation-2002}
+slightly adapted to the categorical setting gives us a very rich \isotopessa
+model supporting a basic notion of refinement types, while at the same time
+giving us a nontrivial notion of nonlinear types.
+
+Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of
+type \(A \to \ms{Heap} \to \ms{Prop}\). Taking as objects pairs \((A, \Phi)\),
+we can define the category \(\ms{Sep}\) to have morphisms
+\begin{equation}
+  \ms{Sep}((A, \Phi), (B, \Psi)) = \{f: A \to \ms{Comp}\;B 
+    \mid \forall \phi: \ms{Heap} \to \ms{Prop}, \ms{hoare}\;f\;(\Phi \ast \phi)\;(\Psi \ast \phi)\}
+\end{equation}
+where
+\begin{equation}
+  \ms{hoare}\;f\;\Phi\;\Psi = \forall a\;h. \Phi\;a\;h \implies \forall \iota_0\;(b, h') \in f\;a\;h. \Psi\;b\;h'
+\end{equation}
+is the usual partial correctness triple definition and the separating
+conjunction of heap-predicates is defined as usual to be 
+\begin{equation}
+  \begin{gathered}
+  \varphi \ast \psi = \lambda h. \exists h_1, h_2. h = h_1 \sqcup h_2 \land \varphi h_1 \land \psi h_2 
+  \qquad
+  \varphi \ast \Phi = \lambda a. \varphi \ast \Phi\;a
+  \qquad
+  \Phi \ast \varphi = \lambda a. \Phi\;a \ast \varphi \\
+  \Phi \ast \Psi = \lambda (a, b). \Phi\;a \ast \Psi\;b: A \times B \to \ms{Heap} \to \ms{Prop}
+  \end{gathered}
+\end{equation}
+We can naturally define a tensor product on objects \((A, \Phi) \otimes (B,
+\Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking
+\begin{itemize}
+  \item Monoidal unit \((\mb{1}, \ms{emp})\), where \(\ms{emp}\) is the
+  predicate which holds only for the empty heap
+  \item Associators and inverse associators given by \(\alpha_{(A, \Phi), (B,
+  \Psi), (C, \Theta)} = \alpha_{A, B, C}\), which are valid since \(\ast\) is
+  associative
+  \item Unitors and inverse unitors given by \(\lambda_{(A, \Phi)} =
+  \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\), which are valid since \(\ms{emp}\)
+  is the unit of \(\ast\)
+  \item \((C, \Theta) \otimes f = C \times f\) and \(f \otimes (C, \Theta) = f
+  \times C\), which are valid definitions by the frame rule
+\end{itemize}
+equips \(\ms{Sep}\) with the structure of a premonoidal category. We can give \(\ms{Sep}\) the structure of a substructural effectful category by taking:
+\begin{itemize}
+  \item Affine objects to be precisely those for which \(\ms{drop} =
+  \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), which for
+  this logic corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend
+  on the heap.
+  \item Relevant objects to be precisely those for which \(\ms{split} =
+  \upg{\lambda a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), again in
+  this case those for which \(\Phi\) does not depend on the heap.
+  \item Central morphisms to be those which do not print, diverge, \textit{or} modify the
+  heap. We will consider how to relax this condition later.
+  \item Relevant and affine morphisms to be the lifts of relevant/affine morphisms in \(\ms{Print}\)
+\end{itemize}
+\(\ms{Sep}\) also has coproducts given by \((A, \Phi) + (B, \Psi) = (A + B, [\Phi, \Psi])\), which allow us to equip it with an Elgot structure by taking advantage of the fact that
+\begin{equation}
+  f \in \ms{Sep}((A, \Phi), (B + A, [\Psi, \Phi])) \implies f^\dagger \in \ms{Sep}((A, \Phi), (B, \Psi))
+\end{equation}
+This implies that \(\ms{Sep}\) is an \isotopessa model supporting the following interesting operations
+\begin{itemize}
+  \item \(\ms{coe}: \mc{I}^\tint_1((A, \Phi), (A, \Psi))\) where \(\forall a,
+  \Phi\;a \implies \Psi\;a\), implemented by the identity morphism
+  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Bytes}, \ms{emp}), (\mb{1},
+  \ms{emp}))\)
+  \item \(\ms{rand}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for \(A\) nonempty
+  \item \(\ms{set}: \mc{I}^\varnothing_0((\nats \times \nats, \lambda (\ell,
+  \cdot). \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell
+  \mapsto v))\), where \(\ms{set}\) is modified slightly to return the address
+  at which the value passed in was stored to overcome a lack of dependency
+  between the source and target objects.
+  \item \(\ms{get}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)
+  \item \(\ms{alloc}: \mc{I}^\varnothing_0((\nats, \ms{emp}), (\nats, \lambda
+  \ell. \exists v. \ell \mapsto v))\)
+  \item \(\ms{free}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\mb{1}, \ms{emp}))\)
+\end{itemize}
+We can clearly see that this is a refinement type system in the sence of
+\cite{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi)
+\mapsto A\) which is simply the identity on morphisms. Unfortunately, this means
+that our notion of equality is a little too stringent, since we need to consider
+equality on inputs which do not match our preconditions. Thankfully, we can fix
+this problem quite easily by quotienting morphisms over the equivalence relation
+\begin{equation}
+  \forall f, g: \ms{Sep}((A, \Phi), (B, \Psi)).
+  f \simeq g \iff \forall a, h. \Phi\;a\;h \implies f\;a\;h = g\;a\;h
+\end{equation}
+yielding the new category \(\ms{SComp}\). In particular, this allows us to widen
+our definition of central morphisms to include all morphisms which do not print
+or diverge, including in particular \(\ms{set}, \ms{get}, \ms{alloc}\) and
+\(\ms{free}\), which we can in fact mark relevant and affine as well, allowing
+us, for example, to make use of intermediate allocations in an otherwise pure
+computation.
+
+\subsection{Weak Memory}
+
+Another interesting use of the trace monad to build a sophisticated \isotopessa model is to build a model of TSO-style weak memory based on \citet{sparky}.
+For simplicity, we will consider an infinite set of named, fixed locations \(x, y, z \in \ms{Loc}\) which are subject to concurrent modification by all processors via TSO reads, writes, and fences.
+
+We begin with the following definitions:
+\begin{definition}[Pomset] 
+  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
+  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
+  \textit{nonempty} partially-ordered \textbf{carrier set} \(P\) such that every
+  \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
+  \to \mc{A}\) quotiented under the equivalence relation \(\alpha: P \to \mc{A}
+  \simeq \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\)
+  from \(P\) and \(Q\), there exists a partial order isomorphism between them
+  which respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite}
+  if its carrier set is. We will write unordered pomsets using multiset
+  notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list
+  notation, e.g. \([a, a, b]\).
+\end{definition}
+(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\),
+which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to
+\mc{A}\), where \(P + Q\) is given the lexicographic ordering and
+\(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many
+predecessors. They also form a monoid under parallel composition \(\alpha ||
+\beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\)
+where \(P + Q\) is given the standard partial ordering (with elements of \(P\)
+and \(Q\) incomparable). Note in both cases the monoidal unit is \(\{\delta\}\),
+since we only consider nonempty pomsets but allow the removal of finitely many
+ticks.
+
+Given a partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n \in N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by the function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\), where the dependent product \(\Sigma_nP_n\) is given the lexicographic order.
+In particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega \to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on finite pomsets.
+
+We define a \textbf{program order pomset} to be a pomset with \(\mc{A}_{\ms{PO}} =
+\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
+\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
+\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
+\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
+\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
+program order pomsets given by sequential composition for \(\mc{A}\). This
+yields an \isotopessa model with support for concurrent read and write
+operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
+\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
+semantics
+\begin{equation}
+  \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
+  \qquad
+  \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
+\end{equation}
+The semantics of \(\ms{read}\) in particular give a hint as to how this model
+works: rather than tracking the state of the heap, we simply \textit{emit} a
+pomset of the events that would have to happen for a given execution (in the
+case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
+the case of a write, the single write event \(x := v\) for a write of \(v\)),
+and later post-filter on the pomset of an entire program to get a sequentially
+consistent execution. Hence, our semantics remains compositional, allowing us to
+reason about each program fragment individually, while still allowing us to add
+arbitrary external state (e.g. another program concurrently modifying the heap)
+before finally considering the execution of an entire program.
+
+On that note, we support the parallel execution of multiple (potentially
+nonterminating) morphisms as follows:
+\begin{equation}
+  \begin{aligned}
+  f_0 || f_1 = \lambda (a_0, a_1). 
+  & \{\iota_0 ((b_0, b_1), \alpha_0 || \alpha_1) 
+    \mid \iota_0 (b_i, \alpha_i) \in f_i\;a_i\} 
+  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0) \land \iota_1 \alpha_1 \in f_1\;a_1\} 
+  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0 \land \iota_0 (a_1, \alpha_1) \in f_1\;a_1\}: \ms{Set}_{\ms{PO}}(A \otimes A', B \otimes B')
+  \end{aligned}
+\end{equation}
+It's tempting to try to use this as a tensor product of morphisms and so
+graduate to a bona fide monoidal category rather than our current premonoidal
+setting, since spatial concurrency is one of the headline use-cases of the
+former, but unfortunately, sliding still fails:
+\(
+  (W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})
+  \neq (\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}} || \ms{id})
+\),
+since the left-hand side will emit the pomset \([x := a, y := b]\) while the
+right-hand side will emit the pomset \([y := b, x := a]\). If we did want this
+behaviour, we'd need to have a significantly more sophisticated model of
+composition which effectively takes nontermination into account, which we will
+leave to future work.
+
+To graduate from SC concurrency to a genuine, if maximally simple, weak memory
+model, we will to introduce a \textit{buffer} of write actions. We will
+implement TSO ordering by buffering all our writes, in which case they are only
+visible to the local thread. On the other hand, reads events will first attempt
+to read from the buffer, and, if there is no corresponding write in the buffer,
+will read an arbitrary value, in both cases pushing an event to the global
+pomset. At any point, we may choose to \textit{flush} some of the writes in
+buffer to the global pomset of events. In particular, we will introduce the set
+\(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write actions}, where an
+action \(\bufloc{x} := v\) by a thread denotes adding a write \(x := v\) to the
+thread's write buffer. We may then define the set of TSO actions
+\(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be defined
+to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret
+as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index (with the empty list corresponding to \(\{\delta\}\)). In particular, we
+define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
+where \(\Sigma\) is the stream action of finite pomsets over
+\(\mc{A}_{\ms{TSO}}\) on pomsets over \(\mc{A}_{\ms{TSO}}\); we can view
+\(\ms{PO}\) as a submonad of this monad in the standard manner.
+
+Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
+\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
+\begin{equation}
+  (L;\{\bufloc{x} := v\})[x] = v
+  \qquad
+  (L;\{\bufloc{y} := v\})[x] = L[x]
+  \qquad
+  [][x] = \bot
+\end{equation}
+Note that writes at the \textit{end} of the buffer are prioritized, since later
+writes overwrite earlier ones! The semantics of reads and writes are then given
+by
+\begin{equation}
+  \begin{aligned}
+  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
+    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
+  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
+    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
+  \end{aligned}
+\end{equation}
+where buffer flushing is implemented via the morphism (called \(\ms{split}\) in
+\cite{sparky})
+\begin{equation}
+  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
+  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
+\end{equation}
+To be able to perform synchronization, we will also need to introduce a \(\ms{fence}: \mc{I}^\varnothing_0(\mb{1}, \mb{1})\) instruction, which simply causes all actions before the fence to be observed before any actions after the fence.
+For \(\ms{TSO}\), implementing this is as simple as flushing the buffer, giving
+\begin{equation}
+  \dnt{\ms{fence}} = \lambda ()\;L. \{((), [], L;\{\delta\})\}
+\end{equation}
+We can now define the parallel composition of morphisms in a "fork-join" style as follows: we first flush the buffer completely (i.e., taking it and sticking it at the beginning of our pomset), then execute \(f\) and \(g\) in parallel with separate buffers, \textit{filtering out executions which completely flush the buffer}.
+Since both threads end up with an empty buffer, the resulting joined buffer is also empty, giving us the following:
+\begin{equation}
+  \begin{aligned}
+    f_0 || f_1 = \lambda (a_0, a_1)\;L. 
+    & \{\iota_0 ((b_0, b_1), [], L;(\alpha_0 || \alpha_1)) 
+      \mid \iota_0 (b_i, [], \alpha_i) \in f_i\;a_i\;[]\} 
+    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, [], \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0\;[]) \land \iota_1 \alpha_1 \in f_1\;a_1\;[]\} 
+    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0\;[] \land \iota_0 (a_1, [], \alpha_1) \in f_1\;a_1\;[]\}
+    \end{aligned}
+\end{equation}
+
+\section{Related Work}
+
+\TODO{\cite{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
+
+\TODO{\cite{schneider-phd}}
+
+\TODO{\cite{chakravarty-functional-ssa-2003}}
+
+\TODO{\cite{barthe-compcert-ssa-2014}}
+
+\TODO{\cite{buchwald-verified-ssa-2016}}
+
+\TODO{\cite{herklotz-gsa-2023}}
+
+\TODO{\cite{pop-ssa-inout-2009}}
+
+\TODO{\cite{garbuzov-structural-cfg-2018}}
+
+\TODO{\cite{hua-explicit-ssa-2010}}
+
+\TODO{\cite{rvsdg}}
+
+\TODO{this}
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{references}
+
+\clearpage
+
+\appendix
+
+\section{Appendix}
+
 \begin{figure}
@@
 
-\TODO{some text about rewriting}
-
-\section{SSA Semantics}
-
-\label{sec:semantics}
-
-\TODO{intro text}
-
-\subsection{Substructural Categories}
-
-\TODO{intro text}
-
-\begin{definition}[Binoidal Category] \
-  \TODO{this}
-\end{definition}
-
-\begin{definition}[Premonoidal Category] \
-  \TODO{this}
-\end{definition}
-
-\begin{definition}[Effectful Category] \
-  \TODO{this}
-\end{definition}
-
-\begin{definition}[(Symmetric) Premonoidal Subcategory]
-  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is closed under whiskering and contains all associators, unitors, and symmetries.
-We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\), \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is \textbf{central} if this generalizes to the case where only one of \(f, g\) is in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
-\end{definition} 
-Note that the set of symmetric premonoidal subcategories, monoidal subcategories, and central subcategories of a category \(\mc{C}\) are all closed under intersections and therefore, in particular, form a complete meet-semilattice with shared bottom element \(\ms{Skeleton}(\mc{C})\), where \(\ms{Skeleton}(C)\) is the smallest possible symmetric premonodial subcategory.
-The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the largest central subcategory is the center of \(\mc{C}\) \(Z(\mb{C})\); the existence of a top-element means that in these cases we may induce the structure of a complete lattice in the usual manner.
-
-\begin{definition}[Strictly Effectful Category] \
-  \TODO{this}
-\end{definition}
-
-\TODO{text}
-
-\begin{definition}[Substructural (Pre)monoidal Category]
-  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A}, \mc{R})\) where:
-  \begin{itemize}
-    \item \(\mc{C}\) is a family of categories \(\mc{C}^q\) parametrized by subsets \(q \subseteq \{\ms{a}, \ms{r}\}\) such that
-    \begin{itemize}
-      \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the same. We will write this set as \(|\mc{C}|\)
-      \item All \(\mc{C}^q\) share symmetric (pre)monoidal structure, and are individually closed under the tensor products of morphisms
-      \item \(\mc{C}^{q \cap q'} = \mc{C}^q \cap \mc{C}^{q'}\)
-    \end{itemize}
-    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects closed under tensor products, such that
-    \begin{itemize}
-      \item \(\mb{1} \in \mc{A}\)
-      \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
-      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
-      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\{\ms{a}\}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
-    \end{itemize}
-    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects closed under tensor products, such that
-    \begin{itemize}
-      \item \(\mb{1} \in \mc{R}\)
-      \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
-      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
-      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
-      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
-      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
-      \item \(\forall A, B \in \mc{R}, \forall f: \mc{C}^{\{\ms{r}\}}(A, B), f;\ms{split}(B) = \ms{split}(A);f \ltimes f = \ms{split}(A);f\rtimes f\)
-    \end{itemize}
-  \end{itemize}
-  We call the morphisms in \(\mc{C}^{\{\ms{a}\}}\) \textbf{affine}, the morphisms in \(\mc{C}^{\{\ms{r}\}}\) \textbf{relevant}, and the morphisms in \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}.
-We call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the other \(\mc{C}^q\) \textbf{linear}.
-We call a substructural \textit{monoidal} category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
-\end{definition}
-\TODO{Define \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) syntax? Make lattice generated by syntax?}
-A simple example of a substructural Cartesian category is the category of relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\).
-In particular, we can define:
-\begin{equation}
-  \begin{aligned}
-    \ms{Rel}^\varnothing(A, B) 
-    &= \ms{Rel}(A, B) 
-    &&= A \to \mc{P}(B) \\
-    \ms{Rel}^{\{\ms{a}\}}(A, B) 
-    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
-    &&= A \to \mc{P}^+(B) \\
-    \ms{Rel}^{\{\ms{r}\}}(A, B) 
-    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
-    &&= A \to \ms{Option}(B) \\
-    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
-    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
-    &&= A \to B
-  \end{aligned}
-\end{equation}
-Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and \(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and that they satisfy the axioms of a commutative comonoid, as desired.
-
-Note that we cannot simply have \(\ms{Rel}^{\{\ms{a}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) = \varnothing\), we have
-\begin{equation}
-  (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a) = \{()\}
-\end{equation}
-We will see that this corresponds to the fact that the following programs are not semantically equal
-\begin{equation}
-  \ms{let}\;x = \ms{loop}(); e \not\simeq e \qquad \text{where} \qquad x \notin \ms{fv}(e)
-\end{equation}
-Similarly, we cannot simply have \(\ms{Rel}^{\{\ms{r}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)| \supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
-\begin{equation}
-  (R;\ms{split}(B))(a) = \{(b, b) \mid b \in R(a)\} \neq (\ms{split}(A);R)(a) = R(a) \times R(a) 
-\end{equation}
-since only the left-hand side contains the pair \((b_0, b_1)\), this corresponding to the fact that the following programs are not semantically equal
-\begin{equation}
-  \ms{let}\;x = \ms{rand}(); (x, x) \not\simeq (\ms{rand}(), \ms{rand}())
-\end{equation}
-To get a substructural premonoidal category, we can apply the state monad transformer to the power set monad to obtain the Kleisli category with morphisms of the form \(\ms{Set}_{\ms{St}\;\mc{P}\;S}(A, B) = A \to S \to \mc{P}(B \times S)\).
-Once again, it is clear that all objects have split and drop morphisms inherited from \(\ms{Set}\), and therefore are affine and relevant
-
-\begin{definition}[Susbtructural Effectful Category]
-  A \textbf{substructural effectful category} \(\mc{C}\) consists of
-  \begin{itemize}
-    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms in this category \textbf{central} or \textbf{linear}
-    \item A substructural premonoidal category \(\mc{C}_0\) sharing the same affine and relevant objects as \(\mc{C}_1\)
-    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0^\varnothing\) preserving all substructural premonoidal structure, i.e., preserves all symmetries, associators, unitors, drops, and splits and sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\).
-    In particular, restricting \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to \mc{C}_0^q\) into effectful categories.
-  \end{itemize}
-  We define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\) and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply be the identity functor on \(\mc{C}_1^q\).
-
-  Morphisms in \(\mc{C}_p^{\{\ms{a}\}}\) are called \textbf{copyable}, morphisms in \(\mc{C}_p^{\{\ms{r}\}}\) are called \textbf{discardable},
-  while morphisms in \(\mc{C}_1^{\{\ms{a}\}}, \mc{C}_1^{\{\ms{r}\}}, \mc{C}_1^{\{\ms{a}, \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and \textbf{intuitionistic} or \textbf{pure} respectively.
-\end{definition}
-
-We can obtain an example of a substructural effectful category quite easily by, considering the power-set monad above, applying the state transformer to yield the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to \mc{P}(A \times S)\).
-Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
-\(
-  \upg{R}{} = \lambda a, s. R(a) \times \{s\}
-\), we can verify that this satisfies the axioms of a substructural effectful category.
-Note that here the \textit{central} morphisms are precisely those which do not read from or write to the state, and therefore which commute with all other morphisms.
+\begin{lemma}[Substitution Coherence] \
+  \begin{itemize}
+    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{lemma}
+
+\TODO{pull down to appendix? just trivial induction.}
+
+\begin{lemma}[Semantic Substitution Splitting] 
+  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
+  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
+  \[
+    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
+      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
+      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
+  \]
+\end{lemma}
+
+\begin{figure}
+  \begin{align*}
+    \dnt{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}
+      &= \dnt{H} \times \mc{C}_p^q(\dnt{\Gamma}, \dnt{A}) 
+      \\
+    \dnt{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}
+      &= \dnt{H} \times \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}}) 
+      \\
+    \dnt{H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
+      &= \dnt{H} \times \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
+      \\
+    \dnt{\cdot} 
+      &= \mb{1}
+  \end{align*}
+  \caption{Semantics for \isotopessa holes}
+  \Description{Semantics for isotope-SSA holes}
+  \label{fig:holes-semantics}
+\end{figure}
+
+\TODO{some text about semantics of holes}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{
+      \dnt{\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}}
+      : \dnt{H} \to \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})}
+      \\
+    \sorry
+  \end{gather*}
+  \caption{Semantics for \isotopessa terms with holes}
+  \Description{Semantics for isotope-SSA blocks with holes}
+  \label{fig:terms-with-holes-semantics}
+\end{figure}
+
+\TODO{fill}
+
+\TODO{some text about semantics of terms with holes}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{
+      \dnt{\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
+      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})}
+      \\
+    \sorry
+      \\  
+    \boxed{
+      \dnt{\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
+      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
+      \\
+    \sorry
+  \end{gather*}
+  \caption{Semantics for \isotopessa blocks with holes}
+  \Description{Semantics for isotope-SSA blocks with holes}
+  \label{fig:blocks-with-holes-semantics}
+\end{figure}
+
+\TODO{fill}
+
+\TODO{some text about semantics of blocks with holes}
+
+\begin{figure}
+  \begin{gather*}
+    \boxed{
+      \dnt{\isrw{\mc{H}}{I}{H}}
+      : \dnt{I} \to \dnt{H} 
+    }
+    \\
+    \dnt{\isrw{\mc{H}}{I}{\cdot}}\;i = ()
+    \qquad
+    \dnt{\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}\;i
+    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}\;i)
+    \\
+    \dnt{\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
+    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}\;i)
+    \\
+    \dnt{\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
+    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}\;i)
+  \end{gather*}
+  \caption{Semantics for \isotopessa rewrites}
+  \Description{Semantics for isotope-SSA rewrites}
+  \label{fig:rewrite-semantics}
+\end{figure}
+
+\TODO{fill}
+
+\TODO{some text about semantics of rewrites}
+
+\begin{lemma}[Rewriting Coherence] \
+  \begin{itemize}
+    \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{lemma}
+
+\begin{theorem}[Semantic Rewriting]
+  Given \(\isrw{\mc{H}}{I}{H}\), we have that
+  \begin{itemize}
+    \item For all \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(m \in \dnt{I}\), we have
+    \[
+      \dnt{\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}}(m)
+      = \dnt{\mhasty{H}{\Gamma}{p}{a}{A}{q}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
+    \]
+    \item For all \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(m \in \dnt{I}\), we have
+    \[
+      \dnt{\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}}(m)
+      = \dnt{\mhaslb{H}{\Gamma}{t}{\ms{L}}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
+    \]
+  \end{itemize}
+
+  Furthermore, we have that
+  \[
+    \dnt{\mhasty{\cdot}{\Gamma}{p}{a}{A}{q}}() = \dnt{\hasty{\Gamma}{p}{a}{A}{q}}
+    \qquad
+    \dnt{\mhaslb{\cdot}{\Gamma}{t}{\ms{L}}}() = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+  \]
+\end{theorem}
 
@@
 
-In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a substructural premonoidal category, with \(\mc{C}^\varnothing = \mc{C}\) and \(\mc{C}^{\{\ms{a}, \ms{r}\}}, \mc{C}^{\{\ms{a}\}}, \mc{C}^{\{\ms{r}\}}\) the subcategory consisting only of the associators, unitors, and symmetry, and only the identity object considered affine and relevant.
-On the other extreme, every \textit{Cartesian} category \(\mc{C}\) can be viewed as a substructural Cartesian category with all morphisms both relevant and affine; in fact, this is the case \textit{if and only if} a category is Cartesian, hence the term "substructural Cartesian category."
-
-We can also deduce the following:
-\begin{proposition}
-  If \(\mc{C}\) is a substructural monoidal category then the Kleisli category \(\mc{C}_{\mb{T}}\) is a substructural premonoidal category with:
-  \begin{itemize}
-    \item \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\), \(\ms{Rel}(\mc{C}_{\mb{T}}) = \ms{Rel}(\mc{C})\)
-    \item \(\mc{C}_{\mb{T}}^q(A, B) = \upg{\mc{C}^q(A, B)}{}\), where \(\upg{f}{} = f;\eta\)
-    \item Drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\)
-  \end{itemize} 
-  In particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a substructural effectful category.
-\end{proposition}
-
-\TODO{work into text?}
-
-\subsection{Distributivity}
-
-\begin{definition}[Distributive (pre)monoidal category] 
-  A \textbf{distributive (pre)monoidal category} is a (pre)monoidal category with all finite coproducts equipped with a natural family of central isomorphisms (the \textbf{distributor}) \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\).
-This isomorphism is subject to rather complex coherence conditions, worked out in full for the more general case of \textit{rig categories} in \cite{laplaza-distributivity}, but for our purposes we will only make use of the following:
-  \begin{equation*}
-    \delta_{A, B, C}^{-1} = [\iota_0 \otimes C, \iota_1 \otimes C] 
-    \qquad
-    \sigma^+_{A, B} \otimes C;\delta_{B, A, C} = \delta_{A, B, C};\sigma^+_{A \otimes C, B \otimes C}
-  \end{equation*}
-  Note in particular that this implies that \(\iota_i \otimes C;\delta_{A, B, C} = \iota_i\).
-A substructural (pre)monodial category is \textbf{distributive} if it is distributive as a (pre)monoidal category, and its distributor is \textit{pure}.
-A substructural effectful category \(\mc{C}\) is \textbf{distributive} if \(\mc{C}_0\) is distributive (as a substructural (pre)monoidal category).
-\end{definition}
-Many (pre)monoidal categories are distributive, including of course \(\ms{Set}\), but also categories which can be derived from monads on more primitive categories such as \(\ms{PFun}\) and \(\ms{Rel}\), as can be deduced from the following proposition:
-\begin{proposition}
-  If \(\mc{C}\) is a distributive substructural monoidal category and \(\mb{T}\) is a strong monad on \(\mc{C}\), then the Kleisli category \(\mc{C}_{\mb{T}}\) is a distributive substructural effectful category (with affine and relevant objects and morphisms inherited from \(\mc{C}\))
-\end{proposition}
-\begin{proof}
-  Let \(\delta_{A, B, C}: \mc{C}((A + B) \otimes C, A \otimes C + B \otimes C)\) denote \(\mc{C}\)'s distributor.
-We will write composition in \(\mc{C}\) as "\(;\)" and composition in the Kleisli category \(\mb{T}\) as "\(\gg\)".
-  We will show that
-  \(\upg{\delta_{A, B, C}}{} = \delta_{A, B, C};\ms{pure}: \mc{C}_{\mb{T}}((A + B) \otimes C, A \otimes C + B \otimes C)\) satisfies the desired properties of a distributor. In particular, it is:
-  \begin{itemize}
-    \item Central: this follows immediately from being the image of a pure morphism
-    \item Affine: \(\upg{\delta}{} \gg \upg{\ms{drop}}{} = \upg{(\delta;\ms{drop})}{} = \upg{\ms{drop}}{}\)
-    \item Relevant: \(\upg{\delta}{} \gg \upg{\ms{split}}{} = \upg{(\delta;\ms{split})}{} = \upg{(\ms{split};\delta \otimes \delta)}{} = \upg{\ms{split}}{} \gg \upg{\delta}{} \otimes \upg{\delta}{}\)
-    \item Left inverse: \(\upg{\delta}{} \gg \upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \upg{\delta;[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \ms{id}\)
-    \item Right inverse: \(\upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} \gg \upg{\delta}{} = \upg{[\iota_0 \otimes C, \iota_1 \otimes C];\delta}{} = \ms{id}\)
-    \item Symmetric: \(\upg{\sigma^+}{} \otimes C \gg \upg{\delta}{} = \upg{\sigma^+ \otimes C;\delta}{} = \upg{\delta;\sigma^+}{} = \upg{\delta}{} \gg \upg{\sigma^+}{}\)
-  \end{itemize}
-\end{proof}
-
-\TODO{text}
-
-\subsection{Elgot Categories}
-
-\TODO{text}
-
-\begin{definition}[Elgot category] 
-  An \textbf{Elgot category} is a category \(\mc{C}\) with coproducts equipped with an operator \(\cdot^\dagger: \mc{C}(A, B + A) \to \mc{C}(A, B)\) satisfying the following properties:
-  \begin{itemize}
-    \item \textbf{Fixpoint:} \(f;[\ms{id}, f^\dagger] = f^\dagger\)
-    \item \textbf{Naturality:} \((f;g + \ms{id})^\dagger = f^\dagger;g\)
-    \item \textbf{Codiagonal:} \(\forall f \in \mc{C}(A, (B + A) + A), (f^\dagger)^\dagger = (f;[\ms{id}, \iota_1])^\dagger\)
-    \item \textbf{Uniformity:}
-    \(
-      h;f = g;\ms{id} + h \implies h;f^\dagger = g^\dagger
-    \)
-  \end{itemize}
-  A monad \(\mb{T}\) on \(\mc{C}\) such that its Kleisli cateogry \(\mc{C}_{\mb{T}}\) is Elgot is called an \textbf{Elgot monad}.
-\end{definition}
-\TODO{why do these matter? they show up in the proofs, but...}
-Some important properties which hold for all Elgot categories include:
-\begin{itemize}
-  \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id}, (h;[\iota_0, g])^\dagger]\)
-  \item \textbf{Squaring:} \((f;[\iota_0, f])^\dagger = f^\dagger\)
-\end{itemize}
-Proofs can be found in Lemma 31 of \cite{goncharov-squaring}.
-
-\TODO{Hyland and Hasegawa traces?}
-
-\TODO{intuition? Can draw diagrams, but those take up space}
-
-\begin{proposition}
-  If \(\mb{T}\) is an Elgot monad on \(\ms{Set}\), then \(\ms{StateT}\;\mb{T}\;S = \lambda A.
-S \to \ms{T}(A \times S)\), \(\ms{ReaderT}\;\mb{T}\;S = \lambda A.
-S \to \ms{T}(A)\), and \(\ms{WriterT}\;\mb{T}\;S = \lambda A.
-\ms{T}(A \times S)\) are all Elgot
-\end{proposition}
-
-\subsection{Denotational Semantics}
-
-\TODO{make aff and rel syntax nicer/more consistent here? Subcategory syntax would be nice for lemmas and just coherence.}
-
-We're now ready to give a semantics for an \isotopessa dialect with a set of types \(\mc{T}\) and instructions \(\mc{I}\). We'll need:
-\begin{itemize}
-  \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot}, equipped with
-  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies \ms{base}(X) \in \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies \ms{base}(X) \in \ms{Rel}(\mc{C})\)
-  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\) to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
-\end{itemize}
-We begin by giving a semantics for types and contexts in terms of objects in \(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}; this allows us to give a semantics for structural judgements in Figure \ref{fig:ssa-structural-semantics}.
-
-\TODO{some text about what the semantics actually are...}
-
-We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics} followed by a semantics for blocks and label-sets in Figure \ref{fig:ssa-block-semantics}.
-
-\TODO{ditto}
-
-\TODO{rename blocks to statements? label-sets to CFGs?}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{A}}: |\mc{C}| \\
-    \dnt{X} = \ms{base}(X) 
-    \qquad
-    \dnt{\mb{1}} = \mb{1}
-    \qquad
-    \dnt{\mb{2}} = \mb{2} = \mb{1} + \mb{1}
-    \qquad \dnt{A \otimes B} = \dnt{A} \otimes \dnt{B} \\
-    \boxed{\dnt{\Gamma}}: |\mc{C}| \\
-    \dnt{\cdot} = \mb{1} \qquad 
-    \dnt{\Gamma, \thyp{x}{A}{q}} = \dnt{\Gamma} \otimes \dnt{A} \\
-    \boxed{\dnt{\ms{L}}}: |\mc{C}| \\
-    \dnt{\cdot} = \mb{0} \qquad
-    \dnt{\ms{L}, \llhyp{\ell}{\Gamma}{A}} = 
-      \dnt{\ms{L}} + \dnt{\Gamma} \otimes \dnt{A}
-  \end{gather*}
-  \caption{Semantics for \isotopessa types and contexts}
-  \Description{Semantics for isotope-SSA types and contexts}
-  \label{fig:ssa-type-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{\tlin{A}{\taff}: \mc{C}_1^\tint(\dnt{A}, \mb{1})}}
-      \\
-      \dnt{\tlin{A}{\taff}} = \ms{drop}(\dnt{A})
-      % \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
-      % \dnt{\tlin{A \otimes B}{\taff}} 
-      %   = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
-      %   ; \lambda 
-      \\
-    \boxed{\dnt{\tlin{A}{\trel}}
-      : \mc{C}_1^\tint(\dnt{A}, \dnt{A} \otimes \dnt{A})}
-      \\
-      \dnt{\tlin{A}{\trel}} = \ms{split}(\dnt{A})
-      % \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
-      % \dnt{\tlin{A \otimes B}{\trel}} = 
-      %   \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
-      %   ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
-      \\
-    \boxed{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})}
-      \\
-    \dnt{\csplits{\cdot}{\cdot}{\cdot}} = \lambda^{-1} \qquad
-    \dnt{\csplits
-      {\Gamma, \thyp{x}{A}{q}}
-      {\Delta, \thyp{x}{A}{q}}
-      {\Xi}} 
-      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha;\dnt{\Delta} \otimes \sigma;\alpha 
-      \\
-    \dnt{\csplits
-      {\Gamma, \thyp{x}{A}{q}}
-      {\Delta}
-      {\Xi, \thyp{x}{A}{q}}} 
-      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha 
-      \\
-    \dnt{\csplits
-      {\Gamma, \thyp{x}{A}{q}}
-      {\Delta}
-      {\Xi, \thyp{x}{A}{q}}} \
-      = \dnt{\Gamma} \otimes \dnt{\tlin{A}{\trel}};
-        \alpha;
-        \dnt{\Delta} \otimes \sigma \otimes \dnt{A};
-        \alpha 
-        \\
-    \dnt{\csplits
-      {\Gamma, \thyp{x}{A}{q}}
-      {\Delta}
-      {\Xi}}
-      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-        \otimes \dnt{\tlin{A}{\taff}}
-      ; \rho 
-      \\
-    \boxed{\dnt{\cwk{\Gamma}{\Delta}}
-      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta})} 
-      \\
-      \dnt{\cwk{\Gamma}{\Delta}} 
-      = \dnt{\csplits{\Gamma}{\Delta}{\cdot}};\rho 
-      \\
-    \boxed{\dnt{\lwk{\ms{L}}{\ms{K}}}
-      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} 
-      \\
-    \dnt{\lwk{\cdot}{\cdot}} = \ms{id} \qquad
-    \dnt{\lwk
-      {\ms{L}, \llhyp{\ell}{\Gamma}{A}}
-      {\ms{K}, \llhyp{\ell}{\Delta}{A}}}
-      = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\cwk{\Gamma}{\Delta} \otimes \dnt{A})}{} 
-      \\
-    \dnt{\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      = \dnt{\lwk{\ms{L}}{\ms{K}}};\rho_+^{-1}
-      ; \dnt{\ms{K}} \oplus 0_{\dnt{\Gamma} \otimes \dnt{A}}
-  \end{gather*}
-  \caption{Semantics for \isotopessa structural judgements}
-  \Description{Semantics for isotope-SSA structural judgements}
-  \label{fig:ssa-structural-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{\hasty{\Gamma}{p}{a}{A}{q}}
-      : \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{x}{A}{q}} 
-      = \dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}
-      \qquad
-    \dnt{\hasty{\Gamma}{p}{f\;a}{B}{q}}
-      = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{p}
-      ; \ms{inst}_p(f) 
-      \\
-    \dnt{\hasty{\Gamma}{p}{(a, b)}{A \otimes B}{q}}
-      = \upg{(
-        \dnt{\csplits{\Gamma}{\Delta}{\Xi}};
-        \dnt{\hasty{\Delta}{1}{a}{A}{q}} \otimes
-        \dnt{\hasty{\Xi}{1}{b}{B}{q}}
-      )}{\uparrow p} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{()}{\mb{1}}{q}}
-      = \upg{\dnt{\cwk{\Gamma}{\cdot}}}{p} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{\ctt}{\mb{2}}{q}}
-      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\ctt)}{p}
-      \qquad
-    \dnt{\hasty{\Gamma}{p}{\cff}{\mb{2}}{q}}
-      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\cff)}{p} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{\letexpr{x}{a}{e}}{B}{q}}
-      = \upg{(
-        \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{q}}
-      )}{p};\dnt{\hasty{\Delta, \thyp{x}{A}{}}{p}{e}{B}{q}} 
-      \\
-    \dnt{\hasty{\Gamma}{p}{\letexpr{(x, y)}{a}{e}}{C}{q}}
-      = \upg{(
-        \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-        ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A \otimes B}{q}}
-      )}{p}
-      ; 
-      \\ \qquad \qquad \qquad \qquad \qquad \alpha
-      ; \dnt{\hasty{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{e}{C}{q}}
-    %   \\
-    % \dnt{\hasty{\Gamma}{0}{\lbsplice{\ell}{A}{t}}{A}{\varnothing}}
-    %   = \dnt{\haslb{\Gamma}{t}{\llhyp{\ell}{\cdot}{A}}};\alpha_+
-  \end{gather*}
-  \caption{Semantics for \isotopessa terms}
-  \Description{Semantics for isotope-SSA terms}
-  \label{fig:ssa-term-semantics}
-\end{figure}
-
-\TODO{some text about term semantics}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      : \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})} \\
-    \dnt{\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}}
-      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{\tint}})}
-      ; \dnt{\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}} \\
-    \dnt{\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}}
-      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\hasty{\Delta}{1}{e}{\mb{2}}{\tint}} \otimes \dnt{\Xi})}
-      ; \ms{ite}_{\dnt{\Xi}} ;
-      \\ \qquad \qquad \qquad \qquad
-      [
-        \dnt{\haslb{\Xi}{s}{\ms{L}}}, 
-        \dnt{\haslb{\Xi}{t}{\ms{L}}} 
-      ]
-      \\
-    \dnt{\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}}
-      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A}{q}}
-      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
-      \\
-    \dnt{\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}}
-      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A \otimes B}{q}}
-      ; \alpha
-      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
-      \\
-    \dnt{\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}}
-      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      ; \dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger
-      \\
-    \boxed{\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}
-      : \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
-      \\
-    \dnt{\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
-      = \dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      ; [[\iota_0, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}], \iota_1]
-      \\
-    \dnt{\lhaslb{\ms{L}}{\cdot}{\ms{L}}}
-      = \iota_0
-  \end{gather*}
-  \caption{Semantics for \isotopessa blocks}
-  \Description{Semantics for isotope-SSA blocks}
-  \label{fig:ssa-block-semantics}
-\end{figure}
-
-\TODO{some text about block semantics}
-
-\subsection{Semantic Metatheory}
-
-\TODO{some text about semantic metatheory}
-
-\begin{lemma}[Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\TODO{some text about coherence, implicit use above}
-
-\begin{lemma}[Semantic Upgrade]
-  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
-  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
-\end{lemma}
-
-\TODO{some text about upgrade}
-
-\begin{lemma}[Semantic Weakening] 
-  If \(\cwk{\Gamma}{\Delta}\), then 
-  \begin{itemize}
-    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
-    \[
-      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
-      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-    \]
-    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
-    \[
-      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
-      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
-    \]
-  \end{itemize}
-
-  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
-  \[
-    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
-    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
-  \]
-\end{lemma}
-
-\TODO{some text about semantic weakening}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_1^\varnothing(\dnt{\Theta}, \dnt{\Gamma})} 
-    \\
-    \dnt{\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
-      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{1}{a}{A}{q}} \\
-    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id}
-    \\
-    \boxed{\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}: \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}})} 
-  \end{gather*}
-  \caption{Semantics for \isotopessa substitutions}
-  \Description{Semantics for isotope-SSA substitutions}
-  \label{fig:ssa-subst-semantics}
-\end{figure}
-
-\TODO{some text about substitution semantics}
-
-\begin{lemma}[Substitution Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\TODO{pull down to appendix? just trivial induction.}
-
-\begin{lemma}[Semantic Substitution Splitting] 
-  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
-  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
-  \[
-    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
-      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
-      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
-  \]
-\end{lemma}
-
-\begin{theorem}[Semantic Substitution] 
-  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  \begin{itemize}
-    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
-    \[
-      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
-      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
-    \]
-    \item For all 
-      \(\haslb{\Gamma}{t}{\ms{L}}\), 
-      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
-      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
-    \[
-      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
-      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
-      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
-    \]
-    \item For all 
-      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
-      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
-      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
-      \[
-        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
-        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
-        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
-        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-      \]
-  \end{itemize}
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
-  \begin{itemize} 
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
-    \[
-      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
-    \]
-    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), we have
-    \[
-      \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      = \dnt{\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{L}}}
-    \]
-  \end{itemize}
-\end{theorem}
-
-\TODO{some text about semantic substitution}
-
-\begin{figure}
-  \begin{align*}
-    \dnt{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}
-      &= \dnt{H} \times \mc{C}_p^q(\dnt{\Gamma}, \dnt{A}) 
-      \\
-    \dnt{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}
-      &= \dnt{H} \times \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}}) 
-      \\
-    \dnt{H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
-      &= \dnt{H} \times \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
-      \\
-    \dnt{\cdot} 
-      &= \mb{1}
-  \end{align*}
-  \caption{Semantics for \isotopessa holes}
-  \Description{Semantics for isotope-SSA holes}
-  \label{fig:holes-semantics}
-\end{figure}
-
-\TODO{some text about semantics of holes}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}}
-      : \dnt{H} \to \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})}
-      \\
-    \sorry
-  \end{gather*}
-  \caption{Semantics for \isotopessa terms with holes}
-  \Description{Semantics for isotope-SSA blocks with holes}
-  \label{fig:terms-with-holes-semantics}
-\end{figure}
-
-\TODO{fill}
-
-\TODO{some text about semantics of terms with holes}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
-      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})}
-      \\
-    \sorry
-      \\  
-    \boxed{
-      \dnt{\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
-      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
-      \\
-    \sorry
-  \end{gather*}
-  \caption{Semantics for \isotopessa blocks with holes}
-  \Description{Semantics for isotope-SSA blocks with holes}
-  \label{fig:blocks-with-holes-semantics}
-\end{figure}
-
-\TODO{fill}
-
-\TODO{some text about semantics of blocks with holes}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\isrw{\mc{H}}{I}{H}}
-      : \dnt{I} \to \dnt{H} 
-    }
-    \\
-    \dnt{\isrw{\mc{H}}{I}{\cdot}}\;i = ()
-    \qquad
-    \dnt{\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}\;i
-    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}\;i)
-    \\
-    \dnt{\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
-    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}\;i)
-    \\
-    \dnt{\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
-    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}\;i)
-  \end{gather*}
-  \caption{Semantics for \isotopessa rewrites}
-  \Description{Semantics for isotope-SSA rewrites}
-  \label{fig:rewrite-semantics}
-\end{figure}
-
-\TODO{fill}
-
-\TODO{some text about semantics of rewrites}
-
-\begin{lemma}[Rewriting Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\begin{theorem}[Semantic Rewriting]
-  Given \(\isrw{\mc{H}}{I}{H}\), we have that
-  \begin{itemize}
-    \item For all \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(m \in \dnt{I}\), we have
-    \[
-      \dnt{\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}}(m)
-      = \dnt{\mhasty{H}{\Gamma}{p}{a}{A}{q}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
-    \]
-    \item For all \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(m \in \dnt{I}\), we have
-    \[
-      \dnt{\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}}(m)
-      = \dnt{\mhaslb{H}{\Gamma}{t}{\ms{L}}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
-    \]
-  \end{itemize}
-
-  Furthermore, we have that
-  \[
-    \dnt{\mhasty{\cdot}{\Gamma}{p}{a}{A}{q}}() = \dnt{\hasty{\Gamma}{p}{a}{A}{q}}
-    \qquad
-    \dnt{\mhaslb{\cdot}{\Gamma}{t}{\ms{L}}}() = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-  \]
-\end{theorem}
-
-\subsection{Normalization}
-
-\label{ssec:ssa-norm}
-
-\TODO{this}
-
-\subsection{Optimizations}
-
-\TODO{this:}
-
-\TODO{substitution \(\implies\): loop hoisting, global value numbering, some dead code elimination}
-
-\TODO{rewriting \(\implies\) strength reduction, E-graph optimization, see \cite{cranelift}}
-
-\TODO{inlining}
-
-\TODO{loop fusion}
-
-\TODO{unreachable code elimination}
-
-\TODO{more dead code elimination}
-
-\section{Concrete Models}
-
-In this section, we will build on the bare-bones models given in Section \ref{sec:semantics} to develop more sophisticated models capable of underpinning \isotopessa dialects supporting features such as weak memory and separation logic.
-
-\subsection{Trace Models}
-
-We will begin by giving a useful foundation for a wide class of \isotopessa models: the "trace models" over \(\mb{Set}\), which accumulate a (potentially infinite) nondeterministic trace of monoidal "events".
-These will later come in handy for reasoning about output behaviour and weak memory.
-
-We begin with some definitions:
-\begin{definition}[Stream Action]
-  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to I\) is a map from streams of \(M\) to \(I\) satisfying
-  \(
-    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
-  \),
-  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the stream \(\ell\).
-\end{definition}
-Some basic examples include:
-\begin{enumerate}
-  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\) forms a stream action on the one element set \(\mathbf{1}\).
-  \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta = \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
-  \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta = \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i \cdot_i l_i)_i\).
-  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite) sequence \(s\).
-  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts on \(\nats \cup \{\infty\}\) by addition.
-\end{enumerate}
-We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted \(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\).
-For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq \omega}) = Z(A^*) = \{[\,]\}\), while \(\mc{SZ}(\nats, \nats \cup \{\infty\}) = \{0\} \subseteq Z(\nats) = \nats\).
-Generally, \(\mc{SZ}(M, I)\) always contains the unit of \(M\).
-
-We can now define the \textit{trace monad} of a stream action as follows:
-\begin{definition}[Trace Monad]
-  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A \times M + I)\), with
-  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
-  and 
-  \[\mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}\]
-  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with iteration operator
-  \(f^\dagger(a) = f^\infty(a) \cup \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t \in f_n(a)\}\), where
-  \[
-    f_0 = f, \qquad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n]
-  \]
-  \[
-    f^\infty(a) = \{\Theta (\lambda n. m_n) \mid a_0 = a \land \forall n, \exists a_{n + 1}. \iota_0 (\iota_1\;a_{n + 1}, m_n) \in f(a_n)\}
-  \]
-  \TODO{clean this up}
-\end{definition}
-
-\TODO{text}
-
-\begin{definition}[Relevant, Affine Monad]
-  A monad \(\mb{T}\) on \(\ms{Set}\) is \textbf{relevant} if its Kleisli
-  category satisfies \(f;\ms{split} = \ms{split};f \ltimes f = \ms{split};f
-  \rtimes f\), where \(\ms{split} = \upg{\lambda a. (a, a)}{}\). \(\mb{T}\) is
-  \textbf{affine} if its Kleisli category satisfies \(f;\ms{drop} = \ms{drop}\),
-  where \(\ms{drop} = \upg{\lambda a.()}{}\). A monad is \textbf{intuitionistic}
-  if it is both relevant and affine.
-\end{definition}
-
-\begin{definition}[Submonad] 
-  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a
-  collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq
-  \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in
-  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say
-  that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and,
-  for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where
-  \(f^\dagger\) is computed as in \(\mb{T}\). Note that \(\mb{S}\) is also an
-  (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
-  from \(\mb{T}\).
-\end{definition}
-Some important examples of submonads include:
-\begin{enumerate}
-  \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad
-  \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\);
-  similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
-  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
-  intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
-  submonads, forming a complete meet-semilattice with bottom element
-  \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete
-  lattice in the standard manner.
-  \item In particular, the intersection of a relevant/affine/commutative
-  submonad and another submonad is relevant/affine/commutative respectively.
-  Since every commutative submonad induces a central subcategory of the Kleisli
-  category of the parent monad, it follows that a triple of a relevant, affine,
-  and commutative submonad make the Kleisli category of a monad \(\mb{T}\) into
-  a substructural effectful category with all objects both affine and relevant.
-  \item If \(C \subseteq Z(M)\) is submonoid of \(M\), then \(A \mapsto
-  \mc{P}^+(A \times C + \mb{0}), \mc{P}^+(A \times C \times I) \subseteq
-  \ms{Traces}\;\Theta\;A\) are submonads
-  \item In particular, if \(M_C \subseteq Z(M)\) is a central submonoid of \(M\)
-  such that \(\forall m \in M_C, \forall i \in I, m \cdot i = i\), then \(A
-  \mapsto \mc{P}^+(A \times M_C + \mb{0}) \subseteq \ms{Traces}\;\Theta\;A\) is
-  a submonad whose Kleisli category is a central subcategory of the Kleisli
-  category of \(\ms{Traces}\;\Theta\;A\). Note \(M_C = \{1\}\) is always such a
-  submonoid.
-  \label{item:central-submonad}
-  \item Similarly, if \(M_R \subseteq M_C\) is a submonoid such that \(\forall m
-  \in M_R, m \cdot m = m\), then \(A \mapsto \{S \in \mc{P}^+(A \times M_R + I)
-  \mid |S| = 1\} \subseteq \ms{Traces}\;\Theta\;A\) is a \textit{relevant}
-  submonad. Note \(M_R = \{1\}\) is always such a submonoid.
-  \label{item:relevant-submonad}
-  \item Finally, \(A \mapsto \mc{P}^+(A \times \{1\} + \mb{0}) \subseteq
-  \ms{Traces}\;\Theta\;A\) is an \textit{affine} submonad.
-  \label{item:affine-submonad}
-\end{enumerate}
-In particular, \ref{item:central-submonad}, \ref{item:relevant-submonad},
-\ref{item:affine-submonad} equip the Kleisli category of
-\(\ms{Traces}\;\Theta\;A\) with the structure of a substructural effectful
-category given choices of \(M_C\), \(M_R\).
-
-\subsection{Heaps, Printing, and Separation Logic}
-
-\label{ssec:separation}
-
-As a concrete example, we can now introduce the \textbf{printing monad}, which
-is simply defined as \(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where
-\(\Sigma: (\ms{Bytes}^*)^\omega \to \ms{Bytes}^{\leq \omega}\) denotes
-concatenation of lists of bytestrings. This can be equipped with the structure
-of a substructural effectful category by choosing \(M_C = M_R = \{[]\}\), giving
-us an \isotopessa model supporting the instructions \(\ms{print} \in
-\mc{I}^\varnothing_0(\ms{Bytes}, \mb{1})\) and \(\ms{rand} \in
-\mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard semantics
-\(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), [b])\}\), \(\dnt{\ms{rand}} =
-\lambda (). \{\iota_0 (a, p[]) \mid a \in A\}\).
-
-We can make things a little more interesting by adding a heap to the mix,
-defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where
-\(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function. The
-Kleisli category of this monad can be equipped with the structure of a
-substructural effectful category by simply taking the affine/relevant morphisms
-to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\);
-as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we
-therefore have an \isotopessa model supporting the instructions \(\ms{set}:
-\mc{I}^\varnothing_0(\nats \times \nats, \mb{1})\), \(\ms{get}:
-\mc{I}^\varnothing_0(\nats, \nats)\), \(\ms{alloc}: \mc{I}^\varnothing_0(\nats,
-\nats)\), and \(\ms{free}: \mc{I}^\varnothing_0(\nats, \mb{1})\). We can assign
-semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} =
-\lambda (p, v)\;h. \{\iota_0((), [p \mapsto v]h, [])\}\). \(\ms{get}\) is a bit
-more tricky, since it is unclear what to do when we try to access uninitialized
-memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}}
-= \lambda p\;h. \{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\); for now,
-this will do. Finally, \(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', [])
-\mid h' = h \sqcup p \mapsto v\}\) simply fills a random empty heap cell with
-the provided value, returning a pointer to the cell, while \(\dnt{\ms{free}} =
-\lambda p\;h. \{\iota_0\;((), h \setminus p, [])\}\)
-
-We can use this construction to, building on the ideas in \cite{mellies-ftrs},
-show how a simple separation logic based on \cite{reynolds-separation-2002}
-slightly adapted to the categorical setting gives us a very rich \isotopessa
-model supporting a basic notion of refinement types, while at the same time
-giving us a nontrivial notion of nonlinear types.
-
-Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of
-type \(A \to \ms{Heap} \to \ms{Prop}\). Taking as objects pairs \((A, \Phi)\),
-we can define the category \(\ms{Sep}\) to have morphisms
-\begin{equation}
-  \ms{Sep}((A, \Phi), (B, \Psi)) = \{f: A \to \ms{Comp}\;B 
-    \mid \forall \phi: \ms{Heap} \to \ms{Prop}, \ms{hoare}\;f\;(\Phi \ast \phi)\;(\Psi \ast \phi)\}
-\end{equation}
-where
-\begin{equation}
-  \ms{hoare}\;f\;\Phi\;\Psi = \forall a\;h. \Phi\;a\;h \implies \forall \iota_0\;(b, h') \in f\;a\;h. \Psi\;b\;h'
-\end{equation}
-is the usual partial correctness triple definition and the separating
-conjunction of heap-predicates is defined as usual to be 
-\begin{equation}
-  \begin{gathered}
-  \varphi \ast \psi = \lambda h. \exists h_1, h_2. h = h_1 \sqcup h_2 \land \varphi h_1 \land \psi h_2 
-  \qquad
-  \varphi \ast \Phi = \lambda a. \varphi \ast \Phi\;a
-  \qquad
-  \Phi \ast \varphi = \lambda a. \Phi\;a \ast \varphi \\
-  \Phi \ast \Psi = \lambda (a, b). \Phi\;a \ast \Psi\;b: A \times B \to \ms{Heap} \to \ms{Prop}
-  \end{gathered}
-\end{equation}
-We can naturally define a tensor product on objects \((A, \Phi) \otimes (B,
-\Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking
-\begin{itemize}
-  \item Monoidal unit \((\mb{1}, \ms{emp})\), where \(\ms{emp}\) is the
-  predicate which holds only for the empty heap
-  \item Associators and inverse associators given by \(\alpha_{(A, \Phi), (B,
-  \Psi), (C, \Theta)} = \alpha_{A, B, C}\), which are valid since \(\ast\) is
-  associative
-  \item Unitors and inverse unitors given by \(\lambda_{(A, \Phi)} =
-  \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\), which are valid since \(\ms{emp}\)
-  is the unit of \(\ast\)
-  \item \((C, \Theta) \otimes f = C \times f\) and \(f \otimes (C, \Theta) = f
-  \times C\), which are valid definitions by the frame rule
-\end{itemize}
-equips \(\ms{Sep}\) with the structure of a premonoidal category. We can give \(\ms{Sep}\) the structure of a substructural effectful category by taking:
-\begin{itemize}
-  \item Affine objects to be precisely those for which \(\ms{drop} =
-  \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), which for
-  this logic corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend
-  on the heap.
-  \item Relevant objects to be precisely those for which \(\ms{split} =
-  \upg{\lambda a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), again in
-  this case those for which \(\Phi\) does not depend on the heap.
-  \item Central morphisms to be those which do not print, diverge, \textit{or} modify the
-  heap. We will consider how to relax this condition later.
-  \item Relevant and affine morphisms to be the lifts of relevant/affine morphisms in \(\ms{Print}\)
-\end{itemize}
-\(\ms{Sep}\) also has coproducts given by \((A, \Phi) + (B, \Psi) = (A + B, [\Phi, \Psi])\), which allow us to equip it with an Elgot structure by taking advantage of the fact that
-\begin{equation}
-  f \in \ms{Sep}((A, \Phi), (B + A, [\Psi, \Phi])) \implies f^\dagger \in \ms{Sep}((A, \Phi), (B, \Psi))
-\end{equation}
-This implies that \(\ms{Sep}\) is an \isotopessa model supporting the following interesting operations
-\begin{itemize}
-  \item \(\ms{coe}: \mc{I}^\tint_1((A, \Phi), (A, \Psi))\) where \(\forall a,
-  \Phi\;a \implies \Psi\;a\), implemented by the identity morphism
-  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Bytes}, \ms{emp}), (\mb{1},
-  \ms{emp}))\)
-  \item \(\ms{rand}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for \(A\) nonempty
-  \item \(\ms{set}: \mc{I}^\varnothing_0((\nats \times \nats, \lambda (\ell,
-  \cdot). \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell
-  \mapsto v))\), where \(\ms{set}\) is modified slightly to return the address
-  at which the value passed in was stored to overcome a lack of dependency
-  between the source and target objects.
-  \item \(\ms{get}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)
-  \item \(\ms{alloc}: \mc{I}^\varnothing_0((\nats, \ms{emp}), (\nats, \lambda
-  \ell. \exists v. \ell \mapsto v))\)
-  \item \(\ms{free}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\mb{1}, \ms{emp}))\)
-\end{itemize}
-We can clearly see that this is a refinement type system in the sence of
-\cite{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi)
-\mapsto A\) which is simply the identity on morphisms. Unfortunately, this means
-that our notion of equality is a little too stringent, since we need to consider
-equality on inputs which do not match our preconditions. Thankfully, we can fix
-this problem quite easily by quotienting morphisms over the equivalence relation
-\begin{equation}
-  \forall f, g: \ms{Sep}((A, \Phi), (B, \Psi)).
-  f \simeq g \iff \forall a, h. \Phi\;a\;h \implies f\;a\;h = g\;a\;h
-\end{equation}
-yielding the new category \(\ms{SComp}\). In particular, this allows us to widen
-our definition of central morphisms to include all morphisms which do not print
-or diverge, including in particular \(\ms{set}, \ms{get}, \ms{alloc}\) and
-\(\ms{free}\), which we can in fact mark relevant and affine as well, allowing
-us, for example, to make use of intermediate allocations in an otherwise pure
-computation.
-
-\subsection{Weak Memory}
-
-Another interesting use of the trace monad to build a sophisticated \isotopessa model is to build a model of TSO-style weak memory based on \citet{sparky}.
-For simplicity, we will consider an infinite set of named, fixed locations \(x, y, z \in \ms{Loc}\) which are subject to concurrent modification by all processors via TSO reads, writes, and fences.
-
-We begin with the following definitions:
-\begin{definition}[Pomset] 
-  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
-  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
-  \textit{nonempty} partially-ordered \textbf{carrier set} \(P\) such that every
-  \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
-  \to \mc{A}\) quotiented under the equivalence relation \(\alpha: P \to \mc{A}
-  \simeq \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\)
-  from \(P\) and \(Q\), there exists a partial order isomorphism between them
-  which respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite}
-  if its carrier set is. We will write unordered pomsets using multiset
-  notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list
-  notation, e.g. \([a, a, b]\).
-\end{definition}
-(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\),
-which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to
-\mc{A}\), where \(P + Q\) is given the lexicographic ordering and
-\(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many
-predecessors. They also form a monoid under parallel composition \(\alpha ||
-\beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\)
-where \(P + Q\) is given the standard partial ordering (with elements of \(P\)
-and \(Q\) incomparable). Note in both cases the monoidal unit is \(\{\delta\}\),
-since we only consider nonempty pomsets but allow the removal of finitely many
-ticks.
-
-Given a partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n \in N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by the function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\), where the dependent product \(\Sigma_nP_n\) is given the lexicographic order.
-In particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega \to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on finite pomsets.
-
-We define a \textbf{program order pomset} to be a pomset with \(\mc{A}_{\ms{PO}} =
-\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
-\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
-\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
-\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
-\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
-program order pomsets given by sequential composition for \(\mc{A}\). This
-yields an \isotopessa model with support for concurrent read and write
-operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
-\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
-semantics
-\begin{equation}
-  \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
-  \qquad
-  \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
-\end{equation}
-The semantics of \(\ms{read}\) in particular give a hint as to how this model
-works: rather than tracking the state of the heap, we simply \textit{emit} a
-pomset of the events that would have to happen for a given execution (in the
-case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
-the case of a write, the single write event \(x := v\) for a write of \(v\)),
-and later post-filter on the pomset of an entire program to get a sequentially
-consistent execution. Hence, our semantics remains compositional, allowing us to
-reason about each program fragment individually, while still allowing us to add
-arbitrary external state (e.g. another program concurrently modifying the heap)
-before finally considering the execution of an entire program.
-
-On that note, we support the parallel execution of multiple (potentially
-nonterminating) morphisms as follows:
-\begin{equation}
-  \begin{aligned}
-  f_0 || f_1 = \lambda (a_0, a_1). 
-  & \{\iota_0 ((b_0, b_1), \alpha_0 || \alpha_1) 
-    \mid \iota_0 (b_i, \alpha_i) \in f_i\;a_i\} 
-  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0) \land \iota_1 \alpha_1 \in f_1\;a_1\} 
-  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0 \land \iota_0 (a_1, \alpha_1) \in f_1\;a_1\}: \ms{Set}_{\ms{PO}}(A \otimes A', B \otimes B')
-  \end{aligned}
-\end{equation}
-It's tempting to try to use this as a tensor product of morphisms and so
-graduate to a bona fide monoidal category rather than our current premonoidal
-setting, since spatial concurrency is one of the headline use-cases of the
-former, but unfortunately, sliding still fails:
-\(
-  (W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})
-  \neq (\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}} || \ms{id})
-\),
-since the left-hand side will emit the pomset \([x := a, y := b]\) while the
-right-hand side will emit the pomset \([y := b, x := a]\). If we did want this
-behaviour, we'd need to have a significantly more sophisticated model of
-composition which effectively takes nontermination into account, which we will
-leave to future work.
-
-To graduate from SC concurrency to a genuine, if maximally simple, weak memory
-model, we will to introduce a \textit{buffer} of write actions. We will
-implement TSO ordering by buffering all our writes, in which case they are only
-visible to the local thread. On the other hand, reads events will first attempt
-to read from the buffer, and, if there is no corresponding write in the buffer,
-will read an arbitrary value, in both cases pushing an event to the global
-pomset. At any point, we may choose to \textit{flush} some of the writes in
-buffer to the global pomset of events. In particular, we will introduce the set
-\(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write actions}, where an
-action \(\bufloc{x} := v\) by a thread denotes adding a write \(x := v\) to the
-thread's write buffer. We may then define the set of TSO actions
-\(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be defined
-to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret
-as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index (with the empty list corresponding to \(\{\delta\}\)). In particular, we
-define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
-where \(\Sigma\) is the stream action of finite pomsets over
-\(\mc{A}_{\ms{TSO}}\) on pomsets over \(\mc{A}_{\ms{TSO}}\); we can view
-\(\ms{PO}\) as a submonad of this monad in the standard manner.
-
-Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
-\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
-\begin{equation}
-  (L;\{\bufloc{x} := v\})[x] = v
-  \qquad
-  (L;\{\bufloc{y} := v\})[x] = L[x]
-  \qquad
-  [][x] = \bot
-\end{equation}
-Note that writes at the \textit{end} of the buffer are prioritized, since later
-writes overwrite earlier ones! The semantics of reads and writes are then given
-by
-\begin{equation}
-  \begin{aligned}
-  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
-    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
-  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
-    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
-  \end{aligned}
-\end{equation}
-where buffer flushing is implemented via the morphism (called \(\ms{split}\) in
-\cite{sparky})
-\begin{equation}
-  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
-  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
-\end{equation}
-To be able to perform synchronization, we will also need to introduce a \(\ms{fence}: \mc{I}^\varnothing_0(\mb{1}, \mb{1})\) instruction, which simply causes all actions before the fence to be observed before any actions after the fence.
-For \(\ms{TSO}\), implementing this is as simple as flushing the buffer, giving
-\begin{equation}
-  \dnt{\ms{fence}} = \lambda ()\;L. \{((), [], L;\{\delta\})\}
-\end{equation}
-We can now define the parallel composition of morphisms in a "fork-join" style as follows: we first flush the buffer completely (i.e., taking it and sticking it at the beginning of our pomset), then execute \(f\) and \(g\) in parallel with separate buffers, \textit{filtering out executions which completely flush the buffer}.
-Since both threads end up with an empty buffer, the resulting joined buffer is also empty, giving us the following:
-\begin{equation}
-  \begin{aligned}
-    f_0 || f_1 = \lambda (a_0, a_1)\;L. 
-    & \{\iota_0 ((b_0, b_1), [], L;(\alpha_0 || \alpha_1)) 
-      \mid \iota_0 (b_i, [], \alpha_i) \in f_i\;a_i\;[]\} 
-    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, [], \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0\;[]) \land \iota_1 \alpha_1 \in f_1\;a_1\;[]\} 
-    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0\;[] \land \iota_0 (a_1, [], \alpha_1) \in f_1\;a_1\;[]\}
-    \end{aligned}
-\end{equation}
-
-\section{Related Work}
-
-\TODO{\cite{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
-
-\TODO{\cite{schneider-phd}}
-
-\TODO{\cite{chakravarty-functional-ssa-2003}}
-
-\TODO{\cite{barthe-compcert-ssa-2014}}
-
-\TODO{\cite{buchwald-verified-ssa-2016}}
-
-\TODO{\cite{herklotz-gsa-2023}}
-
-\TODO{\cite{pop-ssa-inout-2009}}
-
-\TODO{\cite{garbuzov-structural-cfg-2018}}
-
-\TODO{\cite{hua-explicit-ssa-2010}}
-
-\TODO{\cite{rvsdg}}
-
-\TODO{this}
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{references}
-
-\clearpage
-
-\appendix
-
-\section{Appendix}
-
 \end{document}

### 2023-11-14 pldi24/paper.tex
Tap at definitions

@@
 
-\begin{definition}[Binoidal Category] \
-  \TODO{this}
+\begin{definition}[Binoidal Category] 
+  A \textbf{binoidal category} \(\mc{C}\) is a category equipped with a binary
+  operation on objects \(\otimes: \mc{C} \times \mc{C} \to \mc{C}\) which is
+  independently functorial in each argument (i.e., for each object \(A\), \(A
+  \otimes -\) and \(- \otimes A\) are functors).
 \end{definition}
-
-\begin{definition}[Premonoidal Category] \
-  \TODO{this}
+Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
+write \(f \ltimes g = f \otimes A';B \otimes g\) and \(f \rtimes g = A \otimes
+g;f \otimes B'\); we say \(f\) is \textbf{central} if, for all \(g\), \(f
+\ltimes g = f \rtimes g\) and \(g \ltimes f = g \rtimes f\), in which case we
+simply write \(f \otimes g\), \(g \otimes f\) respectively.
+
+\TODO{text}
+
+\begin{definition}[Premonoidal Category] 
+  A \textbf{premonoidal category} is a binoidal category with an identity object
+  \(I: \mc{C}\) such that
+  \begin{itemize}
+    \item There exist central natural isomorphisms \(\lambda_A: I \otimes A
+    \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I \simeq A\) (the
+    \textbf{right unitor})
+    \item There exists a central isomorphism \(\alpha_{A, B, C}: (A \otimes B)
+    \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
+    \item Satisfying the \textbf{pentagon law}: \todo{this}
+    \item Satisfying the \textbf{triangle law}: \todo{this}
+  \end{itemize}
+  We say a premonoidal category is \textbf{symmetric} if it is equipped with a
+  central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
+  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} \todo{this}.
+
+  We say a premonoidal category is \textbf{monoidal} if every morphism is
+  central.
 \end{definition}
+
+\TODO{text}
 
@@
 \end{definition}
+
+\TODO{text}
 
@@
 
-\begin{definition}[Strictly Effectful Category] \
-  \TODO{this}
-\end{definition}
-
-\TODO{text}
+% Should we go down this road?
+% \begin{definition}[Strictly Effectful Category] \
+%   \TODO{this}
+% \end{definition}
+
+% \TODO{text}
 

### 2023-11-14 pldi24/paper.tex
Tap

@@
 strong enough to justify many other essential compiler optimisations. For
-example, hoisting a store out of a loop or conditional involves moving an
-effectful term (an assignment) out of its original position in the program. As
-another example, dead store elimination removes an effectful term from the
-program altogether. \todo{Add link to figure with code for these examples}
+example, hoisting a load out of a loop or conditional involves moving an
+effectful term out of its original position in the program (see Equation
+\ref{eqn:hoist}). As another example, dead store elimination removes an
+effectful term from the program altogether.
+
+%TODO: note that dead store elimination requires refinement, which we're not
+%covering in this paper...
 
@@
 \end{BVerbatim}
+\label{eqn:hoist}
 \end{equation}
@@
 \section{SSA Syntax}
+
+\TODO{standardize "branching to" wordmark? Is wordmark the word here?}
 
@@
 
+\TODO{intro text}
+
 \begin{theorem}[Weakening] \
@@
 \end{theorem}
+
+\TODO{text}
 
@@
         \(\issubst{\gamma}{\Theta}{\Gamma}\) &
-        ... \\
+        \(\gamma\) is a substitution taking \(\Theta\) to \(\Gamma\) \\
         \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) &
-        ... \\
-        \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) &
-        ... \\
-        \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) &
-        ... \\
-        \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) &
-        ... \\
-        \(\isrw{\mc{H}}{I}{H}\) &
-        ... \\
+        \(\mc{L}\) is a label substitution taking \(\ms{L}\) to \(\ms{K}\) \\
+        % \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) &
+        % ... \\
+        % \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) &
+        % ... \\
+        % \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) &
+        % ... \\
+        % \(\isrw{\mc{H}}{I}{H}\) &
+        % ... \\
     \end{tabular}

### 2023-11-14 pldi24/paper.tex
Tap

@@
 
-\TODO{intro text}
+In this section, we will give a categorical semantics for \isotopessa. Starting
+with Freyd categories, we will generalize to \textit{substructural effectful
+categories} to support substructurality. We will then require our substructural
+effectful category to have \textit{distributive coproducts} to implement
+branching. Finally, we will require an \textit{Elgot structure} to implement
+general, potentially nonterminating control flow.
 
@@
 
-\TODO{intro text}
-
+We begin with some basic definitions:
 \begin{definition}[Binoidal Category] 
@@
 simply write \(f \otimes g\), \(g \otimes f\) respectively.
-
-\TODO{text}
-
 \begin{definition}[Premonoidal Category] 
@@
   central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
-  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} \todo{this}.
-
-  We say a premonoidal category is \textbf{monoidal} if every morphism is
-  central.
+  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} \todo{this}. We
+  say a premonoidal category is \textbf{monoidal} if every morphism is central.
 \end{definition}
-
-\TODO{text}
-
-\begin{definition}[Effectful Category] \
-  \TODO{this}
+One issue with the notion of a premonoidal category is that it is unclear what
+exactly a premonoidal functor should be: if we want them to compose, we would
+like to preserve central morphisms, but trying to preserve \textit{all} central
+morphisms is too strict a notion. We want, in essence, to be able to distinguish
+morphisms that are truly "pure" from those which simply happen to be central. To
+do that, we introduce the notion of an \textit{effectful category} as follows:
+\begin{definition}[Effectful Category] 
+  An \textbf{effectful category} \(\mc{V} \to \mc{C}\) is an identity on objects
+  functor \(\upg{\cdot}{}: \mc{V} \to \mc{C}\) from a monoidal category
+  \(\mc{V}\) to a premonoidal category \(\mc{C}\) which preserves all
+  premonoidal structure (i.e. sends associators to associators, unitors to
+  unitors, and tensor products to tensor products). It is \textbf{symmetric} if
+  both \(\mc{V}\) and \(\mc{C}\) are, and \(\upg{\cdot}{}\) preserves the
+  symmetry. A \textbf{Freyd category} is an effectful category where \(\mc{V}\)
+  is Cartesian.
 \end{definition}
 
-\TODO{text}
+\TODO{text on Freyd categories}
 
@@
 
-% \TODO{text}
+\TODO{text}
 
@@
 Proofs can be found in Lemma 31 of \cite{goncharov-squaring}.
-
-\TODO{Hyland and Hasegawa traces?}
-
-\TODO{intuition? Can draw diagrams, but those take up space}
 

### 2023-11-14 pldi24/paper.tex
Tap

@@
   We will show that
-  \(\upg{\delta_{A, B, C}}{} = \delta_{A, B, C};\ms{pure}: \mc{C}_{\mb{T}}((A + B) \otimes C, A \otimes C + B \otimes C)\) satisfies the desired properties of a distributor. In particular, it is:
+  \(\upg{\delta_{A, B, C}}{}: \mc{C}_{\mb{T}}((A + B) \otimes C, A \otimes C + B \otimes C)\) satisfies the desired properties of a distributor. In particular, it is:
   \begin{itemize}
@@
 
-\section{Appendix}
+\section{Proofs}
+
+\TODO{these}
+
+\begin{lemma}[Substitution Coherence] \
+  \begin{itemize}
+    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{lemma}
+
+\begin{lemma}[Semantic Substitution Splitting] 
+  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
+  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
+  \[
+    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
+      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
+      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
+  \]
+\end{lemma}
+
+\section{Rewriting}
 
@@
 \end{theorem}
-
-\begin{lemma}[Substitution Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\TODO{pull down to appendix? just trivial induction.}
-
-\begin{lemma}[Semantic Substitution Splitting] 
-  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
-  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
-  \[
-    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
-      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
-      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
-  \]
-\end{lemma}
 

### 2023-11-14 pldi24/paper.tex
Tap

@@
   \end{subfigure}
-  \caption{A simple program for computing the \texttt{i}th Fibonacci number, represented as 3-address code, basic-blocks with arguments, and in our SSA syntax}
+  \caption{A simple program for computing the \texttt{i}th Fibonacci number,
+  represented as 3-address code, basic-blocks with arguments, and in our SSA
+  syntax} 
   \Description{}
@@
 
-\begin{theorem}[Weakening] \
-  \begin{itemize}
-    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\), \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
+\begin{lemma}[Weakening] \
+  \begin{itemize}
+    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
+    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
     \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
@@
   \end{itemize}
-\end{theorem}
+\end{lemma}
 
@@
   \end{gather*}
-  \caption{
-    Typing rules for \isotopessa substitutions. 
-    We generally assume that for all but finitely many variables \(x\) and labels \(\lbl{\ell}\), \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) = \lbrb{\ell}{x}\).
-  }
+  \caption{ Typing rules for \isotopessa substitutions. We generally assume that
+    for all but finitely many variables \(x\) and labels \(\lbl{\ell}\),
+    \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) = \lbrb{\ell}{x}\). }
     \Description{Typing rules for isotope-SSA substitutions}
@@
 \begin{lemma}[Substitution Splitting] 
-  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
+  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
   \begin{itemize}
@@
 
-\begin{theorem}[Substitution] 
+\begin{lemma}[Substitution] 
   Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
@@
     \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
-    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
-    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
+    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such
+    that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
+    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
+    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
+    \ms{W}'\) such that 
     \begin{itemize}
@@
 
-\begin{theorem}[Label Substitution] 
+\begin{lemma}[Label Substitution] 
   Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
@@
 \begin{definition}[(Symmetric) Premonoidal Subcategory]
-  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is closed under whiskering and contains all associators, unitors, and symmetries.
-We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\), \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is \textbf{central} if this generalizes to the case where only one of \(f, g\) is in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
+  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory
+  \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is
+  closed under whiskering and contains all associators, unitors, and symmetries.
+  We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\),
+  \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is
+  \textbf{central} if this generalizes to the case where only one of \(f, g\) is
+  in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
 \end{definition} 
-Note that the set of symmetric premonoidal subcategories, monoidal subcategories, and central subcategories of a category \(\mc{C}\) are all closed under intersections and therefore, in particular, form a complete meet-semilattice with shared bottom element \(\ms{Skeleton}(\mc{C})\), where \(\ms{Skeleton}(C)\) is the smallest possible symmetric premonodial subcategory.
-The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the largest central subcategory is the center of \(\mc{C}\) \(Z(\mb{C})\); the existence of a top-element means that in these cases we may induce the structure of a complete lattice in the usual manner.
+Note that the set of symmetric premonoidal subcategories, monoidal
+subcategories, and central subcategories of a category \(\mc{C}\) are all closed
+under intersections and therefore, in particular, form a complete
+meet-semilattice with shared bottom element \(\ms{Skeleton}(\mc{C})\), where
+\(\ms{Skeleton}(C)\) is the smallest possible symmetric premonodial subcategory.
+The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the
+largest central subcategory is the center of \(\mc{C}\) \(Z(\mb{C})\); the
+existence of a top-element means that in these cases we may induce the structure
+of a complete lattice in the usual manner.
 
@@
 \begin{definition}[Substructural (Pre)monoidal Category]
-  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A}, \mc{R})\) where:
-  \begin{itemize}
-    \item \(\mc{C}\) is a family of categories \(\mc{C}^q\) parametrized by subsets \(q \subseteq \{\ms{a}, \ms{r}\}\) such that
-    \begin{itemize}
-      \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the same. We will write this set as \(|\mc{C}|\)
-      \item All \(\mc{C}^q\) share symmetric (pre)monoidal structure, and are individually closed under the tensor products of morphisms
+  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A},
+  \mc{R})\) where:
+  \begin{itemize}
+    \item \(\mc{C}\) is a family of categories \(\mc{C}^q\) parametrized by
+    subsets \(q \subseteq \{\ms{a}, \ms{r}\}\) such that
+    \begin{itemize}
+      \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of
+      \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the
+      same. We will write this set as \(|\mc{C}|\)
+      \item All \(\mc{C}^q\) share symmetric (pre)monoidal structure, and are
+      individually closed under the tensor products of morphisms
       \item \(\mc{C}^{q \cap q'} = \mc{C}^q \cap \mc{C}^{q'}\)
     \end{itemize}
-    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects closed under tensor products, such that
+    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects
+    closed under tensor products, such that
     \begin{itemize}
       \item \(\mb{1} \in \mc{A}\)
-      \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
+      \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism
+      \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
       \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
@@
     \end{itemize}
-    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects closed under tensor products, such that
+    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects
+    closed under tensor products, such that
     \begin{itemize}
       \item \(\mb{1} \in \mc{R}\)
-      \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
+      \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism
+      \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
       \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
@@
   \end{itemize}
-  We call the morphisms in \(\mc{C}^{\{\ms{a}\}}\) \textbf{affine}, the morphisms in \(\mc{C}^{\{\ms{r}\}}\) \textbf{relevant}, and the morphisms in \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}.
-We call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the other \(\mc{C}^q\) \textbf{linear}.
-We call a substructural \textit{monoidal} category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
+  We call the morphisms in \(\mc{C}^{\{\ms{a}\}}\) \textbf{affine}, the
+  morphisms in \(\mc{C}^{\{\ms{r}\}}\) \textbf{relevant}, and the morphisms in
+  \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}. We
+  call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the
+  other \(\mc{C}^q\) \textbf{linear}. We call a substructural \textit{monoidal}
+  category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} =
+  |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
 \end{definition}
-\TODO{Define \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) syntax? Make lattice generated by syntax?}
-A simple example of a substructural Cartesian category is the category of relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\).
+\TODO{Define \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) syntax? Make lattice generated
+by syntax?} A simple example of a substructural Cartesian category is the
+category of relations \(\ms{Rel}\), viewed as the Kleisli category of the
+power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\).
 In particular, we can define:
@@
 \end{equation}
-Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and \(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and that they satisfy the axioms of a commutative comonoid, as desired.
-
-Note that we cannot simply have \(\ms{Rel}^{\{\ms{a}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) = \varnothing\), we have
+Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and
+\(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to
+prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and
+that they satisfy the axioms of a commutative comonoid, as desired.
+
+Note that we cannot simply have \(\ms{Rel}^{\{\ms{a}\}} = \ms{Rel}\), since, for
+any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) =
+\varnothing\), we have
 \begin{equation}
@@
 \end{equation}
-We will see that this corresponds to the fact that the following programs are not semantically equal
+We will see that this corresponds to the fact that the following programs are
+not semantically equal
 \begin{equation}
@@
 \end{equation}
-Similarly, we cannot simply have \(\ms{Rel}^{\{\ms{r}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)| \supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
+Similarly, we cannot simply have \(\ms{Rel}^{\{\ms{r}\}} = \ms{Rel}\), since,
+for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)|
+\supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
 \begin{equation}
@@
 \end{equation}
-since only the left-hand side contains the pair \((b_0, b_1)\), this corresponding to the fact that the following programs are not semantically equal
+since only the left-hand side contains the pair \((b_0, b_1)\), this
+corresponding to the fact that the following programs are not semantically equal
 \begin{equation}
@@
 \end{equation}
-To get a substructural premonoidal category, we can apply the state monad transformer to the power set monad to obtain the Kleisli category with morphisms of the form \(\ms{Set}_{\ms{St}\;\mc{P}\;S}(A, B) = A \to S \to \mc{P}(B \times S)\).
-Once again, it is clear that all objects have split and drop morphisms inherited from \(\ms{Set}\), and therefore are affine and relevant
+To get a substructural premonoidal category, we can apply the state monad
+transformer to the power set monad to obtain the Kleisli category with morphisms
+of the form \(\ms{Set}_{\ms{St}\;\mc{P}\;S}(A, B) = A \to S \to \mc{P}(B \times
+S)\). Once again, it is clear that all objects have split and drop morphisms
+inherited from \(\ms{Set}\), and therefore are affine and relevant
 
@@
   \begin{itemize}
-    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms in this category \textbf{central} or \textbf{linear}
-    \item A substructural premonoidal category \(\mc{C}_0\) sharing the same affine and relevant objects as \(\mc{C}_1\)
-    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0^\varnothing\) preserving all substructural premonoidal structure, i.e., preserves all symmetries, associators, unitors, drops, and splits and sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\).
-    In particular, restricting \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to \mc{C}_0^q\) into effectful categories.
-  \end{itemize}
-  We define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\) and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply be the identity functor on \(\mc{C}_1^q\).
-
-  Morphisms in \(\mc{C}_p^{\{\ms{a}\}}\) are called \textbf{copyable}, morphisms in \(\mc{C}_p^{\{\ms{r}\}}\) are called \textbf{discardable},
-  while morphisms in \(\mc{C}_1^{\{\ms{a}\}}, \mc{C}_1^{\{\ms{r}\}}, \mc{C}_1^{\{\ms{a}, \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and \textbf{intuitionistic} or \textbf{pure} respectively.
+    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms
+    in this category \textbf{central} or \textbf{linear}
+    \item A substructural premonoidal category \(\mc{C}_0\) sharing the same
+    affine and relevant objects as \(\mc{C}_1\)
+    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing
+    \to \mc{C}_0^\varnothing\) preserving all substructural premonoidal
+    structure, i.e., preserves all symmetries, associators, unitors, drops, and
+    splits and sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\). In particular,
+    restricting \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to
+    \mc{C}_0^q\) into effectful categories.
+  \end{itemize}
+  We define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\)
+  and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply
+  be the identity functor on \(\mc{C}_1^q\).
+
+  Morphisms in \(\mc{C}_p^{\{\ms{a}\}}\) are called \textbf{copyable}, morphisms
+  in \(\mc{C}_p^{\{\ms{r}\}}\) are called \textbf{discardable}, while morphisms
+  in \(\mc{C}_1^{\{\ms{a}\}}, \mc{C}_1^{\{\ms{r}\}}, \mc{C}_1^{\{\ms{a},
+  \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and
+  \textbf{intuitionistic} or \textbf{pure} respectively.
 \end{definition}
 
-We can obtain an example of a substructural effectful category quite easily by, considering the power-set monad above, applying the state transformer to yield the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to \mc{P}(A \times S)\).
-Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
+We can obtain an example of a substructural effectful category quite easily by,
+considering the power-set monad above, applying the state transformer to yield
+the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to
+\mc{P}(A \times S)\). Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
 \(
@@
 \), we can verify that this satisfies the axioms of a substructural effectful category.
-Note that here the \textit{central} morphisms are precisely those which do not read from or write to the state, and therefore which commute with all other morphisms.
-
-In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a substructural premonoidal category, with \(\mc{C}^\varnothing = \mc{C}\) and \(\mc{C}^{\{\ms{a}, \ms{r}\}}, \mc{C}^{\{\ms{a}\}}, \mc{C}^{\{\ms{r}\}}\) the subcategory consisting only of the associators, unitors, and symmetry, and only the identity object considered affine and relevant.
-On the other extreme, every \textit{Cartesian} category \(\mc{C}\) can be viewed as a substructural Cartesian category with all morphisms both relevant and affine; in fact, this is the case \textit{if and only if} a category is Cartesian, hence the term "substructural Cartesian category."
+Note that here the \textit{central} morphisms are precisely those which do not
+read from or write to the state, and therefore which commute with all other
+morphisms.
+
+In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a
+substructural premonoidal category, with \(\mc{C}^\varnothing = \mc{C}\) and
+\(\mc{C}^{\{\ms{a}, \ms{r}\}}, \mc{C}^{\{\ms{a}\}}, \mc{C}^{\{\ms{r}\}}\) the
+subcategory consisting only of the associators, unitors, and symmetry, and only
+the identity object considered affine and relevant. On the other extreme, every
+\textit{Cartesian} category \(\mc{C}\) can be viewed as a substructural
+Cartesian category with all morphisms both relevant and affine; in fact, this is
+the case \textit{if and only if} a category is Cartesian, hence the term
+"substructural Cartesian category."
 
@@
 \begin{proposition}
-  If \(\mc{C}\) is a substructural monoidal category then the Kleisli category \(\mc{C}_{\mb{T}}\) is a substructural premonoidal category with:
+  If \(\mc{C}\) is a substructural monoidal category then the Kleisli category
+  \(\mc{C}_{\mb{T}}\) is a substructural premonoidal category with:
   \begin{itemize}
@@
   \end{itemize} 
-  In particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a substructural effectful category.
+  In particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a
+  substructural effectful category.
 \end{proposition}
@@
 \begin{definition}[Distributive (pre)monoidal category] 
-  A \textbf{distributive (pre)monoidal category} is a (pre)monoidal category with all finite coproducts equipped with a natural family of central isomorphisms (the \textbf{distributor}) \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\).
-This isomorphism is subject to rather complex coherence conditions, worked out in full for the more general case of \textit{rig categories} in \cite{laplaza-distributivity}, but for our purposes we will only make use of the following:
+  A \textbf{distributive (pre)monoidal category} is a (pre)monoidal category
+  with all finite coproducts equipped with a natural family of central
+  isomorphisms (the \textbf{distributor}) \(\delta_{A, B, C}: (A + B) \otimes C
+  \simeq A \otimes C + B \otimes C\). This isomorphism is subject to rather
+  complex coherence conditions, worked out in full for the more general case of
+  \textit{rig categories} in \cite{laplaza-distributivity}, but for our purposes
+  we will only make use of the following:
   \begin{equation*}
@@
   \end{equation*}
-  Note in particular that this implies that \(\iota_i \otimes C;\delta_{A, B, C} = \iota_i\).
-A substructural (pre)monodial category is \textbf{distributive} if it is distributive as a (pre)monoidal category, and its distributor is \textit{pure}.
-A substructural effectful category \(\mc{C}\) is \textbf{distributive} if \(\mc{C}_0\) is distributive (as a substructural (pre)monoidal category).
+  Note in particular that this implies that \(\iota_i \otimes C;\delta_{A, B, C}
+  = \iota_i\). A substructural (pre)monodial category is \textbf{distributive}
+  if it is distributive as a (pre)monoidal category, and its distributor is
+  \textit{pure}. A substructural effectful category \(\mc{C}\) is
+  \textbf{distributive} if \(\mc{C}_0\) is distributive (as a substructural
+  (pre)monoidal category).
 \end{definition}
-Many (pre)monoidal categories are distributive, including of course \(\ms{Set}\), but also categories which can be derived from monads on more primitive categories such as \(\ms{PFun}\) and \(\ms{Rel}\), as can be deduced from the following proposition:
+Many (pre)monoidal categories are distributive, including of course
+\(\ms{Set}\), but also categories which can be derived from monads on more
+primitive categories such as \(\ms{PFun}\) and \(\ms{Rel}\), as can be deduced
+from the following proposition:
 \begin{proposition}
-  If \(\mc{C}\) is a distributive substructural monoidal category and \(\mb{T}\) is a strong monad on \(\mc{C}\), then the Kleisli category \(\mc{C}_{\mb{T}}\) is a distributive substructural effectful category (with affine and relevant objects and morphisms inherited from \(\mc{C}\))
+  If \(\mc{C}\) is a distributive substructural monoidal category and \(\mb{T}\)
+  is a strong monad on \(\mc{C}\), then the Kleisli category \(\mc{C}_{\mb{T}}\)
+  is a distributive substructural effectful category (with affine and relevant
+  objects and morphisms inherited from \(\mc{C}\))
 \end{proposition}
@@
 \begin{definition}[Elgot category] 
-  An \textbf{Elgot category} is a category \(\mc{C}\) with coproducts equipped with an operator \(\cdot^\dagger: \mc{C}(A, B + A) \to \mc{C}(A, B)\) satisfying the following properties:
+  An \textbf{Elgot category} is a category \(\mc{C}\) with coproducts equipped
+  with an operator \(\cdot^\dagger: \mc{C}(A, B + A) \to \mc{C}(A, B)\)
+  satisfying the following properties:
   \begin{itemize}
@@
   \end{itemize}
-  A monad \(\mb{T}\) on \(\mc{C}\) such that its Kleisli cateogry \(\mc{C}_{\mb{T}}\) is Elgot is called an \textbf{Elgot monad}.
+  A monad \(\mb{T}\) on \(\mc{C}\) such that its Kleisli cateogry
+  \(\mc{C}_{\mb{T}}\) is Elgot is called an \textbf{Elgot monad}.
 \end{definition}
@@
 \begin{itemize}
-  \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id}, (h;[\iota_0, g])^\dagger]\)
+  \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id},
+  (h;[\iota_0, g])^\dagger]\)
   \item \textbf{Squaring:} \((f;[\iota_0, f])^\dagger = f^\dagger\)
@@
 \begin{proposition}
-  If \(\mb{T}\) is an Elgot monad on \(\ms{Set}\), then \(\ms{StateT}\;\mb{T}\;S = \lambda A.
-S \to \ms{T}(A \times S)\), \(\ms{ReaderT}\;\mb{T}\;S = \lambda A.
-S \to \ms{T}(A)\), and \(\ms{WriterT}\;\mb{T}\;S = \lambda A.
-\ms{T}(A \times S)\) are all Elgot
+  If \(\mb{T}\) is an Elgot monad on \(\ms{Set}\), then \(\ms{StateT}\;\mb{T}\;S
+  = \lambda A. S \to \ms{T}(A \times S)\), \(\ms{ReaderT}\;\mb{T}\;S = \lambda
+  A. S \to \ms{T}(A)\), and \(\ms{WriterT}\;\mb{T}\;S = \lambda A. \ms{T}(A
+  \times S)\) are all Elgot
 \end{proposition}
@@
 
-\TODO{make aff and rel syntax nicer/more consistent here? Subcategory syntax would be nice for lemmas and just coherence.}
-
-We're now ready to give a semantics for an \isotopessa dialect with a set of types \(\mc{T}\) and instructions \(\mc{I}\). We'll need:
-\begin{itemize}
-  \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot}, equipped with
-  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies \ms{base}(X) \in \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies \ms{base}(X) \in \ms{Rel}(\mc{C})\)
-  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\) to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
-\end{itemize}
-We begin by giving a semantics for types and contexts in terms of objects in \(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}; this allows us to give a semantics for structural judgements in Figure \ref{fig:ssa-structural-semantics}.
+\TODO{make aff and rel syntax nicer/more consistent here? Subcategory syntax
+would be nice for lemmas and just coherence.}
+
+We're now ready to give a semantics for an \isotopessa dialect with a set of
+types \(\mc{T}\) and instructions \(\mc{I}\). We'll need:
+\begin{itemize}
+  \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to
+  \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot},
+  equipped with
+  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects
+  in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies \ms{base}(X) \in
+  \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies \ms{base}(X) \in
+  \ms{Rel}(\mc{C})\)
+  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\)
+  to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
+\end{itemize}
+We begin by giving a semantics for types and contexts in terms of objects in
+\(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}; this allows us to give a
+semantics for structural judgements in Figure
+\ref{fig:ssa-structural-semantics}.
 
@@
 
-We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics} followed by a semantics for blocks and label-sets in Figure \ref{fig:ssa-block-semantics}.
+We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics}
+followed by a semantics for blocks and label-sets in Figure
+\ref{fig:ssa-block-semantics}.
 
@@
 
-\TODO{these}
+\subsection{Syntactic Metatheory}
+
+\begin{theorem}[Weakening] \
+  \begin{itemize}
+    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\), \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
+    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
+    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then \(\haslb{\Gamma}{t}{\ms{K}}\)
+  \end{itemize}
+\end{theorem}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{lemma}[Substitution Splitting] 
+  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
+  \begin{itemize}
+    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
+    \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
+    \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
+  \end{itemize}
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{lemma}[Substitution] 
+  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  \begin{itemize}
+    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
+    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
+    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
+    \begin{itemize}
+      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
+      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
+      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
+    \end{itemize}
+  \end{itemize}
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{lemma}[Label Substitution] 
+  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
+  \begin{itemize}
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
+    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
+  \end{itemize}
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\subsection{Semantic Metatheory}
+
+\begin{lemma}[Semantic Upgrade]
+  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
+  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{lemma}[Semantic Weakening] 
+  If \(\cwk{\Gamma}{\Delta}\), then 
+  \begin{itemize}
+    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
+    \[
+      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
+      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+    \]
+    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
+    \[
+      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
+      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
+    \]
+  \end{itemize}
+
+  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
+  \[
+    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
+    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
+  \]
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
 
@@
 \end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
 
@@
 \end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{theorem}[Semantic Substitution] 
+  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  \begin{itemize}
+    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
+    \[
+      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
+      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
+    \]
+    \item For all 
+      \(\haslb{\Gamma}{t}{\ms{L}}\), 
+      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
+      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
+    \[
+      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
+      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
+      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
+    \]
+    \item For all 
+      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
+      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
+      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
+      \[
+        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
+        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
+        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
+        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+      \]
+  \end{itemize}
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
+  \begin{itemize} 
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+    \[
+      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
+    \]
+    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), we have
+    \[
+      \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      = \dnt{\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{L}}}
+    \]
+  \end{itemize}
+\end{theorem}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
 

### 2023-11-14 pldi24/paper.tex
Tap

@@
 
-\TODO{intro text}
-
+We can now state some of the basic metatheoretic properties of \isotopessa. We
+begin with weakening, which we state in our setting as follows:
 \begin{lemma}[Weakening] \
@@
     \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
-    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then \(\haslb{\Gamma}{t}{\ms{K}}\)
+    \item If \(\lwk{\ms{L}}{\ms{K}}\), then:
+    \begin{itemize}
+      \item If \(\haslb{\Gamma}{t}{\ms{L}}\) then \(\haslb{\Gamma}{t}{\ms{K}}\)
+      \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\) then \(\lhaslb{\ms{W}}{L}{\ms{L}}\)
+    \end{itemize}
   \end{itemize}
 \end{lemma}
-
-\TODO{text}
+That is, we can do the following and still keep our typing judgements valid
+\begin{itemize}
+  \item Add unused \textit{affine} variables
+  \item \textit{Reduce} the quantity of an expressions type
+  \item Add arbitrary unused labels, and remove unused \textit{affine} variables
+  from existing labels
+\end{itemize}
+This can be proved by a relatively straightforward mutual induction.
+
+Our main metatheoretic results will be \textit{substitution} and
+\textit{label-substitution}, which will require introducing some new
+metatheoretic judgements to state properly, which we do in Figure
+\ref{fig:ssa-meta-judgements}.
+
+\TODO{some text about each metatheoretic judgement}
 
@@
 
-\TODO{some text about each metatheoretic judgement}
-
 \begin{figure}
@@
   \begin{itemize}
-    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
+    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) and \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}\)
     \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such
     that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
-    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
+    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\), \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
     \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},

### 2023-11-14 pldi24/paper.tex
Tap

@@
 The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the
-largest central subcategory is the center of \(\mc{C}\) \(Z(\mb{C})\); the
+largest central subcategory is the center of \(\mc{C}\), \(Z(\mb{C})\); the
 existence of a top-element means that in these cases we may induce the structure
@@
   \begin{itemize}
-    \item \(\mc{C}\) is a family of categories \(\mc{C}^q\) parametrized by
-    subsets \(q \subseteq \{\ms{a}, \ms{r}\}\) such that
-    \begin{itemize}
-      \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of
-      \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the
-      same. We will write this set as \(|\mc{C}|\)
-      \item All \(\mc{C}^q\) share symmetric (pre)monoidal structure, and are
-      individually closed under the tensor products of morphisms
-      \item \(\mc{C}^{q \cap q'} = \mc{C}^q \cap \mc{C}^{q'}\)
-    \end{itemize}
+    \item \(\mc{C}\) is a symmetric premonoidal category equipped with wide
+    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\).
+    We define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)
     \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects
@@
       \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
-      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\{\ms{a}\}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
+      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\ms{a}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
     \end{itemize}
@@
       \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
-      \item \(\forall A, B \in \mc{R}, \forall f: \mc{C}^{\{\ms{r}\}}(A, B), f;\ms{split}(B) = \ms{split}(A);f \ltimes f = \ms{split}(A);f\rtimes f\)
-    \end{itemize}
-  \end{itemize}
-  We call the morphisms in \(\mc{C}^{\{\ms{a}\}}\) \textbf{affine}, the
-  morphisms in \(\mc{C}^{\{\ms{r}\}}\) \textbf{relevant}, and the morphisms in
+      \item \(\forall A, B \in \mc{R}, \forall f: \mc{C}^{\ms{r}}(A, B), f;\ms{split}(B) = \ms{split}(A);f \ltimes f = \ms{split}(A);f\rtimes f\)
+    \end{itemize}
+  \end{itemize}
+  We call the morphisms in \(\mc{C}^{\ms{a}}\) \textbf{affine}, the
+  morphisms in \(\mc{C}^{\ms{r}}\) \textbf{relevant}, and the morphisms in
   \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}. We
@@
 \end{definition}
-\TODO{Define \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) syntax? Make lattice generated
-by syntax?} A simple example of a substructural Cartesian category is the
-category of relations \(\ms{Rel}\), viewed as the Kleisli category of the
-power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\).
-In particular, we can define:
+A simple example of a substructural Cartesian category is the category of
+relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad
+with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\). In particular,
+we can define:
 \begin{equation}
@@
     &&= A \to \mc{P}(B) \\
-    \ms{Rel}^{\{\ms{a}\}}(A, B) 
+    \ms{Rel}^{\ms{a}}(A, B) 
     &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
     &&= A \to \mc{P}^+(B) \\
-    \ms{Rel}^{\{\ms{r}\}}(A, B) 
+    \ms{Rel}^{\ms{r}}(A, B) 
     &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
@@
 
-Note that we cannot simply have \(\ms{Rel}^{\{\ms{a}\}} = \ms{Rel}\), since, for
+Note that we cannot simply have \(\ms{Rel}^{\ms{a}} = \ms{Rel}\), since, for
 any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) =
@@
 \end{equation}
-Similarly, we cannot simply have \(\ms{Rel}^{\{\ms{r}\}} = \ms{Rel}\), since,
+Similarly, we cannot simply have \(\ms{Rel}^{\ms{r}} = \ms{Rel}\), since,
 for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)|

### 2023-11-14 pldi24/paper.tex
Tap
Typo

@@
   \begin{aligned}
-    \ms{Rel}^\varnothing(A, B) 
-    &= \ms{Rel}(A, B) 
-    &&= A \to \mc{P}(B) \\
     \ms{Rel}^{\ms{a}}(A, B) 
@@
 
-  Morphisms in \(\mc{C}_p^{\{\ms{a}\}}\) are called \textbf{copyable}, morphisms
-  in \(\mc{C}_p^{\{\ms{r}\}}\) are called \textbf{discardable}, while morphisms
-  in \(\mc{C}_1^{\{\ms{a}\}}, \mc{C}_1^{\{\ms{r}\}}, \mc{C}_1^{\{\ms{a},
+  Morphisms in \(\mc{C}_p^{\ms{r}}\) are called \textbf{copyable}, morphisms
+  in \(\mc{C}_p^{\ms{a}}\) are called \textbf{discardable}, while morphisms
+  in \(\mc{C}_1^{\ms{a}}, \mc{C}_1^{\ms{r}}, \mc{C}_1^{\{\ms{a},
   \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and

### 2023-11-14 pldi24/paper.tex
Tap

@@
 
-\TODO{some text about substitution}
+In particular, the judgement \(\issubst{\gamma}{\Theta}{\Gamma}\) says that
+\(\gamma\), which is taken to be a map from variable names to expressions, is a
+valid substitution taking \(\Theta\) to \(\Gamma\). What this means is that, for
+each variable \(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted
+as a \textit{pure} term of type \(A^q\) in a subcontext of \(\Theta\), such that
+when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables
+in \(\Theta\) are used up: this is described formally via the typing rules
+\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
+to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
+\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
+\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/discarding
+of substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
+\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).
+
+\TODO{some text about substitution operations}
+
+\TODO{some text about label substitution}
+
+\TODO{some text about label substitution operations}
+
+\TODO{shuffle subst/lsubst and then ops?}
 
@@
 \end{figure}
-
-\TODO{some text about substitution operations}
 

### 2023-11-14 pldi24/paper.tex
Tap

@@
 each variable \(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted
-as a \textit{pure} term of type \(A^q\) in a subcontext of \(\Theta\), such that
-when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables
-in \(\Theta\) are used up: this is described formally via the typing rules
-\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
-to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
+as a \textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such
+that when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine
+variables in \(\Theta\) are used up: this is described formally via the typing
+rules \rle{subst-nil}, which says that any function \(\gamma\) takes the empty
+context to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
 \(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
@@
 
-\TODO{some text about substitution operations}
+Given such a substitution \(\gamma\), there are two ways to use it. The first is
+standard capture-avoiding substitution, which we denote \([\gamma]e\) for
+expressions, \([\gamma]t\) for blocks, and \([\gamma]L\) for label-sets, as is
+standard. The second is to use the substitution to generate a set of
+let-bindings for a context \(\Gamma\), written
+\(\exprletsubst{\substctx{\gamma}{\Gamma}}{e}\) for expressions and
+\(\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}\) for blocks, as defined in Figure
+\ref{fig:ssa-subst-ops}. Note that this operation cannot be easily defined for
+label-sets. In Section \ref{sec:semantics}, we will show that this is
+semantically equivalent to regular capture-avoiding substitution, due to the
+fact that all terms in a substitution are central \textit{and} that only affine
+computations can be used in an expression substituting for an affine variable,
+and only relevant computationbs can be used in an expression substituting
+for a relevant variable.
+
+We can now state our substitution theorem for terms in a straightforward way: if
+\(\substctx{\gamma}{\Theta}{\Gamma}\) and \(\hasty{\Gamma}{p}{e}{A}{q}\), then
+\(\hasty{\Theta}{p}{[\gamma]e}{A}{q}\) and
+\(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{e}}{A}{q}\).
+Unfortunately, the substitution theorem for blocks is a little more complicated:
+the naive choice of attempting to show \(\haslb{\Gamma}{t}{\ms{L}} \implies
+\haslb{\Theta}{[\gamma]t}{\ms{L}}\) does not work, since \(\ms{L}\) may contain
+captured variables from \(\Gamma\) that are no longer available if we are typing
+\([\gamma]t\) in the context of \(\Theta\). To solve this problem, we need to
+introduce the concept of a \textit{label-substitution} \(\mc{L}\) \todo{rest}
 
@@
 \TODO{some text about label substitution operations}
-
-\TODO{shuffle subst/lsubst and then ops?}
 

### 2023-11-14 pldi24/paper.tex
Tap

@@
 
-Our main metatheoretic results will be \textit{substitution} and
-\textit{label-substitution}, which will require introducing some new
-metatheoretic judgements to state properly, which we do in Figure
-\ref{fig:ssa-meta-judgements}.
-
 \begin{figure}
@@
 
-In particular, the judgement \(\issubst{\gamma}{\Theta}{\Gamma}\) says that
-\(\gamma\), which is taken to be a map from variable names to expressions, is a
-valid substitution taking \(\Theta\) to \(\Gamma\). What this means is that, for
-each variable \(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted
-as a \textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such
-that when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine
-variables in \(\Theta\) are used up: this is described formally via the typing
-rules \rle{subst-nil}, which says that any function \(\gamma\) takes the empty
-context to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
+Our main metatheoretic results will be \textit{substitution} and
+\textit{label-substitution}, which will require introducing some new
+metatheoretic judgements to state properly, which we do in Figure
+\ref{fig:ssa-meta-judgements}. In particular, the judgement
+\(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which is taken to be
+a map from variable names to expressions, is a valid substitution taking
+\(\Theta\) to \(\Gamma\). What this means is that, for each variable
+\(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted as a
+\textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such that
+when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables
+in \(\Theta\) are used up: this is described formally via the typing rules
+\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
+to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
 \(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
@@
 We can now state our substitution theorem for terms in a straightforward way: if
-\(\substctx{\gamma}{\Theta}{\Gamma}\) and \(\hasty{\Gamma}{p}{e}{A}{q}\), then
+\(\issubst{\gamma}{\Theta}{\Gamma}\) and \(\hasty{\Gamma}{p}{e}{A}{q}\), then
 \(\hasty{\Theta}{p}{[\gamma]e}{A}{q}\) and

### 2023-11-14 pldi24/paper.tex
Tap

@@
 captured variables from \(\Gamma\) that are no longer available if we are typing
-\([\gamma]t\) in the context of \(\Theta\). To solve this problem, we need to
-introduce the concept of a \textit{label-substitution} \(\mc{L}\) \todo{rest}
-
-\TODO{some text about label substitution}
-
-\TODO{some text about label substitution operations}
+\([\gamma]t\) in the context of \(\Theta\).
 
@@
 
-\begin{lemma}[Substitution Splitting] 
-  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
-  \begin{itemize}
-    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
-    \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
-    \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
-  \end{itemize}
-\end{lemma}
-
-\TODO{some text about substitution weakening}
-
+To solve this problem, we need to introduce the concept of a
+\textit{label-substitution} \(\mc{L}\), which we define as a function that takes
+labels \(\lbl{\ell}\) and variables \(x\) to blocks \(t\) such that, for any
+fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)\). We
+introduce the judgement \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to state that the
+label-substitution \(\mc{L}\) takes the label-set \(\ms{L}\) to the label-set
+\(\ms{K}\); that is, for each \(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have
+that \(\mc{L}(\lbl{\ell}, x)\) is a block that branches to \(\ms{K}\). This is
+formally defined via the rules \rle{lb-subst-nil}, which states that this is
+vacuously true for \(\ms{L} = \cdot\), and \rle{lb-subst-cons}, which says that
+if \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
+\thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{\ms{L},
+\lhyp{\lbl{\ell}}{\Gamma}{A}}{\ms{K}}\), in Figure \ref{fig:ssa-subst-typing}.
+Given a context \(\gamma\) and a label-set \(\ms{L}\), we can convert it to a
+label-substitution \(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
+\ref{fig:ssa-subst-ops}: using this, we can state the full substitution theorem
+as follows:
 \begin{lemma}[Substitution] 
@@
   \begin{itemize}
-    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) and \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}\)
+    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then
+    \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) and
+    \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}\)
     \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such
     that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
-    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\), \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
+    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
+    \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
     \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
@@
 
-\TODO{some text about substitution}
+% \begin{lemma}[Substitution Splitting] 
+%   If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
+%   then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
+%   \begin{itemize}
+%     \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
+%     \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
+%     \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
+%   \end{itemize}
+% \end{lemma}
+
+% \TODO{some text about substitution weakening}
+
+\TODO{label substitution text}
 
@@
   \begin{itemize}
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
-    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\),
+    \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
+    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\),
+    \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
   \end{itemize}
 \end{lemma}
-
-\TODO{some text about label substitution}
 
@@
     \item \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\), \(\ms{Rel}(\mc{C}_{\mb{T}}) = \ms{Rel}(\mc{C})\)
-    \item \(\mc{C}_{\mb{T}}^q(A, B) = \upg{\mc{C}^q(A, B)}{}\), where \(\upg{f}{} = f;\eta\)
+    \item \(\mc{C}_{\mb{T}}^q(A, B) = \upg{\mc{C}^q(A, B)}{}\), where
+    \(\upg{f}{} = f;\eta\)
     \item Drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\)

### 2023-11-14 pldi24/paper.tex
Tap

@@
 label-substitution \(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
-\ref{fig:ssa-subst-ops}: using this, we can state the full substitution theorem
-as follows:
+\ref{fig:ssa-subst-ops}: using this, we can state the full substitution lemma as
+follows:
 \begin{lemma}[Substitution] 
@@
 
-\TODO{label substitution text}
-
+We can also, using the rules in Figure \ref{fig:ssa-subst-ops}, define
+capture-avoiding label-substitution for blocks \([\mc{L}]t\) and label-sets
+\([\mc{L}]L\), allowing us to state the following lemma:
 \begin{lemma}[Label Substitution] 
@@
 
-\TODO{some text about rewriting}
+\TODO{some text about rewriting?}
 

### 2023-11-14 pldi24/paper.tex
Tap

@@
 
+To add support for branching, we need to equip our category with a
+\textit{distributive coproduct structure}, as follows:
 \begin{definition}[Distributive (pre)monoidal category] 
@@
 
-\TODO{text}
-
+The structure so far is enough to support non-looping control flow, but to allow
+general, potentially nonterminating control flow, we need a way to reason about
+iteration. To do so, we introduce the notion of an \textit{Elgot category} as
+follows:
 \begin{definition}[Elgot category] 
@@
 \end{definition}
-\TODO{why do these matter? they show up in the proofs, but...}
 Some important properties which hold for all Elgot categories include:
@@
 
+One important way of generating Elgot categories is through monads: a monad
+whose Kliesli category is Elgot is called an \textbf{Elgot monad}. Examples of
+Elgot monads include \(\ms{Option}\) (where \(\ms{None}\) is interpreted as
+nontermination), the powerset monad (where the empty set is interpreted as
+nontermination), and the trace monads we will explore in Section
+\ref{ssec:trace-monads}.
+
+It turns out that various monad transformers preserve Elgot structure, allowing
+us to build up Elgot monads in a composable fashion. In particular,
 \begin{proposition}
-  If \(\mb{T}\) is an Elgot monad on \(\ms{Set}\), then \(\ms{StateT}\;\mb{T}\;S
+  If \(\mb{T}\) is an Elgot monad on \(\ms{Set}\), then \(\ms{StateT}\;S\;\mb{T}
   = \lambda A. S \to \ms{T}(A \times S)\), \(\ms{ReaderT}\;\mb{T}\;S = \lambda
   A. S \to \ms{T}(A)\), and \(\ms{WriterT}\;\mb{T}\;S = \lambda A. \ms{T}(A
-  \times S)\) are all Elgot
+  \times S)\) are all Elgot, with iteration operators lifted from \(\mb{T}\) in
+  the obvious manner.
 \end{proposition}
@@
 \subsection{Denotational Semantics}
-
-\TODO{make aff and rel syntax nicer/more consistent here? Subcategory syntax
-would be nice for lemmas and just coherence.}
 
@@
 \subsection{Trace Models}
+
+\label{ssec:trace-monads}
 

### 2023-11-14 pldi24/paper.tex
Tap

@@
 two anonymous basic blocks, while nested scopes simply \textit{restrict} where
-we can branch to. In particular, in Section \ref{ssec:ssa-norm}, we will
-formalize interconversion with traditional SSA and show how it is semantically
-sound to simply merge all nested scopes. The point of the slightly modified
-syntax is simply to make some of our metatheoretic results more practical to
-state.
+we can branch to. 
+% In particular, in Section \ref{ssec:ssa-norm}, we will formalize
+% interconversion with traditional SSA and show how it is semantically sound to
+% simply merge all nested scopes. The point of the slightly modified syntax is
+% simply to make some of our metatheoretic results more practical to state.
 
@@
 
-\subsection{Syntactic Metatheory}
+\subsection{Weakening and Substitution}
 
@@
 
-\TODO{some text about rewriting?}
+\TODO{small rewriting section here?}
 
@@
 
-\subsection{Normalization}
-
-\label{ssec:ssa-norm}
-
-\TODO{this}
+% \subsection{Normalization}
+
+% \label{ssec:ssa-norm}
+
+% \TODO{this}
 
@@
 
+\subsection{Syntax}
+
+To properly be able to reason about congruence and congruence-based
+optimization, we need to introduce the concept of \textit{terms with holes},
+which are defined using the grammar in Figure
+\ref{fig:blocks-with-holes-grammar}. This simply extends the three main
+syntactic categories with \textit{holes} of the form \(\lhole{X}\),
+\(\lhole{T}\), \(\lhole{W}\). To \textit{type} terms with holes, we need to
+introduce the concept of a \textit{hole-context} \(H\), which types a hole as being
+one of:
+\begin{itemize}
+  \item \todo{expression}
+  \item \todo{block}
+  \item \todo{label-set}
+\end{itemize}
+\TODO{hole typing rules for expressions, blocks, label-sets; everything else is
+the same}.
+
 \begin{figure}
@@
 \end{figure}
-
-\TODO{some text about terms with holes}
 
@@
 
-\TODO{some text about term with hole typing rules}
-
 \begin{figure}
@@
 
-\TODO{some text about block with hole typing rules}
-
 \begin{figure}
@@
   \end{itemize}
-
   Furthermore, we have that 
@@
 \end{theorem}
+
+\subsection{Semantics}
+
+\TODO{text}
 

### 2023-11-14 pldi24/paper.tex
some changes to the related work

@@
 while (e) {
-  x = read(a);
+  x = load(a);
   body
@@
 \begin{BVerbatim}[baseline=c]
-x = read(a);
+x = load(a);
 while(e) {
@@
 informally, since the "intended model" of the language is a simple sequential
-machine operating over a heap which is a flat array of bytes. Unfortunately,
+machine operating over a flat array of bytes. Unfortunately,
 over time both the intended model and the actual hardware have drifted away from
@@
 Our main metatheoretic results will be \textit{substitution} and
-\textit{label-substitution}, which will require introducing some new
-metatheoretic judgements to state properly, which we do in Figure
+\textit{label-substitution}, which will require introducing some
+substitution judgements to state properly, which we do in Figure
 \ref{fig:ssa-meta-judgements}. In particular, the judgement
-\(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which is taken to be
-a map from variable names to expressions, is a valid substitution taking
-\(\Theta\) to \(\Gamma\). What this means is that, for each variable
-\(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted as a
-\textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such that
-when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables
-in \(\Theta\) are used up: this is described formally via the typing rules
-\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
-to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
-\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
-\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/discarding
-of substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
-\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).
+\(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which is
+taken to be a map from variable names to expressions, is a valid
+substitution taking \(\Theta\) to \(\Gamma\). What this means is that,
+for each variable \(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can
+be interpreted as a \textit{central} term of type \(A^q\) in a
+subcontext of \(\Theta\), such that when this is done for all
+\(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables in \(\Theta\)
+are used up: this is described formally via the typing rules
+\rle{subst-nil}, which says that any function \(\gamma\) takes the
+empty context to itself, and \rle{subst-cons}, which says that if
+\(\gamma\) takes \(\Theta_\Gamma\) to \(\Gamma\),
+\(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\),
+\(q \subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the
+duplication/discarding of substructural variables), and \(\Theta\)
+splits into \(\Theta_\Gamma\) and \(\Theta_x\), then
+\(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).
 
@@
 
-\section{Related Work}
-
-\TODO{\cite{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
-
-\TODO{\cite{schneider-phd}}
-
-\TODO{\cite{chakravarty-functional-ssa-2003}}
-
-\TODO{\cite{barthe-compcert-ssa-2014}}
-
-\TODO{\cite{buchwald-verified-ssa-2016}}
-
-\TODO{\cite{herklotz-gsa-2023}}
-
-\TODO{\cite{pop-ssa-inout-2009}}
-
-\TODO{\cite{garbuzov-structural-cfg-2018}}
-
-\TODO{\cite{hua-explicit-ssa-2010}}
-
-\TODO{\cite{rvsdg}}
+\section{Related Work and Discussion}
+
+There is a vast literature on SSA and related compiler IRs, but for
+space reasons, we are forced to restrict our discussion to the most
+closely related work.
+
+\paragraph{Weak memory}
+
+\TODO{\citet{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
+\TODO{\citet{paviotti-modular-relaxed-dep-20}}
+
+\todo{Main thing here to say is that we need semantics for while loops, plus associativity of semicolon, if statements must support distributivity}
+
+\paragraph{Type Theory for SSA}
+
+\TODO{\citet{hua-explicit-ssa-2010}}
+\TODO{\citet{ssa-types-matsuno-ohori-06}}
+\TODO{\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20}}
+
+\paragraph{Functional SSA}
+
+\TODO{\citet{chakravarty-functional-ssa-2003}}
+
+\subparagraph{Operational Semantics}
+
+\TODO{\citet{barthe-compcert-ssa-2014}}
+\TODO{\citet{buchwald-verified-ssa-2016}}
+\TODO{\citet{herklotz-gsa-2023}}
+\TODO{\citet{garbuzov-structural-cfg-2018}}
+\TODO{\citet{vellum}}
+
+\subparagraph{Denotational Semantics}
+
+\TODO{\citet{pop-ssa-inout-2009}}
+
+% \TODO{\cite{schneider-phd}}
+% \TODO{\cite{rvsdg}}
 

### 2023-11-14 pldi24/paper.tex
some changes to the related work
Tap

@@
   judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
-  is of type \(A\) with \textit{purity} \(p = 1\) if \textit{pure}, or \(p = 0\)
+  is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or \(p = 0\)
   otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). The rules
@@
         \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
-        \(\hasty{\Gamma}{p}{a}{A}{q}\) &
-        \(a\) is a term of type \(A\) in context \(\Gamma\) with purity \(p\) and quantity \(q\) \\
-        \(\haslb{\Gamma}{t}{\ms{L}}\) &
-        \(t\) is a block targeting labels \(\ms{L}\) in context \(\Gamma\) \\
-        \(\lhaslb{\ms{L}}{L}{\ms{K}}\) &
-        The label-set \(L\) sends labels \(\ms{L}\) to labels \(\ms{K}\) \\
-        \(\csplits{\Gamma}{\Delta}{\Xi}\) &
-        The context \(\Gamma\) splits into \(\Delta\) and \(\Xi\) \\
-        \(\lwk{\ms{L}}{\ms{K}}\) &
-        The label-set \(\ms{L}\) weakens to the label-set \(\ms{K}\) \\
+        \(\hasty{\Gamma}{p}{a}{A}{q}\) & \(a\) is a term of type \(A\) in
+        context \(\Gamma\) with centrality \(p\) and quantity \(q\) \\
+        \(\haslb{\Gamma}{t}{\ms{L}}\) & \(t\) is a block targeting labels
+        \(\ms{L}\) in context \(\Gamma\) \\
+        \(\lhaslb{\ms{L}}{L}{\ms{K}}\) & The label-set \(L\) sends labels
+        \(\ms{L}\) to labels \(\ms{K}\) \\
+        \(\csplits{\Gamma}{\Delta}{\Xi}\) & The context \(\Gamma\) splits into
+        \(\Delta\) and \(\Xi\) \\
+        \(\lwk{\ms{L}}{\ms{K}}\) & The label-set \(\ms{L}\) weakens to the
+        label-set \(\ms{K}\) \\
         % \(\tlin{A}{q}\) &
@@
 instruction \(f\) which takes in an argument \(A\) and has a return type of
-\(B\) and a \textit{pure} expression \(\hasty{\Gamma}{1}{a}{A}{q}\), we conclude
-that \(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that \(f\) itself
-have purity \(p\) and quantity \(q\), i.e. that \(f \in \mc{I}_p^q(A, B)\). The
-\rle{pair} rule says that we can type a pair \((a, b)\) in \(\Gamma\) if
-\(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\) and \(\Xi\) types
-\(b\); the quantity of a pair is simply the intersection of the quantity of its
-components (this is expressed in the rule by allowing any quantity which types
-\textit{both} components), and both components must be pure. Finally, we have
-the rules \rle{let} and \rle{let2}, corresponding to let-bindings and
-destructuring let-bindings respectively. In both cases, we split the context
-\(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is used to type a pure
-subterm \(a\), which is bound to variables \(x\) or destructured to variables
-\(x, y\) respectively, in a potentially impure term \(e\) which is otherwise
-typed in \(\Delta\).
+\(B\) and a \textit{central} expression \(\hasty{\Gamma}{1}{a}{A}{q}\), we
+conclude that \(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that
+\(f\) itself have centrality \(p\) and quantity \(q\), i.e. that \(f \in
+\mc{I}_p^q(A, B)\). The \rle{pair} rule says that we can type a pair \((a, b)\)
+in \(\Gamma\) if \(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\)
+and \(\Xi\) types \(b\); the quantity of a pair is simply the intersection of
+the quantity of its components (this is expressed in the rule by allowing any
+quantity which types \textit{both} components), and both components must be
+central. Finally, we have the rules \rle{let} and \rle{let2}, corresponding to
+let-bindings and destructuring let-bindings respectively. In both cases, we
+split the context \(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is
+used to type a central subterm \(a\), which is bound to variables \(x\) or
+destructured to variables \(x, y\) respectively, in a potentially non-central
+term \(e\) which is otherwise typed in \(\Delta\).
 
@@
 The rules for block typing are given in Figure \ref{fig:ssa-block-typing}.
-\rle{br} says that, given a pure term \(a: A\) which is \textit{both} affine and
-relevant (i.e. intuitionistic), we can type a block \(\lbrb{\ell}{a}\) which
-branches to \(\ell\) with argument \(a\), where \(\ms{L}\) contains the context
-\(\ell\) with an argument type of \(A\) and an appropriate set of unused
-variables. The rule \rle{ite} typechecks if-then-else statements by stating
-that, if \(\Gamma\) splits into \(\Delta\) and \(\Xi\), where \(\Xi\) typechecks
-a pure, intuitionistic discriminant \(e: \mb{2}\) and \(\Delta\) typechecks both
-\(s, t\) to branch into \(\ms{L}\), then the whole if-statement
-\(\ite{e}{s}{t}\) also typechecks as branching into \(\ms{L}\) in constext
-\(\Gamma\). \rle{let-blk} and \rle{let2-blk} are quite similar to their
-expression analogues, except that the bound expression \(e\) is allowed to be
-impure (otherwise, programs could only make use of pure functions!), and the
-remainder of the context plus the bound variable is used to type a block rather
-than a term. Finally, probably the most interesting rule is \rle{where}, which
-says that, given a block \(t\) which branches to \(\ms{L}\) in context
-\(\Gamma\), and labels \(L\) mapping \(\ms{L}\) to \(\ms{K}\), \(\ewhere{t}{L}\)
-branches to \(\ms{K}\). Here, \(L\) mapping \(\ms{L}\) to \(\ms{K}\) is defined
-using the typing rules \rle{nil-br}, which says that an empty list of blocks
-leaves the target label-set unchanged, and \rle{cons-br}, which says that if
-\(L\) maps \(\ms{L}\) to either \(\ms{K}\) or the label
-\(\lbl{\ell}[\Gamma](A)\), and \(t\) branches to \(\ms{L}\) in the context
-\(\Gamma, \thyp{x}{A}{}\), then \(L\) maps \(\ms{L}\) to \(\ms{K}\), having
-(potentially recursively) defined \(\lbl{\ell}\).
+\rle{br} says that, given a pure term \(a: A\) (that is, central, relevant, and
+affine), we can type a block \(\lbrb{\ell}{a}\) which branches to \(\ell\) with
+argument \(a\), where \(\ms{L}\) contains the context \(\ell\) with an argument
+type of \(A\) and an appropriate set of unused variables. The rule \rle{ite}
+typechecks if-then-else statements by stating that, if \(\Gamma\) splits into
+\(\Delta\) and \(\Xi\), where \(\Xi\) typechecks a pure, intuitionistic
+discriminant \(e: \mb{2}\) and \(\Delta\) typechecks both \(s, t\) to branch
+into \(\ms{L}\), then the whole if-statement \(\ite{e}{s}{t}\) also typechecks
+as branching into \(\ms{L}\) in constext \(\Gamma\). \rle{let-blk} and
+\rle{let2-blk} are quite similar to their expression analogues, except that the
+bound expression \(e\) is allowed to be impure (otherwise, programs could only
+make use of pure functions!), and the remainder of the context plus the bound
+variable is used to type a block rather than a term. Finally, probably the most
+interesting rule is \rle{where}, which says that, given a block \(t\) which
+branches to \(\ms{L}\) in context \(\Gamma\), and labels \(L\) mapping
+\(\ms{L}\) to \(\ms{K}\), \(\ewhere{t}{L}\) branches to \(\ms{K}\). Here, \(L\)
+mapping \(\ms{L}\) to \(\ms{K}\) is defined using the typing rules \rle{nil-br},
+which says that an empty list of blocks leaves the target label-set unchanged,
+and \rle{cons-br}, which says that if \(L\) maps \(\ms{L}\) to either \(\ms{K}\)
+or the label \(\lbl{\ell}[\Gamma](A)\), and \(t\) branches to \(\ms{L}\) in the
+context \(\Gamma, \thyp{x}{A}{}\), then \(L\) maps \(\ms{L}\) to \(\ms{K}\),
+having (potentially recursively) defined \(\lbl{\ell}\).
 
@@
 \end{definition}
+
+\TODO{confusion with notion of pure in truly "pure" above}
 
@@
 \TODO{\citet{paviotti-modular-relaxed-dep-20}}
+\TODO{\citet{sparky}}
 

### 2023-11-15 pldi24/paper.tex
Tap

@@
 We begin by giving a semantics for types and contexts in terms of objects in
-\(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}; this allows us to give a
-semantics for structural judgements in Figure
-\ref{fig:ssa-structural-semantics}.
-
-\TODO{some text about what the semantics actually are...}
-
-We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics}
-followed by a semantics for blocks and label-sets in Figure
-\ref{fig:ssa-block-semantics}.
-
-\TODO{ditto}
-
-\TODO{rename blocks to statements? label-sets to CFGs?}
+\(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}: we map tensor products to
+tensor products, the unit type to the monoidal unit, and booleans to \(\mb{1} +
+\mb{1}\). Contexts are simply interpreted as the tensor products of the variable
+types which appear in them; in particular, the quantity annotations have
+\textit{no} effect on a context's semantics. We then give a semantics for
+structural judgements in Figure \ref{fig:ssa-structural-semantics}. In
+particular, we begin by interpreting the judgement that a type is affine,
+\(\tlin{A}{\taff}\), as the drop morphism \(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A},
+I)\) allowing us to discard the type by sending it to the monoidal unit; type
+being relevant, \(\tlin{A}{\trel}\) is similarly interpreted as the split
+morphism \(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, \dnt{A} \otimes \dnt{A})\). We
+can then interpret a weakening \(\csplits{\Gamma}{\Delta}{\Xi}\) as a pure
+morphism \(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\) which
+partitions the variables as appropriate, splitting and dropping as necessary;
+weakenings are then just interpreted as the postcomposition of the associated
+split morphism and the unitor.
 
@@
 
-\TODO{some text about term semantics}
+We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics}.
+
+\TODO{some text about what the semantics actually are...}
 

### 2023-11-15 pldi24/paper.tex
Tap

@@
 structural judgements in Figure \ref{fig:ssa-structural-semantics}. In
-particular, we begin by interpreting the judgement that a type is affine,
-\(\tlin{A}{\taff}\), as the drop morphism \(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A},
-I)\) allowing us to discard the type by sending it to the monoidal unit; type
-being relevant, \(\tlin{A}{\trel}\) is similarly interpreted as the split
-morphism \(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, \dnt{A} \otimes \dnt{A})\). We
-can then interpret a weakening \(\csplits{\Gamma}{\Delta}{\Xi}\) as a pure
+particular, we begin by interpreting the judgements that a type is affine (i.e.
+\(\tlin{A}{\taff}\)) or relevant (i.e. \(\tlin{A}{\trel}\)) as the drop morphism
+\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, I)\) and split morphism
+\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, \dnt{A} \otimes \dnt{A})\) respectively.
+We can then interpret a weakening \(\csplits{\Gamma}{\Delta}{\Xi}\) as a pure
 morphism \(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\) which
@@
 \begin{enumerate}
-  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\) forms a stream action on the one element set \(\mathbf{1}\).
-  \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta = \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
-  \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta = \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i \cdot_i l_i)_i\).
-  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite) sequence \(s\).
-  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts on \(\nats \cup \{\infty\}\) by addition.
+  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\)
+  forms a stream action on the one element set \(\mathbf{1}\).
+  \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and
+  stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta =
+  \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action
+  compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
+  \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action
+  \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta =
+  \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to
+  \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i
+  \cdot_i l_i)_i\).
+  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream
+  action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\),
+  the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell
+  \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite)
+  sequence \(s\).
+  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a
+  stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts
+  on \(\nats \cup \{\infty\}\) by addition.
 \end{enumerate}
@@
 \end{equation}
-We can now define the parallel composition of morphisms in a "fork-join" style as follows: we first flush the buffer completely (i.e., taking it and sticking it at the beginning of our pomset), then execute \(f\) and \(g\) in parallel with separate buffers, \textit{filtering out executions which completely flush the buffer}.
-Since both threads end up with an empty buffer, the resulting joined buffer is also empty, giving us the following:
+We can now define the parallel composition of morphisms in a "fork-join" style
+as follows: we first flush the buffer completely (i.e., taking it and sticking
+it at the beginning of our pomset), then execute \(f\) and \(g\) in parallel
+with separate buffers, \textit{filtering out executions which completely flush
+the buffer}. Since both threads end up with an empty buffer, the resulting
+joined buffer is also empty, giving us the following:
 \begin{equation}
@@
 
-There is a vast literature on SSA and related compiler IRs, but for
-space reasons, we are forced to restrict our discussion to the most
-closely related work.
+There is a vast literature on SSA and related compiler IRs, but for space
+reasons, we are forced to restrict our discussion to the most closely related
+work.
 

### 2023-11-15 pldi24/paper.tex
Tap

@@
   if it is distributive as a (pre)monoidal category, and its distributor is
-  \textit{pure}. A substructural effectful category \(\mc{C}\) is
-  \textbf{distributive} if \(\mc{C}_0\) is distributive (as a substructural
-  (pre)monoidal category).
+  \textit{pure}. A substructural effectful category \(\mc{C}_1 \to \mc{C}_0\) is
+  \textbf{distributive} if \(\mc{C}_1, \mc{C}_0\) are distributive (as a
+  substructural (pre)monoidal category) and \(\upg{\cdot}{}\) strictly preserves
+  all coproducts.
 \end{definition}
@@
 \(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, \dnt{A} \otimes \dnt{A})\) respectively.
-We can then interpret a weakening \(\csplits{\Gamma}{\Delta}{\Xi}\) as a pure
-morphism \(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\) which
-partitions the variables as appropriate, splitting and dropping as necessary;
-weakenings are then just interpreted as the postcomposition of the associated
-split morphism and the unitor.
+We can then interpret a context split \(\csplits{\Gamma}{\Delta}{\Xi}\) as a
+pure morphism \(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\)
+which partitions the variables as appropriate, splitting and dropping as
+necessary. We interpret a context weakening \(\cwk{\Gamma}{\Delta}\) as the
+associated split \(\csplits{\Gamma}{\Delta}{\cdot}\) postcomposed with a unitor.
+Finally, we interpret a label-context weakening \(\lwk{\ms{L}}{\ms{K}}\) as a
+pure morphism \(\mc{C}^\tint_1(\dnt{\ms{L}}, \dnt{\ms{K}})\), where
+\(\rle{join-nil}\) is the identity, \(\rle{join-cons}\) is the sum of
+\(\dnt{\lwk{\ms{L}}{\ms{K}}}\) and \(\dnt{\cwk{\Gamma}{\Delta}}\), and
+\(\rle{join-zero}\) is \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) postcomposed with the left
+injection.
 
@@
     \boxed{\dnt{\lwk{\ms{L}}{\ms{K}}}
-      : \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}})} 
+      : \mc{C}_1^\tint(\dnt{\ms{L}}, \dnt{\ms{K}})} 
       \\
@@
       {\ms{K}, \llhyp{\ell}{\Delta}{A}}}
-      = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\cwk{\Gamma}{\Delta} \otimes \dnt{A})}{} 
+      = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\dnt{\cwk{\Gamma}{\Delta}} \otimes \dnt{A})}{} 
       \\
     \dnt{\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      = \dnt{\lwk{\ms{L}}{\ms{K}}};\rho_+^{-1}
-      ; \dnt{\ms{K}} \oplus 0_{\dnt{\Gamma} \otimes \dnt{A}}
+      = \dnt{\lwk{\ms{L}}{\ms{K}}};\iota_0
   \end{gather*}

### 2023-11-15 pldi24/paper.tex
foo
starting text for related work

@@
   for this judgement are given in Figure \ref{fig:ssa-term-typing}.
+
+  So the centrality $p$ indicates whether it is safe to substitute this expression
+  for a variable, and $q$, plus the linearity of $A$, control how many times that
+  variable can occur.
+
   \item \textit{Blocks} \(s, t\), which are typed via the judgement
@@
   \ref{fig:ssa-block-typing}.
+
 \item \textit{Label-sets} or \textit{control-flow graphs} \(L\), which are
@@
 
-There is a vast literature on SSA and related compiler IRs, but for space
-reasons, we are forced to restrict our discussion to the most closely related
-work.
-
-\paragraph{Weak memory}
-
-\TODO{\citet{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
-\TODO{\citet{paviotti-modular-relaxed-dep-20}}
-\TODO{\citet{sparky}}
-
-\todo{Main thing here to say is that we need semantics for while loops, plus associativity of semicolon, if statements must support distributivity}
-
-\paragraph{Type Theory for SSA}
-
-\TODO{\citet{hua-explicit-ssa-2010}}
-\TODO{\citet{ssa-types-matsuno-ohori-06}}
-\TODO{\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20}}
-
-\paragraph{Functional SSA}
-
-\TODO{\citet{chakravarty-functional-ssa-2003}}
-
-\subparagraph{Operational Semantics}
-
-\TODO{\citet{barthe-compcert-ssa-2014}}
+There is a vast literature on SSA and related compiler IRs, but for
+space reasons, we are forced to restrict our discussion to the most
+closely related work.
+
+Originally, \citet{kelsey-ssa-cps} showed how to embed SSA into a CPS
+representation, and \citet{appel-ssa} showed informally how SSA can be
+seen as a collection of mutually-tail-recursive
+procedures. \citet{chakravarty-functional-ssa-2003} made Appel's
+observation formal, by showing how to translate SSA into A-normal
+form~\cite{anf}, and showed how to use the translation to prove the
+correctness of a constant propagation analysis. These translations
+were for an untyped source and target language, and
+\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20} give a typed
+translation from SSA into the lambda calculus, which also tracks the
+effects of expressions. The effect tracking is handled Koka-style,
+where the effect system tracks a list of effect names. (In contrast,
+we track the centrality and duplicability of terms, without reference
+to the specific effect.)
+
+All of these papers give SSA semantics by translation -- that is,
+they defined the semantics of an SSA program to be the semantics of
+the image of the translation, whether CPS or ANF. There have also
+been a number of papers which have studied the semantics of SSA directly.
+
+\citet{ssa-types-matsuno-ohori-06} give a type theory for what appear
+to be ordinary three-address code programs, where the types of the
+variables at each program point represent the control flows from the
+assignment statements reaching that program point. However, the type
+information means that every well-typed program can be placed into
+SSA-form by inserting $\phi$-nodes in a completely type-directed
+way. This lets them avoid having to give an operational reading of
+$\phi$-nodes, since they can use the standard semantics for
+three-address code.
+
+\citet{hua-explicit-ssa-2010} give another type system for SSA
+programs, as well as a direct operational semantics for it, using the
+standard sequential heap semantics, as well as proving progress and
+preservation for their type system.
+
+There have also been many operational semantics for SSA arising from
+compiler verification efforts. \citet{barthe-compcert-ssa-2014} give
+a
+
+\TODO{}}
 \TODO{\citet{buchwald-verified-ssa-2016}}
@@
 \TODO{\citet{pop-ssa-inout-2009}}
+
+\paragraph{Weak memory}
+
+\TODO{\citet{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
+\TODO{\citet{paviotti-modular-relaxed-dep-20}}
+\TODO{\citet{sparky}}
+
+\todo{Main thing here to say is that we need semantics for while loops, plus associativity of semicolon, if statements must support distributivity}
 

### 2023-11-15 pldi24/paper.tex
more related work

@@
 
-All of these papers give SSA semantics by translation -- that is,
-they defined the semantics of an SSA program to be the semantics of
-the image of the translation, whether CPS or ANF. There have also
-been a number of papers which have studied the semantics of SSA directly.
+All of these papers give SSA semantics by translation -- that is, they
+defined the semantics of an SSA program to be the semantics of the
+image of the translation, with the idea that the simpler semantics of
+the functional language can make proofs easier. The Grail language of
+\citet{beringer-imp-fun} makes this idea explicit by giving both a
+functional and imperative semantics to a low-level language, and
+showing the conditions under which they coincide. \citet{schneider-imp-fun} prove
+a similar result, and further mechanise the proof in Coq.
+
+There have also been a number of papers which have studied the
+semantics of SSA directly.
 
@@
 compiler verification efforts. \citet{barthe-compcert-ssa-2014} give
-a
-
-\TODO{}}
-\TODO{\citet{buchwald-verified-ssa-2016}}
-\TODO{\citet{herklotz-gsa-2023}}
-\TODO{\citet{garbuzov-structural-cfg-2018}}
-\TODO{\citet{vellum}}
-
-\subparagraph{Denotational Semantics}
-
-\TODO{\citet{pop-ssa-inout-2009}}
-
-\paragraph{Weak memory}
+an operational semantics of SSA as part of the CompCertSSA project,
+and prove that the translation from three-address code into SSA is
+semantics preserving. \citet{herklotz-gsa-2023} also formalise an
+extension to SSA, ``gated SSA'', and give it a semantics, and give
+semantics-preserving translations between GSA and ordinary SSA. Going
+beyond CompCertSSA, \citet{vellum} have studied the semantics of the LLVM
+IR itself.
+
+There has been much less work on denotational semantics for SSA, or
+directly on its equational theory. \citet{pop-ssa-inout-2009} give an
+unusual denotational model of SSA in terms of the iteration structure
+of a program, which they use to better understand the loop-closing
+$\phi$-nodes found both in the gated SSA representation as well as
+practical compilers such as GCC.  \citet{garbuzov-structural-cfg-2018}
+exhibit a correspondence between an operational semantics for SSA, and
+an operational semantics for the call-by-push-value calculus of
+\citet{cbpv}. Despite working over an operational semantics, they use
+the normal-form bisimulations of \citet{lassen-bisim} to derive an
+equational theory for use in justifying optimisations. \todo{Their
+operational semantics does contain not contain any effects, which
+means that it cannot justify effect-dependent optimisations such as
+eliminating redundant writes.}
+
+We are the first paper to connect a weak memory semantics to the
+semantics of SSA. In order to establish a rich equational theory,
+we needed a number of features of the semantics which are sometimes
+difficult to come by in weak memory semantics. First, sequencing
+must be associative (i.e., $(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).
+Second, the semantics must support loops, including the equation
+that a while-loop is equal to its unrolling. Finally, booleans need
+to be distributive (i.e., $(A \otimes \mb{2}) \simeq (A + A)$).
 

### 2023-11-15 pldi24/paper.tex
finished first draft of related work

@@
 the normal-form bisimulations of \citet{lassen-bisim} to derive an
-equational theory for use in justifying optimisations. \todo{Their
+equational theory for use in justifying optimisations. Their
 operational semantics does contain not contain any effects, which
 means that it cannot justify effect-dependent optimisations such as
-eliminating redundant writes.}
-
-We are the first paper to connect a weak memory semantics to the
-semantics of SSA. In order to establish a rich equational theory,
-we needed a number of features of the semantics which are sometimes
-difficult to come by in weak memory semantics. First, sequencing
-must be associative (i.e., $(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).
-Second, the semantics must support loops, including the equation
-that a while-loop is equal to its unrolling. Finally, booleans need
-to be distributive (i.e., $(A \otimes \mb{2}) \simeq (A + A)$).
-
-\TODO{\citet{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
-\TODO{\citet{paviotti-modular-relaxed-dep-20}}
-\TODO{\citet{sparky}}
-
-\todo{Main thing here to say is that we need semantics for while loops, plus associativity of semicolon, if statements must support distributivity}
-
-% \TODO{\cite{schneider-phd}}
-% \TODO{\cite{rvsdg}}
-
-\TODO{this}
+eliminating redundant writes.
+
+Our work is the first to validate the equational theory of SSA against
+a weak memory semantics. We focused on denotational models of weak
+memory, since they make it easier to establish equational properties
+than the operational and axiomatic models. However, for our proof to
+go through, we needed a number of features of the semantics which are
+sometimes difficult to establish in the weak memory world. First,
+sequencing must be associative (i.e.,
+$(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).  Second, the semantics must
+support loops, including the equation that a while-loop is equal to
+its unrolling. Finally, sums need to be distributive (e.g., in the
+boolean case, $(A \otimes \mb{2}) \simeq (A + A)$).
+
+As a result, we focused on the TSO semantics of \citet{sparky}, which
+does have all the properties we need. For the weaker ARM-style models
+(which admit load buffering), there seem to be a number of models
+which are "near misses", in that they have some, but not all, the
+properties we need. The event structure model of
+\citet{paviotti-modular-relaxed-dep-20} is formulated in terms of
+step-indexing, so satisfies all of our needed properties except the
+loop unrolling property -- step-indexing means it is a refinement
+rather than an equation. The pomsets with preconditions model of
+\citet{jagadeesan-pwp-20} does not exactly satisfy the associativity
+of semicolon, and while the pomsets with predicate transformers model
+of \citet{leaky-semicolon} does satisfy associativity, it does not yet
+support loops.
+
+We do not think it would be prohibitively difficult to extend these
+models to support the required features, and we believe our results
+suggest that it would be have a high payoff: those features would
+suffice to build a model of our IR, giving us tight results
+validating the transformations compilers want against the low-level
+specification of the actual machine.
 

### 2023-11-15 pldi24/paper.tex
Tap

@@
 pure morphism \(\mc{C}^\tint_1(\dnt{\ms{L}}, \dnt{\ms{K}})\), where
-\(\rle{join-nil}\) is the identity, \(\rle{join-cons}\) is the sum of
+\rle{join-nil} is the identity, \rle{join-cons} is the sum of
 \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) and \(\dnt{\cwk{\Gamma}{\Delta}}\), and
-\(\rle{join-zero}\) is \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) postcomposed with the left
+\rle{join-zero} is \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) postcomposed with the left
 injection.
@@
         \dnt{\hasty{\Xi}{1}{b}{B}{q}}
-      )}{\uparrow p} 
+      )}{p} 
       \\
@@
 We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics}.
+Our treatment of \rle{var} is quite standard: the expression \(x\) types in
+\(\Gamma\) if all other variables in \(\Gamma\) and can be discarded, in which
+case the semantics is simply doing so, i.e.
+\(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\), lifting to \(\mc{C}_0\) as necessary.
+Similarly, to type a function application \rle{var}, we first take
+\(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is guarnateed to be central, lift it
+to the appropriate category \(\mc{C}_p\), and then simply postcompose it with
+the denotation of \(f\). \rle{pair} is handled by composing the context split
+\(\csplits{\Gamma}{\Delta}{\Xi}\) taking the tensor product of the components
+\(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\), \(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is
+well-defined since both components are central and \(\mc{C}_1\) is monoidal. We
+then lift the result to the appropriate \(\mc{C}_p\). The semantics of
+\rle{unit} is simply dropping the entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\),
+while \rle{true}, and \rle{false} are each handled by first dropping the context
+and then postcomposing with the appropriate constant; in all cases, this is
+followed by lifting to \(\mc{C}_p\). Finally, \rle{let} and \rle{let2} are
+handled by first splitting the context, then passing the right hand component to
+the semantics of the bound expression \(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or
+\(\dnt{\hasty{\Xi}{1}{a}{A \otimes B}{q}}\), and then, lifting to \(\mc{C}_p\)
+as necessary, after reassociating with \(\alpha\), passing everything to the
+inner expression \(\dnt{\hasty{\Delta, \thyp{x}{A}{q}}{p}{e}{B}{q}}\) or
+\(\dnt{\hasty{\Delta, \thyp{x}{A}{q}, \thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that
+this inner expression is not necessarily central, unlike the bound expression!
 

### 2023-11-16 pldi24/paper.tex
Tap

@@
 Our main metatheoretic results will be \textit{substitution} and
-\textit{label-substitution}, which will require introducing some
-substitution judgements to state properly, which we do in Figure
+\textit{label-substitution}, which will require introducing some substitution
+judgements to state properly, which we do in Figure
 \ref{fig:ssa-meta-judgements}. In particular, the judgement
-\(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which is
-taken to be a map from variable names to expressions, is a valid
-substitution taking \(\Theta\) to \(\Gamma\). What this means is that,
-for each variable \(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can
-be interpreted as a \textit{central} term of type \(A^q\) in a
-subcontext of \(\Theta\), such that when this is done for all
-\(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables in \(\Theta\)
-are used up: this is described formally via the typing rules
-\rle{subst-nil}, which says that any function \(\gamma\) takes the
-empty context to itself, and \rle{subst-cons}, which says that if
-\(\gamma\) takes \(\Theta_\Gamma\) to \(\Gamma\),
-\(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\),
-\(q \subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the
-duplication/discarding of substructural variables), and \(\Theta\)
-splits into \(\Theta_\Gamma\) and \(\Theta_x\), then
-\(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).
+\(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which is taken to be
+a map from variable names to expressions, is a valid substitution taking
+\(\Theta\) to \(\Gamma\). What this means is that, for each variable
+\(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted as a
+\textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such that
+when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables
+in \(\Theta\) are used up: this is described formally via the typing rules
+\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
+to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
+\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
+\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/discarding
+of substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
+\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).
 
@@
 computations can be used in an expression substituting for an affine variable,
-and only relevant computationbs can be used in an expression substituting
-for a relevant variable.
+and only relevant computationbs can be used in an expression substituting for a
+relevant variable.
 
@@
 
-\TODO{some text about what the semantics actually are...}
-
 \begin{figure}
@@
 \end{figure}
+
+We can now give a semantics for blocks in Figure \ref{fig:ssa-block-semantics}.
+Our semantics for break statements \(\lbrb{\ell}{a}\) is given by the semantics
+for splitting the context into variables which pass into dominated blocks and
+variables used to compute the argument, \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\),
+with the latter passed into the semantics of the argument (which is pure, and
+therefore must be lifted to \(\mc{C}_0\))
+\(\dnt{\hasty{\Gamma}{1}{a}{A}{\tint}}\), with the result along with the
+preserved variables passed into \(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to
+obtain the desired target. The semantics of if-statements are given by first
+splitting the context via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\) and then
+calculating the discriminant \(\dnt{\hasty{\Delta}{1}{b}{\mb{2}}{\tint}}\),
+which must be pure and therefore needs to be lifted. This is then passed through
+the distributor \(\delta\), with each branch passed to the semantics of its
+corresponding block \(\dnt{\haslb{\Xi}{s}{\ms{L}}}\),
+\(\dnt{\haslb{\Xi}{t}{\ms{L}}}\).Tap
 

### 2023-11-16 pldi24/paper.tex
Added addresses, formatted, tap

@@
 semantics preserving. \citet{herklotz-gsa-2023} also formalise an
-extension to SSA, ``gated SSA'', and give it a semantics, and give
+extension to SSA, "gated SSA", and give it a semantics, and give
 semantics-preserving translations between GSA and ordinary SSA. Going

### 2023-11-16 pldi24/paper.tex
Tap

@@
 
+\TODO{terms vs expression/blocks/label-sets}
+
 \TODO{standardize "branching to" wordmark? Is wordmark the word here?}
@@
   \end{gather*}
-  \caption{Typing rules for \isotopessa terms}
-  \Description{Typing rules for isotope-SSA terms}
+  \caption{Typing rules for \isotopessa expressions}
+  \Description{Typing rules for isotope-SSA expressions}
   \label{fig:ssa-term-typing}
@@
   \end{gather*}
-  \caption{Semantics for \isotopessa terms}
-  \Description{Semantics for isotope-SSA terms}
+  \caption{Semantics for \isotopessa expressions}
+  \Description{Semantics for isotope-SSA expressions}
   \label{fig:ssa-term-semantics}
@@
 corresponding block \(\dnt{\haslb{\Xi}{s}{\ms{L}}}\),
-\(\dnt{\haslb{\Xi}{t}{\ms{L}}}\).Tap
+\(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The semantics of \rle{let-blk} and \rle{let2-blk}
+are analogous to those of \rle{let} and \rle{let2} in the expression case, with the
 
@@
   \end{grammar}
-  \caption{Grammar for \isotopessa terms and blocks with holes}
+  \caption{Grammar for \isotopessa expressions and blocks with holes}
   \Description{Grammar for isotope-SSA blocks with holes}
@@
   \end{gather*}
-  \caption{Typing rules for \isotopessa terms with holes}
-  \Description{Typing rules for isotope-SSA terms with holes}
+  \caption{Typing rules for \isotopessa expressions with holes}
+  \Description{Typing rules for isotope-SSA expressions with holes}
   \label{fig:terms-with-holes-typing}
@@
   \end{gather*}
-  \caption{Semantics for \isotopessa terms with holes}
-  \Description{Semantics for isotope-SSA blocks with holes}
+  \caption{Semantics for \isotopessa expressions with holes}
+  \Description{Semantics for isotope-SSA expressions with holes}
   \label{fig:terms-with-holes-semantics}

### 2023-11-16 pldi24/paper.tex
Tap

@@
 corresponding block \(\dnt{\haslb{\Xi}{s}{\ms{L}}}\),
-\(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The semantics of \rle{let-blk} and \rle{let2-blk}
-are analogous to those of \rle{let} and \rle{let2} in the expression case, with the
-
-\TODO{some text about block semantics}
+\(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The semantics of \rle{let-blk} and
+\rle{let2-blk} are analogous to those of \rle{let} and \rle{let2} in the
+expression case, except that now the bound expression \(a\) is allowed to be
+impure (and therefore in particular does not need to be upgraded).
+
+Finally, we must give semantics for \rle{where}. To do so, we begin by giving a
+semantics for control-flow graphs: \todo{this}. We can now quite easily state
+the semantics of \rle{where}: we simply precompose the semantics of the inner
+block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the semantics of the associated
+control-flow graph \(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}\); \todo{explain}
 

### 2023-11-16 pldi24/paper.tex
Proved substitution splitting

@@
 \end{itemize}
+
+\TODO{We know \textit{what} these semantics are, unlike below, but maybe clarify
+\textit{why}}
+
 We begin by giving a semantics for types and contexts in terms of objects in
@@
 \end{figure}
+
+\TODO{\textit{what} is the semantics of a term}
 
@@
 
+\TODO{\textit{what} is the semantics of a block}
+
 We can now give a semantics for blocks in Figure \ref{fig:ssa-block-semantics}.
@@
 Finally, we must give semantics for \rle{where}. To do so, we begin by giving a
-semantics for control-flow graphs: \todo{this}. We can now quite easily state
-the semantics of \rle{where}: we simply precompose the semantics of the inner
-block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the semantics of the associated
-control-flow graph \(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}\); \todo{explain}
+semantics for control-flow graphs: \rle{nil-br} simply passes the input context
+\(\ms{L}\) through unchanced, while \rle{cons-br} takes any calls to
+\(\lbl{\ell}\) and passes them to the denotation
+\(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\), passing the rest of \(\ms{L}\) to
+\(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}\). We can now quite easily state the
+semantics of \rle{where}: we simply precompose the semantics of the inner block
+\(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the semantics of the associated
+control-flow graph \(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}\).
+
+\TODO{explain what the above actually means, \textit{what} is the semantics of a
+CFG}
 
@@
   \begin{itemize}
-    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\), \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
-    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
-    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then \(\haslb{\Gamma}{t}{\ms{K}}\)
+    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
+    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
+    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then
+    \(\haslb{\Gamma}{t}{\ms{L}}\)
+    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then
+    \(\haslb{\Gamma}{t}{\ms{K}}\)
   \end{itemize}
@@
 
-\begin{proof} \
-  \TODO{this}
+\begin{proof} 
+  We proceed by induction on derivations \(\csplits{\Gamma}{\Delta}{\Xi}\). In
+  the \rle{split-nil} case, we have that \(\issubst{\gamma}{\cdot}{\cdot}\) by
+  \rle{subst-nil}. For all other cases, since \(\issubst{\gamma}{\Theta}{\Gamma,
+  \thyp{x}{A}{q}}\), we have by \rle{subst-cons} that
+  \(\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}\),
+  \(\issubst{\gamma}{\Theta_\Gamma}{\Gamma}\),
+  \(\hasty{\Theta_\Gamma}{1}{\gamma(x)}{A}{q}\). Since
+  \(\csplits{\Gamma}{\Delta}{\Xi}\), by induction, we have that
+  \(\csplits{\Theta_\Gamma}{\Theta_\Delta}{\Theta_\Xi}\) where \(\Theta_\Delta\)
+  and \(\Theta_\Xi\) have the desired properties above.
+  \begin{itemize}
+    \item \rle{split-left}: by left split association, we have that there exists
+    \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\csplits{\Theta}{\Theta_{\Delta,
+    \thyp{x}{A}{q}}}{\Theta_\Xi}\), \(\csplits{\Theta_{\Delta,
+    \thyp{x}{A}{q}}}{\Theta_\Delta}{\Theta_x}\). Using the fact that
+    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), we can conclude by
+    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Delta,
+    \thyp{x}{A}{q}}}{\Delta, \thyp{x}{A}{q}}\); the desired result hence
+    follows.
+    \item \rle{split-right}: by right split association, we have that there
+    exists \(\Theta_{\Xi, \thyp{x}{A}{q}}\),
+    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_{\Xi, \thyp{x}{A}{q}}}\),
+    \(\csplits{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Theta_\Xi}{\Theta_x}\). Using the
+    fact that \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by
+    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
+    \thyp{x}{A}{q}}\); the desired result hence follows.
+    \item \rle{split-dup}: since \(\tlin{A}{\trel}\), it follows by
+    \rle{subst-cons} that \(\tlin{\Theta_x}{\trel}\), and hence that
+    \(\csplits{\Theta_x}{\Theta_x}{\Theta_x}\). Therefore, by split association,
+    we have that there exist \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\Theta_{\Xi,
+    \thyp{x}{A}{q}}\) such that \(\csplits{\Theta}{\Theta_{\Delta,
+    \thyp{x}{A}{q}}}{\Theta_{\Xi, \thyp{x}{A}{q}}}\). Using the facts that
+    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\),
+    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by \rle{subst-cons}
+    that \(\issubst{\gamma}{\Theta_{\Delta, \thyp{x}{A}{q}}}{\Delta,
+    \thyp{x}{A}{q}}\), \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
+    \thyp{x}{A}{q}}\); the desired result hence follows.
+    \item \rle{split-drop}: since \(\tlin{A}{\taff}\), it follows by
+    \rle{subst-cons} that \(\tlin{\Theta_x}{\taff}\), and hence that
+    \(\cwk{\Theta_x}{\cdot}\). Therefore, by split association,
+    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\), giving the desired result.
+  \end{itemize}
 \end{proof}
@@
 \end{proof}
+
+\section{Loop Hoisting}
+
+\TODO{this?}
 

### 2023-11-16 pldi24/paper.tex
Alas for POSIX!

@@
 is simply defined as \(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where
-\(\Sigma: (\ms{Bytes}^*)^\omega \to \ms{Bytes}^{\leq \omega}\) denotes
-concatenation of lists of bytestrings. This can be equipped with the structure
-of a substructural effectful category by choosing \(M_C = M_R = \{[]\}\), giving
-us an \isotopessa model supporting the instructions \(\ms{print} \in
-\mc{I}^\varnothing_0(\ms{Bytes}, \mb{1})\) and \(\ms{rand} \in
-\mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard semantics
-\(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), [b])\}\), \(\dnt{\ms{rand}} =
-\lambda (). \{\iota_0 (a, p[]) \mid a \in A\}\).
+\(\Sigma: (\ms{Byte}^*)^\omega \to \ms{Byte}^{\leq \omega}\) denotes
+concatenation of bytestrings into a potentially infinite bytestream. This can be
+equipped with the structure of a substructural effectful category by choosing
+\(M_C = M_R = \{[]\}\), giving us an \isotopessa model supporting the
+instructions \(\ms{print} \in \mc{I}^\varnothing_0(\ms{Byte}^*, \mb{1})\) and
+\(\ms{rand} \in \mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard
+semantics \(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), b)\}\),
+\(\dnt{\ms{rand}} = \lambda (). \{\iota_0 (a, []) \mid a \in A\}\).
 
@@
   \Phi\;a \implies \Psi\;a\), implemented by the identity morphism
-  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Bytes}, \ms{emp}), (\mb{1},
+  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Byte}^*, \ms{emp}), (\mb{1},
   \ms{emp}))\)

### 2023-11-16 pldi24/paper.tex
Tap

@@
 
-\TODO{\textit{what} is the semantics of a term}
-
-We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics}.
-Our treatment of \rle{var} is quite standard: the expression \(x\) types in
-\(\Gamma\) if all other variables in \(\Gamma\) and can be discarded, in which
-case the semantics is simply doing so, i.e.
+We can now interpret a term \(\hasty{\Gamma}{p}{a}{A}{q}\) as a
+\textit{morphism} taking \(\dnt{\Gamma}\) into \(\dnt{A}\) with
+\textit{centrality} \(p\) and \textit{quantity} \(q\) (i.e., lying in
+\(\mc{C}_p^q\)). This means that, as expected, a central term becomes a central
+morphism in \(\mc{C}_1\), an affine term becomes an affine morphism, and a
+relevant term becomes a relevant morphism, and so, in particular, a pure term
+becomes a pure morphsim. Our rules for doing so are given in Figure
+\ref{fig:ssa-term-semantics}. Our treatment of \rle{var} is quite standard: the
+expression \(x\) types in \(\Gamma\) if all other variables in \(\Gamma\) and
+can be discarded, in which case the semantics is simply doing so, i.e.
 \(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\), lifting to \(\mc{C}_0\) as necessary.

### 2023-11-16 pldi24/paper.tex
Swapped contexts for ite, tap

@@
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\hasty{\Delta}{1}{e}{\mb{2}}{\tint}}
-      {\haslb{\Xi}{s}{\ms{L}}}
-      {\haslb{\Xi}{t}{\ms{L}}}
+      {\hasty{\Xi}{1}{e}{\mb{2}}{\tint}}
+      {\haslb{\Delta}{s}{\ms{L}}}
+      {\haslb{\Delta}{t}{\ms{L}}}
       {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} 
@@
       = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\hasty{\Delta}{1}{e}{\mb{2}}{\tint}} \otimes \dnt{\Xi})}
-      ; \ms{ite}_{\dnt{\Xi}} ;
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{e}{\mb{2}}{\tint}})}
+      ; \delta ;
       \\ \qquad \qquad \qquad \qquad
       [
-        \dnt{\haslb{\Xi}{s}{\ms{L}}}, 
-        \dnt{\haslb{\Xi}{t}{\ms{L}}} 
+        \dnt{\haslb{\Delta}{s}{\ms{L}}}, 
+        \dnt{\haslb{\Delta}{t}{\ms{L}}} 
       ]
@@
 
-\TODO{\textit{what} is the semantics of a block}
-
-We can now give a semantics for blocks in Figure \ref{fig:ssa-block-semantics}.
-Our semantics for break statements \(\lbrb{\ell}{a}\) is given by the semantics
-for splitting the context into variables which pass into dominated blocks and
-variables used to compute the argument, \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\),
-with the latter passed into the semantics of the argument (which is pure, and
-therefore must be lifted to \(\mc{C}_0\))
-\(\dnt{\hasty{\Gamma}{1}{a}{A}{\tint}}\), with the result along with the
-preserved variables passed into \(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to
-obtain the desired target. The semantics of if-statements are given by first
-splitting the context via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\) and then
-calculating the discriminant \(\dnt{\hasty{\Delta}{1}{b}{\mb{2}}{\tint}}\),
-which must be pure and therefore needs to be lifted. This is then passed through
-the distributor \(\delta\), with each branch passed to the semantics of its
-corresponding block \(\dnt{\haslb{\Xi}{s}{\ms{L}}}\),
-\(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The semantics of \rle{let-blk} and
-\rle{let2-blk} are analogous to those of \rle{let} and \rle{let2} in the
-expression case, except that now the bound expression \(a\) is allowed to be
-impure (and therefore in particular does not need to be upgraded).
+Similarly to terms, \textit{blocks} \(\haslb{\Gamma}{t}{\ms{L}}\) are
+interpreted as \textit{impure} morphisms from \(\dnt{\Gamma}\) to \(\ms{L}\):
+here \(\Gamma\) represents all the variables live on entry to the block, while
+\(\ms{L}\) represents all the external blocks which \(t\) might branch to, each
+annotated with their live variables and their arguments; our rules are given in
+Figure \ref{fig:ssa-block-semantics}. Our semantics for break statements
+\(\lbrb{\ell}{a}\) is as follows: we begin by splitting the context into live
+variables which pass into dominated blocks and variables used in the argument
+via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\). We then pass the latter into the
+semantics of the argument (which is pure, and therefore must be lifted to
+\(\mc{C}_0\)) \(\dnt{\hasty{\Gamma}{1}{a}{A}{\tint}}\), with the result and live
+variables re-associated and passed into
+\(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to obtain the desired target. The
+semantics of if-statements are similarly given by first splitting the context
+via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\) into variables live in the branches,
+\(\Delta\), and variables appearing in the discriminant, \(\Xi\); the latter of
+which are passed into the lifted pure morphism
+\(\dnt{\hasty{\Delta}{1}{b}{\mb{2}}{\tint}}\). The "if" statement itself is then
+simply an application of the distributor \(\delta: A \otimes (\mb{1} + \mb{1})
+\to A + A\), with each branch passed to the semantics of its corresponding block
+\(\dnt{\haslb{\Xi}{s}{\ms{L}}}\), \(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The
+semantics of \rle{let-blk} and \rle{let2-blk} are analogous to those of
+\rle{let} and \rle{let2} in the expression case, except that now the bound
+expression \(a\) is allowed to be impure (and therefore in particular does not
+need to be upgraded).
 
@@
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta}{1}{\mhole{e}}{\mb{2}}{\tint}}
-      {\mhaslb{H}{\Xi}{\mhole{s}}{\ms{L}}}
-      {\mhaslb{H}{\Xi}{\mhole{t}}{\ms{L}}}
+      {\mhasty{H}{\Xi}{1}{\mhole{e}}{\mb{2}}{\tint}}
+      {\mhaslb{H}{\Delta}{\mhole{s}}{\ms{L}}}
+      {\mhaslb{H}{\Delta}{\mhole{t}}{\ms{L}}}
       {\mhaslb{H}{\Gamma}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\

### 2023-11-16 pldi24/paper.tex
Tap

@@
       = \dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      ; [[\iota_0, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}], \iota_1]
+      ; [\dnt{\ms{K}} + \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}, \iota_1]
       \\
@@
 Finally, we must give semantics for \rle{where}. To do so, we begin by giving a
-semantics for control-flow graphs: \rle{nil-br} simply passes the input context
-\(\ms{L}\) through unchanced, while \rle{cons-br} takes any calls to
-\(\lbl{\ell}\) and passes them to the denotation
-\(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\), passing the rest of \(\ms{L}\) to
-\(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}\). We can now quite easily state the
-semantics of \rle{where}: we simply precompose the semantics of the inner block
-\(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the semantics of the associated
-control-flow graph \(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}\).
-
-\TODO{explain what the above actually means, \textit{what} is the semantics of a
-CFG}
+interpreting a control-flow graph \(\lhaslb{\ms{L}}{L}{\ms{K}}\) as a morphism
+from \(\dnt{\ms{L}}\) to \(\dnt{\ms{K}} + \dnt{\ms{L}}\), where the left-hand
+side of the coproduct denotes branching to one of the target labels in
+\(\ms{K}\) (the \textit{output}), while the right-hand side of the coproduct
+denotes a recursive call to a label in \(\ms{L}\) (the \textit{recursive
+input}). In particular, \rle{nil-br} simply treats a call to one of \(\ms{L}\)
+as a call to a \textit{target label}, passing it through the left injection
+unchanged. On the other hand \rle{cons-br} is a little bit more complicated:
+first we execute \(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}\),
+which will give us either an output \(\dnt{\ms{K}, \lhyp{\ell}{\Gamma}{A}}\), or
+a recursive input \(\dnt{\ms{L}}\), the latter of which we can pass through
+unchanged as a recursive input. In the output case, we can interpret this as
+either a value of \(\dnt{\ms{K}}\) or \(\dnt{\Gamma} \otimes \dnt{A}\); the
+former we can pass the value through unchanged as output; in the second case, we
+can pass this in as input to \(\dnt{\haslb{\Gamma,
+\thyp{x}{A}{q}}{t}{\ms{L}}}\), which will give us a value of \(\ms{L}\) we can
+pass in as recursive input. Note that, even if \(t\) branches to an output label
+in \(\ms{K}\), we still return a \textit{recursive input}, not an output; in
+this case, the \textit{next} iteration will return it as an output. We can now
+quite easily state the semantics of \rle{where}: we simply precompose the
+semantics of the inner block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the
+\textit{iterated} semantics of the associated control-flow graph,
+\(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger\).
 

### 2023-11-16 pldi24/paper.tex
Tap

@@
 quite easily state the semantics of \rle{where}: we simply precompose the
-semantics of the inner block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the
+semantics of the entry block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the
 \textit{iterated} semantics of the associated control-flow graph,

### 2023-11-16 pldi24/paper.tex
Tap

@@
 becomes a pure morphsim. Our rules for doing so are given in Figure
-\ref{fig:ssa-term-semantics}. Our treatment of \rle{var} is quite standard: the
-expression \(x\) types in \(\Gamma\) if all other variables in \(\Gamma\) and
-can be discarded, in which case the semantics is simply doing so, i.e.
+\ref{fig:ssa-term-semantics}; we will abuse syntax slightly and write
+\(\dnt{\text{"judgement"}}\) for \(\dnt{D}\), where \(D\) is a derivation of
+"judgement"; this is justified by the coherence lemma:
+\begin{lemma}[Coherence] \
+  \begin{itemize}
+    \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{lemma}
+Our treatment of \rle{var} is quite standard: the expression \(x\) types in
+\(\Gamma\) if all other variables in \(\Gamma\) and can be discarded, in which
+case the semantics is simply doing so, i.e.
 \(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\), lifting to \(\mc{C}_0\) as necessary.
@@
 
+Now that we've given a semantics to \isotopessa programs, we can begin to use it
+to develop an equational theory for performing optimizations. Since we're
+approaching things from a type-theoretic perspective...
+
 \TODO{some text about semantic metatheory}
 
-\begin{lemma}[Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\TODO{some text about coherence, implicit use above}
-
-\begin{lemma}[Semantic Upgrade]
-  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
-  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
-\end{lemma}
-
-\TODO{some text about upgrade}
+\TODO{some text about semantic weakening}
 
@@
 
-\TODO{some text about semantic weakening}
+\TODO{some text about upgrade}
+
+\begin{lemma}[Semantic Upgrade]
+  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
+  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
+\end{lemma}
 
@@
 \TODO{some text about substitution semantics}
+
+\TODO{some text about semantic substitution}
 
@@
 \end{theorem}
-
-\TODO{some text about semantic substitution}
-
-% \subsection{Normalization}
-
-% \label{ssec:ssa-norm}
-
-% \TODO{this}
 

### 2023-11-16 pldi24/paper.tex
Tap

@@
 to develop an equational theory for performing optimizations. Since we're
-approaching things from a type-theoretic perspective...
-
-\TODO{some text about semantic metatheory}
-
-\TODO{some text about semantic weakening}
-
+approaching things from a type-theoretic perspective, we begin by addressing the
+semantic versions of the standard syntactic metatheorems of weakening and
+substitution. In particular, semantic weakening can be stated in the expected
+manner: if we have an expression or block which is well typed in both \(\Gamma\)
+and \(\Delta\), where \(\cwk{\Gamma}{\Delta}\), then evaluating it in \(\Gamma\)
+is the same as first strengthening \(\Gamma\) to \(\Delta\), dropping unused
+affine variables, and then evaluating it in \(\Delta\).
 \begin{lemma}[Semantic Weakening] 
@@
   \end{itemize}
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
-  \begin{itemize} 
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), for all
+  \(\haslb{\Gamma}{t}{\ms{L}}\), we have
   \[
@@
   \]
-    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), we have
-    \[
-      \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      = \dnt{\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{L}}}
-    \]
-  \end{itemize}
+  \TODO{fix broken cfg version}
 \end{theorem}
@@
   \end{itemize}
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
-  \begin{itemize} 
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), for all 
+  \(\haslb{\Gamma}{t}{\ms{L}}\), we have
   \[
@@
   \]
-    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), we have
-    \[
-      \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      = \dnt{\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{L}}}
-    \]
-  \end{itemize}
 \end{theorem}

### 2023-11-16 pldi24/paper.tex
Tap
shrink related work

@@
   \end{itemize}  
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), for all
-  \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
+  \begin{itemize} 
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
     \[
@@
     \]
-  \TODO{fix broken cfg version}
+    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
+    \[
+      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
+      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
+    \]
+    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
+  \end{itemize}
 \end{theorem}
@@
 
-There is a vast literature on SSA and related compiler IRs, but for
-space reasons, we are forced to restrict our discussion to the most
-closely related work.
+For space reasons we, restrict our discussion to the most closely
+related work on SSA.
 
@@
 observation formal, by showing how to translate SSA into A-normal
-form~\cite{anf}, and showed how to use the translation to prove the
-correctness of a constant propagation analysis. These translations
-were for an untyped source and target language, and
+form~\cite{anf}, and used it to prove the correctness of a constant
+propagation analysis. These translations were for an untyped source
+and target language, and
 \citet{typed-effect-ssa-rigon-torrens-vasconcellos-20} give a typed
-translation from SSA into the lambda calculus, which also tracks the
-effects of expressions. The effect tracking is handled Koka-style,
-where the effect system tracks a list of effect names. (In contrast,
-we track the centrality and duplicability of terms, without reference
-to the specific effect.)
+translation from SSA into the lambda calculus, into a type-and-effect
+system. In contrast, we directly track the centrality and duplicability of
+terms, without tracking specific effects.
 
@@
 \citet{ssa-types-matsuno-ohori-06} give a type theory for what appear
-to be ordinary three-address code programs, where the types of the
-variables at each program point represent the control flows from the
-assignment statements reaching that program point. However, the type
-information means that every well-typed program can be placed into
-SSA-form by inserting $\phi$-nodes in a completely type-directed
-way. This lets them avoid having to give an operational reading of
-$\phi$-nodes, since they can use the standard semantics for
-three-address code.
-
-\citet{hua-explicit-ssa-2010} give another type system for SSA
-programs, as well as a direct operational semantics for it, using the
-standard sequential heap semantics, as well as proving progress and
-preservation for their type system.
-
-There have also been many operational semantics for SSA arising from
-compiler verification efforts. \citet{barthe-compcert-ssa-2014} give
-an operational semantics of SSA as part of the CompCertSSA project,
-and prove that the translation from three-address code into SSA is
-semantics preserving. \citet{herklotz-gsa-2023} also formalise an
-extension to SSA, "gated SSA", and give it a semantics, and give
-semantics-preserving translations between GSA and ordinary SSA. Going
-beyond CompCertSSA, \citet{vellum} have studied the semantics of the LLVM
-IR itself.
+to be ordinary three-address code programs. However, every well-typed
+program can be placed into SSA-form by inserting $\phi$-nodes in a
+fully type-directed way. This lets them model SSA without any $\phi$-nodes, lets
+them use the standard
+semantics for three-address code.
+%
+\citet{hua-explicit-ssa-2010} give another type system and
+direct operational semantics for SSA, and prove type safety
+for it.
+
+Many operational semantics for SSA have arisen from compiler
+verification efforts. \citet{barthe-compcert-ssa-2014} give an
+operational semantics as part of the CompCertSSA project, and give a
+semantics-preserving translation from three-address code into
+SSA. \citet{herklotz-gsa-2023} formalise "gated SSA" and give
+semantics-preserving translations between it and ordinary SSA. Going
+beyond CompCertSSA, \citet{vellum} have studied the semantics of the
+LLVM IR itself.
 
@@
 exhibit a correspondence between an operational semantics for SSA, and
-an operational semantics for the call-by-push-value calculus of
-\citet{cbpv}. Despite working over an operational semantics, they use
+an operational semantics for call-by-push-value~\cite{cbpv}. They use
 the normal-form bisimulations of \citet{lassen-bisim} to derive an
 equational theory for use in justifying optimisations. Their
-operational semantics does contain not contain any effects, which
-means that it cannot justify effect-dependent optimisations such as
-eliminating redundant writes.
+operational semantics is effect-free, which means that it cannot
+justify optimisations such as eliminating redundant
+writes.
 
@@
 a weak memory semantics. We focused on denotational models of weak
-memory, since they make it easier to establish equational properties
-than the operational and axiomatic models. However, for our proof to
-go through, we needed a number of features of the semantics which are
-sometimes difficult to establish in the weak memory world. First,
-sequencing must be associative (i.e.,
+memory, since they make it easier to establish equational
+properties. However, we need a number of features of the semantics
+which are sometimes difficult to establish in the weak memory
+world. First, sequencing must be associative (i.e.,
 $(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).  Second, the semantics must
@@
 As a result, we focused on the TSO semantics of \citet{sparky}, which
-does have all the properties we need. For the weaker ARM-style models
-(which admit load buffering), there seem to be a number of models
-which are "near misses", in that they have some, but not all, the
-properties we need. The event structure model of
-\citet{paviotti-modular-relaxed-dep-20} is formulated in terms of
-step-indexing, so satisfies all of our needed properties except the
-loop unrolling property -- step-indexing means it is a refinement
-rather than an equation. The pomsets with preconditions model of
-\citet{jagadeesan-pwp-20} does not exactly satisfy the associativity
-of semicolon, and while the pomsets with predicate transformers model
-of \citet{leaky-semicolon} does satisfy associativity, it does not yet
-support loops.
+validates the properties we need. For ARM-style concurrency, which
+admits load buffering, we only found models which are "near misses":
+they have a subset of the properties we need. The event
+structure model of \citet{paviotti-modular-relaxed-dep-20} is
+formulated in terms of step-indexing, so satisfies all of our needed
+properties except the loop unrolling property -- step-indexing means
+it is a refinement rather than an equation. The pomsets with
+preconditions model of \citet{jagadeesan-pwp-20} does not exactly
+satisfy the associativity of semicolon, and while the pomsets with
+predicate transformers model of \citet{leaky-semicolon} does satisfy
+associativity, it does not yet support loops.
 
@@
   \end{itemize}
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), for all 
-  \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
+  \begin{itemize} 
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
     \[
@@
     \]
+    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
+    \[
+      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
+      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
+    \]
+    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
+  \end{itemize}
 \end{theorem}

### 2023-11-16 pldi24/paper.tex
Added optimisation text

@@
 \item Our abstract semantics justifies a variety of powerful optimisation
-techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and
-strength reduction. 
+techniques, inlining, loop fusion, hoisting, and strength reduction. 
 \item We show how Elgot monads give rise to instances of our categorical
@@
 semantic versions of the standard syntactic metatheorems of weakening and
-substitution. In particular, semantic weakening can be stated in the expected
-manner: if we have an expression or block which is well typed in both \(\Gamma\)
-and \(\Delta\), where \(\cwk{\Gamma}{\Delta}\), then evaluating it in \(\Gamma\)
-is the same as first strengthening \(\Gamma\) to \(\Delta\), dropping unused
-affine variables, and then evaluating it in \(\Delta\).
-\begin{lemma}[Semantic Weakening] 
-  If \(\cwk{\Gamma}{\Delta}\), then 
-  \begin{itemize}
-    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
-    \[
-      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
-      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-    \]
-    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
-    \[
-      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
-      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
-    \]
-  \end{itemize}
-
-  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
-  \[
-    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
-    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
-  \]
-\end{lemma}
-
-\TODO{some text about upgrade}
-
-\begin{lemma}[Semantic Upgrade]
-  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
-  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
-\end{lemma}
+substitution.
+
+\todo{Move weakening and upgrade lemmas to appendix}
+
+% In particular, semantic weakening can be stated in the expected
+% manner: if we have an expression or block which is well typed in both \(\Gamma\)
+% and \(\Delta\), where \(\cwk{\Gamma}{\Delta}\), then evaluating it in \(\Gamma\)
+% is the same as first strengthening \(\Gamma\) to \(\Delta\), dropping unused
+% affine variables, and then evaluating it in \(\Delta\).
+% \begin{lemma}[Semantic Weakening] 
+%   If \(\cwk{\Gamma}{\Delta}\), then 
+%   \begin{itemize}
+%     \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
+%     \[
+%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+%       ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
+%       = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+%     \]
+%     \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
+%     \[
+%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+%       ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
+%       = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
+%     \]
+%   \end{itemize}
+% 
+%   Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
+%   \[
+%     \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+%     ; \dnt{\lwk{\ms{L}}{\ms{K}}}
+%     = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
+%   \]
+% \end{lemma}
+% 
+% \TODO{some text about upgrade}
+% 
+% \begin{lemma}[Semantic Upgrade]
+%   For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
+%   \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
+% \end{lemma}
 
@@
 
-\subsection{Optimizations}
-
-\TODO{this:}
-
-\TODO{substitution \(\implies\): loop hoisting, global value numbering, some dead code elimination}
-
-\TODO{rewriting \(\implies\) strength reduction, E-graph optimization, see \cite{cranelift}}
-
-\TODO{inlining}
-
-\TODO{loop fusion}
-
-\TODO{unreachable code elimination}
-
-\TODO{more dead code elimination}
+\subsection{Optimisations}
+
+Because we have a denotational semantics, it is easy to validate
+equations between terms.
+
+Even the basic proof of the soundness of substitution already lets us
+prove the soundness of many fundamental compiler optimisations, such
+as global value numbering (let-introduction for pure expressions),
+some forms of dead code elimination (i.e., eliminating unused affine
+bindings), and strength reduction with loop hoisting (let-introduction
+for relevant expressions).
+
+Substitution only applies to central terms, because those are the only
+terms for which substitution is sound. Furthermore, there are many
+other optimisations that follow from the beta-eta theory of the
+language, as well as specific properties of model-dependent operations
+(such as memory operations). In the appendix, we prove a general
+rewriting theorem which lifts equations between terms into a full
+congruence. This lets us justify optimisations such as scalar
+replacement of aggregates, eliminating redundant writes, reordering
+independent memory operations, and memory-to-register transformations.
+
+Because we have a type-theoretic presentation of control-flow graphs,
+we also have a substitution principle and equational theory for
+them. This lets us show the soundness of inlining, loop unrolling,
+loop fusion, dead-code elimination, and other optimisations which
+change the structure of the control-flow graph.
+
+Note that we have not yet proved anything about the dataflow analyses
+used to find the rewrites -- we have only shown that the rewrites
+themselves are semantically justified.
 

### 2023-11-16 pldi24/paper.tex
foo

@@
 
-\newcommand{\todo}[1]{\stepcounter{todos} \textcolor{red}{TODO \arabic{todos}:} #1}
+\newcommand{\todo}[1]{\stepcounter{todos} \textcolor{red}{TODO \arabic{todos}: #1}}
 
@@
 
-Now that we've given a semantics to \isotopessa programs, we can begin to use it
-to develop an equational theory for performing optimizations. Since we're
-approaching things from a type-theoretic perspective, we begin by addressing the
-semantic versions of the standard syntactic metatheorems of weakening and
-substitution.
+Now that we have a semantics for \isotopessa programs, we can show
+that it matches the syntax by showing that the syntactic and semantic
+notions of substitution coincide.
 
 \todo{Move weakening and upgrade lemmas to appendix}
-
+% Since we're
+% approaching things from a type-theoretic perspective, we begin by addressing the
+% semantic versions of the standard syntactic metatheorems of weakening and
+% substitution. 
+% 
 % In particular, semantic weakening can be stated in the expected
@@
 
-\TODO{some text about substitution semantics}
-
-\TODO{some text about semantic substitution}
+\TODO{some text about substitution semantics -- substitution = composition of env semantics with therm semantics}
+\TODO{some text about semantic substitution -- it's a function that does the obvious thing}
 

### 2023-11-16 pldi24/paper.tex
Tap

@@
 \end{definition}
-
-\TODO{confusion with notion of pure in truly "pure" above}
-
-\TODO{text on Freyd categories}
-
+Unfortunately, the notion of "purity" given by an effectful category only allows
+us to perform relatively weak optimizations in the presence of effects. To
+remedy this, we want to be able to make a finer distinction between morphisms
+which are merely \textit{central} and those which are also \textit{relevant}
+(i.e., copying the input and applying them twice is the same as applying them
+once and copying the output) and/or \textit{affine} (i.e., applying them and
+discarding the result is the same as never doing them at all). Doing so also
+allows us to add the capability of \textit{distinguishing} between objects which
+are affine, relevant, neither or both, enabling us to reason about substructural type systems categorically.
+
+To do this, we begin by introducing the notion of a \textit{symmetric premonoidal subcategory}, as follows:
 \begin{definition}[(Symmetric) Premonoidal Subcategory]
@@
 
-% Should we go down this road?
-% \begin{definition}[Strictly Effectful Category] \
-%   \TODO{this}
-% \end{definition}
-
-\TODO{text}
-
+We can now define a \textit{substructural premonoidal category}, which is simply
+a premonoidal category equipped with a distinguished set of relevant and affine
+morphisms
 \begin{definition}[Substructural (Pre)monoidal Category]
@@
 S)\). Once again, it is clear that all objects have split and drop morphisms
-inherited from \(\ms{Set}\), and therefore are affine and relevant
-
+inherited from \(\ms{Set}\), and therefore are affine and relevant.
+
+One issue with the above definition is that, just like for regular premonoidal
+categories, it is difficult to distinguish between those morphisms that are
+"truly" central versus those which are just central "by accident", as we only
+want to require any notion of functor we develop subsequently to preserve the
+former. We can solve this the same way, by simply extending the notion of
+effectful category appropriately as follows:
 \begin{definition}[Susbtructural Effectful Category]
@@
 \end{definition}
-
 We can obtain an example of a substructural effectful category quite easily by,
@@
 \mc{P}(A \times S)\). Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
-\(
-  \upg{R}{} = \lambda a, s. R(a) \times \{s\}
-\), we can verify that this satisfies the axioms of a substructural effectful category.
-Note that here the \textit{central} morphisms are precisely those which do not
-read from or write to the state, and therefore which commute with all other
-morphisms.
+\(\upg{R}{} = \lambda a, s. R(a) \times \{s\}\), we can verify that this
+satisfies the axioms of a substructural effectful category. Note that here the
+\textit{central} morphisms are precisely those which do not read from or write
+to the state, and therefore which commute with all other morphisms.
 
@@
     \item Relevant: \(\upg{\delta}{} \gg \upg{\ms{split}}{} = \upg{(\delta;\ms{split})}{} = \upg{(\ms{split};\delta \otimes \delta)}{} = \upg{\ms{split}}{} \gg \upg{\delta}{} \otimes \upg{\delta}{}\)
-    \item Left inverse: \(\upg{\delta}{} \gg \upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \upg{\delta;[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \ms{id}\)
+    \item Left inverse: \(\upg{\delta}{} \gg \upg{[\iota_0 \otimes C, \iota_1
+    \otimes C]}{} = \upg{\delta;[\iota_0 \otimes C, \iota_1 \otimes C]}{} =
+    \ms{id}\)
     \item Right inverse: \(\upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} \gg \upg{\delta}{} = \upg{[\iota_0 \otimes C, \iota_1 \otimes C];\delta}{} = \ms{id}\)
-    \item Symmetric: \(\upg{\sigma^+}{} \otimes C \gg \upg{\delta}{} = \upg{\sigma^+ \otimes C;\delta}{} = \upg{\delta;\sigma^+}{} = \upg{\delta}{} \gg \upg{\sigma^+}{}\)
+    \item Symmetric: \(\upg{\sigma^+}{} \otimes C \gg \upg{\delta}{} =
+    \upg{\sigma^+ \otimes C;\delta}{} = \upg{\delta;\sigma^+}{} = \upg{\delta}{}
+    \gg \upg{\sigma^+}{}\)
   \end{itemize}
@@
 \end{itemize}
-
-\TODO{We know \textit{what} these semantics are, unlike below, but maybe clarify
-\textit{why}}
 
@@
 
-Now that we have a semantics for \isotopessa programs, we can show
-that it matches the syntax by showing that the syntactic and semantic
-notions of substitution coincide.
-
-\todo{Move weakening and upgrade lemmas to appendix}
-% Since we're
-% approaching things from a type-theoretic perspective, we begin by addressing the
-% semantic versions of the standard syntactic metatheorems of weakening and
-% substitution. 
-% 
-% In particular, semantic weakening can be stated in the expected
-% manner: if we have an expression or block which is well typed in both \(\Gamma\)
-% and \(\Delta\), where \(\cwk{\Gamma}{\Delta}\), then evaluating it in \(\Gamma\)
-% is the same as first strengthening \(\Gamma\) to \(\Delta\), dropping unused
-% affine variables, and then evaluating it in \(\Delta\).
-% \begin{lemma}[Semantic Weakening] 
-%   If \(\cwk{\Gamma}{\Delta}\), then 
-%   \begin{itemize}
-%     \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
-%     \[
-%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-%       ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
-%       = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-%     \]
-%     \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
-%     \[
-%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-%       ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
-%       = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
-%     \]
-%   \end{itemize}
-% 
-%   Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
-%   \[
-%     \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-%     ; \dnt{\lwk{\ms{L}}{\ms{K}}}
-%     = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
-%   \]
-% \end{lemma}
-% 
-% \TODO{some text about upgrade}
-% 
-% \begin{lemma}[Semantic Upgrade]
-%   For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
-%   \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
-% \end{lemma}
+Now that we have a semantics for \isotopessa programs, we can show that it
+matches the syntax by showing that the syntactic and semantic notions of
+substitution coincide. We begin by giving a semantics for the judgement
+\(\issubst{\gamma}{\Theta}{\Gamma}\), which we will interpret as a pure morphism
+taking \(\dnt{\Theta}\) to \(\dnt{\Gamma}\), defined via the rules in Figure
+\ref{fig:ssa-subst-semantics}, which can simply be viewed as computing all
+\(\gamma(x)\) for \(\thyp{x}{A}{q}\) in \(\Gamma\). In particular, we interpret
+\rle{subst-nil} as the identity function, and \rle{subst-cons} by splitting the
+context via \(\dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}\), and then
+passing \(\dnt{\Theta_\Gamma}\) to the remainder of the substitution
+\(\dnt{\issubst{\gamma}{\Gamma}{\Theta_\Gamma}}\) and \(\dnt{\Theta_x}\) through
+\(\dnt{\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}}\) to yield \(\dnt{A}\). We give a
+similar semantics for label-substitutions: \rle{lb-subst-nil} is interpreted as
+the zero morphism \(0_{\dnt{\ms{K}}}: \mc{C}_0^\varnothing(\mb{0},
+\dnt{\ms{K}})\), while \rle{lb-subst-cons} takes in the coproduct \(\dnt{\ms{L}}
++ \dnt{\Gamma} \otimes \dnt{A}\), sending the former to
+\(\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{A}}}\) and the latter to the denotation of
+\(\mc{L}(\lbl{\ell}, x)\).
 
@@
     \\
+    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id}
+    \qquad
     \dnt{\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
-      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{1}{a}{A}{q}} \\
-    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id}
+      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{1}{a}{A}{q}}
     \\
     \boxed{\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}: \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}})} 
+    \\
+    \dnt{\lbsubst{\mc{L}}{\cdot}{\ms{K}}} = 0_{\ms{K}}
+    \qquad
+    \dnt{\lbsubst{\mc{L}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
+    = [\dnt{\lbsubst{\ms{L}}{\ms{L}}{\ms{K}}}, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}}]
   \end{gather*}
@@
 
-\TODO{some text about substitution semantics -- substitution = composition of env semantics with therm semantics}
-\TODO{some text about semantic substitution -- it's a function that does the obvious thing}
-
+Given this, we can now state the theorem of semantic substitution, which
+essentially states that the semantics of a substituted expression or block is
+the same as the semantics of that expression or block preceded by let-bindings
+corresponding to the variables in the substitution, which is itself the same as
+the semantics of the original expression or block precomposed with the semantics
+of the substitution as given in Figure \ref{fig:ssa-subst-semantics}. Similarly,
+the semantics of a \textit{label-substituted} block is the same as the semantics
+of the original block \textit{postcomposed} with the semantics of that
+label-substitution. Stated formally, this becomes:
 \begin{theorem}[Semantic Substitution] 
@@
       ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+      = \dnt{\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}}
       = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
@@
       ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+      = \dnt{\haslb{\Theta}{\exprletsubst{\substctx{\gamma}{\Gamma}}}{\ms{K}}} 
+      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
       = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 

### 2023-11-16 pldi24/paper.tex
Tap

@@
 
-\TODO{terms vs expression/blocks/label-sets}
+\TODO{terms vs expression/blocks/control-flow-graphs}
 
@@
 manner being in "SSA form" simply means that every variable name is defined
-exactly once.
-For example, the program in Figure \ref{fig:3-addr} has 3 basic blocks, one of which is a distinguished \textit{entry block}, but is \textit{not} in SSA form, since the variables \(m, n\) and \(i\) are each modified through subsequent assignments (\(t\) is recomputed on every iteration of the loop, but since it only has one definition, it \textit{is} in SSA form).
-To convert this program to SSA form, we would traditionally use \(\phi\)-functions, or, equivalently, annotate each basic block with arguments, as in Figure \ref{fig:bb-arg}.
+exactly once. For example, the program in Figure \ref{fig:3-addr} has 3 basic
+blocks, one of which is a distinguished \textit{entry block}, but is
+\textit{not} in SSA form, since the variables \(m, n\) and \(i\) are each
+modified through subsequent assignments (\(t\) is recomputed on every iteration
+of the loop, but since it only has one definition, it \textit{is} in SSA form).
+To convert this program to SSA form, we would traditionally use
+\(\phi\)-functions, or, equivalently, annotate each basic block with arguments,
+as in Figure \ref{fig:bb-arg}.
 
@@
   judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
-  is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or \(p = 0\)
-  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). The rules
-  for this judgement are given in Figure \ref{fig:ssa-term-typing}.
-
-  So the centrality $p$ indicates whether it is safe to substitute this expression
-  for a variable, and $q$, plus the linearity of $A$, control how many times that
-  variable can occur.
+  is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or
+  \(p = 0\) otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\).
+  So the centrality \(p\) indicates whether it is safe to substitute this
+  expression for a variable, and \(q\), plus the linearity of \(A\), control how
+  many times that variable can occur: if \(\taff \subseteq q\) then the variable
+  is called \textit{affine} or \textit{discardable} and it is permissible to
+  leave it unused, whereas if \(\trel \subseteq q\) it is called
+  \textit{relevant} or \textit{duplicable} and it is permissible to use multiple
+  times. An expression which is central, relevant, and affine is called
+  \textit{pure}. The rules for this judgement are given in Figure
+  \ref{fig:ssa-term-typing}.
 
@@
   \(\haslb{\ms{L}}{t}{\Gamma}\), which says that the block \(t\) is well-formed
-  with respect to the context \(\Gamma\) and branches only to labels in
-  \(\ms{L}\). The rules for this judgement are given in Figure
-  \ref{fig:ssa-block-typing}.
-
-\item \textit{Label-sets} or \textit{control-flow graphs} \(L\), which are
-  typed via the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\)
-  defines a control-flow graph with inputs \(\ms{L}\) and outputs \(\ms{K}\)
-  (some of which may also be in \(\ms{L}\), indicating recursive control-flow).
-  The rules for this judgement are given in Figure \ref{fig:ssa-block-typing}.
+  with respect to the context \(\Gamma\) and only branches to
+  ("\textit{targets}") labels in \(\ms{L}\). The rules for this judgement are
+  given in Figure \ref{fig:ssa-block-typing}.
+
+\item \textit{Control-flow graphs} or \textit{CFG}s \(L\), which are typed via
+  the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\) defines a
+  control-flow graph with input labels \(\ms{L}\) and output labels \(\ms{K}\);
+  recursive control flow is allowed. The rules for this judgement are given in
+  Figure \ref{fig:ssa-block-typing}.
 \end{itemize}
@@
 
-\textit{Label-contexts} \(\ms{L}\) are defined to consist of a list of
+\textit{label-sets} \(\ms{L}\) are defined to consist of a list of
 \textit{labels} \(\lbl{\ell}\) annotated with a context \(\Gamma\) and an
 argument type \(A\). The judgement \(\lwk{\ms{L}}{\ms{K}}\) states that
-\(\ms{L}\) is a subcontext of \(\ms{K}\), i.e., that every program which
-"branches to \(\ms{L}\)" (i.e., finishes by either nontermination or branching
-to one of the labels \(\lbl{\ell}\) with the appropriate preserved variables
-\(\Gamma\) and argument \(A\)) can also be viewed as branching to \(\ms{K}\).
-The conditions for this are formalized in Figure \ref{fig:ssa-structural} as
+\(\ms{L}\) is a subcontext of \(\ms{K}\), i.e., that every program which targets
+\(\ms{L}\) (i.e., finishes by either nontermination or branching to one of the
+labels \(\llhyp{\ell}{\Gamma}{A} \in \ms{L}\) with live variables \(\Gamma\) and
+an argument of type \(A\)) can also be viewed as targeting \(\ms{K}\). The
+conditions for this are formalized in Figure \ref{fig:ssa-structural} as
 follows: the rule \rle{join-nil} provides a base case stating that the empty
@@
 the typing judgement for blocks, and is by analogy to our semantics for blocks
-and label-sets in Section \ref{sec:semantics}.
+and CFGs in Section \ref{sec:semantics}.
 
@@
         context \(\Gamma\) with centrality \(p\) and quantity \(q\) \\
-        \(\haslb{\Gamma}{t}{\ms{L}}\) & \(t\) is a block targeting labels
+        \(\haslb{\Gamma}{t}{\ms{L}}\) & \(t\) is a block targeting label-set
         \(\ms{L}\) in context \(\Gamma\) \\
-        \(\lhaslb{\ms{L}}{L}{\ms{K}}\) & The label-set \(L\) sends labels
-        \(\ms{L}\) to labels \(\ms{K}\) \\
+        \(\lhaslb{\ms{L}}{L}{\ms{K}}\) & The control-flow graph \(L\) sends
+        input labels \(\ms{L}\) to output labels \(\ms{K}\) \\
         \(\csplits{\Gamma}{\Delta}{\Xi}\) & The context \(\Gamma\) splits into
@@
         label-set \(\ms{K}\) \\
-        % \(\tlin{A}{q}\) &
-        % The type \(A\) can be used with linearity \(q\) \\
-        % \(\ltlin{A}{r}{q}\) &
-        % The type \(A\) can be used with linearity \(q \subseteq r\)
-        % (i.e. \(\tlin{A}{q} \land q \subseteq r\)) \\
-        % \(\tlin{\Gamma}{q}\) &
-        % The context \(\Gamma\) has linearity \(q\) \\
-        \(\cwk{\Gamma}{\Delta}\) &
-        \(\Gamma\) is a weakening of \(\Delta\) 
-        (i.e. \(\csplits{\Gamma}{\Delta}{\cdot}\))
+        % \(\tlin{A}{q}\) & The type \(A\) can be used with linearity \(q\) \\
+        % \(\ltlin{A}{r}{q}\) & The type \(A\) can be used with linearity \(q
+        % \subseteq r\) (i.e. \(\tlin{A}{q} \land q \subseteq r\)) \\
+        % \(\tlin{\Gamma}{q}\) & The context \(\Gamma\) has linearity \(q\) \\
+        \(\cwk{\Gamma}{\Delta}\) & \(\Gamma\) is a weakening of \(\Delta\) (i.e.
+        \(\csplits{\Gamma}{\Delta}{\cdot}\))
     \end{tabular}
@@
 mapping \(\ms{L}\) to \(\ms{K}\) is defined using the typing rules \rle{nil-br},
-which says that an empty list of blocks leaves the target label-set unchanged,
-and \rle{cons-br}, which says that if \(L\) maps \(\ms{L}\) to either \(\ms{K}\)
-or the label \(\lbl{\ell}[\Gamma](A)\), and \(t\) branches to \(\ms{L}\) in the
-context \(\Gamma, \thyp{x}{A}{}\), then \(L\) maps \(\ms{L}\) to \(\ms{K}\),
-having (potentially recursively) defined \(\lbl{\ell}\).
+which says that an empty control-flow graph simply forwards the input labels
+directly to the output labels, and \rle{cons-br}, which says that if \(L\) maps
+\(\ms{L}\) to either \(\ms{K}\) or the label \(\lbl{\ell}[\Gamma](A)\), and
+\(t\) branches to \(\ms{L}\) in the context \(\Gamma, \thyp{x}{A}{}\), then
+\(L\) maps \(\ms{L}\) to \(\ms{K}\), having (potentially recursively) defined
+\(\lbl{\ell}\).
 
@@
 standard capture-avoiding substitution, which we denote \([\gamma]e\) for
-expressions, \([\gamma]t\) for blocks, and \([\gamma]L\) for label-sets, as is
+expressions, \([\gamma]t\) for blocks, and \([\gamma]L\) for CFGs, as is
 standard. The second is to use the substitution to generate a set of
@@
 \ref{fig:ssa-subst-ops}. Note that this operation cannot be easily defined for
-label-sets. In Section \ref{sec:semantics}, we will show that this is
-semantically equivalent to regular capture-avoiding substitution, due to the
-fact that all terms in a substitution are central \textit{and} that only affine
-computations can be used in an expression substituting for an affine variable,
-and only relevant computationbs can be used in an expression substituting for a
-relevant variable.
+CFGs. In Section \ref{sec:semantics}, we will show that this is semantically
+equivalent to regular capture-avoiding substitution, due to the fact that all
+terms in a substitution are central \textit{and} that only affine computations
+can be used in an expression substituting for an affine variable, and only
+relevant computationbs can be used in an expression substituting for a relevant
+variable.
 
@@
 \(\ms{K}\); that is, for each \(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have
-that \(\mc{L}(\lbl{\ell}, x)\) is a block that branches to \(\ms{K}\). This is
+that \(\mc{L}(\lbl{\ell}, x)\) is a block that targets \(\ms{K}\). This is
 formally defined via the rules \rle{lb-subst-nil}, which states that this is
@@
 
-% \begin{lemma}[Substitution Splitting] 
-%   If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
-%   then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
-%   \begin{itemize}
-%     \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
-%     \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
-%     \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
-%   \end{itemize}
-% \end{lemma}
-
-% \TODO{some text about substitution weakening}
-
 We can also, using the rules in Figure \ref{fig:ssa-subst-ops}, define
-capture-avoiding label-substitution for blocks \([\mc{L}]t\) and label-sets
+capture-avoiding label-substitution for blocks \([\mc{L}]t\) and CFGs
 \([\mc{L}]L\), allowing us to state the following lemma:
@@
 \end{lemma}
-
-\TODO{small rewriting section here?}
 
@@
     \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
-    \item Satisfying the \textbf{pentagon law}: \todo{this}
-    \item Satisfying the \textbf{triangle law}: \todo{this}
+    \item Satisfying the \textbf{pentagon} and \textbf{triangle laws}; see e.g.
+    \citet{maclane:71}.
   \end{itemize}
@@
   central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
-  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} \todo{this}. We
-  say a premonoidal category is \textbf{monoidal} if every morphism is central.
+  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} (see
+  \cite{maclane:71}). We say a premonoidal category is \textbf{monoidal} if
+  every morphism is central.
 \end{definition}
@@
 \end{proposition}
-\begin{proof}
-  Let \(\delta_{A, B, C}: \mc{C}((A + B) \otimes C, A \otimes C + B \otimes C)\) denote \(\mc{C}\)'s distributor.
+% \begin{proof}
+%   Let \(\delta_{A, B, C}: \mc{C}((A + B) \otimes C, A \otimes C + B \otimes C)\) denote \(\mc{C}\)'s distributor.
 We will write composition in \(\mc{C}\) as "\(;\)" and composition in the Kleisli category \(\mb{T}\) as "\(\gg\)".
-  We will show that
-  \(\upg{\delta_{A, B, C}}{}: \mc{C}_{\mb{T}}((A + B) \otimes C, A \otimes C + B \otimes C)\) satisfies the desired properties of a distributor. In particular, it is:
-  \begin{itemize}
-    \item Central: this follows immediately from being the image of a pure morphism
-    \item Affine: \(\upg{\delta}{} \gg \upg{\ms{drop}}{} = \upg{(\delta;\ms{drop})}{} = \upg{\ms{drop}}{}\)
-    \item Relevant: \(\upg{\delta}{} \gg \upg{\ms{split}}{} = \upg{(\delta;\ms{split})}{} = \upg{(\ms{split};\delta \otimes \delta)}{} = \upg{\ms{split}}{} \gg \upg{\delta}{} \otimes \upg{\delta}{}\)
-    \item Left inverse: \(\upg{\delta}{} \gg \upg{[\iota_0 \otimes C, \iota_1
-    \otimes C]}{} = \upg{\delta;[\iota_0 \otimes C, \iota_1 \otimes C]}{} =
-    \ms{id}\)
-    \item Right inverse: \(\upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} \gg \upg{\delta}{} = \upg{[\iota_0 \otimes C, \iota_1 \otimes C];\delta}{} = \ms{id}\)
-    \item Symmetric: \(\upg{\sigma^+}{} \otimes C \gg \upg{\delta}{} =
-    \upg{\sigma^+ \otimes C;\delta}{} = \upg{\delta;\sigma^+}{} = \upg{\delta}{}
-    \gg \upg{\sigma^+}{}\)
-  \end{itemize}
-\end{proof}
-
-\TODO{text}
+%   We will show that
+%   \(\upg{\delta_{A, B, C}}{}: \mc{C}_{\mb{T}}((A + B) \otimes C, A \otimes C + B \otimes C)\) satisfies the desired properties of a distributor. In particular, it is:
+%   \begin{itemize}
+%     \item Central: this follows immediately from being the image of a pure morphism
+%     \item Affine: \(\upg{\delta}{} \gg \upg{\ms{drop}}{} = \upg{(\delta;\ms{drop})}{} = \upg{\ms{drop}}{}\)
+%     \item Relevant: \(\upg{\delta}{} \gg \upg{\ms{split}}{} = \upg{(\delta;\ms{split})}{} = \upg{(\ms{split};\delta \otimes \delta)}{} = \upg{\ms{split}}{} \gg \upg{\delta}{} \otimes \upg{\delta}{}\)
+%     \item Left inverse: \(\upg{\delta}{} \gg \upg{[\iota_0 \otimes C, \iota_1
+%     \otimes C]}{} = \upg{\delta;[\iota_0 \otimes C, \iota_1 \otimes C]}{} =
+%     \ms{id}\)
+%     \item Right inverse: \(\upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} \gg \upg{\delta}{} = \upg{[\iota_0 \otimes C, \iota_1 \otimes C];\delta}{} = \ms{id}\)
+%     \item Symmetric: \(\upg{\sigma^+}{} \otimes C \gg \upg{\delta}{} =
+%     \upg{\sigma^+ \otimes C;\delta}{} = \upg{\delta;\sigma^+}{} = \upg{\delta}{}
+%     \gg \upg{\sigma^+}{}\)
+%   \end{itemize}
+% \end{proof}
 
@@
 
-The structure so far is enough to support non-looping control flow, but to allow
-general, potentially nonterminating control flow, we need a way to reason about
-iteration. To do so, we introduce the notion of an \textit{Elgot category} as
-follows:
+The structures we've developed so far are enough to support non-looping control
+flow, but to allow general, potentially nonterminating control flow, we need a
+way to reason about iteration. To do so, we introduce the notion of an
+\textit{Elgot category} as follows:
 \begin{definition}[Elgot category] 
@@
 associated split \(\csplits{\Gamma}{\Delta}{\cdot}\) postcomposed with a unitor.
-Finally, we interpret a label-context weakening \(\lwk{\ms{L}}{\ms{K}}\) as a
+Finally, we interpret a label-set weakening \(\lwk{\ms{L}}{\ms{K}}\) as a
 pure morphism \(\mc{C}^\tint_1(\dnt{\ms{L}}, \dnt{\ms{K}})\), where
@@
 \thyp{x}{A}{q}}{t}{\ms{L}}}\), which will give us a value of \(\ms{L}\) we can
-pass in as recursive input. Note that, even if \(t\) branches to an output label
-in \(\ms{K}\), we still return a \textit{recursive input}, not an output; in
-this case, the \textit{next} iteration will return it as an output. We can now
-quite easily state the semantics of \rle{where}: we simply precompose the
-semantics of the entry block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the
-\textit{iterated} semantics of the associated control-flow graph,
+pass in as recursive input. Note that, even if \(t\) targets an output label in
+\(\ms{K}\), we still return a \textit{recursive input}, not an output; in this
+case, the \textit{next} iteration will return it as an output. We can now quite
+easily state the semantics of \rle{where}: we simply precompose the semantics of
+the entry block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the \textit{iterated}
+semantics of the associated control-flow graph,
 \(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger\).
@@
 
-\TODO{text}
-
+We now give a few basic definitions relevant to monads over \(\mb{Set}\):
 \begin{definition}[Relevant, Affine Monad]
@@
 \end{definition}
-
 \begin{definition}[Submonad] 
@@
   a substructural effectful category with all objects both affine and relevant.
+\end{enumerate}
+We can use the trace monad described above to derive some interesting examples
+of submonads:
+\begin{enumerate}
   \item If \(C \subseteq Z(M)\) is submonoid of \(M\), then \(A \mapsto
@@
   \item \todo{block}
-  \item \todo{label-set}
-\end{itemize}
-\TODO{hole typing rules for expressions, blocks, label-sets; everything else is
-the same}.
+  \item \todo{control flow graph}
+\end{itemize}
+\TODO{hole typing rules for expressions, blocks, control-flow graphs; everything
+else is the same}.
 

### 2023-11-16 pldi24/paper.tex
Tap

@@
 terminology, and also call such effects \emph{relevant}. With these distinction
-in place, we can \emph{define} a pure expression as one which is all three of
+in place, we can \emph{define} a pure term as one which is all three of
 central, copyable, and discardable.
@@
 \section{SSA Syntax}
-
-\TODO{terms vs expression/blocks/control-flow-graphs}
-
-\TODO{standardize "branching to" wordmark? Is wordmark the word here?}
 
@@
 
-Our grammar proper is split into three main syntactic categories:
+Our grammar proper splits program fragments into three main syntactic categories:
 \begin{itemize}
   \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
-  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
+  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the term \(a\)
   is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or
@@
   So the centrality \(p\) indicates whether it is safe to substitute this
-  expression for a variable, and \(q\), plus the linearity of \(A\), control how
+  term for a variable, and \(q\), plus the linearity of \(A\), control how
   many times that variable can occur: if \(\taff \subseteq q\) then the variable
@@
   \textit{relevant} or \textit{duplicable} and it is permissible to use multiple
-  times. An expression which is central, relevant, and affine is called
+  times. An term which is central, relevant, and affine is called
   \textit{pure}. The rules for this judgement are given in Figure
@@
 for variables, with the requirement that, other than \(x\), all variables in
-\(\Gamma\) are affine. Note that the quantity \(q\) of the \textit{expression}
+\(\Gamma\) are affine. Note that the quantity \(q\) of the \textit{term}
 \(x\) is must satisfy \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this
@@
 instruction \(f\) which takes in an argument \(A\) and has a return type of
-\(B\) and a \textit{central} expression \(\hasty{\Gamma}{1}{a}{A}{q}\), we
+\(B\) and a \textit{central} term \(\hasty{\Gamma}{1}{a}{A}{q}\), we
 conclude that \(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that
@@
 as branching into \(\ms{L}\) in constext \(\Gamma\). \rle{let-blk} and
-\rle{let2-blk} are quite similar to their expression analogues, except that the
-bound expression \(e\) is allowed to be impure (otherwise, programs could only
+\rle{let2-blk} are quite similar to their term analogues, except that the
+bound term \(e\) is allowed to be impure (otherwise, programs could only
 make use of pure functions!), and the remainder of the context plus the bound
@@
 terms in a substitution are central \textit{and} that only affine computations
-can be used in an expression substituting for an affine variable, and only
-relevant computationbs can be used in an expression substituting for a relevant
-variable.
+can be used in a term substituting for an affine variable, and only relevant
+computationbs can be used in a term substituting for a relevant variable.
 
@@
 \end{lemma}
-Our treatment of \rle{var} is quite standard: the expression \(x\) types in
-\(\Gamma\) if all other variables in \(\Gamma\) and can be discarded, in which
-case the semantics is simply doing so, i.e.
-\(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\), lifting to \(\mc{C}_0\) as necessary.
-Similarly, to type a function application \rle{var}, we first take
-\(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is guarnateed to be central, lift it
-to the appropriate category \(\mc{C}_p\), and then simply postcompose it with
-the denotation of \(f\). \rle{pair} is handled by composing the context split
-\(\csplits{\Gamma}{\Delta}{\Xi}\) taking the tensor product of the components
-\(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\), \(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is
-well-defined since both components are central and \(\mc{C}_1\) is monoidal. We
-then lift the result to the appropriate \(\mc{C}_p\). The semantics of
-\rle{unit} is simply dropping the entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\),
-while \rle{true}, and \rle{false} are each handled by first dropping the context
-and then postcomposing with the appropriate constant; in all cases, this is
-followed by lifting to \(\mc{C}_p\). Finally, \rle{let} and \rle{let2} are
-handled by first splitting the context, then passing the right hand component to
-the semantics of the bound expression \(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or
-\(\dnt{\hasty{\Xi}{1}{a}{A \otimes B}{q}}\), and then, lifting to \(\mc{C}_p\)
-as necessary, after reassociating with \(\alpha\), passing everything to the
-inner expression \(\dnt{\hasty{\Delta, \thyp{x}{A}{q}}{p}{e}{B}{q}}\) or
-\(\dnt{\hasty{\Delta, \thyp{x}{A}{q}, \thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that
-this inner expression is not necessarily central, unlike the bound expression!
+Our treatment of \rle{var} is quite standard: the term \(x\) types in \(\Gamma\)
+if all other variables in \(\Gamma\) and can be discarded, in which case the
+semantics is simply doing so, i.e. \(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\),
+lifting to \(\mc{C}_0\) as necessary. Similarly, to type a function application
+\rle{var}, we first take \(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is
+guarnateed to be central, lift it to the appropriate category \(\mc{C}_p\), and
+then simply postcompose it with the denotation of \(f\). \rle{pair} is handled
+by composing the context split \(\csplits{\Gamma}{\Delta}{\Xi}\) taking the
+tensor product of the components \(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\),
+\(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is well-defined since both components
+are central and \(\mc{C}_1\) is monoidal. We then lift the result to the
+appropriate \(\mc{C}_p\). The semantics of \rle{unit} is simply dropping the
+entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\), while \rle{true}, and \rle{false}
+are each handled by first dropping the context and then postcomposing with the
+appropriate constant; in all cases, this is followed by lifting to \(\mc{C}_p\).
+Finally, \rle{let} and \rle{let2} are handled by first splitting the context,
+then passing the right hand component to the semantics of the bound term
+\(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or \(\dnt{\hasty{\Xi}{1}{a}{A \otimes
+B}{q}}\), and then, lifting to \(\mc{C}_p\) as necessary, after reassociating
+with \(\alpha\), passing everything to the inner term \(\dnt{\hasty{\Delta,
+\thyp{x}{A}{q}}{p}{e}{B}{q}}\) or \(\dnt{\hasty{\Delta, \thyp{x}{A}{q},
+\thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that this inner term is not necessarily
+central, unlike the bound term!
 
@@
 semantics of \rle{let-blk} and \rle{let2-blk} are analogous to those of
-\rle{let} and \rle{let2} in the expression case, except that now the bound
-expression \(a\) is allowed to be impure (and therefore in particular does not
-need to be upgraded).
+\rle{let} and \rle{let2} in the term case, except that now the bound term \(a\)
+is allowed to be impure (and therefore in particular does not need to be
+upgraded).
 
@@
 Given this, we can now state the theorem of semantic substitution, which
-essentially states that the semantics of a substituted expression or block is
-the same as the semantics of that expression or block preceded by let-bindings
-corresponding to the variables in the substitution, which is itself the same as
-the semantics of the original expression or block precomposed with the semantics
-of the substitution as given in Figure \ref{fig:ssa-subst-semantics}. Similarly,
-the semantics of a \textit{label-substituted} block is the same as the semantics
-of the original block \textit{postcomposed} with the semantics of that
-label-substitution. Stated formally, this becomes:
+essentially states that the semantics of a substituted term or block is the same
+as the semantics of that term or block preceded by let-bindings corresponding to
+the variables in the substitution, which is itself the same as the semantics of
+the original term or block precomposed with the semantics of the substitution as
+given in Figure \ref{fig:ssa-subst-semantics}. Similarly, the semantics of a
+\textit{label-substituted} block is the same as the semantics of the original
+block \textit{postcomposed} with the semantics of that label-substitution.
+Stated formally, this becomes:
 \begin{theorem}[Semantic Substitution] 
@@
 
-Because we have a denotational semantics, it is easy to validate
-equations between terms.
-
-Even the basic proof of the soundness of substitution already lets us
-prove the soundness of many fundamental compiler optimisations, such
-as global value numbering (let-introduction for pure expressions),
-some forms of dead code elimination (i.e., eliminating unused affine
-bindings), and strength reduction with loop hoisting (let-introduction
-for relevant expressions).
-
-Substitution only applies to central terms, because those are the only
-terms for which substitution is sound. Furthermore, there are many
-other optimisations that follow from the beta-eta theory of the
-language, as well as specific properties of model-dependent operations
-(such as memory operations). In the appendix, we prove a general
-rewriting theorem which lifts equations between terms into a full
-congruence. This lets us justify optimisations such as scalar
-replacement of aggregates, eliminating redundant writes, reordering
-independent memory operations, and memory-to-register transformations.
-
-Because we have a type-theoretic presentation of control-flow graphs,
-we also have a substitution principle and equational theory for
-them. This lets us show the soundness of inlining, loop unrolling,
-loop fusion, dead-code elimination, and other optimisations which
-change the structure of the control-flow graph.
-
-Note that we have not yet proved anything about the dataflow analyses
-used to find the rewrites -- we have only shown that the rewrites
-themselves are semantically justified.
+Because we have a denotational semantics, it is easy to validate equations
+between terms, blocks, and CFGs.
+
+Even the basic proof of the soundness of substitution already lets us prove the
+soundness of many fundamental compiler optimisations, such as global value
+numbering (let-introduction for pure expressions), some forms of dead code
+elimination (i.e., eliminating unused affine bindings), and strength reduction
+with loop hoisting (let-introduction for relevant expressions).
+
+Substitution only applies to central terms, because those are the only terms for
+which substitution is sound. Furthermore, there are many other optimisations
+that follow from the beta-eta theory of the language, as well as specific
+properties of model-dependent operations (such as memory operations). In the
+appendix, we prove a general rewriting theorem which lifts equations between
+terms into a full congruence. This lets us justify optimisations such as scalar
+replacement of aggregates, eliminating redundant writes, reordering independent
+memory operations, and memory-to-register transformations.
+
+Because we have a type-theoretic presentation of control-flow graphs, we also
+have a substitution principle and equational theory for them. This lets us show
+the soundness of inlining, loop unrolling, loop fusion, dead-code elimination,
+and other optimisations which change the structure of the control-flow graph.
+
+Note that we have not yet proved anything about the dataflow analyses used to
+find the rewrites -- we have only shown that the rewrites themselves are
+semantically justified.
 
@@
 
-For space reasons we, restrict our discussion to the most closely
-related work on SSA.
+For space reasons we, restrict our discussion to the most closely related work
+on SSA.
 
 Originally, \citet{kelsey-ssa-cps} showed how to embed SSA into a CPS
-representation, and \citet{appel-ssa} showed informally how SSA can be
-seen as a collection of mutually-tail-recursive
-procedures. \citet{chakravarty-functional-ssa-2003} made Appel's
-observation formal, by showing how to translate SSA into A-normal
-form~\cite{anf}, and used it to prove the correctness of a constant
-propagation analysis. These translations were for an untyped source
-and target language, and
-\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20} give a typed
-translation from SSA into the lambda calculus, into a type-and-effect
-system. In contrast, we directly track the centrality and duplicability of
-terms, without tracking specific effects.
-
-All of these papers give SSA semantics by translation -- that is, they
-defined the semantics of an SSA program to be the semantics of the
-image of the translation, with the idea that the simpler semantics of
-the functional language can make proofs easier. The Grail language of
-\citet{beringer-imp-fun} makes this idea explicit by giving both a
-functional and imperative semantics to a low-level language, and
-showing the conditions under which they coincide. \citet{schneider-imp-fun} prove
-a similar result, and further mechanise the proof in Coq.
-
-There have also been a number of papers which have studied the
-semantics of SSA directly.
-
-\citet{ssa-types-matsuno-ohori-06} give a type theory for what appear
-to be ordinary three-address code programs. However, every well-typed
-program can be placed into SSA-form by inserting $\phi$-nodes in a
-fully type-directed way. This lets them model SSA without any $\phi$-nodes, lets
-them use the standard
+representation, and \citet{appel-ssa} showed informally how SSA can be seen as a
+collection of mutually-tail-recursive procedures.
+\citet{chakravarty-functional-ssa-2003} made Appel's observation formal, by
+showing how to translate SSA into A-normal form~\cite{anf}, and used it to prove
+the correctness of a constant propagation analysis. These translations were for
+an untyped source and target language, and
+\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20} give a typed translation
+from SSA into the lambda calculus, into a type-and-effect system. In contrast,
+we directly track the centrality and duplicability of terms, without tracking
+specific effects.
+
+All of these papers give SSA semantics by translation -- that is, they defined
+the semantics of an SSA program to be the semantics of the image of the
+translation, with the idea that the simpler semantics of the functional language
+can make proofs easier. The Grail language of \citet{beringer-imp-fun} makes
+this idea explicit by giving both a functional and imperative semantics to a
+low-level language, and showing the conditions under which they coincide.
+\citet{schneider-imp-fun} prove a similar result, and further mechanise the
+proof in Coq.
+
+There have also been a number of papers which have studied the semantics of SSA
+directly.
+
+\citet{ssa-types-matsuno-ohori-06} give a type theory for what appear to be
+ordinary three-address code programs. However, every well-typed program can be
+placed into SSA-form by inserting $\phi$-nodes in a fully type-directed way.
+This lets them model SSA without any $\phi$-nodes, lets them use the standard
 semantics for three-address code.
 %
-\citet{hua-explicit-ssa-2010} give another type system and
-direct operational semantics for SSA, and prove type safety
-for it.
-
-Many operational semantics for SSA have arisen from compiler
-verification efforts. \citet{barthe-compcert-ssa-2014} give an
-operational semantics as part of the CompCertSSA project, and give a
-semantics-preserving translation from three-address code into
-SSA. \citet{herklotz-gsa-2023} formalise "gated SSA" and give
-semantics-preserving translations between it and ordinary SSA. Going
-beyond CompCertSSA, \citet{vellum} have studied the semantics of the
-LLVM IR itself.
-
-There has been much less work on denotational semantics for SSA, or
-directly on its equational theory. \citet{pop-ssa-inout-2009} give an
-unusual denotational model of SSA in terms of the iteration structure
-of a program, which they use to better understand the loop-closing
-$\phi$-nodes found both in the gated SSA representation as well as
-practical compilers such as GCC.  \citet{garbuzov-structural-cfg-2018}
-exhibit a correspondence between an operational semantics for SSA, and
-an operational semantics for call-by-push-value~\cite{cbpv}. They use
-the normal-form bisimulations of \citet{lassen-bisim} to derive an
-equational theory for use in justifying optimisations. Their
-operational semantics is effect-free, which means that it cannot
-justify optimisations such as eliminating redundant
-writes.
-
-Our work is the first to validate the equational theory of SSA against
-a weak memory semantics. We focused on denotational models of weak
-memory, since they make it easier to establish equational
-properties. However, we need a number of features of the semantics
-which are sometimes difficult to establish in the weak memory
-world. First, sequencing must be associative (i.e.,
-$(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).  Second, the semantics must
-support loops, including the equation that a while-loop is equal to
-its unrolling. Finally, sums need to be distributive (e.g., in the
-boolean case, $(A \otimes \mb{2}) \simeq (A + A)$).
-
-As a result, we focused on the TSO semantics of \citet{sparky}, which
-validates the properties we need. For ARM-style concurrency, which
-admits load buffering, we only found models which are "near misses":
-they have a subset of the properties we need. The event
-structure model of \citet{paviotti-modular-relaxed-dep-20} is
-formulated in terms of step-indexing, so satisfies all of our needed
-properties except the loop unrolling property -- step-indexing means
-it is a refinement rather than an equation. The pomsets with
-preconditions model of \citet{jagadeesan-pwp-20} does not exactly
-satisfy the associativity of semicolon, and while the pomsets with
-predicate transformers model of \citet{leaky-semicolon} does satisfy
-associativity, it does not yet support loops.
-
-We do not think it would be prohibitively difficult to extend these
-models to support the required features, and we believe our results
-suggest that it would be have a high payoff: those features would
-suffice to build a model of our IR, giving us tight results
-validating the transformations compilers want against the low-level
-specification of the actual machine.
+\citet{hua-explicit-ssa-2010} give another type system and direct operational
+semantics for SSA, and prove type safety for it.
+
+Many operational semantics for SSA have arisen from compiler verification
+efforts. \citet{barthe-compcert-ssa-2014} give an operational semantics as part
+of the CompCertSSA project, and give a semantics-preserving translation from
+three-address code into SSA. \citet{herklotz-gsa-2023} formalise "gated SSA" and
+give semantics-preserving translations between it and ordinary SSA. Going beyond
+CompCertSSA, \citet{vellum} have studied the semantics of the LLVM IR itself.
+
+There has been much less work on denotational semantics for SSA, or directly on
+its equational theory. \citet{pop-ssa-inout-2009} give an unusual denotational
+model of SSA in terms of the iteration structure of a program, which they use to
+better understand the loop-closing $\phi$-nodes found both in the gated SSA
+representation as well as practical compilers such as GCC.
+\citet{garbuzov-structural-cfg-2018} exhibit a correspondence between an
+operational semantics for SSA, and an operational semantics for
+call-by-push-value~\cite{cbpv}. They use the normal-form bisimulations of
+\citet{lassen-bisim} to derive an equational theory for use in justifying
+optimisations. Their operational semantics is effect-free, which means that it
+cannot justify optimisations such as eliminating redundant writes.
+
+Our work is the first to validate the equational theory of SSA against a weak
+memory semantics. We focused on denotational models of weak memory, since they
+make it easier to establish equational properties. However, we need a number of
+features of the semantics which are sometimes difficult to establish in the weak
+memory world. First, sequencing must be associative (i.e., $(t_1; t_2); t_3
+\equiv t_1; (t_2; t_3)$).  Second, the semantics must support loops, including
+the equation that a while-loop is equal to its unrolling. Finally, sums need to
+be distributive (e.g., in the boolean case, $(A \otimes \mb{2}) \simeq (A +
+A)$).
+
+As a result, we focused on the TSO semantics of \citet{sparky}, which validates
+the properties we need. For ARM-style concurrency, which admits load buffering,
+we only found models which are "near misses": they have a subset of the
+properties we need. The event structure model of
+\citet{paviotti-modular-relaxed-dep-20} is formulated in terms of step-indexing,
+so satisfies all of our needed properties except the loop unrolling property --
+step-indexing means it is a refinement rather than an equation. The pomsets with
+preconditions model of \citet{jagadeesan-pwp-20} does not exactly satisfy the
+associativity of semicolon, and while the pomsets with predicate transformers
+model of \citet{leaky-semicolon} does satisfy associativity, it does not yet
+support loops.
+
+We do not think it would be prohibitively difficult to extend these models to
+support the required features, and we believe our results suggest that it would
+be have a high payoff: those features would suffice to build a model of our IR,
+giving us tight results validating the transformations compilers want against
+the low-level specification of the actual machine.
 
@@
 \begin{itemize}
-  \item \todo{expression}
+  \item \todo{term}
   \item \todo{block}

### 2023-11-16 pldi24/paper.tex
Tap

@@
 subcategory consisting only of the associators, unitors, and symmetry, and only
-the identity object considered affine and relevant. On the other extreme, every
+the identity object considered affine and relevant. Similarly, every
 \textit{Cartesian} category \(\mc{C}\) can be viewed as a substructural
@@
 the case \textit{if and only if} a category is Cartesian, hence the term
-"substructural Cartesian category."
-
-We can also deduce the following:
+"substructural Cartesian category." We also note the following:
 \begin{proposition}
-  If \(\mc{C}\) is a substructural monoidal category then the Kleisli category
-  \(\mc{C}_{\mb{T}}\) is a substructural premonoidal category with:
-  \begin{itemize}
-    \item \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\), \(\ms{Rel}(\mc{C}_{\mb{T}}) = \ms{Rel}(\mc{C})\)
-    \item \(\mc{C}_{\mb{T}}^q(A, B) = \upg{\mc{C}^q(A, B)}{}\), where
-    \(\upg{f}{} = f;\eta\)
-    \item Drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\)
-  \end{itemize} 
-  In particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a
-  substructural effectful category.
+  If \(\mc{C}\) is a substructural monoidal category then we can equip the
+  Kleisli category \(\mc{C}_{\mb{T}}\) with the structure of a substructural
+  effectful category by taking \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\),
+  \(\ms{Rel}(\mc{C}_{\mb{T}}) = \ms{Rel}(\mc{C})\), and \(\mc{C}_{\mb{T}}^q(A,
+  B) = \upg{\mc{C}^q(A, B)}{}\), where \(\upg{f}{} = f;\eta\), and defining
+  drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\). In
+  particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a substructural
+  effectful category.
 \end{proposition}
-
-\TODO{work into text?}
 
@@
   forms a stream action on the one element set \(\mathbf{1}\).
+  % \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action
+  % \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta =
+  % \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to
+  % \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i
+  % \cdot_i l_i)_i\).
+  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream
+  action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\),
+  the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell
+  \cdot s\) prepends the \(\ell\) to \(s\).
+  % \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a
+  % stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts
+  % on \(\nats \cup \{\infty\}\) by addition.
   \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and
@@
   compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
-  \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action
-  \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta =
-  \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to
-  \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i
-  \cdot_i l_i)_i\).
-  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream
-  action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\),
-  the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell
-  \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite)
-  sequence \(s\).
-  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a
-  stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts
-  on \(\nats \cup \{\infty\}\) by addition.
 \end{enumerate}
@@
   where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') =
-  \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an
-  Elgot monad, with iteration operator \(f^\dagger(a) = f^\infty(a) \cup
+  \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). The iteration operator \(f^\dagger(a) = f^\infty(a) \cup
   \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t
-  \in f_n(a)\}\), where
+  \in f_n(a)\}\) yields an Elgot structure, where
   \[
-    f_0 = f, \qquad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n]
-  \]
-  \[
+    f_0 = f, \quad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n], \quad
     f^\infty(a) = \{\Theta (\lambda n. m_n) \mid a_0 = a \land \forall n, \exists a_{n + 1}. \iota_0 (\iota_1\;a_{n + 1}, m_n) \in f(a_n)\}
   \]
-  \TODO{clean this up}
 \end{definition}
@@
   heap. We will consider how to relax this condition later.
-  \item Relevant and affine morphisms to be the lifts of relevant/affine morphisms in \(\ms{Print}\)
+  \item Relevant and affine morphisms to be the lifts of relevant/affine
+  morphisms in \(\ms{Print}\)
 \end{itemize}

### 2023-11-16 pldi24/paper.tex
Tap at rewriting appendix section

@@
         \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) &
-        \(\mc{L}\) is a label substitution taking \(\ms{L}\) to \(\ms{K}\) \\
-        % \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) &
-        % ... \\
-        % \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) &
-        % ... \\
-        % \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) &
-        % ... \\
-        % \(\isrw{\mc{H}}{I}{H}\) &
-        % ... \\
+        \(\mc{L}\) is a label substitution taking \(\ms{L}\) to \(\ms{K}\)
     \end{tabular}
@@
 Unfortunately, the notion of "purity" given by an effectful category only allows
-us to perform relatively weak optimizations in the presence of effects. To
+us to perform relatively weak optimistions in the presence of effects. To
 remedy this, we want to be able to make a finer distinction between morphisms
@@
 properties of model-dependent operations (such as memory operations). In the
-appendix, we prove a general rewriting theorem which lifts equations between
+appendix, we state a general rewriting theorem which lifts equations between
 terms into a full congruence. This lets us justify optimisations such as scalar
@@
 
-\section{Proofs}
-
-\subsection{Weakening and Substitution}
-
-\begin{theorem}[Weakening] \
-  \begin{itemize}
-    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
-    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
-    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then
-    \(\haslb{\Gamma}{t}{\ms{L}}\)
-    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then
-    \(\haslb{\Gamma}{t}{\ms{K}}\)
-  \end{itemize}
-\end{theorem}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{lemma}[Substitution Splitting] 
-  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
-  \begin{itemize}
-    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
-    \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
-    \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
-  \end{itemize}
-\end{lemma}
-
-\begin{proof} 
-  We proceed by induction on derivations \(\csplits{\Gamma}{\Delta}{\Xi}\). In
-  the \rle{split-nil} case, we have that \(\issubst{\gamma}{\cdot}{\cdot}\) by
-  \rle{subst-nil}. For all other cases, since \(\issubst{\gamma}{\Theta}{\Gamma,
-  \thyp{x}{A}{q}}\), we have by \rle{subst-cons} that
-  \(\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}\),
-  \(\issubst{\gamma}{\Theta_\Gamma}{\Gamma}\),
-  \(\hasty{\Theta_\Gamma}{1}{\gamma(x)}{A}{q}\). Since
-  \(\csplits{\Gamma}{\Delta}{\Xi}\), by induction, we have that
-  \(\csplits{\Theta_\Gamma}{\Theta_\Delta}{\Theta_\Xi}\) where \(\Theta_\Delta\)
-  and \(\Theta_\Xi\) have the desired properties above.
-  \begin{itemize}
-    \item \rle{split-left}: by left split association, we have that there exists
-    \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\csplits{\Theta}{\Theta_{\Delta,
-    \thyp{x}{A}{q}}}{\Theta_\Xi}\), \(\csplits{\Theta_{\Delta,
-    \thyp{x}{A}{q}}}{\Theta_\Delta}{\Theta_x}\). Using the fact that
-    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), we can conclude by
-    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Delta,
-    \thyp{x}{A}{q}}}{\Delta, \thyp{x}{A}{q}}\); the desired result hence
-    follows.
-    \item \rle{split-right}: by right split association, we have that there
-    exists \(\Theta_{\Xi, \thyp{x}{A}{q}}\),
-    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_{\Xi, \thyp{x}{A}{q}}}\),
-    \(\csplits{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Theta_\Xi}{\Theta_x}\). Using the
-    fact that \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by
-    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
-    \thyp{x}{A}{q}}\); the desired result hence follows.
-    \item \rle{split-dup}: since \(\tlin{A}{\trel}\), it follows by
-    \rle{subst-cons} that \(\tlin{\Theta_x}{\trel}\), and hence that
-    \(\csplits{\Theta_x}{\Theta_x}{\Theta_x}\). Therefore, by split association,
-    we have that there exist \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\Theta_{\Xi,
-    \thyp{x}{A}{q}}\) such that \(\csplits{\Theta}{\Theta_{\Delta,
-    \thyp{x}{A}{q}}}{\Theta_{\Xi, \thyp{x}{A}{q}}}\). Using the facts that
-    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\),
-    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by \rle{subst-cons}
-    that \(\issubst{\gamma}{\Theta_{\Delta, \thyp{x}{A}{q}}}{\Delta,
-    \thyp{x}{A}{q}}\), \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
-    \thyp{x}{A}{q}}\); the desired result hence follows.
-    \item \rle{split-drop}: since \(\tlin{A}{\taff}\), it follows by
-    \rle{subst-cons} that \(\tlin{\Theta_x}{\taff}\), and hence that
-    \(\cwk{\Theta_x}{\cdot}\). Therefore, by split association,
-    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\), giving the desired result.
-  \end{itemize}
-\end{proof}
-
-\begin{lemma}[Substitution] 
-  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  \begin{itemize}
-    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
-    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
-    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
-    \begin{itemize}
-      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
-      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
-      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
-    \end{itemize}
-  \end{itemize}
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{lemma}[Label Substitution] 
-  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
-  \begin{itemize}
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
-    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
-  \end{itemize}
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\subsection{Semantic Metatheory}
-
-\begin{lemma}[Semantic Upgrade]
-  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
-  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{lemma}[Semantic Weakening] 
-  If \(\cwk{\Gamma}{\Delta}\), then 
-  \begin{itemize}
-    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
-    \[
-      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
-      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-    \]
-    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
-    \[
-      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
-      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
-    \]
-  \end{itemize}
-
-  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
-  \[
-    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
-    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
-  \]
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{lemma}[Substitution Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{lemma}[Semantic Substitution Splitting] 
-  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
-  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
-  \[
-    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
-      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
-      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
-  \]
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{theorem}[Semantic Substitution] 
-  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  \begin{itemize}
-    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
-    \[
-      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
-      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
-    \]
-    \item For all 
-      \(\haslb{\Gamma}{t}{\ms{L}}\), 
-      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
-      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
-    \[
-      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
-      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
-      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
-    \]
-    \item For all 
-      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
-      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
-      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
-      \[
-        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
-        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
-        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
-        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-      \]
-  \end{itemize}
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
-  \begin{itemize} 
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
-    \[
-      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
-    \]
-    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
-    \[
-      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
-      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
-    \]
-    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
-  \end{itemize}
-\end{theorem}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\section{Loop Hoisting}
-
-\TODO{this?}
+% \section{Proofs}
+
+% \subsection{Weakening and Substitution}
+
+% \begin{theorem}[Weakening] \
+%   \begin{itemize}
+%     \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
+%     \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
+%     \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then
+%     \(\haslb{\Gamma}{t}{\ms{L}}\)
+%     \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then
+%     \(\haslb{\Gamma}{t}{\ms{K}}\)
+%   \end{itemize}
+% \end{theorem}
+
+% \begin{proof} \
+%   \TODO{this}
+% \end{proof}
+
+% \begin{lemma}[Substitution Splitting] 
+%   If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
+%   then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
+%   \begin{itemize}
+%     \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
+%     \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
+%     \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
+%   \end{itemize}
+% \end{lemma}
+
+% \begin{proof} 
+%   We proceed by induction on derivations \(\csplits{\Gamma}{\Delta}{\Xi}\). In
+%   the \rle{split-nil} case, we have that \(\issubst{\gamma}{\cdot}{\cdot}\) by
+%   \rle{subst-nil}. For all other cases, since \(\issubst{\gamma}{\Theta}{\Gamma,
+%   \thyp{x}{A}{q}}\), we have by \rle{subst-cons} that
+%   \(\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}\),
+%   \(\issubst{\gamma}{\Theta_\Gamma}{\Gamma}\),
+%   \(\hasty{\Theta_\Gamma}{1}{\gamma(x)}{A}{q}\). Since
+%   \(\csplits{\Gamma}{\Delta}{\Xi}\), by induction, we have that
+%   \(\csplits{\Theta_\Gamma}{\Theta_\Delta}{\Theta_\Xi}\) where \(\Theta_\Delta\)
+%   and \(\Theta_\Xi\) have the desired properties above.
+%   \begin{itemize}
+%     \item \rle{split-left}: by left split association, we have that there exists
+%     \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\csplits{\Theta}{\Theta_{\Delta,
+%     \thyp{x}{A}{q}}}{\Theta_\Xi}\), \(\csplits{\Theta_{\Delta,
+%     \thyp{x}{A}{q}}}{\Theta_\Delta}{\Theta_x}\). Using the fact that
+%     \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), we can conclude by
+%     \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Delta,
+%     \thyp{x}{A}{q}}}{\Delta, \thyp{x}{A}{q}}\); the desired result hence
+%     follows.
+%     \item \rle{split-right}: by right split association, we have that there
+%     exists \(\Theta_{\Xi, \thyp{x}{A}{q}}\),
+%     \(\csplits{\Theta}{\Theta_\Delta}{\Theta_{\Xi, \thyp{x}{A}{q}}}\),
+%     \(\csplits{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Theta_\Xi}{\Theta_x}\). Using the
+%     fact that \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by
+%     \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
+%     \thyp{x}{A}{q}}\); the desired result hence follows.
+%     \item \rle{split-dup}: since \(\tlin{A}{\trel}\), it follows by
+%     \rle{subst-cons} that \(\tlin{\Theta_x}{\trel}\), and hence that
+%     \(\csplits{\Theta_x}{\Theta_x}{\Theta_x}\). Therefore, by split association,
+%     we have that there exist \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\Theta_{\Xi,
+%     \thyp{x}{A}{q}}\) such that \(\csplits{\Theta}{\Theta_{\Delta,
+%     \thyp{x}{A}{q}}}{\Theta_{\Xi, \thyp{x}{A}{q}}}\). Using the facts that
+%     \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\),
+%     \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by \rle{subst-cons}
+%     that \(\issubst{\gamma}{\Theta_{\Delta, \thyp{x}{A}{q}}}{\Delta,
+%     \thyp{x}{A}{q}}\), \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
+%     \thyp{x}{A}{q}}\); the desired result hence follows.
+%     \item \rle{split-drop}: since \(\tlin{A}{\taff}\), it follows by
+%     \rle{subst-cons} that \(\tlin{\Theta_x}{\taff}\), and hence that
+%     \(\cwk{\Theta_x}{\cdot}\). Therefore, by split association,
+%     \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\), giving the desired result.
+%   \end{itemize}
+% \end{proof}
+
+% \begin{lemma}[Substitution] 
+%   Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+%   \begin{itemize}
+%     \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
+%     \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
+%     \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
+%     \begin{itemize}
+%       \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
+%       \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
+%       \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
+%     \end{itemize}
+%   \end{itemize}
+% \end{lemma}
+
+% \begin{proof} \
+%   \TODO{this}
+% \end{proof}
+
+% \begin{lemma}[Label Substitution] 
+%   Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
+%   \begin{itemize}
+%     \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
+%     \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
+%   \end{itemize}
+% \end{lemma}
+
+% \begin{proof} \
+%   \TODO{this}
+% \end{proof}
+
+% \subsection{Semantic Metatheory}
+
+% \begin{lemma}[Semantic Upgrade]
+%   For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
+%   \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
+% \end{lemma}
+
+% \begin{proof} \
+%   \TODO{this}
+% \end{proof}
+
+% \begin{lemma}[Semantic Weakening] 
+%   If \(\cwk{\Gamma}{\Delta}\), then 
+%   \begin{itemize}
+%     \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
+%     \[
+%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+%       ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
+%       = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+%     \]
+%     \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
+%     \[
+%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+%       ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
+%       = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
+%     \]
+%   \end{itemize}
+
+%   Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
+%   \[
+%     \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+%     ; \dnt{\lwk{\ms{L}}{\ms{K}}}
+%     = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
+%   \]
+% \end{lemma}
+
+% \begin{proof} \
+%   \TODO{this}
+% \end{proof}
+
+% \begin{lemma}[Substitution Coherence] \
+%   \begin{itemize}
+%     \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+%     \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+%   \end{itemize}
+% \end{lemma}
+
+% \begin{proof} \
+%   \TODO{this}
+% \end{proof}
+
+% \begin{lemma}[Semantic Substitution Splitting] 
+%   Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
+%   if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
+%   \[
+%     \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+%     = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
+%       \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
+%       \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
+%   \]
+% \end{lemma}
+
+% \begin{proof} \
+%   \TODO{this}
+% \end{proof}
+
+% \begin{theorem}[Semantic Substitution] 
+%   Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+%   \begin{itemize}
+%     \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
+%     \[
+%       \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
+%       ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+%       = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
+%     \]
+%     \item For all 
+%       \(\haslb{\Gamma}{t}{\ms{L}}\), 
+%       \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
+%       \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
+%     \[
+%       \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
+%       ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+%       = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
+%       ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
+%     \]
+%     \item For all 
+%       \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
+%       \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
+%       \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
+%       \[
+%         \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+%         ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
+%         = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
+%         ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
+%         + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+%       \]
+%   \end{itemize}
+%   Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
+%   \begin{itemize} 
+%     \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+%     \[
+%       \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+%       = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
+%     \]
+%     \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
+%     \[
+%       \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+%       + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
+%       = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
+%     \]
+%     where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
+%   \end{itemize}
+% \end{theorem}
+
+% \begin{proof} \
+%   \TODO{this}
+% \end{proof}
 
@@
 
+\label{apx:rewriting}
+
 \subsection{Syntax}
@@
 To properly be able to reason about congruence and congruence-based
-optimization, we need to introduce the concept of \textit{terms with holes},
+optimistion, we need to introduce the concept of \textit{terms with holes},
 which are defined using the grammar in Figure
@@
 \(\lhole{T}\), \(\lhole{W}\). To \textit{type} terms with holes, we need to
-introduce the concept of a \textit{hole-context} \(H\), which types a hole as being
-one of:
+introduce the concept of a \textit{hole-context} \(H\), which is a list of holes
+annotated as corresponding to one of:
 \begin{itemize}
-  \item \todo{term}
-  \item \todo{block}
-  \item \todo{control flow graph}
+  \item A term \(\tyhole{\lhole{X}}{\Delta}{p}{A}{q}\) of type \(A\), quantity
+  \(q\), and purity \(p\) using the variables in \(\Delta\). Such a hole can be
+  filled by any term \(a\) satisfying \(\hasty{\Delta}{p}{a}{A}{q}\)
+  \item A block \(\blkhole{\lhole{T}}{\Delta}{\ms{L}}\) targeting \(\ms{L}\)
+  using the variables in \(\Delta\). Such a hole can be filled by any block
+  \(t\) satisfying \(\haslb{\Delta}{t}{\ms{L}}\).
+  \item A control-flow graph \(\cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\) with input
+  labels \(\ms{L}\) and output labels \(\ms{K}\). Such a hole can be filled by
+  any CFG satisfying \(\lhaslb{\ms{L}}{L}{\ms{K}}\).
 \end{itemize}
-\TODO{hole typing rules for expressions, blocks, control-flow graphs; everything
-else is the same}.
 
@@
 \end{figure}
+
+We introduce new typing judgements for terms, blocks, and CFGs with holes in
+Figure \ref{fig:ssa-holes-judgements}. In each case, they correspond to the
+judgement that, given a collection \(\mc{H}\) terms, blocks, and CFGs to fill
+each hole in \(H\), would satisfy the corresponding judgement without holes in
+\ref{fig:ssa-judgements}.
+\begin{figure}
+  \begin{center}        
+    \begingroup
+    \renewcommand{\arraystretch}{1.5}
+    \setlength{\tabcolsep}{2em}
+    \begin{tabular}{rl}
+        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
+        \(\isrw{\mc{H}}{I}{H}\) & \(\mc{H}\) fills every hole in \(H\), modulo
+        holes \(I\) \\
+        \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) & Given
+        \(\isrw{\mc{H}}{\cdot}{H}\), \(\hasty{\Gamma}{p}{[\mc{H}]\mhole{a}}{A}{q}\) \\
+        \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) & Given
+        \(\isrw{\mc{H}}{\cdot}{H}\), \(\haslb{\Gamma}{[\mc{H}]\mhole{t}}{\ms{L}}\) \\
+        \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) & Given
+        \(\isrw{\mc{H}}{\cdot}{H}\),
+        \(\haslb{\Gamma}{[\mc{H}]\mhole{t}}{\ms{L}}\) \\
+    \end{tabular}
+    \endgroup
+  \end{center}
+  \caption{Typing judgements for \isotopessa terms, blocks, and CFGs with holes}
+  \Description{Typing judgements for isotope-SSA terms, blocks, and CFGs with
+  holes}
+  \label{fig:ssa-holes-judgements}
+\end{figure}
+The typing rules for terms, blocks, and CFGs with holes are exactly the same as
+the corresponding rules without holes, modulo the additional rules
+\rle{hole-tm}, \rle{hole-blk}, and \rle{hole-cfg} to type the holes themselves;
+they are given in full in Figures \ref{fig:ssa-term-holes-typing} and
+\ref{fig:ssa-block-holes-typing}.
 
@@
   \end{gather*}
-  \caption{Typing rules for \isotopessa expressions with holes}
-  \Description{Typing rules for isotope-SSA expressions with holes}
-  \label{fig:terms-with-holes-typing}
+  \caption{Typing rules for \isotopessa terms with holes}
+  \Description{Typing rules for isotope-SSA terms with holes}
+  \label{fig:ssa-term-holes-typing}
 \end{figure}
@@
   \Description{Typing rules for isotope-SSA blocks with holes}
-  \label{fig:blocks-with-holes-typing}
+  \label{fig:ssa-block-holes-typing}
 \end{figure}
@@
 
-\TODO{some text about rewrite typing rules}
-
+This formalism allows us to state the \textit{syntactic rewriting} theorem,
+which says, in essence, that filling holes with valid terms yields valid terms.
 \begin{theorem}[Rewriting]
@@
 
-\TODO{text}
-
+We can interpret a hole context semantically as the \textit{set} of all possible
+tuples of morphisms which would typecheck in place of the holes in the context,
+as specified in Figure \ref{fig:holes-semantics}.
 \begin{figure}
@@
 
-\TODO{some text about semantics of holes}
-
 \begin{figure}
@@
 
-\TODO{fill}
-
-\TODO{some text about semantics of terms with holes}
-
 \begin{figure}
@@
 \end{figure}
-
-\TODO{fill}
-
-\TODO{some text about semantics of blocks with holes}
 
@@
 
-\TODO{fill}
-
 \TODO{some text about semantics of rewrites}
 
-\begin{lemma}[Rewriting Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
+% \begin{lemma}[Rewriting Coherence] \
+%   \begin{itemize}
+%     \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
+%     \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+%     \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
+%   \end{itemize}
+% \end{lemma}
 

### 2023-11-16 pldi24/paper.tex
Microtap

@@
 For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq
-\omega}) = Z(A^*) = \{[\,]\}\), while \(\mc{SZ}(\nats, \nats \cup \{\infty\}) =
-\{0\} \subseteq Z(\nats) = \nats\). Generally, \(\mc{SZ}(M, I)\) always contains
-the unit of \(M\).
-
-We can now define the \textit{trace monad} of a stream action as follows:
+\omega}) = \{[\,]\}\). Generally, \(\mc{SZ}(M, I)\) always contains
+the unit of \(M\). We can now define the \textit{trace monad} of a stream action as follows:
 \begin{definition}[Trace Monad]
-  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the
+  Given a stream action \(\Theta: M^\omega \to I\), we can define the
   \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A
-  \times M + I)\), with \(\eta_A = \lambda a.\iota_0 (a, 1)\) and 
-  \[
+  \times M + I)\) with unit \(\eta_A = \lambda a.\iota_0 (a, 1)\) and
+  multiplication
+  \(
     \mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}
-  \]
+  \),
   where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') =
-  \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). The iteration operator \(f^\dagger(a) = f^\infty(a) \cup
+  \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). We can
+  equip this monad with an iteration operator \(f^\dagger(a) = f^\infty(a) \cup
   \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t
-  \in f_n(a)\}\) yields an Elgot structure, where
+  \in f_n(a)\}\), yielding an Elgot structure, where
   \[

### 2023-11-16 pldi24/paper.tex
Tap

@@
 \ref{fig:ssa-judgements}.
+
+\TODO{note on rewrites}
+
 \begin{figure}
@@
         \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
-        \(\isrw{\mc{H}}{I}{H}\) & \(\mc{H}\) fills every hole in \(H\), modulo
-        holes \(I\) \\
+        \(\isrw{\mc{H}}{I}{H}\) & \(\mc{H}\) is a rewrite from \(I\) to \(H\),
+        i.e. fills every hole in \(H\) modulo \(I\) \\
         \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) & Given
@@
   \begin{align*}
+      \\
     \dnt{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}
@@
 \end{figure}
+We can now quite easily give semantics for terms, blocks, and CFGs with holes
+which simply correspond to "filling in" any holes with the appropriate morphism
+from the hole context, with all other rules interpreted identically to their
+regular variants: this is done in Figures \ref{fig:ssa-term-holes-semantics} and
+\ref{fig:ssa-block-holes-semantics}.
 
@@
   \end{gather*}
-  \caption{Semantics for \isotopessa expressions with holes}
-  \Description{Semantics for isotope-SSA expressions with holes}
-  \label{fig:terms-with-holes-semantics}
+  \caption{Semantics for \isotopessa terms with holes}
+  \Description{Semantics for isotope-SSA terms with holes}
+  \label{fig:ssa-term-holes-semantics}
 \end{figure}
@@
   \Description{Semantics for isotope-SSA blocks with holes}
-  \label{fig:blocks-with-holes-semantics}
+  \label{fig:ssa-block-holes-semantics}
 \end{figure}
@@
 \end{figure}
-
-\TODO{some text about semantics of rewrites}
 

### 2023-11-16 pldi24/paper.tex
Microtap

@@
 \end{figure}
-The typing rules for terms, blocks, and CFGs with holes are exactly the same as
-the corresponding rules without holes, modulo the additional rules
-\rle{hole-tm}, \rle{hole-blk}, and \rle{hole-cfg} to type the holes themselves;
-they are given in full in Figures \ref{fig:ssa-term-holes-typing} and
-\ref{fig:ssa-block-holes-typing}.
 
@@
 
-This formalism allows us to state the \textit{syntactic rewriting} theorem,
-which says, in essence, that filling holes with valid terms yields valid terms.
+The typing rules for terms, blocks, and CFGs with holes are exactly the same as
+the corresponding rules without holes, modulo the additional rules
+\rle{hole-tm}, \rle{hole-blk}, and \rle{hole-cfg} to type the holes themselves;
+they are given in full in Figures \ref{fig:ssa-term-holes-typing} and
+\ref{fig:ssa-block-holes-typing}. This formalism allows us to state the
+\textit{syntactic rewriting} theorem, which says, in essence, that filling holes
+with valid terms yields valid terms.
 \begin{theorem}[Rewriting]

### 2023-11-17 pldi24/paper.tex
Cut a page

@@
   \begin{itemize}
-    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
+    \item If \(\cwk{\Gamma}{\Delta}\), \(q \subseteq r\), and
     \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
-    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{L}}\)
-    \item If \(\lwk{\ms{L}}{\ms{K}}\), then:
-    \begin{itemize}
-      \item If \(\haslb{\Gamma}{t}{\ms{L}}\) then \(\haslb{\Gamma}{t}{\ms{K}}\)
-      \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\) then \(\lhaslb{\ms{W}}{L}{\ms{L}}\)
-    \end{itemize}
+    \item If \(\cwk{\Gamma}{\Delta}\) and \(\haslb{\Delta}{t}{\ms{L}}\), then
+    \(\haslb{\Gamma}{t}{\ms{L}}\)
+    \item If \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{K}}\)
+    % \item If \(\lwk{\ms{L}}{\ms{K}}\) and \(\lhaslb{\ms{W}}{L}{\ms{L}}\) then
+    % \(\lhaslb{\ms{W}}{L}{\ms{K}}\)
   \end{itemize}
 \end{lemma}
-That is, we can do the following and still keep our typing judgements valid
-\begin{itemize}
-  \item Add unused \textit{affine} variables
-  \item \textit{Reduce} the quantity of an expressions type
-  \item Add arbitrary unused labels, and remove unused \textit{affine} variables
-  from existing labels
-\end{itemize}
-This can be proved by a relatively straightforward mutual induction.
+That is, we can add unused \textit{affine} variables to a context,
+\textit{reduce} the quantity of an term's type, add arbitrary unused labels, and
+remove \textit{affine} variables from existing labels and still keep our typing
+judgements valid. This can be proved by a relatively straightforward mutual
+induction.
 
@@
 
-We begin with some basic definitions:
+We begin with some standard definitions:
 \begin{definition}[Binoidal Category] 
@@
   A \textbf{premonoidal category} is a binoidal category with an identity object
-  \(I: \mc{C}\) such that
-  \begin{itemize}
-    \item There exist central natural isomorphisms \(\lambda_A: I \otimes A
-    \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I \simeq A\) (the
-    \textbf{right unitor})
-    \item There exists a central isomorphism \(\alpha_{A, B, C}: (A \otimes B)
+  \(I: \mc{C}\) such that there exist central natural isomorphisms \(\lambda_A:
+  I \otimes A \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I
+  \simeq A\) (the \textbf{right unitor}), and \(\alpha_{A, B, C}: (A \otimes B)
   \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
-    \item Satisfying the \textbf{pentagon} and \textbf{triangle laws}; see e.g.
-    \citet{maclane:71}.
-  \end{itemize}
-  We say a premonoidal category is \textbf{symmetric} if it is equipped with a
-  central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
-  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} (see
-  \cite{maclane:71}). We say a premonoidal category is \textbf{monoidal} if
-  every morphism is central.
+  satisfying the \textit{pentagon} and \textit{triangle laws} (see e.g.
+  \citet{maclane:71}). We say a premonoidal category is \textbf{symmetric} if it
+  is equipped with a central natural isomorphism \(\sigma_{A, B}: A \otimes B
+  \simeq B \otimes A\) (the \textbf{symmetry}) satisfying the \textit{hexagon
+  law} (see \cite{maclane:71}). We say a premonoidal category is
+  \textbf{monoidal} if every morphism is central.
 \end{definition}
@@
 Unfortunately, the notion of "purity" given by an effectful category only allows
-us to perform relatively weak optimistions in the presence of effects. To
-remedy this, we want to be able to make a finer distinction between morphisms
-which are merely \textit{central} and those which are also \textit{relevant}
-(i.e., copying the input and applying them twice is the same as applying them
-once and copying the output) and/or \textit{affine} (i.e., applying them and
-discarding the result is the same as never doing them at all). Doing so also
-allows us to add the capability of \textit{distinguishing} between objects which
-are affine, relevant, neither or both, enabling us to reason about substructural type systems categorically.
-
-To do this, we begin by introducing the notion of a \textit{symmetric premonoidal subcategory}, as follows:
+us to perform relatively weak optimistions in the presence of effects. To remedy
+this, we want to be able to make a finer distinction between morphisms which are
+merely \textit{central} and those which are also \textit{relevant} (i.e.,
+copying the input and applying them twice is the same as applying them once and
+copying the output) and/or \textit{affine} (i.e., applying them and discarding
+the result is the same as never doing them at all). Doing so also allows us to
+add the capability of \textit{distinguishing} between objects which are affine,
+relevant, neither or both, enabling us to reason about substructural type
+systems categorically.
+
+To do this, we begin by introducing the notion of a \textit{symmetric
+premonoidal subcategory}, as follows:
 \begin{definition}[(Symmetric) Premonoidal Subcategory]
@@
 \(\dnt{\text{"judgement"}}\) for \(\dnt{D}\), where \(D\) is a derivation of
-"judgement"; this is justified by the coherence lemma:
-\begin{lemma}[Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
+"judgement"; in general, it can be proven that the semantics of any two such
+derivations are equal.
+% \begin{lemma}[Coherence] \
+%   \begin{itemize}
+%     \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
+%     \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+%     \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
+%     \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
+%     \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
+%   \end{itemize}
+% \end{lemma}
 Our treatment of \rle{var} is quite standard: the term \(x\) types in \(\Gamma\)
@@
 \begin{enumerate}
-  \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad
-  \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\);
-  similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
+  \item The \textbf{unit submonad} \(\eta\) of every monad \(\mb{T}\) with
+  \(\eta(A) = \{\eta_A\;a \mid a \in A\}\).
+  %; similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\). 
   \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
   intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
-  submonads, forming a complete meet-semilattice with bottom element
-  \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete
-  lattice in the standard manner.
+  submonads, forming a complete meet-semilattice with bottom element \(\eta\)
+  and top element \(\mb{T}\),
+  % ; this induces a complete lattice in the standard manner.
   \item In particular, the intersection of a relevant/affine/commutative
@@
 where
-\begin{equation}
+\(
   \ms{hoare}\;f\;\Phi\;\Psi = \forall a\;h. \Phi\;a\;h \implies \forall \iota_0\;(b, h') \in f\;a\;h. \Psi\;b\;h'
-\end{equation}
+\)
 is the usual partial correctness triple definition and the separating
@@
   \item Associators and inverse associators given by \(\alpha_{(A, \Phi), (B,
-  \Psi), (C, \Theta)} = \alpha_{A, B, C}\), which are valid since \(\ast\) is
-  associative
+  \Psi), (C, \Theta)} = \alpha_{A, B, C}\)
+  %, which are valid since \(\ast\) is associative
   \item Unitors and inverse unitors given by \(\lambda_{(A, \Phi)} =
-  \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\), which are valid since \(\ms{emp}\)
+  \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\)
+  %, which are valid since \(\ms{emp}\)
   is the unit of \(\ast\)
@@
   \item Affine objects to be precisely those for which \(\ms{drop} =
-  \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), which for
-  this logic corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend
-  on the heap.
-  \item Relevant objects to be precisely those for which \(\ms{split} =
-  \upg{\lambda a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), again in
-  this case those for which \(\Phi\) does not depend on the heap.
-  \item Central morphisms to be those which do not print, diverge, \textit{or} modify the
-  heap. We will consider how to relax this condition later.
+  \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), and
+  relevant objects to be precisely those for which \(\ms{split} = \upg{\lambda
+  a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), which for this logic
+  corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend on the heap.
+  \item Central morphisms to be those which do not print, diverge, \textit{or}
+  modify the heap. % We will consider how to relax this condition later.
   \item Relevant and affine morphisms to be the lifts of relevant/affine
@@
 \end{equation}
-This implies that \(\ms{Sep}\) is an \isotopessa model supporting the following interesting operations
+This implies that \(\ms{Sep}\) is an \isotopessa model supporting the following
+interesting operations:
 \begin{itemize}
@@
   \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Byte}^*, \ms{emp}), (\mb{1},
-  \ms{emp}))\)
-  \item \(\ms{rand}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for \(A\) nonempty
+  \ms{emp}))\) and \(\ms{rand}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for
+  \(A\) nonempty
   \item \(\ms{set}: \mc{I}^\varnothing_0((\nats \times \nats, \lambda (\ell,
   \cdot). \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell
-  \mapsto v))\), where \(\ms{set}\) is modified slightly to return the address
-  at which the value passed in was stored to overcome a lack of dependency
-  between the source and target objects.
-  \item \(\ms{get}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)
+  \mapsto v))\) and \(\ms{get}: \mc{I}^\varnothing_0((\nats \times \nats,
+  \lambda (-, \ell). \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists
+  v. \ell \mapsto v))\), which are modified slightly to return the input address.
   \item \(\ms{alloc}: \mc{I}^\varnothing_0((\nats, \ms{emp}), (\nats, \lambda
-  \ell. \exists v. \ell \mapsto v))\)
-  \item \(\ms{free}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\mb{1}, \ms{emp}))\)
+  \ell. \exists v. \ell \mapsto v))\) and \(\ms{free}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\mb{1}, \ms{emp}))\)
 \end{itemize}
@@
 
-Given a partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n \in N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by the function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\), where the dependent product \(\Sigma_nP_n\) is given the lexicographic order.
-In particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega \to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on finite pomsets.
-
-We define a \textbf{program order pomset} to be a pomset with \(\mc{A}_{\ms{PO}} =
-\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
+Given a partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n
+\in N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by
+the function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\),
+where the dependent product \(\Sigma_nP_n\) is given the lexicographic order. In
+particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega
+\to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on
+finite pomsets.
+
+We define a \textbf{program order pomset} to be a pomset with \(\mc{A}_{\ms{PO}}
+= \mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
 \textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
@@
 \ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
-\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
+\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is taken as a stream action on finite
 program order pomsets given by sequential composition for \(\mc{A}\). This
@@
 operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
-\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
-semantics
+\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with semantics
 \begin{equation}
@@
 
-On that note, we support the parallel execution of multiple (potentially
-nonterminating) morphisms as follows:
+On that note, we can define the \textit{parallel execution} of two morphisms as
+follows:
 \begin{equation}
@@
 \end{equation}
-It's tempting to try to use this as a tensor product of morphisms and so
-graduate to a bona fide monoidal category rather than our current premonoidal
-setting, since spatial concurrency is one of the headline use-cases of the
-former, but unfortunately, sliding still fails:
-\(
-  (W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})
-  \neq (\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}} || \ms{id})
-\),
-since the left-hand side will emit the pomset \([x := a, y := b]\) while the
-right-hand side will emit the pomset \([y := b, x := a]\). If we did want this
-behaviour, we'd need to have a significantly more sophisticated model of
-composition which effectively takes nontermination into account, which we will
-leave to future work.
+It's tempting to try to use this as a tensor product of morphisms to obtain a
+monoidal (rather than preomonoidal) category, but unfortunately, sliding still
+fails: \((W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})\) emits the
+pomset \([x := a, y := b]\), while \(\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}}
+|| \ms{id})\) emits the pomset \([y := b, x := a]\). 
+% If we did want this behaviour, we'd need to have a significantly more
+% sophisticated model of composition which effectively takes nontermination into
+% account, which we will leave to future work.
 
@@
 implement TSO ordering by buffering all our writes, in which case they are only
-visible to the local thread. On the other hand, reads events will first attempt
+visible to the local thread. On the other hand, read events will first attempt
 to read from the buffer, and, if there is no corresponding write in the buffer,
@@
 thread's write buffer. We may then define the set of TSO actions
-\(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be defined
-to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret
-as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index (with the empty list corresponding to \(\{\delta\}\)). In particular, we
-define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
-where \(\Sigma\) is the stream action of finite pomsets over
-\(\mc{A}_{\ms{TSO}}\) on pomsets over \(\mc{A}_{\ms{TSO}}\); we can view
-\(\ms{PO}\) as a submonad of this monad in the standard manner.
+\(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\); a pomset over this set
+of actions is called a \textbf{TSO pomset}. A buffer will be defined to be a
+list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret as
+linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index (with the empty list
+corresponding to \(\{\delta\}\)). In particular, we define the monad \(\ms{TSO}
+= \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\), where \(\Sigma\) is taken as a
+the stream action of finite TSO pomsets on TSO pomsets. We can view
+\(\ms{PO}\) as a submonad of this monad.
 
 Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
-\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
-\begin{equation}
-  (L;\{\bufloc{x} := v\})[x] = v
-  \qquad
+\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer by induction to be:
+\(
+  (L;\{\bufloc{x} := v\})[x] = v,
+\),
+\(
   (L;\{\bufloc{y} := v\})[x] = L[x]
-  \qquad
+\),
+\(
   [][x] = \bot
-\end{equation}
+\).
 Note that writes at the \textit{end} of the buffer are prioritized, since later
-writes overwrite earlier ones! The semantics of reads and writes are then given
-by
+writes overwrite earlier ones! The semantics of reads and writes can then be
+given by
 \begin{equation}
@@
 \end{equation}
-To be able to perform synchronization, we will also need to introduce a \(\ms{fence}: \mc{I}^\varnothing_0(\mb{1}, \mb{1})\) instruction, which simply causes all actions before the fence to be observed before any actions after the fence.
-For \(\ms{TSO}\), implementing this is as simple as flushing the buffer, giving
-\begin{equation}
+To be able to perform synchronization, we will also need to introduce a
+\(\ms{fence}: \mc{I}^\varnothing_0(\mb{1}, \mb{1})\) instruction, which simply
+causes all actions before the fence to be observed before any actions after the
+fence. For \(\ms{TSO}\), implementing this is as simple as flushing the buffer,
+i.e., we define
+\(
   \dnt{\ms{fence}} = \lambda ()\;L. \{((), [], L;\{\delta\})\}
-\end{equation}
+\).
+
 We can now define the parallel composition of morphisms in a "fork-join" style

### 2023-11-17 pldi24/paper.tex
Tap
Cut half a page

@@
 \begin{abstract}
-  Static single-assignment form (SSA) is one of the most widely used
-  compiler intermediate representations. In this paper, we build on
-  the observation that SSA can be seen as a language of basic blocks
-  with arguments (or first-order procedures tail-calling one another)
-  to give a type theory for SSA programs. Notably, the type system and
-  equational theory of our language is designed to support
-  substituting impure terms for variables, which makes it easy to
-  validate compiler optimisations such as hoisting assignments out of
-  loops.
-
-  We give a categorical axiomatisation of our type theory and show
-  that our equational theory is sound with respect to it. Our
-  categorical semantics generalises the notions of premonoidal
-  category and Freyd category to support substructural features,
-  including both support for substructural types, as well as
-  F\"{u}hrmann's notion of central, copyable and discardable
-  morphisms.
-
+  Static single-assignment form (SSA) is one of the most widely used compiler
+  intermediate representations. In this paper, we build on the observation that
+  SSA can be seen as a language of basic blocks with arguments (or first-order
+  procedures tail-calling one another) to give a type theory for SSA programs.
+  Notably, the type system and equational theory of our language is designed to
+  support substituting impure terms for variables, which makes it easy to
+  validate compiler optimisations such as hoisting assignments out of loops.
+
+  We give a categorical axiomatisation of our type theory and show that our
+  equational theory is sound with respect to it. Our categorical semantics
+  generalises the notions of premonoidal category and Freyd category to support
+  substructural features, including both support for substructural types, as
+  well as F\"{u}hrmann's notion of central, copyable and discardable morphisms.
   Finally, we exhibit a number of concrete models of our categorical
-  axiomatisation, including in particular a model of TSO-style weak
-  memory based on the semantics of \citet{sparky}. This demonstrates
-  that our approach is strong enough to connect machine models of
-  concurrency to compiler IRs supporting a rich collection of sound
-  rewrites.
+  axiomatisation, including in particular a model of TSO-style weak memory based
+  on the semantics of \citet{sparky}. This demonstrates that our approach is
+  strong enough to connect machine models of concurrency to compiler IRs
+  supporting a rich collection of sound rewrites.
 \end{abstract}
@@
 semantically-justified.
-\item Our weak memory semantics and our results about Elgot monads are
-formalised in the Lean theorem prover.
+% \item Our weak memory semantics and our results about Elgot monads are
+% formalised in the Lean theorem prover.
 \end{itemize}
@@
 
-Note that we cannot simply have \(\ms{Rel}^{\ms{a}} = \ms{Rel}\), since, for
-any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) =
-\varnothing\), we have
-\begin{equation}
-  (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a) = \{()\}
-\end{equation}
-We will see that this corresponds to the fact that the following programs are
-not semantically equal
-\begin{equation}
-  \ms{let}\;x = \ms{loop}(); e \not\simeq e \qquad \text{where} \qquad x \notin \ms{fv}(e)
-\end{equation}
-Similarly, we cannot simply have \(\ms{Rel}^{\ms{r}} = \ms{Rel}\), since,
-for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)|
-\supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
+Note that we cannot simply have \(\ms{Rel}^{\ms{a}} = \ms{Rel}\), since, for any
+morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) =
+\varnothing\), we have \( (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a)
+= \{()\} \). We will see that this corresponds to the fact that \( \ms{let}\;x =
+\ms{loop}(); e \) is not semantically equal to \(e\) even where \(x \notin
+\ms{fv}(e)\). Similarly, we cannot simply have \(\ms{Rel}^{\ms{r}} = \ms{Rel}\),
+since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A,
+|R(a)| \supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
 \begin{equation}
@@
 since only the left-hand side contains the pair \((b_0, b_1)\), this
-corresponding to the fact that the following programs are not semantically equal
-\begin{equation}
-  \ms{let}\;x = \ms{rand}(); (x, x) \not\simeq (\ms{rand}(), \ms{rand}())
-\end{equation}
+corresponding to the fact that the program \(\ms{let}\;x = \ms{rand}(); (x, x)\) is
+not semantically equal to \((\ms{rand}(), \ms{rand}())\).
+
 To get a substructural premonoidal category, we can apply the state monad
@@
 In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a
-substructural premonoidal category, with \(\mc{C}^\varnothing = \mc{C}\) and
-\(\mc{C}^{\{\ms{a}, \ms{r}\}}, \mc{C}^{\{\ms{a}\}}, \mc{C}^{\{\ms{r}\}}\) the
-subcategory consisting only of the associators, unitors, and symmetry, and only
-the identity object considered affine and relevant. Similarly, every
-\textit{Cartesian} category \(\mc{C}\) can be viewed as a substructural
-Cartesian category with all morphisms both relevant and affine; in fact, this is
-the case \textit{if and only if} a category is Cartesian, hence the term
-"substructural Cartesian category." We also note the following:
+substructural premonoidal category, taking \(\mc{C}^{\ms{a}}\) and
+\(\mc{C}^{\ms{r}}\) to be the subcategory consisting only of the associators,
+unitors, and symmetry, and only the identity object to be affine and relevant.
+Similarly, every \textit{Cartesian} category \(\mc{C}\) can be viewed as a
+substructural Cartesian category with all morphisms and both relevant and
+affine; in fact, this is the case \textit{if and only if} a category is
+Cartesian. We also note the following:
 \begin{proposition}
@@
   \end{itemize}
-  A monad \(\mb{T}\) on \(\mc{C}\) such that its Kleisli cateogry
-  \(\mc{C}_{\mb{T}}\) is Elgot is called an \textbf{Elgot monad}.
+  % A monad \(\mb{T}\) on \(\mc{C}\) such that its Kleisli cateogry
+  % \(\mc{C}_{\mb{T}}\) is Elgot is called an \textbf{Elgot monad}.
 \end{definition}
-Some important properties which hold for all Elgot categories include:
-\begin{itemize}
-  \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id},
-  (h;[\iota_0, g])^\dagger]\)
-  \item \textbf{Squaring:} \((f;[\iota_0, f])^\dagger = f^\dagger\)
-\end{itemize}
-Proofs can be found in Lemma 31 of \cite{goncharov-squaring}.
+% Some important properties which hold for all Elgot categories include:
+% \begin{itemize}
+%   \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id},
+%   (h;[\iota_0, g])^\dagger]\)
+%   \item \textbf{Squaring:} \((f;[\iota_0, f])^\dagger = f^\dagger\)
+% \end{itemize}
+% Proofs can be found in Lemma 31 of \cite{goncharov-squaring}.
 
 One important way of generating Elgot categories is through monads: a monad
-whose Kliesli category is Elgot is called an \textbf{Elgot monad}. Examples of
-Elgot monads include \(\ms{Option}\) (where \(\ms{None}\) is interpreted as
-nontermination), the powerset monad (where the empty set is interpreted as
-nontermination), and the trace monads we will explore in Section
-\ref{ssec:trace-monads}.
-
-It turns out that various monad transformers preserve Elgot structure, allowing
-us to build up Elgot monads in a composable fashion. In particular,
+\(\mb{T}\) whose Kliesli category \(\mb{C}_{\mb{T}}\) is Elgot is called an
+\textbf{Elgot monad}. Examples of Elgot monads include \(\ms{Option}\) (where
+nontermination is \(\ms{None}\)), the powerset monad (where nontermination is
+the empty set), which can both be viewed as special cases of the trace monads we
+will explore in Section \ref{ssec:trace-monads}. It turns out that various monad
+transformers preserve Elgot structure, allowing us to build up Elgot monads in a
+composable fashion. In particular,
 \begin{proposition}
@@
 
-We're now ready to give a semantics for an \isotopessa dialect with a set of
-types \(\mc{T}\) and instructions \(\mc{I}\). We'll need:
+We can now give a semantics for an \isotopessa dialect with types
+\(\mc{T}\) and instructions \(\mc{I}\). We'll need:
 \begin{itemize}
@@
 associated split \(\csplits{\Gamma}{\Delta}{\cdot}\) postcomposed with a unitor.
-Finally, we interpret a label-set weakening \(\lwk{\ms{L}}{\ms{K}}\) as a
-pure morphism \(\mc{C}^\tint_1(\dnt{\ms{L}}, \dnt{\ms{K}})\), where
-\rle{join-nil} is the identity, \rle{join-cons} is the sum of
-\(\dnt{\lwk{\ms{L}}{\ms{K}}}\) and \(\dnt{\cwk{\Gamma}{\Delta}}\), and
-\rle{join-zero} is \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) postcomposed with the left
-injection.
+Finally, we interpret a label-set weakening \(\lwk{\ms{L}}{\ms{K}}\) as a pure
+morphism \(\mc{C}^\tint_1(\dnt{\ms{L}}, \dnt{\ms{K}})\), where \rle{join-nil} is
+the identity, \rle{join-cons} is the sum of \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) and
+\(\dnt{\cwk{\Gamma}{\Delta}}\), and \rle{join-zero} is
+\(\dnt{\lwk{\ms{L}}{\ms{K}}}\) postcomposed with the left injection.
 

### 2023-11-17 pldi24/paper.tex
Pulled grammar to appendix

@@
 
-The grammar for our syntax is given in Figure \ref{fig:ssa-grammar}, while the
-typing judgements themselves are given in \ref{fig:ssa-judgements}. We begin by
-giving a grammar for \textit{types} \(A, B, C\), parametrized by a set of
-\textit{base types} \(X\); all syntactically well-formed types are also
-semantically valid. We postulate the existence of a \textit{unit type}
-\(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all tensor products \(A
-\otimes B\); we write \(\otimes\) to underscore the fact that types are allowed
-to be substructural, assigning every type a linearity \(\ms{lin}(A)\) satisfying
-the constraints in Figure \ref{fig:ssa-structural}. In particular, the rules
-\rle{unit-lin} and \rle{bool-lin} state that both the unit and boolean types
-must be intuitionistic; the rule \rle{pair-lin} says that the linearity of a
-pair is simply the intersection of the linearity of its components. The rules
-\rle{nil-lin} and \rle{cons-lin} together say that the linearity of a context is
-the intersection of the linearity of its variables, where the linearity of a
-variable is the intersection of the linearity of its type and its quantity
-annotation.
+The formal grammar for our syntax is given in Figure \ref{fig:ssa-grammar} in
+the appendix, while the typing judgements themselves are given in
+\ref{fig:ssa-judgements}. We begin by giving a grammar for \textit{types} \(A,
+B, C\), parametrized by a set of \textit{base types} \(X\); all syntactically
+well-formed types are also semantically valid. We postulate the existence of a
+\textit{unit type} \(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all
+tensor products \(A \otimes B\); we write \(\otimes\) to underscore the fact
+that types are allowed to be substructural, assigning every type a linearity
+\(\ms{lin}(A)\) satisfying the constraints in Figure \ref{fig:ssa-structural}.
+In particular, the rules \rle{unit-lin} and \rle{bool-lin} state that both the
+unit and boolean types must be intuitionistic; the rule \rle{pair-lin} says that
+the linearity of a pair is simply the intersection of the linearity of its
+components. The rules \rle{nil-lin} and \rle{cons-lin} together say that the
+linearity of a context is the intersection of the linearity of its variables,
+where the linearity of a variable is the intersection of the linearity of its
+type and its quantity annotation.
 
@@
 and CFGs in Section \ref{sec:semantics}.
-
-\begin{figure}
-  \begin{center}
-    \begin{grammar}
-      <\(A, B, C\)> ::= 
-      \(X\)
-      \;|\; \(\mathbf{1}\)
-      \;|\; \(\mathbf{2}\)
-      \;|\; \(A \otimes B\)
-
-      <\(a, b, c, e\)> ::= \(x\) 
-      \;|\; \(f\;a\)
-      \;|\; \((a, b)\) 
-      \;|\; \(()\) 
-      \;|\; \(\ctt\) 
-      \;|\; \(\cff\)
-      \;|\; \(\letexpr{x}{a}{e}\)
-      \;|\; \(\letexpr{(x, y)}{a}{e}\)
-      % \;|\; \(\lbsplice{\ell}{x: A}{t}\)
-
-      <\(s, t\)> ::= \(\lbrb{\ell}{a}\) 
-      \;|\; \(\ite{e}{s}{t}\)
-      \;|\; \(\letstmt{x}{a}{t}\)
-      \;|\; \(\letstmt{(x, y)}{a}{t}\)
-      \;|\; \(\ewhere{t}{L}\)
-
-      <\(L\)> ::= \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{t}, L\)
-
-      <\(\Gamma\)> ::= \(\cdot\) \;|\; \(\Gamma, \thyp{x}{A}{q}\)
-
-      <\(\ms{L}\)> ::= \(\cdot\) \;|\; \(\ms{L}, \lbl{\ell}[\Gamma](x: A)\)
-
-      <\(p\)> ::= 0 \;|\; 1
-
-      <\(q\)> ::= \(\varnothing\) 
-      \;|\; \(\taff\) 
-      \;|\; \(\trel\) 
-      \;|\; \(\tint\)
-    \end{grammar}
-  \end{center}
-  \caption{Grammar for \isotopessa, parametrized over a set of instructions \(f \in \mc{I}\)}
-  \Description{Grammar for isotope-SSA}
-  \label{fig:ssa-grammar}
-\end{figure}
 
@@
 
+\begin{figure}
+  \begin{center}
+    \begin{grammar}
+      <\(A, B, C\)> ::= 
+      \(X\)
+      \;|\; \(\mathbf{1}\)
+      \;|\; \(\mathbf{2}\)
+      \;|\; \(A \otimes B\)
+
+      <\(a, b, c, e\)> ::= \(x\) 
+      \;|\; \(f\;a\)
+      \;|\; \((a, b)\) 
+      \;|\; \(()\) 
+      \;|\; \(\ctt\) 
+      \;|\; \(\cff\)
+      \;|\; \(\letexpr{x}{a}{e}\)
+      \;|\; \(\letexpr{(x, y)}{a}{e}\)
+      % \;|\; \(\lbsplice{\ell}{x: A}{t}\)
+
+      <\(s, t\)> ::= \(\lbrb{\ell}{a}\) 
+      \;|\; \(\ite{e}{s}{t}\)
+      \;|\; \(\letstmt{x}{a}{t}\)
+      \;|\; \(\letstmt{(x, y)}{a}{t}\)
+      \;|\; \(\ewhere{t}{L}\)
+
+      <\(L\)> ::= \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{t}, L\)
+
+      <\(\Gamma\)> ::= \(\cdot\) \;|\; \(\Gamma, \thyp{x}{A}{q}\)
+
+      <\(\ms{L}\)> ::= \(\cdot\) \;|\; \(\ms{L}, \lbl{\ell}[\Gamma](x: A)\)
+
+      <\(p\)> ::= 0 \;|\; 1
+
+      <\(q\)> ::= \(\varnothing\) 
+      \;|\; \(\taff\) 
+      \;|\; \(\trel\) 
+      \;|\; \(\tint\)
+    \end{grammar}
+  \end{center}
+  \caption{Grammar for \isotopessa, parametrized over a set of instructions \(f \in \mc{I}\)}
+  \Description{Grammar for isotope-SSA}
+  \label{fig:ssa-grammar}
+\end{figure}
+
 % \section{Proofs}

### 2023-11-17 pldi24/paper.tex
Rest of said page...

@@
 \(\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}\) for blocks, as defined in Figure
-\ref{fig:ssa-subst-ops}. Note that this operation cannot be easily defined for
-CFGs. In Section \ref{sec:semantics}, we will show that this is semantically
-equivalent to regular capture-avoiding substitution, due to the fact that all
-terms in a substitution are central \textit{and} that only affine computations
-can be used in a term substituting for an affine variable, and only relevant
-computationbs can be used in a term substituting for a relevant variable.
+\ref{fig:ssa-subst-ops}. 
+%
+% Note that this operation cannot be easily defined for CFGs. 
+%
+% In Section \ref{sec:semantics}, we will show that this is semantically
+% equivalent to regular capture-avoiding substitution, due to the fact that all
+% terms in a substitution are central \textit{and} that only affine computations
+% can be used in a term substituting for an affine variable, and only relevant
+% computationbs can be used in a term substituting for a relevant variable.
 
@@
 \end{definition}
+
 Note that the set of symmetric premonoidal subcategories, monoidal
 subcategories, and central subcategories of a category \(\mc{C}\) are all closed
-under intersections and therefore, in particular, form a complete
-meet-semilattice with shared bottom element \(\ms{Skeleton}(\mc{C})\), where
-\(\ms{Skeleton}(C)\) is the smallest possible symmetric premonodial subcategory.
-The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the
-largest central subcategory is the center of \(\mc{C}\), \(Z(\mb{C})\); the
-existence of a top-element means that in these cases we may induce the structure
-of a complete lattice in the usual manner.
-
+under intersections.
+%
+% and therefore, in particular, form a complete meet-semilattice with shared
+% bottom element \(\ms{Skeleton}(\mc{C})\), where \(\ms{Skeleton}(C)\) is the
+% smallest possible symmetric premonodial subcategory. The largest symmetric
+% premonoidal subcategory is of course \(\mc{C}\), while the largest central
+% subcategory is the center of \(\mc{C}\), \(Z(\mb{C})\).
+%
+% ; the existence of a top-element means that in these cases we may induce the
+% structure of a complete lattice in the usual manner.
+%
 We can now define a \textit{substructural premonoidal category}, which is simply
@@
     \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects
-    closed under tensor products, such that
+    closed under tensor products, such that \(\mb{1} \in \mc{A}\) and
     \begin{itemize}
-      \item \(\mb{1} \in \mc{A}\)
       \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism
@@
     \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects
-    closed under tensor products, such that
+    closed under tensor products, such that \(\mb{1} \in \mc{R}\) and
     \begin{itemize}
-      \item \(\mb{1} \in \mc{R}\)
       \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism

### 2023-11-17 pldi24/paper.tex
Tap?

@@
 \textit{not} in SSA form, since the variables \(m, n\) and \(i\) are each
-modified through subsequent assignments (\(t\) is recomputed on every iteration
-of the loop, but since it only has one definition, it \textit{is} in SSA form).
+modified through subsequent assignments .
+% (\(t\) is recomputed on every iteration of the loop, but since it only has one
+% definition, it \textit{is} in SSA form).
 To convert this program to SSA form, we would traditionally use
@@
 if-statement can be viewed as syntactic sugar for a conditional branch to one of
-two anonymous basic blocks, while nested scopes simply \textit{restrict} where
-we can branch to. 
-% In particular, in Section \ref{ssec:ssa-norm}, we will formalize
-% interconversion with traditional SSA and show how it is semantically sound to
-% simply merge all nested scopes. The point of the slightly modified syntax is
-% simply to make some of our metatheoretic results more practical to state.
-
-The formal grammar for our syntax is given in Figure \ref{fig:ssa-grammar} in
-the appendix, while the typing judgements themselves are given in
-\ref{fig:ssa-judgements}. We begin by giving a grammar for \textit{types} \(A,
-B, C\), parametrized by a set of \textit{base types} \(X\); all syntactically
-well-formed types are also semantically valid. We postulate the existence of a
-\textit{unit type} \(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all
-tensor products \(A \otimes B\); we write \(\otimes\) to underscore the fact
-that types are allowed to be substructural, assigning every type a linearity
-\(\ms{lin}(A)\) satisfying the constraints in Figure \ref{fig:ssa-structural}.
-In particular, the rules \rle{unit-lin} and \rle{bool-lin} state that both the
-unit and boolean types must be intuitionistic; the rule \rle{pair-lin} says that
-the linearity of a pair is simply the intersection of the linearity of its
-components. The rules \rle{nil-lin} and \rle{cons-lin} together say that the
-linearity of a context is the intersection of the linearity of its variables,
-where the linearity of a variable is the intersection of the linearity of its
-type and its quantity annotation.
-
-Our grammar proper splits program fragments into three main syntactic categories:
+two anonymous basic blocks, while nested scopes simply \textit{restrict}
+branching. Our syntax is composed of three main syntactic categories:
 \begin{itemize}
-  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
-  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the term \(a\)
-  is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or
-  \(p = 0\) otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\).
-  So the centrality \(p\) indicates whether it is safe to substitute this
-  term for a variable, and \(q\), plus the linearity of \(A\), control how
-  many times that variable can occur: if \(\taff \subseteq q\) then the variable
-  is called \textit{affine} or \textit{discardable} and it is permissible to
-  leave it unused, whereas if \(\trel \subseteq q\) it is called
-  \textit{relevant} or \textit{duplicable} and it is permissible to use multiple
-  times. An term which is central, relevant, and affine is called
-  \textit{pure}. The rules for this judgement are given in Figure
-  \ref{fig:ssa-term-typing}.
+  \item \textit{Terms} \(a, b, c, d, e\), which are typed via the judgement
+  \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the term \(a\) is of type
+  \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or \(p = 0\)
+  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). So the
+  centrality \(p\) indicates whether it is safe to substitute this term for a
+  variable, and \(q\), plus the linearity of \(A\), control how many times that
+  variable can occur: if \(\taff \subseteq q\) then the variable is called
+  \textit{affine} or \textit{discardable} and it is permissible to leave it
+  unused, whereas if \(\trel \subseteq q\) it is called \textit{relevant} or
+  \textit{duplicable} and it is permissible to use multiple times. A term which
+  is central, relevant, and affine is called \textit{pure}. The rules for this
+  judgement are given in Figure \ref{fig:ssa-term-typing}.
 
@@
 \end{itemize}
+The formal grammar for our syntax is given in Figure \ref{fig:ssa-grammar} in
+the appendix, while the typing judgements themselves are given in
+\ref{fig:ssa-judgements}.
+
+We postulate the existence of a set of \textit{base types} \(X\), as well as a
+\textit{unit type} \(\mb{1}\) and \textit{boolean type} \(\mb{2}\). We define
+the tensor product of types \(A \otimes B\); we write \(\otimes\) to underscore
+the fact that types are allowed to be substructural, assigning every type a
+linearity \(\ms{lin}(A)\) satisfying the constraints in Figure
+\ref{fig:ssa-structural}. In particular, the rules \rle{unit-lin} and
+\rle{bool-lin} state that both the unit and boolean types must be
+intuitionistic; the rule \rle{pair-lin} says that the linearity of a pair is
+simply the intersection of the linearity of its components. The rules
+\rle{nil-lin} and \rle{cons-lin} together say that the linearity of a context is
+the intersection of the linearity of its variables, where the linearity of a
+variable is the intersection of the linearity of its type and its quantity
+annotation.
+
 Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
@@
 We can now get to the typing judgement for expressions in Figure
-\ref{fig:ssa-term-typing}. The rule \rle{var} is essentially the standard rule
-for variables, with the requirement that, other than \(x\), all variables in
-\(\Gamma\) are affine. Note that the quantity \(q\) of the \textit{term}
-\(x\) is must satisfy \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this
-reflects that we're allowed to \textit{forget} capabilities, but not
-\textit{add} them. The \rle{app} rule is also mostly standard: given an
-instruction \(f\) which takes in an argument \(A\) and has a return type of
-\(B\) and a \textit{central} term \(\hasty{\Gamma}{1}{a}{A}{q}\), we
-conclude that \(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that
-\(f\) itself have centrality \(p\) and quantity \(q\), i.e. that \(f \in
-\mc{I}_p^q(A, B)\). The \rle{pair} rule says that we can type a pair \((a, b)\)
-in \(\Gamma\) if \(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\)
-and \(\Xi\) types \(b\); the quantity of a pair is simply the intersection of
-the quantity of its components (this is expressed in the rule by allowing any
-quantity which types \textit{both} components), and both components must be
-central. Finally, we have the rules \rle{let} and \rle{let2}, corresponding to
-let-bindings and destructuring let-bindings respectively. In both cases, we
-split the context \(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is
-used to type a central subterm \(a\), which is bound to variables \(x\) or
-destructured to variables \(x, y\) respectively, in a potentially non-central
-term \(e\) which is otherwise typed in \(\Delta\).
+\ref{fig:ssa-term-typing}. The rule \rle{var} states that \(x: A^q\) is
+well-typed in \(\Gamma\) if all other variables are affine (i.e., can be
+dropped) and \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this reflects
+that we're allowed to \textit{forget} capabilities, but not \textit{add} them.
+The \rle{app} rule states that, given an instruction \(f\) which takes in an
+argument \(A\) and has a return type of \(B\) and a \textit{central} term
+\(\hasty{\Gamma}{1}{a}{A}{q}\), we can conclude that
+\(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that \(f\) itself have
+centrality \(p\) and quantity \(q\), i.e. that \(f \in \mc{I}_p^q(A, B)\). The
+\rle{pair} rule says that we can type a pair \((a, b)\) in \(\Gamma\) if
+\(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\) and \(\Xi\) types
+\(b\); the quantity of a pair is simply the intersection of the quantity of its
+components (this is expressed in the rule by allowing any quantity which types
+\textit{both} components), and both components must be central. Finally, we have
+the rules \rle{let} and \rle{let2}, correspond to let-bindings and destructuring
+let-bindings respectively. In both cases, we split the context \(\Gamma\) into
+subcontext \(\Delta, \Xi\), one of which is used to type a central subterm
+\(a\), which is bound to variables \(x\) or destructured to variables \(x, y\)
+respectively, in a potentially non-central term \(e\) which is otherwise typed
+in \(\Delta\).
 

### 2023-11-17 pldi24/paper.tex
Tap

@@
 \begin{BVerbatim}[baseline=c]
-let x = rand();
+let x = nondet();
 return 5 
@@
 \begin{BVerbatim}[baseline=c]
-let x = rand();
+let x = nondet();
 let y = x;
@@
 \begin{BVerbatim}[baseline=c]
-let y = rand();
-let z = rand();
+let y = nondet();
+let z = nondet();
 return y == z
@@
 \end{equation}
-In this case, substituting for $x$ duplicates the call to \texttt{rand()}, and
+In this case, substituting for $x$ duplicates the call to \texttt{nondet()}, and
 can change this from a program which always returns true to one which can return
@@
 two anonymous basic blocks, while nested scopes simply \textit{restrict}
-branching. Our syntax is composed of three main syntactic categories:
+branching.
+
+We parametrize an \isotopessa dialect by a set of \textit{base types} \(X \in
+\mc{T}\) and, for each pair of types \(A, B\), a set of \textit{instructions}
+\(f \in \mc{I}^q_p\) with \textit{centrality} \(p \in \{0, 1\}\) and
+\textit{quantity} \(q \subseteq \tint\). As described above, a \textit{central}
+instruction is one for which execution order does not matter, while the quantity
+\(q\) indicates whether it is safe to delete an instruction if its result is
+unused (in which case the instruction is \textit{affine}, and \(\taff \subseteq
+q\)) or duplicate an instruction used multiple times (in which case the
+instruction is \textit{relevant}, and \(\trel \subseteq q\)). An instruction
+which is both affine and relevant is called \textit{pure}. In general, the set
+of types is generated by base types \(X\), the \textit{unit type} \(\mb{1}\),
+the \textit{boolean type} \(\mb{2}\), and tensor products of types \(A \otimes
+B\). We write \(\otimes\) to underscore the fact that types are allowed to be
+substructural: every type is assigned a \textit{linearity} \(\ms{lin}(A)
+\subseteq \tint\) satisfying the constraints in Figure \ref{fig:ssa-structural}.
+In particular, the rules \rle{unit-lin} and \rle{bool-lin} state that both the
+unit and boolean types must be intuitionistic; the rule \rle{pair-lin} says that
+the linearity of a pair is simply the intersection of the linearity of its
+components. 
+% The rules \rle{nil-lin} and \rle{cons-lin} together say that the
+% linearity of a context is the intersection of the linearity of its variables,
+% where the linearity of a variable is the intersection of the linearity of its
+% type and its quantity annotation.
+
+Our syntax is composed of three main syntactic categories:
 \begin{itemize}
@@
 \ref{fig:ssa-judgements}.
-
-We postulate the existence of a set of \textit{base types} \(X\), as well as a
-\textit{unit type} \(\mb{1}\) and \textit{boolean type} \(\mb{2}\). We define
-the tensor product of types \(A \otimes B\); we write \(\otimes\) to underscore
-the fact that types are allowed to be substructural, assigning every type a
-linearity \(\ms{lin}(A)\) satisfying the constraints in Figure
-\ref{fig:ssa-structural}. In particular, the rules \rle{unit-lin} and
-\rle{bool-lin} state that both the unit and boolean types must be
-intuitionistic; the rule \rle{pair-lin} says that the linearity of a pair is
-simply the intersection of the linearity of its components. The rules
-\rle{nil-lin} and \rle{cons-lin} together say that the linearity of a context is
-the intersection of the linearity of its variables, where the linearity of a
-variable is the intersection of the linearity of its type and its quantity
-annotation.
 
@@
     \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} \\
-    \prftree[r]{\rle{nil-lin}}{\tlin{\cdot}{q}} \qquad
-    \prftree[r]{\rle{cons-lin}}{\ltlin{A}{r}{q}}{\tlin{\Gamma}{q}}
-      {\tlin{\Gamma, \thyp{x}{A}{r}}{q}} \\
+    % \prftree[r]{\rle{nil-lin}}{\tlin{\cdot}{q}} \qquad
+    % \prftree[r]{\rle{cons-lin}}{\ltlin{A}{r}{q}}{\tlin{\Gamma}{q}}
+    %   {\tlin{\Gamma, \thyp{x}{A}{r}}{q}} \\
     \prftree[r]{\rle{split-nil}}{\csplits{\cdot}{\cdot}{\cdot}} \qquad
@@
 We can now state some of the basic metatheoretic properties of \isotopessa. We
-begin with weakening, which we state in our setting as follows:
+begin with weakening, which states that we can add unused \textit{affine} variables to a context,
+\textit{reduce} the quantity of an term's type, add arbitrary unused labels, and
+remove \textit{affine} variables from existing labels and still keep our typing
+judgements valid, i.e.
 \begin{lemma}[Weakening] \
@@
 \end{lemma}
-That is, we can add unused \textit{affine} variables to a context,
-\textit{reduce} the quantity of an term's type, add arbitrary unused labels, and
-remove \textit{affine} variables from existing labels and still keep our typing
-judgements valid. This can be proved by a relatively straightforward mutual
-induction.
 
@@
   \end{gather*}
-  \caption{ Typing rules for \isotopessa substitutions. We generally assume that
-    for all but finitely many variables \(x\) and labels \(\lbl{\ell}\),
-    \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) = \lbrb{\ell}{x}\). }
+  \caption{ Typing rules for \isotopessa substitutions. The linearity of a
+    context is defined recursively to be the intersection of the linearities and
+    quantities of its variables: \(\ms{lin}(\cdot) = \tint\),
+    \(\ms{lin}(\Gamma, \thyp{x}{A}{q}) = \ms{lin}(\Gamma) \cap \ms{lin}(A) \cap
+    q\). 
+    % We generally assume that for all but finitely many variables \(x\) and
+    % labels \(\lbl{\ell}\), \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) =
+    % \lbrb{\ell}{x}\). 
+  } 
   \Description{Typing rules for isotope-SSA substitutions}
@@
   \end{itemize}
-\end{lemma}
-
-We can also, using the rules in Figure \ref{fig:ssa-subst-ops}, define
-capture-avoiding label-substitution for blocks \([\mc{L}]t\) and CFGs
-\([\mc{L}]L\), allowing us to state the following lemma:
-\begin{lemma}[Label Substitution] 
-  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
-  \begin{itemize}
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\),
-    \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
-    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\),
-    \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
-  \end{itemize}
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), we have that for all
+  \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
 \end{lemma}
@@
 since only the left-hand side contains the pair \((b_0, b_1)\), this
-corresponding to the fact that the program \(\ms{let}\;x = \ms{rand}(); (x, x)\) is
-not semantically equal to \((\ms{rand}(), \ms{rand}())\).
+corresponding to the fact that the program \(\ms{let}\;x = \ms{nondet}(); (x, x)\) is
+not semantically equal to \((\ms{nondet}(), \ms{nondet}())\).
 
@@
 instructions \(\ms{print} \in \mc{I}^\varnothing_0(\ms{Byte}^*, \mb{1})\) and
-\(\ms{rand} \in \mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard
+\(\ms{nondet} \in \mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard
 semantics \(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), b)\}\),
-\(\dnt{\ms{rand}} = \lambda (). \{\iota_0 (a, []) \mid a \in A\}\).
+\(\dnt{\ms{nondet}} = \lambda (). \{\iota_0 (a, []) \mid a \in A\}\).
 
@@
   \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Byte}^*, \ms{emp}), (\mb{1},
-  \ms{emp}))\) and \(\ms{rand}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for
+  \ms{emp}))\) and \(\ms{nondet}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for
   \(A\) nonempty

### 2023-11-17 pldi24/paper.tex
Tap

@@
 \begin{definition}[Susbtructural Effectful Category]
-  A \textbf{substructural effectful category} \(\mc{C}\) consists of
-  \begin{itemize}
-    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms
-    in this category \textbf{central} or \textbf{linear}
-    \item A substructural premonoidal category \(\mc{C}_0\) sharing the same
-    affine and relevant objects as \(\mc{C}_1\)
-    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing
-    \to \mc{C}_0^\varnothing\) preserving all substructural premonoidal
-    structure, i.e., preserves all symmetries, associators, unitors, drops, and
-    splits and sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\). In particular,
-    restricting \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to
-    \mc{C}_0^q\) into effectful categories.
-  \end{itemize}
-  We define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\)
-  and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply
-  be the identity functor on \(\mc{C}_1^q\).
-
-  Morphisms in \(\mc{C}_p^{\ms{r}}\) are called \textbf{copyable}, morphisms
-  in \(\mc{C}_p^{\ms{a}}\) are called \textbf{discardable}, while morphisms
-  in \(\mc{C}_1^{\ms{a}}, \mc{C}_1^{\ms{r}}, \mc{C}_1^{\{\ms{a},
-  \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and
+  An effectful category \(\mc{C}_0 \to \mc{C}_1\) is said to be
+  \textbf{substructural} if \(\mc{C}_0, \mc{C}_1\) are substructural and
+  \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0^\varnothing\) preserves all
+  symmetries, associators, unitors, drops, and splits and sends \(\mc{C}_1^q\)
+  to \(\mc{C}_0^q\). In particular, restricting \(\upg{\cdot}{}\) appropriately
+  makes all \(\mc{C}_1^q \to \mc{C}_0^q\) into effectful categories. Morphisms
+  in \(\mc{C}_p^{\ms{r}}\) are called \textbf{copyable}, morphisms in
+  \(\mc{C}_p^{\ms{a}}\) are called \textbf{discardable}, while morphisms in
+  \(\mc{C}_1^{\ms{a}}, \mc{C}_1^{\ms{r}}, \mc{C}_1^{\{\ms{a}, \ms{r}\}}\)
+  specifically are called \textbf{affine}, \textbf{relevant}, and
   \textbf{intuitionistic} or \textbf{pure} respectively.
@@
 "judgement"; in general, it can be proven that the semantics of any two such
-derivations are equal.
-% \begin{lemma}[Coherence] \
-%   \begin{itemize}
-%     \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-%     \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-%     \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
-%     \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
-%     \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
-%   \end{itemize}
-% \end{lemma}
+derivations are equal. We also define the syntax sugar \(\upg{f}{p}\), where
+\(\upg{f}{0}\) is \(\upg{f}{}\) for morphisms in \(\mc{C}_1\) and simply \(f\)
+for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) is the identity functor on
+\(\mc{C}_1\).
+
 Our treatment of \rle{var} is quite standard: the term \(x\) types in \(\Gamma\)

### 2023-11-17 pldi24/paper.tex
Tap

@@
   stream \(\ell\).
-\end{definition}
   Some basic examples include:
@@
     % \cdot_i l_i)_i\).
-  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream
-  action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\),
-  the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell
-  \cdot s\) prepends the \(\ell\) to \(s\).
+    \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a
+    stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq
+    \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\),
+    where \(\ell \cdot s\) prepends the \(\ell\) to \(s\).
     % \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a
@@
   \end{enumerate}
-We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted
-\(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of
-those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\).
-For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq
-\omega}) = \{[\,]\}\). Generally, \(\mc{SZ}(M, I)\) always contains
-the unit of \(M\). We can now define the \textit{trace monad} of a stream action as follows:
+\end{definition}
+% We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted
+% \(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of
+% those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\).
+% For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq
+% \omega}) = \{[\,]\}\). Generally, \(\mc{SZ}(M, I)\) always contains
+% the unit of \(M\). 
+Using stream actions to determine what it means to perform a monoidal action
+"infinitely many times," we can now define the \textit{trace monad} of a stream
+action as follows:
 \begin{definition}[Trace Monad]
@@
   (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
-  from \(\mb{T}\).
-\end{definition}
-Some important examples of submonads include:
+  from \(\mb{T}\). Some important examples of submonads include:
   \begin{enumerate}
@@
     submonad and another submonad is relevant/affine/commutative respectively.
-  Since every commutative submonad induces a central subcategory of the Kleisli
-  category of the parent monad, it follows that a triple of a relevant, affine,
-  and commutative submonad make the Kleisli category of a monad \(\mb{T}\) into
-  a substructural effectful category with all objects both affine and relevant.
   \end{enumerate}
+\end{definition} 
 We can use the trace monad described above to derive some interesting examples
@@
 \end{enumerate}
-In particular, \ref{item:central-submonad}, \ref{item:relevant-submonad},
+Since every commutative submonad induces a central subcategory of the Kleisli
+category of the parent monad, it follows that a triple of a relevant, affine,
+and commutative submonad make the Kleisli category of a monad \(\mb{T}\) into a
+substructural effectful category with all objects both affine and relevant. In
+particular, \ref{item:central-submonad}, \ref{item:relevant-submonad},
 \ref{item:affine-submonad} equip the Kleisli category of
@@
 
-Another interesting use of the trace monad to build a sophisticated \isotopessa model is to build a model of TSO-style weak memory based on \citet{sparky}.
-For simplicity, we will consider an infinite set of named, fixed locations \(x, y, z \in \ms{Loc}\) which are subject to concurrent modification by all processors via TSO reads, writes, and fences.
-
-We begin with the following definitions:
+Another interesting use of the trace monad to build a sophisticated \isotopessa
+model is to build a model of TSO-style weak memory based on \citet{sparky}. For
+simplicity, we will consider an infinite set of named locations \(x, y, z \in
+\ms{Loc}\) which are subject to concurrent modification by all threads via TSO
+reads, writes, and fences. We begin with the following definitions:
 \begin{definition}[Pomset] 

### 2023-11-17 pldi24/paper.tex
Tap

@@
 informally, since the "intended model" of the language is a simple sequential
-machine operating over a flat array of bytes. Unfortunately,
-over time both the intended model and the actual hardware have drifted away from
-this simple mental model. Compiler frameworks have to model concurrent accesses
-to memory in order to compile multithreaded code, and furthermore the IR's
-memory model has to be sound with respect to the actual weak memory models
-supported by hardware.
+machine operating over a flat array of bytes. Unfortunately, over time both the
+intended model and the actual hardware have drifted away from this simple mental
+model. Compiler frameworks have to model concurrent accesses to memory in order
+to compile multithreaded code, and furthermore the IR's memory model has to be
+sound with respect to the actual weak memory models supported by hardware.
 
@@
 \item Our abstract semantics justifies a variety of powerful optimisation
-techniques, inlining, loop fusion, hoisting, and strength reduction. 
+techniques, including inlining, loop fusion, hoisting, and strength reduction. 
 \item We show how Elgot monads give rise to instances of our categorical
@@
 
-We parametrize an \isotopessa dialect by a set of \textit{base types} \(X \in
-\mc{T}\) and, for each pair of types \(A, B\), a set of \textit{instructions}
-\(f \in \mc{I}^q_p\) with \textit{centrality} \(p \in \{0, 1\}\) and
-\textit{quantity} \(q \subseteq \tint\). As described above, a \textit{central}
-instruction is one for which execution order does not matter, while the quantity
-\(q\) indicates whether it is safe to delete an instruction if its result is
-unused (in which case the instruction is \textit{affine}, and \(\taff \subseteq
-q\)) or duplicate an instruction used multiple times (in which case the
-instruction is \textit{relevant}, and \(\trel \subseteq q\)). An instruction
-which is both affine and relevant is called \textit{pure}. In general, the set
-of types is generated by base types \(X\), the \textit{unit type} \(\mb{1}\),
-the \textit{boolean type} \(\mb{2}\), and tensor products of types \(A \otimes
-B\). We write \(\otimes\) to underscore the fact that types are allowed to be
-substructural: every type is assigned a \textit{linearity} \(\ms{lin}(A)
-\subseteq \tint\) satisfying the constraints in Figure \ref{fig:ssa-structural}.
-In particular, the rules \rle{unit-lin} and \rle{bool-lin} state that both the
-unit and boolean types must be intuitionistic; the rule \rle{pair-lin} says that
-the linearity of a pair is simply the intersection of the linearity of its
-components. 
+We parametrize our syntax by a set of \textit{base types} \(X \in \mc{T}\) and,
+for each pair of types \(A, B\), a set of \textit{instructions} \(f \in
+\mc{I}^q_p\) with \textit{centrality} \(p \in \{0, 1\}\) and \textit{quantity}
+\(q \subseteq \tint\). As described above, a \textit{central} instruction is one
+for which execution order does not matter, while the quantity \(q\) indicates
+whether it is safe to delete an instruction if its result is unused (in which
+case the instruction is \textit{affine}, and \(\taff \subseteq q\)) or duplicate
+an instruction used multiple times (in which case the instruction is
+\textit{relevant}, and \(\trel \subseteq q\)). An instruction which is both
+affine and relevant is called \textit{pure}. The set of types is generated by
+base types \(X\), the \textit{unit type} \(\mb{1}\), the \textit{boolean type}
+\(\mb{2}\), and tensor products of types \(A \otimes B\). We write \(\otimes\)
+to underscore the fact that types are allowed to be substructural: similarly to
+instructions, every type is assigned a \textit{linearity} \(\ms{lin}(A)
+\subseteq \tint\) satisfying the constraints in Figure \ref{fig:ssa-structural},
+except that a type which is both affine and relevant is called
+\textit{intuitionistic}. In particular, the rules \rle{unit-lin} and
+\rle{bool-lin} state that both the unit and boolean types must be
+intuitionistic; the rule \rle{pair-lin} says that the linearity of a pair is
+simply the intersection of the linearity of its components. 
 % The rules \rle{nil-lin} and \rle{cons-lin} together say that the
@@
     \qquad
-    \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} \\
+    \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} 
+    \\
     % \prftree[r]{\rle{nil-lin}}{\tlin{\cdot}{q}} \qquad
@@
 relevant, neither or both, enabling us to reason about substructural type
-systems categorically.
-
-To do this, we begin by introducing the notion of a \textit{symmetric
+systems categorically. To do this, we begin by introducing the notion of a \textit{symmetric
 premonoidal subcategory}, as follows:
@@
     \item \(\mc{C}\) is a symmetric premonoidal category equipped with wide
-    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\).
-    We define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)
+    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\). We
+    define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell
+    \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)
     \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects
@@
       \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism
-      \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
-      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
-      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\ms{a}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
+      \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\) satisfying
+      \item \(\forall A, B \in \mc{A}. \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\), \(\forall f: \mc{C}^{\ms{a}}(A, B). f;\ms{drop}(B) = \ms{drop}(A)\)
     \end{itemize}
@@
       \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism
-      \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
+      \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\) satisfying
       \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
@@
 \end{itemize}
-
 We begin by giving a semantics for types and contexts in terms of objects in
@@
 \begin{figure}
-  \begin{gather*}
-    \boxed{\dnt{A}}: |\mc{C}| \\
+  \begin{align*}
+    \boxed{\dnt{A}: |\mc{C}|} & \qquad  
     \dnt{X} = \ms{base}(X) 
@@
     \qquad \dnt{A \otimes B} = \dnt{A} \otimes \dnt{B} \\
-    \boxed{\dnt{\Gamma}}: |\mc{C}| \\
+    \boxed{\dnt{\Gamma}: |\mc{C}|} & \qquad 
     \dnt{\cdot} = \mb{1} \qquad 
-    \dnt{\Gamma, \thyp{x}{A}{q}} = \dnt{\Gamma} \otimes \dnt{A} \\
-    \boxed{\dnt{\ms{L}}}: |\mc{C}| \\
+    \dnt{\Gamma, \thyp{x}{A}{q}} = \dnt{\Gamma} \otimes \dnt{A} 
+    \\
+    \boxed{\dnt{\ms{L}}: |\mc{C}|} & \qquad 
     \dnt{\cdot} = \mb{0} \qquad
@@
       \dnt{\ms{L}} + \dnt{\Gamma} \otimes \dnt{A}
-  \end{gather*}
+  \end{align*}
   \caption{Semantics for \isotopessa types and contexts}
@@
   \begin{gather*}
-    \boxed{\dnt{\tlin{A}{\taff}: \mc{C}_1^\tint(\dnt{A}, \mb{1})}}
-      \\
-      \dnt{\tlin{A}{\taff}} = \ms{drop}(\dnt{A})
-      % \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
-      % \dnt{\tlin{A \otimes B}{\taff}} 
-      %   = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
-      %   ; \lambda 
-      \\
-    \boxed{\dnt{\tlin{A}{\trel}}
-      : \mc{C}_1^\tint(\dnt{A}, \dnt{A} \otimes \dnt{A})}
-      \\
-      \dnt{\tlin{A}{\trel}} = \ms{split}(\dnt{A})
-      % \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
-      % \dnt{\tlin{A \otimes B}{\trel}} = 
-      %   \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
-      %   ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
-      \\
+    % \boxed{\dnt{\tlin{A}{\taff}: \mc{C}_1^\tint(\dnt{A}, \mb{1})}}
+    %   \\
+    %   \dnt{\tlin{A}{\taff}} = \ms{drop}(\dnt{A})
+    %   % \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
+    %   % \dnt{\tlin{A \otimes B}{\taff}} 
+    %   %   = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
+    %   %   ; \lambda 
+    %   \\
+    % \boxed{\dnt{\tlin{A}{\trel}}
+    %   : \mc{C}_1^\tint(\dnt{A}, \dnt{A} \otimes \dnt{A})}
+    %   \\
+    %   \dnt{\tlin{A}{\trel}} = \ms{split}(\dnt{A})
+    %   % \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
+    %   % \dnt{\tlin{A \otimes B}{\trel}} = 
+    %   %   \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
+    %   %   ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
+    %   \\
     \boxed{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
@@
       {\Xi, \thyp{x}{A}{q}}} \
-      = \dnt{\Gamma} \otimes \dnt{\tlin{A}{\trel}};
+      = \dnt{\Gamma} \otimes \ms{split}(\dnt{A});
         \alpha;
@@
       = \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-        \otimes \dnt{\tlin{A}{\taff}}
+        \otimes \ms{drop}(\dnt{A})
       ; \rho 
@@
 and other optimisations which change the structure of the control-flow graph.
-
 Note that we have not yet proved anything about the dataflow analyses used to
@@
 infinite) nondeterministic trace of monoidal "events". These will later come in
-handy for reasoning about output behaviour and weak memory.
-
-We begin with some definitions:
+handy for reasoning about output behaviour and weak memory. We start with some
+definitions:
 \begin{definition}[Stream Action]
@@
     \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\)
-    forms a stream action on the one element set \(\mathbf{1}\).
+    forms a stream action on the unit set.
     % \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action

### 2023-11-17 pldi24/paper.tex
Tap

@@
 for each pair of types \(A, B\), a set of \textit{instructions} \(f \in
-\mc{I}^q_p\) with \textit{centrality} \(p \in \{0, 1\}\) and \textit{quantity}
-\(q \subseteq \tint\). As described above, a \textit{central} instruction is one
-for which execution order does not matter, while the quantity \(q\) indicates
-whether it is safe to delete an instruction if its result is unused (in which
-case the instruction is \textit{affine}, and \(\taff \subseteq q\)) or duplicate
-an instruction used multiple times (in which case the instruction is
-\textit{relevant}, and \(\trel \subseteq q\)). An instruction which is both
-affine and relevant is called \textit{pure}. The set of types is generated by
-base types \(X\), the \textit{unit type} \(\mb{1}\), the \textit{boolean type}
-\(\mb{2}\), and tensor products of types \(A \otimes B\). We write \(\otimes\)
-to underscore the fact that types are allowed to be substructural: similarly to
-instructions, every type is assigned a \textit{linearity} \(\ms{lin}(A)
-\subseteq \tint\) satisfying the constraints in Figure \ref{fig:ssa-structural},
-except that a type which is both affine and relevant is called
-\textit{intuitionistic}. In particular, the rules \rle{unit-lin} and
-\rle{bool-lin} state that both the unit and boolean types must be
-intuitionistic; the rule \rle{pair-lin} says that the linearity of a pair is
-simply the intersection of the linearity of its components. 
-% The rules \rle{nil-lin} and \rle{cons-lin} together say that the
-% linearity of a context is the intersection of the linearity of its variables,
-% where the linearity of a variable is the intersection of the linearity of its
-% type and its quantity annotation.
-
-Our syntax is composed of three main syntactic categories:
+\mc{I}^q_p(A, B)\) with \textit{centrality} \(p = 1\) if \textit{central} or \(p
+= 0\) otherwise and \textit{quantity} \(q \subseteq \tint\). As discussed above,
+an instruction is considered \textit{central} if execution order does not
+matter, while the quantity \(q\) indicates whether it is safe to delete an
+instruction if its result is unused (in which case the instruction is
+\textit{affine}, and \(\taff \subseteq q\)) or duplicate an instruction used
+multiple times (in which case the instruction is \textit{relevant}, and \(\trel
+\subseteq q\)). An instruction which is, central, relevant, and affine is called
+\textit{pure}. Note that we assume that \(p \leq p' \and q \subseteq r \implies
+\mc{I}_p^q(A, B) \supseteq \mc{I}_{p'}^{r}(A, B)\). The set of types is
+generated by base types \(X\), the \textit{unit type} \(\mb{1}\), the
+\textit{boolean type} \(\mb{2}\), and tensor products of types \(A \otimes B\).
+We write \(\otimes\) (rather than \(\times\)) to underscore the fact that types
+are allowed to be substructural: given a linearity \(\ms{lin}(X) \subseteq
+\tint\), we can assign every type a \textit{linearity} \(\ms{lin}(A) \subseteq
+\tint\) by taking \(\ms{lin}(A \otimes B) = \ms{lin}(A) \cap \ms{lin}(B)\) and
+\(\ms{lin}(\mb{1}) = \ms{lin}(\mb{2}) = \tint\). We say a type with \(\taff
+\subseteq \ms{lin}(A)\) is \textit{affine}, while a type with \(\trel \subseteq
+\ms{lin}(A)\) is \textit{relevant}; a type which is both affine and relevant is
+called \textit{intuitionistic}. Our syntax is composed of three main syntactic
+categories:
 \begin{itemize}
@@
   \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the term \(a\) is of type
-  \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or \(p = 0\)
-  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). So the
-  centrality \(p\) indicates whether it is safe to substitute this term for a
-  variable, and \(q\), plus the linearity of \(A\), control how many times that
-  variable can occur: if \(\taff \subseteq q\) then the variable is called
-  \textit{affine} or \textit{discardable} and it is permissible to leave it
-  unused, whereas if \(\trel \subseteq q\) it is called \textit{relevant} or
-  \textit{duplicable} and it is permissible to use multiple times. A term which
-  is central, relevant, and affine is called \textit{pure}. The rules for this
-  judgement are given in Figure \ref{fig:ssa-term-typing}.
+  \(A\) with \textit{centrality} \(p\) and \textit{quantity} \(q \subseteq
+  \{\ms{a}, \ms{r}\}\). Here, the centrality \(p\) indicates whether it is safe
+  to substitute this term for a variable, and \(q\), plus the linearity of
+  \(A\), control how many times that variable can occur: if \(\taff \subseteq
+  q\) then the variable is called \textit{affine} or \textit{discardable} and it
+  is permissible to leave it unused, whereas if \(\trel \subseteq q\) it is
+  called \textit{relevant} or \textit{duplicable} and it is permissible to use
+  multiple times. As for instructions, a term which is central, relevant, and
+  affine is called \textit{pure}. The rules for this judgement are given in
+  Figure \ref{fig:ssa-term-typing}.
 
@@
 Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
-type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). A variable
-\(x: A^q\) is called \textit{affine} if its type \(A\) is affine (i.e.
-\(\tlin{A}{\taff}\)) \textit{and} \(\taff \subseteq q\), in which case it can be
-freely ignored. Similarly, \(x\) is \textit{relevant} if its type \(A\) relevant
-(i.e. \(\tlin{A}{\trel}\)) \textit{and} \(\trel \subseteq q\), in which case it
-can be freely ignored. This is formalized in the structural judgement
+type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\) (if \(q\) is
+omitted, we assume \(q = \tint\)). The \textit{linearity} of a variable
+\(\ms{lin}(\thyp{x}{A}{q})\) is defined to be the intersection of the linearity
+of its type \(\ms{lin}(A)\) and its quantity \(q\). We say a variable \(x: A^q\)
+is \textit{affine} if \(\tlin{\thyp{x}{A}{q}}{\taff}\), in which case it can be
+freely ignored. Similarly, \(x\) is \textit{relevant} if
+\(\tlin{\thyp{x}{A}{q}}{\trel}\), in which case it can be used multiple times.
+This is formalized in the structural judgement
 \(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
@@
 typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
-\rle{split-nil} provides a base case by saying that the empty context can be
-split into two empty contexts, while the rules \rle{split-left} and
-\rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively;
-both these rules also allow us to reduce the quantity \(q\) to a quantity \(r
-\subseteq q\), "forgetting" some of our capabilities. The rule \rle{split-dup}
-allows the duplication of a variable which has a relevant type \(A\)
-\textit{and} a relevant quantity \(q\), with the option to weaken the quantity
-to \textit{different} quantities \(r\) and \(s\) on the left-hand side and
-right-hand sides respectively. Finally, the rule \rle{split-aff} simply allows
-us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.
-
-\textit{label-sets} \(\ms{L}\) are defined to consist of a list of
-\textit{labels} \(\lbl{\ell}\) annotated with a context \(\Gamma\) and an
+\rle{split-nil} states that the empty context can be split into two empty
+contexts, while the rules \rle{split-left} and \rle{split-right} send \(x: A^q\)
+to the left or right subcontext repsectively; both these rules also allow us to
+reduce the quantity \(q\) to a quantity \(r \subseteq q\), "forgetting" some of
+our capabilities. The rule \rle{split-dup} allows the duplication of a relevant
+variable, with the option to weaken the quantity to \textit{different}
+quantities \(r\) and \(s\) on the left and right respectively. Finally, the rule
+\rle{split-aff} simply allows us to drop an affine variable.
+
+On the other hand, \textit{label-sets} \(\ms{L}\) are defined to consist of a
+list of \textit{labels} \(\lbl{\ell}\) annotated with a context \(\Gamma\)
+(consisting of variables which must be live on entry to \(\lbl{\ell}\)) and an
 argument type \(A\). The judgement \(\lwk{\ms{L}}{\ms{K}}\) states that
-\(\ms{L}\) is a subcontext of \(\ms{K}\), i.e., that every program which targets
-\(\ms{L}\) (i.e., finishes by either nontermination or branching to one of the
-labels \(\llhyp{\ell}{\Gamma}{A} \in \ms{L}\) with live variables \(\Gamma\) and
-an argument of type \(A\)) can also be viewed as targeting \(\ms{K}\). The
-conditions for this are formalized in Figure \ref{fig:ssa-structural} as
-follows: the rule \rle{join-nil} provides a base case stating that the empty
-label weakens into itself. The rule \rle{join-cons} says that if \(\ms{L}\)
-weakens into \(\ms{K}\) and \(\Delta\) weakens into \(\Gamma\), then
-\(\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}\),
-allowing us to "forget" any extra variables dropped by \(\cwk{\Gamma}{\Delta}\).
-Finally, the rule \rle{join-zero} simply says that if \(\lwk{\ms{L}}{\ms{K}}\),
-then \(\ms{L}\) also weakens to any extension of \(\ms{K}\). Note that this
-judgement is "backwards," in that the "larger" context is on the \textit{right};
-this corresponds to the placement of the label context on the right-hand side of
-the typing judgement for blocks, and is by analogy to our semantics for blocks
-and CFGs in Section \ref{sec:semantics}.
+\(\ms{L}\) is a subcontext of \(\ms{K}\), i.e., has more labels with fewer live
+variables, and hence that every program which targets \(\ms{L}\) can also be
+viewed as targeting \(\ms{K}\). The conditions for this are formalized in Figure
+\ref{fig:ssa-structural} as follows: the rule \rle{join-nil} states that the
+empty label-set is a subcontext of itself. The rule \rle{join-cons} says that if
+\(\ms{L}\) is a subcontext of \(\ms{K}\) and \(\Delta\) weakens into \(\Gamma\),
+then \(\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}\),
+allowing us to "forget" any extra (affine) variables dropped by
+\(\cwk{\Gamma}{\Delta}\). Finally, the rule \rle{join-zero} simply says that if
+\(\lwk{\ms{L}}{\ms{K}}\), then \(\ms{L}\) is also a subcontext of any extension
+of \(\ms{K}\). 
+% Note that this judgement is "backwards," in that the "larger" context is on
+% the \textit{right}; this corresponds to the placement of the label context on
+% the right-hand side of the typing judgement for blocks, and is by analogy to
+% our semantics for blocks and CFGs in Section \ref{sec:semantics}.
 
@@
 We can now get to the typing judgement for expressions in Figure
-\ref{fig:ssa-term-typing}. The rule \rle{var} states that \(x: A^q\) is
-well-typed in \(\Gamma\) if all other variables are affine (i.e., can be
-dropped) and \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this reflects
-that we're allowed to \textit{forget} capabilities, but not \textit{add} them.
-The \rle{app} rule states that, given an instruction \(f\) which takes in an
-argument \(A\) and has a return type of \(B\) and a \textit{central} term
-\(\hasty{\Gamma}{1}{a}{A}{q}\), we can conclude that
-\(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that \(f\) itself have
-centrality \(p\) and quantity \(q\), i.e. that \(f \in \mc{I}_p^q(A, B)\). The
-\rle{pair} rule says that we can type a pair \((a, b)\) in \(\Gamma\) if
+\ref{fig:ssa-term-typing}. \rle{var} states that \(\thyp{x}{A}{q}\) is
+well-typed in \(\Gamma\) if all other variables are affine and \(q \subseteq
+q'\), where \(\thyp{x}{A}{q'} \in \Gamma\); allowing us to \textit{forget}
+capabilities, but not \textit{add} them. rle{app} states that, given an
+instruction \(f \in \mc{I}_p^q(A, B)\) and a \textit{central} term
+\(\hasty{\Gamma}{1}{a}{A}{q}\), we can type \(\hasty{\Gamma}{p}{f\;a}{B}{q}\).
+The \rle{pair} rule says that we can type a pair \((a, b)\) in \(\Gamma\) if
 \(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\) and \(\Xi\) types
 \(b\); the quantity of a pair is simply the intersection of the quantity of its
-components (this is expressed in the rule by allowing any quantity which types
-\textit{both} components), and both components must be central. Finally, we have
-the rules \rle{let} and \rle{let2}, correspond to let-bindings and destructuring
-let-bindings respectively. In both cases, we split the context \(\Gamma\) into
-subcontext \(\Delta, \Xi\), one of which is used to type a central subterm
-\(a\), which is bound to variables \(x\) or destructured to variables \(x, y\)
-respectively, in a potentially non-central term \(e\) which is otherwise typed
-in \(\Delta\).
+components (expressed by allowing any quantity which types \textit{both}
+components), which must both be central. Finally, \rle{let} and \rle{let2},
+allow us to type (destructuring) let-bindings. In both cases, we split the
+context \(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is used to type
+a central term \(a\), which is bound to fresh variables appearing in a
+potentially non-central term \(e\) which may also depend on variables in
+\(\Delta\).
 
@@
 The rules for block typing are given in Figure \ref{fig:ssa-block-typing}.
-\rle{br} says that, given a pure term \(a: A\) (that is, central, relevant, and
-affine), we can type a block \(\lbrb{\ell}{a}\) which branches to \(\ell\) with
-argument \(a\), where \(\ms{L}\) contains the context \(\ell\) with an argument
-type of \(A\) and an appropriate set of unused variables. The rule \rle{ite}
-typechecks if-then-else statements by stating that, if \(\Gamma\) splits into
-\(\Delta\) and \(\Xi\), where \(\Xi\) typechecks a pure, intuitionistic
-discriminant \(e: \mb{2}\) and \(\Delta\) typechecks both \(s, t\) to branch
-into \(\ms{L}\), then the whole if-statement \(\ite{e}{s}{t}\) also typechecks
-as branching into \(\ms{L}\) in constext \(\Gamma\). \rle{let-blk} and
-\rle{let2-blk} are quite similar to their term analogues, except that the
-bound term \(e\) is allowed to be impure (otherwise, programs could only
-make use of pure functions!), and the remainder of the context plus the bound
-variable is used to type a block rather than a term. Finally, probably the most
-interesting rule is \rle{where}, which says that, given a block \(t\) which
-branches to \(\ms{L}\) in context \(\Gamma\), and labels \(L\) mapping
-\(\ms{L}\) to \(\ms{K}\), \(\ewhere{t}{L}\) branches to \(\ms{K}\). Here, \(L\)
-mapping \(\ms{L}\) to \(\ms{K}\) is defined using the typing rules \rle{nil-br},
-which says that an empty control-flow graph simply forwards the input labels
-directly to the output labels, and \rle{cons-br}, which says that if \(L\) maps
-\(\ms{L}\) to either \(\ms{K}\) or the label \(\lbl{\ell}[\Gamma](A)\), and
-\(t\) branches to \(\ms{L}\) in the context \(\Gamma, \thyp{x}{A}{}\), then
-\(L\) maps \(\ms{L}\) to \(\ms{K}\), having (potentially recursively) defined
-\(\lbl{\ell}\).
+\rle{br} says that if \(\lwk{\lhyp{\ell}{\Delta}{A}}{\ms{L}}\), \(\Gamma\)
+splits into \(\Delta\) and \(\Xi\), and \(a\) is a pure term which typechecks in
+\(\Xi\), then \(\lbrb{\ell}{a}\) targets \(\ms{L}\) given \(\Gamma\) is live on
+entry. \rle{ite} states that, if \(\Gamma\) splits into \(\Delta\) and \(\Xi\),
+\(e\) typechecks as a pure, boolean term in \(\Xi\), and that, given all
+variables in \(\Delta\) are live on entry, \(s, t\) both target \(\ms{L}\), then
+the if-statement \(\ite{e}{s}{t}\) targets \(\ms{L}\) given \(\Gamma\) is live
+on entry. \rle{let-blk} and \rle{let2-blk} are quite similar to their term
+analogues, except that the bound term \(e\) is allowed to be impure (otherwise,
+programs could only make use of pure functions!). Finally, \rle{where} says
+that, given a block \(t\) which targets \(\ms{L}\), and a CFG \(L\) with inputs
+\(\ms{L}\) and outputs \(\ms{K}\), \(\ewhere{t}{L}\) targets \(\ms{K}\). CFGs
+are typed using the typing rules \rle{nil-br}, which says that an empty
+control-flow graph simply forwards the input labels directly to the output
+labels, and \rle{cons-br}, which says that if \(L\) maps \(\ms{L}\) to either
+\(\ms{K}\) or \(\lbl{\ell}[\Gamma](A)\), and \(t\) targets \(\ms{L}\) given that
+\(\Gamma, \thyp{x}{A}{}\) is live on entry, then \(L\) maps \(\ms{L}\) to
+\(\ms{K}\), having (potentially recursively) defined \(\lbl{\ell}\).
 
@@
     \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
-    \ms{W}'\) such that 
-    \begin{itemize}
-      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
-      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
-      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
-    \end{itemize}
+    \ms{W}'\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
+    \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\), and
+    \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
   \end{itemize}

### 2023-11-17 pldi24/paper.tex
Tap

@@
 judgements valid, i.e.
-\begin{lemma}[Weakening] \
-  \begin{itemize}
-    \item If \(\cwk{\Gamma}{\Delta}\), \(q \subseteq r\), and
-    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
-    \item If \(\cwk{\Gamma}{\Delta}\) and \(\haslb{\Delta}{t}{\ms{L}}\), then
-    \(\haslb{\Gamma}{t}{\ms{L}}\)
-    \item If \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{K}}\)
-    % \item If \(\lwk{\ms{L}}{\ms{K}}\) and \(\lhaslb{\ms{W}}{L}{\ms{L}}\) then
-    % \(\lhaslb{\ms{W}}{L}{\ms{K}}\)
-  \end{itemize}
+\begin{lemma}[Weakening] 
+  Given \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
+  \(\hasty{\Delta}{p}{a}{A}{r} \implies \hasty{\Gamma}{p}{a}{A}{q}\) and
+  \(\haslb{\Delta}{t}{\ms{L}} \implies \haslb{\Gamma}{t}{\ms{L}}\). Similarly,
+  if \(\lwk{\ms{L}}{\ms{K}}\) then \(\haslb{\Gamma}{t}{\ms{L}} \implies
+  \haslb{\Gamma}{t}{\ms{K}}\).
 \end{lemma}
@@
 
-Our main metatheoretic results will be \textit{substitution} and
-\textit{label-substitution}, which will require introducing some substitution
-judgements to state properly, which we do in Figure
-\ref{fig:ssa-meta-judgements}. In particular, the judgement
-\(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which is taken to be
-a map from variable names to expressions, is a valid substitution taking
-\(\Theta\) to \(\Gamma\). What this means is that, for each variable
-\(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted as a
-\textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such that
-when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables
-in \(\Theta\) are used up: this is described formally via the typing rules
-\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
-to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
-\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
-\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/discarding
-of substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
-\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).
+To state our primary syntactic metatheorems, \textit{substitution} and
+\textit{label-substitution}, we will need to introduce some new metatheoretic
+judgements, which we do in Figure \ref{fig:ssa-meta-judgements}. In particular,
+the judgement \(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which
+is taken to be a map from variable names to expressions, takes each variable
+\(\thyp{x}{A}{q}\) in \(\Gamma\) to a \textit{central} term \(\gamma(x)\) which
+can be interpreted as having type \(A\) and quantity \(q\) in a subcontext of
+\(\Theta\) with linearity \(q\), such that computing all \(\gamma(x)\) uses up
+all non-affine variables in \(\Theta\). This is described formally via the
+typing rules \rle{subst-nil}, which says that any function \(\gamma\) takes the
+empty context to itself, and \rle{subst-cons}, which says that if \(\gamma\)
+takes \(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\),
+\(q \subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the
+duplication/dropping of substructural variables), and \(\Theta\) splits into
+\(\Theta_\Gamma\) and \(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma,
+\thyp{x}{A}{q}}\).
 
@@
 % computationbs can be used in a term substituting for a relevant variable.
-
-We can now state our substitution theorem for terms in a straightforward way: if
-\(\issubst{\gamma}{\Theta}{\Gamma}\) and \(\hasty{\Gamma}{p}{e}{A}{q}\), then
-\(\hasty{\Theta}{p}{[\gamma]e}{A}{q}\) and
+%
+With these definitions, we can now state substitution for terms in a
+straightforward way: if \(\issubst{\gamma}{\Theta}{\Gamma}\) and
+\(\hasty{\Gamma}{p}{e}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]e}{A}{q}\) and
 \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{e}}{A}{q}\).
-Unfortunately, the substitution theorem for blocks is a little more complicated:
-the naive choice of attempting to show \(\haslb{\Gamma}{t}{\ms{L}} \implies
-\haslb{\Theta}{[\gamma]t}{\ms{L}}\) does not work, since \(\ms{L}\) may contain
-captured variables from \(\Gamma\) that are no longer available if we are typing
-\([\gamma]t\) in the context of \(\Theta\).
 
@@
 
-To solve this problem, we need to introduce the concept of a
+Unfortunately, substitution for blocks is a little more complicated: the naive
+choice of attempting to show \(\haslb{\Gamma}{t}{\ms{L}} \implies
+\haslb{\Theta}{[\gamma]t}{\ms{L}}\) does not work, since \(\ms{L}\) may require
+variables from \(\Gamma\) to be live which simply do not exist in \(\Theta\). To
+solve this problem, we need to introduce the concept of a
 \textit{label-substitution} \(\mc{L}\), which we define as a function that takes
@@
 fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)\). We
-introduce the judgement \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to state that the
-label-substitution \(\mc{L}\) takes the label-set \(\ms{L}\) to the label-set
-\(\ms{K}\); that is, for each \(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have
-that \(\mc{L}(\lbl{\ell}, x)\) is a block that targets \(\ms{K}\). This is
-formally defined via the rules \rle{lb-subst-nil}, which states that this is
-vacuously true for \(\ms{L} = \cdot\), and \rle{lb-subst-cons}, which says that
-if \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
+introduce the judgement \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to mean that for
+each \(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have that \(\mc{L}(\lbl{\ell},
+x)\) targets \(\ms{K}\). Formally, \rle{lb-subst-nil} states that this is
+vacuously true for \(\ms{L} = \cdot\), while \rle{lb-subst-cons} says that if
+\(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
 \thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{\ms{L},
-\lhyp{\lbl{\ell}}{\Gamma}{A}}{\ms{K}}\), in Figure \ref{fig:ssa-subst-typing}.
-Given a context \(\gamma\) and a label-set \(\ms{L}\), we can convert it to a
-label-substitution \(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
+\lhyp{\lbl{\ell}}{\Gamma}{A}}{\ms{K}}\). Given a context \(\gamma\) and a
+label-set \(\ms{L}\), we can define an associated label-substitution
+\(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
 \ref{fig:ssa-subst-ops}: using this, we can state the full substitution lemma as

### 2023-11-17 pldi24/paper.tex
Tap

@@
 branching. Finally, we will require an \textit{Elgot structure} to implement
-general, potentially nonterminating control flow.
+potentially nonterminating control flow.
 
@@
 
-We begin with some standard definitions:
+We begin with some standard definitions from the literature:
 \begin{definition}[Binoidal Category] 
@@
   \otimes -\) and \(- \otimes A\) are functors). 
-\end{definition}
   Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
@@
   simply write \(f \otimes g\), \(g \otimes f\) respectively.
+\end{definition}
 \begin{definition}[Premonoidal Category] 
@@
 One issue with the notion of a premonoidal category is that it is unclear what
-exactly a premonoidal functor should be: if we want them to compose, we would
-like to preserve central morphisms, but trying to preserve \textit{all} central
-morphisms is too strict a notion. We want, in essence, to be able to distinguish
-morphisms that are truly "pure" from those which simply happen to be central. To
-do that, we introduce the notion of an \textit{effectful category} as follows:
+exactly a premonoidal functor should be: if we want them to compose, we need to
+preserve central morphisms, but trying to preserve \textit{all} central
+morphisms rules out many important examples. We therefore would to be able to
+distinguish morphisms that are truly "pure" from those which are simply
+"accidentally" central. To do that, we introduce the notion of an
+\textit{effectful category} as follows:
 \begin{definition}[Effectful Category] 
@@
 \end{definition}
-Unfortunately, the notion of "purity" given by an effectful category only allows
-us to perform relatively weak optimistions in the presence of effects. To remedy
-this, we want to be able to make a finer distinction between morphisms which are
-merely \textit{central} and those which are also \textit{relevant} (i.e.,
-copying the input and applying them twice is the same as applying them once and
-copying the output) and/or \textit{affine} (i.e., applying them and discarding
-the result is the same as never doing them at all). Doing so also allows us to
-add the capability of \textit{distinguishing} between objects which are affine,
-relevant, neither or both, enabling us to reason about substructural type
-systems categorically. To do this, we begin by introducing the notion of a \textit{symmetric
+As is, this definition presents us with two extreme options: we can either
+assume \(\mc{V}\) is simply monoidal, and hence that "pure" morphisms are merely
+central, or that \(\mc{V}\) is Cartesian, and hence that all pure morphisms are
+truly \textit{pure}. We would like, however, to make a finer distinction between
+morphisms which are merely \textit{central} and those which are also
+\textit{relevant} and/or \textit{affine}, as defined in the introduction. Doing
+so also allows us to add the capability of define what it means for an
+\textit{object}, which we can interpret as a type, is affine or relevant as
+well. With this aim in mind, we introduce the notion of a \textit{symmetric
 premonoidal subcategory}, as follows:
@@
 \end{definition} 
-
 Note that the set of symmetric premonoidal subcategories, monoidal
 subcategories, and central subcategories of a category \(\mc{C}\) are all closed
-under intersections.
+under intersections, forming a complete meet-semilattice.
 %
@@
 We can now define a \textit{substructural premonoidal category}, which is simply
-a premonoidal category equipped with a distinguished set of relevant and affine
-morphisms
+a premonoidal category equipped with premonoidal subcategories of relevant and
+affine morphisms:
 \begin{definition}[Substructural (Pre)monoidal Category]
-  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A},
-  \mc{R})\) where:
+  A \textbf{substructural (pre)monoidal category} is a symmetric premonoidal
+  category \(\mc{C}\) equipped with
   \begin{itemize}
-    \item \(\mc{C}\) is a symmetric premonoidal category equipped with wide
-    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\). We
-    define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell
-    \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)
-    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects
-    closed under tensor products, such that \(\mb{1} \in \mc{A}\) and
+    \item Wide symmetric premonoidal subcategories \(\mc{C}^{\ms{a}},
+    \mc{C}^{\ms{r}}\). We define \(\mc{C}^q = \bigcap_{\ell \in q}\mc{C}^\ell\),
+    where \(\mc{C}^\varnothing = \mc{C}\)
+    \item A set of \textbf{affine} objects \(\ms{Aff}(\mc{C}) = \mc{A} \subseteq
+    |\mc{C}|\) closed under tensors, such that \(\mb{1} \in \mc{A}\) and
     \begin{itemize}
@@
     \end{itemize}
-    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects
-    closed under tensor products, such that \(\mb{1} \in \mc{R}\) and
+    \item A set of \textbf{relevant} objects \(\ms{Rel}(\mc{C}) = \mc{R}
+    \subseteq |\mc{C}|\) closed under tensors, such that \(\mb{1} \in \mc{R}\)
+    and
     \begin{itemize}
@@
   \end{itemize}
-  We call the morphisms in \(\mc{C}^{\ms{a}}\) \textbf{affine}, the
-  morphisms in \(\mc{C}^{\ms{r}}\) \textbf{relevant}, and the morphisms in
-  \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}. We
-  call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the
-  other \(\mc{C}^q\) \textbf{linear}. We call a substructural \textit{monoidal}
-  category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} =
-  |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
+  We call the morphisms in \(\mc{C}^{\ms{a}}\) \textbf{affine}, the morphisms in
+  \(\mc{C}^{\ms{r}}\) \textbf{relevant}, and the morphisms in
+  \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{intuitionistic}. 
+  % We call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any
+  % of the other \(\mc{C}^q\) \textbf{linear}. 
+  We call a substructural \textit{monoidal} category a \textbf{substructural
+  Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case
+  \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
 \end{definition}
@@
 morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) =
-\varnothing\), we have \( (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a)
-= \{()\} \). We will see that this corresponds to the fact that \( \ms{let}\;x =
-\ms{loop}(); e \) is not semantically equal to \(e\) even where \(x \notin
+\varnothing\), we have \((R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a)
+= \{()\} \). We will see that this corresponds to the fact that "\(\ms{let}\;x =
+\ms{loop}(); e\)" is not semantically equal to "\(e\)" even where \(x \notin
 \ms{fv}(e)\). Similarly, we cannot simply have \(\ms{Rel}^{\ms{r}} = \ms{Rel}\),
-since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A,
-|R(a)| \supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
-\begin{equation}
+since, for any \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)| > 1\),
+\(
   (R;\ms{split}(B))(a) = \{(b, b) \mid b \in R(a)\} \neq (\ms{split}(A);R)(a) = R(a) \times R(a) 
-\end{equation}
-since only the left-hand side contains the pair \((b_0, b_1)\), this
-corresponding to the fact that the program \(\ms{let}\;x = \ms{nondet}(); (x, x)\) is
-not semantically equal to \((\ms{nondet}(), \ms{nondet}())\).
-
-To get a substructural premonoidal category, we can apply the state monad
-transformer to the power set monad to obtain the Kleisli category with morphisms
-of the form \(\ms{Set}_{\ms{St}\;\mc{P}\;S}(A, B) = A \to S \to \mc{P}(B \times
-S)\). Once again, it is clear that all objects have split and drop morphisms
-inherited from \(\ms{Set}\), and therefore are affine and relevant.
-
-One issue with the above definition is that, just like for regular premonoidal
-categories, it is difficult to distinguish between those morphisms that are
-"truly" central versus those which are just central "by accident", as we only
-want to require any notion of functor we develop subsequently to preserve the
-former. We can solve this the same way, by simply extending the notion of
-effectful category appropriately as follows:
+\). This corresponds
+to the fact that the program "\(\ms{let}\;x = \ms{nondet}(); (x, x)\)" is not
+semantically equal to "\((\ms{nondet}(), \ms{nondet}())\)".
+
+% To get a substructural premonoidal category, we can apply the state monad
+% transformer to the power set monad to obtain the Kleisli category with morphisms
+% of the form \(\ms{Set}_{\ms{St}\;\mc{P}\;S}(A, B) = A \to S \to \mc{P}(B \times
+% S)\). It is clear that all objects have split and drop morphisms inherited from
+% \(\ms{Set}\), and therefore are affine and relevant.
+
+Just like for regular premonoidal categories, we would like to distinguish
+between those morphisms that are "truly" central versus those which are just
+central "by accident". We can solve this the same way, by simply extending the
+notion of effectful category appropriately as follows:
 \begin{definition}[Susbtructural Effectful Category]
-  An effectful category \(\mc{C}_0 \to \mc{C}_1\) is said to be
+  An effectful category \(\mc{C}_1 \to \mc{C}_0\) is said to be
   \textbf{substructural} if \(\mc{C}_0, \mc{C}_1\) are substructural and
-  \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0^\varnothing\) preserves all
-  symmetries, associators, unitors, drops, and splits and sends \(\mc{C}_1^q\)
-  to \(\mc{C}_0^q\). In particular, restricting \(\upg{\cdot}{}\) appropriately
-  makes all \(\mc{C}_1^q \to \mc{C}_0^q\) into effectful categories. Morphisms
-  in \(\mc{C}_p^{\ms{r}}\) are called \textbf{copyable}, morphisms in
-  \(\mc{C}_p^{\ms{a}}\) are called \textbf{discardable}, while morphisms in
-  \(\mc{C}_1^{\ms{a}}, \mc{C}_1^{\ms{r}}, \mc{C}_1^{\{\ms{a}, \ms{r}\}}\)
-  specifically are called \textbf{affine}, \textbf{relevant}, and
-  \textbf{intuitionistic} or \textbf{pure} respectively.
+  \(\upg{\cdot}{}: \mc{C}_1 \to \mc{C}_0\) preserves all drops and splits and
+  sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\). In particular, restricting
+  \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to \mc{C}_0^q\) into
+  effectful categories.  We call a morphism in \(\mc{C}_1\) \textbf{central},
+  and a central intuitionistic morphism \textbf{pure}.
+  % Morphisms in \(\mc{C}_p^{\ms{r}}\) are called \textbf{copyable}, morphisms
+  % in \(\mc{C}_p^{\ms{a}}\) are called \textbf{discardable}, while morphisms in
+  % \(\mc{C}_1^{\ms{a}}, \mc{C}_1^{\ms{r}}, \mc{C}_1^{\{\ms{a}, \ms{r}\}}\)
+  % specifically are called \textbf{affine}, \textbf{relevant}, and
+  % \textbf{intuitionistic} or \textbf{pure} respectively.
 \end{definition}
@@
 \mc{P}(A \times S)\). Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
-\(\upg{R}{} = \lambda a, s. R(a) \times \{s\}\), we can verify that this
-satisfies the axioms of a substructural effectful category. Note that here the
-\textit{central} morphisms are precisely those which do not read from or write
-to the state, and therefore which commute with all other morphisms.
+\(\upg{R}{} = \lambda a, s. R(a) \times \{s\}\) to be the standard lifting
+operator, we can verify that this satisfies the axioms of a substructural
+effectful category. Note that here the \textit{central} morphisms are precisely
+those which do not read from or write to the state, and therefore which commute
+with all other morphisms.
 
@@
 affine; in fact, this is the case \textit{if and only if} a category is
-Cartesian. We also note the following:
+Cartesian. We also have that:
 \begin{proposition}
@@
 
-To add support for branching, we need to equip our category with a
-\textit{distributive coproduct structure}, as follows:
+To support branching, we need to add a bit more structure:
 \begin{definition}[Distributive (pre)monoidal category] 
@@
   is a distributive substructural effectful category (with affine and relevant
-  objects and morphisms inherited from \(\mc{C}\))
+  objects and morphisms lifted from \(\mc{C}\))
 \end{proposition}
@@
 % Proofs can be found in Lemma 31 of \cite{goncharov-squaring}.
-
-One important way of generating Elgot categories is through monads: a monad
-\(\mb{T}\) whose Kliesli category \(\mb{C}_{\mb{T}}\) is Elgot is called an
-\textbf{Elgot monad}. Examples of Elgot monads include \(\ms{Option}\) (where
+A monad \(\mb{T}\) whose Kliesli category \(\mb{C}_{\mb{T}}\) is Elgot is called
+an \textbf{Elgot monad}. Examples of Elgot monads include \(\ms{Option}\) (where
 nontermination is \(\ms{None}\)), the powerset monad (where nontermination is
@@
 transformers preserve Elgot structure, allowing us to build up Elgot monads in a
-composable fashion. In particular,
+composable fashion via a monad stack. In particular,
 \begin{proposition}
@@
   \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to
-  \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot},
-  equipped with
+  \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot}
   \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects

### 2023-11-17 pldi24/paper.tex
Another page!

@@
 We begin by giving a semantics for types and contexts in terms of objects in
-\(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}: we map tensor products to
-tensor products, the unit type to the monoidal unit, and booleans to \(\mb{1} +
-\mb{1}\). Contexts are simply interpreted as the tensor products of the variable
-types which appear in them; in particular, the quantity annotations have
-\textit{no} effect on a context's semantics. We then give a semantics for
-structural judgements in Figure \ref{fig:ssa-structural-semantics}. In
-particular, we begin by interpreting the judgements that a type is affine (i.e.
-\(\tlin{A}{\taff}\)) or relevant (i.e. \(\tlin{A}{\trel}\)) as the drop morphism
-\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, I)\) and split morphism
-\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, \dnt{A} \otimes \dnt{A})\) respectively.
-We can then interpret a context split \(\csplits{\Gamma}{\Delta}{\Xi}\) as a
-pure morphism \(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\)
-which partitions the variables as appropriate, splitting and dropping as
-necessary. We interpret a context weakening \(\cwk{\Gamma}{\Delta}\) as the
-associated split \(\csplits{\Gamma}{\Delta}{\cdot}\) postcomposed with a unitor.
-Finally, we interpret a label-set weakening \(\lwk{\ms{L}}{\ms{K}}\) as a pure
-morphism \(\mc{C}^\tint_1(\dnt{\ms{L}}, \dnt{\ms{K}})\), where \rle{join-nil} is
-the identity, \rle{join-cons} is the sum of \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) and
+Figure \ref{fig:ssa-type-semantics}: we map tensor products to tensor products,
+the unit type to the monoidal unit, and booleans to \(\mb{1} + \mb{1}\).
+Contexts are simply interpreted as the tensor products of the variable types
+which appear in them; in particular, quantity annotations have \textit{no}
+effect on a context's semantics. We proceed to give semantics for structural
+judgements in Figure \ref{fig:ssa-structural-semantics}. In particular, we
+interpret a context split \(\csplits{\Gamma}{\Delta}{\Xi}\) as a pure morphism
+\(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\) which
+partitions the variables as appropriate, splitting and dropping as necessary; a
+weakening \(\cwk{\Gamma}{\Delta}\) is interpreted as the associated split
+\(\dnt{\csplits{\Gamma}{\Delta}{\cdot}}\) postcomposed with a unitor. Finally, a
+label-set weakening \(\lwk{\ms{L}}{\ms{K}}\) is interpreted as a pure morphism
+\(\mc{C}^\tint_1(\dnt{\ms{L}}, \dnt{\ms{K}})\), where \rle{join-nil} is the
+identity, \rle{join-cons} is the sum of \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) and
 \(\dnt{\cwk{\Gamma}{\Delta}}\), and \rle{join-zero} is
@@
       : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta})} 
-      \\
+      \qquad \qquad
       \dnt{\cwk{\Gamma}{\Delta}} 
@@
       = \upg{\dnt{\cwk{\Gamma}{\cdot}}}{p} 
-      \\
+      \quad
     \dnt{\hasty{\Gamma}{p}{\ctt}{\mb{2}}{q}}
-      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\ctt)}{p}
-      \qquad
+      = \upg{\dnt{\cwk{\Gamma}{\cdot}}}{p};\iota_0
+      \quad
     \dnt{\hasty{\Gamma}{p}{\cff}{\mb{2}}{q}}
-      = \upg{(\dnt{\cwk{\Gamma}{\cdot}};\cff)}{p} 
+      = \upg{\dnt{\cwk{\Gamma}{\cdot}}}{p};\iota_1 
       \\
@@
 \textit{centrality} \(p\) and \textit{quantity} \(q\) (i.e., lying in
-\(\mc{C}_p^q\)). This means that, as expected, a central term becomes a central
-morphism in \(\mc{C}_1\), an affine term becomes an affine morphism, and a
-relevant term becomes a relevant morphism, and so, in particular, a pure term
-becomes a pure morphsim. Our rules for doing so are given in Figure
-\ref{fig:ssa-term-semantics}; we will abuse syntax slightly and write
+\(\mc{C}_p^q\)); our rules for doing so are given in Figure
+\ref{fig:ssa-term-semantics}. We will abuse syntax slightly and write
 \(\dnt{\text{"judgement"}}\) for \(\dnt{D}\), where \(D\) is a derivation of
@@
 
-Our treatment of \rle{var} is quite standard: the term \(x\) types in \(\Gamma\)
-if all other variables in \(\Gamma\) and can be discarded, in which case the
-semantics is simply doing so, i.e. \(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\),
-lifting to \(\mc{C}_0\) as necessary. Similarly, to type a function application
-\rle{var}, we first take \(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is
-guarnateed to be central, lift it to the appropriate category \(\mc{C}_p\), and
-then simply postcompose it with the denotation of \(f\). \rle{pair} is handled
-by composing the context split \(\csplits{\Gamma}{\Delta}{\Xi}\) taking the
-tensor product of the components \(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\),
-\(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is well-defined since both components
-are central and \(\mc{C}_1\) is monoidal. We then lift the result to the
-appropriate \(\mc{C}_p\). The semantics of \rle{unit} is simply dropping the
-entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\), while \rle{true}, and \rle{false}
-are each handled by first dropping the context and then postcomposing with the
-appropriate constant; in all cases, this is followed by lifting to \(\mc{C}_p\).
-Finally, \rle{let} and \rle{let2} are handled by first splitting the context,
-then passing the right hand component to the semantics of the bound term
+Our treatment of \rle{var} is quite standard: semantics of \(x\) is simply
+discarding all other variables in \(\Gamma\), i.e.
+\(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\). Similarly, to type a function
+application \rle{var}, we first take \(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which
+is guaranteed to be central, lift to \(\mc{C}_p\), and then simply postcompose
+with \(f\). \rle{pair} is handled by precomposing the context split
+\(\csplits{\Gamma}{\Delta}{\Xi}\) with the tensor product of the components
+\(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\), \(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is
+well-defined since both components are central and \(\mc{C}_1\) is monoidal. The
+semantics of \rle{unit} is simply dropping the entire context, while \rle{true}
+and \rle{false} are interpreted as first dropping the context and then injecting
+into the appropriate branch of the boolean type. Finally, \rle{let} and
+\rle{let2} are handled by first splitting the context, then passing the right
+hand component to the semantics of the bound term
 \(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or \(\dnt{\hasty{\Xi}{1}{a}{A \otimes
@@
 \thyp{x}{A}{q}}{p}{e}{B}{q}}\) or \(\dnt{\hasty{\Delta, \thyp{x}{A}{q},
-\thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that this inner term is not necessarily
-central, unlike the bound term!
+\thyp{y}{B}{q}}{p}{e}{C}{q}}\). \
+% Note that this inner term is not necessarily central, unlike the bound term!
 
@@
 
-Similarly to terms, \textit{blocks} \(\haslb{\Gamma}{t}{\ms{L}}\) are
-interpreted as \textit{impure} morphisms from \(\dnt{\Gamma}\) to \(\ms{L}\):
-here \(\Gamma\) represents all the variables live on entry to the block, while
-\(\ms{L}\) represents all the external blocks which \(t\) might branch to, each
-annotated with their live variables and their arguments; our rules are given in
-Figure \ref{fig:ssa-block-semantics}. Our semantics for break statements
-\(\lbrb{\ell}{a}\) is as follows: we begin by splitting the context into live
-variables which pass into dominated blocks and variables used in the argument
-via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\). We then pass the latter into the
-semantics of the argument (which is pure, and therefore must be lifted to
-\(\mc{C}_0\)) \(\dnt{\hasty{\Gamma}{1}{a}{A}{\tint}}\), with the result and live
-variables re-associated and passed into
-\(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to obtain the desired target. The
-semantics of if-statements are similarly given by first splitting the context
-via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\) into variables live in the branches,
-\(\Delta\), and variables appearing in the discriminant, \(\Xi\); the latter of
-which are passed into the lifted pure morphism
-\(\dnt{\hasty{\Delta}{1}{b}{\mb{2}}{\tint}}\). The "if" statement itself is then
-simply an application of the distributor \(\delta: A \otimes (\mb{1} + \mb{1})
-\to A + A\), with each branch passed to the semantics of its corresponding block
-\(\dnt{\haslb{\Xi}{s}{\ms{L}}}\), \(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The
-semantics of \rle{let-blk} and \rle{let2-blk} are analogous to those of
-\rle{let} and \rle{let2} in the term case, except that now the bound term \(a\)
-is allowed to be impure (and therefore in particular does not need to be
-upgraded).
+We intrpret \textit{blocks} \(\haslb{\Gamma}{t}{\ms{L}}\) as \textit{impure}
+morphisms from \(\dnt{\Gamma}\) to \(\ms{L}\): \(\Gamma\) represents all the
+variables live on entry to the block, while \(\ms{L}\) represents all valid
+targets, each annotated with live variables and an argument type; our rules are
+given in Figure \ref{fig:ssa-block-semantics}. \rle{br} is interpreted by first
+splitting the context via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\), and then
+using \(\Xi\) to compute the target's argument. The result, along the unchanged
+live variables \(\Delta\), is then passed to
+\(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to obtain the desired target.
+Similarly, we interpret \rle{ite} by first splitting the context via
+\(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\) and using \(\Xi\) to compute the
+discriminant. The "if" statement itself is then simply an application of the
+distributor \(\delta: \dnt{\Delta} \otimes (\mb{1} + \mb{1}) \to \dnt{\Delta} +
+\dnt{\Delta}\); we pass the left of the coproduct to the true branch
+\(\dnt{\haslb{\Xi}{s}{\ms{L}}}\) and the right to the false branch
+\(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The semantics of \rle{let-blk} and
+\rle{let2-blk} are analogous to those of \rle{let} and \rle{let2}, except that
+now the bound term \(a\) is allowed to be impure.
 

### 2023-11-17 pldi24/paper.tex
Tap

@@
 
-Finally, we must give semantics for \rle{where}. To do so, we begin by giving a
-interpreting a control-flow graph \(\lhaslb{\ms{L}}{L}{\ms{K}}\) as a morphism
-from \(\dnt{\ms{L}}\) to \(\dnt{\ms{K}} + \dnt{\ms{L}}\), where the left-hand
-side of the coproduct denotes branching to one of the target labels in
-\(\ms{K}\) (the \textit{output}), while the right-hand side of the coproduct
-denotes a recursive call to a label in \(\ms{L}\) (the \textit{recursive
-input}). In particular, \rle{nil-br} simply treats a call to one of \(\ms{L}\)
-as a call to a \textit{target label}, passing it through the left injection
-unchanged. On the other hand \rle{cons-br} is a little bit more complicated:
-first we execute \(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}\),
-which will give us either an output \(\dnt{\ms{K}, \lhyp{\ell}{\Gamma}{A}}\), or
-a recursive input \(\dnt{\ms{L}}\), the latter of which we can pass through
-unchanged as a recursive input. In the output case, we can interpret this as
-either a value of \(\dnt{\ms{K}}\) or \(\dnt{\Gamma} \otimes \dnt{A}\); the
-former we can pass the value through unchanged as output; in the second case, we
-can pass this in as input to \(\dnt{\haslb{\Gamma,
-\thyp{x}{A}{q}}{t}{\ms{L}}}\), which will give us a value of \(\ms{L}\) we can
-pass in as recursive input. Note that, even if \(t\) targets an output label in
-\(\ms{K}\), we still return a \textit{recursive input}, not an output; in this
-case, the \textit{next} iteration will return it as an output. We can now quite
-easily state the semantics of \rle{where}: we simply precompose the semantics of
-the entry block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the \textit{iterated}
+Finally, we must give semantics for \rle{where}. To do so, we interpret a
+control-flow graph \(\lhaslb{\ms{L}}{L}{\ms{K}}\) as a morphism from
+\(\dnt{\ms{L}}\) to \(\dnt{\ms{K}} + \dnt{\ms{L}}\), where the left of the
+coproduct denotes branching to one of the labels in \(\ms{K}\) (the
+\textit{output}), while the right of the coproduct denotes a recursive call to a
+label in \(\ms{L}\) (the \textit{recursive input}). In particular, \rle{nil-br}
+simply forwards all calls to one of \(\ms{L}\) to the output unchanged.
+\rle{cons-br} is a little bit more complicated: first we execute
+\(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}\), which yields
+either a recursive input \(\dnt{\ms{L}}\), which we pass through unchanged as a
+recursive input. In the former case, we can interpret this as either a value of
+\(\dnt{\ms{K}}\), which we pass through unchanged as output, or \(\dnt{\Gamma}
+\otimes \dnt{A}\); which we can pass as input to \(\dnt{\haslb{\Gamma,
+\thyp{x}{A}{q}}{t}{\ms{L}}}\), giving us a value of \(\ms{L}\) we can use as
+recursive input. Note that, even if \(t\) yields an output label in \(\ms{K}\),
+we still return a \textit{recursive input}, not an output; in this case, the
+\textit{next} iteration will return it as an output. We can now quite easily
+state the semantics of \rle{where}: we simply precompose the semantics of the
+entry block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the \textit{iterated}
 semantics of the associated control-flow graph,

### 2023-11-17 pldi24/paper.tex
Tap

@@
 matches the syntax by showing that the syntactic and semantic notions of
-substitution coincide. We begin by giving a semantics for the judgement
-\(\issubst{\gamma}{\Theta}{\Gamma}\), which we will interpret as a pure morphism
-taking \(\dnt{\Theta}\) to \(\dnt{\Gamma}\), defined via the rules in Figure
-\ref{fig:ssa-subst-semantics}, which can simply be viewed as computing all
-\(\gamma(x)\) for \(\thyp{x}{A}{q}\) in \(\Gamma\). In particular, we interpret
-\rle{subst-nil} as the identity function, and \rle{subst-cons} by splitting the
-context via \(\dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}\), and then
-passing \(\dnt{\Theta_\Gamma}\) to the remainder of the substitution
-\(\dnt{\issubst{\gamma}{\Gamma}{\Theta_\Gamma}}\) and \(\dnt{\Theta_x}\) through
-\(\dnt{\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}}\) to yield \(\dnt{A}\). We give a
-similar semantics for label-substitutions: \rle{lb-subst-nil} is interpreted as
-the zero morphism \(0_{\dnt{\ms{K}}}: \mc{C}_0^\varnothing(\mb{0},
-\dnt{\ms{K}})\), while \rle{lb-subst-cons} takes in the coproduct \(\dnt{\ms{L}}
-+ \dnt{\Gamma} \otimes \dnt{A}\), sending the former to
-\(\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{A}}}\) and the latter to the denotation of
-\(\mc{L}(\lbl{\ell}, x)\).
+substitution coincide. We begin by giving a semantics for
+\(\issubst{\gamma}{\Theta}{\Gamma}\) in Figure \ref{fig:ssa-subst-semantics},
+which we will interpret as a pure morphism taking \(\dnt{\Theta}\) to
+\(\dnt{\Gamma}\). This morphism can simply be viewed as computing all
+\(\gamma(x)\) for \(\thyp{x}{A}{q}\) in \(\Gamma\); in particular, we interpret
+\rle{subst-cons} by splitting the context via
+\(\dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}\), passing
+\(\dnt{\Theta_\Gamma}\) to the remainder of the substitution
+\(\dnt{\issubst{\gamma}{\Gamma}{\Theta_\Gamma}}\) to compute the rest of the
+\(\gamma(y)\) in \(\Gamma\) recursively, and using \(\dnt{\Theta_x}\) to compute
+\(\gamma(x)\). Similarly, we interpret label-substitutions as morphisms from \(\dnt{\ms{L}}\) to
+\(\dnt{\ms{K}}\) which, for each input label in \(\ms{L}\), performs a
+computation targeting \(\ms{K}\). In particular, \rle{lb-subst-nil} is
+interpreted as the zero morphism \(0_{\dnt{\ms{K}}}:
+\mc{C}_0^\varnothing(\mb{0}, \dnt{\ms{K}})\), while \rle{lb-subst-cons} takes in
+the coproduct \(\dnt{\ms{L}} + \dnt{\Gamma} \otimes \dnt{A}\), sending the
+former to \(\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{A}}}\) and the latter to the
+denotation of \(\mc{L}(\lbl{\ell}, x)\).
 
@@
 
-Given this, we can now state the theorem of semantic substitution, which
-essentially states that the semantics of a substituted term or block is the same
-as the semantics of that term or block preceded by let-bindings corresponding to
-the variables in the substitution, which is itself the same as the semantics of
-the original term or block precomposed with the semantics of the substitution as
-given in Figure \ref{fig:ssa-subst-semantics}. Similarly, the semantics of a
-\textit{label-substituted} block is the same as the semantics of the original
-block \textit{postcomposed} with the semantics of that label-substitution.
-Stated formally, this becomes:
+We can now state the theorem of semantic substitution, which essentially states
+that the semantics of a substituted term or block is the same as the semantics
+of that term or block preceded by let-bindings corresponding to the variables in
+the substitution, which is the same as the semantics of the original term or
+block precomposed with the semantics of the substitution. Similarly, the
+semantics of a \textit{label-substituted} block is the same as the semantics of
+the original block \textit{postcomposed} with the semantics of that
+label-substitution. Stated formally, we obtain:
 \begin{theorem}[Semantic Substitution] 
@@
     \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
-    \[
+    \(
       \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
@@
       = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
-    \]
+    \)
     \item For all 
@@
     \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
-    \[
+    \(
       \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
       = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
-    \]
+    \)
     \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have

### 2023-11-17 pldi24/paper.tex
Tap
Half a page

@@
 Because we have a denotational semantics, it is easy to validate equations
-between terms, blocks, and CFGs.
-
-Even the basic proof of the soundness of substitution already lets us prove the
-soundness of many fundamental compiler optimisations, such as global value
-numbering (let-introduction for pure expressions), some forms of dead code
-elimination (i.e., eliminating unused affine bindings), and strength reduction
-with loop hoisting (let-introduction for relevant expressions).
+between terms, blocks, and CFGs. Even the basic proof of the soundness of
+substitution already lets us prove the soundness of many fundamental compiler
+optimisations, such as global value numbering (let-introduction for pure
+expressions), some forms of dead code elimination (i.e., eliminating unused
+affine bindings), and strength reduction with loop hoisting (let-introduction
+for relevant expressions).
 
@@
 definitions:
+\begin{definition}[Relevant, Affine Monad]
+  A monad \(\mb{T}\) on \(\ms{Set}\) is \textbf{relevant} if its Kleisli
+  category satisfies \(f;\ms{split} = \ms{split};f \ltimes f = \ms{split};f
+  \rtimes f\), where \(\ms{split} = \upg{\lambda a. (a, a)}{}\). \(\mb{T}\) is
+  \textbf{affine} if its Kleisli category satisfies \(f;\ms{drop} = \ms{drop}\),
+  where \(\ms{drop} = \upg{\lambda a.()}{}\). A monad is \textbf{intuitionistic}
+  if it is both relevant and affine.
+\end{definition}
+\begin{definition}[Submonad] 
+  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a
+  collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq
+  \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in
+  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say
+  that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and,
+  for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where
+  \(f^\dagger\) is computed as in \(\mb{T}\). Note that \(\mb{S}\) is also an
+  (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
+  from \(\mb{T}\). Some important examples of submonads include:
+  \begin{enumerate}
+    \item The \textbf{unit submonad} \(\eta\) of every monad \(\mb{T}\) with
+    \(\eta(A) = \{\eta_A\;a \mid a \in A\}\).
+    %; similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\). 
+    \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
+    intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
+    submonads, forming a complete meet-semilattice with bottom \(\eta\) and top
+    \(\mb{T}\),
+    % ; this induces a complete lattice in the standard manner.
+    \item In particular, the intersection of a relevant/affine/commutative
+    submonad and another submonad is relevant/affine/commutative respectively.
+  \end{enumerate}
+\end{definition}
 \begin{definition}[Stream Action]
@@
 \end{definition}
-
-We now give a few basic definitions relevant to monads over \(\mb{Set}\):
-\begin{definition}[Relevant, Affine Monad]
-  A monad \(\mb{T}\) on \(\ms{Set}\) is \textbf{relevant} if its Kleisli
-  category satisfies \(f;\ms{split} = \ms{split};f \ltimes f = \ms{split};f
-  \rtimes f\), where \(\ms{split} = \upg{\lambda a. (a, a)}{}\). \(\mb{T}\) is
-  \textbf{affine} if its Kleisli category satisfies \(f;\ms{drop} = \ms{drop}\),
-  where \(\ms{drop} = \upg{\lambda a.()}{}\). A monad is \textbf{intuitionistic}
-  if it is both relevant and affine.
-\end{definition}
-\begin{definition}[Submonad] 
-  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a
-  collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq
-  \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in
-  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say
-  that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and,
-  for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where
-  \(f^\dagger\) is computed as in \(\mb{T}\). Note that \(\mb{S}\) is also an
-  (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
-  from \(\mb{T}\). Some important examples of submonads include:
-  \begin{enumerate}
-    \item The \textbf{unit submonad} \(\eta\) of every monad \(\mb{T}\) with
-    \(\eta(A) = \{\eta_A\;a \mid a \in A\}\).
-    %; similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\). 
-    \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
-    intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
-    submonads, forming a complete meet-semilattice with bottom element \(\eta\)
-    and top element \(\mb{T}\),
-    % ; this induces a complete lattice in the standard manner.
-    \item In particular, the intersection of a relevant/affine/commutative
-    submonad and another submonad is relevant/affine/commutative respectively.
-  \end{enumerate}
-\end{definition} 
 We can use the trace monad described above to derive some interesting examples
@@
 
-As a concrete example, we can now introduce the \textbf{printing monad}, which
-is simply defined as \(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where
-\(\Sigma: (\ms{Byte}^*)^\omega \to \ms{Byte}^{\leq \omega}\) denotes
-concatenation of bytestrings into a potentially infinite bytestream. This can be
-equipped with the structure of a substructural effectful category by choosing
-\(M_C = M_R = \{[]\}\), giving us an \isotopessa model supporting the
-instructions \(\ms{print} \in \mc{I}^\varnothing_0(\ms{Byte}^*, \mb{1})\) and
-\(\ms{nondet} \in \mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard
-semantics \(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), b)\}\),
-\(\dnt{\ms{nondet}} = \lambda (). \{\iota_0 (a, []) \mid a \in A\}\).
+We can now introduce the \textbf{printing monad}, which is simply defined as
+\(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where \(\Sigma:
+(\ms{Byte}^*)^\omega \to \ms{Byte}^{\leq \omega}\) denotes concatenation of
+bytestrings into a potentially infinite bytestream. This can be made into a
+substructural effectful category by choosing \(M_C = M_R = \{[]\}\), giving us
+an \isotopessa model supporting instructions \(\ms{print} \in
+\mc{I}^\varnothing_0(\ms{Byte}^*, \mb{1})\) and \(\ms{nondet} \in
+\mc{I}^\taff_1(\mb{1}, A)\) (for \(A\) nonempty), with semantics
+\(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), b)\}\), \(\dnt{\ms{nondet}} =
+\lambda (). \{\iota_0 (a, []) \mid a \in A\}\).
 
@@
 defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where
-\(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function. The
-Kleisli category of this monad can be equipped with the structure of a
-substructural effectful category by simply taking the affine/relevant morphisms
-to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\);
-as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we
-therefore have an \isotopessa model supporting the instructions \(\ms{set}:
+\(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function with
+finite support. The Kleisli category of this monad is a substructural effectful
+category and inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\);
+we therefore have an \isotopessa model supporting the instructions \(\ms{set}:
 \mc{I}^\varnothing_0(\nats \times \nats, \mb{1})\), \(\ms{get}:
@@
 memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}}
-= \lambda p\;h. \{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\); for now,
-this will do. Finally, \(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', [])
-\mid h' = h \sqcup p \mapsto v\}\) simply fills a random empty heap cell with
-the provided value, returning a pointer to the cell, while \(\dnt{\ms{free}} =
-\lambda p\;h. \{\iota_0\;((), h \setminus p, [])\}\)
-
-We can use this construction to, building on the ideas in \cite{mellies-ftrs},
-show how a simple separation logic based on \cite{reynolds-separation-2002}
-slightly adapted to the categorical setting gives us a very rich \isotopessa
-model supporting a basic notion of refinement types, while at the same time
-giving us a nontrivial notion of nonlinear types.
-
-Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of
-type \(A \to \ms{Heap} \to \ms{Prop}\). Taking as objects pairs \((A, \Phi)\),
-we can define the category \(\ms{Sep}\) to have morphisms
+= \lambda p\;h. \{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\). Finally,
+\(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', []) \mid h' = h \sqcup p
+\mapsto v\}\) simply fills a random empty heap cell with the provided value,
+returning a pointer to the cell, while \(\dnt{\ms{free}} = \lambda p\;h.
+\{\iota_0\;((), h \setminus p, [])\}\).
+
+Building on the ideas in \citet{mellies-ftrs}, we show how a simple separation
+logic based on \citet{reynolds-separation-2002} slightly adapted to the
+categorical setting gives us a very rich \isotopessa model supporting a basic
+notion of refinement types, while at the same time giving us a nontrivial notion
+of nonlinear types. Taking as objects pairs \((A, \Phi)\), where \(\Phi: A \to
+\ms{Heap} \to \ms{Prop}\), we can define the category \(\ms{Sep}\) to have
+morphisms
 \begin{equation}
@@
 \)
-is the usual partial correctness triple definition and the separating
-conjunction of heap-predicates is defined as usual to be 
+corresponds to the usual notion of a partial correctness triple, and the
+separating conjunction of heap-predicates is defined as
 \begin{equation}
@@
 We can naturally define a tensor product on objects \((A, \Phi) \otimes (B,
-\Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking
+\Psi) = (A \times B, \Phi \ast \Psi)\), making \(\ms{Sep}\) into a substructural
+effectful category with:
 \begin{itemize}
@@
   \times C\), which are valid definitions by the frame rule
-\end{itemize}
-equips \(\ms{Sep}\) with the structure of a premonoidal category. We can give \(\ms{Sep}\) the structure of a substructural effectful category by taking:
-\begin{itemize}
-  \item Affine objects to be precisely those for which \(\ms{drop} =
-  \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), and
-  relevant objects to be precisely those for which \(\ms{split} = \upg{\lambda
-  a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), which for this logic
-  corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend on the heap.
+  \item Affine objects those for which \(\ms{drop} = \upg{\lambda a. ()}{} \in
+  \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), and relevant objects those for
+  which \(\ms{split} = \upg{\lambda a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi
+  * \Phi))\), which for this logic means pairs \((A, \Phi)\) where \(\Phi\) does
+  not depend on the heap.
   \item Central morphisms to be those which do not print, diverge, \textit{or}
@@
   \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
-  \to \mc{A}\) quotiented under the equivalence relation \(\alpha: P \to \mc{A}
-  \simeq \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\)
-  from \(P\) and \(Q\), there exists a partial order isomorphism between them
-  which respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite}
-  if its carrier set is. We will write unordered pomsets using multiset
-  notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list
-  notation, e.g. \([a, a, b]\).
+  \to \mc{A}\) quotiented over the removal of finitely many \(\delta\). A pomset
+  is \textbf{finite} if its carrier set is. We will write unordered pomsets
+  using multiset notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets
+  using list notation, e.g. \([a, a, b]\).
 \end{definition}
@@
 \(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many
-predecessors. They also form a monoid under parallel composition \(\alpha ||
-\beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\)
-where \(P + Q\) is given the standard partial ordering (with elements of \(P\)
-and \(Q\) incomparable). Note in both cases the monoidal unit is \(\{\delta\}\),
-since we only consider nonempty pomsets but allow the removal of finitely many
-ticks.
-
-Given a partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n
-\in N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by
-the function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\),
+predecessors. They also form a monoid under parallel composition, defining
+\(\alpha || \beta = [\alpha, \beta]: P + Q \to \mc{A}\) where \(P + Q\) is given
+the standard partial ordering (with elements of \(P\) and \(Q\) incomparable).
+Note in both cases the monoidal unit is \(\{\delta\}\), since we only consider
+nonempty pomsets but allow the removal of finitely many ticks. Given any
+partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n \in
+N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by the
+function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\),
 where the dependent product \(\Sigma_nP_n\) is given the lexicographic order. In
@@
 \(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with semantics
-\begin{equation}
+\(
   \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
-  \qquad
+\),
+\(
   \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
-\end{equation}
+\).
 The semantics of \(\ms{read}\) in particular give a hint as to how this model
 works: rather than tracking the state of the heap, we simply \textit{emit} a
-pomset of the events that would have to happen for a given execution (in the
+pomset of the events that would have be generated by a given execution (in the
 case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
 the case of a write, the single write event \(x := v\) for a write of \(v\)),
-and later post-filter on the pomset of an entire program to get a sequentially
-consistent execution. Hence, our semantics remains compositional, allowing us to
-reason about each program fragment individually, while still allowing us to add
-arbitrary external state (e.g. another program concurrently modifying the heap)
-before finally considering the execution of an entire program.
-
-On that note, we can define the \textit{parallel execution} of two morphisms as
-follows:
+and later post-filter to obtain a set of valid executions. By doing so, our
+semantics remains compositional, allowing us to reason about each program
+fragment individually, while still allowing other program fragments to perform
+concurrent operations affecting our potential executions. On that note, we can
+define the \textit{parallel execution} of two morphisms as follows:
 \begin{equation}
@@
 
-To graduate from SC concurrency to a genuine, if maximally simple, weak memory
-model, we will to introduce a \textit{buffer} of write actions. We will
+To graduate from sequential consistency to a genuine, if maximally simple, weak
+memory model, we introduce \textit{load buffering} of write actions. We will
 implement TSO ordering by buffering all our writes, in which case they are only
@@
 pomset. At any point, we may choose to \textit{flush} some of the writes in
-buffer to the global pomset of events. In particular, we will introduce the set
-\(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write actions}, where an
-action \(\bufloc{x} := v\) by a thread denotes adding a write \(x := v\) to the
-thread's write buffer. We may then define the set of TSO actions
-\(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\); a pomset over this set
-of actions is called a \textbf{TSO pomset}. A buffer will be defined to be a
-list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret as
-linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index (with the empty list
-corresponding to \(\{\delta\}\)). In particular, we define the monad \(\ms{TSO}
-= \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\), where \(\Sigma\) is taken as a
-the stream action of finite TSO pomsets on TSO pomsets. We can view
-\(\ms{PO}\) as a submonad of this monad.
+buffer. We introduce the set \(\mc{A}_b = \{(\bufloc{x} := v)\}\) of
+\textit{buffer write actions}, where an action \(\bufloc{x} := v\) by a thread
+denotes adding a write \(x := v\) to the thread's write buffer. We may then
+define the set of TSO actions \(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup
+\mc{A}_b\); a pomset over this set is called a \textbf{TSO pomset}. A
+buffer will be defined to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\),
+which we will interpret as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by
+index (with the empty list corresponding to \(\{\delta\}\)). In particular, we
+define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
+where \(\Sigma\) is taken as a the stream action of finite TSO pomsets on TSO
+pomsets. We can view \(\ms{PO}\) as a submonad of this monad.
 

### 2023-11-17 pldi24/paper.tex
Tap

@@
 \end{equation}
-where buffer flushing is implemented via the morphism (called \(\ms{split}\) in
-\cite{sparky})
-\begin{equation}
-  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
-  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
-\end{equation}
-To be able to perform synchronization, we will also need to introduce a
-\(\ms{fence}: \mc{I}^\varnothing_0(\mb{1}, \mb{1})\) instruction, which simply
-causes all actions before the fence to be observed before any actions after the
-fence. For \(\ms{TSO}\), implementing this is as simple as flushing the buffer,
-i.e., we define
+where buffer flushing is implemented via the morphism 
+\(
+  \ms{pflush}_A = \lambda a\;L. \{(a, R, \alpha) | L = \alpha;R\}
+  : \ms{Set}_{\ms{TSO}}(A, A)
+\)
+(called \(\ms{split}\) in \cite{sparky}). To be able to perform synchronization,
+we will also need to introduce a \(\ms{fence}: \mc{I}^\varnothing_0(\mb{1},
+\mb{1})\) instruction, which simply causes all actions before the fence to be
+observed before any actions after the fence. For \(\ms{TSO}\), implementing this
+is as simple as flushing the buffer, i.e., we define
 \(
@@
 \).
-
 We can now define the parallel composition of morphisms in a "fork-join" style
@@
 the buffer}. Since both threads end up with an empty buffer, the resulting
-joined buffer is also empty, giving us the following:
+joined buffer is also empty, giving us the following definition:
 \begin{equation}

### 2023-11-17 pldi24/paper.tex
Tap

@@
       {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} 
+    \quad
+    \prftree[r]{\rle{let-blk}}
+      {\csplits{\Gamma}{\Delta}{\Xi}}
+      {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
+      {\hasty{\Xi}{0}{a}{A}{q}}
+      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} 
     \\
@@
       {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} 
-    \\
-    \prftree[r]{\rle{let-blk}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
-      {\hasty{\Xi}{0}{a}{A}{q}}
-      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} 
     \\
@@
         \(\issubst{\gamma}{\Theta}{\Gamma}\) &
-        \(\gamma\) is a substitution taking \(\Theta\) to \(\Gamma\) \\
+        \(\gamma: \ms{Var} \to \ms{Expr}\) is a substitution taking \(\Theta\) to \(\Gamma\) \\
         \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) &
-        \(\mc{L}\) is a label substitution taking \(\ms{L}\) to \(\ms{K}\)
+        \(\mc{L}: \ms{Label} \to \ms{Expr} \to \ms{Block}\) is a label substitution taking \(\ms{L}\) to \(\ms{K}\)
     \end{tabular}
@@
   \begin{gather*}
-    \boxed{\gamma: \ms{Var} \to \ms{Expr}}
-    \\
+    % \boxed{\gamma: \ms{Var} \to \ms{Expr}}
+    % \\
     \prftree[r]{\rle{subst-nil}}
@@
     \\
-    \boxed{\mc{L}: \ms{Label} \to \ms{Expr} \to \ms{Block} \text{ s.t. for fresh } x, [a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)}
-    \\
+    % \boxed{\mc{L}: \ms{Label} \to \ms{Expr} \to \ms{Block} \text{ s.t. for fresh } x, [a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)}
+    % \\
     \prftree[r]{\rle{lb-subst-nil}}
@@
     \(\ms{lin}(\Gamma, \thyp{x}{A}{q}) = \ms{lin}(\Gamma) \cap \ms{lin}(A) \cap
-    q\). 
+    q\). We assume that a substitution \(\ms{L}\) satisfies, for fresh \(x\),
+    \([a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)\).
     % We generally assume that for all but finitely many variables \(x\) and
@@
 \end{definition} 
-Note that the set of symmetric premonoidal subcategories, monoidal
-subcategories, and central subcategories of a category \(\mc{C}\) are all closed
-under intersections, forming a complete meet-semilattice.
+% Note that the symmetric premonoidal subcategories, monoidal subcategories, and
+% central subcategories of a category \(\mc{C}\) are all closed under
+% intersections, forming a complete meet-semilattice.
 %
@@
 between those morphisms that are "truly" central versus those which are just
-central "by accident". We can solve this the same way, by simply extending the
+central "by accident". We can do this the same way, by simply extending the
 notion of effectful category appropriately as follows:
@@
 \end{definition}
-We can obtain an example of a substructural effectful category quite easily by,
-considering the power-set monad above, applying the state transformer to yield
-the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to
-\mc{P}(A \times S)\). Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
-\(\upg{R}{} = \lambda a, s. R(a) \times \{s\}\) to be the standard lifting
-operator, we can verify that this satisfies the axioms of a substructural
-effectful category. Note that here the \textit{central} morphisms are precisely
-those which do not read from or write to the state, and therefore which commute
-with all other morphisms.
-
-In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a
-substructural premonoidal category, taking \(\mc{C}^{\ms{a}}\) and
-\(\mc{C}^{\ms{r}}\) to be the subcategory consisting only of the associators,
-unitors, and symmetry, and only the identity object to be affine and relevant.
-Similarly, every \textit{Cartesian} category \(\mc{C}\) can be viewed as a
-substructural Cartesian category with all morphisms and both relevant and
-affine; in fact, this is the case \textit{if and only if} a category is
-Cartesian. We also have that:
+Applying the state transformer to the power-set monad of the previous discussion
+yields the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to
+S \to \mc{P}(A \times S)\). Taking \(\mc{C}_1\) to be the original category
+\(\ms{Rel}\), where \(\upg{R}{} = \lambda a, s. R(a) \times \{s\}\) is the
+standard lifting operator, and taking \(\mc{C}_1^q = \ms{Rel}^q\), we can verify
+that this gives us the axioms of a substructural effectful category. Note that
+here the \textit{central} morphisms are precisely those which do not read from
+or write to the state, and therefore which commute with all other morphisms. In
+general,
 \begin{proposition}
@@
 \end{proposition}
+Moreover, \textit{every} premonoidal category \(\mc{C}\) can be viewed a
+substructural premonoidal category, taking \(\mc{C}^{\ms{a}}\) and
+\(\mc{C}^{\ms{r}}\) to be the subcategory consisting only of the associators,
+unitors, and symmetry, and only the identity object to be affine and relevant.
+Similarly, every \textit{Cartesian} category \(\mc{C}\) can be viewed as a
+substructural Cartesian category with all morphisms and objects both relevant
+and affine; in fact, this is the case \textit{if and only if} a category is
+Cartesian.
 

### 2023-11-17 pldi24/paper.tex
Tap
Nearly there...

@@
     % \prftree[r]{\rle{base-lin}}{q \subseteq \ms{lin}(X)}{\tlin{X}{q}} \qquad
-    \prftree[r]{\rle{unit-lin}}{}{\tlin{\mathbf{1}}{q}} 
-    \qquad
-    \prftree[r]{\rle{bool-lin}}{}{\tlin{\mathbf{2}}{q}} 
-    \qquad
-    \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} 
-    \\
+    % \prftree[r]{\rle{unit-lin}}{}{\tlin{\mathbf{1}}{q}} 
+    % \qquad
+    % \prftree[r]{\rle{bool-lin}}{}{\tlin{\mathbf{2}}{q}} 
+    % \qquad
+    % \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} 
+    % \\
     % \prftree[r]{\rle{nil-lin}}{\tlin{\cdot}{q}} \qquad
@@
       {\hasty{\Gamma}{p}{(a, b)}{A \otimes B}{q}} \\
-    \prftree[r]{\rle{unit}}
-      {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
     \prftree[r]{\rle{true}}
@@
       {\cwk{\Gamma}{\cdot}}
-      {\hasty{\Gamma}{p}{\cff}{\mb{2}}{q}} \\
+      {\hasty{\Gamma}{p}{\cff}{\mb{2}}{q}} \qquad
     \prftree[r]{\rle{let}}
@@
       \\
+    \prftree[r]{\rle{unit}}
+      {\cwk{\Gamma}{\cdot}}
+      {\hasty{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
     \prftree[r]{\rle{let2}}
@@
 \ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is taken as a stream action on finite
-program order pomsets given by sequential composition for \(\mc{A}\). This
-yields an \isotopessa model with support for concurrent read and write
-operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
-\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with semantics
+program order pomsets. This yields an \isotopessa model with support for
+concurrent read and write operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1},
+\ms{Word})\), \(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
+semantics
 \(
@@
 \).
-The semantics of \(\ms{read}\) in particular give a hint as to how this model
-works: rather than tracking the state of the heap, we simply \textit{emit} a
-pomset of the events that would have be generated by a given execution (in the
-case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
-the case of a write, the single write event \(x := v\) for a write of \(v\)),
-and later post-filter to obtain a set of valid executions. By doing so, our
-semantics remains compositional, allowing us to reason about each program
-fragment individually, while still allowing other program fragments to perform
-concurrent operations affecting our potential executions. On that note, we can
-define the \textit{parallel execution} of two morphisms as follows:
+% The semantics of \(\ms{read}\) in particular give a hint as to how this model
+% works: rather than tracking the state of the heap, we simply \textit{emit} a
+% pomset of the events that would have be generated by a given execution (in the
+% case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
+% the case of a write, the single write event \(x := v\) for a write of \(v\)),
+% and later post-filter to obtain a set of valid executions. By doing so, our
+% semantics remains compositional, allowing us to reason about each program
+% fragment individually, while still allowing other program fragments to perform
+% concurrent operations affecting our potential executions. 
+% On that note, 
+We can define the \textit{parallel execution} of two morphisms as follows:
 \begin{equation}
@@
 \).
-Note that writes at the \textit{end} of the buffer are prioritized, since later
-writes overwrite earlier ones! The semantics of reads and writes can then be
-given by
+% Note that writes at the \textit{end} of the buffer are prioritized, since
+% later writes overwrite earlier ones! 
+The semantics of reads and writes can then be given by
 \begin{equation}

### 2023-11-17 pldi24/paper.tex
Yes!

@@
   (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
-  from \(\mb{T}\). Some important examples of submonads include:
-  \begin{enumerate}
-    \item The \textbf{unit submonad} \(\eta\) of every monad \(\mb{T}\) with
-    \(\eta(A) = \{\eta_A\;a \mid a \in A\}\).
-    %; similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\). 
-    \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
-    intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
-    submonads, forming a complete meet-semilattice with bottom \(\eta\) and top
-    \(\mb{T}\),
-    % ; this induces a complete lattice in the standard manner.
-    \item In particular, the intersection of a relevant/affine/commutative
-    submonad and another submonad is relevant/affine/commutative respectively.
-  \end{enumerate}
+  from \(\mb{T}\). The \textbf{unit submonad} \(\eta\) of every monad \(\mb{T}\)
+  is given by the map \(\eta(A) = \{\eta_A\;a \mid a \in A\}\). Given submonads
+  \(\mb{S}_i\) of \(\mb{T}\), their elementwise intersections
+  \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are submonads,
+  forming a complete meet-semilattice with bottom \(\eta\) and top \(\mb{T}\).
+  In particular, the intersection of a relevant/affine/commutative submonad and
+  another submonad is relevant/affine/commutative respectively.
 \end{definition}
@@
     % on \(\nats \cup \{\infty\}\) by addition.
-    \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and
-    stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta =
-    \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action
-    compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
+    % \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and
+    % stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta =
+    % \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action
+    % compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
   \end{enumerate}
@@
   \item If \(C \subseteq Z(M)\) is submonoid of \(M\), then \(A \mapsto
-  \mc{P}^+(A \times C + \mb{0}), \mc{P}^+(A \times C \times I) \subseteq
-  \ms{Traces}\;\Theta\;A\) are submonads
+  \mc{P}^+(A \times C + \mb{0}) \subseteq \ms{Traces}\;\Theta\;A\) is a submonad.
   \item In particular, if \(M_C \subseteq Z(M)\) is a central submonoid of \(M\)
@@
 categorical setting gives us a very rich \isotopessa model supporting a basic
-notion of refinement types, while at the same time giving us a nontrivial notion
-of nonlinear types. Taking as objects pairs \((A, \Phi)\), where \(\Phi: A \to
+notion of refinement types. Taking as objects pairs \((A, \Phi)\), where \(\Phi: A \to
 \ms{Heap} \to \ms{Prop}\), we can define the category \(\ms{Sep}\) to have

### 2023-11-17 pldi24/paper.tex
Tap

@@
 Static Single Assignment, or SSA, has been a cornerstone of compiler design ever
-since its introduction by \cite{ssa-original,rosen-gvn-1988} in the late 1980s.
+since its introduction by \citet{ssa-original,rosen-gvn-1988} in the late 1980s.
 Older compiler intermediate representations permitted the same local variable to
-be assigned multiple times, whereas SSA requires each variable bindings to be
+be assigned multiple times, whereas SSA requires each variable binding to be
 immutable -- they cannot be modified after initialisation. This makes many
@@
 $\phi$-node collects together the arguments from every callsite, which lets the
-tail-calls/jumps to omit their arguments. Thus, the main difference between SSA
-and a functional program is where in the program text the arguments to a tail
-call are listed.
+tail-calls/jumps omit their arguments. Thus, the main difference between SSA and
+a functional program is where in the program text the arguments to a tail call
+are listed.
 
@@
 we can just interpret an SSA program as a corresponding functional program, and
-then rewrite the SSA programs using the equational theory of functional
-programs.
+then rewrite SSA programs using the equational theory of functional programs.
 
@@
 variables. Since this is not a generally valid program transformation, we need a
-finer classification of programs than merely pure and effectful to judge when a
-substitution is allowed.
+finer classification of programs than merely value and expression to judge when
+a substitution is allowed.
 
@@
 let y = print("world");
-let x = print("hello");
+let z = print("hello");
 ...
@@
 \end{equation}
-In this case, substituting a term for a variable changes the order the I/O
-actions are performed, which changes the program's observable behaviour. On the
-other hand, suppose we have an effect which increments a counter. Then the same
-transformation is sound, since \(x\) is used exactly once:
+In this case, substituting a term for a variable changes the order of I/O
+actions, which changes the program's observable behaviour. On the other hand,
+suppose we have an effect which increments a counter. Then in the following
+program the same transformation is sound, since \(x\) is used exactly once:
 \begin{equation}
@@
 In this case, substituting for $x$ duplicates the call to \texttt{nondet()}, and
-can change this from a program which always returns true to one which can return
-either true or false. On the other hand, if the effect is the possibility of
-nontermination, then the substitution is sound. 
+can change this from a program which always returns \texttt{true} to one which
+can return either \texttt{true} or \texttt{false}. On the other hand, if the
+effect is the possibility of nontermination, then the substitution is sound. 
 \begin{equation}
@@
 terminology, and also call such effects \emph{relevant}. With these distinction
-in place, we can \emph{define} a pure term as one which is all three of
-central, copyable, and discardable.
+in place, we can \emph{define} a pure term as one which is all three of central,
+copyable, and discardable.
 
@@
 types, which associate aliasing and ownership information with individual
-substructural values. Having aliasing information justifies important
-optimisations such as loop hoisting: 
+values. Having aliasing information justifies important optimisations such as
+loop hoisting: 
 \begin{equation}
@@
 all offer different memory models. This means that we need an abstract semantics
-for compiler IRs, which can permit a separation of concerns. This will let
-compiler writers validate their optimisations against the abstract semantics,
-and will let us separately show that a particular machine model faithfully
-implements the semantics.
+for compiler IRs, which can permit a separation of concerns. Compiler writers
+can validate their optimisations against the abstract semantics, and we can
+separately show that a particular machine model faithfully implements the
+semantics.
 
@@
 \item We give a type system for SSA-style programs, which has a full
-substitution theorem and equational theory. This language includes substructural
-types, both for controlling effects and tracking ownership in values, and our
-strong substitution principle makes formulating inlining and rewriting
-optimisations easier than in the traditional CFG-based formulations. 
+substitution theorem and equational theory for both expressions and control-flow
+graphs. This language includes substructural types, both for controlling effects
+and tracking ownership in values. 
 \item We give an axiomatic semantics for our language, by showing that it has a
-compositional intepretation in any effectful category with an Elgot structure.
-We use this structure to give generic proofs of the soundness of substitution
-and rewriting properties.
+compositional intepretation in any distributive effectful category (a
+substructural variant of a Freyd category) with an Elgot structure. We use this
+structure to give generic proofs of the soundness of substitution and rewriting
+properties.
 \item Our abstract semantics justifies a variety of powerful optimisation
@@
 \item We use these tools to demonstrate the existence of a variety of concrete
-models satisfying these categorical axioms. Starting first from simple languges
+models satisfying these categorical axioms. Starting first from simple languages
 such as state plus printing, we also show how more challenging semantics such as
@@
 In a standard SSA-based compiler, to represent a function, we begin by
-decomposing it into basic blocks, which can be viewed as \textit{label},
+decomposing it into basic blocks, which can be viewed as \textit{labels},
 followed by a sequence of \textit{instructions}, each defining a
@@
 The program in Figure \ref{fig:bb-arg} translated to our syntax is given in
-Figure \ref{fig:our-syn}. These do not introduce any new expressivity: an
+Figure \ref{fig:our-syn}. This does not introduce any new expressivity: an
 if-statement can be viewed as syntactic sugar for a conditional branch to one of
@@
 an instruction is considered \textit{central} if execution order does not
-matter, while the quantity \(q\) indicates whether it is safe to delete an
-instruction if its result is unused (in which case the instruction is
-\textit{affine}, and \(\taff \subseteq q\)) or duplicate an instruction used
-multiple times (in which case the instruction is \textit{relevant}, and \(\trel
-\subseteq q\)). An instruction which is, central, relevant, and affine is called
-\textit{pure}. Note that we assume that \(p \leq p' \and q \subseteq r \implies
-\mc{I}_p^q(A, B) \supseteq \mc{I}_{p'}^{r}(A, B)\). The set of types is
-generated by base types \(X\), the \textit{unit type} \(\mb{1}\), the
-\textit{boolean type} \(\mb{2}\), and tensor products of types \(A \otimes B\).
-We write \(\otimes\) (rather than \(\times\)) to underscore the fact that types
-are allowed to be substructural: given a linearity \(\ms{lin}(X) \subseteq
-\tint\), we can assign every type a \textit{linearity} \(\ms{lin}(A) \subseteq
-\tint\) by taking \(\ms{lin}(A \otimes B) = \ms{lin}(A) \cap \ms{lin}(B)\) and
-\(\ms{lin}(\mb{1}) = \ms{lin}(\mb{2}) = \tint\). We say a type with \(\taff
-\subseteq \ms{lin}(A)\) is \textit{affine}, while a type with \(\trel \subseteq
-\ms{lin}(A)\) is \textit{relevant}; a type which is both affine and relevant is
-called \textit{intuitionistic}. Our syntax is composed of three main syntactic
-categories:
+matter. The quantity \(q\) indicates whether it is safe to duplicate an
+instruction, or to delete it an if its result is unused. The instruction is
+\textit{affine} if \(\ms{a} \in q\), \textit{relevant}, and \(\ms{r} \in q\),
+and, if both affine and relevant, \textit{intuitionistic}. An instruction which
+is, central, relevant, and affine is called \textit{pure}. We assume that \(p
+\leq p'\) and \(q \subseteq r\) then \(\mc{I}_p^q(A, B) \supseteq
+\mc{I}_{p'}^{r}(A, B)\). The set of types is generated by base types \(X\), the
+\textit{unit type} \(\mb{1}\), the \textit{boolean type} \(\mb{2}\), and tensor
+products of types \(A \otimes B\). We write \(\otimes\) (rather than \(\times\))
+to underscore the fact that types are allowed to be substructural. For each base
+type \(X\), if we are given a linearity \(\ms{lin}(X) \subseteq \tint\), we can
+lift this to all types by taking \(\ms{lin}(A \otimes B) = \ms{lin}(A) \cap
+\ms{lin}(B)\) and \(\ms{lin}(\mb{1}) = \ms{lin}(\mb{2}) = \tint\). We say a type
+with \(\ms{a} \in \ms{lin}(A)\) is \textit{affine}, while a type with \(\ms{r}
+\in \ms{lin}(A)\) is \textit{relevant}; a type which is both affine and relevant
+is called \textit{intuitionistic}. Our syntax is composed of three main
+syntactic categories:
 \begin{itemize}
@@
   to substitute this term for a variable, and \(q\), plus the linearity of
-  \(A\), control how many times that variable can occur: if \(\taff \subseteq
-  q\) then the variable is called \textit{affine} or \textit{discardable} and it
-  is permissible to leave it unused, whereas if \(\trel \subseteq q\) it is
-  called \textit{relevant} or \textit{duplicable} and it is permissible to use
-  multiple times. As for instructions, a term which is central, relevant, and
-  affine is called \textit{pure}. The rules for this judgement are given in
-  Figure \ref{fig:ssa-term-typing}.
+  \(A\), control how many times that variable can occur: if \(\ms{a} \in q \cap
+  \ms{lin}(A)\) then the variable is called \textit{affine} or
+  \textit{discardable} and it is permissible to leave it unused, whereas if
+  \(\ms{r} \in q \cap \ms{lin}(A)\) it is called \textit{relevant} or
+  \textit{duplicable} and it is permissible to use multiple times. As for
+  instructions, a term which is central, relevant, and affine is called
+  \textit{pure}. The rules for this judgement are given in Figure
+  \ref{fig:ssa-term-typing}.
 
   \item \textit{Blocks} \(s, t\), which are typed via the judgement
-  \(\haslb{\ms{L}}{t}{\Gamma}\), which says that the block \(t\) is well-formed
+  \(\haslb{\Gamma}{t}{\ms{L}}\), which says that the block \(t\) is well-formed
   with respect to the context \(\Gamma\) and only branches to
@@
 The formal grammar for our syntax is given in Figure \ref{fig:ssa-grammar} in
-the appendix, while the typing judgements themselves are given in
+the appendix, while the typing judgements themselves are given in Figure
 \ref{fig:ssa-judgements}.
@@
 omitted, we assume \(q = \tint\)). The \textit{linearity} of a variable
-\(\ms{lin}(\thyp{x}{A}{q})\) is defined to be the intersection of the linearity
-of its type \(\ms{lin}(A)\) and its quantity \(q\). We say a variable \(x: A^q\)
-is \textit{affine} if \(\tlin{\thyp{x}{A}{q}}{\taff}\), in which case it can be
-freely ignored. Similarly, \(x\) is \textit{relevant} if
-\(\tlin{\thyp{x}{A}{q}}{\trel}\), in which case it can be used multiple times.
-This is formalized in the structural judgement
-\(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
-subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special case
-\(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
+\(\ms{lin}(\thyp{x}{A}{q})\) is defined to be \(\ms{lin}(A) \cap q\). We say a
+variable \(x: A^q\) is \textit{affine} if \(\ms{a} \in
+\ms{lin}(\thyp{x}{A}{q})\), in which case it can be freely ignored. Similarly,
+\(x\) is \textit{relevant} if \(\ms{r} \in \ms{lin}(\thyp{x}{A}{q})\), in which
+case it can be used multiple times. This is formalized in the structural
+judgement \(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be
+split into subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special
+case \(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
 typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
@@
 contexts, while the rules \rle{split-left} and \rle{split-right} send \(x: A^q\)
-to the left or right subcontext repsectively; both these rules also allow us to
+to the left or right subcontext respectively; both these rules also allow us to
 reduce the quantity \(q\) to a quantity \(r \subseteq q\), "forgetting" some of
@@
 empty label-set is a subcontext of itself. The rule \rle{join-cons} says that if
-\(\ms{L}\) is a subcontext of \(\ms{K}\) and \(\Delta\) weakens into \(\Gamma\),
-then \(\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}\),
-allowing us to "forget" any extra (affine) variables dropped by
-\(\cwk{\Gamma}{\Delta}\). Finally, the rule \rle{join-zero} simply says that if
-\(\lwk{\ms{L}}{\ms{K}}\), then \(\ms{L}\) is also a subcontext of any extension
-of \(\ms{K}\). 
+\(\ms{L}\) is a subcontext of \(\ms{K}\) and \(\Gamma\) is a weakening of
+\(\Delta\), then \(\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K},
+\llhyp{\ell}{\Delta}{A}}\), allowing us to "forget" any extra (affine) variables
+dropped by \(\cwk{\Gamma}{\Delta}\). Finally, the rule \rle{join-zero} simply
+says that if \(\lwk{\ms{L}}{\ms{K}}\), then \(\ms{L}\) is also a subcontext of
+any extension of \(\ms{K}\).
 % Note that this judgement is "backwards," in that the "larger" context is on
@@
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\ltlin{A}{q}{\trel}}
+      {\ms{r} \in \ms{lin}(\thyp{x}{A}{q})}
       {r, s \subseteq q}
@@
       {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\ltlin{A}{q}{\taff}}
+      {\ms{a} \in \ms{lin}(\thyp{x}{A}{q})}
       {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta}{\Xi}}

### 2023-11-17 pldi24/paper.tex
Tap

@@
 \begin{equation}
+  \begin{small}
 \begin{BVerbatim}[baseline=c]
@@
 \end{BVerbatim}
+  \end{small}
 \qquad \qquad \centernot\implies \qquad \qquad
+  \begin{small}
 \begin{BVerbatim}[baseline=c]
@@
 \end{BVerbatim}
+  \end{small}
 \end{equation}
@@
 \end{BVerbatim}
+  \end{small}
 \qquad \qquad \implies \qquad \qquad
+  \begin{small}
 \begin{BVerbatim}[baseline=c]
@@
 \end{BVerbatim}
+  \end{small}
 \qquad \qquad \centernot\implies \qquad \qquad
+\begin{small}
 \begin{BVerbatim}[baseline=c]
@@
 \end{BVerbatim}
+\end{small}
 \end{equation}
@@
 \begin{equation}
+\begin{small}
 \begin{BVerbatim}[baseline=c]
@@
 \end{BVerbatim}
+\end{small}
 \qquad \qquad \implies \qquad \qquad
+\begin{small}
 \begin{BVerbatim}[baseline=c]
@@
 \end{BVerbatim}
+\end{small}
 \label{eqn:hoist}
@@
   \begin{subfigure}[t]{0.33\textwidth}
+    \begin{small}
     \begin{BVerbatim}[baseline=t, gobble=3]
@@
     \end{BVerbatim}
+    \end{small}
     \caption{3-address code}
@@
   \begin{subfigure}[t]{0.32\textwidth}
+    \begin{small}
     \begin{BVerbatim}[baseline=t, gobble=3]
@@
     \end{BVerbatim}
+    \end{small}
     \caption{Basic blocks with arguments}
@@
     \end{BVerbatim}
+    \end{small}
     \caption{Our syntax}
@@
 q'\), where \(\thyp{x}{A}{q'} \in \Gamma\); allowing us to \textit{forget}
-capabilities, but not \textit{add} them. rle{app} states that, given an
+capabilities, but not \textit{add} them. \rle{app} states that, given an
 instruction \(f \in \mc{I}_p^q(A, B)\) and a \textit{central} term
@@
 \(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\) and \(\Xi\) types
-\(b\); the quantity of a pair is simply the intersection of the quantity of its
-components (expressed by allowing any quantity which types \textit{both}
-components), which must both be central. Finally, \rle{let} and \rle{let2},
-allow us to type (destructuring) let-bindings. In both cases, we split the
-context \(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is used to type
-a central term \(a\), which is bound to fresh variables appearing in a
-potentially non-central term \(e\) which may also depend on variables in
-\(\Delta\).
+\(b\), both as central terms. The quantity of a pair is simply the intersection
+of the quantity of its components (expressed by allowing any quantity which
+types \textit{both} components). Finally, \rle{let} and \rle{let2}, allow us to
+type (destructuring) let-bindings. In both cases, we split the context
+\(\Gamma\) into subcontexts \(\Delta, \Xi\), then, \(\Xi\) is used to type a
+central term \(a\), which is bound to fresh variables appearing in a potentially
+non-central term \(e\) which may also depend on variables in \(\Delta\).
 
@@
 The rules for block typing are given in Figure \ref{fig:ssa-block-typing}.
-\rle{br} says that if \(\lwk{\lhyp{\ell}{\Delta}{A}}{\ms{L}}\), \(\Gamma\)
-splits into \(\Delta\) and \(\Xi\), and \(a\) is a pure term which typechecks in
-\(\Xi\), then \(\lbrb{\ell}{a}\) targets \(\ms{L}\) given \(\Gamma\) is live on
-entry. \rle{ite} states that, if \(\Gamma\) splits into \(\Delta\) and \(\Xi\),
-\(e\) typechecks as a pure, boolean term in \(\Xi\), and that, given all
-variables in \(\Delta\) are live on entry, \(s, t\) both target \(\ms{L}\), then
-the if-statement \(\ite{e}{s}{t}\) targets \(\ms{L}\) given \(\Gamma\) is live
-on entry. \rle{let-blk} and \rle{let2-blk} are quite similar to their term
+Blocks do not have a type; instead, we judge which label-set \(\ms{L}\) a block
+can branch to given that \(\Gamma\) is live on entry. In particular, \rle{br}
+says that if \(\lwk{\lhyp{\ell}{\Delta}{A}}{\ms{L}}\), \(\Gamma\) splits into
+\(\Delta\) and \(\Xi\), and \(a\) is a pure term which typechecks in \(\Xi\),
+then \(\lbrb{\ell}{a}\) targets \(\ms{L}\) given \(\Gamma\) is live on entry.
+\rle{ite} states that, if \(\Gamma\) splits into \(\Delta\) and \(\Xi\), \(e\)
+typechecks as a pure, boolean term in \(\Xi\), and that, given all variables in
+\(\Delta\) are live on entry, \(s, t\) both target \(\ms{L}\), then the
+if-statement \(\ite{e}{s}{t}\) targets \(\ms{L}\) given \(\Gamma\) is live on
+entry. \rle{let-blk} and \rle{let2-blk} are quite similar to their term
 analogues, except that the bound term \(e\) is allowed to be impure (otherwise,
@@
 We can now state some of the basic metatheoretic properties of \isotopessa. We
-begin with weakening, which states that we can add unused \textit{affine} variables to a context,
-\textit{reduce} the quantity of an term's type, add arbitrary unused labels, and
-remove \textit{affine} variables from existing labels and still keep our typing
-judgements valid, i.e.
+begin with weakening, which states that we can add unused \textit{affine}
+variables to a context, \textit{reduce} the quantity of an term's type, add
+arbitrary unused labels, and remove \textit{affine} variables from existing
+labels and still keep our typing judgements valid, i.e.
 \begin{lemma}[Weakening] 
@@
   \caption{ Typing rules for \isotopessa substitutions. The linearity of a
-    context is defined recursively to be the intersection of the linearities and
-    quantities of its variables: \(\ms{lin}(\cdot) = \tint\),
-    \(\ms{lin}(\Gamma, \thyp{x}{A}{q}) = \ms{lin}(\Gamma) \cap \ms{lin}(A) \cap
-    q\). We assume that a substitution \(\ms{L}\) satisfies, for fresh \(x\),
+    context is defined recursively to be the intersection of the linearities of
+    its variables: \(\ms{lin}(\cdot) = \tint\), \(\ms{lin}(\Gamma,
+    \thyp{x}{A}{q}) = \ms{lin}(\Gamma) \cap \ms{lin}(\thyp{x}{A}{q})\). We
+    assume that a substitution \(\ms{L}\) satisfies, for fresh \(x\),
     \([a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)\).
@@
 is taken to be a map from variable names to expressions, takes each variable
-\(\thyp{x}{A}{q}\) in \(\Gamma\) to a \textit{central} term \(\gamma(x)\) which
-can be interpreted as having type \(A\) and quantity \(q\) in a subcontext of
-\(\Theta\) with linearity \(q\), such that computing all \(\gamma(x)\) uses up
-all non-affine variables in \(\Theta\). This is described formally via the
-typing rules \rle{subst-nil}, which says that any function \(\gamma\) takes the
-empty context to itself, and \rle{subst-cons}, which says that if \(\gamma\)
-takes \(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\),
-\(q \subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the
-duplication/dropping of substructural variables), and \(\Theta\) splits into
-\(\Theta_\Gamma\) and \(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma,
-\thyp{x}{A}{q}}\).
+\(\thyp{x}{A}{q}\) in \(\Gamma\) to a \textit{central} term \(\gamma(x)\). Due
+to linearity, the resources \(\Theta\) must be split into subcontexts
+\(\Theta_x\) which allow us to interpret each \(\gamma(x)\) as a term of type
+\(A\) with quantity \(q\). This is described formally via the typing rules
+\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
+to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
+\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
+\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/dropping of
+substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
+\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).
 
@@
 variables from \(\Gamma\) to be live which simply do not exist in \(\Theta\). To
-solve this problem, we need to introduce the concept of a
-\textit{label-substitution} \(\mc{L}\), which we define as a function that takes
-labels \(\lbl{\ell}\) and variables \(x\) to blocks \(t\) such that, for any
-fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)\). We
-introduce the judgement \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to mean that for
-each \(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have that \(\mc{L}(\lbl{\ell},
-x)\) targets \(\ms{K}\). Formally, \rle{lb-subst-nil} states that this is
-vacuously true for \(\ms{L} = \cdot\), while \rle{lb-subst-cons} says that if
+solve this problem, we need to bind the terms in \(\gamma\) to variables with
+the correct names for the live variable sets of desired targets. To do this, we
+introduce the concept of a \textit{label-substitution} \(\mc{L}\), which we
+define as a function that takes labels \(\lbl{\ell}\) and variables \(x\) to
+blocks \(t\) such that, for any fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell},
+x) = \mc{L}(\lbl{\ell}, a)\). We introduce the judgement
+\(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to mean that for each
+\(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have that \(\mc{L}(\lbl{\ell}, x)\)
+targets \(\ms{K}\). Formally, \rle{lb-subst-nil} states that this is vacuously
+true for \(\ms{L} = \cdot\), while \rle{lb-subst-cons} says that if
 \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
-\thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{\ms{L},
-\lhyp{\lbl{\ell}}{\Gamma}{A}}{\ms{K}}\). Given a context \(\gamma\) and a
+\thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{(\ms{L},
+\lhyp{\lbl{\ell}}{\Gamma}{A})}{\ms{K}}\). Given a context \(\gamma\) and a
 label-set \(\ms{L}\), we can define an associated label-substitution
@@
     that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
-    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
+    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\) and
     \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
@@
 so also allows us to add the capability of define what it means for an
-\textit{object}, which we can interpret as a type, is affine or relevant as
-well. With this aim in mind, we introduce the notion of a \textit{symmetric
-premonoidal subcategory}, as follows:
+\textit{object} to be affine or relevant as well. With this aim in mind, we
+introduce the notion of a \textit{symmetric premonoidal subcategory}, as
+follows:
 \begin{definition}[(Symmetric) Premonoidal Subcategory]
@@
 \end{definition} 
-% Note that the symmetric premonoidal subcategories, monoidal subcategories, and
-% central subcategories of a category \(\mc{C}\) are all closed under
-% intersections, forming a complete meet-semilattice.
+Note that the symmetric premonoidal subcategories, monoidal subcategories, and
+central subcategories of a category \(\mc{C}\) are all closed under
+intersections of hom-sets, forming a complete meet-semilattice.
 %
@@
 \end{definition}
+
+\paragraph{Example}
+
 A simple example of a substructural Cartesian category is the category of
@@
 \end{definition}
-Applying the state transformer to the power-set monad of the previous discussion
-yields the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to
-S \to \mc{P}(A \times S)\). Taking \(\mc{C}_1\) to be the original category
+
+\paragraph{Example}
+
+Applying the FP-style state transformer to the power-set monad yields the
+Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to
+\mc{P}(A \times S)\). Taking \(\mc{C}_1\) to be the original category
 \(\ms{Rel}\), where \(\upg{R}{} = \lambda a, s. R(a) \times \{s\}\) is the
@@
 \begin{proposition}
-  If \(\mc{C}\) is a substructural monoidal category then we can equip the
-  Kleisli category \(\mc{C}_{\mb{T}}\) with the structure of a substructural
-  effectful category by taking \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\),
+  If \(\mc{C}\) is a substructural monoidal category and \(\mb{T}\) is a strong
+  monad on \(\mc{C}\) then we can make the Kleisli category \(\mc{C}_{\mb{T}}\)
+  into a substructural effectful category \(\upg{\cdot}{}: \mc{C} \to
+  \mc{C}_{\mb{T}}\) by taking \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\),
   \(\ms{Rel}(\mc{C}_{\mb{T}}) = \ms{Rel}(\mc{C})\), and \(\mc{C}_{\mb{T}}^q(A,
   B) = \upg{\mc{C}^q(A, B)}{}\), where \(\upg{f}{} = f;\eta\), and defining
-  drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\). In
-  particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a substructural
-  effectful category.
+  drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\).
 \end{proposition}
@@
 
-To support branching, we need to add a bit more structure:
+To support branching, we need a bit more structure:
 \begin{definition}[Distributive (pre)monoidal category] 
@@
   complex coherence conditions, worked out in full for the more general case of
-  \textit{rig categories} in \cite{laplaza-distributivity}, but for our purposes
-  we will only make use of the following:
+  \textit{rig categories} by \citet{laplaza-distributivity}, but for our
+  purposes we will only make use of the following:
   \begin{equation*}
@@
   is a distributive substructural effectful category (with affine and relevant
-  objects and morphisms lifted from \(\mc{C}\))
+  objects and morphisms lifted from \(\mc{C}\)).
 \end{proposition}
@@
 \(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\). Similarly, to type a function
-application \rle{var}, we first take \(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which
+application \rle{app}, we first take \(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which
 is guaranteed to be central, lift to \(\mc{C}_p\), and then simply postcompose
@@
 \thyp{x}{A}{q}}{p}{e}{B}{q}}\) or \(\dnt{\hasty{\Delta, \thyp{x}{A}{q},
-\thyp{y}{B}{q}}{p}{e}{C}{q}}\). \
+\thyp{y}{B}{q}}{p}{e}{C}{q}}\).
 % Note that this inner term is not necessarily central, unlike the bound term!
@@
       = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{\tint}})}
+      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{\tint}})}{}
       ; \dnt{\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}} \\
     \dnt{\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}}
-      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}{}
       ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{e}{\mb{2}}{\tint}})}
@@
     \dnt{\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}}
-      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
+      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}{}
       ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A}{q}}
@@
     \dnt{\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}}
-      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
+      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}{}
       ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A \otimes B}{q}}
@@
 
-We intrpret \textit{blocks} \(\haslb{\Gamma}{t}{\ms{L}}\) as \textit{impure}
-morphisms from \(\dnt{\Gamma}\) to \(\ms{L}\): \(\Gamma\) represents all the
-variables live on entry to the block, while \(\ms{L}\) represents all valid
-targets, each annotated with live variables and an argument type; our rules are
-given in Figure \ref{fig:ssa-block-semantics}. \rle{br} is interpreted by first
-splitting the context via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\), and then
-using \(\Xi\) to compute the target's argument. The result, along the unchanged
-live variables \(\Delta\), is then passed to
+We interpret \textit{blocks} \(\haslb{\Gamma}{t}{\ms{L}}\) as \textit{impure}
+morphisms from \(\dnt{\Gamma}\) to \(\dnt{\ms{L}}\). \(\Gamma\) represents all
+the variables live on entry to the block, while \(\ms{L}\) represents all valid
+targets, each annotated with live variables and an argument type; our semantics
+are given in Figure \ref{fig:ssa-block-semantics}. \rle{br} is interpreted by
+first splitting the context via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\), and
+then using \(\Xi\) to compute the target's argument. The result, along with the
+unchanged live variables \(\Delta\), is then passed to
 \(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to obtain the desired target.
@@
 \(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}\), which yields
-either a recursive input \(\dnt{\ms{L}}\), which we pass through unchanged as a
-recursive input. In the former case, we can interpret this as either a value of
-\(\dnt{\ms{K}}\), which we pass through unchanged as output, or \(\dnt{\Gamma}
-\otimes \dnt{A}\); which we can pass as input to \(\dnt{\haslb{\Gamma,
-\thyp{x}{A}{q}}{t}{\ms{L}}}\), giving us a value of \(\ms{L}\) we can use as
-recursive input. Note that, even if \(t\) yields an output label in \(\ms{K}\),
-we still return a \textit{recursive input}, not an output; in this case, the
-\textit{next} iteration will return it as an output. We can now quite easily
-state the semantics of \rle{where}: we simply precompose the semantics of the
-entry block \(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the \textit{iterated}
-semantics of the associated control-flow graph,
-\(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger\).
+either an output \(\dnt{\ms{K}, \llhyp{\ell}{\Gamma}{A}}\) or a recursive input
+\(\dnt{\ms{L}}\), which we pass through unchanged as a recursive input. In the
+former case, we can interpret this as either a value of \(\dnt{\ms{K}}\), which
+we pass through unchanged as output, or \(\dnt{\Gamma} \otimes \dnt{A}\); which
+we can pass as input to \(\dnt{\haslb{\Gamma, \thyp{x}{A}{q}}{t}{\ms{L}}}\),
+giving us a value of \(\ms{L}\) we can use as recursive input. Note that, even
+if \(t\) yields an output label in \(\ms{K}\), we still return a
+\textit{recursive input}, not an output; in this case, the \textit{next}
+iteration will return it as an output. We can now quite easily state the
+semantics of \rle{where}: we simply precompose the semantics of the entry block
+\(\dnt{\haslb{\Gamma}{t}{\ms{L}}}\) with the \textit{iterated} semantics of the
+associated control-flow graph, \(\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger\).
 
@@
 \(\gamma(y)\) in \(\Gamma\) recursively, and using \(\dnt{\Theta_x}\) to compute
-\(\gamma(x)\). Similarly, we interpret label-substitutions as morphisms from \(\dnt{\ms{L}}\) to
-\(\dnt{\ms{K}}\) which, for each input label in \(\ms{L}\), performs a
-computation targeting \(\ms{K}\). In particular, \rle{lb-subst-nil} is
-interpreted as the zero morphism \(0_{\dnt{\ms{K}}}:
+\(\gamma(x)\). Similarly, we interpret label-substitutions as morphisms from
+\(\dnt{\ms{L}}\) to \(\dnt{\ms{K}}\) which, for each input label in \(\ms{L}\),
+performs a computation targeting \(\ms{K}\). In particular, \rle{lb-subst-nil}
+is interpreted as the zero morphism \(0_{\dnt{\ms{K}}}:
 \mc{C}_0^\varnothing(\mb{0}, \dnt{\ms{K}})\), while \rle{lb-subst-cons} takes in
@@
       ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      = \dnt{\haslb{\Theta}{\exprletsubst{\substctx{\gamma}{\Gamma}}}{\ms{K}}} 
+      = \dnt{\haslb{\Theta}{\exprletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}} 
       ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
@@
 that follow from the beta-eta theory of the language, as well as specific
-properties of model-dependent operations (such as memory operations). In the
-appendix, we state a general rewriting theorem which lifts equations between
-terms into a full congruence. This lets us justify optimisations such as scalar
-replacement of aggregates, eliminating redundant writes, reordering independent
-memory operations, and memory-to-register transformations.
+properties of model-dependent operations (such as memory operations). Because we
+have a compositional denotational semantics, equations automatically lift into a
+full congruence. This lets us justify optimisations such as scalar replacement
+of aggregates, eliminating redundant writes, reordering independent memory
+operations, and memory-to-register transformations.
 
@@
 
-In this section, we will build on the bare-bones models given in Section
-\ref{sec:semantics} to develop more sophisticated models capable of underpinning
-\isotopessa dialects supporting features such as weak memory and separation
-logic.
+In this section, we give cncrete models of the axiomatisation given in Section
+\ref{sec:semantics} to with successively more sophisticated features such as
+weak memory and separation logic.
 
@@
   \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in
-  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say
+  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\ms{bind}\;s\;f \in \mb{S}\;A\). We say
   that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and,
@@
 
-\section{Rewriting}
-
-\label{apx:rewriting}
-
-\subsection{Syntax}
-
-To properly be able to reason about congruence and congruence-based
-optimistion, we need to introduce the concept of \textit{terms with holes},
-which are defined using the grammar in Figure
-\ref{fig:blocks-with-holes-grammar}. This simply extends the three main
-syntactic categories with \textit{holes} of the form \(\lhole{X}\),
-\(\lhole{T}\), \(\lhole{W}\). To \textit{type} terms with holes, we need to
-introduce the concept of a \textit{hole-context} \(H\), which is a list of holes
-annotated as corresponding to one of:
-\begin{itemize}
-  \item A term \(\tyhole{\lhole{X}}{\Delta}{p}{A}{q}\) of type \(A\), quantity
-  \(q\), and purity \(p\) using the variables in \(\Delta\). Such a hole can be
-  filled by any term \(a\) satisfying \(\hasty{\Delta}{p}{a}{A}{q}\)
-  \item A block \(\blkhole{\lhole{T}}{\Delta}{\ms{L}}\) targeting \(\ms{L}\)
-  using the variables in \(\Delta\). Such a hole can be filled by any block
-  \(t\) satisfying \(\haslb{\Delta}{t}{\ms{L}}\).
-  \item A control-flow graph \(\cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\) with input
-  labels \(\ms{L}\) and output labels \(\ms{K}\). Such a hole can be filled by
-  any CFG satisfying \(\lhaslb{\ms{L}}{L}{\ms{K}}\).
-\end{itemize}
-
-\begin{figure}
-  \begin{grammar}
-    <\(\mhole{a}, \mhole{b}, \mhole{c}, \mhole{e}\)> ::= \(\lhole{X}\)
-    \;|\; \(x\) 
-    \;|\; \(f\;\mhole{a}\)
-    \;|\; \((\mhole{a}, \mhole{b})\) 
-    \;|\; \(()\) 
-    \;|\; \(\ctt\) 
-    \;|\; \(\cff\)
-    \;|\; \(\letexpr{x}{\mhole{a}}{\mhole{e}}\)
-    \;|\; \(\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}\)
-    % \alt \(\lbsplice{\ell}{x: A}{\mhole{t}}\)
-
-    <\(\mhole{s}, \mhole{t}\)> ::= \(\lhole{T}\) 
-    \;|\; \(\lbrb{\ell}{\mhole{a}}\) 
-    \;|\; \(\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}\)
-    \;|\; \(\letstmt{x}{\mhole{a}}{\mhole{t}}\)
-    \;|\; \(\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}\)
-    \;|\; \(\ewhere{\mhole{t}}{\mhole{L}}\)
-
-    <\(\mhole{L}\)> ::= \(\lhole{W}\) \;|\; \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{\mhole{t}}, \mhole{L}\)
-
-    <\(H\)> ::= \(\cdot\) 
-    \;|\; \(H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}\)
-    \;|\; \(H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}\)
-    \;|\; \(H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\)
-  \end{grammar}
-  \caption{Grammar for \isotopessa expressions and blocks with holes}
-  \Description{Grammar for isotope-SSA blocks with holes}
-  \label{fig:blocks-with-holes-grammar}
-\end{figure}
-
-We introduce new typing judgements for terms, blocks, and CFGs with holes in
-Figure \ref{fig:ssa-holes-judgements}. In each case, they correspond to the
-judgement that, given a collection \(\mc{H}\) terms, blocks, and CFGs to fill
-each hole in \(H\), would satisfy the corresponding judgement without holes in
-\ref{fig:ssa-judgements}.
-
-\TODO{note on rewrites}
-
-\begin{figure}
-  \begin{center}        
-    \begingroup
-    \renewcommand{\arraystretch}{1.5}
-    \setlength{\tabcolsep}{2em}
-    \begin{tabular}{rl}
-        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
-        \(\isrw{\mc{H}}{I}{H}\) & \(\mc{H}\) is a rewrite from \(I\) to \(H\),
-        i.e. fills every hole in \(H\) modulo \(I\) \\
-        \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) & Given
-        \(\isrw{\mc{H}}{\cdot}{H}\), \(\hasty{\Gamma}{p}{[\mc{H}]\mhole{a}}{A}{q}\) \\
-        \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) & Given
-        \(\isrw{\mc{H}}{\cdot}{H}\), \(\haslb{\Gamma}{[\mc{H}]\mhole{t}}{\ms{L}}\) \\
-        \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) & Given
-        \(\isrw{\mc{H}}{\cdot}{H}\),
-        \(\haslb{\Gamma}{[\mc{H}]\mhole{t}}{\ms{L}}\) \\
-    \end{tabular}
-    \endgroup
-  \end{center}
-  \caption{Typing judgements for \isotopessa terms, blocks, and CFGs with holes}
-  \Description{Typing judgements for isotope-SSA terms, blocks, and CFGs with
-  holes}
-  \label{fig:ssa-holes-judgements}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}    
-    \prftree[r]{\rle{hole-tm}}
-      {\tyhole{\lhole{X}}{\Delta}{p}{A}{q}}
-      {\cwk{\Gamma}{\Delta}}
-      {\mhasty{H}{\Gamma}{p}{\lhole{X}}{A}{q}} \qquad
-    \prftree[r]{\rle{var}}
-      {\cwk{\Gamma}{\thyp{x}{A}{q}}}
-      {\mhasty{H}{\Gamma}{p}{x}{A}{q}} \qquad
-    \prftree[r]{\rle{app}}
-      {f \in \mc{I}_p^q(A, B)}
-      {\mhasty{H}{\Gamma}{1}{\mhole{a}}{A}{q}}
-      {\mhasty{H}{\Gamma}{p}{f\;\mhole{a}}{B}{q}} \\
-    \prftree[r]{\rle{pair}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta}{1}{\mhole{a}}{A}{q}}
-      {\mhasty{H}{\Xi}{1}{\mhole{b}}{B}{q}}
-      {\mhasty{H}{\Gamma}{p}{(\mhole{a}, \mhole{b})}{A \otimes B}{q}} \\
-    \prftree[r]{\rle{unit}}
-      {\cwk{\Gamma}{\cdot}}
-      {\mhasty{H}{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
-    \prftree[r]{\rle{true}}
-      {\cwk{\Gamma}{\cdot}}
-      {\mhasty{H}{\Gamma}{p}{\ctt}{\mb{2}}{q}} \qquad
-    \prftree[r]{\rle{false}}
-      {\cwk{\Gamma}{\cdot}}
-      {\mhasty{H}{\Gamma}{p}{\cff}{\mb{2}}{q}} \\
-    \prftree[r]{\rle{let}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta, \thyp{x}{A}{}}{p}{\mhole{e}}{B}{q}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
-      {\mhasty{H}{\Gamma}{p}{\letexpr{x}{\mhole{a}}{\mhole{e}}}{B}{q}} \qquad
-    % \prftree[r]{\rle{blk}}
-    %   {\mhaslb{H}{\Gamma}{\mhole{t}}{\llhyp{\ell}{\cdot}{A}}}
-    %   {\mhasty{H}{\Gamma}{0}{\lbsplice{\ell}{A}{\mhole{t}}}{A}{\varnothing}} 
-      \\
-    \prftree[r]{\rle{let2}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{\mhole{e}}{C}{q}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
-      {\mhasty{H}{\Gamma}{p}{\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}}{C}{q}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa terms with holes}
-  \Description{Typing rules for isotope-SSA terms with holes}
-  \label{fig:ssa-term-holes-typing}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}    
-    \prftree[r]{\rle{hole-blk}}
-      {\blkhole{\lhole{T}}{\Delta}{\ms{L}} \in H}
-      {\cwk{\Gamma}{\Delta}}
-      {\mhaslb{H}{\Gamma}{\lhole{T}}{\ms{L}}} 
-      \qquad
-    \prftree[r]{\rle{br}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{\tint}}
-      {\mhaslb{H}{\Gamma}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
-      \\
-    \prftree[r]{\rle{ite}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Xi}{1}{\mhole{e}}{\mb{2}}{\tint}}
-      {\mhaslb{H}{\Delta}{\mhole{s}}{\ms{L}}}
-      {\mhaslb{H}{\Delta}{\mhole{t}}{\ms{L}}}
-      {\mhaslb{H}{\Gamma}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\
-    \prftree[r]{\rle{let-blk}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhaslb{H}{\Delta, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{q}}
-      {\mhaslb{H}{\Gamma}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
-    \prftree[r]{\rle{let2-blk}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhaslb{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{\mhole{t}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A \otimes B}{q}}
-      {\mhaslb{H}{\Gamma}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
-    \prftree[r]{\rle{where}}
-      {\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
-      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
-      {\mhaslb{H}{\Gamma}{\ewhere{\mhole{t}}{\mhole{L}}}{\ms{K}}} 
-      \qquad
-    \prftree[r]{\rle{hole-cfg}}
-      {\cfghole{\lhole{W}}{\ms{L}}{\ms{K}} \in H}
-      {\lwk{\ms{M}}{\ms{K}}}
-      {\mlhaslb{\ms{L}}{H}{\lhole{W}}{\ms{K}}}
-      \\
-    \prftree[r]{\rle{nil-br}}
-      {\mlhaslb{H}{\ms{L}}{\cdot}{\ms{L}}} \qquad
-    \prftree[r]{\rle{cons-br}}
-      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      {\mhaslb{H}{\Gamma, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
-      {\mlhaslb{H}{\ms{L}}{\mhole{L}, \lwbranch{\ell}{x: A}{\mhole{t}}}{\ms{K}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa blocks with holes}
-  \Description{Typing rules for isotope-SSA blocks with holes}
-  \label{fig:ssa-block-holes-typing}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{H: \ms{Hole} \to \ms{MExpr} \sqcup \ms{MBlock} \sqcup \ms{MCfg}}
-    \\
-    \prftree[r]{\rle{rw-nil}}
-      {\isrw{\mc{H}}{H}{\cdot}}
-    \qquad
-    \prftree[r]{\rle{rw-tm}}
-      {\isrw{\mc{H}}{I}{H}}
-      {\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}
-      {\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}
-    \\
-    \prftree[r]{\rle{rw-blk}}
-      {\isrw{\mc{H}}{I}{H}}
-      {\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}
-      {\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
-    \qquad
-    \prftree[r]{\rle{rw-cfg}}
-      {\isrw{\mc{H}}{I}{H}}
-      {\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}
-      {\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa rewrites}
-  \Description{Typing rules for isotope-SSA rewrites}
-  \label{fig:rewrite-typing}
-\end{figure}
-
-The typing rules for terms, blocks, and CFGs with holes are exactly the same as
-the corresponding rules without holes, modulo the additional rules
-\rle{hole-tm}, \rle{hole-blk}, and \rle{hole-cfg} to type the holes themselves;
-they are given in full in Figures \ref{fig:ssa-term-holes-typing} and
-\ref{fig:ssa-block-holes-typing}. This formalism allows us to state the
-\textit{syntactic rewriting} theorem, which says, in essence, that filling holes
-with valid terms yields valid terms.
-\begin{theorem}[Rewriting]
-  Given \(\isrw{\mc{H}}{H}{I}\), we have
-  \begin{itemize}
-    \item If \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), then \(\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}\)
-    \item If \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), then \(\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}\)
-  \end{itemize}
-  Furthermore, we have that 
-  \begin{itemize}
-    \item \(\mhasty{\cdot}{\Gamma}{p}{a}{A}{q} \iff \hasty{\Gamma}{p}{a}{A}{q}\)
-    \item \(\mhaslb{\cdot}{\Gamma}{t}{\ms{L}} \iff \haslb{\Gamma}{t}{\ms{L}}\)
-  \end{itemize}
-\end{theorem}
-
-\subsection{Semantics}
-
-We can interpret a hole context semantically as the \textit{set} of all possible
-tuples of morphisms which would typecheck in place of the holes in the context,
-as specified in Figure \ref{fig:holes-semantics}.
-\begin{figure}
-  \begin{align*}
-      \\
-    \dnt{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}
-      &= \dnt{H} \times \mc{C}_p^q(\dnt{\Gamma}, \dnt{A}) 
-      \\
-    \dnt{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}
-      &= \dnt{H} \times \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}}) 
-      \\
-    \dnt{H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
-      &= \dnt{H} \times \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
-      \\
-    \dnt{\cdot} 
-      &= \mb{1}
-  \end{align*}
-  \caption{Semantics for \isotopessa holes}
-  \Description{Semantics for isotope-SSA holes}
-  \label{fig:holes-semantics}
-\end{figure}
-We can now quite easily give semantics for terms, blocks, and CFGs with holes
-which simply correspond to "filling in" any holes with the appropriate morphism
-from the hole context, with all other rules interpreted identically to their
-regular variants: this is done in Figures \ref{fig:ssa-term-holes-semantics} and
-\ref{fig:ssa-block-holes-semantics}.
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}}
-      : \dnt{H} \to \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})}
-      \\
-    \sorry
-  \end{gather*}
-  \caption{Semantics for \isotopessa terms with holes}
-  \Description{Semantics for isotope-SSA terms with holes}
-  \label{fig:ssa-term-holes-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
-      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})}
-      \\
-    \sorry
-      \\  
-    \boxed{
-      \dnt{\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
-      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
-      \\
-    \sorry
-  \end{gather*}
-  \caption{Semantics for \isotopessa blocks with holes}
-  \Description{Semantics for isotope-SSA blocks with holes}
-  \label{fig:ssa-block-holes-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\isrw{\mc{H}}{I}{H}}
-      : \dnt{I} \to \dnt{H} 
-    }
-    \\
-    \dnt{\isrw{\mc{H}}{I}{\cdot}}\;i = ()
-    \qquad
-    \dnt{\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}\;i
-    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}\;i)
-    \\
-    \dnt{\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
-    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}\;i)
-    \\
-    \dnt{\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
-    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}\;i)
-  \end{gather*}
-  \caption{Semantics for \isotopessa rewrites}
-  \Description{Semantics for isotope-SSA rewrites}
-  \label{fig:rewrite-semantics}
-\end{figure}
-
-% \begin{lemma}[Rewriting Coherence] \
+% \section{Rewriting}
+
+% \label{apx:rewriting}
+
+% \subsection{Syntax}
+
+% To properly be able to reason about congruence and congruence-based
+% optimistion, we need to introduce the concept of \textit{terms with holes},
+% which are defined using the grammar in Figure
+% \ref{fig:blocks-with-holes-grammar}. This simply extends the three main
+% syntactic categories with \textit{holes} of the form \(\lhole{X}\),
+% \(\lhole{T}\), \(\lhole{W}\). To \textit{type} terms with holes, we need to
+% introduce the concept of a \textit{hole-context} \(H\), which is a list of holes
+% annotated as corresponding to one of:
 % \begin{itemize}
-%     \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-%     \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-%     \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
+%   \item A term \(\tyhole{\lhole{X}}{\Delta}{p}{A}{q}\) of type \(A\), quantity
+%   \(q\), and purity \(p\) using the variables in \(\Delta\). Such a hole can be
+%   filled by any term \(a\) satisfying \(\hasty{\Delta}{p}{a}{A}{q}\)
+%   \item A block \(\blkhole{\lhole{T}}{\Delta}{\ms{L}}\) targeting \(\ms{L}\)
+%   using the variables in \(\Delta\). Such a hole can be filled by any block
+%   \(t\) satisfying \(\haslb{\Delta}{t}{\ms{L}}\).
+%   \item A control-flow graph \(\cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\) with input
+%   labels \(\ms{L}\) and output labels \(\ms{K}\). Such a hole can be filled by
+%   any CFG satisfying \(\lhaslb{\ms{L}}{L}{\ms{K}}\).
 % \end{itemize}
-% \end{lemma}
-
-\begin{theorem}[Semantic Rewriting]
-  Given \(\isrw{\mc{H}}{I}{H}\), we have that
-  \begin{itemize}
-    \item For all \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(m \in \dnt{I}\), we have
-    \[
-      \dnt{\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}}(m)
-      = \dnt{\mhasty{H}{\Gamma}{p}{a}{A}{q}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
-    \]
-    \item For all \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(m \in \dnt{I}\), we have
-    \[
-      \dnt{\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}}(m)
-      = \dnt{\mhaslb{H}{\Gamma}{t}{\ms{L}}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
-    \]
-  \end{itemize}
-
-  Furthermore, we have that
-  \[
-    \dnt{\mhasty{\cdot}{\Gamma}{p}{a}{A}{q}}() = \dnt{\hasty{\Gamma}{p}{a}{A}{q}}
-    \qquad
-    \dnt{\mhaslb{\cdot}{\Gamma}{t}{\ms{L}}}() = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-  \]
-\end{theorem}
+
+% \begin{figure}
+%   \begin{grammar}
+%     <\(\mhole{a}, \mhole{b}, \mhole{c}, \mhole{e}\)> ::= \(\lhole{X}\)
+%     \;|\; \(x\) 
+%     \;|\; \(f\;\mhole{a}\)
+%     \;|\; \((\mhole{a}, \mhole{b})\) 
+%     \;|\; \(()\) 
+%     \;|\; \(\ctt\) 
+%     \;|\; \(\cff\)
+%     \;|\; \(\letexpr{x}{\mhole{a}}{\mhole{e}}\)
+%     \;|\; \(\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}\)
+%     % \alt \(\lbsplice{\ell}{x: A}{\mhole{t}}\)
+
+%     <\(\mhole{s}, \mhole{t}\)> ::= \(\lhole{T}\) 
+%     \;|\; \(\lbrb{\ell}{\mhole{a}}\) 
+%     \;|\; \(\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}\)
+%     \;|\; \(\letstmt{x}{\mhole{a}}{\mhole{t}}\)
+%     \;|\; \(\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}\)
+%     \;|\; \(\ewhere{\mhole{t}}{\mhole{L}}\)
+
+%     <\(\mhole{L}\)> ::= \(\lhole{W}\) \;|\; \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{\mhole{t}}, \mhole{L}\)
+
+%     <\(H\)> ::= \(\cdot\) 
+%     \;|\; \(H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}\)
+%     \;|\; \(H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}\)
+%     \;|\; \(H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\)
+%   \end{grammar}
+%   \caption{Grammar for \isotopessa expressions and blocks with holes}
+%   \Description{Grammar for isotope-SSA blocks with holes}
+%   \label{fig:blocks-with-holes-grammar}
+% \end{figure}
+
+% We introduce new typing judgements for terms, blocks, and CFGs with holes in
+% Figure \ref{fig:ssa-holes-judgements}. In each case, they correspond to the
+% judgement that, given a collection \(\mc{H}\) terms, blocks, and CFGs to fill
+% each hole in \(H\), would satisfy the corresponding judgement without holes in
+% \ref{fig:ssa-judgements}.
+
+% \TODO{note on rewrites}
+
+% \begin{figure}
+%   \begin{center}        
+%     \begingroup
+%     \renewcommand{\arraystretch}{1.5}
+%     \setlength{\tabcolsep}{2em}
+%     \begin{tabular}{rl}
+%         \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
+%         \(\isrw{\mc{H}}{I}{H}\) & \(\mc{H}\) is a rewrite from \(I\) to \(H\),
+%         i.e. fills every hole in \(H\) modulo \(I\) \\
+%         \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) & Given
+%         \(\isrw{\mc{H}}{\cdot}{H}\), \(\hasty{\Gamma}{p}{[\mc{H}]\mhole{a}}{A}{q}\) \\
+%         \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) & Given
+%         \(\isrw{\mc{H}}{\cdot}{H}\), \(\haslb{\Gamma}{[\mc{H}]\mhole{t}}{\ms{L}}\) \\
+%         \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) & Given
+%         \(\isrw{\mc{H}}{\cdot}{H}\),
+%         \(\haslb{\Gamma}{[\mc{H}]\mhole{t}}{\ms{L}}\) \\
+%     \end{tabular}
+%     \endgroup
+%   \end{center}
+%   \caption{Typing judgements for \isotopessa terms, blocks, and CFGs with holes}
+%   \Description{Typing judgements for isotope-SSA terms, blocks, and CFGs with
+%   holes}
+%   \label{fig:ssa-holes-judgements}
+% \end{figure}
+
+% \begin{figure}
+%   \begin{gather*}    
+%     \prftree[r]{\rle{hole-tm}}
+%       {\tyhole{\lhole{X}}{\Delta}{p}{A}{q}}
+%       {\cwk{\Gamma}{\Delta}}
+%       {\mhasty{H}{\Gamma}{p}{\lhole{X}}{A}{q}} \qquad
+%     \prftree[r]{\rle{var}}
+%       {\cwk{\Gamma}{\thyp{x}{A}{q}}}
+%       {\mhasty{H}{\Gamma}{p}{x}{A}{q}} \qquad
+%     \prftree[r]{\rle{app}}
+%       {f \in \mc{I}_p^q(A, B)}
+%       {\mhasty{H}{\Gamma}{1}{\mhole{a}}{A}{q}}
+%       {\mhasty{H}{\Gamma}{p}{f\;\mhole{a}}{B}{q}} \\
+%     \prftree[r]{\rle{pair}}
+%       {\csplits{\Gamma}{\Delta}{\Xi}}
+%       {\mhasty{H}{\Delta}{1}{\mhole{a}}{A}{q}}
+%       {\mhasty{H}{\Xi}{1}{\mhole{b}}{B}{q}}
+%       {\mhasty{H}{\Gamma}{p}{(\mhole{a}, \mhole{b})}{A \otimes B}{q}} \\
+%     \prftree[r]{\rle{unit}}
+%       {\cwk{\Gamma}{\cdot}}
+%       {\mhasty{H}{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
+%     \prftree[r]{\rle{true}}
+%       {\cwk{\Gamma}{\cdot}}
+%       {\mhasty{H}{\Gamma}{p}{\ctt}{\mb{2}}{q}} \qquad
+%     \prftree[r]{\rle{false}}
+%       {\cwk{\Gamma}{\cdot}}
+%       {\mhasty{H}{\Gamma}{p}{\cff}{\mb{2}}{q}} \\
+%     \prftree[r]{\rle{let}}
+%       {\csplits{\Gamma}{\Delta}{\Xi}}
+%       {\mhasty{H}{\Delta, \thyp{x}{A}{}}{p}{\mhole{e}}{B}{q}}
+%       {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
+%       {\mhasty{H}{\Gamma}{p}{\letexpr{x}{\mhole{a}}{\mhole{e}}}{B}{q}} \qquad
+%     % \prftree[r]{\rle{blk}}
+%     %   {\mhaslb{H}{\Gamma}{\mhole{t}}{\llhyp{\ell}{\cdot}{A}}}
+%     %   {\mhasty{H}{\Gamma}{0}{\lbsplice{\ell}{A}{\mhole{t}}}{A}{\varnothing}} 
+%       \\
+%     \prftree[r]{\rle{let2}}
+%       {\csplits{\Gamma}{\Delta}{\Xi}}
+%       {\mhasty{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{\mhole{e}}{C}{q}}
+%       {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
+%       {\mhasty{H}{\Gamma}{p}{\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}}{C}{q}}
+%   \end{gather*}
+%   \caption{Typing rules for \isotopessa terms with holes}
+%   \Description{Typing rules for isotope-SSA terms with holes}
+%   \label{fig:ssa-term-holes-typing}
+% \end{figure}
+
+% \begin{figure}
+%   \begin{gather*}    
+%     \prftree[r]{\rle{hole-blk}}
+%       {\blkhole{\lhole{T}}{\Delta}{\ms{L}} \in H}
+%       {\cwk{\Gamma}{\Delta}}
+%       {\mhaslb{H}{\Gamma}{\lhole{T}}{\ms{L}}} 
+%       \qquad
+%     \prftree[r]{\rle{br}}
+%       {\csplits{\Gamma}{\Delta}{\Xi}}
+%       {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
+%       {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{\tint}}
+%       {\mhaslb{H}{\Gamma}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
+%       \\
+%     \prftree[r]{\rle{ite}}
+%       {\csplits{\Gamma}{\Delta}{\Xi}}
+%       {\mhasty{H}{\Xi}{1}{\mhole{e}}{\mb{2}}{\tint}}
+%       {\mhaslb{H}{\Delta}{\mhole{s}}{\ms{L}}}
+%       {\mhaslb{H}{\Delta}{\mhole{t}}{\ms{L}}}
+%       {\mhaslb{H}{\Gamma}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\
+%     \prftree[r]{\rle{let-blk}}
+%       {\csplits{\Gamma}{\Delta}{\Xi}}
+%       {\mhaslb{H}{\Delta, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
+%       {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{q}}
+%       {\mhaslb{H}{\Gamma}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
+%     \prftree[r]{\rle{let2-blk}}
+%       {\csplits{\Gamma}{\Delta}{\Xi}}
+%       {\mhaslb{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{\mhole{t}}{\ms{L}}}
+%       {\mhasty{H}{\Xi}{0}{\mhole{a}}{A \otimes B}{q}}
+%       {\mhaslb{H}{\Gamma}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
+%     \prftree[r]{\rle{where}}
+%       {\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
+%       {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
+%       {\mhaslb{H}{\Gamma}{\ewhere{\mhole{t}}{\mhole{L}}}{\ms{K}}} 
+%       \qquad
+%     \prftree[r]{\rle{hole-cfg}}
+%       {\cfghole{\lhole{W}}{\ms{L}}{\ms{K}} \in H}
+%       {\lwk{\ms{M}}{\ms{K}}}
+%       {\mlhaslb{\ms{L}}{H}{\lhole{W}}{\ms{K}}}
+%       \\
+%     \prftree[r]{\rle{nil-br}}
+%       {\mlhaslb{H}{\ms{L}}{\cdot}{\ms{L}}} \qquad
+%     \prftree[r]{\rle{cons-br}}
+%       {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
+%       {\mhaslb{H}{\Gamma, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
+%       {\mlhaslb{H}{\ms{L}}{\mhole{L}, \lwbranch{\ell}{x: A}{\mhole{t}}}{\ms{K}}}
+%   \end{gather*}
+%   \caption{Typing rules for \isotopessa blocks with holes}
+%   \Description{Typing rules for isotope-SSA blocks with holes}
+%   \label{fig:ssa-block-holes-typing}
+% \end{figure}
+
+% \begin{figure}
+%   \begin{gather*}
+%     \boxed{H: \ms{Hole} \to \ms{MExpr} \sqcup \ms{MBlock} \sqcup \ms{MCfg}}
+%     \\
+%     \prftree[r]{\rle{rw-nil}}
+%       {\isrw{\mc{H}}{H}{\cdot}}
+%     \qquad
+%     \prftree[r]{\rle{rw-tm}}
+%       {\isrw{\mc{H}}{I}{H}}
+%       {\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}
+%       {\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}
+%     \\
+%     \prftree[r]{\rle{rw-blk}}
+%       {\isrw{\mc{H}}{I}{H}}
+%       {\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}
+%       {\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
+%     \qquad
+%     \prftree[r]{\rle{rw-cfg}}
+%       {\isrw{\mc{H}}{I}{H}}
+%       {\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}
+%       {\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
+%   \end{gather*}
+%   \caption{Typing rules for \isotopessa rewrites}
+%   \Description{Typing rules for isotope-SSA rewrites}
+%   \label{fig:rewrite-typing}
+% \end{figure}
+
+% The typing rules for terms, blocks, and CFGs with holes are exactly the same as
+% the corresponding rules without holes, modulo the additional rules
+% \rle{hole-tm}, \rle{hole-blk}, and \rle{hole-cfg} to type the holes themselves;
+% they are given in full in Figures \ref{fig:ssa-term-holes-typing} and
+% \ref{fig:ssa-block-holes-typing}. This formalism allows us to state the
+% \textit{syntactic rewriting} theorem, which says, in essence, that filling holes
+% with valid terms yields valid terms.
+% \begin{theorem}[Rewriting]
+%   Given \(\isrw{\mc{H}}{H}{I}\), we have
+%   \begin{itemize}
+%     \item If \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), then \(\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}\)
+%     \item If \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), then \(\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}\)
+%   \end{itemize}
+%   Furthermore, we have that 
+%   \begin{itemize}
+%     \item \(\mhasty{\cdot}{\Gamma}{p}{a}{A}{q} \iff \hasty{\Gamma}{p}{a}{A}{q}\)
+%     \item \(\mhaslb{\cdot}{\Gamma}{t}{\ms{L}} \iff \haslb{\Gamma}{t}{\ms{L}}\)
+%   \end{itemize}
+% \end{theorem}
+
+% \subsection{Semantics}
+
+% We can interpret a hole context semantically as the \textit{set} of all possible
+% tuples of morphisms which would typecheck in place of the holes in the context,
+% as specified in Figure \ref{fig:holes-semantics}.
+% \begin{figure}
+%   \begin{align*}
+%       \\
+%     \dnt{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}
+%       &= \dnt{H} \times \mc{C}_p^q(\dnt{\Gamma}, \dnt{A}) 
+%       \\
+%     \dnt{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}
+%       &= \dnt{H} \times \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}}) 
+%       \\
+%     \dnt{H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
+%       &= \dnt{H} \times \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
+%       \\
+%     \dnt{\cdot} 
+%       &= \mb{1}
+%   \end{align*}
+%   \caption{Semantics for \isotopessa holes}
+%   \Description{Semantics for isotope-SSA holes}
+%   \label{fig:holes-semantics}
+% \end{figure}
+% We can now quite easily give semantics for terms, blocks, and CFGs with holes
+% which simply correspond to "filling in" any holes with the appropriate morphism
+% from the hole context, with all other rules interpreted identically to their
+% regular variants: this is done in Figures \ref{fig:ssa-term-holes-semantics} and
+% \ref{fig:ssa-block-holes-semantics}.
+
+% \begin{figure}
+%   \begin{gather*}
+%     \boxed{
+%       \dnt{\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}}
+%       : \dnt{H} \to \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})}
+%       \\
+%     \sorry
+%   \end{gather*}
+%   \caption{Semantics for \isotopessa terms with holes}
+%   \Description{Semantics for isotope-SSA terms with holes}
+%   \label{fig:ssa-term-holes-semantics}
+% \end{figure}
+
+% \begin{figure}
+%   \begin{gather*}
+%     \boxed{
+%       \dnt{\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
+%       : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})}
+%       \\
+%     \sorry
+%       \\  
+%     \boxed{
+%       \dnt{\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
+%       : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
+%       \\
+%     \sorry
+%   \end{gather*}
+%   \caption{Semantics for \isotopessa blocks with holes}
+%   \Description{Semantics for isotope-SSA blocks with holes}
+%   \label{fig:ssa-block-holes-semantics}
+% \end{figure}
+
+% \begin{figure}
+%   \begin{gather*}
+%     \boxed{
+%       \dnt{\isrw{\mc{H}}{I}{H}}
+%       : \dnt{I} \to \dnt{H} 
+%     }
+%     \\
+%     \dnt{\isrw{\mc{H}}{I}{\cdot}}\;i = ()
+%     \qquad
+%     \dnt{\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}\;i
+%     = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}\;i)
+%     \\
+%     \dnt{\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
+%     = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}\;i)
+%     \\
+%     \dnt{\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
+%     = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}\;i)
+%   \end{gather*}
+%   \caption{Semantics for \isotopessa rewrites}
+%   \Description{Semantics for isotope-SSA rewrites}
+%   \label{fig:rewrite-semantics}
+% \end{figure}
+
+% % \begin{lemma}[Rewriting Coherence] \
+% %   \begin{itemize}
+% %     \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
+% %     \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+% %     \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
+% %   \end{itemize}
+% % \end{lemma}
+
+% \begin{theorem}[Semantic Rewriting]
+%   Given \(\isrw{\mc{H}}{I}{H}\), we have that
+%   \begin{itemize}
+%     \item For all \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(m \in \dnt{I}\), we have
+%     \[
+%       \dnt{\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}}(m)
+%       = \dnt{\mhasty{H}{\Gamma}{p}{a}{A}{q}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
+%     \]
+%     \item For all \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(m \in \dnt{I}\), we have
+%     \[
+%       \dnt{\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}}(m)
+%       = \dnt{\mhaslb{H}{\Gamma}{t}{\ms{L}}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
+%     \]
+%   \end{itemize}
+
+%   Furthermore, we have that
+%   \[
+%     \dnt{\mhasty{\cdot}{\Gamma}{p}{a}{A}{q}}() = \dnt{\hasty{\Gamma}{p}{a}{A}{q}}
+%     \qquad
+%     \dnt{\mhaslb{\cdot}{\Gamma}{t}{\ms{L}}}() = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+%   \]
+% \end{theorem}
 

### 2023-11-17 pldi24/paper.tex
Tap

@@
 \end{itemize}
-We can clearly see that this is a refinement type system in the sence of
-\cite{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi)
+We can clearly see that this is a refinement type system in the sense of
+\citet{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi)
 \mapsto A\) which is simply the identity on morphisms. Unfortunately, this means
@@
 \mc{A}\), where \(P + Q\) is given the lexicographic ordering and
-\(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many
+\(\ms{trim}(R)\) removes all elements of \(R\) with infinitely many
 predecessors. They also form a monoid under parallel composition, defining
@@
 fails: \((W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})\) emits the
-pomset \([x := a, y := b]\), while \(\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}}
+pomset \([x := a, y := b]\), while \((\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}}
 || \ms{id})\) emits the pomset \([y := b, x := a]\). 
@@
 will read an arbitrary value, in both cases pushing an event to the global
-pomset. At any point, we may choose to \textit{flush} some of the writes in
-buffer. We introduce the set \(\mc{A}_b = \{(\bufloc{x} := v)\}\) of
+pomset. At any point, we may choose to \textit{flush} some of the buffered
+writes. We introduce the set \(\mc{A}_b = \{(\bufloc{x} := v)\}\) of
 \textit{buffer write actions}, where an action \(\bufloc{x} := v\) by a thread
@@
 define the set of TSO actions \(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup
-\mc{A}_b\); a pomset over this set is called a \textbf{TSO pomset}. A
-buffer will be defined to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\),
-which we will interpret as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by
-index (with the empty list corresponding to \(\{\delta\}\)). In particular, we
-define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
-where \(\Sigma\) is taken as a the stream action of finite TSO pomsets on TSO
-pomsets. We can view \(\ms{PO}\) as a submonad of this monad.
+\mc{A}_b\); a pomset over this set is called a \textbf{TSO pomset}. A buffer
+will be defined to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which
+we will interpret as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index
+(with the empty list corresponding to \(\{\delta\}\)). In particular, we define
+the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\), where
+\(\Sigma\) is taken as a the stream action of finite TSO pomsets on TSO pomsets.
+We can view \(\ms{PO}\) as a submonad of this monad.
 

### 2023-11-21 pldi24/paper.tex
Tap
Uncommented rewriting appendix

--- 2023-11-17 pldi24/paper.tex
@@
   \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
-  \to \mc{A}\) quotiented over the removal of finitely many \(\delta\). A pomset
-  is \textbf{finite} if its carrier set is. We will write unordered pomsets
-  using multiset notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets
-  using list notation, e.g. \([a, a, b]\).
+  \to \mc{A}\) quotiented over the removal of \(\delta\) such that cardinality
+  is preserved. A pomset is \textbf{finite} if its carrier set is. We will write
+  unordered pomsets using multiset notation, e.g. \(\{a, a, b\}\), and linearly
+  ordered pomsets using list notation, e.g. \([a, a, b]\).
 \end{definition}
@@
 \end{figure}
-
-We introduce new typing judgements for terms, blocks, and CFGs with holes in
-Figure \ref{fig:ssa-holes-judgements}. In each case, they correspond to the
-judgement that, given a collection \(\mc{H}\) terms, blocks, and CFGs to fill
-each hole in \(H\), would satisfy the corresponding judgement without holes in
+We introduce a typing judgement \(\isrw{\mc{H}}{I}{H}\), which says that
+\(\mc{H}\) assigns to each hole in \(H\) a term satisfying the corresponding
+typing judgement within the hole context \(I\). We also introduce new typing
+judgements for terms, blocks, and CFGs with holes in Figure
+\ref{fig:ssa-holes-judgements}. In each case, they correspond to the judgement
+that, given a collection \(\mc{H}\) terms, blocks, and CFGs to fill each hole in
+\(H\), would satisfy the corresponding judgement without holes in
 \ref{fig:ssa-judgements}.
-
-\TODO{note on rewrites}
 
@@
 \end{figure}
-
-% \begin{lemma}[Rewriting Coherence] \
-%   \begin{itemize}
-%     \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-%     \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-%     \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
-%   \end{itemize}
-% \end{lemma}
-
+Of course, as for terms without holes, we have a coherence lemma:
+\begin{lemma}[Rewriting Coherence] \
+  \begin{itemize}
+    \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{lemma}
+We may finally state the semantic rewriting theorem as follows:
 \begin{theorem}[Semantic Rewriting]
@@
 \end{theorem}
-
+This justifies congruence based optimizations, and in particular E-graph rewriting.
+
+% \subsection{State-Passing Category}
+%
 % We can further use this example to get a substructural \textit{monoidal} category by considering a primitive form of separation logic: let's introduce a grammar of predicates \(\varphi, \psi ::= \top, \ms{S}, \bot\) for programs which cannot access state, can access state, and are invalid, respectively.

### 2023-11-21 pldi24/paper.tex
Tap

@@
 
-% \section{Proofs}
-
-% \subsection{Weakening and Substitution}
-
-% \begin{theorem}[Weakening] \
-%   \begin{itemize}
-%     \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
-%     \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
-%     \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then
-%     \(\haslb{\Gamma}{t}{\ms{L}}\)
-%     \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then
-%     \(\haslb{\Gamma}{t}{\ms{K}}\)
-%   \end{itemize}
-% \end{theorem}
-
-% \begin{proof} \
-%   \TODO{this}
-% \end{proof}
-
-% \begin{lemma}[Substitution Splitting] 
-%   If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
-%   then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
-%   \begin{itemize}
-%     \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
-%     \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
-%     \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
-%   \end{itemize}
-% \end{lemma}
-
-% \begin{proof} 
-%   We proceed by induction on derivations \(\csplits{\Gamma}{\Delta}{\Xi}\). In
-%   the \rle{split-nil} case, we have that \(\issubst{\gamma}{\cdot}{\cdot}\) by
-%   \rle{subst-nil}. For all other cases, since \(\issubst{\gamma}{\Theta}{\Gamma,
-%   \thyp{x}{A}{q}}\), we have by \rle{subst-cons} that
-%   \(\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}\),
-%   \(\issubst{\gamma}{\Theta_\Gamma}{\Gamma}\),
-%   \(\hasty{\Theta_\Gamma}{1}{\gamma(x)}{A}{q}\). Since
-%   \(\csplits{\Gamma}{\Delta}{\Xi}\), by induction, we have that
-%   \(\csplits{\Theta_\Gamma}{\Theta_\Delta}{\Theta_\Xi}\) where \(\Theta_\Delta\)
-%   and \(\Theta_\Xi\) have the desired properties above.
-%   \begin{itemize}
-%     \item \rle{split-left}: by left split association, we have that there exists
-%     \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\csplits{\Theta}{\Theta_{\Delta,
-%     \thyp{x}{A}{q}}}{\Theta_\Xi}\), \(\csplits{\Theta_{\Delta,
-%     \thyp{x}{A}{q}}}{\Theta_\Delta}{\Theta_x}\). Using the fact that
-%     \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), we can conclude by
-%     \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Delta,
-%     \thyp{x}{A}{q}}}{\Delta, \thyp{x}{A}{q}}\); the desired result hence
-%     follows.
-%     \item \rle{split-right}: by right split association, we have that there
-%     exists \(\Theta_{\Xi, \thyp{x}{A}{q}}\),
-%     \(\csplits{\Theta}{\Theta_\Delta}{\Theta_{\Xi, \thyp{x}{A}{q}}}\),
-%     \(\csplits{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Theta_\Xi}{\Theta_x}\). Using the
-%     fact that \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by
-%     \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
-%     \thyp{x}{A}{q}}\); the desired result hence follows.
-%     \item \rle{split-dup}: since \(\tlin{A}{\trel}\), it follows by
-%     \rle{subst-cons} that \(\tlin{\Theta_x}{\trel}\), and hence that
-%     \(\csplits{\Theta_x}{\Theta_x}{\Theta_x}\). Therefore, by split association,
-%     we have that there exist \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\Theta_{\Xi,
-%     \thyp{x}{A}{q}}\) such that \(\csplits{\Theta}{\Theta_{\Delta,
-%     \thyp{x}{A}{q}}}{\Theta_{\Xi, \thyp{x}{A}{q}}}\). Using the facts that
-%     \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\),
-%     \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by \rle{subst-cons}
-%     that \(\issubst{\gamma}{\Theta_{\Delta, \thyp{x}{A}{q}}}{\Delta,
-%     \thyp{x}{A}{q}}\), \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
-%     \thyp{x}{A}{q}}\); the desired result hence follows.
-%     \item \rle{split-drop}: since \(\tlin{A}{\taff}\), it follows by
-%     \rle{subst-cons} that \(\tlin{\Theta_x}{\taff}\), and hence that
-%     \(\cwk{\Theta_x}{\cdot}\). Therefore, by split association,
-%     \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\), giving the desired result.
-%   \end{itemize}
-% \end{proof}
-
-% \begin{lemma}[Substitution] 
-%   Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
-%   \begin{itemize}
-%     \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
-%     \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
-%     \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
-%     \begin{itemize}
-%       \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
-%       \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
-%       \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
-%     \end{itemize}
-%   \end{itemize}
-% \end{lemma}
-
-% \begin{proof} \
-%   \TODO{this}
-% \end{proof}
-
-% \begin{lemma}[Label Substitution] 
-%   Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
-%   \begin{itemize}
-%     \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
-%     \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
-%   \end{itemize}
-% \end{lemma}
-
-% \begin{proof} \
-%   \TODO{this}
-% \end{proof}
-
-% \subsection{Semantic Metatheory}
-
-% \begin{lemma}[Semantic Upgrade]
-%   For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
-%   \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
-% \end{lemma}
-
-% \begin{proof} \
-%   \TODO{this}
-% \end{proof}
-
-% \begin{lemma}[Semantic Weakening] 
-%   If \(\cwk{\Gamma}{\Delta}\), then 
-%   \begin{itemize}
-%     \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
-%     \[
-%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-%       ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
-%       = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-%     \]
-%     \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
-%     \[
-%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-%       ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
-%       = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
-%     \]
-%   \end{itemize}
-
-%   Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
-%   \[
-%     \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-%     ; \dnt{\lwk{\ms{L}}{\ms{K}}}
-%     = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
-%   \]
-% \end{lemma}
-
-% \begin{proof} \
-%   \TODO{this}
-% \end{proof}
-
-% \begin{lemma}[Substitution Coherence] \
-%   \begin{itemize}
-%     \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-%     \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-%   \end{itemize}
-% \end{lemma}
-
-% \begin{proof} \
-%   \TODO{this}
-% \end{proof}
-
-% \begin{lemma}[Semantic Substitution Splitting] 
-%   Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
-%   if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
-%   \[
-%     \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-%     = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
-%       \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
-%       \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
-%   \]
-% \end{lemma}
-
-% \begin{proof} \
-%   \TODO{this}
-% \end{proof}
-
-% \begin{theorem}[Semantic Substitution] 
-%   Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
-%   \begin{itemize}
-%     \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
-%     \[
-%       \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
-%       ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-%       = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
-%     \]
-%     \item For all 
-%       \(\haslb{\Gamma}{t}{\ms{L}}\), 
-%       \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
-%       \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
-%     \[
-%       \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
-%       ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-%       = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
-%       ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
-%     \]
-%     \item For all 
-%       \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
-%       \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
-%       \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
-%       \[
-%         \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-%         ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
-%         = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
-%         ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
-%         + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-%       \]
-%   \end{itemize}
-%   Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
-%   \begin{itemize} 
-%     \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
-%     \[
-%       \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-%       = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
-%     \]
-%     \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
-%     \[
-%       \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-%       + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
-%       = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
-%     \]
-%     where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
-%   \end{itemize}
-% \end{theorem}
-
-% \begin{proof} \
-%   \TODO{this}
-% \end{proof}
+\section{Proofs}
+
+\subsection{Weakening and Substitution}
+
+\begin{theorem}[Weakening] \
+  \begin{enumerate}
+    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
+    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
+    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then
+    \(\haslb{\Gamma}{t}{\ms{L}}\)
+    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then
+    \(\haslb{\Gamma}{t}{\ms{K}}\)
+  \end{enumerate}
+\end{theorem}
+
+\begin{proof}
+  By induction: 
+  \begin{enumerate}
+    \item
+    \item
+    \item 
+  \end{enumerate}
+\end{proof}
+
+\begin{lemma}[Substitution Splitting] 
+  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
+  \begin{itemize}
+    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
+    \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
+    \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
+  \end{itemize}
+\end{lemma}
+
+\begin{proof} 
+  We proceed by induction on derivations \(\csplits{\Gamma}{\Delta}{\Xi}\). In
+  the \rle{split-nil} case, we have that \(\issubst{\gamma}{\cdot}{\cdot}\) by
+  \rle{subst-nil}. For all other cases, since \(\issubst{\gamma}{\Theta}{\Gamma,
+  \thyp{x}{A}{q}}\), we have by \rle{subst-cons} that
+  \(\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}\),
+  \(\issubst{\gamma}{\Theta_\Gamma}{\Gamma}\),
+  \(\hasty{\Theta_\Gamma}{1}{\gamma(x)}{A}{q}\). Since
+  \(\csplits{\Gamma}{\Delta}{\Xi}\), by induction, we have that
+  \(\csplits{\Theta_\Gamma}{\Theta_\Delta}{\Theta_\Xi}\) where \(\Theta_\Delta\)
+  and \(\Theta_\Xi\) have the desired properties above.
+  \begin{itemize}
+    \item \rle{split-left}: by left split association, we have that there exists
+    \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\csplits{\Theta}{\Theta_{\Delta,
+    \thyp{x}{A}{q}}}{\Theta_\Xi}\), \(\csplits{\Theta_{\Delta,
+    \thyp{x}{A}{q}}}{\Theta_\Delta}{\Theta_x}\). Using the fact that
+    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), we can conclude by
+    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Delta,
+    \thyp{x}{A}{q}}}{\Delta, \thyp{x}{A}{q}}\); the desired result hence
+    follows.
+    \item \rle{split-right}: by right split association, we have that there
+    exists \(\Theta_{\Xi, \thyp{x}{A}{q}}\),
+    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_{\Xi, \thyp{x}{A}{q}}}\),
+    \(\csplits{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Theta_\Xi}{\Theta_x}\). Using the
+    fact that \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by
+    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
+    \thyp{x}{A}{q}}\); the desired result hence follows.
+    \item \rle{split-dup}: since \(\tlin{A}{\trel}\), it follows by
+    \rle{subst-cons} that \(\tlin{\Theta_x}{\trel}\), and hence that
+    \(\csplits{\Theta_x}{\Theta_x}{\Theta_x}\). Therefore, by split association,
+    we have that there exist \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\Theta_{\Xi,
+    \thyp{x}{A}{q}}\) such that \(\csplits{\Theta}{\Theta_{\Delta,
+    \thyp{x}{A}{q}}}{\Theta_{\Xi, \thyp{x}{A}{q}}}\). Using the facts that
+    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\),
+    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by \rle{subst-cons}
+    that \(\issubst{\gamma}{\Theta_{\Delta, \thyp{x}{A}{q}}}{\Delta,
+    \thyp{x}{A}{q}}\), \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
+    \thyp{x}{A}{q}}\); the desired result hence follows.
+    \item \rle{split-drop}: since \(\tlin{A}{\taff}\), it follows by
+    \rle{subst-cons} that \(\tlin{\Theta_x}{\taff}\), and hence that
+    \(\cwk{\Theta_x}{\cdot}\). Therefore, by split association,
+    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\), giving the desired result.
+  \end{itemize}
+\end{proof}
+
+\begin{lemma}[Substitution] 
+  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  \begin{itemize}
+    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
+    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
+    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
+    \begin{itemize}
+      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
+      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
+      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
+    \end{itemize}
+  \end{itemize}
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{lemma}[Label Substitution] 
+  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
+  \begin{itemize}
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
+    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
+  \end{itemize}
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\subsection{Semantic Metatheory}
+
+\begin{lemma}[Semantic Upgrade]
+  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
+  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{lemma}[Semantic Weakening] 
+  If \(\cwk{\Gamma}{\Delta}\), then 
+  \begin{itemize}
+    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
+    \[
+      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
+      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+    \]
+    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
+    \[
+      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
+      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
+      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
+    \]
+  \end{itemize}
+
+  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
+  \[
+    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
+    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
+  \]
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{lemma}[Substitution Coherence] \
+  \begin{itemize}
+    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
+  \end{itemize}
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{lemma}[Semantic Substitution Splitting] 
+  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
+  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
+  \[
+    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
+    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
+      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
+      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
+  \]
+\end{lemma}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
+
+\begin{theorem}[Semantic Substitution] 
+  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  \begin{itemize}
+    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
+    \[
+      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
+      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
+      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
+    \]
+    \item For all 
+      \(\haslb{\Gamma}{t}{\ms{L}}\), 
+      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
+      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
+    \[
+      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
+      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
+      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
+      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
+    \]
+    \item For all 
+      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
+      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
+      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
+      \[
+        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
+        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
+        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
+        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+      \]
+  \end{itemize}
+  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
+  \begin{itemize} 
+    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+    \[
+      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
+    \]
+    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
+    \[
+      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
+      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
+    \]
+    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
+  \end{itemize}
+\end{theorem}
+
+\begin{proof} \
+  \TODO{this}
+\end{proof}
 

### 2023-11-22 pldi24/paper.tex
Work on terms with holes semantics
Tap...

@@
 
-\begin{theorem}[Weakening] \
+\begin{lemma} We have that
+  \begin{enumerate}
+    \item Splitting commutes: if \(\csplits{\Gamma}{\Delta}{\Xi}\) then \(\csplits{\Gamma}{\Xi}{\Delta}\), and \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}} = \dnt{\csplits{\Gamma}{\Xi}{\Delta}};\sigma\)
+    \item Splitting associates: if \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\csplits{\Xi}{\Theta}{\Phi}\) then there exists \(\Gamma_{\Delta, \Theta}\) such that \(\csplits{\Gamma}{\Gamma_{\Delta, \Theta}}{\Phi}\) and \(\csplits{\Gamma_{\Delta, \Theta}}{\Delta}{\Theta}\), and, for \emph{all} such \(\Gamma_{\Delta, \Theta}\), we have \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}};\dnt{\Delta} \otimes \csplits{\Xi}{\Theta}{\Phi} = \dnt{\csplits{\Gamma}{\Gamma_{\Delta, \Theta}}{\Phi}};\dnt{\csplits{\Gamma_{\Delta, \Theta}}{\Delta}{\Theta}} \otimes \dnt{\Phi};\alpha\)
+    \item Weakening composes: if \(\cwk{\Gamma}{\Delta}\) and \(\cwk{\Delta}{\Xi}\), then \(\cwk{\Gamma}{\Xi}\)
+    \item Splitting and weakening compose on the right: if \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\cwk{\Delta}{\Delta'}\) then \(\csplits{\Gamma}{\Delta'}{\Xi}\), furthermore,
+    \(\dnt{\csplits{\Gamma}{\Delta'}{\Xi}} = \dnt{\csplits{\Gamma}{\Delta}{\Xi}};\dnt{\cwk{\Delta}{\Delta'}} \otimes \dnt{\Xi}\)
+    \item Splitting and weakening compose on the left: if \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\cwk{\Gamma'}{\Gamma}\) then \(\csplits{\Gamma'}{\Delta'}{\Xi}\), furthermore,
+    \(\dnt{\csplits{\Gamma'}{\Delta}{\Xi}} = \dnt{\cwk{\Gamma'}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\)
+  \end{enumerate}
+\end{lemma}
+
+\begin{proof}
+  By induction.
+\end{proof}
+
+\begin{lemma}[Weakening] \
   \begin{enumerate}
@@
   \end{enumerate}
-\end{theorem}
+\end{lemma}
 
 \begin{proof}
-  By induction: 
-  \begin{enumerate}
-    \item
-    \item
-    \item 
-  \end{enumerate}
+  By induction.
 \end{proof}
@@
   If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
-  \begin{itemize}
-    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
-    \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
-    \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
-  \end{itemize}
+  then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that
+  \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\),
+  \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), and
+  \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). In particular, if
+  \(\cwk{\Gamma}{\Delta}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then
+  \(\issubst{\gamma}{\Theta}{\Delta}\).
 \end{lemma}
@@
   Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
+  \begin{enumerate}
+    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) \label{lemma:substitution:term}
+    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
+    \label{lemma:substitution:block}
+    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
+    \ms{W}'\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
+    \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\), and
+    \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
+    \label{lemma:substitution:cfg}
+  \end{enumerate}
+\end{lemma}
+
+\begin{proof} 
+  We begin by proving \ref{lemma:substitution:term} by induction on derivations \(\hasty{\Gamma}{p}{a}{A}{q}\):
   \begin{itemize}
-    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
-    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
-    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
+    \item \rle{var}: we have that \(\cwk{\Gamma}{\thyp{x}{A}{q}}\), and
+    therefore that \(\issubst{\gamma}{\Theta}{\thyp{x}{A}{q}}\); by
+    \rle{subst-var}, we obtain the desired result that
+    \(\hasty{\Theta}{p}{\gamma(x)}{A}{q}\).
+    \item \rle{app}: \todo{this}
+    \item \rle{pair}: \todo{this}
+    \item \rle{true}: \todo{this}
+    \item \rle{false}: \todo{this}
+    \item \rle{let}: \todo{this}
+    \item \rle{unit}: \todo{this}
+    \item \rle{let2}: \todo{this}
+  \end{itemize}
+  Using \ref{lemma:substitution:term}; we can now prove
+  \ref{lemma:substitution:block} and \ref{lemma:substitution:cfg} by
+  mutual induction on derivations as follows:
   \begin{itemize}
-      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
-      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
-      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
+    \item \rle{br}: \todo{this}
+    \item \rle{let-blk}: \todo{this}
+    \item \rle{ite}: \todo{this}
+    \item \rle{let2-blk}: \todo{this}
+    \item \rle{where}: \todo{this}
+    \item \rle{nil-br}: \todo{this}
+    \item \rle{cons-br}: \todo{this}
   \end{itemize}
-  \end{itemize}
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
 \end{proof}
@@
 
-\begin{proof} \
-  \TODO{this}
+\begin{proof} 
+  We proceed by mutual induction on derivations as follows:
+  \begin{itemize}
+    \item \rle{br}: \todo{this}
+    \item \rle{let-blk}: \todo{this}
+    \item \rle{ite}: \todo{this}
+    \item \rle{let2-blk}: \todo{this}
+    \item \rle{where}: \todo{this}
+    \item \rle{nil-br}: \todo{this}
+    \item \rle{cons-br}: \todo{this}
+  \end{itemize}
 \end{proof}
@@
 \begin{lemma}[Semantic Upgrade]
-  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
-  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
+  If \(\hasty{\Gamma}{1}{a}{A}{q}\), we have that
+  \(\dnt{\hasty{\Gamma}{0}{a}{A}{q}} =
+  \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\)
 \end{lemma}
@@
     \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
-    \[
+    \(
       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
@@
       = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-    \]
+    \)
     \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
-    \[
+    \(
       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
@@
       = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
-    \]
+    \)
   \end{itemize}
@@
   Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
-  \[
+  \(
     \dnt{\haslb{\Gamma}{t}{\ms{L}}}
@@
     = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
-  \]
+  \)
 \end{lemma}
@@
     <\(H\)> ::= \(\cdot\) 
-    \;|\; \(H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}\)
-    \;|\; \(H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}\)
-    \;|\; \(H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\)
+    \;|\; \(M\)
+
+    <\(M\)> ::= \(\tyhole{\lhole{X}}{\Gamma}{p}{A}{q}\)
+    \;|\; \(\blkhole{\lhole{T}}{\Gamma}{\ms{L}}\)
+    \;|\; \(\cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\)
   \end{grammar}
@@
     \prftree[r]{\rle{hole-tm}}
-      {\tyhole{\lhole{X}}{\Delta}{p}{A}{q}}
+      {\tyhole{\lhole{X}}{\Delta}{p}{A}{q} \in H}
       {\cwk{\Gamma}{\Delta}}
@@
 \begin{figure}
-  \begin{align*}
-      \\
-    \dnt{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}
-      &= \dnt{H} \times \mc{C}_p^q(\dnt{\Gamma}, \dnt{A}) 
-      \\
-    \dnt{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}
-      &= \dnt{H} \times \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}}) 
-      \\
-    \dnt{H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
-      &= \dnt{H} \times \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
-      \\
-    \dnt{\cdot} 
-      &= \mb{1}
-  \end{align*}
+  \begin{gather*}
+    \dnt{\tyhole{\lhole{X}}{\Gamma}{p}{A}{q}} 
+      = \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})
+    \qquad
+    \dnt{\blkhole{\lhole{T}}{\Gamma}{\ms{L}}} 
+      = \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}})
+      \\
+    \dnt{\cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
+      = \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
+      \\
+    \dnt{H, M} = \dnt{H} \times \dnt{M}
+    \qquad
+    \dnt{\cdot} = \mb{1}
+      \\
+    \boxed{\dnt{M \in H}: \dnt{H} \to \dnt{M}}
+      \qquad
+    \dnt{M \in H, M \in H} 
+      = \pi_1
+    \qquad
+    \dnt{M \in H, N} = \pi_0;\dnt{M \in H}
+  \end{gather*}
   \caption{Semantics for \isotopessa holes}
@@
     \sorry
+      \\
+    \boxed{
+      \dnt{\tyhole{\mhole{X}}{\Gamma}{a}{A}{q} \in H}
+      : H \to \mc{C}^q_p(\dnt{\Gamma}, \dnt{A})
+    }
   \end{gather*}

### 2023-11-22 pldi24/paper.tex
Work on potential appendix structure
Tap

@@
     therefore that \(\issubst{\gamma}{\Theta}{\thyp{x}{A}{q}}\); by
-    \rle{subst-var}, we obtain the desired result that
+    \rle{subst-cons}, we obtain the desired result that
     \(\hasty{\Theta}{p}{\gamma(x)}{A}{q}\).
-    \item \rle{app}: \todo{this}
-    \item \rle{pair}: \todo{this}
-    \item \rle{true}: \todo{this}
-    \item \rle{false}: \todo{this}
-    \item \rle{let}: \todo{this}
-    \item \rle{unit}: \todo{this}
-    \item \rle{let2}: \todo{this}
+    \item \rle{app}: by induction, we have that
+    \(\hasty{\Theta}{1}{[\gamma]a}{A}{q}\), and therefore we can apply \rle{app}
+    to obtain \(\hasty{\Theta}{p}{[\gamma](f\;a)}{B}{q}\) as desired.
+    \item \rle{pair}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\) and
+    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Therefore, by induction, we have that
+    \(\hasty{\Theta_\Delta}{1}{[\gamma]a}{A}{q}\) and
+    \(\hasty{\Theta_\Xi}{1}{[\gamma]b}{B}{q}\); we can simply apply \rle{pair}
+    to obtain our desired result that \(\hasty{\Theta}{p}{[\gamma](a, b)}{A
+    \otimes B}{q}\)
+    \item \rle{true}, \rle{unit}, \rle{false}: these cases hold trivially since
+    \(\cwk{\Theta}{\cdot}\)
+    \item \rle{let}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
+    and \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Taking \(x\) to be a fresh
+    variable (with \(\gamma(x) = x\)), we can conclude by \rle{subst-cons} that
+    \(\issubst{\gamma}{\Theta_\Delta, \thyp{x}{A}{}}{\Delta, \thyp{x}{A}{}}\).
+    Therefore, by induction, we have that \(\hasty{\Theta_\Delta, \thyp{x}{A}{}}{p}{[\gamma]e}{B}{q}\) and \(\hasty{\Theta_\Xi}{1}{[\gamma]a}{A}{q}\); we
+    can hence simply apply \rle{let} to obtain
+    \(\hasty{\Theta}{p}{[\gamma]\letexpr{x}{a}{e}}{B}{q}\), as desired.
+    \item \rle{let2}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
+    and \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Taking \(x, y\) to be fresh
+    variables (with \(\gamma(x) = x\), \(\gamma(y) = y\)), we can conclude by
+    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_\Delta, \thyp{x}{A}{},
+    \thyp{y}{B}{}}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}\). Therefore, by
+    induction, we have that \(\hasty{\Theta_\Delta, \thyp{x}{A}{},
+    \thyp{y}{B}{}}{p}{[\gamma]e}{C}{q}\) and
+    \(\hasty{\Theta_\Xi}{1}{[\gamma]a}{A \otimes B}{q}\); we can hence simply
+    apply \rle{let2} to obtain
+    \(\hasty{\Theta}{p}{[\gamma]\letexpr{(x, y)}{a}{e}}{C}{q}\), as desired.
   \end{itemize}
   Using \ref{lemma:substitution:term}; we can now prove
-  \ref{lemma:substitution:block} and \ref{lemma:substitution:cfg} by
-  mutual induction on derivations as follows:
+  \ref{lemma:substitution:block} and \ref{lemma:substitution:cfg} by mutual
+  induction on derivations as follows:
   \begin{itemize}
@@
 \end{theorem}
-This justifies congruence based optimizations, and in particular E-graph rewriting.
-
-% \subsection{State-Passing Category}
+This justifies congruence based optimizations, and in particular E-graph
+rewriting.
+
+\subsection{Equational Theory}
+
+\TODO{\(\beta\), \(\eta\), etc}
+
+\section{Refinement}
+
+In general, we often want to justify optimizations which \textit{refine} a
+program by, for example, reducing the set of allowable behaviours, rather than
+simply rewrite the program to an equivalent one.
+
+\TODO{freeze example, UB}
+
+\TODO{determinism and local refinement}
+
+\subsection{Poset-Enriched Categories}
+
+\TODO{semantics of refinement via poset-enriched categories}
+
+\TODO{trivial poset-enrichment}
+
+\TODO{powerset-based poset-enrichment}
+
+\subsection{Semantic Refinement}
+
+\TODO{refinement for rewrites}
+
+\TODO{semantic refinement theorem}
+
+\subsection{Equational Theory}
+
+\TODO{this}
+
+\section{Normalisation}
+
+\TODO{this}
+
+\subsection{Strict SSA form}
+
+\TODO{this}
+
+\subsection{Structured Control-Flow}
+
+\TODO{this}
+
+\section{Loop Optimizations}
+
+\TODO{this}
+
+\subsection{Unrolling and Vectorization}
+
+\TODO{this}
+
+\subsection{Dead-code Elimination}
+
+\TODO{this}
+
+\subsection{Loop Hoisting}
+
+\TODO{this}
+
+\subsection{Loop Fusion/Fission}
+
+\TODO{this}
+
+\section{E-Graph Optimizations}
+
+\TODO{this}
+
+\subsection{Floating SSA}
+
+\TODO{this}
+
+\subsection{Whole-program rewriting}
+
+\TODO{this}
+
+% \section{State-Passing Category}
 %

### 2023-11-29 pldi24/paper.tex
Started work on resource algebra
Removed warnings

@@
 
+\author{Jad Ghalayini}
+\email{jeg74@cl.cam.ac.uk}
+\orcid{0000-0002-6905-1303}
+
 \author{Neel Krishnaswami}
@@
 \orcid{0000-0003-2838-5865}
-
-\author{Jad Ghalayini}
-\email{jeg74@cl.cam.ac.uk}
-\orcid{0000-0002-6905-1303}
 
@@
 for $x$ in the program below:
-\begin{equation}
-  \begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{equation*}
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 let x = print("hello");
@@
 \end{BVerbatim}
-  \end{small}
 \qquad \qquad \centernot\implies \qquad \qquad
-  \begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 let y = print("world");
@@
 \end{BVerbatim}
-  \end{small}
-\end{equation}
+\end{equation*}
 In this case, substituting a term for a variable changes the order of I/O
@@
 program the same transformation is sound, since \(x\) is used exactly once:
-\begin{equation}
-  \begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{equation*}
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 let x = increment(a);
@@
 \end{BVerbatim}
-  \end{small}
 \qquad \qquad \implies \qquad \qquad
-  \begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 let y = increment(b);
@@
 \end{BVerbatim}
-  \end{small}
-\end{equation}
+\end{equation*}
 F\"{u}hrmann called such commutative effects \emph{central}, which we also call
@@
 \begin{equation}
-  \begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 let x = may_loop();
@@
 \end{BVerbatim}
-  \end{small}
 \qquad \qquad \centernot\implies \qquad \qquad
-\begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 return 5
 \end{BVerbatim}
-\end{small}
 \end{equation}
@@
 \begin{equation}
-  \begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 let x = nondet();
@@
 \end{BVerbatim}
-  \end{small}
 \qquad \qquad \implies \qquad \qquad
-  \begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 return 5
 \end{BVerbatim}
-  \end{small}
 \end{equation}
@@
 \end{BVerbatim}
-  \end{small}
 \qquad \qquad \centernot\implies \qquad \qquad
-  \begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 let y = nondet();
@@
 \end{BVerbatim}
-  \end{small}
 \end{equation}
@@
 \begin{equation}
-\begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 let x = may_loop();
@@
 \end{BVerbatim}
-\end{small}
 \qquad \qquad \implies \qquad \qquad
-\begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 let y = may_loop();
@@
 \end{BVerbatim}
-\end{small}
 \end{equation}
@@
 \begin{equation}
-  \begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 while (e) {
@@
 \end{BVerbatim}
-\end{small}
 \qquad \qquad \implies \qquad \qquad
-\begin{small}
-\begin{BVerbatim}[baseline=c]
+\begin{BVerbatim}[baseline=c,fontsize=\small]
 x = load(a);
@@
 \end{BVerbatim}
-\end{small}
 \label{eqn:hoist}
@@
 program by, for example, reducing the set of allowable behaviours, rather than
-simply rewrite the program to an equivalent one.
+simply rewrite the program to an equivalent one. One place where refinement
+pops up quite often is in the treatment of undefined values and undefined 
+behaviour; for example, in a language where reading a dangling pointer is UB,
 

### 2023-11-29 pldi24/paper.tex
Tap at refinement

@@
 \subsection{Substructural Categories}
+
+\label{ssec:substruct}
 
@@
 behaviour; for example, in a language where reading a dangling pointer is UB, 
-
-\TODO{freeze example, UB}
-
-\TODO{determinism and local refinement}
+the transformation 
+\begin{equation*}
+\begin{BVerbatim}[baseline=c, fontsize=\small]
+  let x = read(p);
+  ret 5
+\end{BVerbatim}
+  \qquad \qquad \implies \qquad \qquad
+\begin{BVerbatim}[baseline=c, fontsize=\small]
+  ret 5
+\end{BVerbatim}
+\end{equation*}
+is valid, but the reverse is \textit{not}, since that would \textit{introduce},
+rather than \textit{remove}, potential UB. These optimizations are particularly
+important when operations are allowed to generate undefined values, in which
+case, even simple rewrites such as \(a \times 0 \to 0\) and \(a - a \to 0\) are
+refinements rather than equalities. In this appendix section, we will generalize
+our semantics for rewriting from section \ref{apx:rewriting} to support
+reasoning about refinements in a compositional manner.
 
@@
 
-\TODO{semantics of refinement via poset-enriched categories}
-
-\TODO{trivial poset-enrichment}
-
-\TODO{powerset-based poset-enrichment}
+To reason about refinement, we need to introduce the definition of a
+\textit{poset-enriched category}:
+\begin{definition}[Poset-Enriched Category]
+  A category \(\mc{C}\) is \textbf{poset-enriched} if, for all objects \(A, B
+  \in |\mc{C}|\), \(\mc{C}(A, B)\) is equipped with a partial order \(\leq\)
+  compatible with composition, i.e. \(f \leq g \implies f;h \leq g;h, h';f \leq
+  h';g\). A \textbf{poset-enriched premonoidal category} is a poset-enriched
+  category such that the partial order on morphisms is compatible with
+  whiskering, i.e. \(f \leq g \implies f \otimes C \leq g \otimes C, C \otimes f
+  \leq C \otimes g\). A \textbf{poset-enriched substructural category} is a
+  poset-enriched premonoidal category such that the subcategories
+  \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) are downward-closed under \(\leq\). 
+\end{definition}
+Note that every (premonoidal/substructural) category \(\mc{C}\) can be
+considered poset-enriched under the discrete partial order \((\leq) = (=)\). One
+useful example of a poset-enriched substructural category is the category of
+relations from section \ref{ssec:substruct}, taking \(f \leq g \iff \forall a
+\in A, f(a) \subseteq g(a)\).
 

### 2023-12-28 pldi24/paper.tex
Fixed typos in basic blocks with arguments, work on slides

@@
       n0 = 1
-      brz N 
+      brz i
         'exit (m0) 
@@
       i1 = sub i0 1
-      brz z1 
+      brz i1 
         'exit (m2) 

### 2024-02-24 toplas/paper.tex
Tap

--- 2023-12-28 pldi24/paper.tex
@@
 \setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
+\copyrightyear{2024}
+\acmYear{2024}
 \acmDOI{XXXXXXX.XXXXXXX}
@@
 properties.
+\item We show how our syntax quotiented by our equational theory is an initial model
+of our semantics.
 \item Our abstract semantics justifies a variety of powerful optimisation
@@
     \begin{BVerbatim}[baseline=t, gobble=3]
-    'entry:
+    ^entry:
       m = 0
@@
       brz i 'exit 'loop
-    'loop:
+    ^loop:
       t = add m n
@@
       brz i 'exit 'loop
-    'exit:
+    ^exit:
       ret m
@@
     \begin{BVerbatim}[baseline=t, gobble=3]
-    'entry:
+    ^entry:
       m0 = 0
@@
       brz i
-        'exit (m0) 
-        'loop (i, m0, n0)
-    'loop(i0, m1, n1):
+        ^exit (m0) 
+        ^loop (i, m0, n0)
+    ^loop(i0, m1, n1):
       m2 = n1
@@
       brz i1 
-        'exit (m2) 
-        'loop (i1, m2, n2)
-    'exit(m3):
+        ^exit (m2) 
+        ^loop (i1, m2, n2)
+    ^exit(m3):
       ret m3
@@
     if eq i 0 {
-      br 'exit (m0)
+      br ^exit (m0)
     } else {
-      br 'loop (i, m0, n0)
+      br ^loop (i, m0, n0)
     } where
-    'loop(i0, m1, n1) =>
+    ^loop(i0, m1, n1) =>
       let m2 = n1;
@@
       if eq i1 0 {
-        br 'exit (m2)
+        br ^exit (m2)
       } else {
-        br 'loop (i1, m2, n2)
+        br ^loop (i1, m2, n2)
       }
-    'exit(m3) => ret m3
+    ^exit(m3) => ret m3
     \end{BVerbatim}
@@
 
-  \item \textit{Blocks} \(s, t\), which are typed via the judgement
-  \(\haslb{\Gamma}{t}{\ms{L}}\), which says that the block \(t\) is well-formed
-  with respect to the context \(\Gamma\) and only branches to
+  \item \textit{Bodies} \(b\), which are typed via the judgement
+  \(\hasty{\Gamma}{p}{b}{\Delta}{}\), which says that, given variables
+  \(\Gamma\) live on entry, the variables \(\Delta\) are live on exit. The rules
+  for this judgement are given in Figure \ref{fig:ssa-block-typing}.
+
+  \item \textit{Terminators} \(s, t\), which are typed via the judgement
+  \(\haslb{\Gamma}{t}{\ms{L}}\), which says that the terminator \(t\) is
+  well-formed with respect to the context \(\Gamma\) and only branches to
   ("\textit{targets}") labels in \(\ms{L}\). The rules for this judgement are
   given in Figure \ref{fig:ssa-block-typing}.
+
+  \item \textit{Basic blocks} \(\beta\), which are typed via the judgement
+  \(\haslb{\Gamma}{\beta}{\ms{L}}\)...
 
@@
     Figure \ref{fig:ssa-block-typing}.
+
+  \item \textit{Regions} ...
 \end{itemize}

### 2024-02-24 toplas/paper.tex
Tap

@@
   \item \textit{Basic blocks} \(\beta\), which are typed via the judgement
-  \(\haslb{\Gamma}{\beta}{\ms{L}}\)...
+  \(\haslb{\Gamma}{\beta}{\ms{L}}\), which says that, given variables \(\Gamma\)
+  live on entry, \(\beta\) only branches to labels in \(\ms{L}\). The rules for
+  this judgement are given in Figure \ref{fig:ssa-block-typing}.
 
   \item \textit{Control-flow graphs} or \textit{CFG}s \(L\), which are typed via
-    the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\) defines a
-    control-flow graph with input labels \(\ms{L}\) and output labels \(\ms{K}\);
-    recursive control flow is allowed. The rules for this judgement are given in
-    Figure \ref{fig:ssa-block-typing}.
-
-  \item \textit{Regions} ...
+    the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\) defines
+    a control-flow graph with input labels \(\ms{L}\) and output labels
+    \(\ms{K}\); recursive control flow is allowed. The rules for this judgement
+    are given in Figure \ref{fig:ssa-block-typing}.
+
+  \item \textit{Regions}, which are a control-flow graph with a single
+  distinguished entry block \(\beta\), and are typed with the judgement
+  \(\haslb{\Gamma}{R}{\ms{L}}\), which says that, given variables \(\Gamma\)
+  live on entry (which can be viewed as parameters to the region), the region
+  has outputs \(\ms{L}\).
 \end{itemize}
@@
   \end{gather*}
-  \caption{Typing rules for \isotopessa blocks}
+  \caption{Typing rules for \isotopessa terminators, blocks, and regions}
   \Description{Typing rules for isotope-SSA blocks}

### 2024-03-11 syco12/paper.tex
Tap

--- 2023-12-28 pldi24/paper.tex
@@
 
-\documentclass[acmsmall,screen,review]{acmart}
+\documentclass[acmsmall,screen]{acmart}
 
@@
     \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
-    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
-    \ms{W}'\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
-    \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\), and
-    \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
+    % \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
+    % \ms{W}'\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
+    % \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\), and
+    % \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
   \end{itemize}
@@
     \]
-    \item For all 
-      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
-      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
-      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
-      \[
-        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
-        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
-        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
-        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-      \]
+    % \item For all 
+    %   \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
+    %   \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
+    %   \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
+    %   \[
+    %     \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+    %     ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
+    %     = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
+    %     ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
+    %     + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
+    %   \]
   \end{itemize}  
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
-  \begin{itemize} 
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
-    \(
-      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
-    \)
-    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
-    \[
-      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
-      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
-    \]
-    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
-  \end{itemize}
+  % Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
+  % \begin{itemize} 
+  %   \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
+  %   \(
+  %     \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+  %     = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
+  %   \)
+  %   \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
+  %   \[
+  %     \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
+  %     + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
+  %     = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
+  %   \]
+  %   where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
+  % \end{itemize}
 \end{theorem}
@@
 
-\section{Proofs}
-
-\subsection{Weakening and Substitution}
-
-\begin{lemma} We have that
-  \begin{enumerate}
-    \item Splitting commutes: if \(\csplits{\Gamma}{\Delta}{\Xi}\) then \(\csplits{\Gamma}{\Xi}{\Delta}\), and \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}} = \dnt{\csplits{\Gamma}{\Xi}{\Delta}};\sigma\)
-    \item Splitting associates: if \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\csplits{\Xi}{\Theta}{\Phi}\) then there exists \(\Gamma_{\Delta, \Theta}\) such that \(\csplits{\Gamma}{\Gamma_{\Delta, \Theta}}{\Phi}\) and \(\csplits{\Gamma_{\Delta, \Theta}}{\Delta}{\Theta}\), and, for \emph{all} such \(\Gamma_{\Delta, \Theta}\), we have \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}};\dnt{\Delta} \otimes \csplits{\Xi}{\Theta}{\Phi} = \dnt{\csplits{\Gamma}{\Gamma_{\Delta, \Theta}}{\Phi}};\dnt{\csplits{\Gamma_{\Delta, \Theta}}{\Delta}{\Theta}} \otimes \dnt{\Phi};\alpha\)
-    \item Weakening composes: if \(\cwk{\Gamma}{\Delta}\) and \(\cwk{\Delta}{\Xi}\), then \(\cwk{\Gamma}{\Xi}\)
-    \item Splitting and weakening compose on the right: if \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\cwk{\Delta}{\Delta'}\) then \(\csplits{\Gamma}{\Delta'}{\Xi}\), furthermore,
-    \(\dnt{\csplits{\Gamma}{\Delta'}{\Xi}} = \dnt{\csplits{\Gamma}{\Delta}{\Xi}};\dnt{\cwk{\Delta}{\Delta'}} \otimes \dnt{\Xi}\)
-    \item Splitting and weakening compose on the left: if \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\cwk{\Gamma'}{\Gamma}\) then \(\csplits{\Gamma'}{\Delta'}{\Xi}\), furthermore,
-    \(\dnt{\csplits{\Gamma'}{\Delta}{\Xi}} = \dnt{\cwk{\Gamma'}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\)
-  \end{enumerate}
-\end{lemma}
-
-\begin{proof}
-  By induction.
-\end{proof}
-
-\begin{lemma}[Weakening] \
-  \begin{enumerate}
-    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
-    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
-    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then
-    \(\haslb{\Gamma}{t}{\ms{L}}\)
-    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then
-    \(\haslb{\Gamma}{t}{\ms{K}}\)
-  \end{enumerate}
-\end{lemma}
-
-\begin{proof}
-  By induction.
-\end{proof}
-
-\begin{lemma}[Substitution Splitting] 
-  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that
-  \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\),
-  \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), and
-  \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). In particular, if
-  \(\cwk{\Gamma}{\Delta}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then
-  \(\issubst{\gamma}{\Theta}{\Delta}\).
-\end{lemma}
-
-\begin{proof} 
-  We proceed by induction on derivations \(\csplits{\Gamma}{\Delta}{\Xi}\). In
-  the \rle{split-nil} case, we have that \(\issubst{\gamma}{\cdot}{\cdot}\) by
-  \rle{subst-nil}. For all other cases, since \(\issubst{\gamma}{\Theta}{\Gamma,
-  \thyp{x}{A}{q}}\), we have by \rle{subst-cons} that
-  \(\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}\),
-  \(\issubst{\gamma}{\Theta_\Gamma}{\Gamma}\),
-  \(\hasty{\Theta_\Gamma}{1}{\gamma(x)}{A}{q}\). Since
-  \(\csplits{\Gamma}{\Delta}{\Xi}\), by induction, we have that
-  \(\csplits{\Theta_\Gamma}{\Theta_\Delta}{\Theta_\Xi}\) where \(\Theta_\Delta\)
-  and \(\Theta_\Xi\) have the desired properties above.
-  \begin{itemize}
-    \item \rle{split-left}: by left split association, we have that there exists
-    \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\csplits{\Theta}{\Theta_{\Delta,
-    \thyp{x}{A}{q}}}{\Theta_\Xi}\), \(\csplits{\Theta_{\Delta,
-    \thyp{x}{A}{q}}}{\Theta_\Delta}{\Theta_x}\). Using the fact that
-    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), we can conclude by
-    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Delta,
-    \thyp{x}{A}{q}}}{\Delta, \thyp{x}{A}{q}}\); the desired result hence
-    follows.
-    \item \rle{split-right}: by right split association, we have that there
-    exists \(\Theta_{\Xi, \thyp{x}{A}{q}}\),
-    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_{\Xi, \thyp{x}{A}{q}}}\),
-    \(\csplits{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Theta_\Xi}{\Theta_x}\). Using the
-    fact that \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by
-    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
-    \thyp{x}{A}{q}}\); the desired result hence follows.
-    \item \rle{split-dup}: since \(\tlin{A}{\trel}\), it follows by
-    \rle{subst-cons} that \(\tlin{\Theta_x}{\trel}\), and hence that
-    \(\csplits{\Theta_x}{\Theta_x}{\Theta_x}\). Therefore, by split association,
-    we have that there exist \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\Theta_{\Xi,
-    \thyp{x}{A}{q}}\) such that \(\csplits{\Theta}{\Theta_{\Delta,
-    \thyp{x}{A}{q}}}{\Theta_{\Xi, \thyp{x}{A}{q}}}\). Using the facts that
-    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\),
-    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by \rle{subst-cons}
-    that \(\issubst{\gamma}{\Theta_{\Delta, \thyp{x}{A}{q}}}{\Delta,
-    \thyp{x}{A}{q}}\), \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
-    \thyp{x}{A}{q}}\); the desired result hence follows.
-    \item \rle{split-drop}: since \(\tlin{A}{\taff}\), it follows by
-    \rle{subst-cons} that \(\tlin{\Theta_x}{\taff}\), and hence that
-    \(\cwk{\Theta_x}{\cdot}\). Therefore, by split association,
-    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\), giving the desired result.
-  \end{itemize}
-\end{proof}
-
-\begin{lemma}[Substitution] 
-  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  \begin{enumerate}
-    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) \label{lemma:substitution:term}
-    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
-    \label{lemma:substitution:block}
-    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
-    \ms{W}'\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
-    \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\), and
-    \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
-    \label{lemma:substitution:cfg}
-  \end{enumerate}
-\end{lemma}
-
-\begin{proof} 
-  We begin by proving \ref{lemma:substitution:term} by induction on derivations \(\hasty{\Gamma}{p}{a}{A}{q}\):
-  \begin{itemize}
-    \item \rle{var}: we have that \(\cwk{\Gamma}{\thyp{x}{A}{q}}\), and
-    therefore that \(\issubst{\gamma}{\Theta}{\thyp{x}{A}{q}}\); by
-    \rle{subst-cons}, we obtain the desired result that
-    \(\hasty{\Theta}{p}{\gamma(x)}{A}{q}\).
-    \item \rle{app}: by induction, we have that
-    \(\hasty{\Theta}{1}{[\gamma]a}{A}{q}\), and therefore we can apply \rle{app}
-    to obtain \(\hasty{\Theta}{p}{[\gamma](f\;a)}{B}{q}\) as desired.
-    \item \rle{pair}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\) and
-    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Therefore, by induction, we have that
-    \(\hasty{\Theta_\Delta}{1}{[\gamma]a}{A}{q}\) and
-    \(\hasty{\Theta_\Xi}{1}{[\gamma]b}{B}{q}\); we can simply apply \rle{pair}
-    to obtain our desired result that \(\hasty{\Theta}{p}{[\gamma](a, b)}{A
-    \otimes B}{q}\)
-    \item \rle{true}, \rle{unit}, \rle{false}: these cases hold trivially since
-    \(\cwk{\Theta}{\cdot}\)
-    \item \rle{let}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
-    and \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Taking \(x\) to be a fresh
-    variable (with \(\gamma(x) = x\)), we can conclude by \rle{subst-cons} that
-    \(\issubst{\gamma}{\Theta_\Delta, \thyp{x}{A}{}}{\Delta, \thyp{x}{A}{}}\).
-    Therefore, by induction, we have that \(\hasty{\Theta_\Delta, \thyp{x}{A}{}}{p}{[\gamma]e}{B}{q}\) and \(\hasty{\Theta_\Xi}{1}{[\gamma]a}{A}{q}\); we
-    can hence simply apply \rle{let} to obtain
-    \(\hasty{\Theta}{p}{[\gamma]\letexpr{x}{a}{e}}{B}{q}\), as desired.
-    \item \rle{let2}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
-    and \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Taking \(x, y\) to be fresh
-    variables (with \(\gamma(x) = x\), \(\gamma(y) = y\)), we can conclude by
-    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_\Delta, \thyp{x}{A}{},
-    \thyp{y}{B}{}}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}\). Therefore, by
-    induction, we have that \(\hasty{\Theta_\Delta, \thyp{x}{A}{},
-    \thyp{y}{B}{}}{p}{[\gamma]e}{C}{q}\) and
-    \(\hasty{\Theta_\Xi}{1}{[\gamma]a}{A \otimes B}{q}\); we can hence simply
-    apply \rle{let2} to obtain
-    \(\hasty{\Theta}{p}{[\gamma]\letexpr{(x, y)}{a}{e}}{C}{q}\), as desired.
-  \end{itemize}
-  Using \ref{lemma:substitution:term}; we can now prove
-  \ref{lemma:substitution:block} and \ref{lemma:substitution:cfg} by mutual
-  induction on derivations as follows:
-  \begin{itemize}
-    \item \rle{br}: \todo{this}
-    \item \rle{let-blk}: \todo{this}
-    \item \rle{ite}: \todo{this}
-    \item \rle{let2-blk}: \todo{this}
-    \item \rle{where}: \todo{this}
-    \item \rle{nil-br}: \todo{this}
-    \item \rle{cons-br}: \todo{this}
-  \end{itemize}
-\end{proof}
-
-\begin{lemma}[Label Substitution] 
-  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
-  \begin{itemize}
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
-    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
-  \end{itemize}
-\end{lemma}
-
-\begin{proof} 
-  We proceed by mutual induction on derivations as follows:
-  \begin{itemize}
-    \item \rle{br}: \todo{this}
-    \item \rle{let-blk}: \todo{this}
-    \item \rle{ite}: \todo{this}
-    \item \rle{let2-blk}: \todo{this}
-    \item \rle{where}: \todo{this}
-    \item \rle{nil-br}: \todo{this}
-    \item \rle{cons-br}: \todo{this}
-  \end{itemize}
-\end{proof}
-
-\subsection{Semantic Metatheory}
-
-\begin{lemma}[Semantic Upgrade]
-  If \(\hasty{\Gamma}{1}{a}{A}{q}\), we have that
-  \(\dnt{\hasty{\Gamma}{0}{a}{A}{q}} =
-  \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\)
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{lemma}[Semantic Weakening] 
-  If \(\cwk{\Gamma}{\Delta}\), then 
-  \begin{itemize}
-    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
-    \(
-      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
-      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-    \)
-    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
-    \(
-      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
-      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
-      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
-    \)
-  \end{itemize}
-
-  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
-  \(
-    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
-    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
-  \)
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{lemma}[Substitution Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{lemma}[Semantic Substitution Splitting] 
-  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
-  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
-  \[
-    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
-    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
-      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
-      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
-  \]
-\end{lemma}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\begin{theorem}[Semantic Substitution] 
-  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
-  \begin{itemize}
-    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
-    \[
-      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
-      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
-      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
-    \]
-    \item For all 
-      \(\haslb{\Gamma}{t}{\ms{L}}\), 
-      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
-      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
-    \[
-      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
-      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
-      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
-    \]
-    \item For all 
-      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
-      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
-      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
-      \[
-        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
-        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
-        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
-        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
-      \]
-  \end{itemize}
-  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
-  \begin{itemize} 
-    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
-    \[
-      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
-    \]
-    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
-    \[
-      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
-      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
-      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
-    \]
-    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
-  \end{itemize}
-\end{theorem}
-
-\begin{proof} \
-  \TODO{this}
-\end{proof}
-
-\section{Rewriting}
-
-\label{apx:rewriting}
-
-\subsection{Syntax}
-
-To properly be able to reason about congruence and congruence-based
-optimistion, we need to introduce the concept of \textit{terms with holes},
-which are defined using the grammar in Figure
-\ref{fig:blocks-with-holes-grammar}. This simply extends the three main
-syntactic categories with \textit{holes} of the form \(\lhole{X}\),
-\(\lhole{T}\), \(\lhole{W}\). To \textit{type} terms with holes, we need to
-introduce the concept of a \textit{hole-context} \(H\), which is a list of holes
-annotated as corresponding to one of:
-\begin{itemize}
-  \item A term \(\tyhole{\lhole{X}}{\Delta}{p}{A}{q}\) of type \(A\), quantity
-  \(q\), and purity \(p\) using the variables in \(\Delta\). Such a hole can be
-  filled by any term \(a\) satisfying \(\hasty{\Delta}{p}{a}{A}{q}\)
-  \item A block \(\blkhole{\lhole{T}}{\Delta}{\ms{L}}\) targeting \(\ms{L}\)
-  using the variables in \(\Delta\). Such a hole can be filled by any block
-  \(t\) satisfying \(\haslb{\Delta}{t}{\ms{L}}\).
-  \item A control-flow graph \(\cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\) with input
-  labels \(\ms{L}\) and output labels \(\ms{K}\). Such a hole can be filled by
-  any CFG satisfying \(\lhaslb{\ms{L}}{L}{\ms{K}}\).
-\end{itemize}
-
-\begin{figure}
-  \begin{grammar}
-    <\(\mhole{a}, \mhole{b}, \mhole{c}, \mhole{e}\)> ::= \(\lhole{X}\)
-    \;|\; \(x\) 
-    \;|\; \(f\;\mhole{a}\)
-    \;|\; \((\mhole{a}, \mhole{b})\) 
-    \;|\; \(()\) 
-    \;|\; \(\ctt\) 
-    \;|\; \(\cff\)
-    \;|\; \(\letexpr{x}{\mhole{a}}{\mhole{e}}\)
-    \;|\; \(\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}\)
-    % \alt \(\lbsplice{\ell}{x: A}{\mhole{t}}\)
-
-    <\(\mhole{s}, \mhole{t}\)> ::= \(\lhole{T}\) 
-    \;|\; \(\lbrb{\ell}{\mhole{a}}\) 
-    \;|\; \(\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}\)
-    \;|\; \(\letstmt{x}{\mhole{a}}{\mhole{t}}\)
-    \;|\; \(\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}\)
-    \;|\; \(\ewhere{\mhole{t}}{\mhole{L}}\)
-
-    <\(\mhole{L}\)> ::= \(\lhole{W}\) \;|\; \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{\mhole{t}}, \mhole{L}\)
-
-    <\(H\)> ::= \(\cdot\) 
-    \;|\; \(M\)
-
-    <\(M\)> ::= \(\tyhole{\lhole{X}}{\Gamma}{p}{A}{q}\)
-    \;|\; \(\blkhole{\lhole{T}}{\Gamma}{\ms{L}}\)
-    \;|\; \(\cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\)
-  \end{grammar}
-  \caption{Grammar for \isotopessa expressions and blocks with holes}
-  \Description{Grammar for isotope-SSA blocks with holes}
-  \label{fig:blocks-with-holes-grammar}
-\end{figure}
-We introduce a typing judgement \(\isrw{\mc{H}}{I}{H}\), which says that
-\(\mc{H}\) assigns to each hole in \(H\) a term satisfying the corresponding
-typing judgement within the hole context \(I\). We also introduce new typing
-judgements for terms, blocks, and CFGs with holes in Figure
-\ref{fig:ssa-holes-judgements}. In each case, they correspond to the judgement
-that, given a collection \(\mc{H}\) terms, blocks, and CFGs to fill each hole in
-\(H\), would satisfy the corresponding judgement without holes in
-\ref{fig:ssa-judgements}.
-
-\begin{figure}
-  \begin{center}        
-    \begingroup
-    \renewcommand{\arraystretch}{1.5}
-    \setlength{\tabcolsep}{2em}
-    \begin{tabular}{rl}
-        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
-        \(\isrw{\mc{H}}{I}{H}\) & \(\mc{H}\) is a rewrite from \(I\) to \(H\),
-        i.e. fills every hole in \(H\) modulo \(I\) \\
-        \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) & Given
-        \(\isrw{\mc{H}}{\cdot}{H}\), \(\hasty{\Gamma}{p}{[\mc{H}]\mhole{a}}{A}{q}\) \\
-        \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) & Given
-        \(\isrw{\mc{H}}{\cdot}{H}\), \(\haslb{\Gamma}{[\mc{H}]\mhole{t}}{\ms{L}}\) \\
-        \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) & Given
-        \(\isrw{\mc{H}}{\cdot}{H}\),
-        \(\haslb{\Gamma}{[\mc{H}]\mhole{t}}{\ms{L}}\) \\
-    \end{tabular}
-    \endgroup
-  \end{center}
-  \caption{Typing judgements for \isotopessa terms, blocks, and CFGs with holes}
-  \Description{Typing judgements for isotope-SSA terms, blocks, and CFGs with
-  holes}
-  \label{fig:ssa-holes-judgements}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}    
-    \prftree[r]{\rle{hole-tm}}
-      {\tyhole{\lhole{X}}{\Delta}{p}{A}{q} \in H}
-      {\cwk{\Gamma}{\Delta}}
-      {\mhasty{H}{\Gamma}{p}{\lhole{X}}{A}{q}} \qquad
-    \prftree[r]{\rle{var}}
-      {\cwk{\Gamma}{\thyp{x}{A}{q}}}
-      {\mhasty{H}{\Gamma}{p}{x}{A}{q}} \qquad
-    \prftree[r]{\rle{app}}
-      {f \in \mc{I}_p^q(A, B)}
-      {\mhasty{H}{\Gamma}{1}{\mhole{a}}{A}{q}}
-      {\mhasty{H}{\Gamma}{p}{f\;\mhole{a}}{B}{q}} \\
-    \prftree[r]{\rle{pair}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta}{1}{\mhole{a}}{A}{q}}
-      {\mhasty{H}{\Xi}{1}{\mhole{b}}{B}{q}}
-      {\mhasty{H}{\Gamma}{p}{(\mhole{a}, \mhole{b})}{A \otimes B}{q}} \\
-    \prftree[r]{\rle{unit}}
-      {\cwk{\Gamma}{\cdot}}
-      {\mhasty{H}{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
-    \prftree[r]{\rle{true}}
-      {\cwk{\Gamma}{\cdot}}
-      {\mhasty{H}{\Gamma}{p}{\ctt}{\mb{2}}{q}} \qquad
-    \prftree[r]{\rle{false}}
-      {\cwk{\Gamma}{\cdot}}
-      {\mhasty{H}{\Gamma}{p}{\cff}{\mb{2}}{q}} \\
-    \prftree[r]{\rle{let}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta, \thyp{x}{A}{}}{p}{\mhole{e}}{B}{q}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
-      {\mhasty{H}{\Gamma}{p}{\letexpr{x}{\mhole{a}}{\mhole{e}}}{B}{q}} \qquad
-    % \prftree[r]{\rle{blk}}
-    %   {\mhaslb{H}{\Gamma}{\mhole{t}}{\llhyp{\ell}{\cdot}{A}}}
-    %   {\mhasty{H}{\Gamma}{0}{\lbsplice{\ell}{A}{\mhole{t}}}{A}{\varnothing}} 
-      \\
-    \prftree[r]{\rle{let2}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{\mhole{e}}{C}{q}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
-      {\mhasty{H}{\Gamma}{p}{\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}}{C}{q}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa terms with holes}
-  \Description{Typing rules for isotope-SSA terms with holes}
-  \label{fig:ssa-term-holes-typing}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}    
-    \prftree[r]{\rle{hole-blk}}
-      {\blkhole{\lhole{T}}{\Delta}{\ms{L}} \in H}
-      {\cwk{\Gamma}{\Delta}}
-      {\mhaslb{H}{\Gamma}{\lhole{T}}{\ms{L}}} 
-      \qquad
-    \prftree[r]{\rle{br}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{\tint}}
-      {\mhaslb{H}{\Gamma}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
-      \\
-    \prftree[r]{\rle{ite}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhasty{H}{\Xi}{1}{\mhole{e}}{\mb{2}}{\tint}}
-      {\mhaslb{H}{\Delta}{\mhole{s}}{\ms{L}}}
-      {\mhaslb{H}{\Delta}{\mhole{t}}{\ms{L}}}
-      {\mhaslb{H}{\Gamma}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\
-    \prftree[r]{\rle{let-blk}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhaslb{H}{\Delta, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{q}}
-      {\mhaslb{H}{\Gamma}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
-    \prftree[r]{\rle{let2-blk}}
-      {\csplits{\Gamma}{\Delta}{\Xi}}
-      {\mhaslb{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{\mhole{t}}{\ms{L}}}
-      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A \otimes B}{q}}
-      {\mhaslb{H}{\Gamma}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
-    \prftree[r]{\rle{where}}
-      {\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
-      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
-      {\mhaslb{H}{\Gamma}{\ewhere{\mhole{t}}{\mhole{L}}}{\ms{K}}} 
-      \qquad
-    \prftree[r]{\rle{hole-cfg}}
-      {\cfghole{\lhole{W}}{\ms{L}}{\ms{K}} \in H}
-      {\lwk{\ms{M}}{\ms{K}}}
-      {\mlhaslb{\ms{L}}{H}{\lhole{W}}{\ms{K}}}
-      \\
-    \prftree[r]{\rle{nil-br}}
-      {\mlhaslb{H}{\ms{L}}{\cdot}{\ms{L}}} \qquad
-    \prftree[r]{\rle{cons-br}}
-      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
-      {\mhaslb{H}{\Gamma, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
-      {\mlhaslb{H}{\ms{L}}{\mhole{L}, \lwbranch{\ell}{x: A}{\mhole{t}}}{\ms{K}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa blocks with holes}
-  \Description{Typing rules for isotope-SSA blocks with holes}
-  \label{fig:ssa-block-holes-typing}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{H: \ms{Hole} \to \ms{MExpr} \sqcup \ms{MBlock} \sqcup \ms{MCfg}}
-    \\
-    \prftree[r]{\rle{rw-nil}}
-      {\isrw{\mc{H}}{H}{\cdot}}
-    \qquad
-    \prftree[r]{\rle{rw-tm}}
-      {\isrw{\mc{H}}{I}{H}}
-      {\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}
-      {\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}
-    \\
-    \prftree[r]{\rle{rw-blk}}
-      {\isrw{\mc{H}}{I}{H}}
-      {\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}
-      {\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
-    \qquad
-    \prftree[r]{\rle{rw-cfg}}
-      {\isrw{\mc{H}}{I}{H}}
-      {\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}
-      {\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
-  \end{gather*}
-  \caption{Typing rules for \isotopessa rewrites}
-  \Description{Typing rules for isotope-SSA rewrites}
-  \label{fig:rewrite-typing}
-\end{figure}
-
-The typing rules for terms, blocks, and CFGs with holes are exactly the same as
-the corresponding rules without holes, modulo the additional rules
-\rle{hole-tm}, \rle{hole-blk}, and \rle{hole-cfg} to type the holes themselves;
-they are given in full in Figures \ref{fig:ssa-term-holes-typing} and
-\ref{fig:ssa-block-holes-typing}. This formalism allows us to state the
-\textit{syntactic rewriting} theorem, which says, in essence, that filling holes
-with valid terms yields valid terms.
-\begin{theorem}[Rewriting]
-  Given \(\isrw{\mc{H}}{H}{I}\), we have
-  \begin{itemize}
-    \item If \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), then \(\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}\)
-    \item If \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), then \(\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}\)
-  \end{itemize}
-  Furthermore, we have that 
-  \begin{itemize}
-    \item \(\mhasty{\cdot}{\Gamma}{p}{a}{A}{q} \iff \hasty{\Gamma}{p}{a}{A}{q}\)
-    \item \(\mhaslb{\cdot}{\Gamma}{t}{\ms{L}} \iff \haslb{\Gamma}{t}{\ms{L}}\)
-  \end{itemize}
-\end{theorem}
-
-\subsection{Semantics}
-
-We can interpret a hole context semantically as the \textit{set} of all possible
-tuples of morphisms which would typecheck in place of the holes in the context,
-as specified in Figure \ref{fig:holes-semantics}.
-\begin{figure}
-  \begin{gather*}
-    \dnt{\tyhole{\lhole{X}}{\Gamma}{p}{A}{q}} 
-      = \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})
-    \qquad
-    \dnt{\blkhole{\lhole{T}}{\Gamma}{\ms{L}}} 
-      = \mc{C}_0(\dnt{\Gamma}, \dnt{\ms{L}})
-      \\
-    \dnt{\cfghole{\lhole{W}}{\ms{L}}{\ms{K}}}
-      = \mc{C}_0(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}}) 
-      \\
-    \dnt{H, M} = \dnt{H} \times \dnt{M}
-    \qquad
-    \dnt{\cdot} = \mb{1}
-      \\
-    \boxed{\dnt{M \in H}: \dnt{H} \to \dnt{M}}
-      \qquad
-    \dnt{M \in H, M \in H} 
-      = \pi_1
-    \qquad
-    \dnt{M \in H, N} = \pi_0;\dnt{M \in H}
-  \end{gather*}
-  \caption{Semantics for \isotopessa holes}
-  \Description{Semantics for isotope-SSA holes}
-  \label{fig:holes-semantics}
-\end{figure}
-We can now quite easily give semantics for terms, blocks, and CFGs with holes
-which simply correspond to "filling in" any holes with the appropriate morphism
-from the hole context, with all other rules interpreted identically to their
-regular variants: this is done in Figures \ref{fig:ssa-term-holes-semantics} and
-\ref{fig:ssa-block-holes-semantics}.
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}}
-      : \dnt{H} \to \mc{C}_p^q(\dnt{\Gamma}, \dnt{A})}
-      \\
-    \sorry
-      \\
-    \boxed{
-      \dnt{\tyhole{\mhole{X}}{\Gamma}{a}{A}{q} \in H}
-      : H \to \mc{C}^q_p(\dnt{\Gamma}, \dnt{A})
-    }
-  \end{gather*}
-  \caption{Semantics for \isotopessa terms with holes}
-  \Description{Semantics for isotope-SSA terms with holes}
-  \label{fig:ssa-term-holes-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
-      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})}
-      \\
-    \sorry
-      \\  
-    \boxed{
-      \dnt{\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
-      : \dnt{H} \to \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
-      \\
-    \sorry
-  \end{gather*}
-  \caption{Semantics for \isotopessa blocks with holes}
-  \Description{Semantics for isotope-SSA blocks with holes}
-  \label{fig:ssa-block-holes-semantics}
-\end{figure}
-
-\begin{figure}
-  \begin{gather*}
-    \boxed{
-      \dnt{\isrw{\mc{H}}{I}{H}}
-      : \dnt{I} \to \dnt{H} 
-    }
-    \\
-    \dnt{\isrw{\mc{H}}{I}{\cdot}}\;i = ()
-    \qquad
-    \dnt{\isrw{\mc{H}}{I}{H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}}}\;i
-    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhasty{I}{\Gamma}{p}{\mc{H}(\lhole{X})}{A}{q}}\;i)
-    \\
-    \dnt{\isrw{\mc{H}}{I}{H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}}}
-    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mhaslb{I}{\Gamma}{\mc{H}(\lhole{T})}{\ms{L}}}\;i)
-    \\
-    \dnt{\isrw{\mc{H}}{I}{H, \cfghole{\lhole{L}}{\ms{L}}{\ms{K}}}}
-    = (\dnt{\isrw{\mc{H}}{I}{H}}\;i, \dnt{\mlhaslb{I}{\ms{L}}{\mc{H}(\lhole{W})}{\ms{K}}}\;i)
-  \end{gather*}
-  \caption{Semantics for \isotopessa rewrites}
-  \Description{Semantics for isotope-SSA rewrites}
-  \label{fig:rewrite-semantics}
-\end{figure}
-Of course, as for terms without holes, we have a coherence lemma:
-\begin{lemma}[Rewriting Coherence] \
-  \begin{itemize}
-    \item Given any two derivations \(D_1: \mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(D_2: \mhasty{\Gamma}{p}{H}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
-    \item Given any two derivations \(D_1, D_2: \isrw{\mc{H}}{I}{H}\), \(\dnt{D_1} = \dnt{D_2}\)
-  \end{itemize}
-\end{lemma}
-We may finally state the semantic rewriting theorem as follows:
-\begin{theorem}[Semantic Rewriting]
-  Given \(\isrw{\mc{H}}{I}{H}\), we have that
-  \begin{itemize}
-    \item For all \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), \(m \in \dnt{I}\), we have
-    \[
-      \dnt{\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}}(m)
-      = \dnt{\mhasty{H}{\Gamma}{p}{a}{A}{q}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
-    \]
-    \item For all \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), \(m \in \dnt{I}\), we have
-    \[
-      \dnt{\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}}(m)
-      = \dnt{\mhaslb{H}{\Gamma}{t}{\ms{L}}}(\dnt{\isrw{\mc{H}}{I}{H}}(m))
-    \]
-  \end{itemize}
-
-  Furthermore, we have that
-  \[
-    \dnt{\mhasty{\cdot}{\Gamma}{p}{a}{A}{q}}() = \dnt{\hasty{\Gamma}{p}{a}{A}{q}}
-    \qquad
-    \dnt{\mhaslb{\cdot}{\Gamma}{t}{\ms{L}}}() = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
-  \]
-\end{theorem}
-This justifies congruence based optimizations, and in particular E-graph
-rewriting.
-
-\subsection{Equational Theory}
-
-\TODO{\(\beta\), \(\eta\), etc}
-
-\section{Refinement}
-
-In general, we often want to justify optimizations which \textit{refine} a
-program by, for example, reducing the set of allowable behaviours, rather than
-simply rewrite the program to an equivalent one. One place where refinement
-pops up quite often is in the treatment of undefined values and undefined 
-behaviour; for example, in a language where reading a dangling pointer is UB, 
-the transformation 
-\begin{equation*}
-\begin{BVerbatim}[baseline=c, fontsize=\small]
-  let x = read(p);
-  ret 5
-\end{BVerbatim}
-  \qquad \qquad \implies \qquad \qquad
-\begin{BVerbatim}[baseline=c, fontsize=\small]
-  ret 5
-\end{BVerbatim}
-\end{equation*}
-is valid, but the reverse is \textit{not}, since that would \textit{introduce},
-rather than \textit{remove}, potential UB. These optimizations are particularly
-important when operations are allowed to generate undefined values, in which
-case, even simple rewrites such as \(a \times 0 \to 0\) and \(a - a \to 0\) are
-refinements rather than equalities. In this appendix section, we will generalize
-our semantics for rewriting from section \ref{apx:rewriting} to support
-reasoning about refinements in a compositional manner.
-
-\subsection{Poset-Enriched Categories}
-
-To reason about refinement, we need to introduce the definition of a
-\textit{poset-enriched category}:
-\begin{definition}[Poset-Enriched Category]
-  A category \(\mc{C}\) is \textbf{poset-enriched} if, for all objects \(A, B
-  \in |\mc{C}|\), \(\mc{C}(A, B)\) is equipped with a partial order \(\leq\)
-  compatible with composition, i.e. \(f \leq g \implies f;h \leq g;h, h';f \leq
-  h';g\). A \textbf{poset-enriched premonoidal category} is a poset-enriched
-  category such that the partial order on morphisms is compatible with
-  whiskering, i.e. \(f \leq g \implies f \otimes C \leq g \otimes C, C \otimes f
-  \leq C \otimes g\). A \textbf{poset-enriched substructural category} is a
-  poset-enriched premonoidal category such that the subcategories
-  \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) are downward-closed under \(\leq\). 
-\end{definition}
-Note that every (premonoidal/substructural) category \(\mc{C}\) can be
-considered poset-enriched under the discrete partial order \((\leq) = (=)\). One
-useful example of a poset-enriched substructural category is the category of
-relations from section \ref{ssec:substruct}, taking \(f \leq g \iff \forall a
-\in A, f(a) \subseteq g(a)\).
-
-\subsection{Semantic Refinement}
-
-\TODO{refinement for rewrites}
-
-\TODO{semantic refinement theorem}
-
-\subsection{Equational Theory}
-
-\TODO{this}
-
-\section{Normalisation}
-
-\TODO{this}
-
-\subsection{Strict SSA form}
-
-\TODO{this}
-
-\subsection{Structured Control-Flow}
-
-\TODO{this}
-
-\section{Loop Optimizations}
-
-\TODO{this}
-
-\subsection{Unrolling and Vectorization}
-
-\TODO{this}
-
-\subsection{Dead-code Elimination}
-
-\TODO{this}
-
-\subsection{Loop Hoisting}
-
-\TODO{this}
-
-\subsection{Loop Fusion/Fission}
-
-\TODO{this}
-
-\section{E-Graph Optimizations}
-
-\TODO{this}
-
-\subsection{Floating SSA}
-
-\TODO{this}
-
-\subsection{Whole-program rewriting}
-
-\TODO{this}
-
-% \section{State-Passing Category}
-%
-% We can further use this example to get a substructural \textit{monoidal} category by considering a primitive form of separation logic: let's introduce a grammar of predicates \(\varphi, \psi ::= \top, \ms{S}, \bot\) for programs which cannot access state, can access state, and are invalid, respectively.
-The separating conjunction operator on these predicates is then given by
-% \begin{equation}
-%   \top * \varphi = \varphi * \top = \varphi
-%   \qquad
-%   \bot * \varphi = \varphi * \bot = \bot
-%   \qquad
-%   \ms{S} * \ms{S} = \bot
-% \end{equation}
-% We'll then define a category \(\ms{SRel}\) with objects of the form \((A, \varphi)\) and morphisms
-% \begin{equation}
-%   \begin{gathered}
-%     \boxed{
-%       \ms{SRel}((A, \varphi), (B, \psi)) 
-%       \subseteq A \to \ms{S} \to \mc{P}(A \times S) 
-%     }
-%     \\
-%     \begin{aligned}
-%       \ms{SRel}((A, \top), (B, \top)) &= \{f \in A \to S \to \mc{P}(A \times S) \mid \exists g. f = \lambda a, s. g(a) \times \{s\}\} \\
-%       \ms{SRel}((A, \ms{S}), (B, \ms{S})) &= A \to S \to \mc{P}(A \times S) \\
-%       \ms{SRel}((A, \varphi), (B, \psi)) &= \{\lambda a, s. \varnothing\} \quad \text{otherwise}
-%     \end{aligned}
-%   \end{gathered}
-% \end{equation}
-% Note that it is precisely the \textit{central} morphisms \(\ms{C}_0(A, B)\) which lie in \(\ms{SRel}((A, \top), (B, \top))\).
-We can then define the tensor product on objects as \((A, \varphi) \otimes (B, \psi) = (A \otimes B, \varphi * \psi)\), and, for \(f \in \mathsf{SRel}((A, \varphi), (B, \psi))\), and tensor product functors
-% \begin{equation}
-%   \begin{aligned}
-%   f \otimes (C, \phi) &= f \otimes C,
-%   & (C, \phi) \otimes f &= C \otimes f
-%   && \text{if}\; \varphi * \phi \neq \bot, \psi * \phi \neq \bot \\
-%   f \otimes (C, \phi) &= \lambda a, s. \varnothing,
-%   & (C, \phi) \otimes f &=  \lambda a, s. \varnothing
-%   && \text{otherwise}
-%   \end{aligned} 
-% \end{equation}
-% Lifting associators, unitors, and symmetries from \(\mc{C}_0\) in the obvious manner (setting them to \(\lambda a, s.
-\varnothing\) where necessary), we can verify that \(\ms{SRel}\) is a \textit{premonoidal} category.
-we can see that it is in fact \textit{monoidal} by nothing that, for any \(f, g\), \(f \ltimes g = f \rtimes g\) since either \(f\) is central, \(g\) is central, or their product is the empty morphism \(\lambda a, s.
-\varnothing\).
-Hence, to get a \textit{substructural} monoidal category, we simply need to define:
-% \begin{equation}
-%   \begin{gathered}
-%   \ms{Aff}(\ms{SRel}) = \ms{Rel}(\ms{SRel}) = \{(A, \top) | A \in \ms{Set}\} \neq |\ms{SRel}|
-%   \\
-%   \begin{aligned}
-%     \ms{SRel}^\varnothing((A, \phi), (B, \psi))
-%     &= \ms{SRel}((A, \phi), (B, \psi)) \\
-%     \ms{SRel}^{\{\ms{a}\}}((A, \phi), (B, \psi)) 
-%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \geq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
-%     \ms{SRel}^{\{\ms{r}\}}((A, \phi), (B, \psi)) 
-%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \leq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
-%     \implies \ms{SRel}^{\{\ms{a}, \ms{r}\}}((A, \phi), (B, \psi)) 
-%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| = 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
-%   \end{aligned}
-%   \end{gathered}
-% \end{equation}
-% We can verify that the associators, unitors, and symmetry lie in \(\ms{SRel}^{\{\ms{a}, \ms{r}\}}\) and that all other desired axioms are satisfied.
-
-% \TODO{relate separation logic trick to \cite{promonad}, \cite{linear-state-usage}, \cite{mellies-ftrs}}
-
-% \TODO{clean up, and generalize to arbitrary premonoidal categories (w/ initial object? how to deal w/ coproducts?)}
-
 \end{document}
